---
title: "BLAST"
subtitle: "What are the 12 genes and what do they do?"
author: "Sarah Tanja"
date: 2025-06-17
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

# Background

We identified 44 genes that were broadly impacted by leachate. Here we want to know what those genes do!
Those 44 genes grouped in 

Why is goseq different from other enrichment tools? Typical enrichment methods (like Fisher's exact test or hypergeometric tests) assume that every gene has an equal probability of being selected as a DEG. But in RNA-Seq longer genes are more likely to be detected as DEGs just because they accumulate more reads. `goseq` corrects for this bias.

## Inputs

-   `LRT_significant_genes_combined_wald.csv` A dataframe of the 12 genes and their LRT and Wald test statistics
-   functional annotation (version 3) from: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz (EggNog)
-   functional annotation http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz (KEGG)

## Outputs

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  dpi = 600,
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

## Install packages
```{r}
if ("Biostrings" %in% rownames(installed.packages()) == 'FALSE') install.packages('Biostrings')
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')
if ("goseq" %in% rownames(installed.packages()) == 'FALSE') install.packages('goseq')
```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GenomicRanges")
BiocManager::install("goseq")
BiocManager::install("GO.db")
BiocManager::install("rrvgo")
```

## Load packages
```{r}
library(tidyverse)
library(GenomicRanges)
library(Biostrings)
library(goseq)
library(rrvgo)
library(GO.db)
```

## Read in gene lists and counts
```{r}
all_genes <- read.csv("../../output/06_deg/interactions/all_genes.csv")
sig_genes <- read.csv("../../output/06_deg/interactions/sig_genes.csv")
sig_counts <- read.csv("../../output/06_deg/interactions/sig_counts.csv", row.names="gene_id", check.names = FALSE)
```


```{r}
all_genes <- all_genes %>% 
  select(gene_id)

head(all_genes)

write_csv(all_genes, "../../output/06_deg/interactions/all_genes.csv")
```

```{r}
sig_genes <- sig_genes %>% 
  select(!X)

head(sig_genes) 

write_csv(sig_genes, "../../output/06_deg/interactions/sig_genes.csv")
```

```{r}
sig_counts <- sig_counts %>% 
  select(!1)

head(sig_counts)

sig_counts_save <- sig_counts %>% 
 rownames_to_column("gene_id")

write_csv(sig_counts_save, "../../output/06_deg/interactions/sig_counts.csv")
```

## Import fasta

```{r}
# Load your transcriptome
all_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")  

# Filter for the 44 significant interaction genes
sig_seqs <- all_seqs[names(all_seqs) %in% sig_genes]

# Save to new FASTA file
writeXStringSet(sig_seqs, "../../output/07_blast/significant_genes.fasta")
```

## Download annotation & enrichment files

```{r, engine='bash'}
cd ../../input/annotations

wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz
wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz
```

Unzip

```{r, engine='bash'}
cd ../../input/annotations
gunzip *.gz
```

## Import to R environment

```{r}
# Read in the file (it will include "#query" as a column name)
EggNog_anno <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")
```

rename "#query" to `gene_id`

```{r}

EggNog_anno <- EggNog_anno %>%
  rename(gene_id="#query")

nrow(EggNog_anno)
names(EggNog_anno)
```

> This functional annotation has 24,072 genes. This is 44% of total protein predicted genes described in the publication associated with this annotation. https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755

# Annotate significant genes

Match up genes in sig_genes dataframe to annotation dataframe

```{r}
sig_genes_anno <- inner_join(sig_genes, EggNog_anno, by = "gene_id")

length(sig_genes_anno)
```
::: callout-note
34 of the 44 significant genes were annotated! The annotation file `sig_genes_anno` dataframe now only contains genes that were significant and that have annotation information.
:::

Get gene length information.

```{r}
#import file
gff <- rtracklayer::import("../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3") #if this doesn't work, restart R and try again 

transcripts <- subset(gff, type == "transcript") #keep only transcripts 

transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE) #extract length information

transcript_lengths <- width(transcripts_gr) #isolate length of each gene

gene_id<-transcripts_gr$ID #extract list of gene id 

lengths<-cbind(gene_id, transcript_lengths)

lengths<-as.data.frame(lengths) #convert to data frame
```

Add in length to the annotation file.

```{r}
sig_genes_anno <- sig_genes_anno %>% left_join(lengths, by = "gene_id")

sig_genes_anno <- sig_genes_anno %>% select(!length)
```

Format GO terms to remove dashes and quotes.

```{r}
#remove dashes, remove quotes from columns and separate by semicolons (replace , with ;) in KEGG_ko, KEGG_pathway, and Annotation.GO.ID columns
head(sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub(",", ";", sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub('"', "", sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub("-", NA, sig_genes_anno$GOs)
```

# Conduct functional enrichment

Conduct functional enrichment using goseq and rrvgo. Compare against all genes detected in the dataset that we have annotation information for.

## Biological process

Conduct at biological process level. Filter for adjusted p-value \<0.05.

```{r}
# Define your DEG list — the genes you're testing
ID.vector <- sig_genes_anno$gene_id  # Replace with your actual gene vector


# All genes detected in the dataset with annotation info
ALL.vector <- all_genes$gene_id # gene_ids for all 12565 genes that we used to make our DESeq2 object
ALL.vector <- ALL.vector[ALL.vector %in% EggNog_anno$gene_id] #filtered down to only genes that are annotated

length(ALL.vector)
```
::: callout-note
9563 out of the 12564 genes in our dataset are annotated
:::

```{r}
lengths_anno <- lengths %>% 
  filter(gene_id %in% all_genes$gene_id) %>% 
  filter(gene_id %in% EggNog_anno$gene_id)

LENGTH.vector <- as.integer(lengths_anno$transcript_length)
names(LENGTH.vector) <- ALL.vector  # Needed for bias correction

# Gene to GO term mapping
GO.terms <- sig_genes_anno %>%
  dplyr::select(gene_id, GOs)

# Format GO terms to one term per row
split <- strsplit(as.character(GO.terms$GOs), ";")
split2 <- data.frame(
  gene = rep.int(GO.terms$gene, sapply(split, length)),
  GOs = unlist(split)
)

GO.terms <- split2

# Construct named gene vector: 1 = DEG, 0 = not DEG
gene.vector <- as.integer(ALL.vector %in% ID.vector)
names(gene.vector) <- ALL.vector

# Null probability weighting function (bias for gene length)
# nullp is a function from nadiadavidson/goseq
pwf <- nullp(gene.vector, bias.data = LENGTH.vector)
```

About goseq() command:

| Argument                       | Meaning                                                                                                                                                      |
|---------------------------|--------------------------------------------------------|
| `pwf`                          | A **probability weighting function** object from `nullp()`. It contains the DEG status of all genes (1 = DEG, 0 = not DEG), plus gene length bias info.      |
| `gene2cat`                     | A data frame or list mapping **genes to functional categories** (like GO terms). In your case, it's a long-format table of gene-GO pairs.                    |
| `test.cats = "GO:BP"`          | This limits the enrichment to only **biological process** GO terms (you could also use `"GO:MF"` or `"GO:CC"` for molecular function or cellular component). |
| `method = "Wallenius"`         | This uses the **Wallenius noncentral hypergeometric distribution**, which accounts for sampling bias (like gene length).                                     |
| `use_genes_without_cat = TRUE` | Allows the method to consider genes that **do not have a GO annotation** (prevents them from being excluded by default).                                     |

`goseq()` returns a data frame where each row is a GO term, with:

-   The GO ID

-   Category (BP, MF, or CC)

-   Over-representation p-value (for DEGs)

-   Under-representation p-value (less commonly used)

-   Number of DEGs in that GO term

-   Total number of genes in that GO term

-   Other metadata (like GO term name)

```{r}
# Run GO enrichment for biological process terms
GO.bp <- goseq(
  pwf,
  gene2cat = GO.terms,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

# Format and filter results
GO.bp <- GO.bp %>% 
  arrange(over_represented_pvalue)

# Use the Benjamini-Hochberg method to control false discovery rate 
GO.bp$bh_adjust <- p.adjust(GO.bp$over_represented_pvalue, method = "BH")

GO.bp <- GO.bp %>%
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

# Optional: write to file
write_csv(GO.bp, file = "../../output/07_blast/functional_enrichment_biological_process_goseq.csv")
```


## Reduce redundancy
```{r}
# Filter for biological process terms with valid p-values
GO_BP <- GO.bp %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>%
  arrange(bh_adjust)

# Calculate semantic similarity matrix
simMatrix <- calculateSimMatrix(
  GO_BP$GOterm,
  orgdb="org.Ce.eg.db", #c. elegans database
  ont = "BP",
  method = "Rel"
)

# Create named scores vector: -log10 adjusted p-values
scores <- setNames(-log10(go_results$bh_adjust), go_results$GOterm)

# Reduce GO terms
reducedTerms <- reduceSimMatrix(
  simMatrix,
  scores = scores,
  threshold = 0.7,
  orgdb = "org.Ce.eg.db"
)

# Keep only the GO terms in reduced list
go_results <- go_results %>%
  filter(GOterm %in% reducedTerms$go)

# Add parent term (representative term)
go_results$ParentTerm <- reducedTerms$parentTerm[match(go_results$GOterm, reducedTerms$go)]

# Done! You can now save or plot
write_csv(go_results, "reduced_goseq_results.csv")
```

