---
title: "Step 4: Create a Count Matrix"
subtitle: "Using a python script`"
author: "Sarah Tanja"
date: 11/14/2024
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output))        
```

# Background
> DESeq2 and edgeR are two popular Bioconductor packages for analyzing differential expression, which take as input a matrix of read counts mapped to particular genomic features (e.g., genes). We provide a Python script (prepDE.py, or the Python 3 version:prepDE.py3 ) that can be used to extract this read count information directly from the files generated by StringTie (run with the -e parameter).

> prepDE.py derives hypothetical read counts for each transcript from the coverage values estimated by StringTie for each transcript, by using this simple formula: reads_per_transcript = coverage \* transcript_len / read_len

# Goals
Our goal is to generate a count matrix from the gtf files. The count matrix will then be used as input for differential gene expression analysis. 

## Inputs
- *.gtf files for each sample 
  - `../../output/03_align/stringtie/*gtf`
  
## Outputs
- gtf_list.txt
  - `../../output/04_count/gtf_list.txt`
- gene_count_matrix.csv
  - `../../output/04_count/gene_count_matrix.csv`
- transcript_count_matrix.csv
  - `../../output/04_count/transcript_count_matrix.csv`

# Generate count matrix
## Make gtf_list.txt

List the gtf files at this filepath.
```{r, engine='bash'}
ls ../../output/03_align/stringtie/*gtf 
```

How many *.gtf files are there?
```{r, engine="bash"}
cd ../../output/03_align/stringtie 
find . -type f -name "*.gtf" | wc -l
```
There should be 63 files for 63 samples!

Save the list of basenames (sample_names) and their respective file paths to gtf_list.txt 
```{r, engine='bash'}
ls ../../output/03_align/stringtie/*.gtf | awk -F'/' '{path=$0; file=$NF; gsub(".gtf$", "", file); print file "\t" path}' > ../../output/04_count/gtf_list.txt
```

Look at the gtf_list.txt we just made to make sure there is a basename and a filepath for each sample
```{r, engine='bash'}
cat ../../output/04_count/gtf_list.txt
```

## Run python script
Run the python script `prepDR.py3`, which takes the basenames, filepaths to gtf files and churns out a gene count matrix and a transcript count matrix. 
```{r, engine='bash'}
python3 /home/shared/stringtie-2.2.1.Linux_x86_64/prepDE.py3 \
-i ../../output/04_count/gtf_list.txt \
-g ../../output/04_count/gene_count_matrix.csv \
-t ../../output/04_count/transcript_count_matrix.csv
```

# Summary & next steps

We successfully made our gene count matrix! Next up is to do initial exploration of it in Step 5 to see broad trends in gene counts as they relate to our experimental variables: embryonic stage and leachate level.  
