---
title: "Step 5: Final figure of PCA & Heatmap of full gene set"
subtitle: "Samples group largely by stage rather than leachate"
author: "Sarah Tanja"
date: 11/22/2025
date-modified: today
bibliography: "../../references.bib"
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin

editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = ""
  ) # Prevents appending '##' to beginning of lines in code output))        
```

# Background

## Load libraries
```{r}
library(colorspace)
library(shinyjs) # for colorspace hclwizard
library(ggsidekick) # for sean andersons theme_sleek()
library(ComplexHeatmap)
library(tidyverse)
library(patchwork)
```

## Load inputs
```{r}
mat <- readRDS("../../output/05_explore/vst_matrix.rds")
pca <- read_csv("../../output/05_explore/pca_500_or.csv")
pheat_df <- readRDS("../../output/05_explore/pheat_annotation_df.rds")
```

## Custom ggplot theme
```{r}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_blank(),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      plot.title = element_text(
      size  = 14,
      color = "grey40",
      hjust = 0.5,    # center
      margin = margin(b = 20) # add space BELOW title
      ),      ,       
      plot.subtitle = element_text(
      size   = 10,       # smaller than title
      colour = "grey60", # lighter grey
      hjust  = 0.5 ),      # center
      margin = margin(b = 40) # add space BELOW title
      )
}

```


## Set colorschemes

```{r}
#colorspace::hclwizard()
```
### Leachate
Leachate options
Ocean environmental watery tones
blue-purple
```{r}
leachate_0 <- c("#B0F4FA", "#75C165", "#A96C00", "#8B0069")
leachate_1 <- c("#D5ECC0", "#6ACFBA", "#399AC2", "#80146E")
leachate_2 <- c("#E4EBD6", "#8CC4AD", "#3E909F", "#4E4E7A")
leachate_3 <- c("#DAE6FF", "#BBC7FF", "#9CA9FF", "#7D8BFF")

leachate.colors <- c(control = "#AEF1FF", 
                low = "#BBC7FF",
                mid = "#7D8BFF", 
                high = "#592F7D")
```

### Stage
Earthy stony coral tones
yellow-orange-red
```{r}
stage_0 <- c("#DCC14F", "#D89500", "#C86600")
stage_1 <- c("#DCC14F", "#9F3D2F", "#000000")
stage_2 <- c("#A2223C", "#E6AA83",  "#D9685B")
stage_3 <- c("#D9685B", "#D89500",  "#9F3D2F")
stage_4 <- c("#FFE362", "#EBA600", "#8D5300")

stage.colors <- c(cleavage = "#E6AA83", 
                  prawnchip = "#D9685B", 
                  earlygastrula = "#A2223C")
```

### Spawn night

```{r}
night_0 <- c("#FDE725", "#440154", "#21918C")
night_1 <- c("#FFCEF4", "#578B21", "#201158")
night_2 <- c("#E3FAA5", "#B16776", "#1E2440")

night.colors <- c(July_6th = "#E3FAA5", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")
```


```{r}
# Annotation colors for pheatmap
ann_colors = list(
  leachate = leachate.colors,
  stage = stage.colors,
  night = night.colors
)
```

# PCA plot of full gene set
```{r}
pca_flip <- pca %>% 
  mutate(stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula")),
         leachate = factor(leachate, levels = c("control", "low", "mid", "high")))%>%
  dplyr::mutate(PC1 = -PC1)
```


Extract the Percent Variance for the X and Y axis labels
```{r}
pca_comp <- prcomp(t(mat))         # samples in rows

percentVar <- (pca_comp$sdev^2) / sum(pca_comp$sdev^2) * 100

pc1_var <- round(percentVar[1], 1)
pc2_var <- round(percentVar[2], 1)
```


```{r, fig.height= 4, fig.width = 6}
pca_plot <- ggplot(pca_flip, aes(PC1, PC2, color = stage, shape = leachate, label = name)) +
  geom_point(size = 1,  stroke = 1) +
  xlab(paste0("PC1: ", pc1_var, "% variance")) +
  ylab(paste0("PC2: ", pc2_var, "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = stage.colors) +
  stat_ellipse(aes(group = stage, color = stage), linewidth = 0.4, show.legend = FALSE) + 
  labs(
    title = "Variation of gene expression patterns among samples",
    subtitle = "PCA of 500 most variable genes • Each point is a sample",
  )+
  theme_sleek_axe() 

pca_plot
```


# Heatmap of full gene set

Annotations for our heatmap
```{r}

ha_top <- HeatmapAnnotation(
  stage = pheat_df$stage,
  col   = list(stage = ann_colors$stage),
  annotation_name_side = "right"  # label "stage" on the left side
)

ha_bottom <- HeatmapAnnotation(
  `spawn night` = pheat_df$spawn_night,
  leachate      = pheat_df$leachate,
  col = list(
    `spawn night` = ann_colors$night,
    leachate      = ann_colors$leachate
  ),
  annotation_name_side = "right"
)
```

Scale the matrix to be z scores! And then clamp extreme values for better visualization. 
```{r}
mat_scaled <- t(scale(t(mat)))     # row-wise z-scaling
mat_scaled_clamped <- pmin(pmax(mat_scaled, -5), 5)  # clamp extreme values to -5 and 5
```

```{r, fig.height = 8, fig.width= 6}
# 1. build column dendrogram once
col_hc <- hclust(
  dist(t(mat_scaled_clamped), method = "euclidean"),
  method = "complete"
)

# 2. flip it
col_dend_flipped <- rev(as.dendrogram(col_hc))

set.seed(11222025)
Heatmap(
  mat_scaled_clamped,
  clustering_distance_rows    = "euclidean",
  clustering_method_rows      = "complete",
  cluster_columns             = col_dend_flipped,
  name = "zscore of gene expression",
  border = NA,
  column_dend_gp  = gpar(col = "grey50", lwd = 0.7),
  row_names_gp    = gpar(col = "grey50", fontsize = 8),
  column_names_gp = gpar(col = "grey50", fontsize = 8),
  col = rev(colorspace::divergingx_hcl(n = 200, palette = "Roma")),
  cluster_column_slices = TRUE,
  show_row_dend    = FALSE,
  show_column_dend = TRUE,
  top_annotation    = ha_top,
  bottom_annotation = ha_bottom,
  show_row_names    = FALSE,
  show_column_names = TRUE
)
```
> Expression values were transformed using variance-stabilizing transformation (VST) and then standardized on a per-gene basis (z-score transformation). For each gene, the mean expression across samples was subtracted, and the standard deviation was used to scale its expression profile. Thus, a value of 0 indicates average expression for that gene across all samples, positive values indicate above-average expression, and negative values indicate below-average expression. This allows visual comparison of expression patterns, not absolute expression levels between genes.

# Patchwork 

```{r, fig.width = 6, fig.height = 8}
#| include: false
# 1. Your heatmap object (DON'T draw it yet)
set.seed(11222025)
ht <- 
Heatmap(
  mat_scaled_clamped,
  clustering_distance_rows    = "euclidean",
  clustering_method_rows      = "complete",
  cluster_columns             = col_dend_flipped,
  name = "zscore",
  border = NA,
  column_dend_gp  = gpar(col = "grey50", lwd = 0.7),
  row_names_gp    = gpar(col = "grey50", fontsize = 8),
  column_names_gp = gpar(col = "grey50", fontsize = 8),
  col = rev(colorspace::divergingx_hcl(n = 200, palette = "Roma")),
  cluster_column_slices = TRUE,
  show_row_dend    = FALSE,
  show_column_dend = TRUE,
  top_annotation    = ha_top,
  bottom_annotation = ha_bottom,
  show_row_names    = FALSE,
  show_column_names = TRUE
)

# 2. Grab the drawn heatmap as a grob
ht_grob <- grid.grabExpr(draw(ht, 
                              heatmap_legend_side    = "right",
                              annotation_legend_side = "right",
                              merge_legends          = TRUE))

# 3. Wrap it so patchwork can use it
heatmap_wrapped <- patchwork::wrap_elements(ht_grob)

```

Caption 
```{r}
library(ggplot2)
library(patchwork)
library(stringr)

cap_text <- paste(
  "A. Principal Component Analysis (PCA) of VST-normalized gene expression across all samples.",
  "PC1 explains 53.9% of the total variance and primarily separates samples by embryonic stage,",
  "while PC2 explains 18.5% of the variance, accounting for 72.4% of total sample variance.",
  "Each point represents an individual sample, positioned according to its overall expression profile.",
  "B. Heatmap of z-scored variance-stabilized gene count matrix. Values were transformed using",
  "variance-stabilizing transformation (VST) and then standardized on a per-gene basis (z-score).",
  "For each gene, the mean expression across samples was subtracted and the standard deviation",
  "was used to scale its expression profile. A value of 0 indicates no change in expression for",
  "that gene across all samples; positive values indicate above-average expression and negative",
  "values indicate below-average expression. This allows visual comparison of expression patterns",
  "across samples (columns), not absolute expression levels between genes (rows).",
  sep = " "
)

# wrap to, say, 100 characters per line (tweak as needed)
cap_wrapped <- str_wrap(cap_text, width = 100)


caption_plot <- ggplot() +
  annotate(
    "text",
    x     = 0, y = 1,
    label = cap_wrapped,   # << wrapped text
    hjust = 0, vjust = 1,
    size  = 3,
    lineheight = 0.95
  ) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_void() +
  theme(
    plot.margin = margin(t = 5, r = 5, b = 5, l = 5)
  )

caption_plot
```

Join plots and add text annotations

```{r, fig.height= 14, fig.width= 9}
fig_pca_heat <- (
  (pca_plot / heatmap_wrapped) +
  plot_annotation(tag_levels = "A",
                  caption = cap_wrapped) &
  theme(plot.tag = element_text(size = 10, 
                                colour = "grey50"),
        plot.caption = element_text(size   = 8,
                                colour = "grey30",
                                hjust  = 0.22,
                                margin = margin(t = 0, r = 0, b = 5, l = 0))   # << add space ABOVE caption)
  ))+
  labs(
    subtitle = "Z-scored expression of DA tested gene set (n = 12,221) • Each row is a gene",
  )+
  theme(
    plot.subtitle = element_text(
      size   = 10,
      colour = "grey50",
      hjust  = 0.5
    ),
   
    plot.margin = margin(t = 5, r = 5, b = 60, l = 5)   # << extra space at bottom for caption
  )


fig_pca_heat
```

> caption = paste(
        "A. Principal Component Analysis (PCA) of VST-normalized gene expression across all samples.",
        "PC1 explains 53.9% of the total variance and primarily separates samples by embryonic stage,",
        "while PC2 explains 18.5% of the variance, accounting for 72.4% of total sample variance.",
        "Each point represents an individual sample, positioned according to its overall expression profile.",
        "B. Heatmap of z-scored variance-stabilized gene count matrix. Values were transformed using variance-stabilizing",
        "transformation (VST) and then standardized on a per-gene basis (z-score transformation). For each gene, the mean",
        "expression across samples was subtracted, and the standard deviation was used to scale its expression profile.",
        "A value of 0 indicates no change in expression for that gene across all samples, positive values indicate",
        "above-average expression, and negative values indicate below-average expression. This allows visual comparison",
        "of expression patterns across samples (columns), not absolute expression levels between genes (rows).",
        sep = " ")

```{r}
# Save figure
ggsave(
  filename = "../../output/05_explore/fig_pca_heatmap_fullgene.png",
  plot     = fig_pca_heat,
  width    = 9,
  height   = 12,
  units    = "in",
  dpi      = 900
)
```

