---
title: "Step 5: Final figure of PCA & Heatmap of full gene set"
subtitle: "Samples group largely by stage rather than leachate"
author: "Sarah Tanja"
date: 11/22/2025
date-modified: today
bibliography: "../../references.bib"
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin

editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = ""
  ) # Prevents appending '##' to beginning of lines in code output))        
```

# Background

## Load libraries
```{r}
library(colorspace)
library(shinyjs) # for colorspace hclwizard
library(ggsidekick) # for sean andersons theme_sleek()
library(ComplexHeatmap)
library(tidyverse)
library(patchwork)
```

## Load inputs
```{r}
mat <- readRDS("../../output/05_explore/vst_matrix.rds")
pca <- read_csv("../../output/05_explore/pca_500_or.csv")
pheat_df <- readRDS("../../output/05_explore/pheat_annotation_df.rds")
```

## Custom ggplot themes
```{r}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.5
      ),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 12,
        color = "grey30",
        hjust = 0.5,
        margin = margin(b = 2)
      ),
      
      strip.text = element_text(size = 10, color = "grey30"),
      
      plot.subtitle = element_text(
        size   = 10,
        color  = "grey30",
        hjust  = 0.5,
        margin = margin(t = 0)
      ),
      
      axis.title = element_text(size = 10, color = "grey30"),
      axis.text = element_text(size = 10, color = "grey30"),
      legend.title = element_text(size = 10, color = "grey30"),
      legend.text = element_text(size = 10, color = "grey30")
    )
}


theme_pptx <- function() {
  theme_sleek() +
    theme(
      panel.border = element_blank(),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 40,
        color = "grey30",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 24, color = "grey30"),
      
      plot.subtitle = element_text(
        size   = 24,
        color  = "grey30",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 24, color = "grey30"),
      axis.text = element_text(size = 24, color = "grey30"),
      legend.title = element_text(size = 24, color = "grey30"),
      legend.text = element_text(size = 24, color = "grey30")
    )
}
```

```{r, geoms}
geom_defaults <- function() {
  update_geom_defaults("point",      list(size = 1.5, stroke = 0.5))
  update_geom_defaults("line",       list(linewidth = 0.5))
  update_geom_defaults("boxplot",    list(linewidth = 0.5))
  update_geom_defaults("violin",     list(linewidth = 0.5))
  update_geom_defaults("vline",      list(linewidth = 0.5))
  update_geom_defaults("hline",      list(linewidth = 0.5))
  update_geom_defaults("abline",     list(linewidth = 0.5))
  update_geom_defaults("errorbar",   list(linewidth = 0.5))
  update_geom_defaults("pointrange", list(linewidth = 0.5))
  update_geom_defaults("crossbar",   list(linewidth = 0.5))
  update_geom_defaults("linerange",  list(linewidth = 0.5))
  update_geom_defaults("path", list(linewidth = 0.5))
}

geom_pptx <- function() {
  update_geom_defaults("point",   list(size = 5, stroke = 1.2))
  update_geom_defaults("line",    list(linewidth = 1.5))
  update_geom_defaults("boxplot", list(linewidth = 1.2))
  update_geom_defaults("violin",  list(linewidth = 1.2))
  update_geom_defaults("vline", list(linewidth = 1.5))
  update_geom_defaults("hline", list(linewidth = 1.5))
  update_geom_defaults("abline", list(linewidth = 1.5))
  update_geom_defaults("errorbar", list(linewidth = 1.5))
  update_geom_defaults("pointrange", list(linewidth = 1.5))
  update_geom_defaults("crossbar", list(linewidth = 1.5))
  update_geom_defaults("linerange", list(linewidth = 1.5))
  update_geom_defaults("path", list(linewidth = 1.5))
}

```


## Set colorschemes

```{r}
leachate.colors <- c(control = "#AEF1FF", 
                     low     = "#BBC7FF",
                     mid     = "#7D8BFF", 
                     high    = "#592F7D")

stage.5.colors <- c(egg           = "#FFE362",
                    cleavage      = "#EBA600", 
                    morula        = "#E6AA83",
                    prawnchip     = "#D9685B", 
                    earlygastrula = "#A2223C")

stage.3.colors <- c(cleavage      = "#EBA600", 
                    prawnchip     = "#D9685B", 
                    earlygastrula = "#A2223C")

status.colors <- c(typical   = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")

night.colors <- c(July_6th = "#E3FAA5", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")

night.colors.alt <- c(July_6th = "#21918C", 
                      July_7th = "#578B21", 
                      July_8th = "#1E2440")
```


```{r}
# Annotation colors for pheatmap
ann_colors = list(
  leachate = leachate.colors,
  stage = stage.3.colors,
  night = night.colors.alt
)
```

# PCA plot of full gene set
```{r}
pca_flip <- pca %>% 
  mutate(stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula")),
         leachate = factor(leachate, levels = c("control", "low", "mid", "high")))%>%
  dplyr::mutate(PC1 = -PC1)
```


Extract the Percent Variance for the X and Y axis labels
```{r}
pca_comp <- prcomp(t(mat))         # samples in rows

percentVar <- (pca_comp$sdev^2) / sum(pca_comp$sdev^2) * 100

pc1_var <- round(percentVar[1], 1)
pc2_var <- round(percentVar[2], 1)
```


```{r, fig.height= 3, fig.width = 7}
labs_leachate <- c(control = "0 mg/L (control)",
                   low = "0.01 mg/L (low)",
                   mid = "0.1 mg/L (mid)",
                   high = "1 mg/L (high)")

labs_stage <- c(cleavage = "Cleavage", prawnchip = "Prawn chip", earlygastrula = "Early gastrula")

pca_plot <- ggplot(pca_flip, aes(PC1, PC2, color = stage, shape = leachate, fill = leachate, label = name)) +
  geom_point(alpha = 0.7) +
  xlab(paste0("PC1: ", pc1_var, "% variance")) +
  ylab(paste0("PC2: ", pc2_var, "% variance")) + 
  coord_fixed() +

  # Control legend labels + titles here
  scale_color_manual(
    name = "Stage",
    values = stage.3.colors,
    labels = labs_stage
  ) +
  scale_shape_manual(
    name = "PVC Leachate",
    values = c(21, 22, 23, 24), # open circle, square, diamond, triangle
    labels = labs_leachate, 
    #guide = "none"
  ) +
  scale_fill_manual(
    name = "PVC Leachate",
    values = leachate.colors,
    labels = labs_leachate, 
    #guide = "none" # prevents duplicate legend
  )+

  stat_ellipse(aes(group = stage, color = stage), show.legend = FALSE) + 
  labs(
    title = "Variation of gene expression patterns among samples",
    subtitle = "PCA of 500 most variable genes • Each point is a sample"
  ) +
  theme_sleek_axe() +
  theme(panel.border = element_blank())

geom_defaults()
pca_plot
```

```{r}
ggsave(
  "../../figs/pca_gene.png",
  plot=pca_plot,
  width=7,
  height=3,
  units="in",
  dpi=600
)
```


```{r pca-pptx, fig.width=16, fig.height=9}
geom_pptx()
pca_pptx <- pca_plot + theme_pptx()
pca_pptx
```

```{r}
ggsave(
  "../../figs/pca_pptx.png",
  plot=pca_pptx,
  width=16,
  height=9,
  units = "in",
  dpi=600
)
```


# Heatmap of full gene set

Annotations for our heatmap
```{r}

ha_top <- HeatmapAnnotation(
  stage = pheat_df$stage,
  col   = list(stage = ann_colors$stage),
  annotation_name_side = "right"  # label "stage" on the left side
)

ha_bottom <- HeatmapAnnotation(
  `spawn night` = pheat_df$spawn_night,
  leachate      = pheat_df$leachate,
  col = list(
    `spawn night` = ann_colors$night,
    leachate      = ann_colors$leachate
  ),
  annotation_name_side = "right"
)
```

Scale the matrix to be z scores! And then clamp extreme values for better visualization. 
```{r}
mat_scaled <- t(scale(t(mat)))     # row-wise z-scaling
mat_scaled_clamped <- pmin(pmax(mat_scaled, -5), 5)  # clamp extreme values to -5 and 5
```

```{r, fig.height = 8, fig.width= 6}
# 1. build column dendrogram once
col_hc <- hclust(
  dist(t(mat_scaled_clamped), method = "euclidean"),
  method = "complete"
)

# 2. flip it
col_dend_flipped <- rev(as.dendrogram(col_hc))

set.seed(11222025)
Heatmap(
  mat_scaled_clamped,
  clustering_distance_rows    = "euclidean",
  clustering_method_rows      = "complete",
  cluster_columns             = col_dend_flipped,
  name = "zscore of gene expression",
  border = NA,
  column_dend_gp  = gpar(col = "grey50", lwd = 0.7),
  row_names_gp    = gpar(col = "grey50", fontsize = 8),
  column_names_gp = gpar(col = "grey50", fontsize = 8),
  col = rev(colorspace::divergingx_hcl(n = 200, palette = "Roma")),
  cluster_column_slices = TRUE,
  show_row_dend    = FALSE,
  show_column_dend = TRUE,
  top_annotation    = ha_top,
  bottom_annotation = ha_bottom,
  show_row_names    = FALSE,
  show_column_names = TRUE
)
```
> Expression values were transformed using variance-stabilizing transformation (VST) and then standardized on a per-gene basis (z-score transformation). For each gene, the mean expression across samples was subtracted, and the standard deviation was used to scale its expression profile. Thus, a value of 0 indicates average expression for that gene across all samples, positive values indicate above-average expression, and negative values indicate below-average expression. This allows visual comparison of expression patterns, not absolute expression levels between genes.

# Patchwork 

```{r, fig.width = 6, fig.height = 8}
#| include: false
set.seed(11222025)

ht <- 
Heatmap(
  mat_scaled_clamped,
  clustering_distance_rows    = "euclidean",
  clustering_method_rows      = "complete",
  cluster_columns             = col_dend_flipped,
  name = "zscore",
  border = NA,
  column_dend_gp  = gpar(col = "grey50", lwd = 0.7),
  row_names_gp    = gpar(col = "grey50", fontsize = 8),
  column_names_gp = gpar(col = "grey50", fontsize = 8),
  col = rev(colorspace::divergingx_hcl(n = 200, palette = "Roma")),
  cluster_column_slices = TRUE,
  show_row_dend    = FALSE,
  show_column_dend = TRUE,
  top_annotation    = ha_top,
  bottom_annotation = ha_bottom,
  show_row_names    = FALSE,
  show_column_names = TRUE
)

# 2. Grab the drawn heatmap as a grob
ht_grob <- grid.grabExpr(draw(ht, 
                              heatmap_legend_side    = "right",
                              annotation_legend_side = "right",
                              merge_legends          = TRUE))

# 3. Wrap it so patchwork can use it
heatmap_wrapped <- patchwork::wrap_elements(ht_grob)

heatmap_wrapped <- heatmap_wrapped +
  labs(
    title = "Z-scored expression of DESeq2 tested genes (n = 12,221)",
    subtitle = "Brown is increased expression, blue is decreased expression • Each row is a gene "
  ) +
  theme(
    plot.title = element_text(
      size = 12,
      colour = "grey30",
      hjust = 0.5
    ),
    plot.subtitle = element_text(
      size = 10,
      colour = "grey30",
      hjust = 0.5
    )
  )

heatmap_wrapped
```
```{r}
ggsave(
  "../../figs/heatmap_gene.png",
  plot=heatmap_wrapped,
  width=6,
  height=8,
  dpi=600
)
```

```{r, fig.width=16, fig.height=9}
geom_pptx()
heatmap_pptx <- heatmap_wrapped + theme_pptx() 
heatmap_pptx
```

```{r}
ggsave(
  "../../figs/heatmap_pptx.png",
  plot=heatmap_pptx,
  width=16,
  height=9,
  units    = "in",
  dpi=600
)
```


Join plots and add text annotations

```{r, fig.height= 11, fig.width= 8}
geom_defaults()
fig_pca_heat <- (
  (wrap_elements(pca_plot) / heatmap_wrapped) +
    plot_layout(heights = c(0.8, 1.5))+
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 10, 
                                colour = "grey30")
        )
  )  


fig_pca_heat
```

> A. Variance-stabilized gene expression profiles across all samples visualized using principal component analysis (PCA). PC1 explained 53.9% of the total variance and primarily separated samples by embryonic stage, while PC2 explained 18.5% of the variance, together accounting for 72.4% of total sample variation. Each point represents an individual sample positioned according to its overall gene expression profile. B. Relative gene expression patterns across samples based on variance-stabilized and z-score standardized gene counts. For each gene, mean expression across samples was centered to zero and scaled by the standard deviation, allowing comparison of relative expression patterns among samples rather than absolute expression levels between genes. Positive values indicate above-average expression for a given gene, while negative values indicate below-average expression.

```{r}
# Save figure
ggsave(
  filename = "../../figs/pca_heatmap.png",
  plot     = fig_pca_heat,
  width    = 8,
  height   = 11,
  units    = "in",
  dpi      = 900
)
```

