---
title: "Step 3: QAQC RNA sequences"
subtitle: "Use `FastQC`, `MultiQC`, and `Fastp` to clean up sequences"
author: "Sarah Tanja"
date: 10/10/2024
date-format: long
date-modified: today
format: gfm
toc: true
toc-depth: 3
link-external-icon: true
link-external-newwindow: true
reference-location: margin
citation-location: margin
---

# Overview

In this script, we will generate FastQC/MultiQC for raw sequences, conduct trimming and cleaning, then generate reports for cleaned sequences.

## [fastqc](https://anaconda.org/bioconda/fastqc)

*FastQC generates sequence quality information of your reads*

```{r, engine = 'bash'}
/home/shared/FastQC-0.12.1/fastqc -h

```

## [multiqc](https://anaconda.org/bioconda/multiqc) 

*Multiqc summarizes FastQC analysis logs and summarizes results in an html report. Checkout the* [git developer version](Installation)

## [fastp](https://anaconda.org/bioconda/fastp)

*FastP provides fast all-in-one preprocessing for FastQ files*

Reference bash variables from Roberts Lab [Code Snippets](https://robertslab.github.io/resources/code_Snippets/)

How to do a [for loop in bash](https://stackoverflow.com/questions/9612090/how-to-loop-through-file-names-returned-by-find)

::: {.callout-tip appearance="minimal"}
Instead of \``git add –all` execute `find . -type f -size -104M -exec git add {} +` in terminal before making a git commit to only track files that are under the git size limit of 10MB.
:::

# Create a Bash variables file

This allows usage of Bash variables across R Markdown chunks.

## Name directory paths

```{r, engine = bash}
{
echo "#### Assign Variables ####"
echo ""

echo "# Data directories"
echo 'export project_dir=/media/4TB_JPG_ext/stanja/gitprojects/coral-embryo-ecotox'
echo 'export output_dir_top=${project_dir}/output'
echo 'export output_dir_qaqc=${project_dir}/output/03_qaqc'
echo 'export output_dir_raw=${project_dir}/output/03_qaqc/raw_reads'
echo 'export output_dir_trim=${project_dir}/output/03_qaqc/trimmed_reads'
echo 'export raw_reads_dir=${project_dir}/rawfastq'
echo ""
} > .bashvars

cat .bashvars
```

## Program paths

```{r, engine = 'bash'}
{
echo "# Paths to programs"
echo 'export programs_dir="/home/shared"'
echo 'export fastp="${programs_dir}/fastp"'
echo 'export fastqc=${programs_dir}/FastQC-0.12.1/fastqc'
echo 'export multiqc=/home/sam/programs/mambaforge/bin/multiqc'
echo ""
} >> .bashvars

cat .bashvars
```

## Filename patterns

```{r, engine = 'bash'}
{
echo "# Set FastQ filename patterns"
echo "export fastq_pattern='*.fastq.gz'"
echo "export R1_fastq_pattern='*_R1_*.fastq.gz'"
echo "export R2_fastq_pattern='*_R2_*.fastq.gz'"
echo "export trimmed_fastq_pattern='*fastp-trim.fq.gz'"
echo ""
} >> .bashvars

cat .bashvars
```

## Thread count

```{r, engine = 'bash'}
{
echo "# Set number of CPUs to use"
echo 'export threads=40'
echo ""
} >> .bashvars

cat .bashvars
```

## File name arrays

```{r, engine = 'bash'}
{
echo "## Inititalize arrays"
echo 'export fastq_array_R1=()'
echo 'export fastq_array_R2=()'
echo 'export raw_fastqs_array=()'
echo 'export R1_names_array=()'
echo 'export R2_names_array=()'
echo ""
} >> .bashvars

cat .bashvars
```

## Program array

```{r, engine = 'bash'}
{
echo "# Programs associative array"
echo "declare -A programs_array"
echo "programs_array=("
echo '[fastp]="${fastp}" \'
echo '[fastqc]="${fastqc}" \'
echo '[multiqc]="${multiqc}" \'
echo ")"
echo ""
} >> .bashvars

cat .bashvars
```

# Quality check raw reads with `FastQC`

FastQC on Raven lives: `/home/shared/FastQC/fastqc`

```{r, engine = 'bash'}
# Load bash variables into memory
source .bashvars

${fastqc} -h
```

```{r, engine = 'bash'}
# Load bash variables into memory
source .bashvars

# Populate array with FastQ files
fastq_array=(${raw_reads_dir}/*.fastq.gz)

# Pass array contents to new variable
fastqc_list=$(echo "${fastq_array[*]}")

# Run FastQC
# NOTE: Do NOT quote ${fastqc_list}
${fastqc} \
--threads ${threads} \
--outdir ${output_dir_raw} \
${fastqc_list}
```

# Compile a report of raw reads with `MultiQC`

```{r, engine = 'bash'}
# Load bash variables into memory
source .bashvars

cd ../output/03_qaqc/raw_reads
${multiqc} ./
```

# Interpretation of `MultiQC` report for raw reads

Watch a quick [6-min tutorial](https://www.youtube.com/watch?v=qPbIlO_KWN0) on how to navigate in the MultiQC Report

# Trim & Clean reads using `fastp`

`fastp` all [options](https://github.com/OpenGene/fastp?tab=readme-ov-file#:~:text=of%201%20~%206.-,all%20options,-usage%3A%20fastp%20%2Di) can be found in the fastp git README file.

-   `--in1` \| Path to forward read input

-   `--in2` \| Path to reverse read input

-   `--qualified_quality_phred` \| The quality value that a base is qualified. Default 15 means phred quality \>=Q15 is qualified. Chosen value is 30.

-   `--trim_poly_x 6` \| Enable polyX trimming in 3' ends. Trims any repeating sequence (AAAAAA, TTTTTT, CCCCCC, GGGGGG), the default is 10... here we've chosen if it has 6 or more repeating bases to trim them off. This is more conservative than 10 or more repeating bases, and will result in cleaner sequences that have fewer polyA tails leftover.

-   `--detect_adapter_for_pe` \| This automatically detects Illumina adapter sequences for paired end (pe) sequences and trims them off. Adapters can be trimmed by overlap analysis, however, `--detect_adapter_for_pe` will usually result in slightly cleaner output than overlap detection alone. This results in a slightly slower run time.

-   `--thread 16` \| fastp uses up to 16 threads (even if there are more available). If this is not specified, it uses 3 as a default.

-   `--html` \| The html format report file name

-   `--json` \| The json format report file name

-   `-out1` \| Path to forward read output

-   `--out2` \| Path to reverse read output

```{bash}
# Load bash variables into memory
source .bashvars

# Change to raw reads directory
cd "${raw_reads_dir}"

# Create arrays of fastq R1 files and sample names
for fastq in ${R1_fastq_pattern}
do
  fastq_array_R1+=("${fastq}")
  R1_names_array+=("$(echo "${fastq}" | awk -F"_" '{print $1}')")
done

# Create array of fastq R2 files
for fastq in ${R2_fastq_pattern}
do
  fastq_array_R2+=("${fastq}")
  R2_names_array+=("$(echo "${fastq}" | awk -F"_" '{print $1}')")
done

# Create list of fastq files used in analysis
# Create MD5 checksum for reference
if [ ! -f "${raw_reads_dir}"/*raw-fastq-checksums*fastq.gz.md5 ]; then
for fastq in *.gz
  do
    md5sum ${fastq} >> "${output_dir_qaqc}"/checksums/raw-fastq-checksums.md5
  done
fi

# Run fastp on files
# Adds JSON report output for downstream usage by MultiQC
for index in "${!fastq_array_R1[@]}"
do
  R1_sample_name=$(echo "${R1_names_array[index]}")
  R2_sample_name=$(echo "${R2_names_array[index]}")
  ${fastp} \
  --in1 ${fastq_array_R1[index]} \
  --in2 ${fastq_array_R2[index]} \
  --detect_adapter_for_pe \
  --qualified_quality_phred 30 \
  --thread 16 \
  --trim_poly_x 6 \
  --html "${output_dir_trim}"/"${R1_sample_name}".fastp-trim.report.html \
  --json "${output_dir_trim}"/"${R1_sample_name}".fastp-trim.report.json \
  --out1 "${output_dir_trim}"/"${R1_sample_name}"_R1_001.fastp-trim.fq.gz \
  --out2 "${output_dir_trim}"/"${R2_sample_name}"_R2_001.fastp-trim.fq.gz \
  2>> "${output_dir_trim}"/fastp.stderr


  # Generate md5 checksums for newly trimmed files
  cd "${output_dir_trim}"
  md5sum "${R1_sample_name}"_R1_001.fastp-trim.fq.gz > "${R1_sample_name}"_R1_001.fastp-trim.fq.gz.md5
  md5sum "${R2_sample_name}"_R2_001.fastp-trim.fq.gz > "${R2_sample_name}"_R2_001.fastp-trim.fq.gz.md5
  cd -
done
```

# Quality check trimmed reads with `FastQC`

```{r, engine= 'bash'}
# Load bash variables into memory
source .bashvars

############ RUN FASTQC ############

# Create array of trimmed FastQs
trimmed_fastqs_array=(${output_dir_trim}/${trimmed_fastq_pattern})

echo "Files in trimmed_fastqs_array: ${trimmed_fastqs_array[@]}"

# Pass array contents to new variable as space-delimited list
trimmed_fastqc_list=$(echo "${trimmed_fastqs_array[*]}")

echo ""
echo "Beginning FastQC on trimmed reads..."
echo ""
echo "Output directory: ${output_dir_trim}"
echo ""

# Run FastQC
### NOTE: Do NOT quote trimmed_fastqc_list
${programs_array[fastqc]} \
--threads ${threads} \
--outdir ${output_dir_trim} \
--quiet \
${trimmed_fastqc_list}

echo "FastQC on trimmed reads complete!"
echo ""

############ END FASTQC ############
```

```{r, engine = 'bash'}
# Load bash variables into memory
source .bashvars

# Populate array with fastp-trim.fq.gz FastQ files
trimmed_fastq_array=("${output_dir_trimmed}/*.fastp-trim.fq.gz")

echo "Files in trimmed_fastqs_array: ${trimmed_fastqs_array[@]}"

# Pass array contents to new variable
fastqc_list=$(echo "${fastq_array[*]}")

# Run FastQC
# NOTE: Do NOT quote ${fastqc_list}
${fastqc} \
--threads ${threads} \
--outdir ${output_dir_trim} \
${fastqc_list}
```

# Compile report of trimmed reads with `MultiQC`

```{r, engine = 'bash'}
############ RUN MULTIQC ############
echo "Beginning MultiQC on trimmed FastQC..."
echo ""

${programs_array[multiqc]} ${output_dir_trim} -o ${output_dir_trim}

echo ""
echo "MultiQC on trimmed FastQs complete."
echo ""

############ END MULTIQC ############

echo "Removing FastQC zip files."
echo ""
rm ${output_dir_trim}/*.zip
echo "FastQC zip files removed."
echo ""
```

# Summary & Next Steps

Trimmed reads are now ready for alignment in step 4, `04_align`
