---
title: "Control vs High"
subtitle: "DESeq2 across all developmental stages"
author: "Sarah Tanja"
date: 05/08/2025
format:
  gfm: default  # or html if you want to render in HTML
toc: true
toc-depth: 3
link-external-icon: true
link-external-newwindow: true
reference-location: margin
citation-location: margin
---

# Load libraries

```{r}
library(tidyverse)
library(DESeq2)
library(vsn)
```

# Filter count matrix

We want to look at the `control` group (embryos developed in 0.22 um filtered seawater) vs. the `high` group across all embryonic stages. To do this let's start by filtering the count matrix to only the samples of interest.

```{}
```

```{r}
load("../output/06_exploration/gcm_tidyfilt.RData")
```

The genes are the rows. The samples are the columns.

remove outlier samples

```{r}
gcm_filtout <- gcm_tidyfilt %>% 
  select(-"131415L4", -"789C4")
```

remove low & mid

```{r}
gcm_controlvshigh <- gcm_filtout %>% 
  select(-matches("L|M"))
```

# Setup metadata

> It is absolutely critical that the columns of the count matrix and the rows of the column data (information about samples) are in the same order. DESeq2 will not make guesses as to which column of the count matrix belongs to which row of the column data, these must be provided to DESeq2 already in consistent order.

```{r}
load("../output/06_exploration/metadata.RData")
```

The samples are rows

remove outlier samples

```{r}
metadata_out <- metadata %>% 
  filter(!sample_name %in% c("131415L4", "789C4"))
```

remove low & mid

```{r}
metadata_controlvshigh <- metadata_out %>% 
  filter(grepl("C|H", sample_name))
```

confirm metadata sample_name matches gene count matrix columns. The following code checks that all the sample names in the metadata dataframe match the column names in the gene count matrix. It should return 'TRUE'

```{r}
meta <- (metadata_controlvshigh$sample_name)
gene_matrix <- (colnames(gcm_controlvshigh))
meta_check <- data.frame(meta, gene_matrix)

all(meta == gene_matrix)
```

# Make DESeq Object

Make [DESeq Dataset Object from Count Matrix](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input) Create a DESeqDataSet design from gene count matrix and treatment conditions.

```{r}
#Set DESeq2 design for 1 factor with 2 levels
dds_cvh <- DESeqDataSetFromMatrix(countData = gcm_controlvshigh,
                              colData = metadata_controlvshigh, 
                              design = ~ pvc_leachate_level)
```

# Count transformations

```{r}
vsd_cvh <- vst(dds_cvh, blind=FALSE)
rld_cvh <- rlog(dds_cvh, blind=FALSE)

```

```{r}
dds_cvh <- estimateSizeFactors(dds_cvh)

df <- bind_rows(
  as_data_frame(log2(counts(dds_cvh, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd_cvh)[, 1:2]) %>% mutate(transformation = "vst"),
  as_data_frame(assay(rld_cvh)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) +
  geom_hex(bins = 80) +
  coord_fixed() +
  facet_grid(. ~ transformation) +
  ggtitle("Control vs. High across development Count Transformation Comparisons")
```

```{r}
meanSdPlot(assay(vsd_cvh))
meanSdPlot(assay(rld_cvh))
```

# PCA

```{r}
# Perform PCA on raw count matrix
pca_cvh <- prcomp(t(gcm_controlvshigh), scale = TRUE)
```

```{r}
# Perform PCA on 
pca_cvh <- prcomp(t(assay(rld_cvh)), scale = TRUE)
```
