---
title: "Step 6: Filter"
subtitle: "Conduct prefiltering, visualize exploratory PCAs, identify outliers, & format metadata"
author: "Sarah Tanja"
date: 2024-11-18
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
# Background
## Inputs
- `gene_count_matrix.csv`
- `mcap_RNAseq_simplified_metadata.csv`
## Outputs
- `gcm_filt.csv`
- `metadata_filt.csv`
# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  dpi = 600,
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

## Install packages
```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') 
  install.packages('tidyverse')
```
## Load packages
```{r}
library(tidyverse)
```

# Import metadata

```{r}
metadata <- read.csv("../metadata/mcap_RNAseq_simplified_metadata.csv")
str(metadata)
```

## Format metadata

```{r}

metadata$embryonic_stage <- gsub(" ", "", metadata$development_stage) # Remove space in development_stage condition column and rename to embryonic_stage
metadata$leachate_mgL <- metadata$pvc_leachate_concentration_mg.L # Rename to leachate_mgL to simplify
metadata$leachate <- metadata$pvc_leachate_level # Rename to leachate_lvl to simplify

# Make a new column that represents the parent crosses
metadata <- metadata %>% 
  mutate(parents = sub("^(.*?)[CLMH].*", "\\1", sample_name)) %>% 
  relocate(parents, .after = sample_name)

# Make a new column that simplifies the treatment group
metadata <- metadata %>% 
  mutate(group = sub(".*?([CLMH].*)", "\\1", sample_name)) %>% 
  relocate(group, .after = parents)

# Reorder factors
metadata$embryonic_stage <- as.factor(metadata$embryonic_stage)
metadata$embryonic_stage <- fct_relevel(metadata$embryonic_stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata$leachate <- as.factor(metadata$leachate)
metadata$leachate <- fct_relevel(metadata$leachate, "control", "low", "mid", "high")

metadata$hpf <- as.factor(metadata$hours_post_fertilization)
metadata$hpf <- fct_relevel(metadata$hpf, "4", "9", "14")
```

## Filter metadata

```{r}
metadata_filt <- metadata %>% dplyr::select(sample_name, parents, group, hpf, embryonic_stage, leachate, leachate_mgL)

metadata_filt <- metadata_filt %>% 
  filter(!sample_name %in% c("131415L4", "789C4"))
```


# Import gene count matrix (gcm)

> In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were probabilistically assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r}
gcm <- read.csv("../output/05_count/gene_count_matrix.csv", row.names="gene_id", check.names = FALSE)
head(gcm)
```

The count matrix should show counts of whole numbers (integers) ... since each column shows <int> as the data type we know that the counts are already integers

# Exclude outliers 
Two samples had poor hisat alignment and should be removed from the dataset: 131415L4 and 789C4

```{r}
gcm_filt <- gcm %>% 
  dplyr::select(-`131415L4`, -`789C4`)
```

> It is absolutely critical that the columns of the count matrix and the rows of the column data (information about samples) are in the same order. DESeq2 will not make guesses as to which column of the count matrix belongs to which row of the column data, these must be provided to DESeq2 already in consistent order.

The following code checks that all the sample names in the metadata dataframe match the column names in the gene count matrix. It should return 'TRUE'
```{r}
print(paste("There are", nrow(metadata_filt), "samples in total"))

# Check if all matched names are equal (should return TRUE)
all(metadata_filt$sample_name == colnames(gcm_filt))
```

Save gcm & metadata 
```{r}
write_csv(metadata_filt, "../metadata/metadata_filt.csv")
write_csv(rownames_to_column(gcm_filt, var = "gene_id"), "../output/06_tidyup/gcm_filt.csv")
```


# Subset gene count matrix by hpf

## 4hpf

```{r}
# Subset metadata
# Filter rows where 'sample_name' matches the pattern [CLMH]4
metadata_4 <- metadata_filt %>% 
  filter(str_detect(sample_name, "[CLMH]4"))

# Subset gene count matrix
# Select only samples that are 4hpf out of gene count matrix
gcm_4 <- gcm_filt %>% 
  dplyr::select(matches("[CLMH]4"))

print(paste("There are", nrow(metadata_4), "samples at 4hpf"))

# Check if all matched names are equal (should return TRUE)
all(metadata_4$sample_name == colnames(gcm_4))
```
- There should be 18 cleavage (4hpf) samples

Save gcm & metadata 
```{r}
write_csv(metadata_4, "../metadata/metadata_4.csv")
write_csv(rownames_to_column(gcm_4, var = "gene_id"), "../output/06_tidyup/gcm_4.csv")
```


## 9hpf
```{r}
# Subset metadata
# Filter rows where 'sample_name' matches the pattern [CLMH]9
metadata_9 <- metadata_filt %>% 
  filter(str_detect(sample_name, "[CLMH]9"))

# Subset gene count matrix
# Select only samples that are 9hpf out of gene count matrix
gcm_9 <- gcm_filt %>% 
  dplyr::select(matches("[CLMH]9"))

print(paste("There are", nrow(metadata_9), "samples at 9hpf"))

# Check if all matched names are equal (should return TRUE)
all(metadata_9$sample_name == colnames(gcm_9))
```
- There should be 21 prawn chip (9hpf) samples

Save gcm & metadata 
```{r}
write_csv(metadata_9, "../metadata/metadata_9.csv")
write_csv(rownames_to_column(gcm_9, var = "gene_id"), "../output/06_tidyup/gcm_9.csv")
```

## 14hpf

```{r}
# Subset metadata
# Filter rows where 'sample_name' matches the pattern [CLMH]14
metadata_14 <- metadata_filt %>% 
  filter(str_detect(sample_name, "[CLMH]14"))

# Subset gene count matrix
# Select only samples that are 9hpf out of gene count matrix
gcm_14 <- gcm_filt %>% 
  dplyr::select(matches("[CLMH]14"))

print(paste("There are", nrow(metadata_14), "samples at 14hpf"))

# Check if all matched names are equal
all(metadata_14$sample_name == colnames(gcm_14))
```
- There should be 22 early gastrula (14hpf) samples

save metadata_14
```{r}
#output metadata for other analyses 
write_csv(metadata_14, "../metadata/metadata_14.csv")
write_csv(rownames_to_column(gcm_14, var = "gene_id"), "../output/06_tidyup/gcm_14.csv")
```


# Prefiltering

[Prefiltering](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering)

Pre-filtering our dataset reduces the memory size of the dataframe, increases the speed of the transformation and testing functions, and improves quality of statistical analysis by removing low-coverage counts. Removed counts could represent outliers in the data and removing these improves sensitivity of statistical tests.

-   **`across(everything(), ~ . > n)`**: This applies the condition `> n` to all columns of the data frame. For each cell, it returns `TRUE` if the value is greater than n and `FALSE` otherwise.

-   **`rowSums()`**: This counts the number of `TRUE` values for each row. It adds up how many columns in each row have values greater than 10.

-   **`>= n * ncol(df)`**: This checks if the count of `TRUE` values in each row is greater than or equal to n% of the total number of columns (`ncol(df)`).

### All

```{r}
# Filter rows where at least 8% of the columns have a value greater than 10
gcm_filt <- gcm_filt %>%
  filter(rowSums(across(everything(), ~ . > 15)) >= 0.75 * ncol(gcm))

nrow(gcm)
nrow(gcm_filt)
```

::: callout-note
We reduced our gene count matrix from 54,384 to 12,225 genes by selecting genes present (in counts of at least 15!) in at least 75% of the samples
:::

```{r}

ggplot(data.frame(log_counts = log2(rowSums(gcm_filt) + 1)), aes(x = log_counts)) +
  geom_histogram(bins = 100, fill = "steelblue") +
  labs(
    title = "Distribution of Log2-Transformed Total Counts per Gene",
    x = "log2(Total Read Counts + 1)",
    y = "Number of Genes"
  ) +
  theme_minimal()

```


We are going to run DGE analysis on each hpf separately (as well as together) so let's setup gene count matrices for each of 4, 9, and 14 hours post fertilization

### 4hpf
```{r}
# Filter rows where at least 75% of the columns have a value greater than 15
gcm_filt_4 <- gcm_4 %>%
  filter(rowSums(across(everything(), ~ . > 15)) >= 0.75 * ncol(gcm_4))

nrow(gcm_4)
nrow(gcm_filt_4)

write_csv(rownames_to_column(gcm_filt_4, var = "gene_id"), file = "../output/06_tidyup/gcm_filt_4.csv")
```

::: callout-note
54384 genes were filtered down to 10,936 genes that have at least 15 copies in at least 75% of the 4hpf samples
:::
```{r}

ggplot(data.frame(log_counts = log2(rowSums(gcm_filt_4) + 1)), aes(x = log_counts)) +
  geom_histogram(bins = 100, fill = "steelblue") +
  labs(
    title = "4 hpf : Distribution of Log2-Transformed Total Counts per Gene",
    x = "log2(Total Read Counts + 1)",
    y = "Number of Genes"
  ) +
  theme_minimal()
```

### 9hpf
```{r}
# Filter rows where at least 75% of the columns have a value greater than 15
gcm_filt_9 <- gcm_9 %>%
  filter(rowSums(across(everything(), ~ . > 15)) >= 0.75 * ncol(gcm_9))

nrow(gcm_9)
nrow(gcm_filt_9)

write_csv(rownames_to_column(gcm_filt_9, var = "gene_id"), file = "../output/06_tidyup/gcm_filt_9.csv")
```

::: callout-note
54384 genes were filtered down to 13,781 genes that have at least 15 copies in at least 75% of the 9hpf samples
:::

```{r}

ggplot(data.frame(log_counts = log2(rowSums(gcm_filt_9) + 1)), aes(x = log_counts)) +
  geom_histogram(bins = 100, fill = "steelblue") +
  labs(
    title = "9 hpf : Distribution of Log2-Transformed Total Counts per Gene",
    x = "log2(Total Read Counts + 1)",
    y = "Number of Genes"
  ) +
  theme_minimal()
```

### 14hpf
```{r}
# Filter rows where at least 75% of the columns have a value greater than 15
gcm_filt_14 <- gcm_14 %>%
  filter(rowSums(across(everything(), ~ . > 15)) >= 0.75 * ncol(gcm_14))

nrow(gcm_14)
nrow(gcm_filt_14)

write_csv(rownames_to_column(gcm_filt_14, var = "gene_id"), file = "../output/06_tidyup/gcm_filt_14.csv")
```

::: callout-note
54384 genes were filtered down to 15,501 genes that have at least 15 copies in at least 75% of the 14hpf samples
:::

```{r}

ggplot(data.frame(log_counts = log2(rowSums(gcm_filt_14) + 1)), aes(x = log_counts)) +
  geom_histogram(bins = 100, fill = "steelblue") +
  labs(
    title = "14 hpf: Distribution of Log2-Transformed Total Counts per Gene",
    x = "log2(Total Read Counts + 1)",
    y = "Number of Genes"
  ) +
  theme_minimal()
```


::: callout-important
More gene transcripts are active as embryonic development progresses!

4hpf : 14,191

9hpf : 17,368

14hpf : 21,447
:::


# Sample summary
```{r}
# Create a summary table
metadata_filt %>%
  count(hpf, leachate) %>%
  ggplot(aes(x = leachate, y = hpf, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), color = "black", size = 5) +
  scale_fill_gradient(low = "lightblue", high = "steelblue") +
  labs(
    title = "Sample distribution by embryonic phase and pollution exposure",
    x = "Pollution Exposure Level",
    y = "Embryonic Phase",
    fill = "Sample Count"
  ) +
  theme_minimal(base_size = 14)
```

::: callout-note
We have 61 samples spanning three embryonic phases and four pollution exposure levels, yielding 12 distinct treatment combinations. The heatmap above shows how samples are distributed across treatments. Our minimum group size is 4, which guides our gene filtering threshold. We hypothesize that both developmental stage and pollution exposure independently and interactively influence gene expression.
:::

## Common genes

```{r}
# Convert rownames to a column
gcm_4_df <- rownames_to_column(gcm_filt_4, var = "gene_id")
gcm_9_df <- rownames_to_column(gcm_filt_9, var = "gene_id")
gcm_14_df <- rownames_to_column(gcm_filt_14, var = "gene_id")

# Join by gene_id
gcm_common <- gcm_4_df %>%
  inner_join(gcm_9_df, by = "gene_id") %>%
  inner_join(gcm_14_df, by = "gene_id")

nrow(gcm_common)
```

