---
title: "Pairwise Differential Gene Expression"
subtitle: "DESeq2 between control and high treatments across all developmental stages"
author: "Sarah Tanja"
date: 2025-05-08
date-modified: today
format:
  gfm: default
toc: true
toc-depth: 3
link-external-icon: true
link-external-newwindow: true
reference-location: margin
citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Goals
This quick analysis pairs down the complexity of multiple embryonic stages and multiple levels of leachate to a direct comparison of our control group to our highest exposure group. Across all stages of embryonic development, is there any impact of our highest pvc leachate treatment on gene expression? 

# Load libraries

```{r}
library(tidyverse)
library(DESeq2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
```

# Filter count matrix

We want to look at the `control` group (embryos developed in 0.22 um filtered seawater) vs. the `high` group across all embryonic stages. To do this let's start by filtering the count matrix to only the samples of interest.

```{r}
load("../output/07_exploration/gcm_filt_or.RData")
```

The genes are the rows. The samples are the columns.

remove low & mid

```{r}
gcm_controlvshigh <- gcm_filt_or %>% 
  dplyr::select(-matches("L|M"))
```

# Setup metadata

> It is absolutely critical that the columns of the count matrix and the rows of the column data (information about samples) are in the same order. DESeq2 will not make guesses as to which column of the count matrix belongs to which row of the column data, these must be provided to DESeq2 already in consistent order.

```{r}
metadata_or <- read_csv("../metadata/metadata_or.csv")
```

The samples are rows

remove low & mid

```{r}
metadata_controlvshigh <- metadata_or %>% 
  filter(grepl("C|H", sample_name))
```

confirm metadata sample_name matches gene count matrix columns. The following code checks that all the sample names in the metadata dataframe match the column names in the gene count matrix. It should return 'TRUE'

```{r}
meta <- (metadata_controlvshigh$sample_name)
gene_matrix <- (colnames(gcm_controlvshigh))
meta_check <- data.frame(meta, gene_matrix)

all(meta == gene_matrix)
```

# Make DESeq Object

Make [DESeq Dataset Object from Count Matrix](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input) Create a DESeqDataSet design from gene count matrix and treatment conditions.

```{r}
#Set DESeq2 design for 1 factor with 2 levels
dds_cvh <- DESeqDataSetFromMatrix(countData = gcm_controlvshigh,
                              colData = metadata_controlvshigh, 
                              design = ~ pvc_leachate_level)
```


# Visualizations

> In order to test for differential expression, we operate on raw counts and use discrete distributions... However for visualization or clustering -- it might be useful to work with transformed versions of the count data.

## Count transformations

```{r}
# This is now the variance stabilized transformed DESeq Data
vsd_cvh <- vst(dds_cvh, blind=FALSE)

# This is the r-log transformed DESeq Data
rld_cvh <- rlog(dds_cvh, blind=FALSE)
```

```{r}
dds_cvh <- estimateSizeFactors(dds_cvh)

df <- bind_rows(
  as_data_frame(log2(counts(dds_cvh, normalized=TRUE)[, 1:2]+1)) %>%
         mutate(transformation = "log2(x + 1)"),
  as_data_frame(assay(vsd_cvh)[, 1:2]) %>% mutate(transformation = "vst"),
  as_data_frame(assay(rld_cvh)[, 1:2]) %>% mutate(transformation = "rlog"))
  
colnames(df)[1:2] <- c("x", "y")  

lvls <- c("log2(x + 1)", "vst", "rlog")
df$transformation <- factor(df$transformation, levels=lvls)

ggplot(df, aes(x = x, y = y)) +
  geom_hex(bins = 80) +
  coord_fixed() +
  facet_grid(. ~ transformation) +
  ggtitle("Control vs. High across development Count Transformation Comparisons")
```

```{r}
meanSdPlot(assay(vsd_cvh))
```

```{r}
meanSdPlot(assay(rld_cvh))
```

## PCA

```{r}
# Perform PCA on rlog transformed count data
pca_cvh <- prcomp(t(assay(rld_cvh)), scale = TRUE)

pca_cvh_plot <- plotPCA(rld_cvh, intgroup=c("pvc_leachate_level"), returnData=TRUE)

percentVar <- round(100 * attr(pca_cvh_plot, "percentVar"))

# Make the plot
PCA_CvH_plot <- ggplot(pca_cvh_plot, aes(PC1, PC2, 
                              color=pvc_leachate_level)) +
                              #geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              geom_point(size=3) +
                              xlab(paste0("PC1: ",percentVar[1],"% variance")) +
                              ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
                              coord_fixed() + 
                              geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              theme_minimal()

# Display the plot
PCA_CvH_plot

# Save the plot as a png
ggsave("../output/08_deg/PCA_control_vs_high.png", plot = PCA_CvH_plot, width = 8, height = 6, dpi = 600)
```

## Heatmap

```{r}

# 1. Select the top 50 genes by mean expression
top_genes <- order(rowMeans(counts(dds_cvh, normalized = TRUE)), decreasing = TRUE)[1:50]

# 2. Extract rlog-transformed matrix
rlog_matrix <- assay(rld_cvh)[top_genes, ]

# 3. Extract only pvc_leachate_level annotation
annotation_df <- as.data.frame(colData(dds_cvh)[, "pvc_leachate_level", drop = FALSE])

# Ensure it's a factor in correct order: control, high
annotation_df$pvc_leachate_level <- factor(annotation_df$pvc_leachate_level, levels = c("control", "high"))

# 4. Order samples by pvc_leachate_level
ordered_cols <- order(annotation_df$pvc_leachate_level)

# Reorder matrix and annotation
rlog_matrix_ordered <- rlog_matrix[, ordered_cols]
annotation_ordered <- annotation_df[ordered_cols, , drop = FALSE]

# 5. Plot heatmap
heatmap_colors <- colorRampPalette(rev(brewer.pal(n = 9, "RdBu")))(100)

pheatmap(
  rlog_matrix_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  annotation_col = annotation_ordered,
  color = heatmap_colors,
  main = "Top 50 Most Expressed Genes (Control vs High)"
)

```

# Run DESeq2

```{r}
dds_cvh_deseq <- DESeq(dds_cvh)
```

```{r}
summary(results(dds_cvh_deseq))
```

::: callout-important
There are zero differentially expressed genes between the control and high treatment groups across all developmental stages.
:::
