---
title: "Pairwise Differential Gene Expression"
subtitle: "DESeq2 between control and high treatments by each developmental stage"
author: "Sarah Tanja"
date: 2025-05-08
date-modified: today
format:
  gfm: default
toc: true
toc-depth: 3
link-external-icon: true
link-external-newwindow: true
reference-location: margin
citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Goals
This quick analysis pairs down the complexity of multiple embryonic stages and multiple levels of leachate to a direct comparison of our control group to our highest exposure group. We did not find any significantly differently expressed genes across all embryonic stages, we believe, due to the gene expression in our dataset is clustered by stage, and the effect of leachate was 'drowned out' by the much larger patterns of embryonic stage. This code breaks it down to look at control vs high within each stage of embryonic development. This reduces our n to 5 for each group and we have less power with this analysis, but it will help us get a fuller picture of our data and what it's telling us.  

# Load libraries

```{r}
library(tidyverse)
library(DESeq2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
```

# Load count matrix & metadata

We want to look at the `control` group (embryos developed in 0.22 um filtered seawater) vs. the `high` group in each embryonic stage. To do this let's start by filtering the count matrix to only the samples of interest.

```{r}
gcm_filt_4 <- read_csv("../output/06_tidyup/gcm_filt_4.csv")
gcm_filt_9 <- read_csv("../output/06_tidyup/gcm_filt_9.csv")
gcm_filt_14 <- read_csv("../output/06_tidyup/gcm_filt_14.csv")
```

The genes are the rows. The samples are the columns.

# Select groups
Remove low & mid so we just compare control and high 

```{r}
gcm_filt_4_cvh <- gcm_filt_4 %>% 
  dplyr::select(-matches("L|M"))

gcm_filt_9_cvh <- gcm_filt_9 %>% 
  dplyr::select(-matches("L|M"))

gcm_filt_14_cvh <- gcm_filt_14 %>% 
  dplyr::select(-matches("L|M"))
```

# Setup metadata

> It is absolutely critical that the columns of the count matrix and the rows of the column data (information about samples) are in the same order. DESeq2 will not make guesses as to which column of the count matrix belongs to which row of the column data, these must be provided to DESeq2 already in consistent order.

```{r}
metadata_4 <- read_csv("../metadata/metadata_4.csv")
metadata_9 <- read_csv("../metadata/metadata_9.csv")
metadata_14 <- read_csv("../metadata/metadata_14.csv")
```

The samples are rows

remove low & mid

```{r}
metadata_4_cvh <- metadata_4 %>% 
  filter(grepl("C|H", sample_name))

metadata_9_cvh <- metadata_9 %>% 
  filter(grepl("C|H", sample_name))

metadata_14_cvh <- metadata_14 %>% 
  filter(grepl("C|H", sample_name))
```

confirm metadata sample_name matches gene count matrix columns. The following code checks that all the sample names in the metadata dataframe match the column names in the gene count matrix. It should return 'TRUE'

```{r}
all(metadata_4_cvh$sample_name == colnames(gcm_filt_4_cvh))
all(metadata_9_cvh$sample_name == colnames(gcm_filt_9_cvh))
all(metadata_14_cvh$sample_name == colnames(gcm_filt_14_cvh))
```

# Make DESeq Objects

Make [DESeq Dataset Object from Count Matrix](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input) Create a DESeqDataSet design from gene count matrix and treatment conditions.

```{r}
#Set DESeq2 design for 1 factor with 2 levels
dds_4_cvh <- DESeqDataSetFromMatrix(countData = gcm_filt_4_cvh,
                              colData = metadata_4_cvh, 
                              design = ~ 1 + leachate)

dds_9_cvh <- DESeqDataSetFromMatrix(countData = gcm_filt_9_cvh,
                              colData = metadata_9_cvh, 
                              design = ~ 1 + leachate)

dds_14_cvh <- DESeqDataSetFromMatrix(countData = gcm_filt_14_cvh,
                              colData = metadata_14_cvh, 
                              design = ~ 1 + leachate)
```

## Estimate size factors
```{r}
# Estimate size factors
dds_4_cvh <- estimateSizeFactors(dds_4_cvh)
dds_9_cvh <- estimateSizeFactors(dds_9_cvh)
dds_14_cvh <- estimateSizeFactors(dds_14_cvh)
```



# Visualizations

> In order to test for differential expression, we operate on raw counts and use discrete distributions... However for visualization or clustering -- we work with transformed versions of the count data.

## Count transformations

```{r}
# This is now the variance stabilized transformed DESeq Data
vsd_4_cvh <- vst(dds_4_cvh, blind=FALSE)
vsd_9_cvh <- vst(dds_9_cvh, blind=FALSE)
vsd_14_cvh <- vst(dds_14_cvh, blind=FALSE)

# This is the r-log transformed DESeq Data
#rld_cvh <- rlog(dds_cvh, blind=FALSE)
```

```{r}
meanSdPlot(assay(vsd_4_cvh))
meanSdPlot(assay(vsd_9_cvh))
meanSdPlot(assay(vsd_14_cvh))
```

```{r}
#meanSdPlot(assay(rld_cvh))
```

## PCA
### 4 hpf
```{r}
# Perform PCA on vst transformed count data
pca_4_cvh <- prcomp(t(assay(vsd_4_cvh)), scale = TRUE)

pca_4_cvh_plot <- plotPCA(vsd_4_cvh, intgroup=c("leachate"), returnData=TRUE)

percentVar <- round(100 * attr(pca_4_cvh_plot, "percentVar"))

# Make the plot
PCA_4_CvH_plot <- ggplot(pca_4_cvh_plot, aes(PC1, PC2, 
                              color=leachate)) +
                              #geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              geom_point(size=3) +
                              xlab(paste0("PC1: ",percentVar[1],"% variance")) +
                              ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
                              coord_fixed() + 
                              geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              theme_minimal()

# Display the plot
PCA_4_CvH_plot
```

```{r}
# Save the plot as a png
ggsave("../output/08_deg/PCA_4_control_vs_high.png", plot = PCA_4_CvH_plot, width = 8, height = 6, dpi = 600)
```

### 9 hpf
```{r}
# Perform PCA on vst transformed count data
pca_9_cvh <- prcomp(t(assay(vsd_9_cvh)), scale = TRUE)

pca_9_cvh_plot <- plotPCA(vsd_9_cvh, intgroup=c("leachate"), returnData=TRUE)

percentVar <- round(100 * attr(pca_9_cvh_plot, "percentVar"))

# Make the plot
PCA_9_CvH_plot <- ggplot(pca_9_cvh_plot, aes(PC1, PC2, 
                              color=leachate)) +
                              #geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              geom_point(size=3) +
                              xlab(paste0("PC1: ",percentVar[1],"% variance")) +
                              ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
                              coord_fixed() + 
                              geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              theme_minimal()

# Display the plot
PCA_9_CvH_plot
```
### 14 hpf
```{r}
# Perform PCA on vst transformed count data
pca_14_cvh <- prcomp(t(assay(vsd_14_cvh)), scale = TRUE)

pca_14_cvh_plot <- plotPCA(vsd_14_cvh, intgroup=c("leachate"), returnData=TRUE)

percentVar <- round(100 * attr(pca_14_cvh_plot, "percentVar"))

# Make the plot
PCA_14_CvH_plot <- ggplot(pca_14_cvh_plot, aes(PC1, PC2, 
                              color=leachate)) +
                              #geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              geom_point(size=3) +
                              xlab(paste0("PC1: ",percentVar[1],"% variance")) +
                              ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
                              coord_fixed() + 
                              geom_text(aes(label = name), size = 3) +  # Add labels from the 'name' column
                              theme_minimal()

# Display the plot
PCA_14_CvH_plot
```
## Heatmap

### 4 hpf
```{r}

# 1. Select the top 50 genes by mean expression
top_genes <- order(rowMeans(counts(dds_4_cvh, normalized = TRUE)), decreasing = TRUE)[1:50]

# 2. Extract vs-transformed matrix
vst_matrix <- assay(vsd_4_cvh)[top_genes, ]

# 3. Extract only leachate annotation
annotation_df <- as.data.frame(colData(dds_4_cvh)[, "leachate", drop = FALSE])

# Ensure it's a factor in correct order: control, high
annotation_df$leachate <- factor(annotation_df$leachate, levels = c("control", "high"))

# 4. Order samples by pvc_leachate_level
ordered_cols <- order(annotation_df$leachate)

# Reorder matrix and annotation
vst_matrix_ordered <- vst_matrix[, ordered_cols]
annotation_ordered <- annotation_df[ordered_cols, , drop = FALSE]

# 5. Plot heatmap
heatmap_colors <- colorRampPalette(rev(brewer.pal(n = 9, "RdBu")))(100)

pheatmap(
  vst_matrix_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  annotation_col = annotation_ordered,
  color = heatmap_colors,
  main = "Top 50 Most Expressed Genes (Control vs High) at 4 hpf"
)

```
### 9 hpf
```{r}
# 1. Select the top 50 genes by mean expression
top_genes <- order(rowMeans(counts(dds_9_cvh, normalized = TRUE)), decreasing = TRUE)[1:50]

# 2. Extract vs-transformed matrix
vst_matrix <- assay(vsd_9_cvh)[top_genes, ]

# 3. Extract only leachate annotation
annotation_df <- as.data.frame(colData(dds_9_cvh)[, "leachate", drop = FALSE])

# Ensure it's a factor in correct order: control, high
annotation_df$leachate <- factor(annotation_df$leachate, levels = c("control", "high"))

# 4. Order samples by pvc_leachate_level
ordered_cols <- order(annotation_df$leachate)

# Reorder matrix and annotation
vst_matrix_ordered <- vst_matrix[, ordered_cols]
annotation_ordered <- annotation_df[ordered_cols, , drop = FALSE]

# 5. Plot heatmap
heatmap_colors <- colorRampPalette(rev(brewer.pal(n = 9, "RdBu")))(100)

pheatmap(
  vst_matrix_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  annotation_col = annotation_ordered,
  color = heatmap_colors,
  main = "Top 50 Most Expressed Genes (Control vs High) at 9 hpf"
)

```
### 14 hpf 
```{r}
# 1. Select the top 50 genes by mean expression
top_genes <- order(rowMeans(counts(dds_14_cvh, normalized = TRUE)), decreasing = TRUE)[1:50]

# 2. Extract vs-transformed matrix
vst_matrix <- assay(vsd_14_cvh)[top_genes, ]

# 3. Extract only leachate annotation
annotation_df <- as.data.frame(colData(dds_14_cvh)[, "leachate", drop = FALSE])

# Ensure it's a factor in correct order: control, high
annotation_df$leachate <- factor(annotation_df$leachate, levels = c("control", "high"))

# 4. Order samples by pvc_leachate_level
ordered_cols <- order(annotation_df$leachate)

# Reorder matrix and annotation
vst_matrix_ordered <- vst_matrix[, ordered_cols]
annotation_ordered <- annotation_df[ordered_cols, , drop = FALSE]

# 5. Plot heatmap
heatmap_colors <- colorRampPalette(rev(brewer.pal(n = 9, "RdBu")))(100)

pheatmap(
  vst_matrix_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  annotation_col = annotation_ordered,
  color = heatmap_colors,
  main = "Top 50 Most Expressed Genes (Control vs High) at 14 hpf"
)

```

# Run DESeq2
## 4 hpf
```{r}
# Wald test for leachate
dds_4_cvh_deseq <- DESeq(dds_4_cvh, minReplicatesForReplace = Inf)

# Count DEGs due to interaction
DESeq2_4_results <- results(dds_4_cvh_deseq, lfcThreshold = 0, alpha = 0.05)

summary(DESeq2_4_results)
```
## 9 hpf 
```{r}
# Wald test for leachate
dds_9_cvh_deseq <- DESeq(dds_9_cvh, minReplicatesForReplace = Inf)

# Count DEGs due to interaction
DESeq2_9_results <- results(dds_9_cvh_deseq, lfcThreshold = 0, alpha = 0.05)

summary(DESeq2_9_results)
```

## 14 hpf
```{r}
# Wald test for leachate
dds_14_cvh_deseq <- DESeq(dds_14_cvh, minReplicatesForReplace = Inf)

# Count DEGs due to interaction
DESeq2_14_results <- results(dds_14_cvh_deseq, lfcThreshold = 0, alpha = 0.05)

summary(DESeq2_14_results)
```


::: callout-important
There are zero differentially expressed genes between the control and high treatment groups within all developmental stages.
:::
