---
title: "Pairwise Differential Gene Expression"
subtitle: "DESeq2 within each developmental stage"
author: "Sarah Tanja"
date: 2025-05-08
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background
We did not find any significantly differently expressed genes across all embryonic stages, we believe, due to the gene expression in our dataset is clustered by stage, and the effect of leachate was 'drowned out' by the much larger patterns of embryonic stage. This code breaks it down to look at DGEs within each stage of embryonic development. This reduces our n to 5 for each group and we have less power with this analysis, but it will help us get a fuller picture of our data and what it's telling us.  
## Inputs
- `gcm_4.csv` is the unfiltered gene count matrix for all 4hpf samples.
- `gcm_9.csv`
- `gcm_14.csv`
- `metadata_4.csv` is the metadata for 61 samples (two outliers were already removed).
- `metadata_9.csv`
- `metadata_14.csv`
## Outputs
- `volcano_facet_pairwise.png`

# Setup
## Install packages
```{r}
BiocManager::install("RUVSeq")
```

## Load packages
```{r}
library(tidyverse)
library(DESeq2)
library(RUVSeq)
library(pheatmap)
library(RColorBrewer)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
```

## Import gene counts
```{r}
gcm_4 <- read.csv("../output/06_tidyup/gcm_4.csv", row.names="gene_id", check.names = FALSE)
gcm_9 <- read.csv("../output/06_tidyup/gcm_9.csv", row.names="gene_id", check.names = FALSE)
gcm_14 <- read.csv("../output/06_tidyup/gcm_14.csv", row.names="gene_id", check.names = FALSE)
```

## Import metadata 
```{r}
metadata_4 <- read.csv("../metadata/metadata_4.csv")
metadata_9 <- read.csv("../metadata/metadata_9.csv")
metadata_14 <- read.csv("../metadata/metadata_14.csv")
```

# 4 hpf
```{r}
str(metadata_4)
```

```{r}
# Reorder leachate factor
metadata_4$leachate <- as.factor(metadata_4$leachate)
metadata_4$leachate <- fct_relevel(metadata_4$leachate, "control", "low", "mid", "high")

# Make parents a character string
metadata_4$parents <- as.character(metadata_4$parents)
str(metadata_4)
```

## Filter low reads
```{r}
# Filter rows where at least 75% of the columns have a raw count of at least 15
gcm_4_filt <- gcm_4 %>%
  filter(rowSums(across(everything(), ~ . > 50)) >= 0.75 * ncol(gcm_4))
nrow(gcm_4_filt)
```

## Make DESeq Object

Make [DESeq Dataset Object from Count Matrix](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#count-matrix-input) Create a DESeqDataSet design from gene count matrix and treatment conditions.

```{r}
dds_4 <- DESeqDataSetFromMatrix(countData = gcm_4_filt,
                              colData = metadata_4, 
                              design = ~leachate)

dds_4 <- estimateSizeFactors(dds_4)
```

Are all size factors less than 4?
```{r}
all(sizeFactors(dds_4))<4
```
Make count transformations (for use in visualizations & clustering)
```{r}
rld_4 <- rlog(dds_4, blind = FALSE) 
vsd_4 <- vst(dds_4, blind = FALSE)
log2_4 <- as.data.frame(log2(counts(dds_4, normalized = TRUE) + 1), check.names = FALSE)
```

Clusters
```{r}
rdists <- dist(t(assay(rld_4)))
plot(hclust(rdists))
```

```{r}
vdists <- dist(t(assay(vsd_4)))
plot(hclust(vdists))
```


### Count transformations
The distribution should shift via the transformation from a negative binomial to a approximately normal distribution.

```{r}
# Combine all transformations into one long-format dataframe
df_4 <- bind_rows(
  as.data.frame(log2(counts(dds_4, normalized = TRUE) + 1)) %>%
    mutate(transformation = "log2(x + 1)"),
  as.data.frame(assay(vsd_4)) %>%
    mutate(transformation = "vst"),
  as.data.frame(assay(rld_4)) %>%
    mutate(transformation = "rlog")
)

# Pivot to long format: one row per expression value
df_long <- df_4 %>%
  pivot_longer(
    cols = -transformation,
    names_to = "sample",
    values_to = "expression"
  )

# Plot
ggplot(df_long, aes(x = expression)) +
  geom_histogram(bins = 100, fill = "steelblue") +
  facet_wrap(~ transformation) +
  theme_classic() +
  labs(
    title = "Distribution of Transformed Counts Across All Samples (4 hpf)",
    x = "Transformed counts",
    y = "Frequency"
  )

```
```{r}
# VST-transformed matrix
vst_mat <- assay(vst(dds_4, blind = TRUE))

# Convert to long format
df_vst <- as.data.frame(vst_mat)
df_vst$gene <- rownames(df_vst)
df_vst_long <- pivot_longer(df_vst, cols = -gene, names_to = "sample", values_to = "expression")

# Plot
ggplot(df_vst_long, aes(x = expression, group = sample)) +
  geom_density(alpha = 0.2) +
  labs(title = "VST Transformed Counts (Density by Sample)")

```
```{r}
# Regularized log-transformed matrix
rld_mat <- assay(rld_4)

# Convert to long format
df_rld <- as.data.frame(rld_mat)
df_rld$gene <- rownames(df_rld)
df_rld_long <- pivot_longer(df_rld, cols = -gene, names_to = "sample", values_to = "expression")

# Plot
ggplot(df_rld_long, aes(x = expression, group = sample)) +
  geom_density(alpha = 0.2) +
  labs(title = "RLOG Transformed Counts (Density by Sample)")
```


### PCA

```{r}
# Get PCA data (top 500 DEGs by default)
pca_4 <- plotPCA(rld_4, intgroup = "group", returnData = TRUE)
percentVar_4 <- round(100*attr(pca_4, "percentVar"))

# Merge with metadata
pca_4 <- merge(pca_4, metadata_4, by.x = "name", by.y = "sample_name", all.x = TRUE)

# Assign specific colors to each leachate
leachate_colors <- c(
  "control" = flex("blue", sat = 200),
  "low" = flex("magenta", sat = 300),
  "mid" = flex("magenta", sat = 600),
  "high" = flex("magenta", sat = 900)
)

# Plot PCA
PCA_4 <- ggplot(pca_4, aes(PC1, PC2, color=leachate)) + 
  geom_point(size=2, alpha = 0.8) +
  ggtitle("M.cap embryos at 4 hpf exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ",percentVar_4[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_4[2],"% variance")) + 
  coord_fixed() +
  scale_color_manual(values=leachate_colors)+
  stat_ellipse()+
  theme_minimal()

PCA_4
```

```{r}
pca_4$label <- paste0("Sample: ", pca_4$sample_name,
                        "\nLeachate: ", pca_4$leachate)

# Original ggplot with an interactive tooltip aesthetic
PCA_4 <- ggplot(pca_4, aes(PC1, PC2, color = leachate, label = name)) + 
  geom_point(shape = 21, size = 2) +
  ggtitle("M.cap embryos at 4 hpf exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ", percentVar_4[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_4[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = leachate_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_4, tooltip = "label")
```
```{r}
# Plot PCA
PCA_4 <- ggplot(pca_4, aes(PC1, PC2, color=parents)) + 
  geom_point(size=2, alpha = 0.8) +
  ggtitle("M.cap embryos at 4 hpf exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ",percentVar_4[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_4[2],"% variance")) + 
  coord_fixed() +
  stat_ellipse()+
  theme_minimal()

PCA_4
```

```{r}
# Save the plot as a png
ggsave("../output/08_deg/PCA_control_vs_high.png", plot = PCA_CvH_plot, width = 8, height = 6, dpi = 600)
```

### Heatmap

```{r}

# 1. Select the top 50 genes by mean expression
top_genes <- order(rowMeans(counts(dds_cvh, normalized = TRUE)), decreasing = TRUE)[1:50]

# 2. Extract rlog-transformed matrix
rlog_matrix <- assay(rld_cvh)[top_genes, ]

# 3. Extract only pvc_leachate_level annotation
annotation_df <- as.data.frame(colData(dds_cvh)[, "pvc_leachate_level", drop = FALSE])

# Ensure it's a factor in correct order: control, high
annotation_df$pvc_leachate_level <- factor(annotation_df$pvc_leachate_level, levels = c("control", "high"))

# 4. Order samples by pvc_leachate_level
ordered_cols <- order(annotation_df$pvc_leachate_level)

# Reorder matrix and annotation
rlog_matrix_ordered <- rlog_matrix[, ordered_cols]
annotation_ordered <- annotation_df[ordered_cols, , drop = FALSE]

# 5. Plot heatmap
heatmap_colors <- colorRampPalette(rev(brewer.pal(n = 9, "RdBu")))(100)

pheatmap(
  rlog_matrix_ordered,
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_rownames = FALSE,
  annotation_col = annotation_ordered,
  color = heatmap_colors,
  main = "Top 50 Most Expressed Genes (Control vs High)"
)

```

### RUVSeq

```{r}
set <- newSeqExpressionSet(as.matrix(filtered),
                           phenoData = data.frame(x, row.names=colnames(filtered)))
set
```


https://bioconductor.org/packages/release/bioc/vignettes/RUVSeq/inst/doc/RUVSeq.html

## Run DESeq2

```{r}
dds_cvh_deseq <- DESeq(dds_cvh)
```

```{r}
summary(results(dds_cvh_deseq))
```

::: callout-important
There are zero differentially expressed genes between the control and high treatment groups across all developmental stages.
:::
