---
title: "Multifactorial Differential Gene Expression"
subtitle: "using DESeq2 contrasts"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Goals



# Setup


## Install packages

```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') 
  install.packages('tidyverse')
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('DESeq2')
if ("apeglm" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('apeglm')
```

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("apeglm")
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(tidyverse)
library(apeglm)
```

## Import gene counts

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r}
gcm <- read.csv("../output/05_count/gene_count_matrix.csv", row.names="gene_id", check.names = FALSE)
head(gcm)
```

::: callout-note
`gcm` is the unfiltered gene count matrix of all 63 samples
:::

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

```{r}
gcm_filt <- gcm %>% 
  dplyr::select(-`131415L4`, -`789C4`)
head(gcm_filt)
```

::: callout-note
`gcm_filt` is the gene count matrix that has been filtered to remove 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA
:::

## Import metadata

```{r}
metadata_filt <- read_csv("../metadata/metadata_filt.csv")
str(metadata_filt)
```
```{r}
# Reorder factors
metadata_filt$embryonic_stage <- as.factor(metadata_filt$embryonic_stage)
metadata_filt$embryonic_stage <- fct_relevel(metadata_filt$embryonic_stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_filt$pvc_leachate_level <- as.factor(metadata_filt$pvc_leachate_level)
metadata_filt$pvc_leachate_level <- fct_relevel(metadata_filt$pvc_leachate_level, "control", "low", "mid", "high")

metadata_filt$hpf <- as.factor(metadata_filt$hours_post_fertilization)
metadata_filt$hpf <- fct_relevel(metadata_filt$hpf, "4", "9", "14")
str(metadata_filt)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples where explanatory variables `hpf` is a factor with 3 levels (4, 9, 14) and `leachate` is a continuous numerical variable representing leachate concentrations with values 0 mg/L, 0.01 mg/L, 0.1 mg/L, and 1 mg/L 
:::


# DESeq2

### Filter
```{r}
# Filter rows where at least 33% of the columns have a raw count of ~15
gcm_filt_dseq <- gcm_filt %>%
  filter(rowSums(across(everything(), ~ . > 50)) >= 0.60 * ncol(gcm_filt))

nrow(gcm_filt_dseq)
```
## Make DESeq2 object
```{r}

dds <- DESeqDataSetFromMatrix(countData = gcm_filt_dseq,
                              colData = metadata_filt,
                              design = formula( ~ 1 + hpf + leachate + hpf:leachate))

# Transform count (variance stabilized) 
vsd <- vst(dds)
```

```{r}
dim(vsd)
all_genes<-t(assay(vsd))
all_genes<-as.data.frame(all_genes)
dim(all_genes)
all_genes$sample<-rownames(all_genes)
all_genes<-all_genes%>%
  dplyr::select(sample, everything())

write_csv(file="../output/09_deg_multifactorial/all_genes.csv", all_genes)
```


### plot distribution of filtered, normalized reads
```{r}

# Extract VST-transformed expression matrix
vsd_mat <- assay(vsd)

# Flatten it into a vector
vsd_vals <- as.vector(vsd_mat)

# Create a dataframe for ggplot
df_vsd <- data.frame(vsd_count = vsd_vals)

# Plot
ggplot(df_vsd, aes(x = vsd_count)) +
  geom_histogram(fill = "grey") +
  theme_classic() +
  labs(
    title = "Histogram of Variance-Stabilized Counts",
    x = "VST-transformed read count",
    y = "Frequency"
  )

```
VST compresses the high-end of the count distribution and expands the low-end, making the variance roughly constant across the range.

The resulting values are on an approximate log2 scale but not exact log2(count + 1) — VST is more nuanced, especially for low counts.

A range of 5 to 15 corresponds roughly to raw counts ranging from 2³² to 2¹⁵ ≈ 32 to 32,000, so it’s realistic for RNA-seq gene expression data.

### Run DESeq Test
```{r}
dds <- DESeq(dds)
resultsNames(dds)  # see what contrast names are available
```
```{r}
plotDispEsts(dds)
```

```{r}
# Main effect of leachate (baseline hpf level) result & summary
res_leachate <- results(dds, name = "leachate", alpha=0.05)
summary(res_leachate)
```
```{r}
plotMA(res_leachate)
```
In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored blue if the adjusted p value is less than 0.1. Points which fall out of the window are plotted as open triangles pointing either up or down. It is more useful to visualize the MA-plot for the shrunken log2 fold changes, which remove the noise associated with log2 fold changes from low count genes without requiring arbitrary filtering thresholds.
```{r}
citation("apeglm")
```


```{r}
res_leachate_shrunk <- lfcShrink(dds, coef="leachate", type="apeglm")
res_leachate_shrunk
```


```{r}
# Extract significant genes (alpha < 0.05)
sig_leachate <- as.data.frame(subset(res_leachate, padj<0.05))
sig_leachate$gene <- rownames(sig_leachate)
rownames(sig_leachate) <- NULL
# Save the gene list (and test statistics!)
write_csv(sig_leachate, file = "../output/09_deg_multifactorial/sig_leachate_list.csv")


## Make and save a corresponding gene count matrix of only significant genes

# Extract normalized gene count matrix from dds object
dds_counts <- counts(dds, normalized = TRUE)
# Subset to significant genes only
sig_leachate_counts <- as.data.frame(dds_counts[rownames(dds_counts) %in% sig_leachate$gene, ])
# Save gene counts 
write_csv(sig_leachate_counts, file = "../output/09_deg_multifactorial/sig_leachate_counts.csv")
```

```{r}
res <- results(dds)

DEG_results_vst <- dds[unique(res$gene)]
dim(DEG_results_vst)

DEG_results_vst <- varianceStabilizingTransformation(DEG_results_vst)
```




```{r}
dds <- estimateDispersionsGeneEst(dds)
dispersions(dds) <- mcols(dds)$dispGeneEst
dds <- nbinomLRT(dds, reduced=~hpf+leachate) 
```


```{r}
deg_counts<-t(assay(DEG_results_vst))
deg_counts<-as.data.frame(deg_counts)
dim(deg_counts)
deg_counts$sample<-rownames(deg_counts)
deg_counts<-deg_counts%>%
  dplyr::select(sample, everything())
write_csv(deg_counts, "Mcap2020/Output/TagSeq/DESeq/deg_counts.csv")
```

```{r}
dput(head(rownames(dds_counts), 10))
```

```{r}
sig_leachate$gene %in% rownames(dds_counts)
```


```{r}
plotMA(res_leachate, alpha = 0.05, ylim = c(-5, 5), main = "MA Plot: Leachate Effect")
```

```{r}
# Main effect of hpf levels (because hpf is a factor with levels  4, 9, 14)
res_hpf9 <- results(dds, name = "hpf_9_vs_4")  # if reference is hpf4
summary(res_hpf9)

sig_hpf9 <- as.data.frame(subset(res_hpf9, padj<0.05))
sig_hpf9$gene <- rownames(sig_hpf9)
rownames(sig_hpf9) <- NULL
write_csv(sig_hpf9, file = "../output/09_deg_multifactorial/sig_hpf9.csv")

plotMA(res_hpf9, ylim = c(-5, 5), main = "MA Plot: hpf 9 vs 4")
```

```{r}
# Main effect of hpf levels (becuase hpf is a factor with levels 4, 9, 14)
res_hpf14 <- results(dds, name = "hpf_14_vs_4")
summary(res_hpf14)

sig_hpf14 <- as.data.frame(subset(res_hpf14, padj<0.05))
sig_hpf14$gene <- rownames(sig_hpf14)
rownames(sig_hpf14) <- NULL
write_csv(sig_hpf14, file = "../output/09_deg_multifactorial/sig_hpf14.csv")

plotMA(res_hpf14, ylim = c(-5, 5), main = "MA Plot: hpf 14 vs 4")
```

```{r}
# Interaction effects
res_interaction_9 <- results(dds, name = "hpf9.leachate")
summary(res_interaction_9)

sig_interaction_9 <- as.data.frame(subset(res_interaction_9, padj<0.05))
sig_interaction_9$gene <- rownames(sig_interaction_9)
rownames(sig_interaction_9) <- NULL
write_csv(sig_interaction_9, file = "../output/09_deg_multifactorial/sig_interaction_9.csv")

plotMA(res_interaction_9, ylim = c(-5, 5), main = "MA Plot: hpf9 x leachate interaction")
```

```{r}
res_interaction_14 <- results(dds, name = "hpf14.leachate")
summary(res_interaction_14)

sig_interaction_14 <- as.data.frame(subset(res_interaction_14, padj<0.05))
sig_interaction_14$gene <- rownames(sig_interaction_14)
rownames(sig_interaction_14) <- NULL
write_csv(sig_interaction_14, file = "../output/09_deg_multifactorial/sig_interaction_14.csv")


plotMA(res_interaction_14, ylim = c(-5, 5), main = "MA Plot: hpf14 x leachate interaction")
```

