---
title: "Multifactorial Differential Gene Expression"
subtitle: "using DESeq2 contrasts"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background
## Inputs
- `gcm.csv` is the unfiltered gene count matrix for all 63 samples.
- `metadata_filt.csv` is the metadata for 61 samples (two outliers were already removed).
## Outputs
- `volcano_facet.png`
# Setup

## Install packages

```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') 
  install.packages('tidyverse')
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('DESeq2')
if ("ashr" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('ashr')
if ("plotly" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('plotly')
if ("cowplot" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('cowplot')
if ("pak" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('pak')
```

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("EnhancedVolcano")
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(tidyverse)
library(EnhancedVolcano)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
pak::pak("mdscheuerell/flexoki")
```

## Import gene counts

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r}
gcm <- read.csv("../output/05_count/gene_count_matrix.csv", row.names="gene_id", check.names = FALSE)
head(gcm)
```

::: callout-note
`gcm` is the unfiltered gene count matrix of all 63 samples
:::

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

```{r}
gcm_filt <- gcm %>% 
  dplyr::select(-`131415L4`, -`789C4`)
head(gcm_filt)
```

::: callout-note
`gcm_filt` is the gene count matrix that has been filtered to remove 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA
:::

## Import metadata

```{r}
metadata_filt <- read_csv("../metadata/metadata_filt.csv")
str(metadata_filt)
```

```{r}
# Reorder factors
metadata_filt$embryonic_stage <- as.factor(metadata_filt$embryonic_stage)
metadata_filt$embryonic_stage <- fct_relevel(metadata_filt$embryonic_stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_filt$leachate <- as.factor(metadata_filt$leachate_lvl)
metadata_filt$leachate <- fct_relevel(metadata_filt$leachate_lvl, "control", "low", "mid", "high")

metadata_filt$hpf <- as.factor(metadata_filt$hpf)
metadata_filt$hpf <- fct_relevel(metadata_filt$hpf, "4", "9", "14")
str(metadata_filt)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples where explanatory variables `hpf` is a factor with 3 levels (4, 9, 14) and `leachate` is a continuous numerical variable representing leachate concentrations with values 0 mg/L, 0.01 mg/L, 0.1 mg/L, and 1 mg/L
:::

# Filter

```{r}
# Filter rows where at least 75% of the columns have a raw count of ~50
gcm_filt <- gcm_filt %>%
  filter(rowSums(across(everything(), ~ . > 120)) >= 0.90 * ncol(gcm_filt))

nrow(gcm_filt)
```

::: callout-note
This is an aggressive filter!
:::

# Make DESeq2 object

```{r}

dds <- DESeqDataSetFromMatrix(countData = gcm_filt,
                              colData = metadata_filt,
                              design = formula( ~ 1 + hpf + leachate + hpf:leachate))
```

## Explore PCAs

::: callout-note
73% of variance explained by dim1 (0.57) and dim2 (0.16) in untransformed counts
:::

```{r}
plotPCA(vst(dds), intgroup = "group", returnData = TRUE)
```

pca_all \<- plotPCA(vst(dds), intgroup = "group", returnData = TRUE, ntop=nrow(all))

```{r}
# Get PCA data (top 500 DEGs by default)
pca_500 <- plotPCA(vst(dds), intgroup = "group", returnData = TRUE)
percentVar_500 <- round(100*attr(pca_500, "percentVar"))

# Merge with metadata
pca_500 <- merge(pca_500, metadata_filt, by.x = "name", by.y = "sample_name", all.x = TRUE)


# Assign specific colors to each hpf
hpf_colors <- c(
  "4" = flex("cyan", sat=500),
  "9" = flex("green", sat =500),
  "14" = flex("yellow", sat=500)
)

# Plot PCAs
PCA_500 <- ggplot(pca_500, aes(PC1, PC2, color=hpf, shape=leachate)) + 
  geom_point(size=2, alpha = 0.8) +
  ggtitle("M.cap embryos exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ",percentVar_500[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_500[2],"% variance")) + 
  coord_fixed() +
  scale_color_manual(values=hpf_colors)+
  stat_ellipse()+
  theme_minimal()

PCA_500
```

```{r}
pca_500$label <- paste0("Sample: ", pca_500$sample_name,
                        "\nHPF: ", pca_500$hpf,
                        "\nLeachate: ", pca_500$leachate)

leachate_alpha_map <- c("control" = 0.3, "low" = 0.5, "mid" = 0.7, "high" = 1)
pca_500$alpha_val <- leachate_alpha_map[pca_500$leachate]
```

```{r}
# Original ggplot with a tooltip aesthetic
PCA_500 <- ggplot(pca_500, aes(PC1, PC2, color = hpf, label = name, alpha = alpha_val)) + 
  geom_point(shape = 21, size = 2) +
  ggtitle("M.cap embryos exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ", percentVar_500[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_500[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_500, tooltip = "label")
```

## Plot distribution

```{r}

# Extract VST-transformed expression matrix
vsd_mat <- assay(vst(dds))

# Flatten it into a vector
vsd_vals <- as.vector(vsd_mat)

# Create a dataframe for ggplot
df_vsd <- data.frame(vsd_count = vsd_vals)

# Plot
ggplot(df_vsd, aes(x = vsd_count)) +
  geom_histogram(fill = "grey") +
  theme_classic() +
  labs(
    title = "Histogram of Variance-Stabilized Counts",
    x = "VST-transformed read count",
    y = "Frequency"
  )

```

VST compresses the high-end of the count distribution and expands the low-end, making the variance roughly constant across the range. The resulting values are on an approximate log2 scale but not exact log2(count + 1) --- VST is more nuanced, especially for low counts. A range of 5 to 15 corresponds roughly to raw counts ranging from 2³² to 2¹⁵ ≈ 32 to 32,000, so it's realistic for RNA-seq gene expression data.

# Run DESeq Test

This calculates the log2fold change between factors

```{r}
dds <- DESeq(dds, test="LRT", reduced=~1)
```

## Plot dispersion

```{r}
plotDispEsts(dds)
```

::: callout-note
Dispersion trend line is hovering around 0.1, and bulk spread of tag-wise gene estimates is from 0.02 to 0.5
:::

# Build contrasts

```{r}
resultsNames(dds)  # see what contrast names are available
```

| Coef | Name                         | Meaning                                                              |
|------|------------------------------|----------------------------------------------------------------------|
| 1    | `"Intercept"`                | Baseline expression: `hpf = 4` and `leachate = control`              |
| 2    | `"hpf_9_vs_4"`               | Change from 4 hpf → 9 hpf, at **control** exposure                   |
| 3    | `"hpf_14_vs_4"`              | Change from 4 hpf → 14 hpf, at **control** exposure                  |
| 4    | `"leachate_low_vs_control"`  | Change from control → low exposure, at **4 hpf**                     |
| 5    | `"leachate_mid_vs_control"`  | Change from control → mid exposure, at **4 hpf**                     |
| 6    | `"leachate_high_vs_control"` | Change from control → high exposure, **4 hpf**                       |
| 7    | `"hpf9.leachatelow"`         | **Interaction term**: extra effect of low exposure at 9 hpf vs 4 hpf |
| 8    | `"hpf14.leachatelow"`        | Extra effect at 14 hpf vs 4 hpf                                      |
| 9    | `"hpf9.leachatemid"`         | Extra effect of mid exposure at 9 hpf vs 4 hpf                       |
| 10   | `"hpf14.leachatemid"`        | ... at 14 hpf                                                        |
| 11   | `"hpf9.leachatehigh"`        | ... for high                                                         |
| 12   | `"hpf14.leachatehigh"`       | ... for high at 14 hpf                                               |

To get the effect of leachate at a specific time: 

| What you want            | Use this contrast                                    | 
| -------------------------|------------------------------------------------------| 
| Low vs Control at 4 hpf  | `"leachate_low_vs_control"`                          | 
| Low vs Control at 9 hpf  | `c("leachate_low_vs_control", "hpf9.leachatelow")`   | 
| Mid vs Control at 14 hpf | `c("leachate_mid_vs_control", "hpf14.leachatemid")`  | 
| High vs Control at 9 hpf | `c("leachate_high_vs_control", "hpf9.leachatehigh")` |

## Effect of high exposure at 14 hpf

When you test for main effects of leachate (e.g., "leachate_high_vs_control"), it's at the reference timepoint (hpf = 4). The interaction terms model how the leachate effect changes relative to that baseline as development progresses.

What's the effect of high exposure at 14 hpf? You must combine: - "leachate_high_vs_control": main leachate effect at 4 hpf - "hpf14.leachatehigh": additional effect at 14 hpf

```{r}
# output results of contrast
res_14hpf_high <- (results(dds, contrast = list(
  c("leachate_high_vs_control", "hpf14.leachatehigh")
)))

summary(res_14hpf_high)
```

### Extract significant genes

```{r}
padj.cutoff <- 0.05
lfc.cutoff <- 0.5
```

```{r}
# Convert results table into tibble
res_14hpf_high_tb <- res_14hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_high_noLFCcutoff <- res_14hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_high <- sig_14hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at high exposure level:", nrow(sig_14hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at high exposure level:", nrow(sig_14hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_high, "../output/09_deg_multifactorial/sig_14hpf_high.csv")
write_csv(sig_14hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_14hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf14.leachatehigh")), type = "ashr")

plotMA(res_14hpf_high)
plotMA(res_14hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_high,
    lab = rownames(res_14hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14high <- 
  ggplot(res_14hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_high[sig_14hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_high[sig_14hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14high
```

## Effect of mid exposure at 14 hpf

```{r}
# output results of contrast
res_14hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf14.leachatemid")
)))

summary(res_14hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_14hpf_mid_tb <- res_14hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_mid_noLFCcutoff <- res_14hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_mid <- sig_14hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at mid exposure level:", nrow(sig_14hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at mid exposure level:", nrow(sig_14hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_mid, "../output/09_deg_multifactorial/sig_14hpf_mid.csv")
write_csv(sig_14hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}

# shrink lfc for visualizations
res_14hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf14.leachatemid")), type = "ashr")

plotMA(res_14hpf_mid)
plotMA(res_14hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_mid,
    lab = rownames(res_14hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14mid <- 
  ggplot(res_14hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_mid[sig_14hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_mid[sig_14hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14mid
```

## Effect of low exposure at 14 hpf

```{r}
# output results of contrast
res_14hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf14.leachatelow")
)))

summary(res_14hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_14hpf_low_tb <- res_14hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_low_noLFCcutoff <- res_14hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_low <- sig_14hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at low exposure level:", nrow(sig_14hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at low exposure level:", nrow(sig_14hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_low, "../output/09_deg_multifactorial/sig_14hpf_low.csv")
write_csv(sig_14hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_14hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf14.leachatelow")), type = "ashr")

plotMA(res_14hpf_low)
plotMA(res_14hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_low,
    lab = rownames(res_14hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14low <- 
  ggplot(res_14hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_low[sig_14hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_low[sig_14hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14low
```

## Effect of high exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_high <- (results(dds, contrast = list(
  c("leachate_high_vs_control", "hpf9.leachatehigh")
)))

summary(res_9hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_high_tb <- res_9hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_high_noLFCcutoff <- res_9hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_high <- sig_9hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_high, "../output/09_deg_multifactorial/sig_9hpf_high.csv")
write_csv(sig_9hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf9.leachatehigh")), type = "ashr")

plotMA(res_9hpf_high)
plotMA(res_9hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_high,
    lab = rownames(res_9hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9high <- 
  ggplot(res_9hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9high
```

## Effect of mid exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf9.leachatemid")
)))

summary(res_9hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_mid_tb <- res_9hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_mid_noLFCcutoff <- res_9hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_mid <- sig_9hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_mid, "../output/09_deg_multifactorial/sig_9hpf_mid.csv")
write_csv(sig_9hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf9.leachatemid")), type = "ashr")

plotMA(res_9hpf_mid)
plotMA(res_9hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_mid,
    lab = rownames(res_9hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9mid <- 
  ggplot(res_9hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9mid
```

## Effect of low exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf9.leachatelow")
)))

summary(res_9hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_low_tb <- res_9hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_low_noLFCcutoff <- res_9hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_low <- sig_9hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_low, "../output/09_deg_multifactorial/sig_9hpf_low.csv")
write_csv(sig_9hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf9.leachatelow")), type = "ashr")

plotMA(res_9hpf_low)
plotMA(res_9hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_low,
    lab = rownames(res_9hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9low <- 
  ggplot(res_9hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9low
```

## Effect of high exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_high <- (results(dds, name = "leachate_high_vs_control")
)

summary(res_4hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_high_tb <- res_4hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_high_noLFCcutoff <- res_4hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_high <- sig_4hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_high, "../output/09_deg_multifactorial/sig_4hpf_high.csv")
write_csv(sig_4hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_high_ashr <- lfcShrink(dds, coef = 6 , type = "ashr")

plotMA(res_4hpf_high)
plotMA(res_4hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_high,
    lab = rownames(res_4hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4high <- 
  ggplot(res_4hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4high
```

## Effect of mid exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_mid <- (results(dds, name = "leachate_mid_vs_control")
)

summary(res_4hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_mid_tb <- res_4hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_mid_noLFCcutoff <- res_4hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_mid <- sig_4hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_mid, "../output/09_deg_multifactorial/sig_4hpf_mid.csv")
write_csv(sig_4hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_mid_ashr <- lfcShrink(dds, coef = 5 , type = "ashr")

plotMA(res_4hpf_mid)
plotMA(res_4hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_mid,
    lab = rownames(res_4hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4mid <- 
  ggplot(res_4hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4mid
```

## Effect of low exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_low <- (results(dds, name = "leachate_low_vs_control", alpha=0.05)
)

summary(res_4hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_low_tb <- res_4hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_low_noLFCcutoff <- res_4hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_low <- sig_4hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_low, "../output/09_deg_multifactorial/sig_4hpf_low.csv")
write_csv(sig_4hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_low_ashr <- lfcShrink(dds, coef = 4 , type = "ashr")
plotMA(res_4hpf_low)
plotMA(res_4hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_low,
    lab = rownames(res_4hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4low <- 
  ggplot(res_4hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4low
```

# Summary

## Facet volcano plot

```{r}
# First, strip labels from all individual ggplot objects
# (You can do this in a loop if the plots are in a list)

remove_labels <- function(p) {
  p +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 8)  # adjust title size here
    )
}

# Example: if your plots are named v4control, v4low, etc.
plots <- list( 
   v4low, v9low, v14low,
   v4mid, v9mid, v14mid,
   v4high, v9high, v14high
)

plots_clean <- lapply(plots, remove_labels)

# Now plot in 3 rows × 3 columns
volcano_facet <- plot_grid(
  plotlist = plots_clean,
  nrow = 3,
  ncol = 3,
  labels = NULL,
  label_size = 0,
  align = "hv"
)

volcano_facet
```

```{r}
ggsave("../output/09_deg_multifactorial/volcano_facet.png", volcano_facet, width = 16, height = 12, dpi = 300)
```

::: callout-warning
Why do volcano plots not look like volcanos!?

Batch Effects or Low Biological Signal If your experimental design has confounding factors (e.g., batch effects), or the leachate treatment has low impact, then few genes will show strong changes.

Extreme p-values Your plot shows points with -log10(padj) values above 20--30. That means padj is very close to 0, which could indicate: - Very large sample size - Very consistent effect - Or something off with the dispersion estimate or p-value adjustment (e.g., if some genes had zero variance across groups)
:::
