---
title: "Functional Enrichment of Differentially Expressed Genes"
subtitle: "working on the pipeline"
author: "Sarah Tanja"
date: 2025-06-17
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Goals
Functional enrichment of genes identified as differentially expressed

## Prepare functional enrichment dataframes 

Read in files 

```{r}
all_genes<-read.csv("../output/09_deg_multifactorial/all_genes.csv")
```

Read in file of DEG's 
```{r}
DEG_list<-read.csv("../output/09_deg_multifactorial/sig_leachate.csv")
DEG_counts<-read.csv("Mcap2020/Output/TagSeq/DESeq/deg_counts.csv")
```

Get gene length information.  
```{r}
#import file
gff <- rtracklayer::import("Mcap2020/Data/TagSeq/Montipora_capitata_HIv3.genes.gff3") #if this doesn't work, restart R and try again 

transcripts <- subset(gff, type == "transcript") #keep only transcripts 

transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE) #extract length information

transcript_lengths <- width(transcripts_gr) #isolate length of each gene

seqnames<-transcripts_gr$ID #extract list of gene id 

lengths<-cbind(seqnames, transcript_lengths)

lengths<-as.data.frame(lengths) #convert to data frame
```

Read in annotation file. 

Downloaded functional annotation (version 3) from: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz (EggNog) and http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz (KEGG). 
Unzipped on desktop. I then removed the # in #query in the first column, otherwise it does not read in column names.  

```{r}
Mcap.annot <- read.table("Mcap2020/Data/TagSeq/Montipora_capitata_HIv3.genes.EggNog_results.txt",  quote="", sep="\t", header=TRUE)
dim(Mcap.annot)

# column names
head(Mcap.annot)
tail(Mcap.annot)

Mcap.annot<-Mcap.annot%>%
  rename(gene=query)

names(Mcap.annot)
```

This functional annotation has 24,072 genes. This is 44% of total protein predicted genes described in the publication associated with this annotation. https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755 

Match up genes in gene list file to annotation file
```{r}
names(Mcap.annot)

#remove sample column name 
probes = names(all_genes)
#probes <- gsub("sample", "", probes)
#probes <- probes[probes != ""]

probes2annot = match(probes, Mcap.annot$gene) #match genes in datExpr to genes in annotation file, note I removed the # before query in this file before loading

# The following is the number of probes without annotation 
sum(is.na(probes2annot))

row_nas<-which(is.na(probes2annot))

#view the genes that do not have a match in the annotation file
missing<-as.data.frame(probes[row_nas])
#print(missing)
```

There are 3395 of our genes that were expressed that are not present in the annotation file.  

Reduce annotation file to only contain genes detected in our dataset.  

```{r}
filtered_Mcap.annot <- Mcap.annot[Mcap.annot$gene %in% probes, ]
dim(filtered_Mcap.annot)
```

The annotation file now only contains genes that were detected in our dataset that have annotation information. This is 8081 out of the 11475 genes in our dataset. 

Add in length to the annotation file.    

```{r}
filtered_Mcap.annot$length<-lengths$transcript_lengths[match(filtered_Mcap.annot$gene, lengths$seqnames)]

which(is.na(filtered_Mcap.annot$length)) #all genes have lengths 
```

All have length information.  

Format GO terms to remove dashes and quotes.  
```{r}
#remove dashes, remove quotes from columns and separate by semicolons (replace , with ;) in KEGG_ko, KEGG_pathway, and Annotation.GO.ID columns
head(filtered_Mcap.annot$GOs)
filtered_Mcap.annot$GOs <- gsub(",", ";", filtered_Mcap.annot$GOs)
filtered_Mcap.annot$GOs <- gsub('"', "", filtered_Mcap.annot$GOs)
filtered_Mcap.annot$GOs <- gsub("-", NA, filtered_Mcap.annot$GOs)
```

## Conduct enrichment for individual clusters 

Conduct enrichment for clusters 

### Conduct enrichment for each cluster  

Merge in relevant annotation information to the DEG list file.  

Merge to only keep genes that were identified to belong to a cluster. 
```{r}
group_list<-left_join(group, filtered_Mcap.annot, by="gene")
```

Conduct functional enrichment using goseq and rrvgo. Compare against all genes detected in the dataset that we have annotation information for.

Conduct at biological process level. Filter for adjusted p-value <0.05.  
```{r}
cluster_list<-levels(as.factor(group_list$cluster))

for (cluster_id in cluster_list) {
  
    ### Generate vector with names of all genes detected in our dataset 
    ALL.vector <- c(filtered_Mcap.annot$gene)
  
    ### Generate length vector for all genes 
    LENGTH.vector <- as.integer(filtered_Mcap.annot$length)
    
    ### Generate vector with names in just the contrast we are analyzing
    ID.vector <- group_list%>%
      filter(cluster==cluster_id)%>%
      pull(gene)
    
    ##Get a list of GO Terms for all genes detected in our dataset 
    GO.terms <- filtered_Mcap.annot%>%
      dplyr::select(gene, GOs)
    
    ##Format to have one goterm per row with gene ID repeated
    split <- strsplit(as.character(GO.terms$GOs), ";") 
    split2 <- data.frame(v1 = rep.int(GO.terms$gene, sapply(split, length)), v2 = unlist(split)) 
    colnames(split2) <- c("gene", "GOs")
    GO.terms<-split2

    ##Construct list of genes with 1 for genes in contrast and 0 for genes not in the contrast
    gene.vector=as.integer(ALL.vector %in% ID.vector) 
    names(gene.vector)<-ALL.vector#set names

    #weight gene vector by bias for length of gene: load in the gene.vector(a list of all genes with 1 indicating it is a DEG in the group of interest and 0 meaning it is not a DEG) and bias.data (list of lengths for all genes)
    pwf<-nullp(gene.vector, bias.data=LENGTH.vector) 
    
    #run goseq using Wallenius method for all categories of GO terms: load in the pwf object (gene name, a 1 or 0 to indicate DEG, the length, and the pwf function; generated from pwf above), and the list of GO terms for all genes
    GO.wall<-goseq(pwf, gene2cat=GO.terms, test.cats=c("GO:BP"), method="Wallenius", use_genes_without_cat=TRUE)

    GO <- GO.wall[order(GO.wall$over_represented_pvalue),]
    colnames(GO)[1] <- "GOterm"
    
    #adjust p-values 
    GO$bh_adjust <- p.adjust(GO$over_represented_pvalue, method="BH") #add adjusted p-values
  
    #Filtering for p < 0.05
    GO <- GO %>%
        dplyr::filter(bh_adjust<0.05) %>%
        dplyr::arrange(., ontology, bh_adjust)

    #Write file of results 
    write_csv(GO, file = paste0("Mcap2020/Output/TagSeq/DESeq/functional_enrichment/goseq_cluster", cluster_id, ".csv"))
    
}
```

```{r}
length(ALL.vector)
length(LENGTH.vector)
dim(pwf)
length(unique(GO.terms$gene))
length(gene.vector)
length(ID.vector)
```
All the dimensions of the input dataframes count to 8081, the expected number of genes. ID vector is your vector of DEGs, which is used to make the gene.vector. 

Clusters 1, 2, 5, 11, 13, 14, 15, 18, 19, 21, 32, 40, 60, 62, 70 have at least one enriched term when analyzed alone. Proceed with conducting enrichment at the level of developmental pattern instead. 

Check the number of genes in each cluster. 
```{r}
group_list%>%
      group_by(cluster, pattern)%>%
      summarise(n=length(unique(gene)))%>%
      write_csv(file="Mcap2020/Output/TagSeq/DESeq/cluster_counts.csv")

names(group_list)
```


## Conduct enrichment for cluster developmental patterns 

Combine cluster gene lists for early, mid, and late developmental patterns and conduct enrichment on these three categories.  

Conduct at biological process level. Filter for adjusted p-value <0.05.  
```{r}
pattern_list<-levels(as.factor(group_list$pattern))

for (pattern_id in pattern_list) {
  
    ### Generate vector with names of all genes detected in our dataset 
    ALL.vector <- c(filtered_Mcap.annot$gene)
    
    ### Generate length vector for all genes 
    LENGTH.vector <- as.integer(filtered_Mcap.annot$length)
    
    ### Generate vector with names in just the contrast we are analyzing
    ID.vector <- group_list%>%
      filter(pattern==pattern_id)%>%
      pull(gene)
    
    ##Get a list of GO Terms for all genes detected in our dataset 
    GO.terms <- filtered_Mcap.annot%>%
      dplyr::select(gene, GOs)
    
    ##Format to have one goterm per row with gene ID repeated
    split <- strsplit(as.character(GO.terms$GOs), ";") 
    split2 <- data.frame(v1 = rep.int(GO.terms$gene, sapply(split, length)), v2 = unlist(split)) 
    colnames(split2) <- c("gene", "GOs")
    GO.terms<-split2

    ##Construct list of genes with 1 for genes in contrast and 0 for genes not in the contrast
    gene.vector=as.integer(ALL.vector %in% ID.vector) 
    names(gene.vector)<-ALL.vector#set names

    #weight gene vector by bias for length of gene 
    pwf<-nullp(gene.vector, bias.data=LENGTH.vector) 
    
    #run goseq using Wallenius method for all categories of GO terms 
    GO.wall<-goseq(pwf, gene2cat=GO.terms, test.cats=c("GO:BP"), method="Wallenius", use_genes_without_cat=TRUE)

    GO <- GO.wall[order(GO.wall$over_represented_pvalue),]
    colnames(GO)[1] <- "GOterm"
    
    #adjust p-values 
    GO$bh_adjust <-  p.adjust(GO$over_represented_pvalue, method="BH") #add adjusted p-values
  
    #Filtering for p < 0.05
    GO <- GO %>%
        dplyr::filter(bh_adjust<0.05) %>%
        dplyr::arrange(., ontology, bh_adjust)

    #Write file of results 
    write_csv(GO, file = paste0("Mcap2020/Output/TagSeq/DESeq/functional_enrichment/goseq_", pattern_id, ".csv"))
    
}
```

Create an empty dataframe to populate (only do this once).  

```{r}
pattern_go_results <- data.frame()
```

Run a loop to filter enrichment for each temperature contrast using rrvgo. 

```{r}
pattern_list<-c("Early", "Mid", "Late")

for (pattern_id in pattern_list) {

    #Read relevant file of results from goseq analysis  
    go_results<-read_csv(file = paste0("Mcap2020/Output/TagSeq/DESeq/functional_enrichment/goseq_", pattern_id, ".csv"))
    
    go_results<-go_results%>%
      filter(ontology=="BP")%>%
      filter(bh_adjust != "NA") %>%
      arrange(., bh_adjust)
    
    print(pattern_id)
    print(length(unique(go_results$GOterm)))
    
    #Reduce/collapse GO term set with the rrvgo package 
      simMatrix <- calculateSimMatrix(go_results$GOterm,
                                orgdb="org.Ce.eg.db", #c. elegans database
                                ont=c("BP"),
                                method="Rel")
    #calculate similarity 
    scores <- setNames(-log10(go_results$bh_adjust), go_results$GOterm)
    reducedTerms <- reduceSimMatrix(simMatrix,
                                scores,
                                threshold=0.7,
                                orgdb="org.Ce.eg.db")
    
    #keep only the goterms from the reduced list
    go_results<-go_results%>%
      filter(GOterm %in% reducedTerms$go)
    
    #add in parent terms to list of go terms 
    go_results$ParentTerm<-reducedTerms$parentTerm[match(go_results$GOterm, reducedTerms$go)]
    
    #add in contrast and regulation
    go_results$pattern<-pattern_id
    
    print(length(unique(go_results$GOterm)))
    
    #add to data frame 
    go_df <- go_results
    pattern_go_results <- rbind(pattern_go_results, go_df) 

}
```

Semantic similarity reduced GO terms for each cluster.    

```{r}
length(unique(pattern_go_results$GOterm))
```
There are 465 total enriched GO terms detected when analyzed by pattern groups. 

## Plot enrichment for each developmental pattern

Early development
```{r}
early_data<-pattern_go_results %>% 
  filter(pattern=="Early")

# First, let's find the unique ParentTerm values
unique_early_terms <- unique(early_data$ParentTerm)

# Determine the number of unique ParentTerm values
num_unique <- length(unique_early_terms)

# Determine the midpoint for splitting into two batches
midpoint <- ceiling(num_unique / 2)
midpoint <- 9

early_data <- early_data %>%
  mutate(batch = ifelse(match(ParentTerm, unique_early_terms) <= midpoint, 1, 2))

early_data$ParentTerm <- factor(early_data$ParentTerm) 
early_data$ParentTerm <- factor(early_data$ParentTerm, levels = rev(levels(early_data$ParentTerm)))

go_plot_earlyA <- early_data %>% 
  filter(batch=="1")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)%>%
  
  ggplot(aes(x=percHits, y=term, fill=bh_adjust, colour=bh_adjust, size=percHits)) +
  facet_grid(ParentTerm ~ ., scales = "free", space='free', labeller = label_wrap_gen(width = 30, multi_line = TRUE))+
  geom_point(aes(group=term))+
  geom_segment(aes(x=0, xend=percHits, y=term, yend=term),linewidth=1) +
  scale_colour_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_fill_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_size(name="Hits (%)", range = c(1, 5))+
  theme_classic() +
  ylab("GO Term")+
  xlab("Hits (%)")+
  ggtitle("Early Development (1-38 hpf); 2,679 genes")+
  theme(legend.position = "none",
        axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=18),
        axis.title.y = element_text(size=24, face="bold"),
        axis.title.x = element_text(size = 24, face = "bold"),
        plot.title = element_text(size = 24, face = "bold"), 
        strip.text.y.right=element_text(angle=0, size=18), 
        legend.title=element_text(size=22),
        legend.text=element_text(size=18)); go_plot_earlyA

go_plot_earlyB <- early_data %>% 
  filter(batch=="2")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)%>%
  
  ggplot(aes(x=percHits, y=term, fill=bh_adjust, colour=bh_adjust, size=percHits)) +
  facet_grid(ParentTerm ~ ., scales = "free", space='free', labeller = label_wrap_gen(width = 30, multi_line = TRUE))+
  geom_point(aes(group=term))+
  geom_segment(aes(x=0, xend=percHits, y=term, yend=term),linewidth=1) +
  scale_colour_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_fill_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_size(name="Hits (%)", range = c(1, 5))+
  theme_classic() +
  ylab("")+
  xlab("Hits (%)")+
  ggtitle("")+
  theme(legend.position = "right",
        axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=18),
        axis.title.y = element_text(size=24, face="bold"),
        axis.title.x = element_text(size = 24, face = "bold"),
        plot.title = element_text(size = 24, face = "bold"), 
        strip.text.y.right=element_text(angle=0, size=18), 
        legend.title=element_text(size=22),
        legend.text=element_text(size=18)); go_plot_earlyB

go_plot_early<-plot_grid(go_plot_earlyA, go_plot_earlyB, ncol=2, rel_widths=c(0.85,1))

ggsave(filename="Mcap2020/Figures/TagSeq/DESeq/functional_enrichment/go_early.pdf", go_plot_early, width=35, height=48)

early_data<-pattern_go_results %>% 
  filter(pattern=="Early")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)

write_csv(early_data, file="Mcap2020/Output/TagSeq/DESeq/functional_enrichment/enrichment_results_early.csv")
```

Mid development
```{r}
go_plot_mid <- pattern_go_results %>% 
  filter(pattern=="Mid")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)%>%
  
  ggplot(aes(x=percHits, y=term, fill=bh_adjust, colour=bh_adjust, size=percHits)) +
  facet_grid(ParentTerm ~ ., scales = "free", space='free', labeller = label_wrap_gen(width = 22, multi_line = TRUE))+
  geom_point(aes(group=term))+
  geom_segment(aes(x=0, xend=percHits, y=term, yend=term),linewidth=1) +
  scale_colour_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_fill_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_size(name="Hits (%)", range = c(1, 5))+
  theme_classic() +
  ylab("GO Term")+
  xlab("Hits (%)")+
  ggtitle("Mid-Development (65-163 hpf); 3,201 genes")+
  theme(legend.position = "right",
        axis.text.y = element_text(size=14),
        axis.text.x = element_text(size=14),
        axis.title.y = element_text(size=18, face="bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 18, face = "bold"), 
        strip.text.y.right=element_text(angle=0, size=14)); go_plot_mid

ggsave(filename="Mcap2020/Figures/TagSeq/DESeq/functional_enrichment/go_mid.pdf", go_plot_mid, width=12, height=15)

mid_data<-pattern_go_results %>% 
  filter(pattern=="Mid")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)

write_csv(mid_data, file="Mcap2020/Output/TagSeq/DESeq/functional_enrichment/enrichment_results_mid.csv")
```

Late development 
```{r}
late_data<-pattern_go_results %>% 
  filter(pattern=="Late")

# First, let's find the unique ParentTerm values
unique_late_terms <- unique(late_data$ParentTerm)

# Determine the number of unique ParentTerm values
num_unique <- length(unique_late_terms)

# Determine the midpoint for splitting into two batches
midpoint <- ceiling(num_unique / 2)
midpoint <- 6

late_data <- late_data %>%
  mutate(batch = ifelse(match(ParentTerm, unique_late_terms) <= midpoint, 1, 2))

late_data$ParentTerm <- factor(late_data$ParentTerm) 
#late_data$ParentTerm <- factor(late_data$ParentTerm, levels = rev(levels(late_data$ParentTerm)))

go_plot_lateA <- late_data %>% 
  filter(batch=="1")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)%>%
  
  ggplot(aes(x=percHits, y=term, fill=bh_adjust, colour=bh_adjust, size=percHits)) +
  facet_grid(ParentTerm ~ ., scales = "free", space='free', labeller = label_wrap_gen(width = 20, multi_line = TRUE))+
  geom_point(aes(group=term))+
  geom_segment(aes(x=0, xend=percHits, y=term, yend=term),linewidth=1) +
  scale_colour_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_fill_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_size(name="Hits (%)", range = c(1, 5))+
  theme_classic() +
  ylab("GO Term")+
  xlab("Hits (%)")+
  ggtitle("Late Development (183-231 hpf); 3,192 genes")+
  theme(legend.position = "none",
        axis.text.y = element_text(size=14),
        axis.text.x = element_text(size=16),
        legend.text = element_text(size=16),
        legend.title = element_text(size=18),
        axis.title.y = element_text(size=18, face="bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"), 
        strip.text.y.right=element_text(angle=0, size=14)); go_plot_lateA

go_plot_lateB <- late_data %>% 
  filter(batch=="2")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)%>%
  
  ggplot(aes(x=percHits, y=term, fill=bh_adjust, colour=bh_adjust, size=percHits)) +
  facet_grid(ParentTerm ~ ., scales = "free", space='free', labeller = label_wrap_gen(width = 20, multi_line = TRUE))+
  geom_point(aes(group=term))+
  geom_segment(aes(x=0, xend=percHits, y=term, yend=term),linewidth=1) +
  scale_colour_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_fill_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_size(name="Hits (%)", range = c(1, 5))+
  theme_classic() +
  ylab("")+
  xlab("Hits (%)")+
  ggtitle("")+
  theme(legend.position = "right",
        axis.text.y = element_text(size=14),
        axis.text.x = element_text(size=16),
        legend.text = element_text(size=16),
        legend.title = element_text(size=18),
        axis.title.y = element_text(size=18, face="bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"), 
        strip.text.y.right=element_text(angle=0, size=14)); go_plot_lateB

go_plot_late<-plot_grid(go_plot_lateA, go_plot_lateB, ncol=2, rel_widths=c(0.8,1))
 
ggsave(filename="Mcap2020/Figures/TagSeq/DESeq/functional_enrichment/go_late.pdf", go_plot_late, width=22, height=12)

late_data<-pattern_go_results %>% 
  filter(pattern=="Late")%>%
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)

write_csv(late_data, file="Mcap2020/Output/TagSeq/DESeq/functional_enrichment/enrichment_results_late.csv")
```