---
title: "KEGG Ortholog (KO) pathway analysis for gene expression clusters"
subtitle: "Characterizing the 3 gene expression patterns found in clustering analysis"
author: "Sarah Tanja"
date: 2025-11-17
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

KEGGREST package to access KEGG database: https://www.bioconductor.org/packages/release/bioc/vignettes/KEGGREST/inst/doc/KEGGREST-vignette.html

For each of GO and KEGG you want:

Background / universe

universe_genes <- rownames(dds) (or similar)

DEG + annotation table in long form

One row per (gene_id, GO_id) and (gene_id, KEGG_id).

TERM2GENE & TERM2NAME tables

GO: (GO_id, gene_id) and (GO_id, GO_term)

KEGG: (KEGG_id, gene_id) and (KEGG_id, KEGG_term)

Clustered gene lists

gene_clusters <- split(deg_annot$gene_id, deg_annot$cluster) (with dedup)

Run compareCluster(..., fun="enricher", TERM2GENE=..., TERM2NAME=...)

One call for GO, one for KEGG.

Visualise with dotplot(), emapplot(), cnetplot(), etc.

Breakdown of the gene ID format

Montipora_capitata_HIv3___TS.g9699.t2

`Montipora_capitata`	Species (your experimental species)
`HIv3`	Genome assembly version — Hawai‘i version 3 (the one commonly used in RNA-seq pipelines now)
`___TS`	Annotation source / transcript set (often "TS" = transcript set, sometimes “TS” refers to Trinity/STRINGTIE if that was used in comprehensive assemblies)
`g9699`	Gene ID (numbered gene)
`t2`	Transcript number 2 (transcript isoform of g9699)

# Setup
## Set paths
```{r}
output_path <- "../../output/07_enrichment/clusterProfiler/"
```


## Load libraries
```{r}
#BiocManager::install("clusterProfiler")
#BiocManager::install("enrichplot")
#BiocManager::install("KEGGREST")

library(clusterProfiler)
library(enrichplot)
library(tidyverse)
library(GO.db)
library(AnnotationDbi)
library(KEGGREST)
```

## Load inputs

Read in gene lists and annotations
```{r}
# Full gene set tested in DESeq2 analysis
all_genes <- read.csv("../../output/06_deg/spawnnight/all_genes.csv")

# significant genes (130 total) (only those that are significant for leachate or stage:leachate)
sig_genes <- read.csv("../../output/06_deg/spawnnight/sig_genes.csv")
sig_summary <- read.csv("../../output/06_deg/spawnnight/sig_summary.csv")

# Gene cluster assignments 
gene_clusters <- read.csv("../../output/06_deg/cluster/gene_clusters.csv")

# GO/KEGG Annotation to gene table
all_genes_anno <- read.csv("../../output/07_enrichment/goseq/all_genes_anno.csv")
gene2kegg <- read.csv("../../output/07_enrichment/goseq/gene2kegg.csv")
gene2go <- read.csv("../../output/07_enrichment/goseq/gene2go.csv")

# gene universe named vector
universe.genes <- all_genes$gene_id
names(universe.genes) <- universe.genes

# DEG named vector 
deg.vector <- sig_genes$gene_id
names(deg.vector) <- deg.vector
```


# Get KEGG Pathways for KO terms

KO = KEGG Orthology ID
Example: K01900 (Enolase)

Represents gene-level functional units.

One KO = one “orthologous gene family” with a well-defined function

KO is the atomic KEGG annotation

Pathways are large biological processes composed of many KOs.

High-level, very interpretable (e.g., apoptosis, TCA cycle, Wnt signaling)

Based on curated networks

Pathways can be metabolic or signaling
# Build kegg TERM2NAME table

KEGG has two montipora species in the organisms list, Montipora capricornis (leaf plate montipora) with organismal code [mcaz](https://www.kegg.jp/kegg-bin/show_organism?org=mcaz) and Montipora foliosa with organismal code [mfoo](https://www.kegg.jp/kegg-bin/show_organism?org=mfoo). 

Use KEGGREST to access mcaz and mfoo pathways
```{r}
kegg_mcaz <- keggList("pathway", "mcaz")
head(kegg_mcaz)
```

```{r}
kegg_mfoo <- keggList("pathway", "mfoo")
head(kegg_mfoo)
```

Use KEGGREST to access KO pathways via function `keggList("pathway", "ko")`! This queries the KO database and returns all KEGG pathways with their names. As a reminder, we got these ko pathways from eggNOG-mapper annotations.
```{r}
kegg_ko_paths <- keggList("pathway", "ko")

head(kegg_ko_paths)
```

keggList returns a named character vector, so now we have to turn it into a tibble dataframe with columns term and name.
```{r}
# kegg_ko_paths is a named character vector
kegg2name <- tibble(
  term = sub("^path:", "", names(kegg_ko_paths)),  # "ko04714"
  name = unname(kegg_ko_paths)                    # "Dopaminergic synapse"
)

# kegg2name
head(kegg2name)
# should look like:
#   term     name
#   ko04714  Dopaminergic synapse
```

We built a gene2kegg table earlier from eggNOG-mapper annotations for goseq. Now we need to turn that into TERM2GENE format for clusterProfiler.
```{r}
kegg2gene <- gene2kegg %>%
  dplyr::select(category, gene) %>%
  rename(term = category)
# gene2kegg 
head(kegg2gene)
# should look like:
#   term      gene
# 	ko04714	  Montipora_capitata_HIv3___RNAseq.g4581.t1
```

# enricher (for KO KEGG pathways)

```{r}
# How many of the 130 significant DE genes are mapped to kegg pathways?
sum(sig_genes$gene_id %in% kegg2gene$gene)
```
::: callout-note
only 44 out of the 130 DEGs have KEGG pathway annotation 
:::

We already know from our goseq results that we don't have any enriched go terms. Therefore, I don't expect to see any enrichment in KEGG Pathways either. But I still want to see what our DEGs do, so we will run the `clusterProfiler::enricher()` function but we won't filter for pvalueCutoff. This means we'll just show everything, and explain that what is shown is **not enriched**, but is descriptive of what the DEGs do.

```{r}
# sig_genes$gene_id = your DEGs (all patterns combined)
# gene2kegg = gene, category (ko####) as you built
# kegg2name = data.frame(term = "ko04714", name = "Dopaminergic synapse")

kegg_all <- enricher(
  gene         = deg.vector,
  TERM2GENE    = kegg2gene,
  TERM2NAME    = kegg2name,
  pvalueCutoff = 1,   # keep everything
  qvalueCutoff = 1
)

# Look at the table
kegg_df <- as.data.frame(kegg_all)
nrow(kegg_df)   # pathways that have at least 1 DEG
```

Make a named numeric “cluster code” for your DEGs
You already have clusters (1/2/3) for DEGs. You can feed that to cnetplot() via foldChange= to color gene nodes by pattern.
```{r}
# sig_summary: gene_id, cluster (1/2/3)
cluster_code <- gene_clusters %>%
  distinct(gene_id, cluster) %>%
  # encode clusters as numeric just for the color scale
  mutate(code = case_when(
    cluster == 1 ~ -1,   # e.g. cleavage-high
    cluster == 2 ~  0,   # prawnchip-high
    cluster == 3 ~  1    # early-gastrula-high
  )) %>%
  dplyr::select(gene_id, code) %>%
  tibble::deframe()   # named numeric: names = gene_id, values = code
```
Now cluster_code["Montipora_capitata_HIv3___TS.g1234.t1"] gives -1/0/1 depending on pattern.

Cnetplot: KEGG pathways with gene nodes colored by pattern
```{r, fig.height=10, fig.width=10, warning=FALSE}
p_cnet <- cnetplot(
  kegg_all,
  foldChange  = cluster_code,  # this colors gene nodes
  showCategory = 200,           # number of pathways to show
  node_label   = "category"    # label pathways, not genes
)

p_cnet+
  scale_colour_gradient2(
    name = "Expression pattern",
    low  = "#8E6B01",  # cluster 1
    mid  = "#CECDC3",  # cluster 2
    high = "#5B8E8A",  # cluster 3
    midpoint = 0
  )
```
> “KEGG pathways annotated to DEGs. Nodes represent pathways (circles) and genes (smaller nodes), with gene nodes colored by expression pattern cluster (cleavage-, prawnchip-, or early-gastrula–biased). Pathway nodes are sized by the number of genes that map to that pathway. No pathways were significantly enriched after multiple-testing correction; visualization is descriptive of 44 DEGs that had KEGG Pathway annotation.”

```{r, fig.width=10, fig.height=10}
kegg_sim <- pairwise_termsim(kegg_all)  # adds term-term similarity
p_emap <- emapplot(
  kegg_sim,
  showCategory = 200
)

p_emap 
```
# enrich GO

# Build GO term to name dataframe

Reformat for clusterProfiler
```{r}
go2gene <- gene2go %>% 
  dplyr::select(category, gene) %>% 
  rename(term = category)

head(go2gene)
# term gene
```

```{r}
go_terms <- AnnotationDbi::select(
  GO.db,
  keys    = go2gene$term,
  keytype = "GOID",
  columns = c("TERM", "ONTOLOGY")
)

head(go_terms)
#      GOID        TERM                               ONTOLOGY
```
```{r}
goterms <- go_terms %>% 
  dplyr::select(GOID, TERM) %>% 
  rename(name = TERM, 
         term = GOID)

head(goterms)
```

```{r}
go2name <- go2gene %>%
  left_join(go_terms, by = c("term" = "GOID")) %>%
  rename(name = TERM) %>%
  dplyr::select(term, name)

head(go2name)
```

```{r}
go_all <- enricher(
  gene         = sig_genes$gene_id,   # your DEGs
  TERM2GENE    = go2gene,
  TERM2NAME    = go2name,
  universe     = universe.genes,
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  minGSSize    = 1
)

go_all
as.data.frame(go_all) %>% head()
```


# ORA
https://chatgpt.com/share/691d3dec-fff0-8008-8a0b-fe8f672fa716


For ORA in clusterProfiler, whether GO, KEGG, or custom gene sets, the core inputs are:

#1. A gene list (or lists)
  - e.g. genes in pattern 1, pattern 2, pattern 3
  - This is a character vector of gene IDs.
  
For comparing expression patterns 1/2/3 in one go, we can pass a named list of gene vectors into compareCluster(geneClusters = ...)  


```{r}
identical(unique(sig_summary$gene_id), sig_genes$gene_id)

sig_genes_vector <- sig_genes$gene_id
```


```{r}
sig_genes_cluster1 <- sig_summary %>% 
  filter(cluster == 1) %>% 
  pull(unique(gene_id)) # does this collapse any duplicate gene_id?

sig_genes_cluster2 <- sig_summary %>%
  filter(cluster == 2) %>% 
  pull(unique(gene_id))

sig_genes_cluster3 <- sig_summary %>%
  filter(cluster == 3) %>% 
  pull(unique(gene_id))

gene_clusters <- list(
  Pattern_1 = sig_genes_cluster1,
  Pattern_2 = sig_genes_cluster2,
  Pattern_3 = sig_genes_cluster3
)
```
  
  

# 2. A background / universe 
  - All genes that could possibly have been called DE in that analysis (e.g. all genes in your filtered DESeq2 object, ~10k).
  - Passed as universe = all_genes in enricher() / enrichGO() etc.
```{r}
universe_genes <- all_genes$gene_id
```

# 3. Annotation mapping: TERM2GENE (and optionally TERM2NAME)
  - For GO ORA: a 2-column data frame of term = GO ID (e.g. "GO:0006355"), gene = gene_id
  - For KEGG ORA: a 2-column data frame of term = KEGG pathway ID or KO ID (whichever you want to treat as “gene set”), gene = gene_id
  - Built from your annotation table (gff, eggNOG), then restricted to the genes in `universe_genes`

Xu et al. explicitly demonstrate using enricher()/compareCluster() with custom TERM2GENE/TERM2NAME for non-model GO and TF annotations, which is exactly our situation.
```{r}
go2gene <- all_genes_GO %>% 
  dplyr::select(GOs, gene_id, Preferred_name) %>% 
  separate_rows(GOs, sep = ",") %>%        # separate the GOs column into rows by comma
  mutate(term = trimws(GOs)) %>%                  # trim whitespace 
  rename(gene = gene_id) %>%               
  dplyr::select(term, gene)
```

## Get GO terms 
```{r}
# vector of GO IDs you care about
go_ids <- unique(go2gene$term)   # or whatever your df is called

go_terms <- AnnotationDbi::select(
  GO.db,
  keys    = go_ids,
  keytype = "GOID",
  columns = c("TERM", "ONTOLOGY")    # TERM = name, ONTOLOGY = BP/CC/MF
)

go2name <- go2gene %>%
  left_join(go_terms, by = c("term" = "GOID")) %>% 
  rename(name = TERM) %>% 
  dplyr::select(term, name)
```

In the KEGG_Pathway columns we basically have two representations of the same KEGG pathway:
- ko04714 = “global KEGG Orthology” pathway ID
- map04714 = “reference pathway map”

For enrichment, if you keep both for the same gene, you’lld ouble-count that gene-pathway association (once as ko04714, once as map04714).

In KEGG, map refers to the pathway map, a graphical representation of a metabolic or functional pathway network, while KO is the KEGG Orthology identifier, a specific gene ortholog that acts as a building block (node) for these maps. A map provides a systems-level view of how various biological processes work together, whereas a KO is the unique identifier for a specific gene or protein that performs a function, such as catalyzing a reaction. 

```{r}
kegg_pairs <- all_genes_KEGG %>%
  rename(gene = Preferred_name) %>%
  # 1) split comma-separated IDs into separate rows
  separate_rows(KEGG_Pathway, sep = ",") %>%
  mutate(kegg_pathway = trimws(KEGG_Pathway)) %>%
  filter(kegg_pathway != "") %>%
  
  # 2) keep only ko/map-style KEGG IDs
  filter(str_detect(kegg_pathway, "^(ko|map)\\d+")) %>%
  
  # 3) tag each as ko or map and pull out the numeric part
  mutate(
    type   = if_else(str_starts(kegg_pathway, "ko"), "ko", "map"),
    id_num = str_extract(kegg_pathway, "\\d+")   # 04714, 05230, etc.
  ) %>%
  select(kegg_pathway, gene, gene_id, type, id_num) %>%
  distinct()

kegg2gene <- kegg_pairs %>%
  pivot_wider(
    id_cols    = c(gene, gene_id, id_num),   # one row per gene + numeric ID
    names_from = type,                 # columns: ko, map
    values_from = kegg_pathway
  ) %>%
  # optional: drop rows where only ko or only map exists
  filter(!is.na(ko) & !is.na(map)) %>%
  select(ko, gene, gene_id, map)
```


```{r}
ego <- enricher(
  gene         = sig_genes_vector,
  TERM2GENE    = go2gene,
  TERM2NAME    = go2name,
  universe     = universe_genes,
  pvalueCutoff = 1,
  qvalueCutoff = 1
)

summary(ego)
```

```{r}
dotplot(ego)
```



```{r}
ego1 <- enricher(
  gene        = sig_genes_cluster1,
  TERM2GENE   = go2gene,
  universe    = universe_genes,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.2
)

ego2 <- enricher(
  gene        = cluster2_genes,
  TERM2GENE   = go_term2gene,
  TERM2NAME   = go_term2name,
  universe    = universe_genes
)

ego3 <- enricher(
  gene        = cluster3_genes,
  TERM2GENE   = go_term2gene,
  TERM2NAME   = go_term2name,
  universe    = universe_genes
)
```




```{r}
ego_clusters <- compareCluster(
  geneClusters = sig_genes,
  fun          = "enricher",
  TERM2GENE    = go2gene,
  universe     = universe_genes,  # optional but good practice
  pAdjustMethod = "BH",
  qvalueCutoff  = 1
)

# quick visualization
library(enrichplot)
dotplot(ego_clusters, showCategory = 10)
```


# GSEA
```{r}
xx <- compareCluster(
  geneCluster   = gene_lists,
  fun           = "enrichKEGG",        # or enrichKEGG, etc.
  OrgDb         = orgdb_object,
  keyType       = "ENSEMBL",         # or whatever your IDs are
  ont           = "BP",
  pvalueCutoff  = 0.05
)
```


