---
title: "Gene ontology (GO) term analysis for gene expression clusters"
subtitle: "Characterizing the 3 gene expression patterns found in clustering analysis"
author: "Sarah Tanja"
date: 2025-10-21
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

In the clustering analysis (07_cluster.qmd), we identified 3 distinct gene expression patterns among the 130 significant genes affected by leachate. 
These genes cluster based on their expression patterns across leachate concentrations (control, low, mid, high).

Gene cluster breakdown:
- **Cluster 1**: 72 genes, peak in early gastrula
- **Cluster 2**: 25 genes, peak in prawnchip 
- **Cluster 3**: 31 genes, peak in cleavage

We now perform functional enrichment analysis on each cluster separately to understand the biological processes associated with each expression pattern.

We use the R package `goseq` for GO enrichment. 
`goseq` corrects for gene length bias - longer genes are more likely to be detected as DEGs in RNA-Seq simply because they accumulate more reads. 
`goseq` accounts for this by using a probability weighting function.

## Inputs

- `sig_genes.csv` - A dataframe (list) of the 130 significant genes
- `gene_clusters.csv` - A dataframe mapping 128 genes to an expression cluster (1, 2, or 3)
- `all_genes.csv` - a dataframe (list) of all genes used in the DESeq2 analysis
- functional annotation: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz (EggNog)
- functional annotation: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz (KEGG)

## Outputs

- GO enrichment results for each of the 3 clusters
- Reduced GO terms (with redundancy removed) for each cluster
- Summary tables and visualizations

```{r}
output_path <- "../../output/07_go/"
```


# Setup

## Install packages

```{r, eval=FALSE}
# CRAN packages
#if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')

# Bioconductor packages
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("GenomicRanges")
#BiocManager::install("Biostrings")
#BiocManager::install("goseq")
#BiocManager::install("rrvgo")
#BiocManager::install("org.Ce.eg.db") # c.elegans genome-wide annotation
```

## Load packages

```{r}
library(GenomicRanges)
library(Biostrings)
library(goseq)
library(rrvgo)
library(org.Ce.eg.db)
library(tidyverse)
```

## Load inputs

Read in gene lists

```{r}
# significant genes (130 total) (only those that are significant for leachate or stage:leachate)
sig_genes <- read.csv("../../output/06_deg/spawnnight/sig_all.csv")

# Gene cluster assignments 
gene_clusters <- read.csv("../../output/06_deg/cluster/gene_clusters.csv")

# Full gene set tested in DESeq2 analysis
all_genes <- read.csv("../../output/06_deg/spawnnight/all_genes.csv")
```

Import fasta 

```{r, eval=FALSE}
# Load transcriptome
all_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")

# Filter for the 130 significant genes
sig_seqs <- all_seqs[names(all_seqs) %in% sig_genes$gene_id]
```

Import EggNog annotation

```{r}
# Read in the file
EggNog_anno <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")

# Rename "#query" to gene_id
EggNog_anno <- EggNog_anno %>%
  rename(gene_id="#query")

paste(nrow(EggNog_anno), "genes have functional annotation")
```

# Annotate significant genes

Match genes in sig_genes to annotation dataframe

```{r}
# Annotate significant genes
sig_genes_anno <- inner_join(sig_genes, EggNog_anno, by = "gene_id")

# Filter out any that lack GO term annotation
sig_genes_annofilt <- sig_genes_anno %>% 
  mutate(GOs = na_if(str_trim(GOs), "-")) %>%
  filter(!is.na(GOs), GOs != "")

paste("Annotated genes:", 
      nrow(sig_genes_anno), "/", 
      nrow(sig_genes), "=",
      round(100*nrow(sig_genes_anno)/nrow(sig_genes),1),"%")

paste("GO annotated genes:", 
      nrow(sig_genes_annofilt), "/", 
      nrow(sig_genes), "=",
      round(100*nrow(sig_genes_annofilt)/nrow(sig_genes),1),"%")
```

## Add gene lengths

::: callout-note
Gene lengths are required for goseq bias correction.
:::

```{r}
# Import GFF file to get gene lengths
# Check if file exists before importing
gff_file <- "../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3"

if (file.exists(gff_file)) {
  gff <- rtracklayer::import(gff_file)
  
  transcripts <- subset(gff, type == "transcript")
  
  transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE)
  
  transcript_lengths <- width(transcripts_gr)
  
  gene_id <- transcripts_gr$ID
  
  lengths <- cbind(gene_id, transcript_lengths)
  
  lengths <- as.data.frame(lengths)
  
  paste("Gene lengths extracted for", nrow(lengths), "transcripts")
} else {
  warning("GFF file not found. Gene length bias correction will not be available.")
  lengths <- data.frame(gene_id = character(), transcript_lengths = integer())
}
```

Add length to the annotation dataframe

```{r}
sig_genes_af <- sig_genes_annofilt %>% left_join(lengths, by = "gene_id")
```

## Add cluster assignments

Join cluster information with annotated genes

```{r}
sig_genes_afc <- inner_join(sig_genes_af, gene_clusters, by = "gene_id")
str(sig_genes_afc)
```
::: callout-note
`sig_genes_afc` is now the dataframe with GO annotated, filtered, and matched to gene pattern clusters
:::

Save GO annotated dataframe

```{r}
write_csv(sig_genes_afc, file.path(output_path, "sig_genes_GOannotated.csv")
```

## Subset by cluster

Create separate dataframes for each cluster

```{r}
sig_cluster1 <- sig_genes_afc %>% filter(cluster == 1)
sig_cluster2 <- sig_genes_afc %>% filter(cluster == 2)
sig_cluster3 <- sig_genes_afc %>% filter(cluster == 3)

paste("Cluster 1:", nrow(sig_cluster1), "genes with GO annotation")
paste("Cluster 2:", nrow(sig_cluster2), "genes with GO annotation")
paste("Cluster 3:", nrow(sig_cluster3), "genes with GO annotation")
paste("All together:", nrow(sig_genes_afc), "genes with GO annotation")
```

# Functional enrichment

Conduct functional enrichment using goseq and rrvgo. 
Compare against all genes in the dataset that have GO annotation information.

Conduct at biological process level. Filter for adjusted p-value < 0.05.

## Turn gene lists into vectors

```{r}
# Define gene ID vectors for each cluster & for the annotated significant genes 
ID.vector.c1 <- sig_cluster1$gene_id
ID.vector.c2 <- sig_cluster2$gene_id
ID.vector.c3 <- sig_cluster3$gene_id
ID.vector.sig <- sig_genes_afc$gene_id
```

## Create gene universe for bias correction

```{r}
# Universe for bias correction
# All genes detected in our filtered DEG tested genes with GO annotation info
ALL.vector <- EggNog_anno %>%
  mutate(GOs = na_if(str_trim(GOs), "-")) %>%
  filter(!is.na(GOs), GOs != "") %>%
  pull(gene_id) %>%
  intersect(all_genes$gene_id)

paste(length(ALL.vector), "genes out of", nrow(all_genes), "genes in our DESeq2 dataset have GO annotations")
paste("Cluster 1:", length(ID.vector.c1), "GO annotated genes")
paste("Cluster 2:", length(ID.vector.c2), "GO annotated genes")
paste("Cluster 3:", length(ID.vector.c3), "GO annotated genes")

paste("All together:", length(ID.vector.sig), "GO annotated genes")

```

## GO term mapping

```{r}
# Gene to GO term mapping for each cluster
GO.terms.c1 <- sig_cluster1 %>%
  dplyr::select(gene_id, GOs)

GO.terms.c2 <- sig_cluster2 %>%
  dplyr::select(gene_id, GOs)

GO.terms.c3 <- sig_cluster3 %>%
  dplyr::select(gene_id, GOs)

GO.terms.sig <- sig_genes_afc %>%
  dplyr::select(gene_id, GOs)
```

Function to split GO terms to one term per row

```{r}
split_go <- function(df, gene_col = "gene_id", go_col = "GOs") {
  df %>%
    mutate(across(all_of(c(gene_col, go_col)), as.character)) %>%
    separate_rows(!!rlang::sym(go_col), sep = ",") %>%
    mutate(!!rlang::sym(go_col) := str_trim(!!rlang::sym(go_col))) %>%
    filter(!is.na(!!rlang::sym(go_col)), !!rlang::sym(go_col) != "") %>%
    dplyr::select(all_of(c(gene_col, go_col))) %>%
    as.data.frame(stringsAsFactors = FALSE)
}
```

Split GO terms

```{r}
# Format GO terms to one term per row
GO.terms.c1 <- split_go(GO.terms.c1)
GO.terms.c2 <- split_go(GO.terms.c2)
GO.terms.c3 <- split_go(GO.terms.c3)
GO.terms.sig <- split_go(GO.terms.sig)
```

Save GO term mappings

```{r}
write_csv(GO.terms.c1, file.path(output_path, "GO.terms.cluster1.csv"))
write_csv(GO.terms.c2, file.path(output_path, "GO.terms.cluster2.csv"))
write_csv(GO.terms.c3, file.path(output_path, "GO.terms.cluster3.csv"))
write_csv(GO.terms.sig, file.path(output_path, "GO.terms.sig.csv"))
```

Get vector of gene lengths

```{r}
# Only proceed if we have length information
if (nrow(lengths) > 0) {
  lengths_anno <- lengths %>% 
    dplyr::filter(gene_id %in% all_genes$gene_id) %>% 
    dplyr::filter(gene_id %in% EggNog_anno$gene_id)
  
  LENGTH.vector <- as.integer(lengths_anno$transcript_length)
  names(LENGTH.vector) <- ALL.vector
  
  use_length_bias <- TRUE
  paste("Length bias correction enabled with", length(LENGTH.vector), "genes")
} else {
  use_length_bias <- FALSE
  warning("Gene lengths not available. Proceeding without length bias correction.")
}
```

Make gene vector function

```{r}
make_gene_vector <- function(all_ids, de_ids, label = "DE set") {
  # Normalize to clean character vectors
  all_ids <- as.character(all_ids)
  de_ids  <- unique(trimws(as.character(de_ids)))

  # Hard checks on ALL.vector
  if (anyNA(all_ids)) stop("ALL.vector contains NA gene IDs.")
  if (anyDuplicated(all_ids)) {
    dups <- unique(all_ids[duplicated(all_ids)])
    stop("ALL.vector has duplicated gene IDs (e.g., ", paste(head(dups, 5), collapse = ", "), ").")
  }

  # Soft check: DE IDs not in ALL
  extras <- setdiff(de_ids, all_ids)
  if (length(extras) > 0) {
    warning(length(extras), " ", label, " IDs not found in ALL.vector (showing up to 5): ",
            paste(head(extras, 5), collapse = ", "))
    de_ids <- intersect(de_ids, all_ids)
  }

  # Construct 0/1 vector aligned to ALL.vector
  vec <- as.integer(all_ids %in% de_ids)
  names(vec) <- all_ids

  # Final sanity
  stopifnot(length(vec) == length(all_ids), all(vec %in% c(0L, 1L)))
  vec
}
```

Create gene vectors for each cluster

```{r}
gene.vector.c1 <- make_gene_vector(ALL.vector, ID.vector.c1, "cluster 1")
gene.vector.c2 <- make_gene_vector(ALL.vector, ID.vector.c2, "cluster 2")
gene.vector.c3 <- make_gene_vector(ALL.vector, ID.vector.c3, "cluster 3")
gene.vector.sig <- make_gene_vector(ALL.vector, ID.vector.sig, "sig genes")
```

## Null probability weighting function (bias for gene length)

::: callout-note
Gene length bias correction accounts for the fact that longer genes are more likely to be detected as differentially expressed in RNA-seq data simply because they accumulate more reads. If length data is not available, goseq can still run but without this correction.
:::

```{r}
# Keep only genes present in ALL.vector and align the order
LENGTH.vector <- LENGTH.vector[ALL.vector]
stopifnot(identical(names(LENGTH.vector), ALL.vector))

# Calculate probability weighting function for each cluster
pwf.c1 <- nullp(gene.vector.c1, bias.data = LENGTH.vector)
pwf.c2 <- nullp(gene.vector.c2, bias.data = LENGTH.vector)
pwf.c3 <- nullp(gene.vector.c3, bias.data = LENGTH.vector)
pwf.sig <- nullp(gene.vector.sig, bias.data = LENGTH.vector)
```

```{r}
identical(pwf.sig, pwf.c3)
```


## GO BP enrichment

Run GO enrichment for biological process terms

```{r}
GO.bp.c1 <- goseq(
  pwf.c1,
  gene2cat = GO.terms.c1,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO.bp.c2 <- goseq(
  pwf.c2,
  gene2cat = GO.terms.c2,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO.bp.c3 <- goseq(
  pwf.c3,
  gene2cat = GO.terms.c3,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO.bp.sig <- goseq(
  pwf.sig,
  gene2cat = GO.terms.sig,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)
```

Use the Benjamini-Hochberg method to control false discovery rate

```{r}
GO.bp.c1$bh_adjust <- p.adjust(GO.bp.c1$over_represented_pvalue, method = "BH")
GO.bp.c2$bh_adjust <- p.adjust(GO.bp.c2$over_represented_pvalue, method = "BH")
GO.bp.c3$bh_adjust <- p.adjust(GO.bp.c3$over_represented_pvalue, method = "BH")
GO.bp.sig$bh_adjust <- p.adjust(GO.bp.sig$over_represented_pvalue, method = "BH")
```

Filter for biological process terms with valid p-values

```{r}
GO.BP.c1 <- GO.bp.c1 %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

GO.BP.c2 <- GO.bp.c2 %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

GO.BP.c3 <- GO.bp.c3 %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

GO.BP.sig <- GO.bp.sig %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)
```

Write to csv

```{r}
write_csv(GO.BP.c1, file = file.path(output_path, "GO.BP.cluster1.csv"))
write_csv(GO.BP.c2, file = file.path(output_path, "GO.BP.cluster2.csv"))
write_csv(GO.BP.c3, file = file.path(output_path, "GO.BP.cluster3.csv"))
write_csv(GO.BP.sig, file = file.path(output_path, "GO.BP.sig.csv"))
```

Display results

```{r}
paste("Cluster 1:", nrow(GO.BP.c1), "enriched BP terms (BH-adjusted p < 0.05)")
paste("Cluster 2:", nrow(GO.BP.c2), "enriched BP terms (BH-adjusted p < 0.05)")
paste("Cluster 3:", nrow(GO.BP.c3), "enriched BP terms (BH-adjusted p < 0.05)")
paste("All together:", nrow(GO.BP.sig), "enriched BP terms (BH-adjusted p < 0.05)")
```

## Reduce redundancy

Calculate semantic similarity matrix for each cluster

```{r}
simMatrix.c1 <- calculateSimMatrix(
    GO.BP.c1$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

simMatrix.c2 <- calculateSimMatrix(
    GO.BP.c2$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

simMatrix.c3 <- calculateSimMatrix(
    GO.BP.c3$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

simMatrix.sig <- calculateSimMatrix(
    GO.BP.sig$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

```

Create named scores vectors

```{r}
# Create scores for each cluster (handle potential zeros in p-values)
scores.c1 <- setNames(
    -log10(pmax(GO.BP.c1$bh_adjust, .Machine$double.xmin)),
    GO.BP.c1$category
  )

scores.c2 <- setNames(
    -log10(pmax(GO.BP.c2$bh_adjust, .Machine$double.xmin)),
    GO.BP.c2$category
  )

scores.c3 <- setNames(
    -log10(pmax(GO.BP.c3$bh_adjust, .Machine$double.xmin)),
    GO.BP.c3$category
  )

scores.sig <- setNames(
    -log10(pmax(GO.BP.sig$bh_adjust, .Machine$double.xmin)),
    GO.BP.sig$category
  )

```

Reduce terms with similarity matrix

```{r}
# Reduce terms for each cluster
reducedTerms.c1 <- reduceSimMatrix(
    simMatrix.c1,
    scores = scores.c1,
    threshold = 0.8,
    orgdb = "org.Ce.eg.db"
  )

reducedTerms.c2 <- reduceSimMatrix(
    simMatrix.c2,
    scores = scores.c2,
    threshold = 0.8,
    orgdb = "org.Ce.eg.db"
  )

reducedTerms.c3 <- reduceSimMatrix(
    simMatrix.c3,
    scores = scores.c3,
    threshold = 0.8,
    orgdb = "org.Ce.eg.db"
  )

reducedTerms.sig <- reduceSimMatrix(
    simMatrix.sig,
    scores = scores.sig,
    threshold = 0.8,
    orgdb = "org.Ce.eg.db"
  )
```

# Save results

Save reduced terms and similarity matrices

```{r}
# Save reduced terms
write_csv(reducedTerms.c1, file.path(output_path, "reducedTerms_cluster1.csv"))

write_csv(reducedTerms.c2, file.path(output_path, "reducedTerms_cluster2.csv"))

write_csv(reducedTerms.c3, file.path(output_path, "reducedTerms_cluster3.csv"))

write_csv(reducedTerms.sig, file.path(output_path, "reducedTerms_sig.csv"))


# Save similarity matrices
simMatrix_c1 <- simMatrix.c1 %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GO")
  write_csv(simMatrix_c1, file.path(output_path, "simMatrix_cluster1.csv"))

simMatrix_c2 <- simMatrix.c2 %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GO")
  write_csv(simMatrix_c2, file.path(output_path, "simMatrix_cluster2.csv"))

simMatrix_c3 <- simMatrix.c3 %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GO")
  write_csv(simMatrix_c3, file.path(output_path, "simMatrix_cluster3.csv"))

simMatrix_sig <- simMatrix.sig %>%
    as.data.frame() %>% 
    rownames_to_column(var = "GO")
  write_csv(simMatrix_sig, file.path(output_path, "simMatrix_sig.csv"))
```

# Summary

This analysis performed GO enrichment on the 3 gene expression clusters identified in the clustering analysis.

## Cluster characterization

**Cluster 1** (97 genes): The largest cluster, this group of genes peak at early gastrula.

**Cluster 2** (28 genes): This group of genes peak at prawnchip.

**Cluster 3** (24 genes): The smallest cluster, this group of genes peak at cleavage.


A higher score means the term is more significantly enriched (because smaller p-values become larger –log10(p) values). If your score = –log10(p), then the parent is the most significantly enriched term in that cluster.

## Next steps

- Visualize enriched GO terms for each cluster
- Compare biological processes across clusters
- Integrate with other analyses to understand biological significance

```{r}
sessionInfo()
```
