---
title: "Gene ontology (GO) term analysis"
subtitle: "Now that we've identified the DEGs, what do they do?"
author: "Sarah Tanja"
date: 2025-06-17
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

# Background

We identified 44 genes that were broadly impacted by leachate. Here we want to know what those genes do! Those 44 genes grouped in

Why is goseq different from other enrichment tools? Typical enrichment methods (like Fisher's exact test or hypergeometric tests) assume that every gene has an equal probability of being selected as a DEG. But in RNA-Seq longer genes are more likely to be detected as DEGs just because they accumulate more reads. `goseq` corrects for this bias.

## Inputs

-   `LRT_significant_genes_combined_wald.csv` A dataframe of the 12 genes and their LRT and Wald test statistics
-   functional annotation (version 3) from: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz (EggNog)
-   functional annotation http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz (KEGG)

## Outputs

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 12,
  fig.height = 8,
  fig.align = "center",
  dpi = 600,
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

## Install packages

```{r, eval=FALSE}
if ("Biostrings" %in% rownames(installed.packages()) == 'FALSE') install.packages('Biostrings')
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')
if ("goseq" %in% rownames(installed.packages()) == 'FALSE') install.packages('goseq')
install.packages("GOplot")
```

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GenomicRanges")
BiocManager::install("goseq")
BiocManager::install("GO.db")
BiocManager::install("rrvgo")
BiocManager::install("org.Ce.eg.db") # c.elegens genome-wide annotation
```

## Load packages

```{r}
library(GenomicRanges)
library(Biostrings)
library(goseq)
library(rrvgo)
library(org.Ce.eg.db)
library(GO.db)
library(GOplot)
library(tidyverse)
```

## Read in gene lists and counts

```{r}
all_genes <- read.csv("../../output/06_deg/interactions/all_genes.csv")
sig_genes <- read.csv("../../output/06_deg/interactions/sig_genes.csv")
sig_counts <- read.csv("../../output/06_deg/interactions/sig_counts.csv", row.names="gene_id", check.names = FALSE)
```

## Import fasta

```{r}
# Load your transcriptome
all_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")  

# Filter for the 44 significant interaction genes
sig_seqs <- all_seqs[names(all_seqs) %in% sig_genes]

# Save to new FASTA file
writeXStringSet(sig_seqs, "../../output/07_go/significant_genes.fasta")
```

## Download annotation & enrichment files

```{r, engine='bash', eval=FALSE}
cd ../../input/annotations

wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz
wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz
```

Unzip

```{r, engine='bash', eval=FALSE}
cd ../../input/annotations
gunzip *.gz
```

## Import to R environment

```{r}
# Read in the file (it will include "#query" as a column name)
EggNog_anno <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")
```

rename "#query" to `gene_id`

```{r}

EggNog_anno <- EggNog_anno %>%
  rename(gene_id="#query")

nrow(EggNog_anno)
names(EggNog_anno)
```

> This functional annotation has 24,072 genes. This is 44% of total protein predicted genes described in the publication associated with this annotation. https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755

# Annotate significant genes

Match up genes in sig_genes dataframe to annotation dataframe

```{r}
sig_genes_anno <- inner_join(sig_genes, EggNog_anno, by = "gene_id")

nrow(sig_genes_anno)
```

::: callout-note
37 of the 44 significant genes were annotated! The annotation file `sig_genes_anno` dataframe now only contains genes that were significant and that have annotation information.
:::

Get gene length information.

```{r}
#import file
gff <- rtracklayer::import("../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3") #if this doesn't work, restart R and try again 

transcripts <- subset(gff, type == "transcript") #keep only transcripts 

transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE) #extract length information

transcript_lengths <- width(transcripts_gr) #isolate length of each gene

gene_id <- transcripts_gr$ID #extract list of gene id 

lengths <- cbind(gene_id, transcript_lengths)

lengths <- as.data.frame(lengths) #convert to data frame
```

Add in length to the annotation file (`sig_genes_anno$transcript_lengths`).

```{r}
sig_genes_anno <- sig_genes_anno %>% left_join(lengths, by = "gene_id")
```

Format GO terms to remove dashes and quotes.

```{r}
#remove dashes, remove quotes from columns and separate by semicolons (replace , with ;) in KEGG_ko, KEGG_pathway, and Annotation.GO.ID columns
head(sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub(",", ";", sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub('"', "", sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub("-", NA, sig_genes_anno$GOs)
```

# Conduct functional enrichment

Conduct functional enrichment using goseq and [rrvgo](https://www.bioconductor.org/packages/release/bioc/vignettes/rrvgo/inst/doc/rrvgo.html). Compare against all genes detected in the dataset that we have annotation information for.

Conduct at biological process level. Filter for adjusted p-value \<0.05.

```{r}
# Define your DEG list — the genes you're testing
ID.vector <- sig_genes_anno$gene_id  # Replace with your actual gene vector


# All genes detected in the dataset with annotation info
ALL.vector <- all_genes$gene_id # gene_ids for all 12565 genes that we used to make our DESeq2 object
ALL.vector <- ALL.vector[ALL.vector %in% EggNog_anno$gene_id] #filtered down to only genes that are annotated

length(ALL.vector)
```

::: callout-note
9563 out of the 12564 genes in our dataset are annotated
:::

```{r}
lengths_anno <- lengths %>% 
  filter(gene_id %in% all_genes$gene_id) %>% 
  filter(gene_id %in% EggNog_anno$gene_id)

LENGTH.vector <- as.integer(lengths_anno$transcript_length)
names(LENGTH.vector) <- ALL.vector  # Needed for bias correction

# Gene to GO term mapping
GO.terms <- sig_genes_anno %>%
  dplyr::select(gene_id, GOs)

# Format GO terms to one term per row
split <- strsplit(as.character(GO.terms$GOs), ";")
split2 <- data.frame(
  gene = rep.int(GO.terms$gene, sapply(split, length)),
  GOs = unlist(split)
)

GO.terms <- split2

# Construct named gene vector: 1 = DEG, 0 = not DEG
gene.vector <- as.integer(ALL.vector %in% ID.vector)
names(gene.vector) <- ALL.vector

# Null probability weighting function (bias for gene length)
# nullp is a function from nadiadavidson/goseq
pwf <- nullp(gene.vector, bias.data = LENGTH.vector)
```

About goseq() command:

| Argument                       | Meaning                                                                                                                                                      |
|-------------------|-----------------------------------------------------|
| `pwf`                          | A **probability weighting function** object from `nullp()`. It contains the DEG status of all genes (1 = DEG, 0 = not DEG), plus gene length bias info.      |
| `gene2cat`                     | A data frame or list mapping **genes to functional categories** (like GO terms). In your case, it's a long-format table of gene-GO pairs.                    |
| `test.cats = "GO:BP"`          | This limits the enrichment to only **biological process** GO terms (you could also use `"GO:MF"` or `"GO:CC"` for molecular function or cellular component). |
| `method = "Wallenius"`         | This uses the **Wallenius noncentral hypergeometric distribution**, which accounts for sampling bias (like gene length).                                     |
| `use_genes_without_cat = TRUE` | Allows the method to consider genes that **do not have a GO annotation** (prevents them from being excluded by default).                                     |

`goseq()` returns a data frame where each row is a GO term, with:

-   The GO ID

-   Category (BP, MF, or CC)

-   Over-representation p-value (for DEGs)

-   Under-representation p-value (less commonly used)

-   Number of DEGs in that GO term

-   Total number of genes in that GO term

-   Other metadata (like GO term name)

## GO enrichment

```{r}
# Run GO enrichment for biological process terms
GO.bp <- goseq(
  pwf,
  gene2cat = GO.terms,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

# Format and filter results
GO.bp <- GO.bp %>% 
  arrange(over_represented_pvalue)

# Use the Benjamini-Hochberg method to control false discovery rate 
GO.bp$bh_adjust <- p.adjust(GO.bp$over_represented_pvalue, method = "BH")

GO.bp <- GO.bp %>%
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

# Optional: write to file
write_csv(GO.bp, file = "../../output/07_go/functional_enrichment_biological_process_goseq.csv")
```

> The Benjamini-Hochberg procedure controls the proportion of false positives among all rejected hypotheses, which is especially important in high-throughput data like gene ontology enrichment. 

> Why Controlling False Discovery Rate Matters in Enrichment Analysis
Controlling the false discovery rate (FDR) is critical in enrichment analysis, especially when thousands of statistical tests are performed simultaneously (e.g., testing many gene sets, pathways, or terms for enrichment):
  - Multiple Comparisons Increase False Positives: In enrichment analyses, each hypothesis test (such as whether a specific gene set is enriched) carries a risk of yielding a false positive—detecting significance where there is none. When testing hundreds or thousands of hypotheses, the probability of encountering several false positives increases dramatically.
  - FDR Balances Discovery and Error: Unlike stricter corrections (e.g., Bonferroni), which often result in missing true findings because they are overly conservative, controlling FDR allows researchers to detect more significant results while maintaining a specified proportion of false positives among all significant findings.
  - Improves Interpretability of Results: In practical biological applications, such as genomics and proteomics, FDR control means that of all reported significant gene sets or pathways, only a small, known proportion are expected to be false discoveries. This provides confidence in the reported biological findings and enhances reproducibility.

> The adjusted p-values (bh_adjust) reflect the minimum FDR at which each test can be called significant.
Use these values to determine significance by comparing them to your desired FDR threshold (e.g., 0.05).

## Reduce redundancy

What is a semantic similarity matrix? \>Calculating the similarity matrix and reducing GO terms First step is to get the similarity matrix between terms. The function calculateSimMatrix takes a list of GO terms for which the semantic simlarity is to be calculated, an OrgDb object for an organism, the ontology of interest and the method to calculate the similarity scores.

> NOTE:rrvgo uses the similarity between pairs of terms to compute a distance matrix, defined as (1-simMatrix). The terms are then hierarchically clustered using complete linkage, and the tree is cut at the desired threshold, picking the term with the highest score as the representative of each group. Therefore, higher thresholds lead to fewer groups, and the threshold should be read as the minimum similarity between group representatives.

> If the organism is not supported in Bioconductor, you can still build your own OrgDb object usign the AnnotationForge package and rendering the necessary data for semantic similarity using the GOSemSim package with:

> `my_new_fancy_orgdb_object <- 'org.Zz.eg.db'` `hsGO <- GOSemSim::godata(my_new_fancy_orgdb_object, ont="MF")` Create named scores vector: -log10 adjusted p-values

scores \<- setNames(-log10(GO_BP\$bh_adjust), GO_BP\$category)

```{r}
# Filter for biological process terms with valid p-values
GO_BP <- GO.bp %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>%
  arrange(bh_adjust)

# Calculate semantic similarity matrix
simMatrix <- calculateSimMatrix(
  GO_BP$category, # the column of go terms
  orgdb="org.Ce.eg.db", #c. elegans database
  ont = "BP",
  method = "Rel"
)

```

Create named scores vector.

::: callout-caution The first time I ran this as: `scores <- setNames(-log10(GO_BP$bh_adjust), GO_BP$category)` but some of the bh_adjust values are \~0, and the -log10 of 0 is Inf. This messes with interpretation and visualization down the line, so instead we generate the scores using `.Machine$double.xmin`, which is `~2.225074e-308`, the smallest positive double in R. `pmax()` replaces any 0 with that minimum before transforming it with `-log10()`.

```{r}
# Create named scores vector
scores <- setNames(
  -log10(pmax(GO_BP$bh_adjust, .Machine$double.xmin)),
  GO_BP$category
)

# Reduce GO terms
reducedTerms <- reduceSimMatrix(
  simMatrix,
  scores = scores,
  threshold = 0.7,
  orgdb = "org.Ce.eg.db"
)



# Done! You can now save and plot
write_csv(reducedTerms, "../../output/07_go/reducedTerms.csv")

simMatrix_save <- simMatrix %>%
  as.data.frame() %>% 
  rownames_to_column(var = "GO")

write_csv(simMatrix_save, "../../output/07_go/simMatrix.csv")
```

# Plotting & interpretation

## Similarity matrix heatmap

Plot similarity matrix as a heatmap, with clustering of columns of rows turned on by default (thus arranging together similar terms).

```{r}
heatmapPlot(simMatrix,
            reducedTerms,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

## Scatter plot depicting groups and distance between terms

Plot GO terms as scattered points. Distances between points represent the similarity between terms, and axes are the first 2 components of applying a PCoA to the (di)similarity matrix. Size of the point represents the provided scores or, in its absence, the number of genes the GO term contains.

```{r}
scatterPlot(simMatrix, reducedTerms)
```

## Treemap plot

Treemaps are space-filling visualization of hierarchical structures. The terms are grouped (colored) based on their parent, and the space used by the term is proportional to the score. Treemaps can help with the interpretation of the summarized results and also comparing different sets of GO terms.

```{r}
treemapPlot(reducedTerms)
```

## Word cloud

Word clouds are visualizations which reproduce a text putting emphasis to words which appear frequently in a text. They can help to identify processes and functions that happen more commonly in a set of enriched GO terms, as well as comparing between different sets.

```{r}
wordcloudPlot(reducedTerms, min.freq=1, colors="black")
```

## Chord diagram

Using [GOplot](https://cran.r-project.org/web/packages/GOplot/vignettes/GOplot_vignette.html)

### Create circ object using `circle_dat()`

> `circle_dat()` takes two data frames as an input. The first one contains the results of the functional analysis and should have at least four columns (category, term, genes, adjusted p-value). Additionally, a data frame of the selected genes and their logFC is needed.

Data object for plotting needs two data frame inputs:


1.  **result** of functional analysis with
    1. category: GO term annotation type (e.g., "BP", "MF", "CC")
    2. ID: GO term identifier (e.g., "GO:0008150")
    3. term: GO term description
    4. adj_pval: the `bh_adjust` column from the GO enrichment analysis

> The significance of a term is indicated by the adjusted p-value (adj_pval). Terms with an adjusted p-value < 0.05 are considered as **significantly enriched** and are more likely to provide reliable information.

`GO_BP` is the result of our functional analysis...
 
```{r}
str(GO_BP)
```


```{r}
results_circ <- GO_BP %>% 
  rename(ID = category,
         category = ontology,
         adj_pval = bh_adjust) %>% 
  select(category, ID, term, adj_pval)

head(results_circ)
```

2.  list of selected **genes** and log2FC from a DESeq2 result
    1. genes: (Your gene ID (use your long gene IDs directly))
    2. log2FC: from one of the DESeq2 contrasts

```{r}
genes_circ <- sig_genes_anno %>% 
  select(genes = gene_id, 
         logFC = earlygastrula_low_log2FC, 
         adj_pval = earlygastrula_low_padj)

head(genes_circ)
```

The circle_dat() fucntion will generate a count and a zscore column
    - count: Number of genes per GO term
    - zscore: GO term z-score, typically calculated as (number of upregulated genes minus downregulated genes)/sqrt(count) for that term




```{r}
terms_circ <- GO.terms %>% 
  rename(ID = GOs,
         genes = gene)

head(terms_circ)
```

So we have to join results_circ to terms_circ

```{r}
circ_df <- results_circ %>%  
  inner_join(terms_circ, by = "ID")
```

```{r}
circ <- circle_dat(circ_df, genes_circ)
```

```{r}
circ_df <- circ_df %>% 
  inner_join(genes_egl_circ, by = "genes") %>% 
  rename(logFC = earlygastrula_low_log2FC,
         adj_pval = earlygastrula_low_padj) 
str(circ_df)
```

Add zscore column

> The last column contains the zscore, an easy to calculate value to give you a hint if the biological process (/molecular function/cellular components) is more likely to be decreased (negative value) or increased (positive value). $$
> zscore = \frac{(up - down)}{\sqrt(count)}
> $$ Whereas up and down are the number of assigned genes up-regulated (logFC\>0) in the data or down- regulated(logFC\<0), respectively.

```{r}
circ_df <- circ_df %>%
  group_by(ID) %>%
  mutate(zscore = (sum(logFC > 0) - sum(logFC < 0)) / sqrt(count)) %>%
  ungroup()
```

```{r}
# 4. Select and arrange columns to match GOplot 'circ' format
circ <- circ_df %>%
  select(category,
         ID,
         term,
         count,
         genes,
         logFC,
         adj_pval,
         zscore) %>% 
  as.data.frame()

# circ is now ready for GOChord
str(circ)
```

> Therefore, you want to include more than one logFC value per gene. To adjust to this situation you should use the nlfc argument. It is a numeric value and it defines the number of logFC columns within your binary membership matrix. The default is '1' assuming that most of the time you just have one contrast and one logFC value per gene.

```{r}
circ_df <- circ %>% mutate(
  genes = trimws(as.character(genes)),
  ID   = trimws(as.character(ID))
) %>% as.data.frame()

genes_df <- circ_df %>%
  select(genes, logFC) %>%
  distinct() %>%
  mutate(genes = trimws(as.character(genes))) %>%
  as.data.frame()

processes_df <- circ_df %>%
  select(category, ID, term, count, zscore) %>%
  distinct() %>%
  mutate(ID = trimws(as.character(ID))) %>%
  as.data.frame()
```

### Create binary matrix using `chord_dat()`
```{r}
chord <- chord_dat(
  data    = circ_df,
  genes   = genes_df,
  process = processes_df
)

# Plot
GOChord(chord, gene.order = "logFC")
```

```{r}
# Filter circ to keep GO terms with count >= 2 (or a threshold you choose)
circ_filtered <- circ %>% 
  filter(count >= 2)

# Build the summarized 'processes' data frame for chord plotting
processes_filtered <- circ_filtered %>%
  select(category, ID, term, count, zscore) %>%
  distinct()

# Get unique genes and logFC values for the filtered dataset
genes_filtered <- circ_filtered %>%
  select(genes, logFC) %>%
  distinct()

# Convert to data.frame if needed
circ_filtered_df <- as.data.frame(circ_filtered)
processes_filtered_df <- as.data.frame(processes_filtered)
genes_filtered_df <- as.data.frame(genes_filtered)

# Now create chord input with filtered data
chord_filtered <- chord_dat(
  data = circ_filtered_df,
  genes = genes_filtered_df,
  process = processes_filtered_df
)

# Plot the chord diagram
GOChord(chord_filtered, gene.order = "logFC")
```

> If the number of genes in your genes argument doesn't exactly match the rows in data, or the number of processes in process doesn't match the columns in data, you get subscript out of bounds.

```{r}
# Load the included dataset
data(EC)

# Building the circ object
excirc <- circle_dat(EC$david, EC$genelist)

# Building the binary matrix
exchord <- chord_dat(excirc, EC$genes, EC$process)

GOChord(exchord, gene.order = "logFC")
```

```{r}
str(excirc)
```
