---
title: "Gene ontology (GO) term analysis"
subtitle: "Now that we've identified the DEGs, what do they do?"
author: "Sarah Tanja"
date: 2025-06-17
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

# Background

We identified 44 genes that were broadly impacted by leachate across all stages (meaning transcript abundance of those genes were significantly different in at least one dose of leachate compared to the control in at least one stage). Here we want to know what those genes do!

We do this using the R package `goseq`. Why is `goseq` different from other enrichment tools? Typical enrichment methods (like Fisher's exact test or hypergeometric tests) assume that every gene has an equal probability of being selected as a DEG. But in RNA-Seq longer genes are more likely to be detected as DEGs just because they accumulate more reads. `goseq` corrects for this bias.

## Inputs

-   `LRT_significant_genes_combined_wald.csv` A dataframe of the 44 genes and their LRT and Wald test statistics
-   functional annotation: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz (EggNog)
-   functional annotation http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz (KEGG)

## Outputs

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  dpi = 600,
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

## Install packages

```{r, eval=FALSE}
if ("Biostrings" %in% rownames(installed.packages()) == 'FALSE') install.packages('Biostrings')
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')
if ("goseq" %in% rownames(installed.packages()) == 'FALSE') install.packages('goseq')
install.packages("GOplot")
remotes::install_github("mdscheuerell/flexoki")
```

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("GenomicRanges")
BiocManager::install("goseq")
BiocManager::install("GO.db")
BiocManager::install("rrvgo")
BiocManager::install("org.Ce.eg.db") # c.elegens genome-wide annotation
BiocManager::install("AnnotationForge")
BiocManager::install("GOSemSim")
```

## Load packages

```{r}
library(GenomicRanges)
library(Biostrings)
library(goseq)
library(rrvgo)
library(org.Ce.eg.db)
library(GO.db)
library(GOplot)
library(AnnotationForge)
library(GOSemSim)
library(tidyverse)
library(flexoki)
library(plotly)
```

## Read in gene lists and counts

```{r}
all_genes <- read.csv("../../output/06_deg/interactions/all_genes.csv")
sig_genes <- read.csv("../../output/06_deg/interactions/sig_genes.csv")
sig_counts <- read.csv("../../output/06_deg/interactions/sig_counts.csv", row.names="gene_id", check.names = FALSE)
```

## Import fasta

```{r}
# Load your transcriptome
all_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")  

# Filter for the 44 significant interaction genes
sig_seqs <- all_seqs[names(all_seqs) %in% sig_genes]

# Save to new FASTA file
writeXStringSet(sig_seqs, "../../output/07_go/significant_genes.fasta")
```

## Download annotation & enrichment files

```{r, engine='bash', eval=FALSE}
cd ../../input/annotations

wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz
wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz
```

Unzip

```{r, engine='bash', eval=FALSE}
cd ../../input/annotations
gunzip *.gz
```

## Import to R environment

```{r}
# Read in the file (it will include "#query" as a column name)
EggNog_anno <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")
```

rename "#query" to `gene_id`

```{r}

EggNog_anno <- EggNog_anno %>%
  rename(gene_id="#query")

nrow(EggNog_anno)
names(EggNog_anno)
```

> This functional annotation has 24,072 genes. This is 44% of total protein predicted genes described in the publication associated with this annotation. https://academic.oup.com/gigascience/article/doi/10.1093/gigascience/giac098/6815755

# Annotate significant genes

Match up genes in sig_genes dataframe to annotation dataframe

```{r}
sig_genes_anno <- inner_join(sig_genes, EggNog_anno, by = "gene_id")

nrow(sig_genes_anno)
```

::: callout-note
37 of the 44 significant genes were annotated! The annotation file `sig_genes_anno` dataframe now only contains genes that were significant and that have annotation information.
:::

Get gene length information.

```{r}
#import file
gff <- rtracklayer::import("../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3") #if this doesn't work, restart R and try again 

transcripts <- subset(gff, type == "transcript") #keep only transcripts 

transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE) #extract length information

transcript_lengths <- width(transcripts_gr) #isolate length of each gene

gene_id <- transcripts_gr$ID #extract list of gene id 

lengths <- cbind(gene_id, transcript_lengths)

lengths <- as.data.frame(lengths) #convert to data frame
```

Add in length to the annotation file (`sig_genes_anno$transcript_lengths`).

```{r}
sig_genes_anno <- sig_genes_anno %>% left_join(lengths, by = "gene_id")
```

Format GO terms to remove dashes and quotes.

```{r}
#remove dashes, remove quotes from columns and separate by semicolons (replace , with ;) in KEGG_ko, KEGG_pathway, and Annotation.GO.ID columns
head(sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub(",", ";", sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub('"', "", sig_genes_anno$GOs)
sig_genes_anno$GOs <- gsub("-", NA, sig_genes_anno$GOs)
```

# Conduct functional enrichment

Conduct functional enrichment using goseq and [rrvgo](https://www.bioconductor.org/packages/release/bioc/vignettes/rrvgo/inst/doc/rrvgo.html). Compare against all genes detected in the dataset that we have annotation information for.

Conduct at biological process level. Filter for adjusted p-value \<0.05.

```{r}
# Define your DEG list — the genes you're testing
ID.vector <- sig_genes_anno$gene_id  # Replace with your actual gene vector


# All genes detected in the dataset with annotation info
ALL.vector <- all_genes$gene_id # gene_ids for all 12565 genes that we used to make our DESeq2 object
ALL.vector <- ALL.vector[ALL.vector %in% EggNog_anno$gene_id] #filtered down to only genes that are annotated

length(ALL.vector)
```

::: callout-note
9563 out of the 12564 genes in our dataset are annotated
:::

```{r}
lengths_anno <- lengths %>% 
  filter(gene_id %in% all_genes$gene_id) %>% 
  filter(gene_id %in% EggNog_anno$gene_id)

LENGTH.vector <- as.integer(lengths_anno$transcript_length)
names(LENGTH.vector) <- ALL.vector  # Needed for bias correction

# Gene to GO term mapping
GO.terms <- sig_genes_anno %>%
  dplyr::select(gene_id, GOs)

# Format GO terms to one term per row
split <- strsplit(as.character(GO.terms$GOs), ";")
split2 <- data.frame(
  gene = rep.int(GO.terms$gene, sapply(split, length)),
  GOs = unlist(split)
)

GO.terms <- split2

# Construct named gene vector: 1 = DEG, 0 = not DEG
gene.vector <- as.integer(ALL.vector %in% ID.vector)
names(gene.vector) <- ALL.vector

# Null probability weighting function (bias for gene length)
# nullp is a function from nadiadavidson/goseq
pwf <- nullp(gene.vector, bias.data = LENGTH.vector)
```

About goseq() command:

| Argument                       | Meaning                                                                                                                                                      |
|--------------------|----------------------------------------------------|
| `pwf`                          | A **probability weighting function** object from `nullp()`. It contains the DEG status of all genes (1 = DEG, 0 = not DEG), plus gene length bias info.      |
| `gene2cat`                     | A data frame or list mapping **genes to functional categories** (like GO terms). In your case, it's a long-format table of gene-GO pairs.                    |
| `test.cats = "GO:BP"`          | This limits the enrichment to only **biological process** GO terms (you could also use `"GO:MF"` or `"GO:CC"` for molecular function or cellular component). |
| `method = "Wallenius"`         | This uses the **Wallenius noncentral hypergeometric distribution**, which accounts for sampling bias (like gene length).                                     |
| `use_genes_without_cat = TRUE` | Allows the method to consider genes that **do not have a GO annotation** (prevents them from being excluded by default).                                     |

`goseq()` returns a data frame where each row is a GO term, with:

-   The GO ID

-   Category (BP, MF, or CC)

-   Over-representation p-value (for DEGs)

-   Under-representation p-value (less commonly used)

-   Number of DEGs in that GO term

-   Total number of genes in that GO term

-   Other metadata (like GO term name)

## GO enrichment

```{r}
# Run GO enrichment for biological process terms
GO.bp <- goseq(
  pwf,
  gene2cat = GO.terms,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

# Format and filter results
GO.bp <- GO.bp %>% 
  arrange(over_represented_pvalue)

# Use the Benjamini-Hochberg method to control false discovery rate 
GO.bp$bh_adjust <- p.adjust(GO.bp$over_represented_pvalue, method = "BH")

GO.bp <- GO.bp %>%
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

# Optional: write to file
write_csv(GO.bp, file = "../../output/07_go/functional_enrichment_biological_process_goseq.csv")
```

::: column-margin
##### Why Does Controlling False Discovery Rate Matter in Enrichment Analysis?

The Benjamini-Hochberg procedure controls the proportion of false positives among all rejected hypotheses, which is especially important in high-throughput data like gene ontology enrichment.

Controlling the false discovery rate (FDR) is critical in enrichment analysis, especially when thousands of statistical tests are performed simultaneously (e.g., testing many gene sets, pathways, or terms for enrichment): - Multiple Comparisons Increase False Positives: In enrichment analyses, each hypothesis test (such as whether a specific gene set is enriched) carries a risk of yielding a false positive---detecting significance where there is none. When testing hundreds or thousands of hypotheses, the probability of encountering several false positives increases dramatically. - FDR Balances Discovery and Error: Unlike stricter corrections (e.g., Bonferroni), which often result in missing true findings because they are overly conservative, controlling FDR allows researchers to detect more significant results while maintaining a specified proportion of false positives among all significant findings. - Improves Interpretability of Results: In practical biological applications, such as genomics and proteomics, FDR control means that of all reported significant gene sets or pathways, only a small, known proportion are expected to be false discoveries. This provides confidence in the reported biological findings and enhances reproducibility.

The adjusted p-values (bh_adjust) reflect the minimum FDR at which each test can be called significant. Use these values to determine significance by comparing them to your desired FDR threshold (e.g., 0.05).

*Generated from ChatGPT on 09AUG2025*
:::

## Reduce redundancy

What is a semantic similarity matrix?

> Calculating the similarity matrix and reducing GO terms First step is to get the similarity matrix between terms. The function calculateSimMatrix takes a list of GO terms for which the semantic simlarity is to be calculated, an OrgDb object for an organism, the ontology of interest and the method to calculate the similarity scores.

> NOTE:rrvgo uses the similarity between pairs of terms to compute a distance matrix, defined as (1-simMatrix). The terms are then hierarchically clustered using complete linkage, and the tree is cut at the desired threshold, picking the term with the highest score as the representative of each group. Therefore, higher thresholds lead to fewer groups, and the threshold should be read as the minimum similarity between group representatives.

```{r}
# Filter for biological process terms with valid p-values
GO_BP <- GO.bp %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>%
  arrange(bh_adjust)

# Calculate semantic similarity matrix
simMatrix <- calculateSimMatrix(
  GO_BP$category, # the column of go terms
  orgdb="org.Ce.eg.db", #c. elegans database org.Ce.eg.db
  ont = "BP",
  method = "Rel"
)

```

Create named scores vector.

::: callout-caution
The first time I ran this as: `scores <- setNames(-log10(GO_BP$bh_adjust), GO_BP$category)` but some of the bh_adjust values are \~0, and the -log10 of 0 is Inf. This messes with interpretation and visualization down the line, so instead we generate the scores using `.Machine$double.xmin`, which is `~2.225074e-308`, the smallest positive double in R. `pmax()` replaces any 0 with that minimum before transforming it with `-log10()`.
:::

```{r}
# Create named scores vector
scores <- setNames(
  -log10(pmax(GO_BP$bh_adjust, .Machine$double.xmin)),
  GO_BP$category
)

# Reduce GO terms
reducedTerms <- reduceSimMatrix(
  simMatrix,
  scores = scores,
  threshold = 0.8,
  orgdb = "org.Ce.eg.db"
)

# Done! You can now save and plot
write_csv(reducedTerms, "../../output/07_go/reducedTerms.csv")

simMatrix_save <- simMatrix %>%
  as.data.frame() %>% 
  rownames_to_column(var = "GO")

write_csv(simMatrix_save, "../../output/07_go/simMatrix.csv")
```


# Plotting & interpretation

[Using the rrvgo package](https://bioconductor.org/packages/devel/bioc/vignettes/rrvgo/inst/doc/rrvgo.html)

## Similarity matrix heatmap

Plot similarity matrix as a heatmap, with clustering of columns of rows turned on by default (thus arranging together similar terms).

```{r}
heatmapPlot(simMatrix,
            reducedTerms,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

## Scatter plot depicting groups and distance between terms

Plot GO terms as scattered points. Distances between points represent the similarity between terms, and axes are the first 2 components of applying a PCoA to the (di)similarity matrix. Size of the point represents the provided scores or, in its absence, the number of genes the GO term contains.
```{r}
?scatterPlot
```

- addLabel	
add labels with the most representative term of the group.
```{r, fig.height=10, fig.width=14}
scatterPlot(simMatrix, reducedTerms,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

I want a dataframe that groups my GO terms based on these 5 main clusters displayed in this PCA.
The scatterPlot uses a dimensionality reduction (PCoA) approach to project GO terms into 2D space and then visually groups the points into a smaller number of major clusters. These visual clusters are typically broader groupings, which correspond to higher-level semantic similarity rather than the narrow clusters from the cut tree. A possible approach if you want to create a cluster assignment dataframe matching the 5 visual clusters is to:
- Extract the 2D coordinates from the scatterPlot dimensionality reduction output,
- Apply a clustering method on the 2D coordinates (e.g., k-means with k=5),
- Create a new dataframe linking each GO term to its visual cluster.
This approach aligns the clustering with the displayed scatterPlot clusters rather than the original fine-grained clusters.

```{r}
# Distance from similarity matrix
dist <- as.dist(1 - simMatrix)

# PCoA to 2 dimensions
coord <- cmdscale(dist, k = 2)

# Convert to dataframe with GO terms
df_coords <- data.frame(go = rownames(simMatrix), Dim1 = coord[,1], Dim2 = coord[,2])

```

Cluster the terms visually in 2D space
Use k-means clustering on the 2D coordinates with k = 5 (number of desired clusters).
```{r}
set.seed(123)  # for reproducibility
clusters <- kmeans(df_coords[, c("Dim1", "Dim2")], centers = 5)
df_coords$cluster <- as.factor(clusters$cluster)
```

Combine cluster info with other GO term data
Merge this dataframe with your reducedTerms or other GO annotation information as needed.
```{r}
final_df <- merge(df_coords, reducedTerms, by.x = "go", by.y = "go")
```
This dataframe `final_df` will now include each GO term, its 2D coordinates used for visualization, and its assigned cluster corresponding to the 5 major groups in the scatterPlot.

```{r}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, color = cluster.x, size = score, alpha = 0.4)) +
  geom_point() +
  coord_equal() +
  labs(color = "Cluster (k=5)", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)
```
```{r, fig.height=10, fig.width=15}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, color = parentTerm, size = score)) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)
```
```{r}
p <- ggplot(final_df, aes(
    Dim1, Dim2,
    color = parentTerm,
    size  = score,
    text  = paste0(
      "GO: ", go,
      "<br>Parent Term: ", parentTerm,
      "<br>Term: ", term,
      "<br>Score: ", round(score, 2)
    )
  )) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term",
       size  = "Score",
       title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)

ggplotly(p, tooltip = "text")
```
How to interpret distances

Closer points = more semantically similar GO terms.
That means the gene sets they represent overlap more or describe conceptually related processes.

Farther points = less similar biologically.
Their gene sets don’t overlap much, and the ontology sees them as describing different processes.
The PCoA projection does not use your logFCs or sample-level values — it uses GO term relationships (semantic similarity).

1. What your clusters represent

The clusters from the PCoA + k-means are based on the pairwise semantic similarity scores across all GO terms.

These are data-driven groupings that capture overall gene overlap + ontology proximity.

The algorithm doesn’t “know” about your chosen parentTerm column — it just groups terms that have similar meaning.

2. What your parentTerm variable is

The “parent term” you merged in is a single hierarchical label from the GO DAG (graph).

A GO term can belong under multiple higher-level categories in the ontology, but in your dataset it’s probably annotated with just one chosen parent (often the most specific or most frequent).

Because the GO graph is not a strict tree, related terms may have different parents — so they don’t collapse neatly into one parentTerm cluster.

3. Why they don’t line up

Ontology is many-to-many: two GO terms might be semantically very close (high overlap) but come from different parents in the hierarchy.

Cluster ≠ hierarchy: k-means clusters cut across the graph structure; they group by similarity, not by the official ontology tree.

Parent terms are broad: “cellular process” or “metabolic process” can cover huge swathes of biology, so those categories span multiple similarity clusters.

4. How to check alignment

Look at your clusters and check: are terms within a cluster biologically coherent even if they span different parents? Often they are — they share overlapping gene sets or related functions.

The clusters are “data-driven neighborhoods” of related GO terms.
The parent terms are “administrative labels” from the GO DAG.
They don’t always coincide because biology (and the GO ontology) is more tangled than a simple tree.
Clusters in the 2D space: show neighborhoods of semantically similar terms (terms that describe similar biology and often share many genes).

Point size (score): tells you which terms were more strongly enriched.

Color by parentTerm: just tags each dot with the one “parent” you merged in. The fact that colors are mixed within a cluster shows that ontology categories don’t always line up with semantic neighborhoods.

### Summarize cluster content

```{r}
# Summarize cluster content by parent term
cluster_summary <- final_df %>%
  group_by(cluster.x, parentTerm) %>%
  summarise(
    n_terms = n(),                     # how many GO terms in this parent within cluster
    mean_score = mean(score, na.rm = TRUE),  # average enrichment strength
    max_score = max(score, na.rm = TRUE),    # strongest term in that parent
    .groups = "drop"
  ) %>%
  arrange(cluster.x, desc(n_terms))

# For each cluster, show top 5 parent terms by count (or max score)
top_terms_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = n_terms, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(n_terms))

top_terms_by_cluster
```




## Treemap plot

Treemaps are space-filling visualization of hierarchical structures. The terms are grouped (colored) based on their parent, and the space used by the term is proportional to the score. Treemaps can help with the interpretation of the summarized results and also comparing different sets of GO terms.

```{r}
treemapPlot(reducedTerms)
```

## Word cloud

Word clouds are visualizations which reproduce a text putting emphasis to words which appear frequently in a text. They can help to identify processes and functions that happen more commonly in a set of enriched GO terms, as well as comparing between different sets.

```{r}
wordcloudPlot(reducedTerms, min.freq=1, colors="black")
```

## Chord diagram

Using [GOplot](https://cran.r-project.org/web/packages/GOplot/vignettes/GOplot_vignette.html)

### Create `circ` object using `circle_dat()`

> `circle_dat()` takes two data frames as an input. The first one contains the results of the functional analysis and should have at least four columns (category, term, genes, adjusted p-value). Additionally, a data frame of the selected genes and their logFC is needed.

Data object for plotting needs two data frame inputs:

1.  **result** of functional analysis with
    1.  category: GO term annotation type (e.g., "BP", "MF", "CC")
    2.  ID: GO term identifier (e.g., "GO:0008150")
    3.  term: GO term description
    4.  adj_pval: the `bh_adjust` column from the GO enrichment analysis

> The significance of a term is indicated by the adjusted p-value (adj_pval). Terms with an adjusted p-value \< 0.05 are considered as **significantly enriched** and are more likely to provide reliable information.

`GO_BP` is the result of our functional analysis...

```{r}
str(GO_BP)
```

```{r}
results_circ <- GO_BP %>% 
  rename(ID = category,
         category = ontology,
         adj_pval = bh_adjust,
         count = numInCat) %>% 
  select(category, ID, term, adj_pval, count)
```

And even though the vignette doesn't say you need it, you need a column of genes for each go term.

```{r}
terms_circ <- GO.terms %>% 
  rename(ID = GOs,
         genes = gene)

str(terms_circ)
unique(terms_circ$genes)
```

So we have to join results_circ to terms_circ

```{r}
go_circ <- results_circ %>%  
  full_join(terms_circ, by = "ID") %>% 
  select(category, ID, term, genes, adj_pval, count)

str(go_circ)
unique(go_circ$genes)
```

2.  list of selected **genes** and log2FC from a DESeq2 result
    1.  ID: (Your gene ID (use your long gene IDs directly))
    2.  log2FC: from one of the DESeq2 contrasts

::: callout-note
The Log2FC will be displayed on the chord diagram, and you can add in additional values from other contrasts after the initial circ object is made. For the time being we're going to select the early gastrula low contrast because it had the most DEGs
:::

```{r}
genes_circ <- sig_genes_anno %>% 
  select(ID = gene_id, 
         logFC = earlygastrula_low_log2FC)

str(genes_circ)
```

The circle_dat() function will generate a count and a zscore column - count: Number of genes per GO term - zscore: GO term z-score, typically calculated as (number of upregulated genes minus downregulated genes)/sqrt(count) for that term

::: callout-caution
The `circle_dat()` function expects your gene ids to be named `genes` in your GO results dataframe and `ID` in your gene list dataframe... this is tricky, watch out!
:::

The following code checks that the gene id column names are the same in the two dataframes with differently named columns

```{r}
length(intersect(go_circ$genes, genes_circ$ID))
```

```{r}
circ <- circle_dat(go_circ, genes_circ)
```

`circ` is the object that will now be used for plotting!

```{r}
# circ from circle_dat()
str(circ)
```

::: callout-caution
`circle_dat()` reformatted the gene ids to be all uppercase
:::

```{r}
summary(circ$count)
```

::: callout-caution
Did not appear to calculate `count` correctly... it isn't counting the genes per term but rather the number of times a term is assigned to a gene... which is always 1, This also throws off the zscore calculation which depends on count.
:::

### JK Manually create `circ_df`

```{r}
circ_df <- go_circ %>%  
  full_join(genes_circ, by = c("genes" = "ID")) %>% 
  drop_na(term)

circ_df_parent <- circ_df %>% 
  left_join(reducedTerms, by = c("ID" = "go", "term" = "term"))
```

Add zscore column

```{r}
circ_dfp <- circ_df_parent %>%
  group_by(ID) %>%
  mutate(zscore = (sum(logFC > 0) - sum(logFC < 0)) / sqrt(count)) %>%
  ungroup()
```

::: column-margin
::: callout-note
### What does the zscore tell us?

The z-score in your gene ontology (GO) analysis context is a statistical measure that helps indicate whether a GO term (biological process, molecular function, or cellular component) is more likely to be biologically increased or decreased based on the direction of gene expression changes.

Specifically, your z-score formula $$
zscore = \frac{(up - down)}{\sqrt(count)}
$$ means:

-   "up" is the number of genes assigned to that GO term that are up-regulated (logFC \> 0).

-   "down" is the number of genes assigned to that GO term that are down-regulated (logFC \< 0).

-   "count" is the total number of genes assigned to that GO term.

Interpretation of z-scores:

A positive z-score indicates that the GO term has more up-regulated genes than down-regulated genes relative to its size and is therefore more likely to be biologically increased or activated.

A negative z-score indicates more down-regulated genes and suggests that the GO term is more likely to be biologically decreased or inhibited.

The magnitude of the z-score tells you how strong this trend is, adjusted for the total number of genes in that GO term.

In your data, z-scores ranging approximately from +2 to -1.73 mean:

Values toward +2 imply that some GO terms show a notable prevalence of up-regulated genes, suggesting activation or up-regulation of those biological functions or processes.

Values near -1.73 indicate other GO terms have a moderate dominance of down-regulated genes, implying suppression or down-regulation.

Typically, a z-score greater than 2 or less than -2 is considered significant evidence of activation or inhibition, respectively, though borderline values like yours still provide useful directional hints.

This z-score metric provides a simple, normalized way to capture the overall regulatory direction of each GO term based on the gene expression data from your DESeq2 contrast, giving you an intuitive measure to prioritize and visualize the biological processes likely impacted in your dataset.

Thus, in your chord diagram or other visualizations, z-score can be used to color or weight GO terms and gene links to quickly highlight activated vs. repressed functions.

*Generated via perplexity AI on 15AUG2025*
:::
:::

```{r}
# Select and arrange columns to match GOplot 'circ' format
circ_dfp <- circ_dfp %>%
  select(!term) %>% # remove term colunm
  mutate(term = parentTerm) %>% # replace it with the parent term column
  select(category,
         ID,
         term,
         count,
         genes,
         logFC,
         adj_pval,
         zscore) %>% 
  as.data.frame()
```


### Create binary matrix using `chord_dat()`

::: callout-caution
`chord_dat()` expects the `genes =` to have genes in a column named `ID` with a `logFC` column
The matrix should be: genes as rows, terms as columns (binary matrix), and one column named “logFC”.
:::

```{r}
circ_genes <- circ_dfp %>% 
  select(genes, logFC) %>% 
  rename(ID = genes) %>% 
  distinct()
```


```{r binary-matrix-chord_dat}
chord <- chord_dat(
  data = circ_dfp,
  genes = circ_genes,
  process = c("regulation of RNA metabolic process", 
              "positive regulation of biological process", 
              "cellular metabolic process", 
              "response to stress", 
              "response to chemical",
              "cell differentiation",
              "regulation of signaling",
              "organonitrogen compound metabolic process"))
```


### Plot chord diagram
```{r}
library(viridis)
library(RColorBrewer)
```

```{r}
display.brewer.all()
```
viridis
magma
plasma
inferno
cividis
mako
rocket
turbo

```{r}
library(scales)
show_col(turbo(5))
```

```{r}
ribbons <- turbo(5)
ribbons_viridis <- viridis(5)
ribbons_pastel <- brewer.pal(5, "Pastel2")
```

```{r plot-chord, fig.height=10, fig.width=9}
chordplot <- GOChord(chord, 
                     gene.order = "logFC",
                     border.size = 0.2,
                     gene.size = 3,
                     title = "Enriched GO terms across significantly different interaction genes",
                     process.label = 10,
                     space = 0.02,
                     gene.space = 0.25)
chordplot
```

```{r}
ggsave("../../output/07_go/chord_plot.png", width = 12, height = 12, dpi = 600 )
```


> Therefore, if you want to include more than one logFC value per gene. To adjust to this situation you should use the nlfc argument. It is a numeric value and it defines the number of logFC columns within your binary membership matrix. The default is '1' assuming that most of the time you just have one contrast and one logFC value per gene.



# GOHeat

```{r}
GOHeat(chord, nlfc = 1, fill.col = c('red', 'white', 'blue'))
```

