---
title: "Gene ontology (GO) enrichment plots"
subtitle: "We've matched the DEGs with GO terms, and identified enriched GOs... let's visualize them!"
author: "Sarah Tanja"
date: 2025-09-30
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

The size column in the reduced results represents the number of genes annotated to each GO term. size \> 0 → that term is associated with at least one gene in your gene universe (and was used in the GOseq enrichment test). size = 0 → no genes in your input set are annotated to that term, even though the term appears in the GO semantic similarity matrix.

Why size = 0 can appear: You built the similarity matrix (calculateSimMatrix) from all GO terms, but then reduced it using a filtered subset (e.g., only enriched terms). Some leftover terms in the similarity matrix may not actually appear in your results' gene annotations. The term was pruned or filtered by GOseq or rrvgo during processing --- for example, if a cluster "parent" term wasn't part of the original enrichment table, but was added as a representative. Annotation differences --- GOseq uses your gene-to-GO mapping (often via gene2cat), while rrvgo's internal lookup (from GO.db) may include additional terms that don't exist in your mapping.

Relationship between score and size

There's no inherent correlation between score and size because they describe different things. A small, specific GO term with few genes (size = 3) might have a very low p-value, if all 3 of those genes are strongly enriched --- → high score (e.g. 307 ≈ --log₁₀(10⁻³⁰⁷)). A large, general term with many genes (size = 321) might not be strongly overrepresented --- → modest p-value → low score (≈ 1.895 → p ≈ 0.0126). That's why score and size often move in opposite directions: small, specific terms can show stronger enrichment signals. GOseq adjusts for gene length bias, so the resulting *p*-values already account for uneven annotation probabilities. Still, a term with few genes can have a small *p*-value if those genes are all DE, whereas a large term with hundreds of genes often dilutes that signal.

Score ≈ significance, not size.

A small size + high score = "narrow but very strong enrichment."

A large size + low score = "broad but weak enrichment."

When reducing with rrvgo, the highest-scoring term in each cluster becomes the representative --- typically the most statistically significant, not necessarily the largest. I**n GOseq (and therefore in rrvgo)**, the `size` column does **not** represent the number of DEGs in that GO term. It represents the **number of genes in your background (gene universe)** that are annotated to that GO term.

🧬 1. What GOseq tests

GOseq's enrichment test works like this:

You provide a set of DEGs and a background (usually all genes tested by DESeq2 or edgeR).

For each GO term, GOseq asks:

-   Are DEGs overrepresented among genes annotated to this term compared to what I'd expect by chance (accounting for gene length bias)?

-   *Given the set of genes that **could have been called DE**, are DEGs over-represented in this GO term compared to what I'd expect by chance?*

So each GO term has:

numDEInCat = number of DEGs annotated to that term.

numInCat = number of total genes (background) annotated to that term.

Then it computes a p-value for enrichment.

## Inputs

-   reduced terms
-   semantic similarity matrix

## Outputs

# Setup
## Set paths
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

## Install packages

```{r}
## CRAN packages-----------
#install.packages("GOplot")
#install.packages("plotly")
#install.packages("tidyverse")

## Bioconductor packages---------
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("rrvgo")
```

## Load packages

```{r}
library(GOplot)
library(plotly)
library(tidyverse)
library(rrvgo)
library(kableExtra)
```

## Load inputs

```{r}
## Reduced Terms
reducedTerms <- read.csv(file.path(input_path, "reducedTerms_GO.csv"))
reducedTerms.anno <- read.csv(file.path(input_path, "reducedTerms_GO_anno.csv"))

## Similarity matrices
simMatrix <- read_csv(file.path(input_path, "simMatrix.csv")) %>% 
  column_to_rownames("GO") %>% as.matrix()  # uses the GO column as rownames

```
> The rows of simMatrix should correspond (in order and identity) to the rows of reducedTerms.
Ensure that the go column is character, not factor. Ensure simMatrix is a data.frame or matrix, and reducedTerms is a data.frame, not a tibble! \*which is why we load it with read.csv, (which loads a data.frame) and not read_csv (which loads a tibble).


# Plotting & interpretation

[Using the rrvgo package](https://bioconductor.org/packages/devel/bioc/vignettes/rrvgo/inst/doc/rrvgo.html)

## Similarity matrix heatmap

Plot similarity matrix as a heatmap, with clustering of columns of rows turned on by default (thus arranging together similar terms).

```{r}
heatmapPlot(simMatrix,
            reducedTerms,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```


## Scatter plot

Plot GO terms as scattered points. Distances between points represent the similarity between terms, and axes are the first 2 components of applying a PCoA to the (di)similarity matrix. Size of the point represents the provided scores or, in its absence, the number of genes the GO term contains.

```{r, fig.height = 7}
scatterPlot(simMatrix, reducedTerms,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```


# Custom PCA

## Significant DEG

I want a dataframe that groups my GO terms based on the 5 main clusters displayed in this PCA. The scatterPlot uses a dimensionality reduction (PCoA) approach to project GO terms into 2D space and then visually groups the points into a smaller number of major clusters. These visual clusters are typically broader groupings, which correspond to higher-level semantic similarity rather than the narrow clusters from the cut tree. A possible approach if you want to create a cluster assignment dataframe matching the 5 visual clusters is to: - Extract the 2D coordinates from the scatterPlot dimensionality reduction output, - Apply a clustering method on the 2D coordinates (e.g., k-means with k=5), - Create a new dataframe linking each GO term to its visual cluster. This approach aligns the clustering with the displayed scatterPlot clusters rather than the original fine-grained clusters.

```{r}
# Distance from similarity matrix
dist <- as.dist(1 - simMatrix.sig)

# PCoA to 2 dimensions
coord <- cmdscale(dist, k = 2)

# Convert to dataframe with GO terms
df_coords <- data.frame(go = rownames(simMatrix.sig), Dim1 = coord[,1], Dim2 = coord[,2])
```

Cluster the terms visually in 2D space Use k-means clustering on the 2D coordinates with k = 5 (number of desired clusters).

```{r}
set.seed(123)  # for reproducibility
clusters <- kmeans(df_coords[, c("Dim1", "Dim2")], centers = 5)
df_coords$cluster <- as.factor(clusters$cluster)
```

Combine cluster info with other GO term data Merge this dataframe with your reducedTerms or other GO annotation information as needed.

```{r}
final_df <- merge(df_coords, reducedTerms.sig, by.x = "go", by.y = "go")
```

This dataframe `final_df` will now include each GO term, its 2D coordinates used for visualization, and its assigned cluster corresponding to the 5 major groups in the scatterPlot.

```{r}
head(final_df)
```

Drop size = 0

```{r}
sigdeg_pcoa <- final_df %>% 
  filter(size > 0)
```

### Interactive plot

For exploration

```{r, fig.width=12}
p <- ggplot(sigdeg_pcoa, aes(
    Dim1, Dim2,
    color = parentTerm,
    size  = score,
    text  = paste0(
      "GO: ", go,
      "<br>Parent Term: ", parentTerm,
      "<br>Term: ", term,
      "<br>Score: ", round(score, 2),
      "<br>Size: ", size
    )
  )) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term",
       size  = "Score",
       title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)

ggplotly(p, tooltip = "text")
```

### Table of top parent terms

```{r}
# Summarize cluster content by parent term
cluster_summary <- sigdeg_pcoa %>%
  group_by(cluster.x, parentTerm) %>%
  summarise(
    n_terms = n(),                     # how many GO terms in this parent within cluster
    mean_score = mean(score, na.rm = TRUE),  # average enrichment strength
    max_score = max(score, na.rm = TRUE),    # strongest term in that parent
    .groups = "drop"
  ) %>%
  arrange(cluster.x, desc(n_terms))

# For each cluster, show top 5 parent terms by count (or max score)
top_parent_terms_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = n_terms, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(n_terms))

knitr::kable(top_parent_terms_by_cluster)
```

### Table of top terms

```{r}
# Summarize cluster content by parent term
cluster_summary <- sigdeg_pcoa %>%
  group_by(cluster.x, term) %>%
  summarise(
    n_terms = n(),                     # how many GO terms in this parent within cluster
    mean_score = mean(score, na.rm = TRUE),  # average enrichment strength
    max_score = max(score, na.rm = TRUE),    # strongest term in that parent
    .groups = "drop"
  ) %>%
  arrange(cluster.x, desc(n_terms))

# For each cluster, show top 5 terms by count (or max score)
top_terms_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = n_terms, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(n_terms))

knitr::kable(top_terms_by_cluster)
```

### Table of top scores

```{r}
# For each cluster, show top 5 parent terms by count (or max score)
top_score_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = mean_score, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(mean_score))

knitr::kable(top_score_by_cluster)
```

## Manuscript plots

```{r}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(sigdeg_pcoa, aes(Dim1, Dim2, size = score, label = parentTerm)) +
  geom_point(alpha = 0.3) +
  geom_text_repel()+
  coord_equal()+
  labs(title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)
```

```{r, fig.height=10, fig.width=15}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, color = parentTerm, size = score)) +
  geom_jitter(alpha = 0.5)+
  coord_equal() +
  labs(color = "Parent Term", title = "GO parent terms clustered in PCoA space") +
  theme_minimal(base_size = 12)+
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9))
```

```{r, fig.height=17}
ggplot(sigdeg_pcoa, aes(Dim1, Dim2, color = term, size = score)) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",      # move below
    legend.box = "vertical",         # stack if multiple guides
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  ) +
  guides(                                  # hide size legend (optional)
    color = guide_legend(nrow = 40, byrow = TRUE)     # wrap legend into rows
  )
```

::: column-margin
::: callout-note
How to interpret distances

Closer points = more semantically similar GO terms. That means the gene sets they represent overlap more or describe conceptually related processes.

Farther points = less similar biologically. Their gene sets don't overlap much, and the ontology sees them as describing different processes. The PCoA projection does not use your logFCs or sample-level values --- it uses GO term relationships (semantic similarity).

1.  What your clusters represent

The clusters from the PCoA + k-means are based on the pairwise semantic similarity scores across all GO terms.

These are data-driven groupings that capture overall gene overlap + ontology proximity.

The algorithm doesn't "know" about your chosen parentTerm column --- it just groups terms that have similar meaning.

2.  What your parentTerm variable is

The "parent term" you merged in is a single hierarchical label from the GO DAG (graph).

A GO term can belong under multiple higher-level categories in the ontology, but in your dataset it's probably annotated with just one chosen parent (often the most specific or most frequent).

Because the GO graph is not a strict tree, related terms may have different parents --- so they don't collapse neatly into one parentTerm cluster.

3.  Why they don't line up

Ontology is many-to-many: two GO terms might be semantically very close (high overlap) but come from different parents in the hierarchy.

Cluster ≠ hierarchy: k-means clusters cut across the graph structure; they group by similarity, not by the official ontology tree.

Parent terms are broad: "cellular process" or "metabolic process" can cover huge swathes of biology, so those categories span multiple similarity clusters.

4.  How to check alignment

Look at your clusters and check: are terms within a cluster biologically coherent even if they span different parents? Often they are --- they share overlapping gene sets or related functions.

The clusters are "data-driven neighborhoods" of related GO terms. The parent terms are "administrative labels" from the GO DAG. They don't always coincide because biology (and the GO ontology) is more tangled than a simple tree. Clusters in the 2D space: show neighborhoods of semantically similar terms (terms that describe similar biology and often share many genes).

Point size (score): tells you which terms were more strongly enriched.

Color by parentTerm: just tags each dot with the one "parent" you merged in. The fact that colors are mixed within a cluster shows that ontology categories don't always line up with semantic neighborhoods.
:::
:::

::: callout-note

Based on frequency of enriched terms, the clusters can be described: - 1

:   Developmental, cellular, nuclear division - 2 : Regulation of metabolic processes - 3 : cellular stress response , response to chemical - 4 : metabolic / catabolic process - 5 : regulation cell cycle communication
:::

```{r}
tbl_min <- top_score_by_cluster %>%
  transmute(
    Cluster = cluster.x,
    `Parent term` = parentTerm,
    n = as.integer(n_terms),
    `Mean score` = round(mean_score, 1),
    `Max score` = round(max_score, 1)
  ) %>%
  arrange(Cluster, desc(`Mean score`), desc(`Max score`))

kable(
  tbl_min,
  format = "pipe",                            # <- Pandoc table -> native Word/Docs
  align  = c("r", "l", "r", "r", "r"),
  caption = "Top GO terms by cluster"
)
```

## Treemap plot

Treemaps are space-filling visualization of hierarchical structures. The terms are grouped (colored) based on their parent, and the space used by the term is proportional to the score. Treemaps can help with the interpretation of the summarized results and also comparing different sets of GO terms.

### Interactions

```{r}
treemapPlot(reducedTerms.sig)
```

#### pattern 1

```{r}
treemapPlot(reducedTerms.1)
```

#### pattern 2

```{r}
treemapPlot(reducedTerms.2)
```

#### pattern 3

```{r}
treemapPlot(reducedTerms.3)
```

# Lollipop plot?

```{r}
go_lolli <- early_data %>% 
  dplyr::arrange(bh_adjust) %>%
  mutate(percHits=(numDEInCat/numInCat)*100)%>%
  
  ggplot(aes(x=percHits, y=term, fill=bh_adjust, colour=bh_adjust, size=percHits)) +
  facet_grid(ParentTerm ~ ., scales = "free", space='free', labeller = label_wrap_gen(width = 30, multi_line = TRUE))+
  geom_point(aes(group=term))+
  geom_segment(aes(x=0, xend=percHits, y=term, yend=term),linewidth=1) +
  scale_colour_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_fill_gradientn(name = "Adj. p-value", colours = colorspace::heat_hcl(7))+
  scale_size(name="Hits (%)", range = c(1, 5))+
  theme_classic() +
  ylab("GO Term")+
  xlab("Hits (%)")+
  ggtitle("Early Development (1-38 hpf); 2,679 genes")+
  theme(legend.position = "none",
        axis.text.y = element_text(size=18),
        axis.text.x = element_text(size=18),
        axis.title.y = element_text(size=24, face="bold"),
        axis.title.x = element_text(size = 24, face = "bold"),
        plot.title = element_text(size = 24, face = "bold"), 
        strip.text.y.right=element_text(angle=0, size=18), 
        legend.title=element_text(size=22),
        legend.text=element_text(size=18)); go_plot_earlyA
```

# Summary & next steps

Produced visualizations: - Heatmap of GO term similarity. - Scatter plot grouping GO terms into 5 major clusters. - Treemap to summarize enriched functions. Outputs:

-   Tables of enriched GO terms, reduced clusters, semantic similarity matrices, and gene-to-term assignments.
-   Multiple graphical summaries (heatmap, PCA scatter, treemap)

```{r}
sessionInfo()
```
