---
title: "Gene ontology (GO) enrichment plots"
subtitle: "We've identified the DEGs... let's visualize what they do!"
author: "Sarah Tanja"
date: 2025-09-30
date-modified: today
format:
  docx: default
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

# Background

Resources: A little help from Grace Crandall's code
[here](https://github.com/grace-ac/paper-pycno-sswd-2021-2022/blob/main/code/28-chord-plots.Rmd). -
\[A tutorial using
`circlize`\](https://bioinfo4all.wordpress.com/2021/03/13/tutorial-7-how-to-do-chord-diagram-using-r/ -
[Another `circlize`
example](https://jokergoo.github.io/circlize_book/book/the-chorddiagram-function.html)
\## Inputs - GO.BP.xx - GO.terms.xx - sig.genes.anno

## Outputs

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  fig.height=7, 
  fig.width=9,
  dpi = 600,
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

## Install packages

```{r}
## CRAN packages-----------
#install.packages("GOplot")
#install.packages("plotly")
#install.packages("tidyverse")

## Bioconductor packages---------
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("rrvgo")
```

## Load packages

```{r}
library(GOplot)
library(plotly)
library(tidyverse)
library(rrvgo)
library(kableExtra)
```

# GOplot

## Chord diagram

Using
[GOplot](https://cran.r-project.org/web/packages/GOplot/vignettes/GOplot_vignette.html)

### Create `circ` object using `circle_dat()`

> `circle_dat()` takes two data frames as an input. The first one
> contains the results of the functional analysis and should have at
> least four columns (category, term, genes, adjusted p-value).
> Additionally, a data frame of the selected genes and their logFC is
> needed.

Data object for plotting needs two data frame inputs:

1.  **result** of functional analysis with
    1.  category: GO term annotation type (e.g., "BP", "MF", "CC")
    2.  ID: GO term identifier (e.g., "GO:0008150")
    3.  term: GO term description
    4.  adj_pval: the `bh_adjust` column from the GO enrichment analysis

> The significance of a term is indicated by the adjusted p-value
> (adj_pval). Terms with an adjusted p-value \< 0.05 are considered as
> **significantly enriched** and are more likely to provide reliable
> information.

`GO.BP.int` is the result of our functional analysis...

```{r}
str(GO.BP.int)
```

```{r}
results_circ <- GO.BP.int %>% 
  rename(ID = category,
         category = ontology,
         adj_pval = bh_adjust,
         count = numInCat) %>% 
  select(category, ID, term, adj_pval, count)
```

And even though the vignette doesn't say you need it, you need a column
of genes for each go term.

```{r}
terms_circ <- GO.terms.int %>% 
  rename(ID = GOs,
         genes = gene_id)

str(terms_circ)
unique(terms_circ$genes)
```

So we have to join results_circ to terms_circ

```{r}
go_circ <- results_circ %>%  
  full_join(terms_circ, by = "ID") %>% 
  select(category, ID, term, genes, adj_pval, count)

str(go_circ)
unique(go_circ$genes)
```

2.  list of selected **genes** and log2FC from a DESeq2 result
    1.  ID: (Your gene ID (use your long gene IDs directly))
    2.  log2FC: from one of the DESeq2 contrasts

::: callout-note
The Log2FC will be displayed on the chord diagram, and you can add in
additional values from other contrasts after the initial circ object is
made. For the time being we're going to select the early gastrula low
contrast because it had the most DEGs
:::

sig_genes_anno is the annotated results from DESeq2

```{r}
genes_circ <- sig_genes_anno %>% 
  select(ID = gene_id, 
         logFC = earlygastrula_high_log2FC)

str(genes_circ)
```

The circle_dat() function will generate a count and a zscore column -
count: Number of genes per GO term - zscore: GO term z-score, typically
calculated as (number of upregulated genes minus downregulated
genes)/sqrt(count) for that term

::: callout-caution
The `circle_dat()` function expects your gene ids to be named `genes` in
your GO results dataframe and `ID` in your gene list dataframe... this
is tricky, watch out!
:::

The following code checks that the gene id column names are the same in
the two dataframes with differently named columns

```{r}
length(intersect(go_circ$genes, genes_circ$ID))
```

```{r}
circ <- circle_dat(go_circ, genes_circ)
```

`circ` is the object that will now be used for plotting!

```{r}
# circ from circle_dat()
str(circ)
```

::: callout-caution
`circle_dat()` reformatted the gene ids to be all uppercase
:::

```{r}
summary(circ$count)
```

::: callout-caution
Did not appear to calculate `count` correctly... it isn't counting the
genes per term but rather the number of times a term is assigned to a
gene... which is always 1, This also throws off the zscore calculation
which depends on count.
:::

### JK Manually create `circ_df`

```{r}
circ_df <- go_circ %>%  
  full_join(genes_circ, by = c("genes" = "ID")) %>% 
  drop_na(term)

circ_df_parent <- circ_df %>% 
  left_join(reducedTerms.all, by = c("ID" = "go", "term" = "term"))
```

Add zscore column

```{r}
circ_dfp <- circ_df_parent %>%
  group_by(ID) %>%
  mutate(zscore = (sum(logFC > 0) - sum(logFC < 0)) / sqrt(count)) %>%
  ungroup()
```

::: column-margin
::: callout-note
### What does the zscore tell us?

The z-score in your gene ontology (GO) analysis context is a statistical
measure that helps indicate whether a GO term (biological process,
molecular function, or cellular component) is more likely to be
biologically increased or decreased based on the direction of gene
expression changes.

Specifically, your z-score formula $$
zscore = \frac{(up - down)}{\sqrt(count)}
$$ means:

-   "up" is the number of genes assigned to that GO term that are
    up-regulated (logFC \> 0).

-   "down" is the number of genes assigned to that GO term that are
    down-regulated (logFC \< 0).

-   "count" is the total number of genes assigned to that GO term.

Interpretation of z-scores:

A positive z-score indicates that the GO term has more up-regulated
genes than down-regulated genes relative to its size and is therefore
more likely to be biologically increased or activated.

A negative z-score indicates more down-regulated genes and suggests that
the GO term is more likely to be biologically decreased or inhibited.

The magnitude of the z-score tells you how strong this trend is,
adjusted for the total number of genes in that GO term.

In your data, z-scores ranging approximately from +2 to -1.73 mean:

Values toward +2 imply that some GO terms show a notable prevalence of
up-regulated genes, suggesting activation or up-regulation of those
biological functions or processes.

Values near -1.73 indicate other GO terms have a moderate dominance of
down-regulated genes, implying suppression or down-regulation.

Typically, a z-score greater than 2 or less than -2 is considered
significant evidence of activation or inhibition, respectively, though
borderline values like yours still provide useful directional hints.

This z-score metric provides a simple, normalized way to capture the
overall regulatory direction of each GO term based on the gene
expression data from your DESeq2 contrast, giving you an intuitive
measure to prioritize and visualize the biological processes likely
impacted in your dataset.

Thus, in your chord diagram or other visualizations, z-score can be used
to color or weight GO terms and gene links to quickly highlight
activated vs. repressed functions.

*Generated via perplexity AI on 15AUG2025*
:::
:::

```{r}
# Select and arrange columns to match GOplot 'circ' format
circ_dfp <- circ_dfp %>%
  select(!term) %>% # remove term colunm
  mutate(term = parentTerm) %>% # replace it with the parent term column
  select(category,
         ID,
         term,
         count,
         genes,
         logFC,
         adj_pval,
         zscore) %>% 
  as.data.frame()
```

### Create binary matrix using `chord_dat()`

::: callout-caution
`chord_dat()` expects the `genes =` to have genes in a column named `ID`
with a `logFC` column The matrix should be: genes as rows, terms as
columns (binary matrix), and one column named "logFC".
:::

```{r}
circ_genes <- circ_dfp %>% 
  select(genes, logFC) %>% 
  rename(ID = genes) %>% 
  distinct()
```

```{r binary-matrix-chord_dat}
chord <- chord_dat(
  data = circ_dfp,
  genes = circ_genes,
  process = c("regulation of RNA metabolic process", 
              "positive regulation of biological process", 
              "cellular metabolic process", 
              "response to stress", 
              "response to chemical",
              "cell differentiation",
              "regulation of signaling",
              "organonitrogen compound metabolic process"))
```

### Plot chord diagram

```{r}
library(viridis)
library(RColorBrewer)
```

```{r}
display.brewer.all()
```

viridis magma plasma inferno cividis mako rocket turbo

```{r}
library(scales)
show_col(turbo(5))
```

```{r}
ribbons <- turbo(5)
ribbons_viridis <- viridis(5)
ribbons_pastel <- brewer.pal(5, "Pastel2")
```

```{r plot-chord, fig.height=10, fig.width=9}
chordplot <- GOChord(chord, 
                     gene.order = "logFC",
                     border.size = 0.2,
                     gene.size = 3,
                     title = "Enriched GO terms across significantly different interaction genes",
                     process.label = 10,
                     space = 0.02,
                     gene.space = 0.25)
chordplot
```

```{r}
ggsave("../../output/07_go/chord_plot.png", width = 12, height = 12, dpi = 600 )
```

> Therefore, if you want to include more than one logFC value per gene.
> To adjust to this situation you should use the nlfc argument. It is a
> numeric value and it defines the number of logFC columns within your
> binary membership matrix. The default is '1' assuming that most of the
> time you just have one contrast and one logFC value per gene.

## GOHeat heatmap

```{r}
GOHeat(chord, nlfc = 1, fill.col = c('red', 'white', 'blue'))
```

# Circos plot
