---
title: "Gene ontology (GO) enrichment plots"
subtitle: "We've identified the DEGs... let's visualize what they do!"
author: "Sarah Tanja"
date: 2025-09-30
date-modified: today
format:
  gfm: default 
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

# Background
## Inputs
## Outputs
# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.align = "center",
  fig.height=4, 
  fig.width=6,
  dpi = 600,
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```
## Install packages
```{r}
## CRAN packages-----------
#install.packages("GOplot")
#remotes::install_github("mdscheuerell/flexoki")

## Bioconductor packages---------
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("rrvgo")
```

## Load packages
```{r}
library(rrvgo)
library(GOplot)
library(flexoki)
library(plotly)
library(tidyverse)
```

## Load inputs
```{r}
## Reduced Terms
### Interaction
reducedTerms.int <- read_csv("../../output/08_go/reducedTerms_int.csv")

### Interaction patterns
reducedTerms.1 <- write_csv("../../output/08_go/reducedTerms_1.csv")
reducedTerms.2 <- write_csv("../../output/08_go/reducedTerms_2.csv")
reducedTerms.3 <- write_csv("../../output/08_go/reducedTerms_3.csv")

### Contrasts
reducedTerms.lvccl <- write_csv("../../output/08_go/reducedTerms_lvccl.csv")
reducedTerms.mvccl <- write_csv("../../output/08_go/reducedTerms_mvccl.csv")
reducedTerms.hvccl <- write_csv("../../output/08_go/reducedTerms_hvccl.csv")
reducedTerms.hvceg <- write_csv("../../output/08_go/reducedTerms_hvceg.csv")

## Similarity matrices
### Interaction
simMatrix.int <- read_csv("../../output/08_go/simMatrix_int.csv")

### Interaction patterns
simMatrix.1 <- read_csv("../../output/08_go/simMatrix_1.csv")
simMatrix.2 <- read_csv("../../output/08_go/simMatrix_2.csv")
simMatrix.3 <- read_csv("../../output/08_go/simMatrix_3.csv")

### Contrasts
simMatrix.lvccl <- read_csv("../../output/08_go/simMatrix_lvccl.csv")
simMatrix.mvccl <- read_csv("../../output/08_go/simMatrix_mvccl.csv")
simMatrix.hvccl <- read_csv("../../output/08_go/simMatrix_hvccl.csv")
simMatrix.hvceg <- read_csv("../../output/08_go/simMatrix_hvceg.csv")
```


# Plotting & interpretation

[Using the rrvgo
package](https://bioconductor.org/packages/devel/bioc/vignettes/rrvgo/inst/doc/rrvgo.html)

## Similarity matrix heatmap

Plot similarity matrix as a heatmap, with clustering of columns of rows
turned on by default (thus arranging together similar terms).

```{r}
heatmapPlot(simMatrix.int,
            reducedTerms.int,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

```{r}
heatmapPlot(simMatrix.1,
            reducedTerms.1,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

```{r}
heatmapPlot(simMatrix.2,
            reducedTerms.2,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

```{r}
heatmapPlot(simMatrix.3,
            reducedTerms.3,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```


## Scatter plot

Plot GO terms as scattered points. Distances between points represent
the similarity between terms, and axes are the first 2 components of
applying a PCoA to the (di)similarity matrix. Size of the point
represents the provided scores or, in its absence, the number of genes
the GO term contains.

```{r}
scatterPlot(simMatrix.int, reducedTerms.int,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

```{r}
scatterPlot(simMatrix.1, reducedTerms.1,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

```{r}
scatterPlot(simMatrix.2, reducedTerms.2,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

```{r}
scatterPlot(simMatrix.3, reducedTerms.3,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```


### Cluster summaries
I want a dataframe that groups my GO terms based on these 5 main
clusters displayed in this PCA. The scatterPlot uses a dimensionality
reduction (PCoA) approach to project GO terms into 2D space and then
visually groups the points into a smaller number of major clusters.
These visual clusters are typically broader groupings, which correspond
to higher-level semantic similarity rather than the narrow clusters from
the cut tree. A possible approach if you want to create a cluster
assignment dataframe matching the 5 visual clusters is to: - Extract the
2D coordinates from the scatterPlot dimensionality reduction output, -
Apply a clustering method on the 2D coordinates (e.g., k-means with
k=5), - Create a new dataframe linking each GO term to its visual
cluster. This approach aligns the clustering with the displayed
scatterPlot clusters rather than the original fine-grained clusters.

```{r}
# Distance from similarity matrix
dist <- as.dist(1 - simMatrix.all)

# PCoA to 2 dimensions
coord <- cmdscale(dist, k = 2)

# Convert to dataframe with GO terms
df_coords <- data.frame(go = rownames(simMatrix.all), Dim1 = coord[,1], Dim2 = coord[,2])

```

Cluster the terms visually in 2D space Use k-means clustering on the 2D
coordinates with k = 5 (number of desired clusters).

```{r}
set.seed(123)  # for reproducibility
clusters <- kmeans(df_coords[, c("Dim1", "Dim2")], centers = 5)
df_coords$cluster <- as.factor(clusters$cluster)
```

Combine cluster info with other GO term data Merge this dataframe with
your reducedTerms or other GO annotation information as needed.

```{r}
final_df <- merge(df_coords, reducedTerms.all, by.x = "go", by.y = "go")
```

This dataframe `final_df` will now include each GO term, its 2D
coordinates used for visualization, and its assigned cluster
corresponding to the 5 major groups in the scatterPlot.

```{r}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, color = cluster.x, size = score, alpha = 0.4)) +
  geom_point() +
  coord_equal() +
  labs(color = "Cluster (k=5)", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)
```

```{r, fig.height=10, fig.width=15}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, color = parentTerm, size = score)) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)
```

```{r}
p <- ggplot(final_df, aes(
    Dim1, Dim2,
    color = parentTerm,
    size  = score,
    text  = paste0(
      "GO: ", go,
      "<br>Parent Term: ", parentTerm,
      "<br>Term: ", term,
      "<br>Score: ", round(score, 2)
    )
  )) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term",
       size  = "Score",
       title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)

ggplotly(p, tooltip = "text")
```

::: column-margin
::: callout-note
How to interpret distances

Closer points = more semantically similar GO terms. That means the gene
sets they represent overlap more or describe conceptually related
processes.

Farther points = less similar biologically. Their gene sets don't
overlap much, and the ontology sees them as describing different
processes. The PCoA projection does not use your logFCs or sample-level
values --- it uses GO term relationships (semantic similarity).

1.  What your clusters represent

The clusters from the PCoA + k-means are based on the pairwise semantic
similarity scores across all GO terms.

These are data-driven groupings that capture overall gene overlap +
ontology proximity.

The algorithm doesn't "know" about your chosen parentTerm column --- it
just groups terms that have similar meaning.

2.  What your parentTerm variable is

The "parent term" you merged in is a single hierarchical label from the
GO DAG (graph).

A GO term can belong under multiple higher-level categories in the
ontology, but in your dataset it's probably annotated with just one
chosen parent (often the most specific or most frequent).

Because the GO graph is not a strict tree, related terms may have
different parents --- so they don't collapse neatly into one parentTerm
cluster.

3.  Why they don't line up

Ontology is many-to-many: two GO terms might be semantically very close
(high overlap) but come from different parents in the hierarchy.

Cluster ≠ hierarchy: k-means clusters cut across the graph structure;
they group by similarity, not by the official ontology tree.

Parent terms are broad: "cellular process" or "metabolic process" can
cover huge swathes of biology, so those categories span multiple
similarity clusters.

4.  How to check alignment

Look at your clusters and check: are terms within a cluster biologically
coherent even if they span different parents? Often they are --- they
share overlapping gene sets or related functions.

The clusters are "data-driven neighborhoods" of related GO terms. The
parent terms are "administrative labels" from the GO DAG. They don't
always coincide because biology (and the GO ontology) is more tangled
than a simple tree. Clusters in the 2D space: show neighborhoods of
semantically similar terms (terms that describe similar biology and
often share many genes).

Point size (score): tells you which terms were more strongly enriched.

Color by parentTerm: just tags each dot with the one "parent" you merged
in. The fact that colors are mixed within a cluster shows that ontology
categories don't always line up with semantic neighborhoods.
:::
:::

```{r}
# Summarize cluster content by parent term
cluster_summary <- final_df %>%
  group_by(cluster.x, parentTerm) %>%
  summarise(
    n_terms = n(),                     # how many GO terms in this parent within cluster
    mean_score = mean(score, na.rm = TRUE),  # average enrichment strength
    max_score = max(score, na.rm = TRUE),    # strongest term in that parent
    .groups = "drop"
  ) %>%
  arrange(cluster.x, desc(n_terms))

# For each cluster, show top 5 parent terms by count (or max score)
top_terms_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = n_terms, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(n_terms))

top_terms_by_cluster
```

## Treemap plot

Treemaps are space-filling visualization of hierarchical structures. The
terms are grouped (colored) based on their parent, and the space used by
the term is proportional to the score. Treemaps can help with the
interpretation of the summarized results and also comparing different
sets of GO terms.

```{r}
treemapPlot(reducedTerms.int)
```

```{r}
treemapPlot(reducedTerms.1)
```

```{r}
treemapPlot(reducedTerms.2)
```

```{r}
treemapPlot(reducedTerms.3)
```

```{r}
treemapPlot(reducedTerms.lvccl)
```

```{r}
treemapPlot(reducedTerms.mvccl)
```

```{r}
treemapPlot(reducedTerms.hvccl)
```

```{r}
treemapPlot(reducedTerms.hvceg)
```

## Word cloud

Word clouds are visualizations which reproduce a text putting emphasis
to words which appear frequently in a text. They can help to identify
processes and functions that happen more commonly in a set of enriched
GO terms, as well as comparing between different sets.

```{r}
wordcloudPlot(reducedTerms.int, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.1, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.2, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.3, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.lvccl, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.mvccl, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.hvccl, min.freq=1, colors="black")
```
```{r}
wordcloudPlot(reducedTerms.hvceg, min.freq=1, colors="black")
```

## Chord diagram

Using
[GOplot](https://cran.r-project.org/web/packages/GOplot/vignettes/GOplot_vignette.html)

### Create `circ` object using `circle_dat()`

> `circle_dat()` takes two data frames as an input. The first one
> contains the results of the functional analysis and should have at
> least four columns (category, term, genes, adjusted p-value).
> Additionally, a data frame of the selected genes and their logFC is
> needed.

Data object for plotting needs two data frame inputs:

1.  **result** of functional analysis with
    1.  category: GO term annotation type (e.g., "BP", "MF", "CC")
    2.  ID: GO term identifier (e.g., "GO:0008150")
    3.  term: GO term description
    4.  adj_pval: the `bh_adjust` column from the GO enrichment analysis

> The significance of a term is indicated by the adjusted p-value
> (adj_pval). Terms with an adjusted p-value \< 0.05 are considered as
> **significantly enriched** and are more likely to provide reliable
> information.

`GO.BP.all` is the result of our functional analysis...

```{r}
str(GO.BP.all)
```

```{r}
results_circ <- GO.BP.all %>% 
  rename(ID = category,
         category = ontology,
         adj_pval = bh_adjust,
         count = numInCat) %>% 
  select(category, ID, term, adj_pval, count)
```

And even though the vignette doesn't say you need it, you need a column
of genes for each go term.

```{r}
terms_circ <- GO.terms.all %>% 
  rename(ID = GOs,
         genes = gene_id)

str(terms_circ)
unique(terms_circ$genes)
```

So we have to join results_circ to terms_circ

```{r}
go_circ <- results_circ %>%  
  full_join(terms_circ, by = "ID") %>% 
  select(category, ID, term, genes, adj_pval, count)

str(go_circ)
unique(go_circ$genes)
```

2.  list of selected **genes** and log2FC from a DESeq2 result
    1.  ID: (Your gene ID (use your long gene IDs directly))
    2.  log2FC: from one of the DESeq2 contrasts

::: callout-note
The Log2FC will be displayed on the chord diagram, and you can add in
additional values from other contrasts after the initial circ object is
made. For the time being we're going to select the early gastrula low
contrast because it had the most DEGs
:::

```{r}
genes_circ <- sig_genes_anno %>% 
  select(ID = gene_id, 
         logFC = earlygastrula_low_log2FC)

str(genes_circ)
```

The circle_dat() function will generate a count and a zscore column -
count: Number of genes per GO term - zscore: GO term z-score, typically
calculated as (number of upregulated genes minus downregulated
genes)/sqrt(count) for that term

::: callout-caution
The `circle_dat()` function expects your gene ids to be named `genes` in
your GO results dataframe and `ID` in your gene list dataframe... this
is tricky, watch out!
:::

The following code checks that the gene id column names are the same in
the two dataframes with differently named columns

```{r}
length(intersect(go_circ$genes, genes_circ$ID))
```

```{r}
circ <- circle_dat(go_circ, genes_circ)
```

`circ` is the object that will now be used for plotting!

```{r}
# circ from circle_dat()
str(circ)
```

::: callout-caution
`circle_dat()` reformatted the gene ids to be all uppercase
:::

```{r}
summary(circ$count)
```

::: callout-caution
Did not appear to calculate `count` correctly... it isn't counting the
genes per term but rather the number of times a term is assigned to a
gene... which is always 1, This also throws off the zscore calculation
which depends on count.
:::

### JK Manually create `circ_df`

```{r}
circ_df <- go_circ %>%  
  full_join(genes_circ, by = c("genes" = "ID")) %>% 
  drop_na(term)

circ_df_parent <- circ_df %>% 
  left_join(reducedTerms.all, by = c("ID" = "go", "term" = "term"))
```

Add zscore column

```{r}
circ_dfp <- circ_df_parent %>%
  group_by(ID) %>%
  mutate(zscore = (sum(logFC > 0) - sum(logFC < 0)) / sqrt(count)) %>%
  ungroup()
```

::: column-margin
::: callout-note
### What does the zscore tell us?

The z-score in your gene ontology (GO) analysis context is a statistical
measure that helps indicate whether a GO term (biological process,
molecular function, or cellular component) is more likely to be
biologically increased or decreased based on the direction of gene
expression changes.

Specifically, your z-score formula $$
zscore = \frac{(up - down)}{\sqrt(count)}
$$ means:

-   "up" is the number of genes assigned to that GO term that are
    up-regulated (logFC \> 0).

-   "down" is the number of genes assigned to that GO term that are
    down-regulated (logFC \< 0).

-   "count" is the total number of genes assigned to that GO term.

Interpretation of z-scores:

A positive z-score indicates that the GO term has more up-regulated
genes than down-regulated genes relative to its size and is therefore
more likely to be biologically increased or activated.

A negative z-score indicates more down-regulated genes and suggests that
the GO term is more likely to be biologically decreased or inhibited.

The magnitude of the z-score tells you how strong this trend is,
adjusted for the total number of genes in that GO term.

In your data, z-scores ranging approximately from +2 to -1.73 mean:

Values toward +2 imply that some GO terms show a notable prevalence of
up-regulated genes, suggesting activation or up-regulation of those
biological functions or processes.

Values near -1.73 indicate other GO terms have a moderate dominance of
down-regulated genes, implying suppression or down-regulation.

Typically, a z-score greater than 2 or less than -2 is considered
significant evidence of activation or inhibition, respectively, though
borderline values like yours still provide useful directional hints.

This z-score metric provides a simple, normalized way to capture the
overall regulatory direction of each GO term based on the gene
expression data from your DESeq2 contrast, giving you an intuitive
measure to prioritize and visualize the biological processes likely
impacted in your dataset.

Thus, in your chord diagram or other visualizations, z-score can be used
to color or weight GO terms and gene links to quickly highlight
activated vs. repressed functions.

*Generated via perplexity AI on 15AUG2025*
:::
:::

```{r}
# Select and arrange columns to match GOplot 'circ' format
circ_dfp <- circ_dfp %>%
  select(!term) %>% # remove term colunm
  mutate(term = parentTerm) %>% # replace it with the parent term column
  select(category,
         ID,
         term,
         count,
         genes,
         logFC,
         adj_pval,
         zscore) %>% 
  as.data.frame()
```

### Create binary matrix using `chord_dat()`

::: callout-caution
`chord_dat()` expects the `genes =` to have genes in a column named `ID`
with a `logFC` column The matrix should be: genes as rows, terms as
columns (binary matrix), and one column named "logFC".
:::

```{r}
circ_genes <- circ_dfp %>% 
  select(genes, logFC) %>% 
  rename(ID = genes) %>% 
  distinct()
```

```{r binary-matrix-chord_dat}
chord <- chord_dat(
  data = circ_dfp,
  genes = circ_genes,
  process = c("regulation of RNA metabolic process", 
              "positive regulation of biological process", 
              "cellular metabolic process", 
              "response to stress", 
              "response to chemical",
              "cell differentiation",
              "regulation of signaling",
              "organonitrogen compound metabolic process"))
```

### Plot chord diagram

```{r}
library(viridis)
library(RColorBrewer)
```

```{r}
display.brewer.all()
```

viridis magma plasma inferno cividis mako rocket turbo

```{r}
library(scales)
show_col(turbo(5))
```

```{r}
ribbons <- turbo(5)
ribbons_viridis <- viridis(5)
ribbons_pastel <- brewer.pal(5, "Pastel2")
```

```{r plot-chord, fig.height=10, fig.width=9}
chordplot <- GOChord(chord, 
                     gene.order = "logFC",
                     border.size = 0.2,
                     gene.size = 3,
                     title = "Enriched GO terms across significantly different interaction genes",
                     process.label = 10,
                     space = 0.02,
                     gene.space = 0.25)
chordplot
```

```{r}
ggsave("../../output/07_go/chord_plot.png", width = 12, height = 12, dpi = 600 )
```

> Therefore, if you want to include more than one logFC value per gene.
> To adjust to this situation you should use the nlfc argument. It is a
> numeric value and it defines the number of logFC columns within your
> binary membership matrix. The default is '1' assuming that most of the
> time you just have one contrast and one logFC value per gene.

# GOHeat

```{r}
GOHeat(chord, nlfc = 1, fill.col = c('red', 'white', 'blue'))
```

# Summary & next steps

Inputs: - List of 44 genes significant in the leachate × stage
interaction test. - Montipora capitata genome FASTA and GFF3
annotation. - EggNog and KEGG functional annotation files.

Setup: - Installed and loaded R packages (goseq, rrvgo, GOplot, flexoki,
etc.). - Imported gene count data, annotation tables, and genome
sequences. - Extracted FASTA sequences of the 44 significant genes.

Annotation: - Mapped 37 of 44 significant genes to EggNog/KEGG
functional annotations. - Linked each gene to transcript length
information from the GFF3 file. - Formatted GO term columns for
downstream use.

Enrichment Analysis: - Used goseq to perform GO enrichment (biological
process level) while correcting for gene length bias. - Significance was
assessed with Wallenius' distribution and Benjamini--Hochberg FDR
correction (padj \< 0.05). - Identified enriched biological process
terms.

Redundancy Reduction & Visualization: - Applied rrvgo to cluster
semantically similar GO terms and reduce redundancy.

Produced visualizations: - Heatmap of GO term similarity. - Scatter plot
grouping GO terms into 5 major clusters. - Treemap, word cloud, and
chord diagrams to summarize enriched functions. - Calculated z-scores
per GO term to indicate whether functions were enriched for upregulated
or downregulated genes.

Outputs:

-   Tables of enriched GO terms, reduced clusters, semantic similarity
    matrices, and gene-to-term assignments.
-   Multiple graphical summaries (heatmap, PCA scatter, treemap, word
    cloud, chord plot)
    
```{r}
sessionInfo()
```

