---
title: "Upset Plot: Overlap of Significant DEGs"
subtitle: "Visualizing overlap between LRT, interaction, and main leachate effects"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output))        
```

# Background

This analysis creates an upset plot to visualize the overlap of significant genes identified across different DESeq2 analyses:

1. **LRT (Likelihood Ratio Test)**: Tests for any genes with differential expression across the full model
2. **Interaction Effects**: Genes showing significant interaction between developmental stage and leachate concentration
3. **Main Leachate Effects**: Genes showing significant main effect of leachate concentration

# Setup

## Install packages

```{r install, eval=FALSE}
# Install UpSetR if not already installed
if (!require("UpSetR", quietly = TRUE)) {
  install.packages("UpSetR")
}

if (!require("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
```

## Load packages

```{r load}
library(tidyverse)
library(UpSetR)
```

# Load Data

## Load significant gene lists

```{r load-data}
# LRT significant genes (genes with any DE across full model)
lrt_genes <- read_csv("../../output/06_deg/interactions/LRT_sig_genes.csv") %>%
  pull(gene_id)

# Interaction effect genes
interaction_genes <- read_csv("../../output/06_deg/interactions/sig_interaction_genes.csv") %>%
  pull(x)

# Main leachate effect genes
leachate_genes <- read_csv("../../output/09_deg_multifactorial/sig_leachate.csv") %>%
  pull(gene)

# Print counts
cat("Number of LRT significant genes:", length(lrt_genes), "\n")
cat("Number of interaction effect genes:", length(interaction_genes), "\n")
cat("Number of main leachate effect genes:", length(leachate_genes), "\n")
```

# Prepare Data for Upset Plot

```{r prepare-upset-data}
# Create a list for UpSetR
gene_list <- list(
  LRT = lrt_genes,
  Interaction = interaction_genes,
  Leachate = leachate_genes
)

# Summary of set sizes
cat("Set sizes:\n")
cat("LRT:", length(lrt_genes), "\n")
cat("Interaction:", length(interaction_genes), "\n")
cat("Leachate:", length(leachate_genes), "\n")

# Get all unique genes
all_genes <- unique(c(lrt_genes, interaction_genes, leachate_genes))
cat("Total unique genes:", length(all_genes), "\n")
```

# Create Upset Plot

```{r upset-plot, fig.width=10, fig.height=6}
# Create the upset plot using UpSetR
upset_plot <- upset(
  fromList(gene_list),
  nsets = 3,
  order.by = "freq",
  sets = c("Leachate", "Interaction", "LRT"),
  keep.order = TRUE,
  point.size = 3.5,
  line.size = 1.5,
  mainbar.y.label = "Intersection Size",
  sets.x.label = "Set Size",
  text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
  mb.ratio = c(0.6, 0.4)
)

print(upset_plot)
```

# Save Plot

```{r save-plot}
# Save as PNG
png(
  filename = "../../output/09_deg_multifactorial/upset_plot_deg_overlap.png",
  width = 10,
  height = 6,
  units = "in",
  res = 300,
  bg = "white"
)

upset(
  fromList(gene_list),
  nsets = 3,
  order.by = "freq",
  sets = c("Leachate", "Interaction", "LRT"),
  keep.order = TRUE,
  point.size = 3.5,
  line.size = 1.5,
  mainbar.y.label = "Intersection Size",
  sets.x.label = "Set Size",
  text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
  mb.ratio = c(0.6, 0.4)
)

dev.off()

cat("Upset plot saved to: output/09_deg_multifactorial/upset_plot_deg_overlap.png\n")
```

# Summary

```{r summary}
# Calculate overlaps using fromList
upset_data_matrix <- fromList(gene_list)

# Calculate set operations
lrt_only <- sum(upset_data_matrix$LRT == 1 & upset_data_matrix$Interaction == 0 & upset_data_matrix$Leachate == 0)
interaction_only <- sum(upset_data_matrix$LRT == 0 & upset_data_matrix$Interaction == 1 & upset_data_matrix$Leachate == 0)
leachate_only <- sum(upset_data_matrix$LRT == 0 & upset_data_matrix$Interaction == 0 & upset_data_matrix$Leachate == 1)
lrt_interaction <- sum(upset_data_matrix$LRT == 1 & upset_data_matrix$Interaction == 1 & upset_data_matrix$Leachate == 0)
lrt_leachate <- sum(upset_data_matrix$LRT == 1 & upset_data_matrix$Interaction == 0 & upset_data_matrix$Leachate == 1)
interaction_leachate <- sum(upset_data_matrix$LRT == 0 & upset_data_matrix$Interaction == 1 & upset_data_matrix$Leachate == 1)
all_three <- sum(upset_data_matrix$LRT == 1 & upset_data_matrix$Interaction == 1 & upset_data_matrix$Leachate == 1)

cat("Overlap summary:\n")
cat("LRT only:", lrt_only, "\n")
cat("Interaction only:", interaction_only, "\n")
cat("Leachate only:", leachate_only, "\n")
cat("LRT & Interaction:", lrt_interaction, "\n")
cat("LRT & Leachate:", lrt_leachate, "\n")
cat("Interaction & Leachate:", interaction_leachate, "\n")
cat("All three:", all_three, "\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
