---
title: "Multifactorial Differential Gene Expression"
subtitle: "using DESeq2 contrasts"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

Here we are looking at a multifactorial differential gene expression analysis for developing coral embryos at 3 developmental stages (4 hours post fertilization , 9 hours post fertilization, and 14 hours post fertiliation). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of microplastic pollution on coral reefs.

## Inputs

-   `gcm_filt.csv` is the unfiltered gene count matrix for 61 samples (two outlier samples were already removed).
-   `metadata_filt.csv` is the metadata for 61 samples

## Outputs

-   `volcano_facet.png`

# Setup

## Install packages

```{r}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') 
  install.packages('tidyverse')
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('DESeq2')
if ("ashr" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('ashr')
if ("plotly" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('plotly')
if ("cowplot" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('cowplot')
if ("pak" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('pak')
```

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("EnhancedVolcano")
#pak::pak("mdscheuerell/flexoki")
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(tidyverse)
library(EnhancedVolcano)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
```

## Import gene counts

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r}
gcm_filt <- read.csv("../output/06_tidyup/gcm_filt.csv", row.names="gene_id", check.names = FALSE)
head(gcm_filt)
```

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filt` is the gene count matrix that has been filtered to remove 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA
:::

## Import metadata

```{r}
metadata_filt <- read_csv("../metadata/metadata_filt.csv")
str(metadata_filt)
```
Format metadata
```{r}
# Reorder factors
metadata_filt$embryonic_stage <- as.factor(metadata_filt$embryonic_stage)
metadata_filt$embryonic_stage <- fct_relevel(metadata_filt$embryonic_stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_filt$leachate <- as.factor(metadata_filt$leachate)
metadata_filt$leachate <- fct_relevel(metadata_filt$leachate, "control", "low", "mid", "high")

metadata_filt$hpf <- as.factor(metadata_filt$hpf)
metadata_filt$hpf <- fct_relevel(metadata_filt$hpf, "4", "9", "14")

metadata_filt$parents <- as.factor(metadata_filt$parents)
str(metadata_filt)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples where explanatory variables `hpf` is a factor with 3 levels (4, 9, 14) and `leachate` is a continuous numerical variable representing leachate concentrations with values 0 mg/L, 0.01 mg/L, 0.1 mg/L, and 1 mg/L
:::

## Explore `parent colony` batches
```{r}
table(metadata_filt$parents, metadata_filt$leachate)
```
```{r}
chisq.test(table(metadata_filt$parents, metadata_filt$leachate))
```
::: callout-note
Pearson's Chi-squared test result was insignificant `(p = 0.8237)`, meaning that the parent colony designation is not confounded with leachate exposure level
:::

What your sample table shows
You have several distinct parent pools (e.g. 101112, 131415, 789, etc.) and some are well-represented across all leachate conditions, while others are very sparse (e.g., only 1 sample in 1 condition).
- 101112, 131415, 123, 456	✅ Yes	Represented in most/all leachate levels
- 13, 45, 6, 67, 7, 89	⚠️ Sparse	Limited to 1–3 samples in few conditions
- 13, 6, 67, 7	❌ Bad	Only 1 sample total → can't estimate reliably

```{r}
# Example: drop very sparse groups
keep_parents <- c("101112", "123", "131415", "456", "789")
dds <- dds[, dds$parents %in% keep_parents]
dds$parents <- droplevels(dds$parents)

design(dds) <- ~ parents + hpf + leachate

```

When to use RUVSeq?
- You suspect variation from unmodeled effects (e.g., parents)
- You can’t include those effects directly in the model due to **sparsity** or confounding
- You want to retain all samples

```{r}
chisq.test(table(metadata_filt$parents, metadata_filt$hpf))
```
::: callout-note
Pearson's Chi-squared test result was insignificant `(p = 0.8869)`, meaning that the parent colony designation is not confounded with hpf 
:::

# Filter

```{r}
# Filter rows where at least 75% of the columns have a raw count of ~50
gcm_filt <- gcm_filt %>%
  filter(rowSums(across(everything(), ~ . > 50)) >= 0.75 * ncol(gcm_filt))

nrow(gcm_filt)
```

::: callout-note
This is an aggressive filter!
:::

# Parents as batch effect?
```{r}
dds_p <- DESeqDataSetFromMatrix(countData = gcm_filt, 
                                colData = metadata_filt,
                                design = formula(~ 1 + parents + hpf + leachate))
```

Here we run a likelihood ratio test to determine if `parents` as a batch effect variable influences gene expression. 
```{r}
dds_p <- DESeq(dds_p, test = "LRT", reduced = ~ hpf + leachate)
```

```{r}
summary(results(dds_p))
```
::: callout-note
Yup! It really does... and this isn't that surprising given that the samples appeared to group by `parents` in the PCAs, dendrograms, and heatmaps. Furthermore, I broadly pooled the crosses by date (or night of spawning). Parental crosses 1-5 were coral colonies that spawned on July 5th, 6-10 spawned on July 6th, and 11-15 spawned on July 7th. Most pools are confounded with spawn night (123 = July 5th, 789 = July 6th, 131415 = July 7th), but not all are pooled with crosses of the same spawn night (ex. 456, 101112)... but still reflect the differences in gene expression due to night of spawning! An interesting aside to this experiment is that night of spawning has an effect on embryonic gene expression. Dr. Kevin Marquez Johnson indicated he wasn't surprised by this either, and indicated that as gametes mature across days of spawning gene expression is highly likely to be differential. So, we need to include `parents` in the model to account for this, and it's effecively a 'batch effect'. 
:::

But wait — is parents nested?
Before deciding, consider:
- If each parent group appears only once per hpf × leachate combination → bad idea to include it: that's near-singular.
- If parents are replicated across groups, and you're curious whether there's inter-individual variation → good to test.

# Make DESeq2 objects

```{r}

dds <- DESeqDataSetFromMatrix(countData = gcm_filt,
                              colData = metadata_filt,
                              design = formula( ~ 1 + parents + hpf + leachate))
```



## Explore PCAs
```{r}
plotPCA(vst(dds), intgroup = "hpf")

plotPCA(vst(dds), intgroup = "leachate")
```
::: callout-note
Data groups largely by embryonic stage (hours post fertilization = hpf)!
:::

```{r}
# Get PCA data (top 500 DEGs by default)
pca_500 <- plotPCA(vst(dds), intgroup = "group", returnData = TRUE)
percentVar_500 <- round(100*attr(pca_500, "percentVar"))

# Merge with metadata
pca_500 <- merge(pca_500, metadata_filt, by.x = "name", by.y = "sample_name", all.x = TRUE)


# Assign specific colors to each hpf
hpf_colors <- c(
  "4" = flex("cyan", sat=500),
  "9" = flex("green", sat =500),
  "14" = flex("yellow", sat=500)
)

# Plot PCAs
PCA_500 <- ggplot(pca_500, aes(PC1, PC2, color=hpf, shape=leachate)) + 
  geom_point(size=2, alpha = 0.8) +
  ggtitle("M.cap embryos exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ",percentVar_500[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar_500[2],"% variance")) + 
  coord_fixed() +
  scale_color_manual(values=hpf_colors)+
  stat_ellipse()+
  theme_minimal()

PCA_500
```

```{r}
pca_500$label <- paste0("Sample: ", pca_500$sample_name,
                        "\nHPF: ", pca_500$hpf,
                        "\nLeachate: ", pca_500$leachate)

# Indicate leachate level by saturation (alpha)
leachate_alpha_map <- c("control" = 0.3, "low" = 0.5, "mid" = 0.7, "high" = 1)
pca_500$alpha_val <- leachate_alpha_map[pca_500$leachate]

# Original ggplot with a tooltip aesthetic
PCA_500 <- ggplot(pca_500, aes(PC1, PC2, color = hpf, label = name, alpha = alpha_val)) + 
  geom_point(shape = 21, size = 2) +
  ggtitle("M.cap embryos exposed to pvc leachate, top 500 most variable genes") +
  xlab(paste0("PC1: ", percentVar_500[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_500[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_500, tooltip = "label")
```

## Plot distribution

```{r}

# Extract VST-transformed expression matrix
vsd_mat <- assay(vst(dds))

# Flatten it into a vector
vsd_vals <- as.vector(vsd_mat)

# Create a dataframe for ggplot
df_vsd <- data.frame(vsd_count = vsd_vals)

# Plot
ggplot(df_vsd, aes(x = vsd_count)) +
  geom_histogram(fill = "grey") +
  theme_classic() +
  labs(
    title = "Histogram of Variance-Stabilized Counts",
    x = "VST-transformed read count",
    y = "Frequency"
  )

```

VST compresses the high-end of the count distribution and expands the low-end, making the variance roughly constant across the range. The resulting values are on an approximate log2 scale but not exact log2(count + 1) --- VST is more nuanced, especially for low counts. A range of 5 to 15 corresponds roughly to raw counts ranging from 2³² to 2¹⁵ ≈ 32 to 32,000, so it's realistic for RNA-seq gene expression data.

# Run DESeq Tests

## LRT
The likelihood ratio test
```{r}
dds_lrt <- DESeq(dds, test = "LRT", reduced = ~ parents + hpf)
```
```{r}
summary(results(dds_lrt))
```

::: Callout-note
This compares the full model `~ parents + hpf + leachate` to the reduced model `~ hpf` and genes with significant p-values can be interpreted as ones that are responding to `leachate`, regardless of direction (up or down) or level (low, mid, high). Another way to view it is what does the `+ leachate` do that isn't explained by `hpf` alone? 

This design is to detect any effect of `leachate` — even if small or complex — while controlling for `hpf`.

In the results summary, 12 genes show significant evidence of being affected by `leachate` (FDR < 0.1), according to the LRT model.

Half the genes (~5,111) were filtered out due to low mean counts.

Many of our p-values are close to zero, but LFCs are small — which can happen when there are consistent but subtle differences with low variance.
These 12 genes are significant across all leachate levels and across all developmental stages! 
:::

### Plot dispersion
```{r}
plotDispEsts(dds_lrt)
```
x-axis:
Mean of normalized counts for each gene (on a log10 scale)

y-axis:
Estimated dispersion (how variable gene expression is across samples), also on a log scale

## Wald
In order to find out which genes are significant for each `leachate` level we have to use the Wald test. 
```{r}
dds_wald <- DESeq(dds, test = "Wald")
```


### Plot dispersion

```{r}
plotDispEsts(dds_wald)
```

::: callout-note
Dispersion trend line is hovering around 0.1, and bulk spread of tag-wise gene estimates is from 0.02 to 0.5
:::

# Understanding coefficients

This is a likelihood ratio test, not a contrast — it tells you:
Does leachate (as a whole factor) significantly affect gene expression?

```{r}
resultsNames(dds)  # see what contrast names are available
```

| Coef | Name                         | Meaning                                                               |
|------|------------------------------|-----------------------------------------------------------------------|
| 1    | `"Intercept"`                | Baseline expression: `hpf = 4` and `leachate = control`               |
| 2    | `"hpf_9_vs_4"`               | Change from 4 hpf → 9 hpf, at **control** exposure                    |
| 3    | `"hpf_14_vs_4"`              | Change from 4 hpf → 14 hpf, at **control** exposure                   |
| 4    | `"leachate_low_vs_control"`  | Change from control → low exposure, at **4 hpf**                      |
| 5    | `"leachate_mid_vs_control"`  | Change from control → mid exposure, at **4 hpf**                      |
| 6    | `"leachate_high_vs_control"` | Change from control → high exposure, at **4 hpf**                     |


| `resultsName()`              | Interpretation                                                        |
| ---------------------------- | --------------------------------------------------------------------- |
| `"Intercept"`                | Baseline expression: samples with `hpf = 4` and `leachate = control`  |
| `"hpf_9_vs_4"`               | Main effect of `hpf = 9` vs `hpf = 4` (averaged over leachate levels) |
| `"hpf_14_vs_4"`              | Main effect of `hpf = 14` vs `hpf = 4`                                |
| `"leachate_low_vs_control"`  | Main effect of `low` vs `control`, averaged across all `hpf` levels   |
| `"leachate_mid_vs_control"`  | Main effect of `mid` vs `control`                                     |
| `"leachate_high_vs_control"` | Main effect of `high` vs `control`                                    |

# Leachate effects

## Effect of high exposure across all hpf levels
```{r}
# output results of contrast
res_high <- (results(dds, name = "leachate_high_vs_control"))

summary(res_high)
```

### Extract significant genes

```{r}
padj.cutoff <- 0.05
lfc.cutoff <- 0.5
```

```{r}
# Convert results table into tibble
res_high_tb <- res_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_high_noLFCcutoff <- res_high_tb %>%
  filter(padj < padj.cutoff)

sig_high <- sig_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs across hpf at high exposure level:", nrow(sig_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs across hpf at high exposure level:", nrow(sig_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_high, "../output/09_deg_multifactorial/noint/sig_high.csv")
write_csv(sig_high_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_high_ashr <- lfcShrink(dds, coef = 6, type = "ashr")

plotMA(res_high)
plotMA(res_high_ashr)
```

```{r}
EnhancedVolcano(res_high,
    lab = rownames(res_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure across developmental stages',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
vhigh <- 
  ggplot(res_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_high[sig_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_high[sig_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure across developmental stages") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(0,3)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

vhigh
```

## Effect of mid exposure across developmental stages

```{r}
# output results 
res_mid <- (results(dds, name = "leachate_mid_vs_control"))

summary(res_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_mid_tb <- res_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_mid_noLFCcutoff <- res_mid_tb %>%
  filter(padj < padj.cutoff)

sig_mid <- sig_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs across hpf at mid exposure level:", nrow(sig_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs across hpf at mid exposure level:", nrow(sig_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_mid, "../output/09_deg_multifactorial/noint/sig_mid.csv")
write_csv(sig_mid_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}

# shrink lfc for visualizations
res_mid_ashr <- lfcShrink(dds, coef = 5, type = "ashr")

plotMA(res_mid)
plotMA(res_mid_ashr)
```

```{r}
EnhancedVolcano(res_mid,
    lab = rownames(res_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure across hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
vmid <- 
  ggplot(res_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_mid[sig_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_mid[sig_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure across hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(0,3)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

vmid
```

## Effect of low exposure across hpf

```{r}
# output results of contrast
res_low <- (results(dds, name = "leachate_low_vs_control"))

summary(res_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_low_tb <- res_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_low_noLFCcutoff <- res_low_tb %>%
  filter(padj < padj.cutoff)

sig_low <- sig_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs across hpf at low exposure level:", nrow(sig_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs across hpf at low exposure level:", nrow(sig_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_low, "../output/09_deg_multifactorial/noint/sig_low.csv")
write_csv(sig_low_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_low_ashr <- lfcShrink(dds, coef = 4, type = "ashr")

plotMA(res_low)
plotMA(res_low_ashr)
```

```{r}
EnhancedVolcano(res_low,
    lab = rownames(res_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure across hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
vlow <- 
  ggplot(res_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_low[sig_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_low[sig_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure across hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-3,3)) +
  scale_y_continuous(limits = c(0,3)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

vlow
```


## Facet volcano plot


```{r}
# First, strip labels from all individual ggplot objects
# (You can do this in a loop if the plots are in a list)

remove_labels <- function(p) {
  p +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 8)  # adjust title size here
    )
}

# Example: if your plots are named v4control, v4low, etc.
plots <- list(vlow, vmid, vhigh)

plots_clean <- lapply(plots, remove_labels)

# Now plot in 3 rows × 3 columns
volcano_facet_leachate <- plot_grid(
  plotlist = plots_clean,
  nrow = 1,
  ncol = 3,
  labels = NULL,
  label_size = 0,
  align = "hv"
)

volcano_facet_leachate
```

```{r}
ggsave("../output/09_deg_multifactorial/noint/volcano_facet_leachate.png", volcano_facet_leachate, width = 16, height = 12, dpi = 300)
```










# Developmental (hpf) effects


## Effect of progressing from 4 to 9 hpf

```{r}
# output results of contrast
res_9 <- (results(dds, name = "hpf_9_vs_4"))

summary(res_9)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_high_tb <- res_9hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_high_noLFCcutoff <- res_9hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_high <- sig_9hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_high, "../output/09_deg_multifactorial/noint/sig_9hpf_high.csv")
write_csv(sig_9hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_9hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf9.leachatehigh")), type = "ashr")

plotMA(res_9hpf_high)
plotMA(res_9hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_high,
    lab = rownames(res_9hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9high <- 
  ggplot(res_9hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9high
```

## Effect of mid exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf9.leachatemid")
)))

summary(res_9hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_mid_tb <- res_9hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_mid_noLFCcutoff <- res_9hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_mid <- sig_9hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_mid, "../output/09_deg_multifactorial/noint/sig_9hpf_mid.csv")
write_csv(sig_9hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_9hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf9.leachatemid")), type = "ashr")

plotMA(res_9hpf_mid)
plotMA(res_9hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_mid,
    lab = rownames(res_9hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9mid <- 
  ggplot(res_9hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9mid
```

## Effect of low exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf9.leachatelow")
)))

summary(res_9hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_low_tb <- res_9hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_low_noLFCcutoff <- res_9hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_low <- sig_9hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_low, "../output/09_deg_multifactorial/noint/sig_9hpf_low.csv")
write_csv(sig_9hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_9hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf9.leachatelow")), type = "ashr")

plotMA(res_9hpf_low)
plotMA(res_9hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_low,
    lab = rownames(res_9hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9low <- 
  ggplot(res_9hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9low
```

## Effect of high exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_high <- (results(dds, name = "leachate_high_vs_control")
)

summary(res_4hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_high_tb <- res_4hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_high_noLFCcutoff <- res_4hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_high <- sig_4hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_high, "../output/09_deg_multifactorial/noint/sig_4hpf_high.csv")
write_csv(sig_4hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_4hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_high_ashr <- lfcShrink(dds, coef = 6 , type = "ashr")

plotMA(res_4hpf_high)
plotMA(res_4hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_high,
    lab = rownames(res_4hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4high <- 
  ggplot(res_4hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4high
```

## Effect of mid exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_mid <- (results(dds, name = "leachate_mid_vs_control")
)

summary(res_4hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_mid_tb <- res_4hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_mid_noLFCcutoff <- res_4hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_mid <- sig_4hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_mid, "../output/09_deg_multifactorial/noint/sig_4hpf_mid.csv")
write_csv(sig_4hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_4hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_mid_ashr <- lfcShrink(dds, coef = 5 , type = "ashr")

plotMA(res_4hpf_mid)
plotMA(res_4hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_mid,
    lab = rownames(res_4hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4mid <- 
  ggplot(res_4hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4mid
```

## Effect of low exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_low <- (results(dds, name = "leachate_low_vs_control", alpha=0.05)
)

summary(res_4hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_low_tb <- res_4hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_low_noLFCcutoff <- res_4hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_low <- sig_4hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_low, "../output/09_deg_multifactorial/noint/sig_4hpf_low.csv")
write_csv(sig_4hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/noint/sig_4hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_low_ashr <- lfcShrink(dds, coef = 4 , type = "ashr")
plotMA(res_4hpf_low)
plotMA(res_4hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_low,
    lab = rownames(res_4hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4low <- 
  ggplot(res_4hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4low
```

# Summary 

## Facet volcano plot

```{r}
# First, strip labels from all individual ggplot objects
# (You can do this in a loop if the plots are in a list)

remove_labels <- function(p) {
  p +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 8)  # adjust title size here
    )
}

# Example: if your plots are named v4control, v4low, etc.
plots <- list( 
   v4low, v9low, v14low,
   v4mid, v9mid, v14mid,
   v4high, v9high, v14high
)

plots_clean <- lapply(plots, remove_labels)

# Now plot in 3 rows × 3 columns
volcano_facet <- plot_grid(
  plotlist = plots_clean,
  nrow = 3,
  ncol = 3,
  labels = NULL,
  label_size = 0,
  align = "hv"
)

volcano_facet
```

```{r}
ggsave("../output/09_deg_multifactorial/noint/volcano_facet.png", volcano_facet, width = 16, height = 12, dpi = 300)
```

::: callout-warning
Why do volcano plots not look like volcanos!?

Batch Effects or Low Biological Signal If your experimental design has confounding factors (e.g., batch effects), or the leachate treatment has low impact, then few genes will show strong changes.

Extreme p-values Your plot shows points with -log10(padj) values above 20--30. That means padj is very close to 0, which could indicate: - Very large sample size - Very consistent effect - Or something off with the dispersion estimate or p-value adjustment (e.g., if some genes had zero variance across groups)
:::

# Next steps

Re-run the model without the interaction term.
The likelihood ratio test (LRT) shows that the interaction (hpf:leachate) is not significant!
We can now test main effects of leachate across stages (i.e., an overall leachate effect) using `~ hpf + leachate`

## Exlpore cooks distance

> DESeq2 relies on the negative binomial distribution to make estimates and perform statistical inference on differences. While the negative binomial is versatile in having a mean and dispersion parameter, extreme counts in individual samples might not fit well to the negative binomial. For this reason, we perform automatic detection of count outliers. We use Cook's distance, which is a measure of how much the fitted coefficients would change if an individual sample were removed (Cook 1977). For more on the implementation of Cook's distance see the manual page for the *results* function. Below we plot the maximum value of Cook's distance for each row over the rank of the test statistic to justify its use as a filtering criterion.

## Explore RUVSeq
