---
title: "Gene ontology (GO) term analysis for differentially expressed genes"
subtitle: "using `goseq` with gene length bias correction"
author: "Sarah Tanja"
date: 2025-10-21
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

In the clustering analysis (`06_DEGreport_WGCNA_cluster.qmd`), we
identified 3 distinct gene expression patterns among the 130 significant
genes affected by leachate, 128 genes out of the 130 DEGs fit into one
of the three patterns.

These genes cluster based on their expression patterns across stage &
leachate concentrations (control, low, mid, high).

Leachate responsive DEGs cluster breakdown:

-   **Cluster 1**: 72 genes, peak in early gastrula –\> leachate_up

-   **Cluster 2**: 25 genes, peak in prawnchip –\> leachate_hump

-   **Cluster 3**: 31 genes, peak in cleavage –\> leachate_down

We also binned the 10,793 stage DEGs into 4 developmental programs via
WGCNA module (indicated by color).

Stage responsive DEGs module breakdown:

-   **stage_down**

    -   turquoise: 2805 genes

-   **stage_hump**

    -   yellow: 1381 genes

-   **stage_valley**

    -   green: 1108 genes
        -   red: 326 genes

-   **stage_up**

    -   blue: 2588 genes
        -   brown: 2436 genes

-   unassigned "grey module"

    -   grey: 149 genes

::: callout-important
Our goal is to now perform functional enrichment analysis on each module
& cluster separately to understand the biological processes associated
with each expression pattern.

Functional enrichment tests whether a biological category contains more
genes from a given gene set than expected under random sampling from the
expressed genome.

$$
\text{Fold Enrichment} = \frac{k/n}{K/N}
$$ Where:

-   n = number of genes in our gene set (e.g. DEGs or a WGCNA module)

-   k = number of genes in our set annotated to a GO term

-   K = number of genes in the background annotated to a GO term

-   N = total number of genes in the background (e.g. all genes tested
    for differential expression, the "universe")

Fold enrichment can be:

-   `>1`: over-represented

-   `=1`: expected

-   `<1`: under-represented

GO / pathway enrichment tests ask:

> Is a functional category over-represented in my gene list compared to
> the background?

H~o~: The null hypothesis is that genes in our gene set are a random
sample from the background genome.

H~a~: The alternative hypothesis is that genes in our gene set(s) are
enriched for a particular function.

Enrichment p-value is the probability of observing k or more annotated
genes by chance $$
p_{\text{enrich}} = P(X \ge k)
$$

**Functional enrichment analysis depends therefore on how we group our
gene set.**

For example, if we run GO enrichment on all 130 leachate DEGs together,
we will get one set of GO terms that describe the overall functions of
all these genes combined.

If we run GO enrichment on each of the 3 leachate DEG clusters
separately, we will get 3 different sets of GO terms, each describing
the functions of genes in that specific co-expression pattern.

Similarly, running GO enrichment on each of the 4 stage DEG modules
separately yields 4 different sets of GO terms, each reflecting the
functions of genes in that developmental program.
:::

We use the R package `goseq` for GO enrichment. `goseq` corrects for
gene length bias - longer genes are more likely to be detected as DEGs
in RNA-Seq simply because they accumulate more reads. `goseq` accounts
for this by using a probability weighting function.

Run GOseq using:

DEG vector Background (all tested genes) Vector of gene lengths (from
your annotation, GTF, eggNOG)

Get the corrected p-values and enriched GO terms.

## Inputs

DEG vectors

-   `stage_DEG_WGCNA_modules.csv` : A dataframe mapping 10793 stage DEGs
    to a WGCNA module. *the grey module (genes that do not cluster) need
    not be passed through enrichment analysis*

-   `leachate_DEG_clusters.csv` : A dataframe mapping 128 genes to an
    expression cluster (1, 2, or 3)

-   `all_genes.csv` : a dataframe (list) of all genes used in the DESeq2
    analysis

Gene lengths - functional annotation:
http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz
(EggNog) - functional annotation:
http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz
(KEGG)

> “We performed GO enrichment using GOseq for stage-program
> clusters/modules derived from co-expression patterns. Genes not
> assigned to a module (WGCNA grey) were retained in the DEG set but
> excluded from module-level interpretation.”

## Outputs

-   GO enrichment results for each of the modules/clusters
-   Reduced GO terms (redundancy removed) for gene universe, and for
    each module/cluster
-   Semantic similarity matrix for gene universe, and for each
    module/cluster

# Setup

## Paths

```{r}
output_path <- "../../output/07_enrichment/goseq/"
```

## Install packages

```{r, eval=FALSE}
# CRAN packages
#if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')

# Bioconductor packages
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("GenomicRanges")
#BiocManager::install("Biostrings")
#BiocManager::install("goseq")
#BiocManager::install("rrvgo")
#BiocManager::install("org.Ce.eg.db") # c.elegans genome-wide annotation
```

## Load packages

```{r}
library(GenomicRanges)
library(Biostrings)
library(goseq)
library(rrvgo)
library(org.Ce.eg.db)
library(tidyverse)
```

## Load inputs

Read in gene lists

```{r}
# significant genes lists
sig_genes_leachate <- read.csv("../../output/06_deg/spawnnight/sig_genes_leachate.csv")

sig_genes_stage <- read.csv("../../output/06_deg/spawnnight/sig_genes_stage.csv")

# Full gene set tested in DESeq2 analysis
universe_genes <- read.csv("../../output/06_deg/spawnnight/all_genes.csv")

# leachate cluster assignments 
leachate_DEG_clusters <- read.csv("../../output/06_deg/cluster/leachate_DEG_clusters.csv")

# stage WGCNA module assignments
stage_DEG_modules <- read.csv("../../output/06_deg/cluster/stage_DEG_WGCNA_modules.csv")
```

### Rename values for simplicity & clarity

Make leachate_DEG_clusters column named gene_id for joining, and rename
clusters to be descriptive: Leachate DEG cluster breakdown: - **Cluster
1**: 72 genes, peak in early gastrula, pattern is up - **Cluster 2**: 25
genes, peak in prawnchip, pattern is hump - **Cluster 3**: 31 genes,
peak in cleavage, pattern is down

```{r}
leachate_DEG_clusters <- leachate_DEG_clusters %>%
  dplyr::rename(gene_id = genes)
```

```{r}

leachate_deg_patterns <- sig_genes_leachate %>%
  left_join(leachate_DEG_clusters, by = "gene_id") %>%
  mutate(pattern_leachate = case_when(
                      cluster == 1 ~ "up",
                      cluster == 2 ~ "hump",
                      cluster == 3 ~ "down",
                      is.na(cluster) ~ "unassigned"),
          peak_leachate = case_when(
                      cluster == 1 ~ "earlygastrula",
                      cluster == 2 ~ "prawnchip",
                      cluster == 3 ~ "cleavage",
                      is.na(cluster) ~ "unassigned"
                    ), 
          deg_leachate = "leachate") %>% 
  rename(DEGpattern = cluster)

leachate_deg_patterns %>% 
count(DEGpattern, peak_leachate, pattern_leachate, sort = TRUE) %>% knitr::kable()  
```

```{r}
stage_deg_modules <- stage_DEG_modules %>% 
  mutate(
    peak_stage = if_else(WGCNAmodule == "grey", 
                         "unassigned", peak_stage),
    pattern_type = if_else(WGCNAmodule == "grey", 
                           "unassigned", pattern_type)) %>% 
  mutate(
    pattern_type = if_else(pattern_type == "monotonic_up", 
                           "up", pattern_type),
    pattern_type = if_else(pattern_type == "monotonic_down", 
                           "down", pattern_type)
  ) %>% 
  dplyr::rename(pattern_stage = pattern_type) %>% 
  dplyr::select(!module_stage) %>%
  mutate(deg_stage = "stage")


stage_deg_modules %>%
count(peak_stage, pattern_stage, sort = TRUE) %>% 
  knitr::kable()
```

```{r}
stage_deg_modules %>%
count(WGCNAmodule, pattern_stage, sort = TRUE) %>% 
  knitr::kable()
```

We should see 149 rows renamed with grey as unassigned. These are the
genes that are differentially expressed due to stage, but did not
cluster in the WGCNA analysis with any other expression patterns. We
want big picture results, and these genes are "noise" that don't follow
a strong expression pattern.

> Genes assigned to the WGCNA grey module were classified as unassigned,
> while remaining genes grouped into cleavage-, prawn-chip–, and early
> gastrula–associated expression programs.

How many of the leachate responsive and stage responsive genes overlap?

```{r}
length(intersect(stage_deg_modules$gene_id, leachate_deg_patterns$gene_id))
```

::: callout-important
ALL of our leachate responsive DEGs are ALSO stage responsive.
:::

What does this mean... this implies that the genes impacted by leachate
are the ones that are changing...? I mean out of \~12k genes tested
though, \~10k were changing across stage.

Import EggNog for GO and KEGG annotation

```{r}
# Read in the file
EggNog <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")

# Rename "#query" to gene_id
EggNog <- EggNog %>%
  rename(gene_id="#query")

paste(nrow(EggNog), "genes have functional annotation")
```

Import fasta for sequence length annotation

```{r, eval=FALSE}
# Load transcriptome sequences to get gene lengths
genome_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")

# Filter for the 12221 background universe genes
universe_seqs <- genome_seqs[names(genome_seqs) %in% universe_genes$gene_id]

# Filter for the 130 leachate DEGs 
#leachate_seqs <- genome_seqs[names(genome_seqs) %in% sig_genes_leachate$gene_id]

# Filter for the 10793 stage DEGs
#stage_seqs <- genome_seqs[names(genome_seqs) %in% sig_genes_stage$gene_id]
```

# Annotate gene lists

Match genes in universe_genes to EggNog annotation dataframe

```{r}
# Annotate all genes with EggNog annotations
universe_genes_anno <- left_join(universe_genes, EggNog, by = "gene_id")
```

## Add gene lengths

::: callout-note
Gene lengths are required for goseq bias correction.
:::

```{r}
# Import GFF file to get gene lengths
# Check if file exists before importing
gff_file <- "../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3"

if (file.exists(gff_file)) {
  gff <- rtracklayer::import(gff_file)
  
  transcripts <- subset(gff, type == "transcript")
  
  transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE)
  
  transcript_lengths <- width(transcripts_gr)
  
  gene_id <- transcripts_gr$ID
  
  lengths <- cbind(gene_id, transcript_lengths)
  
  lengths <- as.data.frame(lengths)
  
  paste("Gene lengths extracted for", nrow(lengths), "transcripts")
} else {
  warning("GFF file not found. Gene length bias correction will not be available.")
  lengths <- data.frame(gene_id = character(), transcript_lengths = integer())
}
```

## Join gene length, leachate pattern & stage module info

Add length & module/cluster info to the annotation dataframe

```{r}
universe_anno <- universe_genes_anno %>% 
  left_join(lengths, by = "gene_id") %>% 
  left_join(leachate_deg_patterns, by = "gene_id") %>% 
  left_join(stage_deg_modules, by = "gene_id")
```

Write out the annotated universe dataframe

```{r}
write_csv(universe_anno, file.path(output_path, "universe_anno.csv"))
```

## Explore GO

### GO coverage

How many of the genes are GO annotated?

```{r}
# Filter out any that lack GO term annotation
universe_genes_GO <- universe_anno %>% 
  mutate(GOs = na_if(str_trim(GOs), "-")) %>%
  filter(!is.na(GOs), GOs != "")

# What is GO coverage in the background universe?
paste("GO annotated genes in DESeq background universe set:", 
      nrow(universe_genes_GO), "/", 
      nrow(universe_genes), "=",
      round(100*nrow(universe_genes_GO)/nrow(universe_genes),1),"%")
```

-   12,221 genes are in our DESeq2 tested background universe
-   6,846 of those have at least one GO annotation
-   56% coverage is very typical for non-model or semi-annotated
    organisms

> “Of the 12,221 genes included in the DESeq2 testing universe, 6,846
> (56%) had at least one Gene Ontology annotation and were eligible for
> GO-based analyses.”
>
> “Gene Ontology annotations were available for 6,846 of the 12,221
> genes tested in DESeq2 (56%), consistent with partial functional
> annotation of the transcriptome.”
>
> “GO enrichment analyses were restricted to the subset of tested genes
> with available GO annotations (6,846 of 12,221 genes; 56%), which
> constituted the effective background universe for functional
> analyses.”

Calculate GO coverage within clusters for descriptive purposes $$
\text{GO coverage} = \frac{\#\,\text{genes with GO annotation in the set}}{\#\,\text{genes in the set}}
$$

This answers: “how well annotated is this DEGpattern?” or what is the GO
coverage for all leachate responsive DEG expression patterns?

```{r}

leachate_go_summary <- universe_anno %>%
  filter(gene_id %in% sig_genes_leachate$gene_id) %>%
  mutate(has_GO = ifelse(!is.na(GOs) & GOs != "", 1, 0)) %>%
  group_by(DEGpattern) %>%
  summarise(
    n_genes = n(),
    n_genes_with_GO = sum(has_GO),
    GO_coverage = round(100 * n_genes_with_GO / n_genes, 1)
  )

leachate_go_summary %>% knitr::kable()
```

What is the GO coverage for all DEGs due to developental stage WGCNA
modules?

```{r}
# What is GO coverage for all stage WGCNA modules?
stage_go_summary <- universe_anno %>%
  filter(gene_id %in% sig_genes_stage$gene_id) %>%
  mutate(has_GO = ifelse(!is.na(GOs) & GOs != "", 1, 0)) %>%
  group_by(WGCNAmodule) %>%
  summarise(
    n_genes = n(),
    n_genes_with_GO = sum(has_GO),
    GO_coverage = round(100 * n_genes_with_GO / n_genes, 1)
  )

stage_go_summary %>% knitr::kable()
```

```{r}
# What is GO coverage for all stage patterns?
stage_pattern_go_summary <- universe_anno %>%
  filter(gene_id %in% sig_genes_stage$gene_id) %>%
  mutate(has_GO = ifelse(!is.na(GOs) & GOs != "", 1, 0)) %>%
  group_by(pattern_stage) %>%
  summarise(
    n_genes = n(),
    n_genes_with_GO = sum(has_GO),
    GO_coverage = round(100 * n_genes_with_GO / n_genes, 1)
  )

stage_pattern_go_summary %>% knitr::kable()
```

# Make gene universe vectors

For bias correction

The gene universe should be the full set of genes that were tested in
the DESeq2 analysis (universe_genes), including those that were not
found to be differentially expressed. For goseq, we need to create a
vector of all gene IDs in the DESeq2 dataset (include **all** genes,
regardless of whether or not they have GO annotations), with a `1` value
if DE and a `0` value if not.

nullp() models the probability of being DE as a function of length using
the full DE universe.

Later, goseq() only tests GO terms using genes in gene2cat (the \~6846
genes we have functional annotation for), but the PWF (bias correction)
was learned from all the data (all 12,221 genes).

So we need a background universe vectors for all DEGs, and each pattern
/ module.

```{r}
# Universe for bias correction
# All 12221 genes from our filtered DEG test
universe_vector <- universe_genes$gene_id
names(universe_vector) <- universe_vector
```

These are annotated dataframes subset to just DEGs and to each cluster

```{r}
leachate_anno <- universe_anno %>% 
  filter(gene_id %in% sig_genes_leachate$gene_id)

leachate_anno_up <- universe_anno %>% 
  filter(pattern_leachate == "up")

leachate_anno_hump <- universe_anno %>%
  filter(pattern_leachate == "hump")

leachate_anno_down <- universe_anno %>%
  filter(pattern_leachate == "down")
```

```{r}
stage_anno <- universe_anno %>% 
  filter(gene_id %in% sig_genes_stage$gene_id)

stage_anno_up <- universe_anno %>%
  filter(pattern_stage == "up")

stage_anno_hump <- universe_anno %>%
  filter(pattern_stage == "hump")

stage_anno_valley <- universe_anno %>%
  filter(pattern_stage == "valley")

stage_anno_down <- universe_anno %>%
  filter(pattern_stage == "down")
```

Create named vectors for DE status (1 = DE, 0 = not DE) for each
leachate responsive pattern and for each WGCNA cluster

#### Vectors for leachate responsive gene patterns

```{r}
vector_leachate <- as.integer(universe_vector %in% leachate_anno$gene_id)
names(vector_leachate) <- universe_vector

vector_leachate_up <- as.integer(universe_vector %in% leachate_anno_up$gene_id)
names(vector_leachate_up) <- universe_vector

vector_leachate_hump <- as.integer(universe_vector %in% leachate_anno_hump$gene_id)
names(vector_leachate_hump) <- universe_vector

vector_leachate_down <- as.integer(universe_vector %in% leachate_anno_down$gene_id)
names(vector_leachate_down) <- universe_vector
```

#### Vectors for DEGs due to developmental stage

```{r}
vector_stage <- as.integer(universe_vector %in% stage_anno$gene_id)
names(vector_stage) <- universe_vector

vector_stage_up <- as.integer(universe_vector %in% stage_anno_up$gene_id)
names(vector_stage_up) <- universe_vector

vector_stage_hump <- as.integer(universe_vector %in% stage_anno_hump$gene_id)
names(vector_stage_hump) <- universe_vector

vector_stage_valley <- as.integer(universe_vector %in% stage_anno_valley$gene_id)
names(vector_stage_valley) <- universe_vector

vector_stage_down <- as.integer(universe_vector %in% stage_anno_down$gene_id)
names(vector_stage_down) <- universe_vector
```

# Make gene lengths vectors

Make named vectors of genes and gene lengths, only keep the genes
present in the background universe (our 12221 DESeq2 tested genes!)

```{r}
gene_lengths <- as.integer(lengths$transcript_lengths)
names(gene_lengths) <- lengths$gene_id

gene_lengths <- gene_lengths[universe_vector]
paste("Transcript lengths for", length(gene_lengths), "genes")
```

Check for NAs in gene_lengths - there should be none

```{r}
# Are there NAs in gene_lengths?
sum(is.na(gene_lengths))
```

The order of genes in gene_lengths must match the order in
universe_genes, should return TRUE

```{r}
identical(names(gene_lengths), names(universe_vector))

identical(names(gene_lengths), names(vector_leachate))   
identical(names(gene_lengths), names(vector_leachate_up))  
identical(names(gene_lengths), names(vector_leachate_hump))
identical(names(gene_lengths), names(vector_leachate_down))

identical(names(gene_lengths), names(vector_stage))
identical(names(gene_lengths), names(vector_stage_up))
identical(names(gene_lengths), names(vector_stage_hump))
identical(names(gene_lengths), names(vector_stage_valley))
identical(names(gene_lengths), names(vector_stage_down))
```

# Null probability weighting function (bias for gene length)

::: callout-note
Gene length bias correction accounts for the fact that longer genes are
more likely to be detected as differentially expressed in RNA-seq data
simply because they accumulate more reads. If length data is not
available, goseq can still run but without this correction.
:::

The function `nullp()` takes as input: - a named vector of 1s and 0s
indicating which genes are DE (1) and which are not (0) for all genes in
the universe - a named vector of gene lengths for all genes in the
universe (background)

The probability weighting function is defined for all genes we tested in
DESeq2 (our full “universe” of 12221 genes), including genes with no GO
annotation.

```{r}
# Calculate probability weighting function for each leachate responsive DEG pattern
pwf_leachate <- nullp(vector_leachate, bias.data = gene_lengths) # all leachate responsive DEGs
pwf_leachate_up <- nullp(vector_leachate_up, bias.data = gene_lengths)
pwf_leachate_hump <- nullp(vector_leachate_hump, bias.data = gene_lengths)
pwf_leachate_down <- nullp(vector_leachate_down, bias.data = gene_lengths)

# Calculate probability weighting function for DEG stage modules
pwf_stage <- nullp(vector_stage, bias.data = gene_lengths) # all DEGs across stage
pwf_stage_up <- nullp(vector_stage_up, bias.data = gene_lengths)
pwf_stage_hump <- nullp(vector_stage_hump, bias.data = gene_lengths)
pwf_stage_valley <- nullp(vector_stage_valley, bias.data = gene_lengths)
pwf_stage_down <- nullp(vector_stage_down, bias.data = gene_lengths)
```

X axis is gene length Y axis is probability of being DE given length

For genes in each length bin, GOseq estimates the fraction of genes that
are DE. Points are `#DE genes in bin/#genes in bin` for genes with
similar length. GOseq then fits a smooth curve through those points –
this is the probability weighting function (PWF). This fitted curve is
what GOseq uses internally to correct for length bias when doing GO
enrichment.

Strong slope / curvature → there is substantial length bias → long genes
are much more (or less) likely to be called DE than short ones → length
correction is important.

Nearly flat line → little or no detectable length bias → the enrichment
results will be similar to a standard hypergeometric ORA.

# GO term mapping

This creates a dataframe that maps gene IDs to GO terms for all genes in
the testing universe. There will be 2 columns, gene_id and GO term, with
one GO term mapped to one gene. Since there are multiple GO terms per
gene, genes will appear in multiple rows in these dataframes. Likewise,
more than one gene may map to the same GO term.

### gene2go

Build the gene to GO term mappings for all genes in the universe

```{r}
gene2go <- universe_genes_anno %>%
  dplyr::select(gene_id, GOs) %>%
  separate_rows(GOs, sep = ",") %>%
  filter(!is.na(GOs), GOs != "") %>%
  distinct(gene_id, GOs) %>%
  rename(gene = gene_id,
         category = GOs) %>%
  filter(gene %in% universe_vector) %>%
  as.data.frame()
```

Save GO term mapping

```{r}
write_csv(gene2go, file.path(output_path, "gene2go.csv"))
```

### gowide

```{r}
gowide <- gene2go %>%
  distinct(gene, category) %>%          # drop duplicates if any
  group_by(category) %>%
  summarise(
    universegenes = list(sort(unique(gene))),  # list-column of genes
    universecount = n_distinct(gene),          # how many genes map
    .groups  = "drop"
  )

```

```{r}
write_csv(gowide, file.path(output_path, "gowide.csv"))
```

## Run `goseq`

Run GO enrichment for biological process terms

```{r}
GO_leachate_up <- goseq(
  pwf = pwf_leachate_up,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_leachate_hump <- goseq(
  pwf = pwf_leachate_hump,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_leachate_down <- goseq(
  pwf = pwf_leachate_down,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_leachate <- goseq(
  pwf = pwf_leachate,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)
```

```{r}
GO_stage_up <- goseq(
  pwf = pwf_stage_up,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_stage_hump <- goseq(
  pwf = pwf_stage_hump,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_stage_valley <- goseq(
  pwf = pwf_stage_valley,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_stage_down <- goseq(
  pwf = pwf_stage_down,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO_stage <- goseq(
  pwf = pwf_stage,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)
```

#### Control for FDR

Use the Benjamini-Hochberg method to control false discovery rate

Control for FDR with over represented p values

```{r}
# leachate responsive
GO_leachate$over_represented_bh <- 
  p.adjust(GO_leachate$over_represented_pvalue, method = "BH")

GO_leachate_up$over_represented_bh <-
  p.adjust(GO_leachate_up$over_represented_pvalue, method = "BH")

GO_leachate_hump$over_represented_bh <-
  p.adjust(GO_leachate_hump$over_represented_pvalue, method = "BH")

GO_leachate_down$over_represented_bh <-
  p.adjust(GO_leachate_down$over_represented_pvalue, method = "BH")


# DEGs across stage
GO_stage$over_represented_bh <- 
  p.adjust(GO_stage$over_represented_pvalue, method = "BH")

GO_stage_up$over_represented_bh <- 
  p.adjust(GO_stage_up$over_represented_pvalue, method = "BH")

GO_stage_hump$over_represented_bh <- 
  p.adjust(GO_stage_hump$over_represented_pvalue, method = "BH")

GO_stage_valley$over_represented_bh <-
  p.adjust(GO_stage_valley$over_represented_pvalue, method = "BH")

GO_stage_down$over_represented_bh <- 
  p.adjust(GO_stage_down$over_represented_pvalue, method = "BH")
```

Control for FDR with under represented p values

```{r}
# leachate responsive
GO_leachate$under_represented_bh <- 
  p.adjust(GO_leachate$under_represented_pvalue, method = "BH")

GO_leachate_up$under_represented_bh <-
  p.adjust(GO_leachate_up$under_represented_pvalue, method = "BH")

GO_leachate_hump$under_represented_bh <-
  p.adjust(GO_leachate_hump$under_represented_pvalue, method = "BH")

GO_leachate_down$under_represented_bh <-
  p.adjust(GO_leachate_down$under_represented_pvalue, method = "BH")


# DEGs across stage
GO_stage$under_represented_bh <- 
  p.adjust(GO_stage$under_represented_pvalue, method = "BH")

GO_stage_up$under_represented_bh <- 
  p.adjust(GO_stage_up$under_represented_pvalue, method = "BH")

GO_stage_hump$under_represented_bh <- 
  p.adjust(GO_stage_hump$under_represented_pvalue, method = "BH")

GO_stage_valley$under_represented_bh <-
  p.adjust(GO_stage_valley$under_represented_pvalue, method = "BH")

GO_stage_down$under_represented_bh <- 
  p.adjust(GO_stage_down$under_represented_pvalue, method = "BH")
```

::: callout-note
This should increase the number of columns in the GO\_\*\*\* output
dataframes from 7 to 9
:::

### Filter for BP terms only

```{r}
GO_leachate <- GO_leachate %>% filter(ontology == "BP")
GO_leachate_up <- GO_leachate_up %>% filter(ontology == "BP")
GO_leachate_hump <- GO_leachate_hump %>% filter(ontology == "BP")
GO_leachate_down <- GO_leachate_down %>% filter(ontology == "BP")

GO_stage <- GO_stage %>% filter(ontology == "BP")
GO_stage_up <- GO_stage_up %>% filter(ontology == "BP")
GO_stage_hump <- GO_stage_hump %>% filter(ontology == "BP")
GO_stage_valley <- GO_stage_valley %>% filter(ontology == "BP")
GO_stage_down <- GO_stage_down %>% filter(ontology == "BP")
```

#### Save GO results

```{r}
write_csv(GO_leachate, 
          file.path(output_path, "GO_leachate.csv"))

write_csv(GO_leachate_up, 
          file.path(output_path, "GO_leachate_up.csv"))

write_csv(GO_leachate_hump, 
          file.path(output_path, "GO_leachate_hump.csv"))

write_csv(GO_leachate_down, 
          file.path(output_path, "GO_leachate_down.csv"))

write_csv(GO_stage, 
          file.path(output_path, "GO_bp_stage.csv"))

write_csv(GO_stage_up, 
          file.path(output_path, "GO_stage_up.csv"))

write_csv(GO_stage_hump, 
          file.path(output_path, "GO_stage_hump.csv"))

write_csv(GO_stage_valley, 
          file.path(output_path, "GO_stage_valley.csv"))

write_csv(GO_stage_down, 
          file.path(output_path, "GO_stage_down.csv"))
```

#### Filter for significantly enriched GO terms

Filter for biological process terms with adjusted p-values \< 0.05

##### Are leachate-responsive DEGs enriched?

```{r}
GO_BPsig_leachate <- GO_leachate %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for leachate DEGs:", nrow(GO_BPsig_leachate))

GO_BPsig_leachate_up <- GO_leachate_up %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for leachate up DEGs:", nrow(GO_BPsig_leachate_up))

GO_BPsig_leachate_hump <- GO_leachate_hump %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for leachate hump DEGs:", nrow(GO_BPsig_leachate_up))

GO_BPsig_leachate_down <- GO_leachate_down %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for leachate down DEGs:", nrow(GO_BPsig_leachate_up))
```

Nothing! Nothing is enriched for leachate... fuck.

Ok. Well. Now we know. This indicates that our DEGs impacted by leachate
don't all do particular functions, or that there is no particular type
of biological process that these genes all do which are impacted by
leachate.

> For our situation (≈12k background genes, ≈130 DEGs): Small DEG set +
> custom annotation If only, say, 40–60 of our 130 DEGs have GO
> annotations, and those are spread across many terms, there just may
> not be enough clustering in any single GO category. Length bias
> correction is conservative, GOseq is explicitly correcting for bias. A
> lot of “borderline”-looking enrichments from simple hypergeometric
> tests disappear once you account for length. *Montipora capitata*
> annotation is sparse. With non-model genomes, GO coverage is often
> patchy. Some explanations are that we could have a true biological
> pattern that doesn’t map cleanly to GO terms. OR our DE list is sparse
> so DEGs represent many different processes at once (e.g., general
> stress + development + metabolism), each individual term may only get
> 2–3 hits → not enough to reach significance threshold for enrichment,
> especially after adjusting p-values for multiple comparisons.

##### What about stage?

How many DEGs are there across stage and in each module?

```{r}
numDE_stage <- nrow(stage_anno)
numDE_stage_up <- nrow(stage_anno_up)
numDE_stage_hump <- nrow(stage_anno_hump)
numDE_stage_valley <- nrow(stage_anno_valley)
numDE_stage_down <- nrow(stage_anno_down)

numUniverse <- nrow(universe_genes_anno)
```

```{r}
# stage
GO_BPsig_stage <- GO_stage %>%
  filter(ontology == "BP") %>%
    filter(over_represented_bh < 0.05 |
             under_represented_bh < 0.05) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for stage DEGs:", nrow(GO_BPsig_stage))

# stage up 
GO_BPsig_stage_up <- GO_stage_up %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for stage up DEGs:", nrow(GO_BPsig_stage_up))

# stage hump 
GO_BPsig_stage_hump <- GO_stage_hump %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for stage hump DEGs:", nrow(GO_BPsig_stage_hump))

# stage valley
GO_BPsig_stage_valley <- GO_stage_valley %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for stage valley DEGs:", nrow(GO_BPsig_stage_valley))

# stage down 
GO_BPsig_stage_down <- GO_stage_down %>%
  filter(ontology == "BP") %>%
    filter(
    coalesce(over_represented_bh, 1) < 0.05 |
    coalesce(under_represented_bh, 1) < 0.05
  ) %>% 
  mutate(
    enrichment_direction = case_when(
      over_represented_bh  < 0.05 ~ "over",
      under_represented_bh < 0.05 ~ "under",
      TRUE                         ~ "none"
    )
  )

paste("Number of significantly enriched GO BP terms for stage down DEGs:", nrow(GO_BPsig_stage_down))
```

##### Over vs. Under

How many of each were over vs under represented?

```{r}
# make a list
GO_BPsig_stage_list <- list(
  stage = GO_BPsig_stage,
  stage_up = GO_BPsig_stage_up,
  stage_hump = GO_BPsig_stage_hump,
  stage_valley = GO_BPsig_stage_valley,
  stage_down = GO_BPsig_stage_down
)

GOBPsig_stage_summary <- imap_dfr(
  GO_BPsig_stage_list,
  ~ .x %>%
      group_by(enrichment_direction) %>%
      summarise(n_terms = n(), .groups = "drop") %>%
      mutate(dataset = .y,
             n_total = nrow(.x))
)

GOBPsig_stage_summary %>% knitr::kable()
```

::: callout-important
Weird that each module either ONLY has over or under represented terms,
not a mix.
:::

Save the filtered significant GO terms that had enrichment

```{r}
write_csv(GO_BPsig_stage, file.path(output_path, "GO_BPsigENRICHED_stage.csv"))

write_csv(GO_BPsig_stage_up, file.path(output_path, "GO_BPsigENRICHED_stage_up.csv"))

write_csv(GO_BPsig_stage_hump, file.path(output_path, "GO_BPsigENRICHED_stage_hump.csv"))

write_csv(GO_BPsig_stage_valley, file.path(output_path, "GO_BPsigENRICHED_stage_valley.csv"))

write_csv(GO_BPsig_stage_down, file.path(output_path, "GO_BPsigENRICHED_stage_down.csv"))

```

## Reduce redundancy in significantly enriched GO terms

Calculate semantic similarity matrix with significantly enriched GO
terms for each stage WGCNA cluster

### stage

```{r}
# Make similarity matrix
simMatrix_stage <- calculateSimMatrix(
    GO_BPsig_stage$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

# Create scores
scores_stage <- setNames(
    -log10(pmax(GO_BPsig_stage$over_represented_bh, .Machine$double.xmin)),
    GO_BPsig_stage$category
  )

# Reduce terms for all DEGs 
reducedTerms_stage <- reduceSimMatrix(
    simMatrix_stage,
    scores = scores_stage,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )

# save similarity matrix and reduced terms
write_csv(as.data.frame(simMatrix_stage), file.path(output_path, "simMatrix_stage.csv"))
write_csv(reducedTerms_stage, file.path(output_path, "reducedTerms_stage.csv"))
```

### stage_up

```{r}
# Make similarity matrix
simMatrix_stage_up <- calculateSimMatrix(
    GO_BPsig_stage_up$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

# Create scores
scores_stage_up <- setNames(
    -log10(pmax(GO_BPsig_stage_up$over_represented_bh, .Machine$double.xmin)),
    GO_BPsig_stage_up$category
  )

# Reduce terms for all DEGs 
reducedTerms_stage_up <- reduceSimMatrix(
    simMatrix_stage_up,
    scores = scores_stage_up,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )

# save similarity matrix and reduced terms
write_csv(as.data.frame(simMatrix_stage_up), file.path(output_path, "simMatrix_stage_up.csv"))
write_csv(reducedTerms_stage_up, file.path(output_path, "reducedTerms_stage_up.csv"))
```

### stage_hump

```{r}
# Make similarity matrix
simMatrix_stage_hump <- calculateSimMatrix(
    GO_BPsig_stage_hump$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

# Create scores
scores_stage_hump <- setNames(
    -log10(pmax(GO_BPsig_stage_hump$under_represented_bh, .Machine$double.xmin)),
    GO_BPsig_stage_hump$category
  )

# Reduce terms for all DEGs 
reducedTerms_stage_hump <- reduceSimMatrix(
    simMatrix_stage_hump,
    scores = scores_stage_hump,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )

# save similarity matrix and reduced terms
write_csv(as.data.frame(simMatrix_stage_hump), file.path(output_path, "simMatrix_stage_hump.csv"))
write_csv(reducedTerms_stage_hump, file.path(output_path, "reducedTerms_stage_hump.csv"))
```

### stage_valley

```{r}
# Make similarity matrix
simMatrix_stage_valley <- calculateSimMatrix(
    GO_BPsig_stage_valley$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

# Create scores
scores_stage_valley <- setNames(
    -log10(pmax(GO_BPsig_stage_valley$over_represented_bh, .Machine$double.xmin)),
    GO_BPsig_stage_valley$category
  )

# Reduce terms for all DEGs 
reducedTerms_stage_valley <- reduceSimMatrix(
    simMatrix_stage_valley,
    scores = scores_stage_valley,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )

# save similarity matrix and reduced terms
write_csv(as.data.frame(simMatrix_stage_valley), file.path(output_path, "simMatrix_stage_valley.csv"))
write_csv(reducedTerms_stage_valley, file.path(output_path, "reducedTerms_stage_valley.csv"))
```

### stage_down

```{r}
# Make similarity matrix
simMatrix_stage_down <- calculateSimMatrix(
    GO_BPsig_stage_down$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

# Create scores
scores_stage_down <- setNames(
    -log10(pmax(GO_BPsig_stage_down$under_represented_bh, .Machine$double.xmin)),
    GO_BPsig_stage_down$category
  )

# Reduce terms for all DEGs 
reducedTerms_stage_down <- reduceSimMatrix(
    simMatrix_stage_down,
    scores = scores_stage_down,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )

# save similarity matrix and reduced terms
write_csv(as.data.frame(simMatrix_stage_down), file.path(output_path, "simMatrix_stage_down.csv"))
write_csv(reducedTerms_stage_down, file.path(output_path, "reducedTerms_stage_down.csv"))
```

# Reduce redundancy in non enriched GO terms

### Filter GO results to leachate responsive DEGs

Filter GO results for only terms that had a DEG annotated to them, even
if they weren’t significantly enriched. This will allow us to visualize
the full set of terms that had at least evidence of differential
expression (even if enrichment of those associated terms are not
significant after correction), and to see how they relate to each other
semantically.

```{r}
GO_leachate_DEG <- GO_leachate %>%
  filter(ontology == "BP") %>%
  filter(numDEInCat > 0)

GO_leachate_up_DEG <- GO_leachate_up %>%
  filter(ontology == "BP") %>%
  filter(numDEInCat > 0)

GO_leachate_hump_DEG <- GO_leachate_hump %>%
  filter(ontology == "BP") %>%
  filter(numDEInCat > 0)

GO_leachate_down_DEG <- GO_leachate_down %>%
  filter(ontology == "BP") %>%
  filter(numDEInCat > 0)
```

```{r}
any(duplicated(GO_leachate_DEG$category))
any(duplicated(GO_leachate_up_DEG$category))
any(duplicated(GO_leachate_hump_DEG$category))
any(duplicated(GO_leachate_down_DEG$category))
```

```{r}
GO_leachate_DEG %>% 
   mutate(term = reorder(term, numDEInCat)) %>% 
  slice_head(n = 15) %>%
ggplot(., aes(y = term)) +
  # Background: universe genes
  geom_col(
    aes(x = numInCat),
    color = "grey80",
    fill = "grey90",
    alpha = 0.3,
    width = 0.5
  ) +

  # Foreground: DEG genes
  geom_col(
    aes(x = numDEInCat),
    fill = "steelblue",
    alpha = 0.5,
    width = 0.5
  ) +

  labs(
    x = "Number of genes",
    y = NULL
  ) +

  theme_minimal()
  
```

### leachate

```{r}
# Make similarity matrix
simMatrix_leachate <- calculateSimMatrix(
    GO_leachate_DEG$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )

# Create scores
scores_leachate <- setNames(
    GO_leachate_DEG$numDEInCat,
    GO_leachate_DEG$category
  )

# Reduce terms 
reducedTerms_leachate <- reduceSimMatrix(
    simMatrix_leachate,
    scores = scores_leachate,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )

# save similarity matrix and reduced terms
write_csv(as.data.frame(simMatrix_leachate), file.path(output_path, "simMatrix_leachate.csv"))
write_csv(reducedTerms_leachate, file.path(output_path, "reducedTerms_leachate.csv"))
```

### leachate_up

### leachate_hump

### leachate_down

```{r}
go_scores <- universe_go_counts %>%
  dplyr::select(category, n_genes) %>%
  distinct() %>%
  tibble::deframe()

head(go_scores)
```

```{r, eval=FALSE}
simMatrix <- calculateSimMatrix(
  x       = names(go_scores),
  orgdb   = "org.Ce.eg.db",
  ont     = "BP",
  method  = "Rel"   # best general-purpose choice
)
```

```{r, eval=FALSE}
write_csv(as.data.frame(simMatrix), file.path(output_path, "universe_simMatrix.csv"))
```

## Reduce redundancy

```{r}
reducedTerms <- reduceSimMatrix(
  simMatrix,
  scores    = go_scores,
  threshold = 0.7,            # adjust as needed
  orgdb     = "org.Ce.eg.db"
)
```

```{r, eval=FALSE}
write_csv(reducedTerms, file.path(output_path, "universe_reducedTerms_GO.csv"))
```

## rrvgo plots

```{r}
library(rrvgo)
```

### Universe

```{r}
heatmapPlot(simMatrix,
            reducedTerms,
            annotateParent=TRUE,
            annotationLabel="parentTerm")
```

```{r}
scatterPlot(simMatrix, reducedTerms,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

# Summary

No GO Terms were significantly enriched among the DEGs impacted by
leachate exposure, either overall (all \~128/130 leachate responsive
DEGs) or within the specific gene expression patterns (down, hump, up).

We did find enriched GO Terms in the

## Next steps

-   Visualize GO terms & KEGG Pathways for each cluster with `rrvgo`
    plotting functions
-   Convert the GOseq output into a clusterProfiler-compatible object
    for visualization (dotplot, emapplot).

```{r}
sessionInfo()
```
