---
title: "07_goslim"
format: html
---

# Background

I matched clustered/module grouped DEGs to GO terms, and tested for enrichment using goseq().

I found that DEGs due to large shifts by stage had enriched GO terms only for prawnchip dip, earlygastrula up, or all stage DEGs combined. 
No other cluster/ module had enrichment.

I want to be able to show however what the DEGs for each cluster/moddule functionally represent, even if not statistically enriched.

For each cluster/module:

keep BP terms only

rank by raw p-value

take top N (e.g. 20–30)

require a minimum gene count

# Load libraries
```{r}
library(GSEABase)
library(GO.db)
library(knitr)
library(tidyverse)
```


# Inputs
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "Gene.Ontology.IDs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- "https://raw.githubusercontent.com/grace-ac/paper-pycno-sswd-2021-2022/d1cdf13c36085868df4ef4b75d2b7de03ef08d1c/analyses/25-compare-2021-2022/DEGlist_same_2021-2022_forGOslim.tab"


##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```

```{r}
GO.bp.1 <- read_csv(file.path(input_path,"GO_bp_cluster1.csv"))
```

```{r}
head(GO.bp.1) %>% knitr::kable()
```
## Semantic similarity
## Reduce redundancy

Calculate semantic similarity matrix with **all GO terms** that are present in DEGs
```{r}
# filter GO terms to those that are BP only
GO.bp.1.filt <- GO.bp.1 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

simMatrix.1 <- calculateSimMatrix(
    GO.bp.1.filt$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )
```

12599 GO terms... 

```{r}
write_csv(as.data.frame(simMatrix.1), file.path(output_path, "simMatrix_1.csv"))
```


## Scores from DEInCat

Create named scores vectors for each module/cluster

```{r}
# Create scores
scores <- setNames(
    -log10(pmax(GO.bp.1.filt$bh_adjust, .Machine$double.xmin)),
    GO.bp.deg.filt$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms <- reduceSimMatrix(
    simMatrix,
    scores = scores,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```

Add in gene and cluster annotation to reduced GO terms present in DEGs
```{r}
reducedTerms_anno <- reducedTerms %>%
  left_join(GO.bp.deg.filt, by = c("go" = "category", "term")) %>% 
  left_join(gene2go, by = c("go" = "category")) %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id")) %>% 
  filter(!is.na(cluster))
```

```{r}
write_csv(reducedTerms, file.path(output_path, "reducedTerms_GO.csv"))
write_csv(reducedTerms_anno, file.path(output_path, "reducedTerms_GO_anno.csv"))
```




## GO.db
```{r}
?select
```

```{r}
# see what GO slim objects exist in your GO.db
ls("package:GO.db", pattern = "GOSLIM")

# extract the generic GO slim IDs (this is what we will intersect with)
goslim_ids <- names(GOSLIMGENERIC)

length(goslim_ids)
head(goslim_ids)
```


```{r}
go_to_slim <- as.list(GOBPANCESTOR)

top_terms_slim <- top_terms %>%
  rowwise() %>%
  mutate(
    goslim = list(
      intersect(
        go_to_slim[[category]],
        keys(GOSLIM_GENERIC)
      )
    )
  ) %>%
  ungroup() %>%
  tidyr::unnest(goslim)
```


