---
title: "Map GO IDs to GOslims"
subtitle: " Using `GSEABase` and `GO.db`"
author: "Sarah Tanja"
date: 01/16/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

[Map GO IDs to GOslims](https://robertslab.github.io/resources/bio-Annotation/#:~:text=of%20the%20above.-,Map%20GO%20IDs%20to%20GOslims,-Below%20are%20a)


# Load libraries

```{r}
#BiocManager::install("GSEABase")
#BiocManager::install("GO.db")
```

```{r}
library(GSEABase)
library(tidyverse)
library(GO.db)
library(purrr)
library(UpSetR)
```

# Inputs
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

The expected input file has at least two columns. 
One each with:
- gene ID
- Gene Ontology (GO) ID

::: callout-note
The GO IDs in the GO ID column **should be separated with a semi-colon**.
gene to go term mapping
::: 

## GO counts
```{r}
gene2go <- read_csv(file.path(input_path, "gene2go.csv"))
gowide <- read_csv(file.path(input_path, "gowide.csv"))
```

```{r}
glimpse(gene2go)
```

```{r}
glimpse(gowide)
```

```{r}
GO_stage_up <- read_csv(file.path(input_path, "GO_stage_up.csv"))
GO_stage_hump <- read_csv(file.path(input_path, "GO_stage_hump.csv"))
GO_stage_valley <- read_csv(file.path(input_path, "GO_stage_valley.csv"))
GO_stage_down <- read_csv(file.path(input_path, "GO_stage_down.csv"))


GO_leachate_up <- read_csv(file.path(input_path, "GO_leachate_up.csv"))
GO_leachate_hump <- read_csv(file.path(input_path, "GO_leachate_hump.csv"))
GO_leachate_down <- read_csv(file.path(input_path, "GO_leachate_down.csv"))
```

```{r}
GO_counts <- gowide %>% 
  left_join(GO_stage_up %>% 
              dplyr::select(numDEInCat_stage_up = numDEInCat,
                            category, 
                            ontology),
            by = "category") %>% 
  left_join(GO_stage_hump %>% 
              dplyr::select(numDEInCat_stage_hump = numDEInCat,
                            category, 
                            ontology),
            by = c("category", "ontology")) %>%
  left_join(GO_stage_valley %>% 
              dplyr::select(numDEInCat_stage_valley = numDEInCat,
                            category, 
                            ontology),
            by = c("category", "ontology")) %>%
  left_join(GO_stage_down %>% 
              dplyr::select(numDEInCat_stage_down = numDEInCat,
                            category, 
                            ontology),
            by = c("category", "ontology")) %>%
  left_join(GO_leachate_up %>% 
              dplyr::select(numDEInCat_leachate_up = numDEInCat,
                            category, 
                            ontology),
            by = c("category", "ontology")) %>%
  left_join(GO_leachate_hump %>% 
              dplyr::select(numDEInCat_leachate_hump = numDEInCat,
                            category, 
                            ontology),
            by = c("category", "ontology")) %>%
  left_join(GO_leachate_down %>% 
              dplyr::select(numDEInCat_leachate_down = numDEInCat,
                            category, 
                            ontology),
            by = c("category", "ontology")) %>% 
  filter(ontology == "BP")
```


## Load GO annotated genes
```{r}
universe_anno <- read_csv(file.path(input_path, "universe_anno.csv"))
dim(universe_anno)
```

```{r}
glimpse(universe_anno)
```


Change comma separated GO IDs to semi-colon separated GO IDs
```{r}
gene2GOs <- universe_anno %>%
  mutate(GOs = str_replace_all(GOs, ",", ";")) %>% 
  dplyr::select(gene_id, GOs, pattern_leachate, pattern_stage)

glimpse(gene2GOs)
```

::: callout-important
The user needs to provide:
- names of the columns containing the GO IDs and the gene IDs!
- Column name for GO IDs : `GOs`
- Column name for gene IDs : `gene_id`
- Path or URL to input file.
:::

# Make a gene set for each cluster and module

## Leachate DEGs
```{r}
gene2GOs_leachate_up <- gene2GOs %>%
  filter(pattern_leachate == "up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(DEG_pattern = "leachate_up")
```

```{r}
gene2GOs_leachate_hump <- gene2GOs  %>%
  filter(pattern_leachate == "hump") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(DEG_pattern = "leachate_hump")
```

```{r}
gene2GOs_leachate_down <- gene2GOs  %>%
  filter(pattern_leachate == "down") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(DEG_pattern = "leachate_down")
```

## Stage DEGs
```{r}
gene2GOs_stage_up <- gene2GOs  %>%
  filter(pattern_stage == "up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(DEG_pattern = "stage_up")
```

```{r}
gene2GOs_stage_hump <- gene2GOs %>%
  filter(pattern_stage == "hump") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(DEG_pattern = "stage_hump")
```

```{r}
gene2GOs_stage_valley <- gene2GOs %>%
  filter(pattern_stage == "valley") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(DEG_pattern = "stage_valley")
```

```{r}
gene2GOs_stage_down <- gene2GOs %>%
  filter(pattern_stage == "down") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_down")
```

## Universe
```{r}
gene2GOs_universe <- gene2GOs%>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "universe")
```

## Save gene2GO dataframes as a named list
```{r}
gene2GOs_list <- list(
  leachate_up   = gene2GOs_leachate_up,
  leachate_hump = gene2GOs_leachate_hump,
  leachate_down = gene2GOs_leachate_down,
  stage_up      = gene2GOs_stage_up,
  stage_hump    = gene2GOs_stage_hump,
  stage_valley  = gene2GOs_stage_valley,
  stage_down    = gene2GOs_stage_down,
  universe      = gene2GOs_universe
)
```

# Outputs

```{r}
output_path <- "../../output/07_enrichment/goslim/"
```

Save input file with semi-colon separated GO IDs
```{r}
write_csv(gene2GOs_universe, file.path(output_path, "gene2GOs_universe.csv"))
```


### Set variables
```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "GOs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- file.path(output_path, "gene2GOs_universe.csv")

##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```

### Set GSEAbase location and download goslim_generic.obo
```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destintation <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destintation)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

### Clean: remove NAs

First check how many rows (aka genes) do not have GO terms annotated to them. If there are no GO terms annotated to a gene, the cell will be NA. We want to remove those rows because they will not be useful for mapping to GOslims.


The following is a function that will check how many rows in a given dataframe have NA values in the GO ID column. It returns a tibble with the total number of rows and the number of rows with NA values in the GO ID column.
```{r}
check_na_go <- function(df, GO.ID.column) {
  tibble(
    n_rows = nrow(df),
    n_na_go = sum(is.na(df[[GO.ID.column]]))
  )
}
```


Now we use the function we created above to check how many rows in each of our gene2GO dataframes have NA values in the GO ID column. We use `imap_dfr()` to apply the function to each dataframe in our list and combine the results into a single tibble with an additional column indicating the dataset name.
```{r}
na_summary <- imap_dfr(
  gene2GOs_list,
  ~ check_na_go(.x, GO.ID.column),
  .id = "dataset"
)

na_summary
```

Now we make a function that will clean the gene2GO dataframes by removing rows with NA values in the GO ID column and also removing any extra whitespace around the GO IDs. The function takes a dataframe, the name of the gene ID column, and the name of the GO ID column as arguments, and returns a cleaned dataframe with only those two columns.
```{r}
clean_gene2go <- function(df, gene.ID.column, GO.ID.column) {
  df %>%
    mutate(
      !!sym(GO.ID.column) :=
        str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")
    ) %>%
    filter(
      !is.na(.data[[gene.ID.column]]),
      !is.na(.data[[GO.ID.column]]),
      .data[[GO.ID.column]] != ""
    ) %>%
    select(all_of(c(gene.ID.column, GO.ID.column)))
}
```


We use the function we created above across our list of gene2GO dataframes to clean them. We use `imap()` to apply the function to each dataframe in our list and keep the names of the datasets in the resulting list of cleaned dataframes, which we name `gene2GOs_clean`.
```{r}
gene2GOs_clean <- imap(
  gene2GOs_list,
  ~ clean_gene2go(.x,
                  gene.ID.column = gene.ID.column,
                  GO.ID.column   = GO.ID.column)
)

```


Here we show that in our new list of dataframes, there are no NA values in the GO ID column and we also show the total number of rows for each dataframe. This is a quick check to confirm that our cleaning function worked as expected.
```{r}
imap_dfr(
  gene2GOs_clean,
  ~ tibble(
    dataset = .y,
    n_rows  = nrow(.x),
    n_na_go = sum(is.na(.x[[GO.ID.column]]))
  )
)
```

### Map 1GO to 1Gene   

This is a function that will map gene 2 GO data frame so each row is single gene ID with single GO ID
```{r}
map_gene2go <- function(df, GO.ID.column, sep = ";") {
  df %>%
    separate_rows(!!sym(GO.ID.column), sep = sep)
}
```

```{r}
gene2GO_map <- imap(
  gene2GOs_clean,   # or whatever list holds your cleaned dataframes
  ~ map_gene2go(.x, GO.ID.column = GO.ID.column)
)
```

Check that mapping worked
```{r}
imap_dfr(
  gene2GO_map,
  ~ tibble(
    dataset = .y,
    n_rows  = nrow(.x)
  )
)
```
Let's see the structure of one of the mapped dataframes
```{r}
head(gene2GO_map[[1]])
```

These dataframes just contain two columns, one with gene IDs and one with GO IDs, with one GO ID per row (and gene_id is repeated for as many GO IDs there are associated with that gene). This is a one to one dataframe! One gene to one GO term! 



### Group

Groups the genes by GO ID (i.e. lists all genes associated with each unique GO ID)

```{r}
group_gene2go <- function(df, gene.ID.column, GO.ID.column) {
      df %>%
    group_by(!!sym(GO.ID.column)) %>%
    summarise(
      !!sym(gene.ID.column) := paste(.data[[gene.ID.column]], collapse = ","),
      .groups = "drop"
    )
}
```

```{r}
GO2genes_grouped <- imap(
  gene2GO_map,
  ~ .x %>%
      dplyr::filter(
        !is.na(.data[[GO.ID.column]]),
        .data[[GO.ID.column]] != "-"
      ) %>%
      group_gene2go(
        gene.ID.column = gene.ID.column,
        GO.ID.column   = GO.ID.column
      )
)

```
What this does
- `!is.na(...)` → removes missing GO IDs
- `!= "-"` → removes placeholder dashes
- `.data[[GO.ID.column]]` → tidy-eval safe, works with column names stored as strings


Check that grouping worked
```{r}
imap_dfr(
  GO2genes_grouped,
  ~ tibble(
    dataset = .y,
    n_rows  = nrow(.x)
  )
)
```
So one row should have one unique GO ID, with all associated gene IDs listed in a comma delimited format. Let's look at one dataframe:
```{r}
head(GO2genes_grouped[[1]])
```


```{r}
dim(GO2genes_grouped[[1]])
dim(GO2genes_grouped[[2]])
dim(GO2genes_grouped[[3]])
dim(GO2genes_grouped[[4]])
dim(GO2genes_grouped[[5]])
dim(GO2genes_grouped[[6]])
dim(GO2genes_grouped[[7]])
dim(GO2genes_grouped[[8]])
```

::: callout-note
How many unique genes are associated with each unique GO term?
:::

Let's make a function that will add a column that counts the number of genes associated with each GO term.
```{r}
count_genes_per_go <- function(df, gene.ID.column) {
  df %>%
    mutate(
      DEG_count = str_count(.data[[gene.ID.column]], ",") + 1
    )
}
```

Now we can imap that function across our list
```{r}
GO2genes_grouped_counted <- imap(
  GO2genes_grouped,
  ~ count_genes_per_go(.x, gene.ID.column = gene.ID.column)
)
```

```{r}
head(GO2genes_grouped_counted[[1]])
```

::: callout-note
This is similar to the `GO_counts` dataframe we made above, but for some reason we lost the list of genes associated with each GO term in that dataframe....
:::

# Map GO IDs to GOslims
The mapping steps were derived from this [bioconductor forum](https://support.bioconductor.org/p/128407/#128408) response

## Extract GO IDs as vectors

This is a function to extract GO IDs as a vector from the grouped gene-GO dataframe
```{r}
vectorize_go_ids <- function(df, GO.ID.column) {
  df[[GO.ID.column]]
}
```


This loops through each grouped gene-GO dataframe in the list and extracts the GO IDs as a vector
```{r}
go_ids_list <- imap(
  GO2genes_grouped,
  ~ vectorize_go_ids(.x, GO.ID.column = GO.ID.column)
)
```

Each vector of GO IDs can now be used to create a `GOCollection` object in GSEABase.
```{r}
head(go_ids_list[[1]])
```

> The mapping steps were derived from this [bioconductor forum response](https://support.bioconductor.org/p/128407/#128408)


## Create OBO Collections

This extracts GOslims from GSEABase

This function creates a GOCollection object of GOslims based on the provided GO IDs. 
```{r}
extract_go_slims <- function(go_ids, gseabase_files) {
  myCollection <- GOCollection(go_ids)
  slim <- getOBOCollection(gseabase_files)
  list(
    myCollection = myCollection,
    slim       = slim
  )
}
```

Here we apply the above function across our list of GO ID vectors
```{r}
go_slims_list <- imap(
  go_ids_list,
  ~ extract_go_slims(
      go_ids = .x,
      gseabase_files = gseabase_files
    )
)
```

```{r}
str(go_slims_list[[1]])
```

GO ⟶ GO-slim
(many-to-one mapping via the ontology DAG)

Note how for our first "module" leachate_up there are 2216 total GO term ids in `$collection`, but 141 GOslims in `$slim`

A `GOCollection` object:
`$collection` → regular GO terms
These are the specific, fine-grained GO annotations associated with your genes.
- knows how to query GO.db
- understands parent/child relationships
- can be passed to functions like goSlim()


An `OBOCollection` object:
`$slim` → GO-slim terms
These are high-level, curated summary categories used to collapse many GO terms into interpretable bins.
- GO-slim terms (a subset of GO)
- Their relationships
- Subset labels (generic, yeast, plant, etc.)
📌 This object defines which high-level categories you allow.

## Get Biological Process (BP) GOslims

```{r}
get_bp_goslims <- function(x, verbose = FALSE) {
  goSlim(x$myCollection, x$slim, "BP", verbose)
}
```

```{r}
bp_goslims_list <- imap(
  go_slims_list,
  ~ get_bp_goslims(.x, verbose = FALSE)
)
```

```{r}
bp_goslims_list[[1]]
```
::: callout-important
There seem to only be 72 GOslim BP terms across all of the dataframes... note that count is 0 in some of these rows indicating those GOslim terms are not represented by the GO IDs in that respective DEG dataframe.
:::

The columns in the resulting dataframes are:
- GOslim ID (as rownames)
- Count
- Percent 
- Term

```{r}

bp_slim_list <- imap(go_slims_list, function(myCollection, nm) {
  message("Running goSlim for: ", nm)

  out <- goSlim(myCollection, slim, "BP", verbose = FALSE)

  # keep track of which group it came from
  out$group <- nm
  out
})

head(bp_slim_list[[1]])

```
Now we have added the group column



# Perform mapping of of GOIDs to GOslims

Returns:

GOslim IDs (as rownames)
GOslim terms
Counts of GO IDs matching to corresponding GOslim
Percentage of GO IDs matching to corresponding GOslim
GOIDs mapped to corresponding GOslim, in a semi-colon delimited format


```{r}
# ---- function: map GO IDs to GOslims for ONE slimdf -------------------------
map_go_to_goslims_one <- function(slimdf, myCollection, go_ids) {

  # List of GOslims and all GO IDs from `go_ids`
  gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

  # Maps `go_ids` to matching GOslims (kept if you want to inspect it)
  mapped <- lapply(gomap, intersect, ids(myCollection))

  # Append all mapped GO IDs to `slimdf` as semicolon-delimited string
  slimdf$GO.IDs <- sapply(mapped, paste, collapse = ";")

  # Remove literal "character(0)" strings (usually you'd get "" already, but keeping your intent)
  slimdf$GO.IDs[slimdf$GO.IDs == "character(0)"] <- ""

  # Add self-matching GOIDs to "GO.IDs" column, if not present
  for (go_id in go_ids) {

    if (go_id %in% rownames(slimdf)) {

      existing <- slimdf[go_id, "GO.IDs"]

      # split safely; handle "" -> character(0)
      existing_vec <- trimws(toupper(unlist(strsplit(existing, ";", fixed = TRUE))))
      existing_vec <- existing_vec[nzchar(existing_vec)]

      if (!toupper(go_id) %in% existing_vec) {

        if (nzchar(existing)) {
          slimdf[go_id, "GO.IDs"] <- paste0(existing, "; ", go_id)
        } else {
          slimdf[go_id, "GO.IDs"] <- go_id
        }
      }
    }
  }

  slimdf
}
```

```{r}
# ---- apply function to LIST of slim dfs ---------------------------------------------
bp_slim_list_mapped <- imap(bp_goslims_list, \(slimdf, nm) {
  map_go_to_goslims_one(
    slimdf        = slimdf,
    myCollection = go_slims_list[[nm]]$myCollection,  # <- S4 GOCollection
    go_ids       = go_ids_list[[nm]]                  # <- character vector
  )
})

# keep names (lapply preserves them, but this makes it explicit)
bp_slim_list_mapped <- setNames(bp_slim_list_mapped, names(bp_goslims_list))

# quick check
str(bp_slim_list_mapped, max.level = 1)

```

# Perform mapping of gene_ids to GOIDS to GOslims

Checkout Sam White's Bioconductor forum post [here](https://support.bioconductor.org/p/128430/)

## Build a `GeneSetCollection` object with the GO annotations.
Create a GO → gene mapping as a GeneSetCollection
Goal

Turn your raw annotation pairs (GO term, gene/transcript ID) into a Bioconductor object where each GO term “owns” a set of genes.

Inputs

  GO.db so you can fetch human-readable GO term names

  A 2-column GO annotation table (from a file or data.frame)

    What it must look like (conceptually):
    
      Column 1: GO IDs like "GO:0016874"
      
      Column 2: gene/transcript IDs like "Montipora_capitata_HIv3___RNAseq.g5198.t1"
      
      One row per association (i.e. if a gene is annotated to multiple GO terms, it will appear in multiple rows). 
    
    Example:
    
    ```{r}
    gene2GO_map[[8]]
    ```



- input one GO term to multiple genes
```{r}
head(GO2genes_grouped[[8]])
```

- fetch GO terms into a named vector
```{r}
head(GO2genes_grouped$universe)
```

```{r}
genesbygo_from_collapsed <- function(df,
                                     go_col = "GOs",
                                     genes_col = "gene_id",
                                     sep = ",") {
  go_ids <- as.character(df[[go_col]])

  # split the comma-separated string into character vectors
  gene_vecs <- strsplit(as.character(df[[genes_col]]), sep, fixed = TRUE)
  gene_vecs <- lapply(gene_vecs, \(x) {
    x <- trimws(x)
    x <- x[nzchar(x)]
    unique(x)
  })

  setNames(gene_vecs, go_ids)
}

```


```{r}
genesbygo_list <- lapply(GO2genes_grouped, genesbygo_from_collapsed)

```


```{r}
universe_go_ids <- unique(as.character(GO2genes_grouped$universe$GOs))

## fetch GO terms into a named vector
goterms <- AnnotationDbi::select(GO.db, keys=universe_go_ids, columns="TERM")

goterms <- setNames(goterms$TERM, goterms$GOID)
```

```{r}
get_goterms_list <- function(go_ids, GOdb = GO.db) {

  # ensure clean character vector
  go_ids <- unique(as.character(go_ids))

  # fetch GO term labels
  goterms_df <- AnnotationDbi::select(
    GOdb,
    keys    = go_ids,
    columns = "TERM",
    keytype = "GOID"
  )

  # named character vector: names = GOID, values = TERM
  goterms_vec <- setNames(goterms_df$TERM, goterms_df$GOID)

  # convert to named list (one element per GO)
  as.list(goterms_vec)
}
```


```{r}
goterms_list <- lapply(GO2genes_grouped, \(df) {
  get_goterms_list(df$GOs)
})
```



## Build `GeneSetCollection` object

Function to build a `GeneSetCollection` object from a list of GO → gene mappings and a list of GOID → GOTERM mappings. The function checks that the names of the two lists are identical, then uses `mapply()` to create a `GeneSet` for each GO term, which is then combined into a `GeneSetCollection` using `do.call()`.
```{r}
make_gsc <- function(genesbygo, goterms) {
  stopifnot(identical(names(genesbygo), names(goterms)))

  do.call(
    "GeneSetCollection",
    mapply(
      FUN = function(geneSetID, geneIDs) {
        term <- goterms[[geneSetID]]  # <- character(1) or NA

        GeneSet(
          NullIdentifier(),
          geneIds = geneIDs,
          setName = geneSetID,
          shortDescription = if (is.na(term)) "" else term
        )
      },
      geneSetID = names(genesbygo),
      geneIDs   = genesbygo,
      USE.NAMES = FALSE,
      SIMPLIFY  = FALSE
    )
  )
}

```

Now we apply the function to build a `GeneSetCollection` object for each of our datasets (leachate_up, leachate_hump, etc.) using the corresponding GO → gene mappings and GOID → GOTERM mappings.
```{r}
gsc_list <- mapply(
  FUN = function(gbg, gtm) make_gsc(gbg, gtm),
  gbg = genesbygo_list,
  gtm = goterms_list,
  SIMPLIFY = FALSE,
  USE.NAMES = TRUE
)

# example:
gsc_list$leachate_up

```


# GOslim dataframe with GOID and gene_id mapping


```{r}
parse_goids <- function(x) {
  if (is.na(x) || !nzchar(x)) return(character(0))
  x <- gsub("\\s+", "", x)               # remove spaces
  ids <- unlist(strsplit(x, ";|,", perl = TRUE))
  ids <- ids[nzchar(ids)]
  unique(ids)
}

```

```{r}

summarize_goslims <- function(bp_slim_df, gsc, keep_genes = FALSE) {

  # GOID -> gene vector lookup
  go2genes <- setNames(
    lapply(gsc, geneIds),
    names(gsc)
  )

  bp_slim_df$n_goids <- NA_integer_
  bp_slim_df$n_genes <- NA_integer_

  # OPTIONAL column (created up front)
  if (keep_genes) {
    bp_slim_df$genes <- NA_character_
  }

  for (i in seq_len(nrow(bp_slim_df))) {

    # parse GO IDs mapped to this GOslim
    goids <- parse_goids(bp_slim_df$GO.IDs[i])

    bp_slim_df$n_goids[i] <- length(goids)

    # collect genes for those GO IDs
    genes_list <- go2genes[intersect(goids, names(go2genes))]

    # UNIQUE genes within this GOslim
    genes_unique <- unique(unlist(genes_list, use.names = FALSE))

    bp_slim_df$n_genes[i] <- length(genes_unique)

    # >>> THIS IS WHERE IT GETS ADDED <<<
    if (keep_genes) {
      bp_slim_df$genes[i] <- paste(genes_unique, collapse = ";")
    }
  }

  bp_slim_df
}


```

```{r}
stopifnot(identical(names(bp_slim_list_mapped), names(gsc_list)))

goslim_summary_list <- mapply(
  FUN = summarize_goslims,
  bp_slim_df = bp_slim_list_mapped,
  gsc        = gsc_list,
  keep_genes = TRUE,
  SIMPLIFY   = FALSE,
  USE.NAMES  = TRUE
)

```


# Combine

```{r}
goslim_summary_df <- imap_dfr(
  goslim_summary_list,
  ~ mutate(.x, module = .y, .before = 1)
)
```

# Wrangle
```{r}
goslim_summary <- goslim_summary_df %>%
  mutate(
    DEdueto = case_when(
      str_detect(module, "leachate") ~ "leachate",
      str_detect(module, "stage") ~ "stage"
    ),
    pattern = case_when(
      str_detect(module, "up") ~ "up",
      str_detect(module, "hump") ~ "hump",
      str_detect(module, "valley") ~ "valley",
      str_detect(module, "down") ~ "down",
      str_detect(module, "universe") ~ "universe",
  ))
```


# Plot 

```{r}
goslim_summary %>% 
  drop_na(pattern)%>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10, with_ties = FALSE) %>%
  ungroup() %>% 
ggplot(., aes(x = n_genes, y = Term, fill = DEdueto)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.85) +
  facet_wrap(~ pattern, scales = "free_y", ncol = 1) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    title = "GOslim BP Terms: Number of Genes Mapped",
    x = "Number of genes mapped to GOslim (n_genes)",
    y = "GOslim term"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```


### Save outputs
```{r}
output_path <- "../../output/07_enrichment/goslim/"
```

```{r}
write_csv(goslim_summary, file.path(output_path, "goslim_summary.csv")) 
```


