---
title: "Map GO IDs to GOslims"
subtitle: " Using `GSEABase` and `GO.db`"
author: "Sarah Tanja"
date: 01/16/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

[Map GO IDs to GOslims](https://robertslab.github.io/resources/bio-Annotation/#:~:text=of%20the%20above.-,Map%20GO%20IDs%20to%20GOslims,-Below%20are%20a)


# Load libraries

```{r}
#BiocManager::install("GSEABase")
```

```{r}
library(GSEABase)
library(tidyverse)
library(GO.db)
library(purrr)
```

# Inputs
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

> The expected input file has at least two columns. One each with:
gene ID
Gene Ontology (GO) ID

::: callout-note
The GO IDs in the GO ID column **should be separated with a semi-colon**.
gene to go term mapping
::: 

## Universe gene set
```{r}
universe_genes_anno <- read_csv(file.path(input_path, "universe_genes_anno.csv"))

head(universe_genes_anno[ , c("gene_id", "GOs")])
```
Change comma seperated GO IDs to semi-colon separated GO IDs
```{r}
gene2GOs <- universe_genes_anno %>%
  mutate(GOs = str_replace_all(GOs, ",", ";")) %>% 
  dplyr::select(gene_id, GOs, cluster_stage, module_stage)

head(gene2GOs)
```

::: callout-important
The user needs to provide:
- names of the columns containing the GO IDs and the gene IDs!
- Column name for GO IDs : `GOs`
- Column name for gene IDs : `gene_id`
- Path or URL to input file.
:::

A gene set for each cluster and module

Leachate DEGs
```{r}
gene2GOs_leachate.cleavage <- gene2GOs %>%
  filter(cluster_stage == "cleavage_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_cleavage")
```

```{r}
gene2GOs_leachate.prawnchip <- gene2GOs  %>%
  filter(cluster_stage == "prawnchip_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_prawnchip")
```

```{r}
gene2GOs_leachate.earlygastrula <- gene2GOs  %>%
  filter(cluster_stage == "early_gastrula_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_earlygastrula")
```

Stage DEGs
```{r}
gene2GOs_stage.cleavage <- gene2GOs  %>%
  filter(module_stage == "cleavage_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_cleavage")
```

```{r}
gene2GOs_stage.prawnchipup <- gene2GOs %>%
  filter(module_stage == "prawnchip_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_prawnchip_up")
```

```{r}
gene2GOs_stage.prawnchipdip <- gene2GOs %>%
  filter(module_stage == "prawnchip_dip") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_prawnchip_dip")
```

```{r}
gene2GOs_stage.earlygastrulaup <- gene2GOs %>%
  filter(module_stage == "earlygastrula_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_earlygastrula_up")
```

Universe
```{r}
gene2GOs_universe <- gene2GOs%>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "universe")
```

Save gene2GO dataframes as a named list
```{r}
gene2GOs_list <- list(
  gene2GOs_leachate.cleavage      = gene2GOs_leachate.cleavage,
  gene2GOs_leachate.prawnchip     = gene2GOs_leachate.prawnchip,
  gene2GOs_leachate.earlygastrula = gene2GOs_leachate.earlygastrula,
  gene2GOs_stage.cleavage         = gene2GOs_stage.cleavage,
  gene2GOs_stage.prawnchipup      = gene2GOs_stage.prawnchipup,
  gene2GOs_stage.prawnchipdip     = gene2GOs_stage.prawnchipdip,
  gene2GOs_stage.earlygastrulaup  = gene2GOs_stage.earlygastrulaup,
  gene2GOs_universe               = gene2GOs_universe
)
```

Save input file with semi-colon separated GO IDs
```{r}
write_csv(universe_gene2GOs, file.path(input_path, "universe_gene2GOs.csv"))
```

### Set variables
```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "GOs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- "../../output/07_enrichment/goseq/universe_gene2GOs.csv"

##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```

### Set GSEAbase location and download goslim_generic.obo
```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destintation <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destintation)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

### Clean: remove NAs

```{r}
check_na_go <- function(df, GO.ID.column) {
  tibble(
    n_rows = nrow(df),
    n_na_go = sum(is.na(df[[GO.ID.column]]))
  )
}
```

```{r}
na_summary <- imap_dfr(
  gene2GOs_list,
  ~ check_na_go(.x, GO.ID.column),
  .id = "dataset"
)

na_summary
```

```{r read-in-gene-file, eval=TRUE}
full.gene.df <- read_csv(file = input.file)

str(full.gene.df)

# check for any NA values in GO ID column
sum(is.na(full.gene.df[[GO.ID.column]]))
```

```{r}
clean_gene2go <- function(df, gene.ID.column, GO.ID.column) {
  df %>%
    mutate(
      !!sym(GO.ID.column) :=
        str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")
    ) %>%
    filter(
      !is.na(.data[[gene.ID.column]]),
      !is.na(.data[[GO.ID.column]]),
      .data[[GO.ID.column]] != ""
    ) %>%
    select(all_of(c(gene.ID.column, GO.ID.column)))
}
```

```{r}
gene2GOs_clean <- imap(
  gene2GOs_list,
  ~ clean_gene2go(.x,
                  gene.ID.column = gene.ID.column,
                  GO.ID.column   = GO.ID.column)
)

```

```{r}
imap_dfr(
  gene2GOs_clean,
  ~ tibble(
    dataset = .y,
    n_rows  = nrow(.x),
    n_na_go = sum(is.na(.x[[GO.ID.column]]))
  )
)
```


```{r remove-NA-and-uniprotIDs, eval=TRUE}

# Clean whitespace, filter NA/empty rows, select columns, and split GO terms using column name variables
gene.GO.df <- full.gene.df %>%
  mutate(!!GO.ID.column := str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")) %>% # Clean up spaces around ";"
  filter(!is.na(.data[[gene.ID.column]]) & !is.na(.data[[GO.ID.column]]) & .data[[GO.ID.column]] != "") %>% 
  select(all_of(c(gene.ID.column, GO.ID.column)))

```

### Flatten 
Map gene 2 GO data frame so each row is single gene ID with single GO ID
```{r}
flatten_gene2go <- function(gene.GO.df, GO.ID.column, sep = ";") {
  gene.GO.df %>%
    separate_rows(!!sym(GO.ID.column), sep = sep)
}
```


```{r}
gene2GOs_flat <- imap(
  gene2GOs_clean,   # or whatever list holds your cleaned gene.GO.df objects
  ~ flatten_gene2go(.x, GO.ID.column = GO.ID.column)
)

names(gene2GOs_flat) <- paste0(names(gene2GOs_flat), "_flat")

```


This is the same as the gene2GO mapping file we did in the GO enrichment step.
```{r flatten-gene-and-GO-IDs, eval=TRUE}
flat.gene.GO.df <- gene.GO.df %>% separate_rows(!!sym(GO.ID.column), sep = ";")

str(flat.gene.GO.df)
```

### Group
Groups the genes by GO ID (i.e. lists all genes associated with each unique GO ID)

```{r}
group_gene2go <- function(flat.gene.GO.df, gene.ID.column, GO.ID.column) {
  flat.gene.GO.df %>%
    group_by(!!sym(GO.ID.column)) %>%
    summarise(
      !!sym(gene.ID.column) := paste(.data[[gene.ID.column]], collapse = ","),
      .groups = "drop"
    )
}
```

```{r}
gene2GOs_grouped <- imap(
  gene2GOs_flat,
  ~ group_gene2go(
      .x,
      gene.ID.column = gene.ID.column,
      GO.ID.column   = GO.ID.column
    )
)

names(gene2GOs_grouped) <- paste0(names(gene2GOs_grouped), "_grouped")

```


```{r group-by-GO, eval=TRUE}
grouped.gene.GO.df <- flat.gene.GO.df %>%
  group_by(!!sym(GO.ID.column)) %>%
  summarise(!!gene.ID.column := paste(.data[[gene.ID.column]], collapse = ","))

str(grouped.gene.GO.df)
```

# Map GO IDs to GOslims

## Extract GO IDs as vector
```{r}
vectorize_go_ids <- function(grouped.gene.GO.df, GO.ID.column) {
  grouped.gene.GO.df[[GO.ID.column]]
}
```

```{r}
go_ids_list <- imap(
  gene2GOs_grouped,
  ~ vectorize_go_ids(.x, GO.ID.column = GO.ID.column)
)

names(go_ids_list) <- paste0(names(go_ids_list), "_go_ids")
```


The mapping steps were derived from this [bioconductor forum response](https://support.bioconductor.org/p/128407/#128408)
```{r vectorize-GOIDs, eval=TRUE}
# Vector of GO IDs
go_ids <- grouped.gene.GO.df[[GO.ID.column]]

str(go_ids)
```

## Retrieve GOslims from GSEABase

```{r}
extract_go_slims <- function(go_ids, gseabase_files) {
  myCollection <- GOCollection(go_ids)
  slim <- getOBOCollection(gseabase_files)
  list(
    collection = myCollection,
    slim       = slim
  )
}
```

```{r}
go_slims_list <- imap(
  go_ids_list,
  ~ extract_go_slims(
      go_ids = .x,
      gseabase_files = gseabase_files
    )
)

names(go_slims_list) <- paste0(names(go_slims_list), "_goslims")

```

Creates new OBO Collection object of just GOslims, based on provided GO IDs.
```{r extract-GOslims-from-OBO, eval=TRUE}
# Create GSEAbase GOCollection using `go_ids`
myCollection <- GOCollection(go_ids)

# Retrieve GOslims from GO OBO file set
slim <- getOBOCollection(gseabase_files)

```

## Get Biological Process (BP) GOslims

```{r}
get_bp_goslims <- function(x, verbose = FALSE) {
  goSlim(x$collection, x$slim, "BP", verbose)
}
```

```{r}
bp_goslims_list <- map(go_slims_list, get_bp_goslims, verbose = FALSE)
names(bp_goslims_list) <- paste0(names(go_slims_list), "_BP")
```


```{r}
bp_goslims_list <- imap(
  go_slims_list,
  ~ get_bp_goslims(.x, ontology = "BP", verbose = FALSE)
)

names(bp_goslims_list) <- paste0(names(bp_goslims_list), "_BP")
```
```{r}
names(go_slims_list)
```
```{r}
names(go_slims_list[[1]])
```
```{r}
str(go_slims_list[[1]])
```

Are any of the dataframes missing the "slim" column?
```{r}
which(!vapply(go_slims_list, function(x) "slim" %in% names(x), logical(1)))
```


Get Biological Process (BP) GOslims associated with provided GO IDs
```{r retrieve-BP-GOslims, eval=TRUE}
# Retrieve Biological Process (BP) GOslims
slimdf <- goSlim(myCollection, slim, "BP", verbose)
str(slimdf)
```

Performs mapping of of GOIDs to GOslims

Returns:

GOslim IDs (as rownames)
GOslim terms
Counts of GO IDs matching to corresponding GOslim
Percentage of GO IDs matching to corresponding GOslim
GOIDs mapped to corresponding GOslim, in a semi-colon delimited format

```{r map-GO-to-GOslims, eval=TRUE}
# List of GOslims and all GO IDs from `go_ids`
gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

# Maps `go_ids` to matching GOslims
mapped <- lapply(gomap, intersect, ids(myCollection))

# Append all mapped GO IDs to `slimdf`
# `sapply` needed to apply paste() to create semi-colon delimited values
slimdf$GO.IDs <- sapply(lapply(gomap, intersect, ids(myCollection)), paste, collapse=";")

# Remove "character(0) string from "GO.IDs" column
slimdf$GO.IDs[slimdf$GO.IDs == "character(0)"] <- ""

# Add self-matching GOIDs to "GO.IDs" column, if not present
for (go_id in go_ids) {
  # Check if the go_id is present in the row names
  if (go_id %in% rownames(slimdf)) {
    # Check if the go_id is not present in the GO.IDs column
    # Also removes white space "trimws()" and converts all to upper case to handle
    # any weird, "invisible" formatting issues.
    if (!go_id %in% trimws(toupper(strsplit(slimdf[go_id, "GO.IDs"], ";")[[1]]))) {
      # Append the go_id to the GO.IDs column with a semi-colon separator
      if (length(slimdf$GO.IDs) > 0 && nchar(slimdf$GO.IDs[nrow(slimdf)]) > 0) {
        slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
      } else {
        slimdf[go_id, "GO.IDs"] <- go_id
      }
    }
  }
}

str(slimdf)
```
"Flatten" file so each row is single GO ID with corresponding GOslim rownames_to_column needed to retain row name info

```{r flatten-GOslims-file, eval=TRUE}
# "Flatten" file so each row is single GO ID with corresponding GOslim
# rownames_to_column needed to retain row name info
slimdf_separated <- as.data.frame(slimdf %>%
  rownames_to_column('GOslim') %>%
  separate_rows(GO.IDs, sep = ";"))

# Group by unique GO ID
grouped_slimdf <- slimdf_separated %>%
  filter(!is.na(GO.IDs) & GO.IDs != "") %>%
  group_by(GO.IDs) %>%
  summarize(GOslim = paste(GOslim, collapse = ";"),
            Term = paste(Term, collapse = ";"))


str(grouped_slimdf)
```

Sorts GOslims by Count, in descending order and then selects just the Term and Count columns.
```{r sort-and-select-slimdf-counts, eval=TRUE}
slimdf.sorted <- slimdf %>% arrange(desc(Count))

slim.count.df <- slimdf.sorted %>% 
  select(Term, Count)

str(slim.count.df)
```

```{r}
head(slim.count.df)
```
### Save outputs
```{r}
output_path <- "../../output/07_enrichment/goslim/"
```

```{r}
write_csv(slim.count.df, file.path(output_path, "gene2goslim_mapping.csv")) 
```

