---
title: "Visualize GOslims by numbers of DEGs"
subtitle: "What do they genes do!? Can we see it with geom_bar and Upset plots?"
date: 02/06/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background
We identified GOslim terms associated with DEGs in each cluster/module. This is a descriptive summary plot, and does not reflect statistical enrichment!

With this plot we simply want to know what the DE genes do, and how many of them are associated with each GOslim term.

`goslim_summary.csv` was generated in 
# Load libraries
```{r}
library(tidyverse)
```

# Load inputs
```{r}
input_path <- "../../output/07_enrichment/goslim/"
```

```{r}
goslim_summary <- read_csv(file.path(input_path, "goslim_summary.csv"))
```

```{r}
glimpse(goslim_summary)
```

```{r}
summary(goslim_summary)
```
```{r}
colSums(is.na(goslim_summary))
```

# Wrangle
remove rows that are 0 or NA
```{r, fig.height = 10}
goslim_summary %>%
  drop_na(genes) %>% 
  filter(DEdueto == "leachate") %>% 
  filter(n_genes > 2) %>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10) %>%
  ungroup() %>% 
ggplot(., aes(x = n_genes, y = Term, fill = Term)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.85) +
  facet_wrap(~ pattern, scales = "free_y", ncol = 1) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    title = "GOslim BP Terms: Number of Genes Mapped",
    x = "Number of differentially expressed genes mapped to GOslim terms",
    y = "Top 10 GOslim terms per gene expression module"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
ggsave(
  filename = "goslim_summary_barplot.png",
  plot = last_plot(),
  path = "../../output/07_enrichment/goslim/",
  width = 10,
  height = 8,
  units = "in",
  dpi = 900
)
```


# How do I show this in context of background universe and stage signals?

# How many GOSlims are there? 
```{r}
length(unique(goslim_summary$Term))
```
There are 72 unique Go Slim Terms. 


# How many terms are represented? 
```{r}
goslim_summary %>%
  filter(DEdueto == "leachate", n_genes > 0) %>%
  group_by(module) %>%
  arrange(n_genes) %>%
  ungroup() %>%
  count(module, Term) %>%
  count(module, name = "n_unique_terms")
```
There are 32 terms in the down pattern, 37 in the hump patterns, and 53 in the up pattern. 


# Top 15 terms
Which are the top ~15 terms of *leachate-responsive DEGs* represented in each pattern by percent?
```{r}
goslim_summary %>%
  filter(DEdueto == "leachate", n_genes > 2) %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 15) %>%
  ungroup() %>%
  count(module, Term) %>%
  count(module, name = "n_unique_terms")
```
There are 14 unique terms represented in the down pattern, 9 in the hump pattern, and 16 in the up pattern. 

# Which are unique? 
Across down, hump, and up.. which terms are only found in one pattern (e.g. only in down, or only in up)?
```{r}
unique_terms <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(Term) %>%
  summarise(
    n_modules = n_distinct(module),
    modules = paste(sort(unique(module)), collapse = ", "),
    patterns = paste(sort(unique(pattern)), collapse = ", "),
    mean_percent = mean(Percent, na.rm = TRUE),
    mean_n_genes = mean(n_genes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_modules == 1) %>%
  arrange(desc(mean_percent), Term) %>% 
  dplyr::select(Term, patterns, mean_percent, mean_n_genes) %>% 
  arrange(patterns, desc(mean_percent))

unique_terms

```
There are 13 unique terms

Unique to down:
- meitoc nuclear division, n_DEG = 3
- amino acid metabolic process, n_DEG = 1
- chromatin organization, n_DEG = 3

Unique to hump:
- lipid metablolic process, n_DEG = 4
- carbohydrate derivative metabolic process
- DNA-templated transcription

Unique to up:
- vesicle-mediated transport, n_DEG = 
- 

# Which are shared?
```{r}
top_terms_percent <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 10, with_ties = FALSE) %>%
  ungroup()

top_terms_genes <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10, with_ties = FALSE) %>%
  ungroup()

setdiff(top_terms_percent$Term, top_terms_genes$Term) # percent only 
setdiff(top_terms_genes$Term, top_terms_percent$Term) # gene count only
intersect(top_terms_percent$Term, top_terms_genes$Term) # shared between both
```

Do I want top 10 by percent, or top 10 by gene count? ... I guess both, and then see how much overlap there is between them.

How do I average top terms across all patterns? 

```{r}
overlap_terms <- top_terms_percent %>%
  group_by(Term) %>%
  summarise(
    n_modules = n_distinct(module),
    modules = paste(sort(unique(module)), collapse = ", "),
    patterns = paste(sort(unique(pattern)), collapse = ", "),
    mean_percent = mean(Percent, na.rm = TRUE),
    mean_n_genes = mean(n_genes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_modules > 1) %>%
  arrange(desc(mean_n_genes), desc(mean_percent))

overlap_terms
```

```{r}
overlap_detail <- top_terms_percent %>%
  semi_join(overlap_terms, by = "Term") %>%
  arrange(Term, module, desc(Percent)) %>%
  select(Term, module, pattern, Percent, n_genes, Count, GO.IDs)

overlap_detail
```

# Make plotting dataframe
```{r, fig.height = 12}
gos_plot <- goslim_summary %>%
  drop_na(genes) %>%
  filter(DEdueto == "leachate", n_genes > 2)

# top 10 terms by number of genes across all patterns
term_levels <- gos_plot %>%
  group_by(Term) %>%
  summarise(total_n_genes = sum(n_genes, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_n_genes), Term) %>%
  slice_head(n = 10) %>%
  pull(Term)

circ_df <- gos_plot %>%
  filter(Term %in% term_levels) %>%
  group_by(pattern, Term) %>%
  summarise(n_genes = sum(n_genes),
            Percent = max(Percent, na.rm = TRUE),
            .groups = "drop") %>%
  complete(pattern, Term = term_levels, 
           fill = list(n_genes = 0, Percent = 0)) %>%
  mutate(
    pattern = factor(pattern, levels = c("down", "hump", "up")),
    Term = factor(Term, levels = term_levels),
    Term_wrap = str_wrap(as.character(Term), width = 12),
    percent_scaled = Percent / max(Percent) * max(n_genes)
  )
```

# N genes wedge
```{r}
ngenes_wedge <- ggplot(circ_df, aes(x = factor(Term_wrap, levels = unique(Term_wrap)), y = n_genes, fill = Term)) +
  geom_col(width = 0.8, alpha = 0.9) +
  coord_polar(start = -pi / 2) +
  facet_wrap(~pattern, nrow = 1) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Bar length reflects number of DE genes mapped to term",
    x = NULL,
    y = "Number of DE genes mapped to term"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )

ngenes_wedge
```

# % enriched wedge

```{r}
ggplot(
  circ_df,
  aes(
    x = factor(Term_wrap, levels = unique(Term_wrap)),
    y = Percent,
    fill = Term
  )
) +
  geom_col(width = 0.95, alpha = 0.9) +
  coord_polar(start = -pi / 2) +
  facet_wrap(~pattern, nrow = 1) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Bar length reflects percent representation",
    x = NULL,
    y = "Percent"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )
```

```{r}
ggplot(
  circ_df,
  aes(
    x = factor(Term_wrap, levels = unique(Term_wrap)),
    y = Percent,
    fill = Term
  )
) +
  geom_col(width = 0.95, alpha = 0.9) +
  coord_polar(start = -pi / 2) +
  facet_wrap(~pattern, nrow = 1) +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Bar length reflects percent representation",
    x = NULL,
    y = "Percent of genes mapped to GOslim term"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    legend.position = "none"
  )
```

# Spoke plot
```{r}
ngenes_spoke <- ggplot(
  circ_df,
  aes(
    x = factor(Term_wrap, levels = unique(Term_wrap)),
    y = 0,
    xend = factor(Term_wrap, levels = unique(Term_wrap)),
    yend = n_genes,
    color = Term
  )
) +
  geom_segment(linewidth = 1, alpha = 0.9) +
  coord_polar(start = -pi / 2) +
  facet_wrap(~pattern, nrow = 1) +
  scale_y_continuous(
    limits = c(0, 21),
    breaks = 1:20,
    expand = c(0, 0)
  ) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Spoke length reflects number of DE genes mapped to term",
    x = NULL,
    y = "Number of DE genes mapped to term"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    #axis.text.x = element_blank(),
    legend.position = "none"
  )

ngenes_spoke
```
# Spider / Radar
 
 inspo from R graph gallery [here](https://r-graph-gallery.com/spider-or-radar-chart.html)
 
```{r, fig.height = 4, fig.width = 12}
plt <- ggplot(circ_df) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:3) * 5),
    color = "lightgrey"
  ) + 
  # Add bars to represent the number of genes mapped to each GO term
  # str_wrap(Term, 12) wraps the text so each line has at most 12 characters
  # (but it doesn't break long words!)
  geom_col(
    aes(
      x = str_wrap(Term, 12),
      y = n_genes,
      fill = Term
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = str_wrap(Term, 12),
      y = 0,
      xend = str_wrap(Term, 12),
      yend = 21,
      color = Term
    ),
    linetype = "dashed"
  ) + 
  # Add dots to represent the percent enriched
  #geom_point(
  #  aes(
  #    x = str_wrap(Term, 12),
  #    y = percent_scaled
  #  ),
  #  size = 1.5, 
  #  color = "grey30"
  #) +
  facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0),
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 0.75, 0.2, 0.5, "cm")
  )
  
  
plt  
```
```{r}

plt <- plt +
# Annotate the bars and the lollipops so the reader understands the scaling
  annotate(
    x = 11, 
    y = 10,
    label = "Percent Enriched",
    geom = "text",
    angle = -67.5,
    color = "gray12",
    size = 2.5,
    family = "Bell MT"
  ) +
  annotate(
    x = 11, 
    y = 30,
    label = "Count of DE Genes",
    geom = "text",
    angle = 23,
    color = "gray12",
    size = 2.5,
    family = "Bell MT"
  ) +
  # Annotate custom scale inside plot
  annotate(
    x = 11.7, 
    y = 12, 
    label = "5", 
    geom = "text", 
    color = "gray12", 
    family = "Bell MT"
  ) +
  annotate(
    x = 11.7, 
    y = 12, 
    label = "10", 
    geom = "text", 
    color = "gray12", 
    family = "Bell MT"
  ) +
  annotate(
    x = 11.7, 
    y = 12, 
    label = "15", 
    geom = "text", 
    color = "gray12", 
    family = "Bell MT"
  ) +
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0),
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
  ) + 
  theme(
    # Remove axis ticks and text
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    # Use gray text for the region names
    axis.text.x = element_text(color = "grey40", size = 12),
  )

plt
```



# Faceted lollipop
```{r}
ggplot(circ_df, aes(x = Percent, y = Term, color = Term)) +
  geom_segment(aes(xend = 0, yend = Term), size = 3) +
  geom_point(size = 4) +
  facet_wrap(~pattern, ncol = 3) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Lollipop length reflects percent representation",
    x = "Percent of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

