---
title: "Map GO IDs to GOslims"
subtitle: " Using `GSEABase` and `GO.db`"
author: "Sarah Tanja"
date: 01/16/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

[Map GO IDs to GOslims](https://robertslab.github.io/resources/bio-Annotation/#:~:text=of%20the%20above.-,Map%20GO%20IDs%20to%20GOslims,-Below%20are%20a)


# Load libraries
```{r}
library(GSEABase)
library(tidyverse)
library(GO.db)
```

# Inputs
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

gene to go term mapping
```{r}
gene2go <- read_csv(file.path(input_path, "gene2go.csv"))
```

```{r}
GO.bp.1 <- read_csv(file.path(input_path,"GO_bp_cluster1.csv"))
```

```{r}
head(GO.bp.1) %>% knitr::kable()
```
## Semantic similarity
## Reduce redundancy

Calculate semantic similarity matrix with **all GO terms** that are present in universe

```{r}
# filter GO terms to those that are BP only
GO.bp.1.filt <- GO.bp.1 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))
```

12599 GO BP terms... 

```{r}
simMatrix_12599BP <- calculateSimMatrix(
    GO.bp.1.filt$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )
```
```{r}
identical(simMatrix, simMatrix.1)
```



```{r}
write_csv(as.data.frame(simMatrix.1), file.path(output_path, "simMatrix_1.csv"))
```


## leachate, cluster 1 (earlygastrula)
Create named scores vector
```{r}
# Create scores
scores.1 <- setNames(
    -log10(pmax(GO.bp.1.filt$bh_adjust, .Machine$double.xmin)),
    GO.bp.1.filt$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.1 <- reduceSimMatrix(
    simMatrix,
    scores = scores.1,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Add in gene and cluster annotation to reduced GO terms 
```{r}
reducedTerms.1_anno <- reducedTerms.1 %>%
  left_join(GO.bp.1.filt, by = c("go" = "category", "term")) %>% 
  left_join(gene2go, by = c("go" = "category")) %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id"))
```

```{r}
write_csv(reducedTerms, file.path(output_path, "reducedTerms_GO.csv"))
write_csv(reducedTerms_anno, file.path(output_path, "reducedTerms_GO_anno.csv"))
```

Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.1,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```

## leachate, cluster 2 (prawnchip)

```{r}
# filter GO terms to those that are BP only
GO.BP.2 <- GO.bp.2 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.2 <- setNames(
    -log10(pmax(GO.BP.2$bh_adjust, .Machine$double.xmin)),
    GO.BP.2$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.2 <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.2,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```

Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.2,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```
## leachate, cluster 3 (cleavage)

```{r}
# filter GO terms to those that are BP only
GO.BP.3 <- GO.bp.3 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.3 <- setNames(
    -log10(pmax(GO.BP.3$bh_adjust, .Machine$double.xmin)),
    GO.BP.3$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.3 <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.3,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Add in gene and cluster annotation to reduced GO terms 
```{r}
reducedTerms.1_anno <- reducedTerms.1 %>%
  left_join(GO.bp.1.filt, by = c("go" = "category", "term")) %>% 
  left_join(gene2go, by = c("go" = "category")) %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id"))
```

```{r}
write_csv(reducedTerms, file.path(output_path, "reducedTerms_GO.csv"))
write_csv(reducedTerms_anno, file.path(output_path, "reducedTerms_GO_anno.csv"))
```

Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.3,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```
## stage, cleavage

```{r}
# filter GO terms to those that are BP only
GO.BP.cleavage <- GO.bp.cleavage %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.cleavage <- setNames(
    -log10(pmax(GO.BP.cleavage$bh_adjust, .Machine$double.xmin)),
    GO.BP.cleavage$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.cleavage <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.cleavage,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.cleavage,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```

## stage, prawnchip up 

```{r}
# filter GO terms to those that are BP only
GO.BP.prawnchip_up <- GO.bp.prawnchip_up %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.prawnchip_up <- setNames(
    -log10(pmax(GO.BP.prawnchip_up$bh_adjust, .Machine$double.xmin)),
    GO.BP.prawnchip_up$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.prawnchip_up <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.prawnchip_up,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.prawnchip_up,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```
## stage, prawnchip down 

```{r}
# filter GO terms to those that are BP only
GO.BP.prawnchip_dip <- GO.bp.prawnchip_dip %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.prawnchip_dip <- setNames(
    -log10(pmax(GO.BP.prawnchip_dip$bh_adjust, .Machine$double.xmin)),
    GO.BP.prawnchip_dip$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.prawnchip_dip <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.prawnchip_dip,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.prawnchip_dip,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```

## stage, earlygastrula 

```{r}
# filter GO terms to those that are BP only
GO.BP.earlygastrula <- GO.bp.earlygastrula %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.earlygastrula <- setNames(
    -log10(pmax(GO.BP.earlygastrula$bh_adjust, .Machine$double.xmin)),
    GO.BP.earlygastrula$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.earlygastrula <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.earlygastrula,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.earlygastrula,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```


# Background universe
```{r}
# Distance from similarity matrix
dist <- as.dist(1 - simMatrix.1)

# PCoA to 2 dimensions
coord <- cmdscale(dist, k = 2)

# Convert to dataframe with GO terms
df_coords <- data.frame(go = rownames(simMatrix.1), Dim1 = coord[,1], Dim2 = coord[,2])
```

