---
title: "Map GO IDs to GOslims"
subtitle: " Using `GSEABase` and `GO.db`"
author: "Sarah Tanja"
date: 01/16/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

[Map GO IDs to GOslims](https://robertslab.github.io/resources/bio-Annotation/#:~:text=of%20the%20above.-,Map%20GO%20IDs%20to%20GOslims,-Below%20are%20a)


# Load libraries

```{r}
#BiocManager::install("GSEABase")
#BiocManager::install("GO.db")
```

```{r}
library(GSEABase)
library(tidyverse)
library(GO.db)
library(purrr)
library(UpSetR)
```

# Inputs
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

> The expected input file has at least two columns. One each with:
gene ID
Gene Ontology (GO) ID

::: callout-note
The GO IDs in the GO ID column **should be separated with a semi-colon**.
gene to go term mapping
::: 

## Universe gene set
```{r}
universe_genes_anno <- read_csv(file.path(input_path, "universe_genes_anno.csv"))

head(universe_genes_anno[ , c("gene_id", "GOs")])
```

Change comma seperated GO IDs to semi-colon separated GO IDs
```{r}
gene2GOs <- universe_genes_anno %>%
  mutate(GOs = str_replace_all(GOs, ",", ";")) %>% 
  dplyr::select(gene_id, GOs, cluster_stage, module_stage)

head(gene2GOs)
```

::: callout-important
The user needs to provide:
- names of the columns containing the GO IDs and the gene IDs!
- Column name for GO IDs : `GOs`
- Column name for gene IDs : `gene_id`
- Path or URL to input file.
:::

A gene set for each cluster and module

Leachate DEGs
```{r}
gene2GOs_leachate.cleavage <- gene2GOs %>%
  filter(cluster_stage == "cleavage_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_cleavage")
```

```{r}
gene2GOs_leachate.prawnchip <- gene2GOs  %>%
  filter(cluster_stage == "prawnchip_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_prawnchip")
```

```{r}
gene2GOs_leachate.earlygastrula <- gene2GOs  %>%
  filter(cluster_stage == "early_gastrula_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_earlygastrula")
```

Stage DEGs
```{r}
gene2GOs_stage.cleavage <- gene2GOs  %>%
  filter(module_stage == "cleavage_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_cleavage")
```

```{r}
gene2GOs_stage.prawnchipup <- gene2GOs %>%
  filter(module_stage == "prawnchip_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_prawnchip_up")
```

```{r}
gene2GOs_stage.prawnchipdip <- gene2GOs %>%
  filter(module_stage == "prawnchip_dip") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_prawnchip_dip")
```

```{r}
gene2GOs_stage.earlygastrulaup <- gene2GOs %>%
  filter(module_stage == "earlygastrula_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_earlygastrula_up")
```

Universe
```{r}
gene2GOs_universe <- gene2GOs%>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "universe")
```

Save gene2GO dataframes as a named list
```{r}
gene2GOs_list <- list(
  gene2GOs_leachate.cleavage      = gene2GOs_leachate.cleavage,
  gene2GOs_leachate.prawnchip     = gene2GOs_leachate.prawnchip,
  gene2GOs_leachate.earlygastrula = gene2GOs_leachate.earlygastrula,
  gene2GOs_stage.cleavage         = gene2GOs_stage.cleavage,
  gene2GOs_stage.prawnchipup      = gene2GOs_stage.prawnchipup,
  gene2GOs_stage.prawnchipdip     = gene2GOs_stage.prawnchipdip,
  gene2GOs_stage.earlygastrulaup  = gene2GOs_stage.earlygastrulaup,
  gene2GOs_universe               = gene2GOs_universe
)
```

# Outputs

```{r}
output_path <- "../../output/07_enrichment/goslim/"
```

Save input file with semi-colon separated GO IDs
```{r}
write_csv(gene2GOs_universe, file.path(output_path, "gene2GOs_universe.csv"))
```

### Set variables
```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "GOs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- "../../output/07_enrichment/goseq/universe_gene2GOs.csv"

##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```

### Set GSEAbase location and download goslim_generic.obo
```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destintation <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destintation)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

### Clean: remove NAs

```{r}
check_na_go <- function(df, GO.ID.column) {
  tibble(
    n_rows = nrow(df),
    n_na_go = sum(is.na(df[[GO.ID.column]]))
  )
}
```

```{r}
na_summary <- imap_dfr(
  gene2GOs_list,
  ~ check_na_go(.x, GO.ID.column),
  .id = "dataset"
)

na_summary
```

```{r read-in-gene-file, eval=TRUE}
full.gene.df <- read_csv(file = input.file)

str(full.gene.df)

# check for any NA values in GO ID column
sum(is.na(full.gene.df[[GO.ID.column]]))
```

```{r}
clean_gene2go <- function(df, gene.ID.column, GO.ID.column) {
  df %>%
    mutate(
      !!sym(GO.ID.column) :=
        str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")
    ) %>%
    filter(
      !is.na(.data[[gene.ID.column]]),
      !is.na(.data[[GO.ID.column]]),
      .data[[GO.ID.column]] != ""
    ) %>%
    select(all_of(c(gene.ID.column, GO.ID.column)))
}
```

```{r}
gene2GOs_clean <- imap(
  gene2GOs_list,
  ~ clean_gene2go(.x,
                  gene.ID.column = gene.ID.column,
                  GO.ID.column   = GO.ID.column)
)

```

```{r}
imap_dfr(
  gene2GOs_clean,
  ~ tibble(
    dataset = .y,
    n_rows  = nrow(.x),
    n_na_go = sum(is.na(.x[[GO.ID.column]]))
  )
)
```


```{r remove-NA-and-uniprotIDs, eval=TRUE}

# Clean whitespace, filter NA/empty rows, select columns, and split GO terms using column name variables
gene.GO.df <- full.gene.df %>%
  mutate(!!GO.ID.column := str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")) %>% # Clean up spaces around ";"
  filter(!is.na(.data[[gene.ID.column]]) & !is.na(.data[[GO.ID.column]]) & .data[[GO.ID.column]] != "") %>% 
  select(all_of(c(gene.ID.column, GO.ID.column)))

```

### Flatten 
Map gene 2 GO data frame so each row is single gene ID with single GO ID
```{r}
flatten_gene2go <- function(gene.GO.df, GO.ID.column, sep = ";") {
  gene.GO.df %>%
    separate_rows(!!sym(GO.ID.column), sep = sep)
}
```


```{r}
gene2GOs_flat <- imap(
  gene2GOs_clean,   # or whatever list holds your cleaned gene.GO.df objects
  ~ flatten_gene2go(.x, GO.ID.column = GO.ID.column)
)

names(gene2GOs_flat) <- paste0(names(gene2GOs_flat), "_flat")

```


This is the same as the gene2GO mapping file we did in the GO enrichment step.
```{r flatten-gene-and-GO-IDs, eval=TRUE}
#flat.gene.GO.df <- gene.GO.df %>% separate_rows(!!sym(GO.ID.column), sep = ";")

#str(flat.gene.GO.df)
```

### Group
Groups the genes by GO ID (i.e. lists all genes associated with each unique GO ID)

```{r}
group_gene2go <- function(flat.gene.GO.df, gene.ID.column, GO.ID.column) {
  flat.gene.GO.df %>%
    group_by(!!sym(GO.ID.column)) %>%
    summarise(
      !!sym(gene.ID.column) := paste(.data[[gene.ID.column]], collapse = ","),
      .groups = "drop"
    )
}
```

```{r}
gene2GOs_grouped <- imap(
  gene2GOs_flat,
  ~ group_gene2go(
      .x,
      gene.ID.column = gene.ID.column,
      GO.ID.column   = GO.ID.column
    )
)

names(gene2GOs_grouped) <- paste0(names(gene2GOs_grouped), "_grouped")

```


```{r group-by-GO, eval=TRUE}
#grouped.gene.GO.df <- flat.gene.GO.df %>%
#  group_by(!!sym(GO.ID.column)) %>%
#  summarise(!!gene.ID.column := paste(.data[[gene.ID.column]], collapse = ","))

#str(grouped.gene.GO.df)
```

# Map GO IDs to GOslims

## Extract GO IDs as vector
```{r}
vectorize_go_ids <- function(grouped.gene.GO.df, GO.ID.column) {
  grouped.gene.GO.df[[GO.ID.column]]
}
```

```{r}
go_ids_list <- imap(
  gene2GOs_grouped,
  ~ vectorize_go_ids(.x, GO.ID.column = GO.ID.column)
)

names(go_ids_list) <- paste0(names(go_ids_list), "_go_ids")
```


The mapping steps were derived from this [bioconductor forum response](https://support.bioconductor.org/p/128407/#128408)
```{r vectorize-GOIDs, eval=TRUE}
# Vector of GO IDs
#go_ids <- grouped.gene.GO.df[[GO.ID.column]]

#str(go_ids)
```

## Retrieve GOslims from GSEABase

```{r}
extract_go_slims <- function(go_ids, gseabase_files) {
  myCollection <- GOCollection(go_ids)
  slim <- getOBOCollection(gseabase_files)
  list(
    collection = myCollection,
    slim       = slim
  )
}
```

```{r}
go_slims_list <- imap(
  go_ids_list,
  ~ extract_go_slims(
      go_ids = .x,
      gseabase_files = gseabase_files
    )
)

names(go_slims_list) <- paste0(names(go_slims_list), "_goslims")

```

Creates new OBO Collection object of just GOslims, based on provided GO IDs.

```{r}
# 1) Load the GOslim OBO collection ONCE
slim <- getOBOCollection(gseabase_files)

# helper: clean GO IDs
clean_go_ids <- function(x) {
  x <- unique(na.omit(as.character(x)))
  x[grepl("^GO:\\d{7}$", x)]
}
```

```{r}
# 2) For each element of go_ids_list, make a GOCollection
go_collections <- map(go_ids_list, ~ GOCollection(clean_go_ids(.x)))

# nice names if you want them
names(go_collections) <- names(go_ids_list)
```


```{r extract-GOslims-from-OBO, eval=TRUE}
# Create GSEAbase GOCollection using `go_ids`
#myCollection <- GOCollection(go_ids)

# Retrieve GOslims from GO OBO file set
#slim <- getOBOCollection(gseabase_files)

```

## Get Biological Process (BP) GOslims

```{r}
get_bp_goslims <- function(x, verbose = FALSE) {
  goSlim(x$collection, x$slim, "BP", verbose)
}
```

```{r}
bp_goslims_list <- map(go_slims_list, get_bp_goslims, verbose = FALSE)
names(bp_goslims_list) <- paste0(names(go_slims_list), "_BP")
```


```{r}
bp_goslims_list <- imap(
  go_slims_list,
  ~ get_bp_goslims(.x, ontology = "BP", verbose = FALSE)
)

names(bp_goslims_list) <- paste0(names(bp_goslims_list), "_BP")
```
```{r}
names(go_slims_list)
```

```{r}
names(go_slims_list[[1]])
```
```{r}
names(bp_goslims_list)
```

```{r}
bp_goslims_list[[1]]
```

Are any of the dataframes missing the "slim" column?
```{r}
which(!vapply(go_slims_list, function(x) "slim" %in% names(x), logical(1)))
```


Get Biological Process (BP) GOslims associated with provided GO IDs

```{r}
# assumes you already created:
# slim <- getOBOCollection(gseabase_files)
# go_collections <- map(go_ids_list, ~ GOCollection(clean_go_ids(.x)))

bp_slim_list <- imap(go_collections, function(myCollection, nm) {
  message("Running goSlim for: ", nm)

  out <- goSlim(myCollection, slim, "BP", verbose = FALSE)

  # keep track of which group it came from
  out$group <- nm
  out
})

str(bp_slim_list[[1]])

```

```{r retrieve-BP-GOslims, eval=TRUE}
# Retrieve Biological Process (BP) GOslims
#slimdf <- goSlim(myCollection, slim, "BP", verbose)
#str(slimdf)
```

Performs mapping of of GOIDs to GOslims

Returns:

GOslim IDs (as rownames)
GOslim terms
Counts of GO IDs matching to corresponding GOslim
Percentage of GO IDs matching to corresponding GOslim
GOIDs mapped to corresponding GOslim, in a semi-colon delimited format

```{r}
clean_go_ids <- function(x) {
  x <- unique(na.omit(as.character(x)))
  x[grepl("^GO:\\d{7}$", x)]
}

map_go_to_goslims <- function(go_ids, slim, ontology = "BP", verbose = FALSE) {
  go_ids <- clean_go_ids(go_ids)

  # Build GOCollection for this group's GO IDs
  myCollection <- GOCollection(go_ids)

  # Get GOslim summary table (rownames are GOslim IDs)
  slimdf <- goSlim(myCollection, slim, ontology, verbose = verbose)

  # --- Map GO IDs -> GOslims (semi-colon delimited) --------------------------
  # For each GOslim term, get its "offspring" terms from GO.db, then intersect with this group's GO IDs
  gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

  mapped_ids <- lapply(gomap, intersect, ids(myCollection))

  slimdf$GO.IDs <- vapply(
    mapped_ids,
    function(x) paste(x, collapse = "; "),
    FUN.VALUE = character(1)
  )

  # remove literal "character(0)" (usually shouldn’t happen with vapply, but keep it safe)
  slimdf$GO.IDs[slimdf$GO.IDs %in% c("character(0)", "")] <- ""

  # Add self-matching GOIDs to its own GOslim row, if applicable
  for (go_id in go_ids) {
    if (go_id %in% rownames(slimdf)) {
      existing <- trimws(toupper(unlist(strsplit(slimdf[go_id, "GO.IDs"], ";"))))
      existing <- existing[nzchar(existing)]

      if (!toupper(go_id) %in% existing) {
        if (nzchar(slimdf[go_id, "GO.IDs"])) {
          slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
        } else {
          slimdf[go_id, "GO.IDs"] <- go_id
        }
      }
    }
  }

  slimdf
}

```

```{r}
bp_slim_mapped_list <- imap(go_ids_list, ~ {
  message("Mapping: ", .y)
  map_go_to_goslims(.x, slim = slim, ontology = "BP", verbose = FALSE)
})

str(bp_slim_mapped_list[[1]])

```

```{r}
bp_slim_mapped_df <- imap_dfr(bp_slim_mapped_list, ~ {
  out <- .x
  out$group <- .y
  out$GOslimID <- rownames(out)
  out
})

dplyr::glimpse(bp_slim_mapped_df)

```

```{r}
# Do your GO IDs look valid?
length(go_ids_list[[1]])
sum(grepl("^GO:\\d{7}$", go_ids_list[[1]]))

# Did we actually map anything?
table(nzchar(bp_slim_mapped_list[[1]]$GO.IDs))

```


```{r map-GO-to-GOslims, eval=TRUE}
# List of GOslims and all GO IDs from `go_ids`
#gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

# Maps `go_ids` to matching GOslims
#mapped <- lapply(gomap, intersect, ids(myCollection))

# Append all mapped GO IDs to `slimdf`
# `sapply` needed to apply paste() to create semi-colon delimited values
#slimdf$GO.IDs <- sapply(lapply(gomap, intersect, ids(myCollection)), paste, collapse=";")

# Remove "character(0) string from "GO.IDs" column
#slimdf$GO.IDs[slimdf$GO.IDs == "character(0)"] <- ""

# Add self-matching GOIDs to "GO.IDs" column, if not present
#for (go_id in go_ids) {
  # Check if the go_id is present in the row names
 # if (go_id %in% rownames(slimdf)) {
    # Check if the go_id is not present in the GO.IDs column
    # Also removes white space "trimws()" and converts all to upper case to handle
    # any weird, "invisible" formatting issues.
#    if (!go_id %in% trimws(toupper(strsplit(slimdf[go_id, "GO.IDs"], ";")[[1]]))) {
      # Append the go_id to the GO.IDs column with a semi-colon separator
#      if (length(slimdf$GO.IDs) > 0 && nchar(slimdf$GO.IDs[nrow(slimdf)]) > 0) {
#        slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
#      } else {
 #       slimdf[go_id, "GO.IDs"] <- go_id
#      }
#    }
#  }
#}

#str(slimdf)
```
"Flatten" file so each row is single GO ID with corresponding GOslim rownames_to_column needed to retain row name info

```{r flatten-GOslims-file, eval=TRUE}
# "Flatten" file so each row is single GO ID with corresponding GOslim
# rownames_to_column needed to retain row name info
slimdf_separated <- as.data.frame(slimdf %>%
  rownames_to_column('GOslim') %>%
  separate_rows(GO.IDs, sep = ";"))

# Group by unique GO ID
grouped_slimdf <- slimdf_separated %>%
  filter(!is.na(GO.IDs) & GO.IDs != "") %>%
  group_by(GO.IDs) %>%
  summarize(GOslim = paste(GOslim, collapse = ";"),
            Term = paste(Term, collapse = ";"))


str(grouped_slimdf)
```


```{r}
flatten_goslim_df <- function(slimdf) {
  slimdf %>%
    rownames_to_column("GOslim") %>%
    separate_rows(GO.IDs, sep = ";") %>%
    mutate(GO.IDs = trimws(GO.IDs)) %>%
    filter(!is.na(GO.IDs), GO.IDs != "") %>%
    group_by(GO.IDs) %>%
    summarize(
      GOslim = paste(unique(GOslim), collapse = "; "),
      Term   = paste(unique(Term),   collapse = "; "),
      .groups = "drop"
    )
}

```

```{r}
grouped_slimdf_list <- map(bp_slim_mapped_list, flatten_goslim_df)

str(grouped_slimdf_list[[1]])

```
```{r}
grouped_slimdf_all <- imap_dfr(bp_slim_mapped_list, ~ {
  flatten_goslim_df(.x) %>%
    mutate(group = .y, .before = 1)
})

dplyr::glimpse(grouped_slimdf_all)

```


Sorts GOslims by Count, in descending order and then selects just the Term and Count columns.
```{r sort-and-select-slimdf-counts, eval=TRUE}
slimdf.sorted <- slimdf %>% arrange(desc(Count))

slim.count.df <- slimdf.sorted %>% 
  select(Term, Count)

str(slim.count.df)
```

```{r}
head(slim.count.df)
```
### Save outputs
```{r}
output_path <- "../../output/07_enrichment/goslim/"
```

```{r}
write_csv(slim.count.df, file.path(output_path, "gene2goslim_mapping.csv")) 
```

# Data Visualization

## Horizontal Bar Chart of Top GOslim Terms

Create a horizontal bar chart showing the top GOslim terms by number of associated genes.

```{r visualize-top-goslims, eval=TRUE, fig.width=10, fig.height=8}
# Select top 20 GOslim terms by count
top_goslims <- bp_slim_mapped_df %>%
  slice_max(Count, n = 50) %>%
  mutate(Term = fct_reorder(Term, Count))

# Create horizontal bar chart
ggplot(top_goslims, aes(x = Count, y = Term, fill = group)) +
  geom_col(alpha = 0.7) +
  labs(
    title = "Top 20 GOslim Terms by Number of Associated Genes",
    x = "Number of Genes",
    y = "GOslim Term"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```
```{r, fig.width=10, fig.height=8}
top_goslims <- bp_slim_mapped_df %>%
  slice_max(Count, n = 30) %>%
  mutate(Term = fct_reorder(Term, Count))

ggplot(top_goslims, aes(x = Count, y = Term, fill = group)) +
  geom_col(alpha = 0.7, position = "dodge") +   # <- grouped bars
  scale_x_continuous(expand = c(0, 0)) +        # <- bars start at 0
  #coord_flip() +                                # <- horizontal
  labs(
    title = "Top GOslim Terms by Number of Associated Genes",
    x = "Number of Genes",
    y = "GOslim Term"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(size = 10),
    axis.text.x = element_text(size = 10),
    axis.title  = element_text(size = 12, face = "bold"),
    plot.title  = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )

```
```{r}
library(dplyr)
library(forcats)

top_n <- 6

top_by_group <- bp_slim_mapped_df %>%
  group_by(group) %>%
  slice_max(order_by = Count, n = top_n, with_ties = FALSE) %>%
  ungroup()

```

```{r, fig.width=18}
top_by_group %>%
  group_by(group) %>%
  mutate(Term = fct_reorder(Term, Count)) %>%
  ungroup() %>%
  ggplot(aes(x = Count, y = Term)) +
  geom_col(alpha = 0.8) +
  scale_x_continuous(expand = c(0, 0)) +
  #coord_flip() +
  facet_wrap(~ group, scales = "free_y", ncol = 1) +
  labs(
    title = "Top GOslim BP Terms by Group",
    x = "Number of genes (Count)",
    y = "GOslim term"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())

```
```{r}
terms_to_compare <- top_by_group %>%
  distinct(Term) %>%
  pull(Term)

compare_df <- bp_slim_mapped_df %>%
  filter(Term %in% terms_to_compare) %>%
  group_by(Term) %>%
  mutate(max_count = max(Count, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(Term = fct_reorder(Term, max_count))

```

```{r, fig.width=12}
ggplot(compare_df, aes(x = Count, y = Term, fill = group)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.85) +
  scale_x_continuous(expand = c(0, 0)) +
  #coord_flip() +
  labs(
    title = "GOslim BP Terms: Counts Compared Across Groups",
    x = "Number of genes (Count)",
    y = "GOslim term"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())

```

This visualization helps identify the most prevalent biological processes represented in the GOslim mapping.

## UpSet Plots: Unique and Shared GOslim Terms

UpSet plots provide a powerful visualization to show which GOslim terms are unique to specific groups and which are shared across multiple groups.

### Prepare data for UpSet plots

```{r}
# Create a list of GOslim terms for each group
# Each group will contain the Term names that are present in that group

# Check if bp_slim_mapped_df has data
if (nrow(bp_slim_mapped_df) > 0) {
  # Extract unique terms for each group
  goslim_term_lists <- bp_slim_mapped_df %>%
    filter(Count > 0) %>%  # Only include terms with at least one gene
    group_by(group) %>%
    summarise(terms = list(unique(Term)), .groups = "drop") %>%
    deframe()
  
  # Print summary
  cat("Number of terms per group:\n")
  iwalk(goslim_term_lists, ~ cat("  ", .y, ": ", length(.x), "\n"))
  
  # Pre-filter lists for different plot types
  term_list_all <- goslim_term_lists[!grepl("universe", names(goslim_term_lists))]
  term_list_leachate <- goslim_term_lists[grepl("leachate", names(goslim_term_lists))]
  term_list_stage <- goslim_term_lists[grepl("stage", names(goslim_term_lists))]
} else {
  cat("Warning: bp_slim_mapped_df is empty. No UpSet plots will be generated.\n")
  goslim_term_lists <- list()
  term_list_all <- list()
  term_list_leachate <- list()
  term_list_stage <- list()
}
```

### UpSet Plot: All Groups

Visualize the overlap of GOslim terms across all groups (leachate effects and stage effects).

```{r upset-all-groups, fig.width=12, fig.height=8}
# Create upset plot for all groups (excluding universe)
# Note: keep.order = FALSE allows reordering by frequency for better visualization
if (length(term_list_all) > 0) {
  upset(
    fromList(term_list_all),
    nsets = length(term_list_all),
    order.by = "freq",
    keep.order = FALSE,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "Intersection Size\n(Number of Shared Terms)",
    sets.x.label = "Set Size\n(Total Terms per Group)",
    text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
    mb.ratio = c(0.6, 0.4)
  )
}
```

### UpSet Plot: Leachate Effects Only

Visualize the overlap of GOslim terms across leachate-responsive genes at different developmental stages.

```{r upset-leachate-groups, fig.width=10, fig.height=6}
# Filter to leachate groups only
# Note: keep.order = TRUE maintains the input order of groups
if (length(term_list_leachate) > 0) {
  upset(
    fromList(term_list_leachate),
    nsets = length(term_list_leachate),
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "Intersection Size\n(Number of Shared Terms)",
    sets.x.label = "Set Size\n(Total Terms per Group)",
    text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
    mb.ratio = c(0.6, 0.4)
  )
}
```

### UpSet Plot: Stage Effects Only

Visualize the overlap of GOslim terms across stage-specific gene expression patterns.

```{r upset-stage-groups, fig.width=10, fig.height=6}
# Filter to stage groups only
# Note: keep.order = TRUE maintains the input order of groups
if (length(term_list_stage) > 0) {
  upset(
    fromList(term_list_stage),
    nsets = length(term_list_stage),
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "Intersection Size\n(Number of Shared Terms)",
    sets.x.label = "Set Size\n(Total Terms per Group)",
    text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
    mb.ratio = c(0.6, 0.4)
  )
}
```

### Save UpSet Plots

```{r save-upset-plots}
# Save all groups plot
png(
  filename = file.path(output_path, "upset_plot_all_groups.png"),
  width = 12,
  height = 8,
  units = "in",
  res = 300,
  bg = "white"
)

if (length(term_list_all) > 0) {
  upset(
    fromList(term_list_all),
    nsets = length(term_list_all),
    order.by = "freq",
    keep.order = FALSE,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "Intersection Size\n(Number of Shared Terms)",
    sets.x.label = "Set Size\n(Total Terms per Group)",
    text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
    mb.ratio = c(0.6, 0.4)
  )
}

dev.off()

# Save leachate groups plot
png(
  filename = file.path(output_path, "upset_plot_leachate_groups.png"),
  width = 10,
  height = 6,
  units = "in",
  res = 300,
  bg = "white"
)

if (length(term_list_leachate) > 0) {
  upset(
    fromList(term_list_leachate),
    nsets = length(term_list_leachate),
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "Intersection Size\n(Number of Shared Terms)",
    sets.x.label = "Set Size\n(Total Terms per Group)",
    text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
    mb.ratio = c(0.6, 0.4)
  )
}

dev.off()

# Save stage groups plot
png(
  filename = file.path(output_path, "upset_plot_stage_groups.png"),
  width = 10,
  height = 6,
  units = "in",
  res = 300,
  bg = "white"
)

if (length(term_list_stage) > 0) {
  upset(
    fromList(term_list_stage),
    nsets = length(term_list_stage),
    order.by = "freq",
    keep.order = TRUE,
    point.size = 3.5,
    line.size = 1.5,
    mainbar.y.label = "Intersection Size\n(Number of Shared Terms)",
    sets.x.label = "Set Size\n(Total Terms per Group)",
    text.scale = c(1.5, 1.3, 1.2, 1.2, 1.5, 1.2),
    mb.ratio = c(0.6, 0.4)
  )
}

dev.off()

cat("UpSet plots saved to:", output_path, "\n")
```

### Interpretation

The UpSet plots show:

- **Left bars**: Total number of GOslim terms in each group (set size)
- **Top bars**: Number of terms in each intersection (how many terms are shared)
- **Dots and lines**: Which groups are included in each intersection

Key insights:
- Terms appearing in only one group represent biological processes unique to that developmental stage or treatment response
- Terms shared across multiple groups indicate common biological processes across conditions
- The largest intersections reveal the most conserved biological processes

