---
title: "Create Gene to GOslim Mapping Dataframe"
subtitle: "Map GOslim terms → GO terms → DEGs"
author: "Sarah Tanja"
date: "01/29/2026"
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

This script creates a dataframe that maps GO slim terms to GO terms to differentially expressed genes (DEGs).

The output dataframe has the following columns:
- `group`: The gene set group (e.g., leachate.cleavage, stage.prawnchipup)
- `GOslimID`: The GO ID of the GO slim term
- `goslimterm`: GO slim term names
- `goterms`: comma-separated list of GO terms that map to each GO slim term (union of all GO terms)
- `deg`: comma-separated list of all genes that map to any of the GO terms (union of all genes)

The dataframe can then be transformed to "long" format with one row per (GO slim term, GO term, gene) combination.

# Load libraries

```{r}
library(GSEABase)
library(tidyverse)
library(GO.db)
library(purrr)
```

# Inputs

```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

## Load GO annotated genes

```{r}
universe_genes_anno <- read_csv(file.path(input_path, "universe_genes_anno.csv"))

glimpse(universe_genes_anno)
```

Change comma separated GO IDs to semi-colon separated GO IDs

```{r}
gene2GOs <- universe_genes_anno %>%
  mutate(GOs = str_replace_all(GOs, ",", ";"))

head(gene2GOs)
```

# Make a gene set for each cluster and module

## Leachate DEGs

```{r}
gene2GOs_leachate.cleavage <- gene2GOs %>%
  filter(cluster_stage == "cleavage_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_cleavage")
```

```{r}
gene2GOs_leachate.prawnchip <- gene2GOs  %>%
  filter(cluster_stage == "prawnchip_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_prawnchip")
```

```{r}
gene2GOs_leachate.earlygastrula <- gene2GOs  %>%
  filter(cluster_stage == "early_gastrula_peak") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "leachate_earlygastrula")
```

## Stage DEGs

```{r}
gene2GOs_stage.cleavage <- gene2GOs  %>%
  filter(module_stage == "cleavage_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_cleavage")
```

```{r}
gene2GOs_stage.prawnchipup <- gene2GOs %>%
  filter(module_stage == "prawnchip_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_prawnchip_up")
```

```{r}
gene2GOs_stage.prawnchipdip <- gene2GOs %>%
  filter(module_stage == "prawnchip_dip") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_prawnchip_dip")
```

```{r}
gene2GOs_stage.earlygastrulaup <- gene2GOs %>%
  filter(module_stage == "earlygastrula_up") %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "stage_earlygastrula_up")
```

## Universe

```{r}
gene2GOs_universe <- gene2GOs %>%
  dplyr::select(gene_id, GOs) %>% 
  mutate(list = "universe")
```

## Save gene2GO dataframes as a named list

```{r}
gene2GOs_list <- list(
  leachate.cleavage      = gene2GOs_leachate.cleavage,
  leachate.prawnchip     = gene2GOs_leachate.prawnchip,
  leachate.earlygastrula = gene2GOs_leachate.earlygastrula,
  stage.cleavage         = gene2GOs_stage.cleavage,
  stage.prawnchipup      = gene2GOs_stage.prawnchipup,
  stage.prawnchipdip     = gene2GOs_stage.prawnchipdip,
  stage.earlygastrulaup  = gene2GOs_stage.earlygastrulaup,
  universe               = gene2GOs_universe
)
```

# Set variables

```{r set-variables, eval=TRUE}
# Column names corresponding to gene name/ID and GO IDs
GO.ID.column <- "GOs"
gene.ID.column <- "gene_id"

# Relative path or URL to input file
input.file <- "../../output/07_enrichment/goseq/universe_gene2GOs.csv"

##### Official GO info - no need to change #####
goslims_obo <- "goslim_generic.obo"
goslims_url <- "http://current.geneontology.org/ontology/subsets/goslim_generic.obo"
```

# Download GOslim OBO file

```{r download-generic-goslim-obo, eval=TRUE}
# Find GSEAbase installation location
gseabase_location <- find.package("GSEABase")

# Load path to GOslim OBO file
goslim_obo_destination <- file.path(gseabase_location, "extdata", goslims_obo, fsep = "/")

# Download the GOslim OBO file
download.file(url = goslims_url, destfile = goslim_obo_destination)

# Loads package files
gseabase_files <- system.file("extdata", goslims_obo, package="GSEABase")
```

# Clean and prepare data

## Helper functions

```{r}
# Clean NA values
check_na_go <- function(df, GO.ID.column) {
  tibble(
    n_rows = nrow(df),
    n_na_go = sum(is.na(df[[GO.ID.column]]))
  )
}

# Clean gene2GO data
clean_gene2go <- function(df, gene.ID.column, GO.ID.column) {
  df %>%
    mutate(
      !!sym(GO.ID.column) :=
        str_replace_all(.data[[GO.ID.column]], "\\s*;\\s*", ";")
    ) %>%
    filter(
      !is.na(.data[[gene.ID.column]]),
      !is.na(.data[[GO.ID.column]]),
      .data[[GO.ID.column]] != ""
    ) %>%
    select(all_of(c(gene.ID.column, GO.ID.column)))
}

# Map gene2GO to one row per gene-GO pair
map_gene2go <- function(df, GO.ID.column, sep = ";") {
  df %>%
    separate_rows(!!sym(GO.ID.column), sep = sep)
}

# Clean GO IDs
clean_go_ids <- function(x) {
  x <- unique(na.omit(as.character(x)))
  x[grepl("^GO:\\d{7}$", x)]
}
```

## Apply cleaning functions

```{r}
# Check for NAs
na_summary <- imap_dfr(
  gene2GOs_list,
  ~ check_na_go(.x, GO.ID.column),
  .id = "dataset"
)

na_summary
```

```{r}
# Clean the data
gene2GOs_clean <- imap(
  gene2GOs_list,
  ~ clean_gene2go(.x,
                  gene.ID.column = gene.ID.column,
                  GO.ID.column   = GO.ID.column)
)
```

```{r}
# Map to one row per gene-GO pair
gene2GO_map <- imap(
  gene2GOs_clean,
  ~ map_gene2go(.x, GO.ID.column = GO.ID.column)
)
```

```{r}
# Check mapping
imap_dfr(
  gene2GO_map,
  ~ tibble(
    dataset = .y,
    n_rows  = nrow(.x)
  )
)
```

# Map GO IDs to GOslims

## Retrieve GOslims from GSEABase

```{r extract-GOslims-from-OBO, eval=TRUE}
# Retrieve GOslims from GO OBO file set
slim <- getOBOCollection(gseabase_files)
```

## Extract GO IDs as vectors

```{r}
# Extract GO IDs as vectors for each group
go_ids_list <- imap(
  gene2GO_map,
  ~ unique(.x[[GO.ID.column]])
)
```

## Create GOCollection objects

```{r}
# For each element of go_ids_list, make a GOCollection
go_collections <- map(go_ids_list, ~ GOCollection(clean_go_ids(.x)))

# Name each collection by module (list)
names(go_collections) <- names(go_ids_list)
```

## Map GO IDs to GOslims function

```{r}
map_go_to_goslims <- function(go_ids, slim, ontology = "BP", verbose = FALSE) {
  go_ids <- clean_go_ids(go_ids)

  # Build GOCollection for this group's GO IDs
  myCollection <- GOCollection(go_ids)

  # Get GOslim summary table (rownames are GOslim IDs)
  slimdf <- goSlim(myCollection, slim, ontology, verbose = verbose)

  # --- Map GO IDs -> GOslims (semi-colon delimited) --------------------------
  # For each GOslim term, get its "offspring" terms from GO.db, then intersect with this group's GO IDs
  gomap <- as.list(GOBPOFFSPRING[rownames(slimdf)])

  mapped_ids <- lapply(gomap, intersect, ids(myCollection))

  slimdf$GO.IDs <- vapply(
    mapped_ids,
    function(x) paste(x, collapse = "; "),
    FUN.VALUE = character(1)
  )

  # Remove literal "character(0)" (usually shouldn't happen with vapply, but keep it safe)
  slimdf$GO.IDs[slimdf$GO.IDs %in% c("character(0)", "")] <- ""

  # Add self-matching GOIDs to its own GOslim row, if applicable
  for (go_id in go_ids) {
    if (go_id %in% rownames(slimdf)) {
      existing <- trimws(toupper(unlist(strsplit(slimdf[go_id, "GO.IDs"], ";"))))
      existing <- existing[nzchar(existing)]

      if (!toupper(go_id) %in% existing) {
        if (nzchar(slimdf[go_id, "GO.IDs"])) {
          slimdf[go_id, "GO.IDs"] <- paste0(slimdf[go_id, "GO.IDs"], "; ", go_id)
        } else {
          slimdf[go_id, "GO.IDs"] <- go_id
        }
      }
    }
  }

  slimdf
}
```

## Apply mapping function

```{r}
bp_slim_mapped_list <- imap(go_ids_list, ~ {
  message("Mapping: ", .y)
  map_go_to_goslims(.x, slim = slim, ontology = "BP", verbose = FALSE)
})

str(bp_slim_mapped_list[[1]])
```

# Create gene2goslim dataframe

Now we'll create a dataframe with the following columns:
- `group`: The gene set group
- `GOslimID`: The GO ID of the GO slim term
- `goslimterm`: GO slim term name
- `goterms`: comma-separated GO terms that map to the GO slim term
- `deg`: comma-separated genes that map to any of the GO terms (union)

## Map genes to GO terms to GOslim terms

```{r}
# Function to create the gene2goslim mapping
create_gene2goslim <- function(gene2go_df, goslim_df, gene.ID.column = "gene_id", GO.ID.column = "GOs") {
  
  # Add GOslim ID as a column
  goslim_with_id <- goslim_df %>%
    rownames_to_column("GOslimID")
  
  # Separate GO.IDs (semicolon-delimited from goslim mapping) into individual rows
  # Note: goslim_df has GO terms in semicolon-delimited format, we'll convert to comma-separated in output
  goslim_long <- goslim_with_id %>%
    separate_rows(GO.IDs, sep = ";") %>%
    mutate(GO.IDs = trimws(GO.IDs)) %>%
    filter(!is.na(GO.IDs), GO.IDs != "")
  
  # Join with gene2GO mapping to get genes for each GO term
  # Note: gene2go_df has one row per gene-GO pair (already separated)
  # The GO.ID.column in gene2go_df should contain individual GO IDs
  gene2goslim <- gene2go_df %>%
    inner_join(goslim_long, by = setNames("GO.IDs", GO.ID.column)) %>%
    # Group by GOslim term and aggregate GO terms and genes
    group_by(GOslimID, Term) %>%
    summarise(
      goterms = paste(unique(!!sym(GO.ID.column)), collapse = ","),
      deg = paste(unique(!!sym(gene.ID.column)), collapse = ","),
      .groups = "drop"
    ) %>%
    # Rename columns to match requested names
    rename(goslimterm = Term)
  
  return(gene2goslim)
}
```

## Apply to all gene sets

```{r}
gene2goslim_list <- imap(names(gene2GO_map), ~ {
  message("Creating gene2goslim for: ", .x)
  create_gene2goslim(
    gene2go_df = gene2GO_map[[.x]],
    goslim_df = bp_slim_mapped_list[[.x]],
    gene.ID.column = gene.ID.column,
    GO.ID.column = GO.ID.column
  ) %>%
    mutate(group = .x, .before = 1)
})

names(gene2goslim_list) <- names(gene2GO_map)
```

## Combine all groups

```{r}
gene2goslim_all <- bind_rows(gene2goslim_list)

glimpse(gene2goslim_all)
```

## Preview the dataframe

```{r}
head(gene2goslim_all)
```

# Transform to long format

Create a "long" dataframe with one row per (GO slim term, GO term, gene) combination.

```{r}
gene2goslim_long <- gene2goslim_all %>%
  # Separate GO terms
  separate_rows(goterms, sep = ",") %>%
  mutate(goterms = trimws(goterms)) %>%
  # Separate genes
  separate_rows(deg, sep = ",") %>%
  mutate(deg = trimws(deg)) %>%
  # Remove any empty values
  filter(!is.na(goslimterm), goslimterm != "",
         !is.na(goterms), goterms != "",
         !is.na(deg), deg != "")

glimpse(gene2goslim_long)
```

## Preview long format

```{r}
head(gene2goslim_long, 20)
```

# Summary statistics

```{r}
# Count entries by group
gene2goslim_all %>%
  group_by(group) %>%
  summarise(
    n_goslim_terms = n(),
    n_go_terms = sum(str_count(goterms, ",") + 1),
    n_genes = sum(str_count(deg, ",") + 1)
  )
```

```{r}
# Count entries in long format by group
gene2goslim_long %>%
  group_by(group) %>%
  summarise(
    n_unique_goslim_terms = n_distinct(goslimterm),
    n_unique_go_terms = n_distinct(goterms),
    n_unique_genes = n_distinct(deg),
    n_total_rows = n()
  )
```

# Save outputs

```{r}
output_path <- "../../output/07_enrichment/goslim/"
```

## Save wide format (comma-separated)

```{r}
write_csv(gene2goslim_all, file.path(output_path, "gene2goslim_wide.csv"))
```

## Save long format

```{r}
write_csv(gene2goslim_long, file.path(output_path, "gene2goslim_long.csv"))
```

## Save individual group files (optional)

```{r}
# Save each group separately if needed
iwalk(gene2goslim_list, ~ {
  write_csv(.x, file.path(output_path, paste0("gene2goslim_", .y, ".csv")))
})
```

```{r}
cat("Output files saved to:", output_path, "\n")
cat("- gene2goslim_wide.csv: Wide format with comma-separated lists\n")
cat("- gene2goslim_long.csv: Long format with one row per goslim-go-gene combination\n")
```

```{r}
SessionInfo()
```

