---
title: "Visualize GOslims by numbers of DEGs"
subtitle: "What do they genes do!? Can we see it with geom_bar and Upset plots?"
date: 02/06/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

We identified GOslim terms associated with DEGs in each cluster/module.
This is a descriptive summary plot, and does not reflect statistical
enrichment!

With this plot we simply want to know what the DE genes do, and how many
of them are associated with each GOslim term.

`goslim_summary.csv` was generated in \# Load libraries

```{r load-libraries}
library(tidyverse)
library(ggsidekick)
library(PNWColors)
library(kableExtra)
```

# Setup

## Custom ggplot themes

```{r theme}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.5
      ),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 12,
        color = "grey30",
        hjust = 0.5,
        margin = margin(b = 2)
      ),
      
      strip.text = element_text(size = 10, color = "grey30"),
      
      plot.subtitle = element_text(
        size   = 10,
        color  = "grey30",
        hjust  = 0.5,
        margin = margin(t = 0)
      ),
      
      axis.title = element_text(size = 10, color = "grey30"),
      axis.text = element_text(size = 10, color = "grey30"),
      legend.title = element_text(size = 10, color = "grey30"),
      legend.text = element_text(size = 10, color = "grey30")
    )
}


theme_pptx <- function() {
  theme_sleek() +
    theme(
      panel.border = element_blank(),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 40,
        color = "grey30",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 24, color = "grey30"),
      
      plot.subtitle = element_text(
        size   = 24,
        color  = "grey30",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 24, color = "grey30"),
      axis.text = element_text(size = 24, color = "grey30"),
      legend.title = element_text(size = 24, color = "grey30"),
      legend.text = element_text(size = 24, color = "grey30")
    )
}

```

## Geoms

```{r geoms}
geom_defaults <- function() {
  update_geom_defaults("point",      list(size = 1.5, stroke = 0.5))
  update_geom_defaults("line",       list(linewidth = 0.5))
  update_geom_defaults("boxplot",    list(linewidth = 0.5))
  update_geom_defaults("violin",     list(linewidth = 0.5))
  update_geom_defaults("vline",      list(linewidth = 0.5))
  update_geom_defaults("hline",      list(linewidth = 0.5))
  update_geom_defaults("abline",     list(linewidth = 0.5))
  update_geom_defaults("errorbar",   list(linewidth = 0.5))
  update_geom_defaults("pointrange", list(linewidth = 0.5))
  update_geom_defaults("crossbar",   list(linewidth = 0.5))
  update_geom_defaults("linerange",  list(linewidth = 0.5))
  update_geom_defaults("path", list(linewidth = 0.5))
  update_geom_defaults("segment", list(linewidth = 0.5))
  update_geom_defaults("text", list(size = 3))
}

geom_pptx <- function() {
  update_geom_defaults("point",   list(size = 5, stroke = 1.2))
  update_geom_defaults("line",    list(linewidth = 1.5))
  update_geom_defaults("boxplot", list(linewidth = 1.2))
  update_geom_defaults("violin",  list(linewidth = 1.2))
  update_geom_defaults("vline", list(linewidth = 1.5))
  update_geom_defaults("hline", list(linewidth = 1.5))
  update_geom_defaults("abline", list(linewidth = 1.5))
  update_geom_defaults("errorbar", list(linewidth = 1.5))
  update_geom_defaults("pointrange", list(linewidth = 1.5))
  update_geom_defaults("crossbar", list(linewidth = 1.5))
  update_geom_defaults("linerange", list(linewidth = 1.5))
  update_geom_defaults("path", list(linewidth = 1.5))
  update_geom_defaults("segment", list(linewidth = 1.5))
  update_geom_defaults("text", list(size = 7))
}

```

## Colorschemes

```{r colors}
leachate.colors <- c(control = "#AEF1FF", 
                     low     = "#BBC7FF",
                     mid     = "#7D8BFF", 
                     high    = "#592F7D")

stage.5.colors <- c(egg           = "#FFE362",
                    cleavage      = "#EBA600", 
                    morula        = "#E6AA83",
                    prawnchip     = "#D9685B", 
                    earlygastrula = "#A2223C")

stage.3.colors <- c(cleavage      = "#EBA600", 
                    prawnchip     = "#D9685B", 
                    earlygastrula = "#A2223C")

status.colors <- c(typical   = "#75C165", 
                   uncertain = "#E3FAA5", 
                   malformed = "#8B0069")

night.colors <- c(July_6th = "#E3FAA5", 
                  July_7th = "#578B21", 
                  July_8th = "#1E2440")

night.colors.alt <- c(July_6th = "#21918C", 
                      July_7th = "#578B21", 
                      July_8th = "#1E2440")
```

# Load inputs

```{r input-path}
input_path <- "../../output/07_enrichment/goslim/"
```

```{r load-goslim-data}
goslim_summary <- read_csv(file.path(input_path, "goslim_summary.csv"))
```

```{r glimpse-goslim}
glimpse(goslim_summary)
```

```{r summary-goslim}
summary(goslim_summary)
```

```{r count-nas}
colSums(is.na(goslim_summary))
```

# Wrangle

remove rows that are 0 or NA

```{r horiz-bar, fig.height = 10}
goslim_summary %>%
  drop_na(genes) %>% 
  filter(DEdueto == "leachate") %>% 
  filter(n_genes > 2) %>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10) %>%
  ungroup() %>% 
ggplot(., aes(x = n_genes, y = Term, fill = Term)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.85) +
  facet_wrap(~ pattern, scales = "free_y", ncol = 1) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    title = "GOslim BP Terms: Number of Genes Mapped",
    x = "Number of differentially expressed genes mapped to GOslim terms",
    y = "Top 10 GOslim terms per gene expression module"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

```{r save-bar}
ggsave(
  filename = "goslim_summary_barplot.png",
  plot = last_plot(),
  path = "../../output/07_enrichment/goslim/",
  width = 10,
  height = 8,
  units = "in",
  dpi = 900
)
```

# How do I show this in context of background universe and stage signals?

# How many GOSlims are there?

```{r length-unique}
length(unique(goslim_summary$Term))
```

There are 72 unique Go Slim Terms.

# How many terms are represented?

```{r count-goslims}
goslim_summary %>%
  filter(DEdueto == "leachate", n_genes > 0) %>%
  group_by(module) %>%
  arrange(n_genes) %>%
  ungroup() %>%
  count(module, Term) %>%
  count(module, name = "n_unique_terms")
```

There are 32 terms in the down pattern, 37 in the hump patterns, and 53
in the up pattern.

# Top 15 terms

Which are the top \~15 terms of *leachate-responsive DEGs* represented
in each pattern by percent?

```{r unique-by-module}
goslim_summary %>%
  filter(DEdueto == "leachate", n_genes > 2) %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 15) %>%
  ungroup() %>%
  count(module, Term) %>%
  count(module, name = "n_unique_terms")
```

There are 14 unique terms represented in the down pattern, 9 in the hump
pattern, and 16 in the up pattern.

# Which are unique?

Across down, hump, and up.. which terms are only found in one pattern
(e.g. only in down, or only in up)?

```{r unique}
unique_terms <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(Term) %>%
  summarise(
    n_modules = n_distinct(module),
    modules = paste(sort(unique(module)), collapse = ", "),
    patterns = paste(sort(unique(pattern)), collapse = ", "),
    mean_percent = mean(Percent, na.rm = TRUE),
    mean_n_genes = mean(n_genes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_modules == 1) %>%
  arrange(desc(mean_percent), Term) %>% 
  dplyr::select(Term, patterns, mean_percent, mean_n_genes) %>% 
  arrange(patterns, desc(mean_percent))

unique_terms

```

There are 6 unique terms

Unique to down (aka cleavage):

-   **meitoc nuclear division, n_DEG = 3**

Unique to hump:

-   lipid metabolic process, n_DEG = 4

-   carbohydrate derivative metabolic process n_DEG = 2

-   DNA-templated transcription, n_DEG = 5

-   protein catabolic process, n_DEG = 1

Unique to up (aka early gastrula):

-   **vesicle-mediated transport, n_DEG = 11**

# Which are shared across counts vs percents?

```{r top10s}
top_terms_percent <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 10, with_ties = FALSE) %>%
  ungroup()

top_terms_genes <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10, with_ties = FALSE) %>%
  ungroup()
```

```{r percent-only}
setdiff(top_terms_percent$Term, top_terms_genes$Term) # percent only 
```

```{r count-only}
setdiff(top_terms_genes$Term, top_terms_percent$Term) # gene count only
```

```{r intersect}
intersect(top_terms_percent$Term, top_terms_genes$Term) # shared between both
```

#### 1. Cell division / early embryogenesis

-   **meiotic nuclear division**
-   **mitotic cell cycle**
-   chromosome segregation

> Early embryos rely on rapid, tightly coordinated cell divisions.
> Changes here may reflect abnormal timing, spindle disruption, or DNA
> segregation stress. This provides further evidence that PVC leachate
> may subtly perturb cell cycle progression without causing overt
> mortality.

#### 2. Developmental patterning

-   **anatomical structure development**
-   **cell differentiation**

these support your observation of subtle developmental timing changes.

Strong angle: transcriptional changes may precede visible malformations.

#### 3. Gene regulation / epigenetic control

-   **regulation of DNA-templated transcription**
-   DNA-templated transcription
-   chromatin organization

> Embryos are executing tightly programmed developmental gene networks.
> Disturbance here suggests reprogramming of developmental trajectories.

Especially strong if responses strongest at 4 hpf.

#### 4. Cellular signaling / communication

-   **signaling**
-   **vesicle-mediated transport**

> Cells respond to external stressors through signaling pathways;
> vesicle trafficking is central to membrane dynamics, secretion, and
> stress responses.

#### 5. Metabolism / energetic shifts

-   **lipid metabolic process**
-   carbohydrate derivative metabolic process
-   protein catabolic process

Why they matter: indicates resource reallocation, stress metabolism,
membrane remodeling, or energetic costs of exposure.

Cell cycle + chromosome segregation → altered developmental timing
Transcription/chromatin regulation → early molecular reprogramming
Metabolism + signaling → physiological stress compensation

Programmed cell death - Useful if multiple apoptosis-related genes
present. If not, mention briefly as potential stress response.

Immune system process \> DEGs included genes annotated to immune-related
processes, potentially reflecting early innate defense or generalized
cellular stress pathways.

Reproductive process - A broad GOslim annotation. In embryos maybe this
reflects maternally provisioned transcripts... indicator for maternal to
zygote transition?

> Leachate-responsive DEGs were concentrated in functions related to
> cell division, developmental regulation, transcriptional control, and
> metabolism, suggesting that PVC leachate may subtly disrupt early
> embryogenic processes before overt morphological abnormalities become
> apparent.

```{r full-join}
top_terms <- full_join(top_terms_genes, top_terms_percent) %>% 
  mutate(
    response = case_when(
      
      # Early development, cleavage, morphogenesis
      Term %in% c(
        "mitotic cell cycle",
        "anatomical structure development",
        "cell differentiation",
        "cytoskeleton organization",
        "meiotic nuclear division"
      ) ~ "Early embryogenesis, cell division, and fate",
      
      # Maternal-effect / early cell-cycle regulatory processes
      Term %in% c(
        "chromosome segregation",
        "reproductive process"
      ) ~ "Cell cycle and maternal-effect processes",
      
      # Transcriptional control / chromatin state
      Term %in% c(
        "regulation of DNA-templated transcription",
        "DNA-templated transcription",
        "chromatin organization"
      ) ~ "Gene regulation and epigenetic control",
      
      # Communication, trafficking, signaling cascades
      Term %in% c(
        "signaling",
        "vesicle-mediated transport"
      ) ~ "Cellular signaling and communication",
      
      # Cellular defense / damage response
      Term %in% c(
        "immune system process",
        "programmed cell death"
      ) ~ "Stress responses",
      
      # Resource use / biochemical processing
      Term %in% c(
        "lipid metabolic process",
        "carbohydrate derivative metabolic process",
        "protein catabolic process"
      ) ~ "Metabolism and energetic shifts",
      
      TRUE ~ "Other"
    ),
    
    response = factor(
      response,
      levels = c(
        "Early embryogenesis, cell division, and fate",
        "Cell cycle and maternal-effect processes",
        "Gene regulation and epigenetic control",
        "Cellular signaling and communication",
        "Stress responses",
        "Metabolism and energetic shifts",
        "Other"
      ),
      ordered = TRUE
    )
  )

top_terms %>% 
  group_by(response) %>%
  summarise(
    sum_genes = sum(n_genes),
    sum_percent = sum(Percent)
  ) %>%
  arrange(desc(sum_genes), desc(sum_percent)) %>% 
  kable() 
```

```{r top17-summary}
top_terms %>% 
  group_by(Term) %>%
  summarise(
    sum_genes = sum(n_genes),
    sum_percent = sum(Percent)
  ) %>%
  arrange(desc(sum_genes), desc(sum_percent)) %>% 
  kable() 
```

# Make plotting dataframe

```{r}
universe <- goslim_summary %>% 
  dplyr::filter(module == "universe") %>% 
  rename(universe_n_genes = n_genes, 
         goterm = Term) %>% 
  dplyr::select(goterm, universe_n_genes)

```

```{r}

plot_terms <- c(
  "anatomical structure development",
  "cell differentiation",
  "meiotic nuclear division",
  "mitotic cell cycle",
  "signaling",
  "vesicle-mediated transport",
  "regulation of DNA-templated transcription",
  "lipid metabolic process"
)

response_levels <- c(
  "Early embryogenesis, cell division, and fate",
  "Gene regulation and epigenetic control",
  "Cellular signaling and communication",
  "Stress responses",
  "Metabolism and energetic shifts"
)

top_plot <- top_terms %>%
  filter(Term %in% plot_terms) %>% 
  mutate(goterm = factor(Term, levels=plot_terms, ordered=TRUE)) %>%
  left_join(., universe) %>% 
  complete(goterm, pattern, fill = list(n_genes = 0, 
                                             n_goids = 0, 
                                             Percent = 0, 
                                             Count = 0)) %>% 
   group_by(goterm) %>% 
  fill(response, .direction = "downup") %>% 
  ungroup() %>% 
  dplyr::select(goterm, pattern, response, n_genes, Percent)

  

str(top_plot)
```

::: callout-note
-   Percent values were linearly rescaled to the range of gene counts to
    allow simultaneous visualization of enrichment percentage and DEG
    counts on the same axis.

-   Percent was normalized to the maximum observed value and scaled to
    match the range of DEG counts for plotting.
:::

# Spider / Radar

inspo from R graph gallery
[here](https://r-graph-gallery.com/spider-or-radar-chart.html)

```{r palette}
bay <- pnw_palette(name="Bay", n=5, type="discrete") #early embryo cell stuff
pnw_palette(name="Spring", n=2, type="discrete") #signaling
pnw_palette(name="Lake", n=2, type="discrete") # gene reg & metabolism
```

```{r radar_palette}
radar_pal <- c(
  "anatomical structure development" = "firebrick4",
  "cell differentiation" = "firebrick",
  "meiotic nuclear division" = "firebrick1",
  "mitotic cell cycle" = "firebrick3",
  "signaling" = "deepskyblue4",
  "vesicle-mediated transport" = "deepskyblue2",
  "regulation of DNA-templated transcription" = "olivedrab4",
  "lipid metabolic process" = "orchid3"
)
```

## Counts

```{r plot-radar-count, fig.height = 4, fig.width = 12}
plt_count <- ggplot(top_plot) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:4) * 5),
    color = "lightgrey"
  ) + 
  # Add bars to represent the number of genes mapped to each GO term
  # str_wrap(Term, 12) wraps the text so each line has at most 12 characters
  # (but it doesn't break long words!)
  geom_col(
    aes(
      x = goterm,
      y = n_genes,
      fill = goterm
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = goterm,
      y = 0,
      xend = goterm,
      yend = 21,
      color = goterm
    ),
    linetype = "dashed"
  ) + 
  facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 25),
    expand = c(0, 0)
  ) + 
  # Wrap text labels
   scale_x_discrete(labels = \(x) stringr::str_wrap(x, 15)
  )+
  # Add color palette
  scale_color_manual(values = radar_pal)+
  scale_fill_manual(values = radar_pal)+
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 12, color="grey30"),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.border = element_blank(),
    strip.text = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 1, 0.2, 1, "cm")
  )
  
  
plt_count  
```

```{r plot-count-annotated-axis, fig.width = 12, fig.height = 4}

plt_count_anno <- plt_count +
  # Annotate custom scale inside plot
  annotate(
    x = 8.5, 
    y = 6, 
    label = "5", 
    geom = "text", 
    color = "gray30", 
    size = 4
  ) +
  annotate(
    x = 8.5, 
    y = 11, 
    label = "10", 
    geom = "text", 
    color = "gray30", 
    size = 4
  ) +
  annotate(
    x = 8.5, 
    y = 16, 
    label = "15", 
    geom = "text", 
    color = "gray30", 
    size = 4
  )

plt_count_anno


```

##### Save radar plot

```{r save-radar}
ggsave("../../figs/radar_counts.png", plt_count_anno, width=12, height=4, units="in", dpi=900)
```

#### radar for pptx

```{r plot-count-pptx, fig.width=16, fig.height=9}
geom_pptx()
radar_count_pptx <- plt_count_anno + 
  theme(axis.text.x = element_text(size = 16, color="grey30"))
radar_count_pptx
ggsave(
  "../../figs/radar_count_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_count_pptx
)
```

### highlight per GoSlim

```{r}

geom_pptx()

terms_to_highlight <- levels(factor(top_plot$goterm))

radar_plots <- purrr::map(
  terms_to_highlight,
  \(term_i) {

    top_plot_hi <- top_plot %>%
      mutate(
        hi_fill = if_else(goterm == term_i, as.character(goterm), "Other"),
        hi_color = if_else(goterm == term_i, as.character(goterm), "Other")
      )

    ggplot(top_plot_hi) +
      geom_hline(
        aes(yintercept = y),
        data = data.frame(y = c(0:4) * 5),
        color = "lightgrey"
      ) +
      geom_col(
        aes(x = goterm, y = n_genes, fill = hi_fill),
        alpha = 0.7,
        show.legend = FALSE
      ) +
      geom_segment(
        aes(
          x = goterm,
          y = 0,
          xend = goterm,
          yend = 21,
          color = hi_color
        ),
        linetype = "dashed",
        show.legend = FALSE
      ) +
      facet_wrap(~pattern, ncol = 3) +
      coord_polar(clip = "off") +
      scale_y_continuous(
        limits = c(-2, 25),
        expand = c(0, 0)
      ) +
      scale_x_discrete(labels = \(x) stringr::str_wrap(x, 15)) +
      scale_fill_manual(values = c(radar_pal, Other = "grey80")) +
      scale_color_manual(values = c(radar_pal, Other = "grey80")) +
      #labs(title = term_i) +
  # Annotate custom scale inside plot
      annotate(
        x = 8.5, 
        y = 6, 
        label = "5", 
        geom = "text", 
        color = "gray30", 
        size = 4
      ) +
      annotate(
        x = 8.5, 
        y = 11, 
        label = "10", 
        geom = "text", 
        color = "gray30", 
        size = 4
      ) +
      annotate(
        x = 8.5, 
        y = 16, 
        label = "15", 
        geom = "text", 
        color = "gray30", 
        size = 4
      )+
      theme(
        panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_text(size = 16, color = "grey30"),
        axis.ticks = element_blank(),
        legend.position = "none",
        axis.title.x = element_blank(),
        panel.background = element_blank(),
        panel.border = element_blank(),
        strip.text = element_blank(),
        panel.spacing.x = unit(2, "cm"),
        plot.margin = margin(0.2, 1, 0.2, 1, "cm")
      )
  }
)
```

```{r radar-highlight fig.width=16, fig.height=9}
radar_plots[[1]]
radar_plots[[2]]
radar_plots[[3]]
radar_plots[[4]]
radar_plots[[5]]
radar_plots[[6]]
radar_plots[[7]]
radar_plots[[8]]

```

```{r radar-anatomical, fig.width=16, fig.height=9}
radar_anatomical <- radar_plots[[1]]
radar_anatomical
ggsave(
  "../../figs/radar_anatomical_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_anatomical
)
```

```{r radar-celldiff, fig.width=16, fig.height=9}
radar_celldiff <- radar_plots[[2]]
radar_celldiff
ggsave(
  "../../figs/radar_celldiff_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_celldiff
)
```

```{r radar-lipid, fig.width=16, fig.height=9}
radar_lipid <- radar_plots[[3]]
radar_lipid
ggsave(
  "../../figs/radar_lipid_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_lipid
)
```
```{r radar-meiotic, fig.width=16, fig.height=9}
radar_meiotic <- radar_plots[[4]]
radar_meiotic
ggsave(
  "../../figs/radar_meiotic_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_meiotic
)
```


```{r radar-mitotic, fig.width=16, fig.height=9}
radar_mitotic <- radar_plots[[5]]
radar_mitotic
ggsave(
  "../../figs/radar_mitotic_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_mitotic
)
```
```{r radar-dnareg, fig.width=16, fig.height=9}
radar_dnareg <- radar_plots[[6]]
radar_dnareg
ggsave(
  "../../figs/radar_dnareg_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_dnareg
)
```
```{r radar-signaling, fig.width=16, fig.height=9}
radar_signaling <- radar_plots[[7]]
radar_signaling
ggsave(
  "../../figs/radar_signaling_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_signaling
)
```
```{r radar-vesicle, fig.width=16, fig.height=9}
radar_vesicle <- radar_plots[[8]]
radar_vesicle
ggsave(
  "../../figs/radar_vesicle_pptx.png",
  width=16,
  height=9,
  dpi=600,
  plot=radar_vesicle
)
```
## Enriched

```{r, fig.width=12, fig.height=4}
plt_enriched <- ggplot(top_plot) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:5) * 5),
    color = "lightgrey"
  ) + 
  geom_col(
    aes(
      x = str_wrap(goterm, 12),
      y = n_genes,
      fill = goterm
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = str_wrap(goterm, 12),
      y = 0,
      xend = str_wrap(goterm, 12),
      yend = 21,
      color = goterm
    ),
    linetype = "dashed"
  ) + 
  # Add dots to represent the percent enriched
  geom_point(
    aes(
      x = str_wrap(goterm, 12),
      y = percent_scaled
    ),
    size = 1.5, 
    color = "grey30"
  ) +
  facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0),
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 1, 0.2, 1, "cm")
  )
  
  
plt_enriched  
```

## Universe

```{r}
uni_df <- goslim_summary %>%
  drop_na(genes) %>%
  filter(module == "universe") %>%
  group_by(Term) %>%
  arrange(desc(n_genes))
```

```{r, include=FALSE}
plt_universe <- ggplot(uni_df) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:5) * 5),
    color = "lightgrey"
  ) + 
  geom_col(
    aes(
      x = str_wrap(Term, 12),
      y = n_genes,
      fill = Term
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = str_wrap(Term, 12),
      y = 0,
      xend = str_wrap(Term, 12),
      yend = 21,
      color = Term
    ),
    linetype = "dashed"
  ) + 
  # Add dots to represent the percent enriched
  #geom_point(
  #  aes(
  #    x = str_wrap(Term, 12),
  #    y = percent_scaled
  #  ),
  #  size = 1.5, 
   # color = "grey30"
  #) +
  #facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0),
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 1, 0.2, 1, "cm")
  )
  
  
plt_universe
```

# Faceted lollipop ALL DEGs

```{r, fig.height= 12}
library("colorspace")
hcl_palettes(plot = TRUE)
```

```{r lollipop-plotting-df}
goslim_lollipop <- goslim_summary %>%
  drop_na(genes) %>% 
  filter(DEdueto == "leachate")

# number of goslim terms in goslim_lollipop
length(unique(goslim_lollipop$Term))
```

## Count

```{r, fig.width= 18, fig.height = 12}
lolli_counts <- 
  ggplot(goslim_lollipop, aes(x = n_genes, y = Term, color = Percent)) +
  geom_segment(aes(xend = 0, yend = Term), size = 3) +
  geom_point(size = 4) +
  facet_wrap(~pattern, ncol = 3) +
  scale_color_continuous_sequential(palette = "Hawaii") +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Lollipop length reflects counts of DEGs",
    x = "Counts of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

lolli_counts
```

```{r save-lolli}
#save to fig 
ggsave("../../figs/lolli_counts.png", lolli_counts, width=18, height=12, units="in", dpi=900)
```

## Enriched

```{r, fig.height = 12}
ggplot(goslim_lollipop, aes(x = Percent, y = Term, color = n_genes)) +
  geom_segment(aes(xend = 0, yend = Term), size = 2) +
  geom_point(size = 3) +
  facet_wrap(~pattern, ncol = 3) +
  scale_color_continuous_sequential(palette = "Heat2", begin = 0.3, end = 1) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Lollipop length reflects percent representation",
    x = "Percent of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Universe

```{r, fig.height=12}
ggplot(uni_df, aes(x = n_genes, y = Term)) +
  geom_segment(aes(xend = 0, yend = Term, color = n_genes), size = 2) +
  geom_point(aes(color = n_genes), size = 3) +
  scale_color_continuous_sequential(palette = "Heat2", begin = 0.2, end = 1) +
  labs(
    title = "GOslim BP terms in the background universe",
    subtitle = "Lollipop length reflects counts of genes mapped to GOslim term",
    x = "Counts of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
