---
title: "Visualize GOslims by numbers of DEGs"
subtitle: "What do they genes do!? Can we see it with geom_bar and Upset plots?"
date: 02/06/2026
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background
We identified GOslim terms associated with DEGs in each cluster/module. This is a descriptive summary plot, and does not reflect statistical enrichment!

With this plot we simply want to know what the DE genes do, and how many of them are associated with each GOslim term.

`goslim_summary.csv` was generated in 
# Load libraries
```{r}
library(tidyverse)
library(kableExtra)
```

# Load inputs
```{r}
input_path <- "../../output/07_enrichment/goslim/"
```

```{r}
goslim_summary <- read_csv(file.path(input_path, "goslim_summary.csv"))
```

```{r}
glimpse(goslim_summary)
```

```{r}
summary(goslim_summary)
```
```{r}
colSums(is.na(goslim_summary))
```

# Wrangle
remove rows that are 0 or NA
```{r, fig.height = 10}
goslim_summary %>%
  drop_na(genes) %>% 
  filter(DEdueto == "leachate") %>% 
  filter(n_genes > 2) %>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10) %>%
  ungroup() %>% 
ggplot(., aes(x = n_genes, y = Term, fill = Term)) +
  geom_col(position = position_dodge(width = 0.9), alpha = 0.85) +
  facet_wrap(~ pattern, scales = "free_y", ncol = 1) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(
    title = "GOslim BP Terms: Number of Genes Mapped",
    x = "Number of differentially expressed genes mapped to GOslim terms",
    y = "Top 10 GOslim terms per gene expression module"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
ggsave(
  filename = "goslim_summary_barplot.png",
  plot = last_plot(),
  path = "../../output/07_enrichment/goslim/",
  width = 10,
  height = 8,
  units = "in",
  dpi = 900
)
```


# How do I show this in context of background universe and stage signals?

# How many GOSlims are there? 
```{r}
length(unique(goslim_summary$Term))
```
There are 72 unique Go Slim Terms. 


# How many terms are represented? 
```{r}
goslim_summary %>%
  filter(DEdueto == "leachate", n_genes > 0) %>%
  group_by(module) %>%
  arrange(n_genes) %>%
  ungroup() %>%
  count(module, Term) %>%
  count(module, name = "n_unique_terms")
```
There are 32 terms in the down pattern, 37 in the hump patterns, and 53 in the up pattern. 


# Top 15 terms
Which are the top ~15 terms of *leachate-responsive DEGs* represented in each pattern by percent?
```{r}
goslim_summary %>%
  filter(DEdueto == "leachate", n_genes > 2) %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 15) %>%
  ungroup() %>%
  count(module, Term) %>%
  count(module, name = "n_unique_terms")
```
There are 14 unique terms represented in the down pattern, 9 in the hump pattern, and 16 in the up pattern. 

# Which are unique? 
Across down, hump, and up.. which terms are only found in one pattern (e.g. only in down, or only in up)?
```{r}
unique_terms <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  group_by(Term) %>%
  summarise(
    n_modules = n_distinct(module),
    modules = paste(sort(unique(module)), collapse = ", "),
    patterns = paste(sort(unique(pattern)), collapse = ", "),
    mean_percent = mean(Percent, na.rm = TRUE),
    mean_n_genes = mean(n_genes, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_modules == 1) %>%
  arrange(desc(mean_percent), Term) %>% 
  dplyr::select(Term, patterns, mean_percent, mean_n_genes) %>% 
  arrange(patterns, desc(mean_percent))

unique_terms

```
There are 6 unique terms

Unique to down (aka cleavage):
- meitoc nuclear division, n_DEG = 3

Unique to hump:
- lipid metablolic process, n_DEG = 4
- carbohydrate derivative metabolic process n_DEG = 2
- DNA-templated transcription, n_DEG = 5
- protein catabolic process, n_DEG = 1

Unique to up (aka early gastrula):
- vesicle-mediated transport, n_DEG = 11

# Which are shared across counts vs percents?
```{r}
top_terms_percent <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = Percent, n = 10, with_ties = FALSE) %>%
  ungroup()

top_terms_genes <- goslim_summary %>%
  filter(DEdueto == "leachate") %>%
  group_by(module) %>%
  slice_max(order_by = n_genes, n = 10, with_ties = FALSE) %>%
  ungroup()
```

```{r}
setdiff(top_terms_percent$Term, top_terms_genes$Term) # percent only 
```
```{r}
setdiff(top_terms_genes$Term, top_terms_percent$Term) # gene count only
```
```{r}
intersect(top_terms_percent$Term, top_terms_genes$Term) # shared between both
```


#### 1. Cell division / early embryogenesis
- meiotic nuclear division
- mitotic cell cycle
- chromosome segregation

> Early embryos rely on rapid, tightly coordinated cell divisions. Changes here may reflect abnormal timing, spindle disruption, or DNA segregation stress. This provides further evidence that PVC leachate may subtly perturb cell cycle progression without causing overt mortality.

#### 2. Developmental patterning 
- anatomical structure development
- cell differentiation

these support your observation of subtle developmental timing changes.

Strong angle: transcriptional changes may precede visible malformations.

#### 3. Gene regulation / epigenetic control
- regulation of DNA-templated transcription
- DNA-templated transcription
- chromatin organization

> Embryos are executing tightly programmed developmental gene networks. Disturbance here suggests reprogramming of developmental trajectories.

Especially strong if responses strongest at 4 hpf.

#### 4. Cellular signaling / communication
- signaling
- vesicle-mediated transport

> Cells respond to external stressors through signaling pathways; vesicle trafficking is central to membrane dynamics, secretion, and stress responses.

#### 5. Metabolism / energetic shifts
- lipid metabolic process
- carbohydrate derivative metabolic process
- protein catabolic process

Why they matter: indicates resource reallocation, stress metabolism, membrane remodeling, or energetic costs of exposure.

Cell cycle + chromosome segregation → altered developmental timing
Transcription/chromatin regulation → early molecular reprogramming
Metabolism + signaling → physiological stress compensation

Programmed cell death
- Useful if multiple apoptosis-related genes present. If not, mention briefly as potential stress response.

Immune system process
> DEGs included genes annotated to immune-related processes, potentially reflecting early innate defense or generalized cellular stress pathways.

Reproductive process
- A broad GOslim annotation. In embryos maybe this reflects maternally provisioned transcripts... indicator for maternal to zygote transition? 

> Leachate-responsive DEGs were concentrated in functions related to cell division, developmental regulation, transcriptional control, and metabolism, suggesting that PVC leachate may subtly disrupt early embryogenic processes before overt morphological abnormalities become apparent.


```{r}
top_terms <- full_join(top_terms_genes, top_terms_percent) %>% 
  mutate(
    response = case_when(
      Term %in% c("meiotic nuclear division",
                  "mitotic cell cycle",
                  "chromosome segregation",
                  "anatomical structure development",
                  "reproductive process",
                  "cell differentiation", 
                  "cytoskeleton organization") ~ "Early embryogenesis, cell division, and fate",
      Term %in% c("regulation of DNA-templated transcription",
                  "DNA-templated transcription",
                  "chromatin organization") ~ "Gene regulation and epigenetic control",
      Term %in% c("signaling", "vesicle-mediated transport") ~ "Cellular signaling and communication",
      Term %in% c("lipid metabolic process", "carbohydrate derivative metabolic process", "protein catabolic process") ~ "Metabolism and energetic shifts",
      Term %in% c("immune system process", "programmed cell death") ~ "Stress responses" 
    )
  )

top_terms %>% 
  group_by(response) %>%
  summarise(
    sum_genes = sum(n_genes),
    sum_percent = sum(Percent)
  ) %>%
  arrange(desc(sum_genes), desc(sum_percent)) %>% 
  kable() 
```
```{r}
top_terms %>% 
  group_by(Term) %>%
  summarise(
    sum_genes = sum(n_genes),
    sum_percent = sum(Percent)
  ) %>%
  arrange(desc(sum_genes), desc(sum_percent)) %>% 
  kable() 
```



# Make plotting dataframe
```{r, fig.height = 12}
plot_df <- top_terms %>%
  filter(DEdueto == "leachate") %>% 
  mutate(goterm = factor(Term, levels = unique(Term)),
    pattern = factor(pattern, levels = c("down", "hump", "up")),
    percent_scaled = Percent / max(Percent) * max(n_genes),
    response = factor(
      response,
      levels = c(
        "Early embryogenesis, cell division, and fate",
        "Gene regulation and epigenetic control",
        "Cellular signaling and communication",
        "Stress responses",
        "Metabolism and energetic shifts"
      ), 
      ordered = TRUE
    )
  ) %>%
  arrange(response, desc(n_genes)) %>%
  mutate(
    goterm = factor(Term, levels = unique(Term)),
    pattern = factor(pattern, levels = c("down", "hump", "up")),
    percent_scaled = Percent / max(Percent) * max(n_genes)
  )

plot_df %>% group_by(response) %>%
  summarise(
    sum_genes = sum(n_genes),
    sum_percent = sum(Percent)
  ) %>%
  arrange(desc(sum_genes), desc(sum_percent)) %>%
  kable()
```

```{r}
plot_terms <- c()

top_plot <- top_terms %>% 
  dplyr::filter
```

::: callout-note
- Percent values were linearly rescaled to the range of gene counts to allow simultaneous visualization of enrichment percentage and DEG counts on the same axis.

- Percent was normalized to the maximum observed value and scaled to match the range of DEG counts for plotting.
:::

# Spider / Radar
 
 inspo from R graph gallery [here](https://r-graph-gallery.com/spider-or-radar-chart.html)
 
## Counts 
```{r, fig.height = 4, fig.width = 18}
plt_count <- ggplot(plot_df) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:4) * 5),
    color = "lightgrey"
  ) + 
  # Add bars to represent the number of genes mapped to each GO term
  # str_wrap(Term, 12) wraps the text so each line has at most 12 characters
  # (but it doesn't break long words!)
  geom_col(
    aes(
      x = str_wrap(goterm, 12),
      y = n_genes,
      fill = response
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = str_wrap(goterm, 12),
      y = 0,
      xend = str_wrap(goterm, 12),
      yend = 21,
      color = response
    ),
    linetype = "dashed"
  ) + 
  # Add dots to represent the percent enriched
  #geom_point(
  #  aes(
  #    x = str_wrap(Term, 12),
  #    y = percent_scaled
  #  ),
  #  size = 1.5, 
  #  color = "grey30"
  #) +
  facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0)
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 10),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.border = element_blank(),
    strip.text = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 1, 0.2, 1, "cm")
  )
  
  
plt_count  
```

```{r, fig.width = 12, fig.height = 4}

plt_anno <- plt_count +
# Annotate the bars and the lollipops so the reader understands the scaling
  annotate(
    x = 11, 
    y = 11,
    label = "Count of DE Genes",
    geom = "text",
    angle = -75,
    color = "gray30",
    size = 3
  ) +
  # Annotate custom scale inside plot
  annotate(
    x = 11.4, 
    y = 6, 
    label = "5", 
    geom = "text", 
    color = "gray40", 
    size = 3
  ) +
  annotate(
    x = 11.4, 
    y = 11, 
    label = "10", 
    geom = "text", 
    color = "gray40", 
    size = 3
  ) +
  annotate(
    x = 11.4, 
    y = 16, 
    label = "15", 
    geom = "text", 
    color = "gray40", 
    size = 3
  )+ 
  theme(
    panel.background = element_blank()
  )

plt_anno
```
## Enriched
```{r, fig.width=12, fig.height=4}
plt_enriched <- ggplot(circ_df) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:5) * 5),
    color = "lightgrey"
  ) + 
  geom_col(
    aes(
      x = str_wrap(Term, 12),
      y = n_genes,
      fill = Term
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = str_wrap(Term, 12),
      y = 0,
      xend = str_wrap(Term, 12),
      yend = 21,
      color = Term
    ),
    linetype = "dashed"
  ) + 
  # Add dots to represent the percent enriched
  geom_point(
    aes(
      x = str_wrap(Term, 12),
      y = percent_scaled
    ),
    size = 1.5, 
    color = "grey30"
  ) +
  facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0),
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 1, 0.2, 1, "cm")
  )
  
  
plt_enriched  
```

## Universe
```{r}
uni_df <- goslim_summary %>%
  drop_na(genes) %>%
  filter(module == "universe") %>%
  group_by(Term) %>%
  arrange(desc(n_genes))
```

```{r, include=FALSE}
plt_universe <- ggplot(uni_df) +
  # Make custom panel grid
  geom_hline(
    aes(yintercept = y), 
    data.frame(y = c(0:5) * 5),
    color = "lightgrey"
  ) + 
  geom_col(
    aes(
      x = str_wrap(Term, 12),
      y = n_genes,
      fill = Term
    ),
    position = "dodge2",
    show.legend = TRUE,
    alpha = .7
  ) +
  # Lollipop shaft for percent enriched
  geom_segment(
    aes(
      x = str_wrap(Term, 12),
      y = 0,
      xend = str_wrap(Term, 12),
      yend = 21,
      color = Term
    ),
    linetype = "dashed"
  ) + 
  # Add dots to represent the percent enriched
  #geom_point(
  #  aes(
  #    x = str_wrap(Term, 12),
  #    y = percent_scaled
  #  ),
  #  size = 1.5, 
   # color = "grey30"
  #) +
  #facet_wrap(~pattern, ncol = 3) +
  # Make it circular!
  coord_polar(clip = "off")+
  # Scale y axis so bars don't start in the center
  scale_y_continuous(
    limits = c(-2, 22),
    expand = c(0, 0),
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20)
  ) + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_text(size = 8),
    axis.ticks = element_blank(),
    legend.position = "none", 
    axis.title.x = element_blank(), 
    panel.background = element_blank(),
    panel.spacing.x = unit(2, "cm"),
    plot.margin = margin(0.2, 1, 0.2, 1, "cm")
  )
  
  
plt_universe
```


# Faceted lollipop ALL DEGs
```{r, fig.height= 12}
library("colorspace")
hcl_palettes(plot = TRUE)
```

```{r}
hclwizard()
```


```{r}
goslim_lollipop <- goslim_summary %>%
  drop_na(genes) %>% 
  filter(DEdueto == "leachate") %>% 
  mutate()

# number of goslim terms in goslim_lollipop
length(unique(goslim_lollipop$Term))
```

## Count
```{r, fig.height = 12}
ggplot(goslim_lollipop, aes(x = n_genes, y = Term, color = Percent)) +
  geom_segment(aes(xend = 0, yend = Term), size = 3) +
  geom_point(size = 4) +
  facet_wrap(~pattern, ncol = 3) +
  scale_color_continuous_sequential(palette = "Heat2", begin = 0.3, end = 1) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Lollipop length reflects counts of DEGs",
    x = "Counts of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Enriched
```{r, fig.height = 12}
ggplot(goslim_lollipop, aes(x = Percent, y = Term, color = n_genes)) +
  geom_segment(aes(xend = 0, yend = Term), size = 2) +
  geom_point(size = 3) +
  facet_wrap(~pattern, ncol = 3) +
  scale_color_continuous_sequential(palette = "Heat2", begin = 0.3, end = 1) +
  labs(
    title = "GOslim BP terms across expression patterns",
    subtitle = "Lollipop length reflects percent representation",
    x = "Percent of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Universe
```{r, fig.height=12}
ggplot(uni_df, aes(x = n_genes, y = Term)) +
  geom_segment(aes(xend = 0, yend = Term, color = n_genes), size = 2) +
  geom_point(aes(color = n_genes), size = 3) +
  scale_color_continuous_sequential(palette = "Heat2", begin = 0.2, end = 1) +
  labs(
    title = "GOslim BP terms in the background universe",
    subtitle = "Lollipop length reflects counts of genes mapped to GOslim term",
    x = "Counts of genes mapped to GOslim term",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

