---
title: "Gene ontology (GO) term analysis for differentially expressed genes"
subtitle: "using `goseq` with gene length bias correction"
author: "Sarah Tanja"
date: 2025-10-21
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

In the clustering analysis (07_cluster.qmd), we identified 3 distinct gene expression patterns among the 130 significant genes affected by leachate, 128 genes out of the 130 DEGs fit into one of the three patterns. 
These genes cluster based on their expression patterns across stage & leachate concentrations (control, low, mid, high).

Gene cluster breakdown:
- **Cluster 1**: 72 genes, peak in early gastrula
- **Cluster 2**: 25 genes, peak in prawnchip 
- **Cluster 3**: 31 genes, peak in cleavage

We now perform functional enrichment analysis on each cluster separately to understand the biological processes associated with each expression pattern.

We use the R package `goseq` for GO enrichment. 
`goseq` corrects for gene length bias - longer genes are more likely to be detected as DEGs in RNA-Seq simply because they accumulate more reads. 
`goseq` accounts for this by using a probability weighting function.

Run GOseq using:

DEG vector
Background (all tested genes)
Vector of gene lengths (from your annotation, GTF, eggNOG, etc.)

Get the corrected p-values and enriched GO terms.


## Inputs

- `sig_genes.csv` - A dataframe (list) of the 130 significant genes
- `gene_clusters.csv` - A dataframe mapping 128 genes to an expression cluster (1, 2, or 3)
- `all_genes.csv` - a dataframe (list) of all genes used in the DESeq2 analysis
- functional annotation: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.EggNog_results.txt.gz (EggNog)
- functional annotation: http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.genes.KEGG_results.txt.gz (KEGG)

## Outputs

- GO enrichment results for each of the 3 clusters
- Reduced GO terms (redundancy removed) for all DEGs and for each cluster
- Semantic similarity matrix for all DEGs and for each cluster

# Setup
##  Path
## Install packages

```{r}
output_path <- "../../output/07_enrichment/goseq/"
```


```{r, eval=FALSE}
# CRAN packages
#if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse')

# Bioconductor packages
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("GenomicRanges")
#BiocManager::install("Biostrings")
#BiocManager::install("goseq")
#BiocManager::install("rrvgo")
#BiocManager::install("org.Ce.eg.db") # c.elegans genome-wide annotation
```

## Load packages

```{r}
library(GenomicRanges)
library(Biostrings)
library(goseq)
library(rrvgo)
library(org.Ce.eg.db)
library(tidyverse)
```

## Load inputs

Read in gene lists

```{r}
# significant genes (130 total) (only those that are significant for leachate or stage:leachate)
sig_genes <- read.csv("../../output/06_deg/spawnnight/sig_genes.csv")

# Gene cluster assignments 
gene_clusters <- read.csv("../../output/06_deg/cluster/gene_clusters.csv")

# Full gene set tested in DESeq2 analysis
all_genes <- read.csv("../../output/06_deg/spawnnight/all_genes.csv")
```

Import EggNog for GO and KEGG annotation

```{r}
# Read in the file
EggNog <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")

# Rename "#query" to gene_id
EggNog <- EggNog %>%
  rename(gene_id="#query")

paste(nrow(EggNog), "genes have functional annotation")
```

Import fasta for sequence length annotation

```{r, eval=FALSE}
# Load transcriptome sequences to get gene lengths
all_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")

# Filter for the 130 significant genes
sig_seqs <- all_seqs[names(all_seqs) %in% sig_genes$gene_id]
```

Import DEG cluster annotations
...Add gene cluster annotation to gene list summaries

```{r}
# leachate
sig_summary_leachate <- sig_summary_leachate %>%
  left_join(gene_clusters, by = "gene_id")

write_csv(sig_summary_leachate, file.path(output_path, "sig_summary_leachate.csv"))
```

# Annotate gene lists

Match genes in all_genes to EggNog annotation dataframe

```{r}
# Annotate all genes with EggNog annotations
all_genes_anno <- left_join(all_genes, EggNog, by = "gene_id")
```

## Add gene lengths
::: callout-note
Gene lengths are required for goseq bias correction.
:::

```{r}
# Import GFF file to get gene lengths
# Check if file exists before importing
gff_file <- "../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3"

if (file.exists(gff_file)) {
  gff <- rtracklayer::import(gff_file)
  
  transcripts <- subset(gff, type == "transcript")
  
  transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE)
  
  transcript_lengths <- width(transcripts_gr)
  
  gene_id <- transcripts_gr$ID
  
  lengths <- cbind(gene_id, transcript_lengths)
  
  lengths <- as.data.frame(lengths)
  
  paste("Gene lengths extracted for", nrow(lengths), "transcripts")
} else {
  warning("GFF file not found. Gene length bias correction will not be available.")
  lengths <- data.frame(gene_id = character(), transcript_lengths = integer())
}
```

Add length and cluster info to the annotation dataframe
```{r}
all_genes_anno <- all_genes_anno %>% 
  left_join(lengths, by = "gene_id")%>% 
  left_join(gene_clusters, by = "gene_id")
```

```{r}
write_csv(all_genes_anno, file.path(output_path, "all_genes_anno.csv"))
```

## Explore GO and KEGG 
### GO coverage
How many of the genes are GO annotated?
```{r}
# Filter out any that lack GO term annotation
all_genes_GO <- all_genes_anno %>% 
  mutate(GOs = na_if(str_trim(GOs), "-")) %>%
  filter(!is.na(GOs), GOs != "")

# What is GO coverage in the background universe?
paste("GO annotated genes in DESeq background universe set:", 
      nrow(all_genes_GO), "/", 
      nrow(all_genes), "=",
      round(100*nrow(all_genes_GO)/nrow(all_genes),1),"%")

# What is GO coverage for all DEGs and for each cluster?
go_DE_genes <- all_genes_GO %>% filter(gene_id %in% sig_genes$gene_id)
go_cluster1 <- all_genes_GO %>% filter(cluster == 1)
go_cluster2 <- all_genes_GO %>% filter(cluster == 2)
go_cluster3 <- all_genes_GO %>% filter(cluster == 3)

paste("All significant genes:", nrow(go_DE_genes), "/", nrow(all_genes), "=", round(100*nrow(go_DE_genes)/nrow(all_genes),1),"%", "genes with GO annotation")

paste("Cluster 1:", nrow(go_cluster1), "/", nrow(all_genes), "=", round(100*nrow(go_cluster1)/nrow(all_genes),1),"%", "genes with GO annotation")

paste("Cluster 2:", nrow(go_cluster2), "/", nrow(all_genes), "=", round(100*nrow(go_cluster2)/nrow(all_genes),1),"%", "genes with GO annotation")

paste("Cluster 3:", nrow(go_cluster3), "/", nrow(all_genes), "=", round(100*nrow(go_cluster3)/nrow(all_genes),1),"%", "genes with GO annotation")
```

### KEGG coverage
How many of the background universe set are KEGG_Pathway annotated?
```{r}
# Filter out any that lack GO term annotation
all_genes_KEGG <- all_genes_anno %>% 
  mutate(KEGG_Pathway = na_if(str_trim(KEGG_Pathway), "-")) %>%
  filter(!is.na(KEGG_Pathway), KEGG_Pathway != "")

paste("KEGG annotated genes in DESeq background universe set:", 
      nrow(all_genes_KEGG), "/", 
      nrow(all_genes), "=",
      round(100*nrow(all_genes_KEGG)/nrow(all_genes),1),"%")
```


# Make gene universe vectors

For bias correction

The gene universe should be the full set of genes that were tested in the DESeq2 analysis (all_genes), including those that were not found to be differentially expressed. For goseq, we need to create a vector of all gene IDs in the DESeq2 dataset (include **all** genes, regardless of whether or not they have GO annotations), with a `1` value if DE and a `0` value if not.

nullp() models the probability of being DE as a function of length using the full DE universe.

Later, goseq() only tests GO terms using genes in gene2cat, but the PWF (bias correction) was learned from all the data.

So we need a background universe vectors for all DEGs, and each cluster. 
```{r}
# Universe for bias correction
# All 12221 genes from our filtered DEG test
universe_genes <- all_genes$gene_id
names(universe_genes) <- universe_genes
```

These are annotated dataframes subset to just DEGs and to each cluster
```{r}
DE_genes <- all_genes_anno %>% filter(gene_id %in% sig_genes$gene_id)
cluster1 <- all_genes_anno %>% filter(cluster == 1)
cluster2 <- all_genes_anno %>% filter(cluster == 2)
cluster3 <- all_genes_anno %>% filter(cluster == 3)
```

Create named vectors for DE status (1 = DE, 0 = not DE) for all DEGs and for each cluster
```{r}
vector_deg <- as.integer(universe_genes %in% DE_genes$gene_id)
names(vector_deg) <- universe_genes
vector_deg <- vector_deg[universe_genes]

vector_1 <- as.integer(universe_genes %in% cluster1$gene_id)
names(vector_1) <- universe_genes

vector_2 <- as.integer(universe_genes %in% cluster2$gene_id)
names(vector_2) <- universe_genes

vector_3 <- as.integer(universe_genes %in% cluster3$gene_id)
names(vector_3) <- universe_genes
```

# Make gene lengths vector

Make named vector of genes and gene lengths, only keep the genes present in the background universe (our 12221 DESeq2 tested genes!)
```{r}
gene_lengths <- as.integer(lengths$transcript_lengths)
names(gene_lengths) <- lengths$gene_id

gene_lengths <- gene_lengths[universe_genes]
paste("Transcript lengths for", length(gene_lengths), "genes")
```
```{r}
# Are there NAs in gene_lengths?
sum(is.na(gene_lengths))
```

The order of genes in gene_lengths must match the order in universe_genes, should return TRUE
```{r}
identical(names(gene_lengths), names(universe_genes))
identical(names(gene_lengths), names(vector_deg))   
identical(names(gene_lengths), names(vector_1))  
identical(names(gene_lengths), names(vector_2))
identical(names(gene_lengths), names(vector_3))
```

# Null probability weighting function (bias for gene length)

::: callout-note
Gene length bias correction accounts for the fact that longer genes are more likely to be detected as differentially expressed in RNA-seq data simply because they accumulate more reads. If length data is not available, goseq can still run but without this correction.
:::

DEgenes and gene_lengths (for nullp()) defined for all genes you tested in DESeq2 (our full “universe” of 12221 genes), including genes with no GO annotation.

A numeric vector of the same length as DEgenes

One number per gene, typically:

total gene length (sum of exons), or

transcript length

The names are again the same gene IDs
```{r}
# Calculate probability weighting function for all DEGs and for each cluster
pwf.1 <- nullp(vector_1, bias.data = gene_lengths)
pwf.2 <- nullp(vector_2, bias.data = gene_lengths)
pwf.3 <- nullp(vector_3, bias.data = gene_lengths)
pwf.deg <- nullp(vector_deg, bias.data = gene_lengths)
```
X axis is gene length
Y axis is probability of being DE given length

For genes in each length bin, GOseq estimates the fraction of genes that are DE. 
Points are `#DE genes in bin/#genes in bin` for genes with similar length. GOseq then fits a smooth curve through those points – this is the probability weighting function (PWF). This fitted curve is what GOseq uses internally to correct for length bias when doing GO enrichment.

Strong slope / curvature
→ there is substantial length bias
→ long genes are much more (or less) likely to be called DE than short ones
→ length correction is important.

Nearly flat line
→ little or no detectable length bias
→ the enrichment results will be similar to a standard hypergeometric ORA.

# GO Enrichment
## GO term mapping

This creates a dataframe that maps gene IDs to GO terms for all genes in the testing universe. There will be 2 columns, gene_id and GO term, with one GO term mapped to one gene. Since there are multiple GO terms per gene, genes will appear in multiple rows in these dataframes. Likewise, more than one gene may map to the same GO term.

Build the gene to GO term mappings for all genes in the universe
```{r}
gene2go <- all_genes_anno %>%
  dplyr::select(gene_id, GOs) %>%
  separate_rows(GOs, sep = ",") %>%
  filter(!is.na(GOs), GOs != "") %>%
  distinct(gene_id, GOs) %>%
  rename(gene = gene_id,
         category = GOs) %>%
  filter(gene %in% universe_genes) %>%
  as.data.frame()
```

Save GO term mapping
```{r}
write_csv(gene2go, file.path(output_path, "gene2go.csv"))
```

## Run `goseq`

Run GO enrichment for biological process terms

```{r}
GO.bp.1 <- goseq(
  pwf = pwf.1,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO.bp.2 <- goseq(
  pwf = pwf.2,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO.bp.3 <- goseq(
  pwf = pwf.3,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)

GO.bp.deg <- goseq(
  pwf = pwf.deg,
  gene2cat = gene2go,
  test.cats = c("GO:BP"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)
```

Use the Benjamini-Hochberg method to control false discovery rate

```{r}
GO.bp.1$bh_adjust <- p.adjust(GO.bp.1$over_represented_pvalue, method = "BH")
GO.bp.2$bh_adjust <- p.adjust(GO.bp.2$over_represented_pvalue, method = "BH")
GO.bp.3$bh_adjust <- p.adjust(GO.bp.3$over_represented_pvalue, method = "BH")
GO.bp.deg$bh_adjust <- p.adjust(GO.bp.deg$over_represented_pvalue, method = "BH")
```

Filter for biological process terms with valid p-values

```{r}
GO.BP.1.sig <- GO.bp.1 %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

GO.BP.2.sig <- GO.bp.2 %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

GO.BP.3.sig <- GO.bp.3 %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)

GO.BP.deg.sig <- GO.bp.deg %>%
  filter(ontology == "BP") %>%
  filter(!is.na(bh_adjust)) %>% 
  filter(bh_adjust < 0.05) %>%
  arrange(ontology, bh_adjust)
```

How many GO terms pass threshold for significantly enriched?
```{r}
paste("Cluster 1:", nrow(GO.BP.1.sig), "enriched BP terms (BH-adjusted p < 0.05)")
paste("Cluster 2:", nrow(GO.BP.2.sig), "enriched BP terms (BH-adjusted p < 0.05)")
paste("Cluster 3:", nrow(GO.BP.3.sig), "enriched BP terms (BH-adjusted p < 0.05)")
paste("All together:", nrow(GO.BP.deg.sig), "enriched BP terms (BH-adjusted p < 0.05)")
```

Nothing! Nothing is enriched... fuck. 

Ok. Well. Now we know. This indicates that our DEGs impacted by leachate don't all do particuler functions, or that there is no particular type of thing that these genes all do which are impacted by leachate.

> For our situation (≈12k background genes, ≈130 DEGs):
Small DEG set + custom annotation
If only, say, 40–60 of your 130 DEGs have GO annotations, and those are spread across many terms, there just may not be enough clustering in any single GO category.
Length bias correction is conservative
GOseq is explicitly correcting for bias. A lot of “borderline”-looking enrichments from simple hypergeometric tests disappear once you account for length.
Montipora annotation is sparse
With non-model genomes, GO coverage is often patchy. You could have a true biological pattern that doesn’t map cleanly to GO terms.
DE list is “diffuse”
If your DEGs represent many different processes at once (e.g., general stress + development + metabolism), each individual term may only get 2–3 hits → not enough to come out.

We can still describe what the DEGs do!

## Reduce redundancy

Calculate semantic similarity matrix with **all GO terms** that are present in DEGs
```{r}
# filter GO terms to those that are present in DEGs
GO.bp.deg.filt <- GO.bp.deg %>%
  filter(ontology == "BP") %>%
  filter(numDEInCat > 0) %>% 
  arrange(desc(numDEInCat))

simMatrix <- calculateSimMatrix(
    GO.bp.deg.filt$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )
```

```{r}
write_csv(as.data.frame(simMatrix), file.path(output_path, "simMatrix.csv"))
```

Create named scores vector
```{r}
# Create scores
scores <- setNames(
    -log10(pmax(GO.bp.deg.filt$bh_adjust, .Machine$double.xmin)),
    GO.bp.deg.filt$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms <- reduceSimMatrix(
    simMatrix,
    scores = scores,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```

Add in gene and cluster annotation to reduced GO terms present in DEGs
```{r}
reducedTerms_anno <- reducedTerms %>%
  left_join(GO.bp.deg.filt, by = c("go" = "category", "term")) %>% 
  left_join(gene2go, by = c("go" = "category")) %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id")) %>% 
  filter(!is.na(cluster))
```

```{r}
write_csv(reducedTerms, file.path(output_path, "reducedTerms_GO.csv"))
write_csv(reducedTerms_anno, file.path(output_path, "reducedTerms_GO_anno.csv"))
```

# KEGG Enrichment
## KEGG term mapping

This creates a dataframe that maps gene IDs to KEGG pathways for all genes in the testing universe. There will be 2 columns, gene_id and KEGG Pathway, with one KEGG Pathway mapped to one gene. 

Build the gene to KEGG Pathway mappings for all genes in the universe
```{r}
gene2kegg <- all_genes_anno %>%
  dplyr::select(gene_id, KEGG_Pathway) %>%
  separate_rows(KEGG_Pathway, sep = ",") %>%
  mutate(KEGG_Pathway = trimws(KEGG_Pathway)) %>%
  filter(!is.na(KEGG_Pathway), KEGG_Pathway != "") %>%
  # normalize: map04714 -> ko04714
  mutate(KEGG_Pathway = sub("^map", "ko", KEGG_Pathway)) %>%
  # (optional but recommended) keep only proper ko IDs
  filter(str_detect(KEGG_Pathway, "^ko\\d+")) %>%
  # now deduplicate gene–pathway pairs
  distinct(gene_id, KEGG_Pathway) %>%
  rename(
    gene     = gene_id,
    category = KEGG_Pathway
  ) %>%
  filter(gene %in% universe_genes) %>%
  as.data.frame()
```
Save KEGG pathway mapping
```{r}
write_csv(gene2kegg, file.path(output_path, "gene2kegg.csv"))
```

## Run `goseq`
```{r}
KEGG.deg <- goseq(
  pwf = pwf.deg,
  gene2cat = gene2kegg,
  test.cats = c("KEGG"),
  method = "Wallenius",
  use_genes_without_cat = TRUE
)
```

Adjust for multiple comparisons with Benjamini-Hochberg correction
```{r}
KEGG.deg$bh_adjust <- p.adjust(KEGG.deg$over_represented_pvalue, method = "BH")
```

Annotate KEGG pathways via genes and gene expression patterns
```{r}
KEGG.deg.anno <- KEGG.deg %>%
  left_join(gene2kegg, by = "category") %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id")) %>% 
  filter(!is.na(cluster))
```

```{r}
write_csv(KEGG.deg, file.path(output_path, "KEGG_deg.csv"))
write_csv(KEGG.deg.anno, file.path(output_path, "KEGG_deg_anno.csv"))
```

# Summary
No GO Terms or KEGG pathways were significantly enriched among the DEGs impacted by leachate exposure, either overall or within specific gene expression clusters.

## Next steps

- Visualize GO terms & KEGG Pathways for each cluster with `rrvgo` plotting functions
- Convert the GOseq output into a clusterProfiler-compatible object for visualization (dotplot, emapplot).


```{r}
sessionInfo()
```
