---
title: "07_goslim"
format: html
---

# Background

I matched clustered/module grouped DEGs to GO terms, and tested for enrichment using goseq().

I found that DEGs due to large shifts by stage had enriched GO terms only for prawnchip dip, earlygastrula up, or all stage DEGs combined. 
No other cluster/ module had enrichment.

I want to be able to show however what the DEGs for each cluster/moddule functionally represent, even if not statistically enriched.

For each cluster/module:

keep BP terms only

rank by raw p-value

take top N (e.g. 20–30)

require a minimum gene count

# Load libraries
```{r}
library(tidyverse)
library(GO.db)
```


# Inputs
```{r}
input_path <- "../../output/07_enrichment/goseq/"
```

gene to go term mapping
```{r}
gene2go <- read_csv(file.path(input_path, "gene2go.csv"))
```

```{r}
GO.bp.1 <- read_csv(file.path(input_path,"GO_bp_cluster1.csv"))
```

```{r}
head(GO.bp.1) %>% knitr::kable()
```
## Semantic similarity
## Reduce redundancy

Calculate semantic similarity matrix with **all GO terms** that are present in universe

```{r}
# filter GO terms to those that are BP only
GO.bp.1.filt <- GO.bp.1 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))
```

12599 GO BP terms... 

```{r}
simMatrix_12599BP <- calculateSimMatrix(
    GO.bp.1.filt$category,
    orgdb="org.Ce.eg.db",
    ont = "BP",
    method = "Rel"
  )
```
```{r}
identical(simMatrix, simMatrix.1)
```



```{r}
write_csv(as.data.frame(simMatrix.1), file.path(output_path, "simMatrix_1.csv"))
```


## leachate, cluster 1 (earlygastrula)
Create named scores vector
```{r}
# Create scores
scores.1 <- setNames(
    -log10(pmax(GO.bp.1.filt$bh_adjust, .Machine$double.xmin)),
    GO.bp.1.filt$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.1 <- reduceSimMatrix(
    simMatrix,
    scores = scores.1,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Add in gene and cluster annotation to reduced GO terms 
```{r}
reducedTerms.1_anno <- reducedTerms.1 %>%
  left_join(GO.bp.1.filt, by = c("go" = "category", "term")) %>% 
  left_join(gene2go, by = c("go" = "category")) %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id"))
```

```{r}
write_csv(reducedTerms, file.path(output_path, "reducedTerms_GO.csv"))
write_csv(reducedTerms_anno, file.path(output_path, "reducedTerms_GO_anno.csv"))
```

Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.1,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```

## leachate, cluster 2 (prawnchip)

```{r}
# filter GO terms to those that are BP only
GO.BP.2 <- GO.bp.2 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.2 <- setNames(
    -log10(pmax(GO.BP.2$bh_adjust, .Machine$double.xmin)),
    GO.BP.2$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.2 <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.2,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```

Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.2,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```
## leachate, cluster 3 (cleavage)

```{r}
# filter GO terms to those that are BP only
GO.BP.3 <- GO.bp.3 %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.3 <- setNames(
    -log10(pmax(GO.BP.3$bh_adjust, .Machine$double.xmin)),
    GO.BP.3$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.3 <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.3,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Add in gene and cluster annotation to reduced GO terms 
```{r}
reducedTerms.1_anno <- reducedTerms.1 %>%
  left_join(GO.bp.1.filt, by = c("go" = "category", "term")) %>% 
  left_join(gene2go, by = c("go" = "category")) %>% 
  left_join(gene_clusters, by = c("gene" = "gene_id"))
```

```{r}
write_csv(reducedTerms, file.path(output_path, "reducedTerms_GO.csv"))
write_csv(reducedTerms_anno, file.path(output_path, "reducedTerms_GO_anno.csv"))
```

Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.3,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```
## stage, cleavage

```{r}
# filter GO terms to those that are BP only
GO.BP.cleavage <- GO.bp.cleavage %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.cleavage <- setNames(
    -log10(pmax(GO.BP.cleavage$bh_adjust, .Machine$double.xmin)),
    GO.BP.cleavage$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.cleavage <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.cleavage,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.cleavage,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```

## stage, prawnchip up 

```{r}
# filter GO terms to those that are BP only
GO.BP.prawnchip_up <- GO.bp.prawnchip_up %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.prawnchip_up <- setNames(
    -log10(pmax(GO.BP.prawnchip_up$bh_adjust, .Machine$double.xmin)),
    GO.BP.prawnchip_up$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.prawnchip_up <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.prawnchip_up,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.prawnchip_up,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```
## stage, prawnchip down 

```{r}
# filter GO terms to those that are BP only
GO.BP.prawnchip_dip <- GO.bp.prawnchip_dip %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.prawnchip_dip <- setNames(
    -log10(pmax(GO.BP.prawnchip_dip$bh_adjust, .Machine$double.xmin)),
    GO.BP.prawnchip_dip$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.prawnchip_dip <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.prawnchip_dip,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.prawnchip_dip,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```

## stage, earlygastrula 

```{r}
# filter GO terms to those that are BP only
GO.BP.earlygastrula <- GO.bp.earlygastrula %>%
  filter(ontology == "BP") %>%
  arrange(desc(numDEInCat))

```

Create named scores vector
```{r}
# Create scores
scores.earlygastrula <- setNames(
    -log10(pmax(GO.BP.earlygastrula$bh_adjust, .Machine$double.xmin)),
    GO.BP.earlygastrula$category
  )
```

Reduce terms with similarity matrix
```{r}
# Reduce terms for all DEGs 
reducedTerms.earlygastrula <- reduceSimMatrix(
    simMatrix.1,
    scores = scores.earlygastrula,
    threshold = 0.7,
    orgdb = "org.Ce.eg.db"
  )
```
Visualize reduced terms
```{r}
scatterPlot(simMatrix.1, reducedTerms.earlygastrula,
            onlyParents = TRUE,
            addLabel = TRUE,
            labelSize = 4)
```


# Background universe
```{r}
# Distance from similarity matrix
dist <- as.dist(1 - simMatrix.1)

# PCoA to 2 dimensions
coord <- cmdscale(dist, k = 2)

# Convert to dataframe with GO terms
df_coords <- data.frame(go = rownames(simMatrix.1), Dim1 = coord[,1], Dim2 = coord[,2])
```

```{r}
anno <- reducedTerms.anno %>%
  select(!gene) %>% 
  group_by(go) %>%
  mutate(pattern = paste0("GEP", pattern)) %>% 
  mutate(value = 1L) %>% 
  distinct(go, pattern, .keep_all = TRUE) %>% 
  pivot_wider(
    names_from  = pattern,
    values_from = value,
    values_fill = 0
  )
```