---
title: "Step 1: Get *Montipora capitata* genome files"
subtitle: "Use `curl` and `wget` to pull reference genome"
author: "Sarah Tanja"
date: 10/14/2024
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output))        
```

# Background

*Montipora capitata* Genome version V3, Rutgers Univrsity: <http://cyanophora.rutgers.edu/montipora/>

[`Montipora_capitata_HIv3.assembly.fasta`](https://owl.fish.washington.edu/halfshell/genomic-databank/Montipora_capitata_HIv3.assembly.fasta) (745MB)

-   MD5 checksum: `99819eadba1b13ed569bb902eef8da08`

Genome publication @Stephens2022-sn

::: callout-note
A compiled list of genomic resources for *Montipora capitata* can be found on the Robert's Lab handbook [here](https://robertslab.github.io/resources/Genomic-Resources/#montipora-capitata:~:text=mmag_pilon_scaffolds.fasta.fai-,Montipora%20capitata,-Genomes%3A)
:::

# Goals

Download, check file integrity, and unzip genome reference files to our `input/genome` directory.

## Inputs

-   none

## Outputs

-   Montipora_capitata_HIv3.assembly.fasta

    -   `../../input/genome/Montipora_capitata_HIv3.assembly.fasta`

-   Montipora_capitata_HIv3.genes_fixed.gff3

    -   `../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3`

# Get reference genome files

This code grabs the *Montipora capitata* fasta file (rna.fna) of genes.

```{r, engine='bash'}
# change to work in data directory
cd ../../input/genome
wget http://cyanophora.rutgers.edu/montipora/Montipora_capitata_HIv3.assembly.fasta.gz
```

## GFF

[`Montipora_capitata_HIv3.genes.gff3`](https://owl.fish.washington.edu/halfshell/genomic-databank/Montipora_capitata_HIv3.genes.gff3) (67MB)

-   MD5 checksum: `5f6b80ba2885471c8c1534932ccb7e84`

Obtain `Montipora_capitata_HIv3.genes_fixed.gff3` file by downloading from GitHub.

```{r, engine='bash'}
cd ../../input/genome
wget https://github.com/AHuffmyer/EarlyLifeHistory_Energetics/raw/master/Mcap2020/Data/TagSeq/Montipora_capitata_HIv3.genes_fixed.gff3.gz
```

This was generated by running the original gff file `Montipora_capitata_HIv3.genes.gff3.gz` through this script in R: <https://github.com/AHuffmyer/EarlyLifeHistory_Energetics/blob/master/Mcap2020/Scripts/TagSeq/Genome_V3/fix_gff_format.Rmd>

Unzip gff and genome file

```{r, engine='bash'}
cd ../../input/genome
gunzip Montipora_capitata_HIv3.genes_fixed.gff3.gz
gunzip Montipora_capitata_HIv3.assembly.fasta.gz
```

## Check file integrity with `md5sum`

::: callout-info
[Learn How to Generate and Verify Files with MD5 Checksum in Linux](https://www.tecmint.com/generate-verify-check-files-md5-checksum-linux/)
:::

```{r, engine='bash', eval=TRUE}
cd ../../input/genome
md5sum *Montipora_capitata*
```

## Checkout genome GFF format

```{r, engine = 'bash', eval=TRUE}
cd ../../input/genome
head -10 Montipora_capitata_HIv3.genes_fixed.gff3
```

# Summary & Next Steps

We now have the V3 reference genome files located at `input/genome`!

::: callout-info
**Next Step:** Go to `01.2_get_sequences` to download raw FASTQ files from Azenta
:::
