---
title: "Gene ontology (GO) enrichment plots"
subtitle: "We've identified the DEGs... let's visualize what they do!"
author: "Sarah Tanja"
date: 2025-09-30
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output))        
```

# Background

## Inputs

-   reduced terms
-   semantic similarity matrix

## Outputs

# Setup

## Install packages

```{r}
## CRAN packages-----------
#install.packages("GOplot")
#install.packages("plotly")
#install.packages("tidyverse")

## Bioconductor packages---------
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("rrvgo")
```

## Load packages

```{r}
library(GOplot)
library(plotly)
library(tidyverse)
library(rrvgo)
library(kableExtra)
```

## Load inputs

```{r}
## Reduced Terms
### Interaction
reducedTerms.int <- read.csv("../../output/08_go/reducedTerms_int.csv")

### Interaction patterns
reducedTerms.1 <- read.csv("../../output/08_go/reducedTerms_1.csv")
reducedTerms.2 <- read.csv("../../output/08_go/reducedTerms_2.csv")
reducedTerms.3 <- read.csv("../../output/08_go/reducedTerms_3.csv")

### Contrasts
reducedTerms.lvccl <- read.csv("../../output/08_go/reducedTerms_lvccl.csv")
reducedTerms.mvccl <- read.csv("../../output/08_go/reducedTerms_mvccl.csv")
reducedTerms.hvccl <- read.csv("../../output/08_go/reducedTerms_hvccl.csv")
reducedTerms.hvceg <- read.csv("../../output/08_go/reducedTerms_hvceg.csv")

## Similarity matrices
### Interaction
simMatrix.int <- read_csv("../../output/08_go/simMatrix_int.csv") %>% 
  column_to_rownames("GO") %>% as.matrix()  # uses the GO column as rownames
  

### Interaction patterns
simMatrix.1 <- read_csv("../../output/08_go/simMatrix_1.csv")%>% 
  column_to_rownames("GO") %>% as.matrix()  # uses the GO column as rownames
simMatrix.2 <- read_csv("../../output/08_go/simMatrix_2.csv")%>% 
  column_to_rownames("GO") %>% as.matrix() # uses the GO column as rownames
simMatrix.3 <- read_csv("../../output/08_go/simMatrix_3.csv")%>% 
  column_to_rownames("GO") %>% as.matrix() # uses the GO column as rownames

### Contrasts
simMatrix.lvccl <- read_csv("../../output/08_go/simMatrix_lvccl.csv")%>% 
  column_to_rownames("GO") %>% as.matrix() # uses the GO column as rownames
simMatrix.mvccl <- read_csv("../../output/08_go/simMatrix_mvccl.csv")%>% 
  column_to_rownames("GO") %>% as.matrix() # uses the GO column as rownames
simMatrix.hvccl <- read_csv("../../output/08_go/simMatrix_hvccl.csv")%>% 
  column_to_rownames("GO") %>% as.matrix() # uses the GO column as rownames
simMatrix.hvceg <- read_csv("../../output/08_go/simMatrix_hvceg.csv")%>% 
  column_to_rownames("GO") %>% as.matrix() # uses the GO column as rownames
```

> The rows of simMatrix.xxx should correspond (in order and identity) to
> the rows of reducedTerms.xxx. Ensure that the go column is character,
> not factor. Ensure simMatrix.xxx is a data.frame or matrix, and
> reducedTerms.xxx is a data.frame, not a tibble!\*which is why we load
> it wirh read.csv, (which loads a data.frame) and not read_csv (which
> loads a tibble).

# Plotting & interpretation

[Using the rrvgo
package](https://bioconductor.org/packages/devel/bioc/vignettes/rrvgo/inst/doc/rrvgo.html)

## Similarity matrix heatmap

Plot similarity matrix as a heatmap, with clustering of columns of rows
turned on by default (thus arranging together similar terms).

```{r}
heatmapPlot(simMatrix.int,
            reducedTerms.int,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

```{r}
heatmapPlot(simMatrix.1,
            reducedTerms.1,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

```{r}
heatmapPlot(simMatrix.2,
            reducedTerms.2,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

```{r}
heatmapPlot(simMatrix.3,
            reducedTerms.3,
            annotateParent=TRUE,
            annotationLabel="parentTerm",
            fontsize=6)
```

## `rrvgo` scatter plots

Plot GO terms as scattered points. Distances between points represent
the similarity between terms, and axes are the first 2 components of
applying a PCoA to the (di)similarity matrix. Size of the point
represents the provided scores or, in its absence, the number of genes
the GO term contains.

### Interaction

```{r}
scatterPlot(simMatrix.int, reducedTerms.int,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### Expression pattern 1

```{r}
scatterPlot(simMatrix.1, reducedTerms.1,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### Expression pattern 2

```{r}
scatterPlot(simMatrix.2, reducedTerms.2,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### Expression pattern 3

```{r}
scatterPlot(simMatrix.3, reducedTerms.3,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### Low vs control: cleavage

```{r}
scatterPlot(simMatrix.lvccl, reducedTerms.lvccl,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### Mid vs control: cleavage

```{r}
scatterPlot(simMatrix.mvccl, reducedTerms.mvccl,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### High vs control: cleavage

```{r}
scatterPlot(simMatrix.hvccl, reducedTerms.hvccl,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

### High vs control: early gastrula

```{r}
scatterPlot(simMatrix.hvceg, reducedTerms.hvceg,
            onlyParents = FALSE,
            addLabel = TRUE,
            labelSize = 4)
```

# Custom PCA scatter plots

## Interaction "5 clusters" summary

I want a dataframe that groups my GO terms based on the 5 main clusters
displayed in this PCA. The scatterPlot uses a dimensionality reduction
(PCoA) approach to project GO terms into 2D space and then visually
groups the points into a smaller number of major clusters. These visual
clusters are typically broader groupings, which correspond to
higher-level semantic similarity rather than the narrow clusters from
the cut tree. A possible approach if you want to create a cluster
assignment dataframe matching the 5 visual clusters is to: - Extract the
2D coordinates from the scatterPlot dimensionality reduction output, -
Apply a clustering method on the 2D coordinates (e.g., k-means with
k=5), - Create a new dataframe linking each GO term to its visual
cluster. This approach aligns the clustering with the displayed
scatterPlot clusters rather than the original fine-grained clusters.

```{r}
# Distance from similarity matrix
dist <- as.dist(1 - simMatrix.int)

# PCoA to 2 dimensions
coord <- cmdscale(dist, k = 2)

# Convert to dataframe with GO terms
df_coords <- data.frame(go = rownames(simMatrix.int), Dim1 = coord[,1], Dim2 = coord[,2])
```

Cluster the terms visually in 2D space Use k-means clustering on the 2D
coordinates with k = 5 (number of desired clusters).

```{r}
set.seed(123)  # for reproducibility
clusters <- kmeans(df_coords[, c("Dim1", "Dim2")], centers = 5)
df_coords$cluster <- as.factor(clusters$cluster)
```

Combine cluster info with other GO term data Merge this dataframe with
your reducedTerms or other GO annotation information as needed.

```{r}
final_df <- merge(df_coords, reducedTerms.int, by.x = "go", by.y = "go")
```

This dataframe `final_df` will now include each GO term, its 2D
coordinates used for visualization, and its assigned cluster
corresponding to the 5 major groups in the scatterPlot.

### Interactive plot

For exploration

```{r}
p <- ggplot(final_df, aes(
    Dim1, Dim2,
    color = parentTerm,
    size  = score,
    text  = paste0(
      "GO: ", go,
      "<br>Parent Term: ", parentTerm,
      "<br>Term: ", term,
      "<br>Score: ", round(score, 2)
    )
  )) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term",
       size  = "Score",
       title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)

ggplotly(p, tooltip = "text")
```

### Table of top parent terms

```{r}
# Summarize cluster content by parent term
cluster_summary <- final_df %>%
  group_by(cluster.x, parentTerm) %>%
  summarise(
    n_terms = n(),                     # how many GO terms in this parent within cluster
    mean_score = mean(score, na.rm = TRUE),  # average enrichment strength
    max_score = max(score, na.rm = TRUE),    # strongest term in that parent
    .groups = "drop"
  ) %>%
  arrange(cluster.x, desc(n_terms))

# For each cluster, show top 5 parent terms by count (or max score)
top_parent_terms_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = n_terms, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(n_terms))

knitr::kable(top_parent_terms_by_cluster)
```

### Table of top terms

```{r}
# Summarize cluster content by parent term
cluster_summary <- final_df %>%
  group_by(cluster.x, term) %>%
  summarise(
    n_terms = n(),                     # how many GO terms in this parent within cluster
    mean_score = mean(score, na.rm = TRUE),  # average enrichment strength
    max_score = max(score, na.rm = TRUE),    # strongest term in that parent
    .groups = "drop"
  ) %>%
  arrange(cluster.x, desc(n_terms))

# For each cluster, show top 5 terms by count (or max score)
top_terms_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = n_terms, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(n_terms))

knitr::kable(top_terms_by_cluster)
```

### Table of top scores

```{r}
# For each cluster, show top 5 parent terms by count (or max score)
top_score_by_cluster <- cluster_summary %>%
  group_by(cluster.x) %>%
  slice_max(order_by = mean_score, n = 5) %>% # top 5 terms for each of the 5 clusters
  arrange(cluster.x, desc(mean_score))

knitr::kable(top_score_by_cluster)
```

## Manuscript plots

```{r}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, size = score, alpha = 0.4)) +
  geom_point() +
  coord_equal() +
  labs(color = "Cluster (k=5)", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12)
```

```{r, fig.height=10, fig.width=15}
# 5) Quick visual check (should resemble your scatterPlot clusters)
ggplot(final_df, aes(Dim1, Dim2, color = parentTerm, size = score)) +
  geom_jitter(alpha = 0.5)+
  coord_equal() +
  labs(color = "Parent Term", title = "GO parent terms clustered in PCoA space") +
  theme_minimal(base_size = 12)+
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 9))
```

```{r, fig.height=17}
ggplot(final_df, aes(Dim1, Dim2, color = term, size = score)) +
  geom_point(alpha = 0.5) +
  coord_equal() +
  labs(color = "Parent Term", title = "GO terms clustered in PCoA space") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",      # move below
    legend.box = "vertical",         # stack if multiple guides
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9)
  ) +
  guides(                                  # hide size legend (optional)
    color = guide_legend(nrow = 40, byrow = TRUE)     # wrap legend into rows
  )
```

::: column-margin
::: callout-note
How to interpret distances

Closer points = more semantically similar GO terms. That means the gene
sets they represent overlap more or describe conceptually related
processes.

Farther points = less similar biologically. Their gene sets don't
overlap much, and the ontology sees them as describing different
processes. The PCoA projection does not use your logFCs or sample-level
values --- it uses GO term relationships (semantic similarity).

1.  What your clusters represent

The clusters from the PCoA + k-means are based on the pairwise semantic
similarity scores across all GO terms.

These are data-driven groupings that capture overall gene overlap +
ontology proximity.

The algorithm doesn't "know" about your chosen parentTerm column --- it
just groups terms that have similar meaning.

2.  What your parentTerm variable is

The "parent term" you merged in is a single hierarchical label from the
GO DAG (graph).

A GO term can belong under multiple higher-level categories in the
ontology, but in your dataset it's probably annotated with just one
chosen parent (often the most specific or most frequent).

Because the GO graph is not a strict tree, related terms may have
different parents --- so they don't collapse neatly into one parentTerm
cluster.

3.  Why they don't line up

Ontology is many-to-many: two GO terms might be semantically very close
(high overlap) but come from different parents in the hierarchy.

Cluster ≠ hierarchy: k-means clusters cut across the graph structure;
they group by similarity, not by the official ontology tree.

Parent terms are broad: "cellular process" or "metabolic process" can
cover huge swathes of biology, so those categories span multiple
similarity clusters.

4.  How to check alignment

Look at your clusters and check: are terms within a cluster biologically
coherent even if they span different parents? Often they are --- they
share overlapping gene sets or related functions.

The clusters are "data-driven neighborhoods" of related GO terms. The
parent terms are "administrative labels" from the GO DAG. They don't
always coincide because biology (and the GO ontology) is more tangled
than a simple tree. Clusters in the 2D space: show neighborhoods of
semantically similar terms (terms that describe similar biology and
often share many genes).

Point size (score): tells you which terms were more strongly enriched.

Color by parentTerm: just tags each dot with the one "parent" you merged
in. The fact that colors are mixed within a cluster shows that ontology
categories don't always line up with semantic neighborhoods.
:::
:::

::: callout-note

Based on frequency of enriched terms, the clusters can be described: - 1

:   Developmental, cellular, nuclear division - 2 : Regulation of
    metabolic processes - 3 : cellular stress response , response to
    chemical - 4 : metabolic / catabolic process - 5 : regulation cell
    cycle communication
:::

```{r}
tbl_min <- top_score_by_cluster %>%
  transmute(
    Cluster = cluster.x,
    `Parent term` = parentTerm,
    n = as.integer(n_terms),
    `Mean score` = round(mean_score, 1),
    `Max score` = round(max_score, 1)
  ) %>%
  arrange(Cluster, desc(`Mean score`), desc(`Max score`))

kable(
  tbl_min,
  format = "pipe",                            # <- Pandoc table -> native Word/Docs
  align  = c("r", "l", "r", "r", "r"),
  caption = "Top GO terms by cluster"
)
```

## Treemap plot

Treemaps are space-filling visualization of hierarchical structures. The
terms are grouped (colored) based on their parent, and the space used by
the term is proportional to the score. Treemaps can help with the
interpretation of the summarized results and also comparing different
sets of GO terms.

### Interactions

```{r}
treemapPlot(reducedTerms.int)
```

#### pattern 1

```{r}
treemapPlot(reducedTerms.1)
```

#### pattern 2

```{r}
treemapPlot(reducedTerms.2)
```

#### pattern 3

```{r}
treemapPlot(reducedTerms.3)
```

### Contrasts

#### Low vs control: cleavage

```{r}
treemapPlot(reducedTerms.lvccl)
```

#### Mid vs control: cleavage

```{r}
treemapPlot(reducedTerms.mvccl)
```

#### High vs control: cleavage

```{r}
treemapPlot(reducedTerms.hvccl)
```

#### High vs control: early gastrula

```{r}
treemapPlot(reducedTerms.hvceg)
```

# Summary & next steps

Produced visualizations: - Heatmap of GO term similarity. - Scatter plot
grouping GO terms into 5 major clusters. - Treemap to summarize enriched
functions. Outputs:

-   Tables of enriched GO terms, reduced clusters, semantic similarity
    matrices, and gene-to-term assignments.
-   Multiple graphical summaries (heatmap, PCA scatter, treemap)

```{r}
sessionInfo()
```
