---
title: "KEGG Ortholog (KO) pathway analysis for gene expression clusters"
subtitle: "Characterizing the 3 gene expression patterns found in clustering analysis"
author: "Sarah Tanja"
date: 2025-11-17
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

```{r}
BiocManager::install("clusterProfiler")
BiocManager::install("enrichplot")
library(clusterProfiler)
library(enrichplot)
library(GOSemSim)
library(tidyverse)
```
## Load inputs

## Load inputs

Read in gene lists

```{r}
# significant genes (130 total) (only those that are significant for leachate or stage:leachate)
sig_genes <- read.csv("../../output/06_deg/spawnnight/sig_all.csv")

# Gene cluster assignments 
gene_clusters <- read.csv("../../output/06_deg/cluster/gene_clusters.csv")

# Full gene set tested in DESeq2 analysis
all_genes <- read.csv("../../output/06_deg/spawnnight/all_genes.csv")
```

Import fasta 

```{r, eval=FALSE}
# Load transcriptome
all_seqs <- readDNAStringSet("../../input/genome/Montipora_capitata_HIv3.assembly.fasta")

# Filter for the 130 significant genes
sig_seqs <- all_seqs[names(all_seqs) %in% sig_genes$gene_id]
```

Import EggNog annotation

```{r}
# Read in the file
EggNog_anno <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")

# Rename "#query" to gene_id
EggNog_anno <- EggNog_anno %>%
  rename(gene_id="#query")

paste(nrow(EggNog_anno), "genes have functional annotation")
```

# Annotate significant genes

Match genes in sig_genes to annotation dataframe

```{r}
# Annotate significant genes
sig_genes_anno <- inner_join(sig_genes, EggNog_anno, by = "gene_id")

# Filter out any that lack GO term annotation
sig_genes_annofilt <- sig_genes_anno %>% 
  mutate(GOs = na_if(str_trim(GOs), "-")) %>%
  filter(!is.na(GOs), GOs != "")

paste("Annotated genes:", 
      nrow(sig_genes_anno), "/", 
      nrow(sig_genes), "=",
      round(100*nrow(sig_genes_anno)/nrow(sig_genes),1),"%")

paste("GO annotated genes:", 
      nrow(sig_genes_annofilt), "/", 
      nrow(sig_genes), "=",
      round(100*nrow(sig_genes_annofilt)/nrow(sig_genes),1),"%")
```

## Add gene lengths

::: callout-note
Gene lengths are required for goseq bias correction.
:::

```{r}
# Import GFF file to get gene lengths
# Check if file exists before importing
gff_file <- "../../input/genome/Montipora_capitata_HIv3.genes_fixed.gff3"

if (file.exists(gff_file)) {
  gff <- rtracklayer::import(gff_file)
  
  transcripts <- subset(gff, type == "transcript")
  
  transcripts_gr <- makeGRangesFromDataFrame(transcripts, keep.extra.columns=TRUE)
  
  transcript_lengths <- width(transcripts_gr)
  
  gene_id <- transcripts_gr$ID
  
  lengths <- cbind(gene_id, transcript_lengths)
  
  lengths <- as.data.frame(lengths)
  
  paste("Gene lengths extracted for", nrow(lengths), "transcripts")
} else {
  warning("GFF file not found. Gene length bias correction will not be available.")
  lengths <- data.frame(gene_id = character(), transcript_lengths = integer())
}
```

Add length to the annotation dataframe

```{r}
sig_genes_af <- sig_genes_annofilt %>% left_join(lengths, by = "gene_id")
```

## Add cluster assignments

Join cluster information with annotated genes

```{r}
sig_genes_afc <- inner_join(sig_genes_af, gene_clusters, by = "gene_id")
str(sig_genes_afc)
```
::: callout-note
`sig_genes_afc` is now the dataframe with GO annotated, filtered, and matched to gene pattern clusters
:::

Save GO annotated dataframe

```{r}
write_csv(sig_genes_afc, file.path(output_path, "sig_genes_GOannotated.csv")
```



Everything from output/07_go/
```{r}
path <- "../../output/07_go/"
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
nm    <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}
```

```{r}
gene_clusters <- read_csv("../../output/06_deg/cluster/gene_clusters.csv")

gene_lists <- split(gene_clusters$gene_id, gene_clusters$cluster)
```


```{r}
xx <- compareCluster(
  geneCluster   = gene_lists,
  fun           = "enrichKEGG",        # or enrichKEGG, etc.
  OrgDb         = orgdb_object,
  keyType       = "ENSEMBL",         # or whatever your IDs are
  ont           = "BP",
  pvalueCutoff  = 0.05
)
```


