---
title: "KEGG Ortholog (KO) pathway analysis for gene expression clusters"
subtitle: "Characterizing the 3 gene expression patterns found in clustering analysis"
author: "Sarah Tanja"
date: 2025-11-17
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

For each of GO and KEGG you want:

Background / universe

universe_genes <- rownames(dds) (or similar)

DEG + annotation table in long form

One row per (gene_id, GO_id) and (gene_id, KEGG_id).

TERM2GENE & TERM2NAME tables

GO: (GO_id, gene_id) and (GO_id, GO_term)

KEGG: (KEGG_id, gene_id) and (KEGG_id, KEGG_term)

Clustered gene lists

gene_clusters <- split(deg_annot$gene_id, deg_annot$cluster) (with dedup)

Run compareCluster(..., fun="enricher", TERM2GENE=..., TERM2NAME=...)

One call for GO, one for KEGG.

Visualise with dotplot(), emapplot(), cnetplot(), etc.

Breakdown of the gene ID format

Montipora_capitata_HIv3___TS.g9699.t2

`Montipora_capitata`	Species (your experimental species)
`HIv3`	Genome assembly version — Hawai‘i version 3 (the one commonly used in RNA-seq pipelines now)
`___TS`	Annotation source / transcript set (often "TS" = transcript set, sometimes “TS” refers to Trinity/STRINGTIE if that was used in comprehensive assemblies)
`g9699`	Gene ID (numbered gene)
`t2`	Transcript number 2 (transcript isoform of g9699)

# Setup
## Set paths
```{r}
output_path <- "../../output/08_kegg/"
```


## Load libraries
```{r}
#BiocManager::install("clusterProfiler")
#BiocManager::install("enrichplot")

library(clusterProfiler)
library(enrichplot)
library(GOSemSim)
library(GenomicRanges)
library(Biostrings)
library(tidyverse)
```

## Load inputs

Read in gene lists
```{r}
# Full gene set tested in DESeq2 analysis
all_genes <- read.csv("../../output/06_deg/spawnnight/all_genes.csv")

# significant genes (130 total) (only those that are significant for leachate or stage:leachate)
sig_genes <- read.csv("../../output/06_deg/spawnnight/sig_genes.csv")
sig_summary <- read.csv("../../output/06_deg/spawnnight/sig_summary.csv")

# Gene cluster assignments 
gene_clusters <- read.csv("../../output/06_deg/cluster/gene_clusters.csv")
```

Add cluster to sig_summary
```{r}
sig_summary <- sig_summary %>% 
  left_join(gene_clusters, by = "gene_id")
```

Import EggNog annotation
```{r}
# Read in the file
EggNog <- read_tsv("../../input/annotations/Montipora_capitata_HIv3.genes.EggNog_results.txt")

# Rename "#query" to gene_id
EggNog <- EggNog %>%
  rename(gene_id="#query")

paste(nrow(EggNog), "genes have functional annotation")
```

KO = KEGG Orthology ID
Example: K01900 (Enolase)

Represents gene-level functional units.

One KO = one “orthologous gene family” with a well-defined function

KO is the atomic KEGG annotation

Pathways are large biological processes composed of many KOs.

High-level, very interpretable (e.g., apoptosis, TCA cycle, Wnt signaling)

Based on curated networks

Pathways can be metabolic or signaling

# Annotate genes

Match genes in all_genes to EggNog annotation dataframe
```{r}
# Annotate significant genes
all_genes_anno <- inner_join(all_genes, EggNog, by = "gene_id")

# How many of the background genes are annotated by either GO or KEGG?
paste("Annotated genes from DESeq background universe set:", 
      nrow(all_genes_anno), "/", 
      nrow(all_genes), "=",
      round(100*nrow(all_genes_anno)/nrow(all_genes),1),"%")
```

How many of the background universe set are GO annotated?
```{r}
# Filter out any that lack GO term annotation
all_genes_GO <- all_genes_anno %>% 
  mutate(GOs = na_if(str_trim(GOs), "-")) %>%
  filter(!is.na(GOs), GOs != "")

paste("GO annotated genes in DESeq background universe set:", 
      nrow(all_genes_GO), "/", 
      nrow(all_genes), "=",
      round(100*nrow(all_genes_GO)/nrow(all_genes),1),"%")
```
Save GO annotated dataframe
```{r}
write_csv(all_genes_GO, file.path(output_path, "all_genes_GO.csv")
```

How many of the background universe set are KEGG_Pathway annotated?
```{r}
# Filter out any that lack GO term annotation
all_genes_KEGG <- all_genes_anno %>% 
  mutate(KEGG_Pathway = na_if(str_trim(KEGG_Pathway), "-")) %>%
  filter(!is.na(KEGG_Pathway), KEGG_Pathway != "")

paste("KEGG annotated genes in DESeq background universe set:", 
      nrow(all_genes_KEGG), "/", 
      nrow(all_genes), "=",
      round(100*nrow(all_genes_KEGG)/nrow(all_genes),1),"%")
```
Save KEGG_Pathway annotated dataframe
```{r}
write_csv(all_genes_KEGG, file.path(output_path, "all_genes_KEGG_Pathway.csv")
```
## Add cluster assignments

Join cluster information with annotated genes
```{r}
sig_genes_afc <- inner_join(sig_genes_af, gene_clusters, by = "gene_id")
str(sig_genes_afc)
```

::: callout-note
`sig_genes_afc` is now the dataframe with GO annotated, filtered, and matched to gene pattern clusters
:::





Everything from output/07_go/
```{r}
path <- "../../output/07_go/"
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
nm    <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}
```

```{r}
gene_clusters <- read_csv("../../output/06_deg/cluster/gene_clusters.csv")

gene_lists <- split(gene_clusters$gene_id, gene_clusters$cluster)
```

# ORA
https://chatgpt.com/share/691d3dec-fff0-8008-8a0b-fe8f672fa716

## What ORA needs (per Xu et al.)

```{r}
gene_clusters <- read_csv("../../output/06_deg/cluster/gene_clusters.csv")

gene_lists <- split(gene_clusters$gene_id, gene_clusters$cluster)
```

gene to GOterm to description
```{r}
term_gene <- GO.terms.sig %>% 
  select(GOs, gene_id) %>% 
  rename(GO = GOs)

term_name <- GO.bp.sig %>% 
  select(category, term) %>% 
  rename(GO = category)
```

```{r}
cc_go <- compareCluster(
  geneCluster   = gene_lists,
  fun           = "enricher",
  TERM2GENE     = term_gene,
  TERM2NAME     = term_name,
  pvalueCutoff  = 0.05,
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.2
)
```


For ORA in clusterProfiler, whether GO, KEGG, or custom gene sets, the core inputs are:

1. A gene list (or lists)
  - e.g. genes in pattern 1, pattern 2, pattern 3
  - This is a character vector of gene IDs.
  
For comparing expression patterns 1/2/3 in one go, we can pass a named list of gene vectors into compareCluster(geneClusters = ...)  

```{r}
sig_genes_cluster1 <- sig_summary %>% 
  filter(cluster == 1) %>% 
  pull(unique(gene_id)) # does this collapse any duplicate gene_id?

sig_genes_cluster2 <- sig_summary %>%
  filter(cluster == 2) %>% 
  pull(unique(gene_id))

sig_genes_cluster3 <- sig_summary %>%
  filter(cluster == 3) %>% 
  pull(unique(gene_id))
```
  
  

2. A background / universe (recommended)
  - All genes that could possibly have been called DE in that analysis (e.g. all genes in your filtered DESeq2 object, ~10k).
  - Passed as universe = all_genes in enricher() / enrichGO() etc.
```{r}
universe_genes <- all_genes$gene_id
```

3. Annotation mapping: TERM2GENE (and optionally TERM2NAME)
  - For GO ORA: a 2-column data frame of term = GO ID (e.g. "GO:0006355"), gene = gene_id
  - For KEGG ORA: a 2-column data frame of term = KEGG pathway ID or KO ID (whichever you want to treat as “gene set”), gene = gene_id
  - Built from your annotation table (gff, eggNOG), then restricted to the genes in `universe_genes`

Xu et al. explicitly demonstrate using enricher()/compareCluster() with custom TERM2GENE/TERM2NAME for non-model GO and TF annotations, which is exactly our situation.
```{r}
go2gene <- GO.terms.sig %>% 
  select(GOs, gene_id) %>%
  )
```


# GSEA
```{r}
xx <- compareCluster(
  geneCluster   = gene_lists,
  fun           = "enrichKEGG",        # or enrichKEGG, etc.
  OrgDb         = orgdb_object,
  keyType       = "ENSEMBL",         # or whatever your IDs are
  ont           = "BP",
  pvalueCutoff  = 0.05
)
```


