---
title: "Clustering gene lists"
subtitle: "Grouping genes by metadata and expression patterns"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---
# Background
## Inputs
- `sig_cat`: List of the 44 genes that show significant interactions between leachate and stage 
- `vsd_sig`: The variance stabilized transformed gene count matrix of 44 significant interaction genes
- `LRT_sig_genes`: DESeq results list of LRT significant genes, LFC, & padj values 
- `Wald_sig_genes`: DESeq results list of Wald test results, LFC & padj values for each interaction contrast
- `sig_lvc_cl`: DESeq results list of the 39 genes that are sig diff from control at low cleavage 
- `sig_mvc_cl`: DESeq results list of the 77 genes that are sig diff from control at mid cleavage
- `sig_hvc_cl`: DESeq results list of the 18 genes that are sig diff from control at high cleavage
- `sig_hvc_eg `:DESeq results list of the 13 genes that are sig diff from control at high early gastrula


## Outputs

# Setup

## Install packages
## Load packages
```{r}
library(ComplexUpset)
library(patchwork)
library(tidyverse)
library(DEGreport)
```

## Load inputs

```{r}
# 44 interaction genes
LRT_sig_genes <- read_csv("../../output/06_deg/interactions/LRT_sig_genes.csv")
Wald_sig_genes <- read_csv("../../output/06_deg/interactions/Wald_sig_genes.csv")
vsd_sig <- read_csv("../../output/06_deg/interactions/vsd_sig.csv")

# contrasts
sig_lvc_cl <- read_csv("../../output/06_deg/contrasts/sig_lvc_cl.csv")
sig_mvc_cl <- read_csv("../../output/06_deg/contrasts/sig_mvc_cl.csv")
sig_hvc_cl <- read_csv("../../output/06_deg/contrasts/sig_hvc_cl.csv")
sig_hvc_eg <- read_csv("../../output/06_deg/contrasts/sig_hvc_eg.csv")
```

```{r}
# metadata
metadata_or <- read_csv("../../metadata/metadata_or.csv")

# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")
```

# Gene lists exploration

## Common significant genes across cleavage
```{r}
common_genes <- Reduce(intersect, list(
  sig_lvc_cl$gene_id,
  sig_mvc_cl$gene_id,
  sig_hvc_cl$gene_id
))

length(common_genes)  # Number of genes common to all cleavage, just 1!
print(common_genes) # And it is...
```

::: callout-note
Montipora_capitata_HIv3\_\_\_TS.g50653.t2" is the only gene common across cleavage with differential expression due to low, mid, and high leachate treatments
:::

## Venn Diagram with ComplexUpset

```{r}
genes1 <- sig_lvc_cl$gene_id
genes2 <- sig_mvc_cl$gene_id
genes3 <- sig_hvc_cl$gene_id
genes4 <- sig_hvc_eg$gene_id
genes5 <- LRT_sig_genes$gene_id
```

Upset requires a binary membership table (presence/absence dataframe)

```{r}
all_genes <- unique(c(genes1, genes2, genes3, genes4, genes5))

upset_data <- data.frame(
  gene_id = all_genes,
  LVC_CL  = all_genes %in% genes1,
  MVC_CL  = all_genes %in% genes2,
  HVC_CL  = all_genes %in% genes3,
  HVC_EG  = all_genes %in% genes4,
  LRT     = all_genes %in% genes5
)
```

```{r}
install.packages("ComplexUpset")
```

```{r}
ComplexUpset::upset(
  upset_data,
    intersect = c("LVC_CL", "MVC_CL", "HVC_CL", "HVC_EG", "LRT"),
    base_annotations = list('Intersection size'= 
                              ComplexUpset::intersection_size(counts = TRUE)))

```

## Gene set overlaps with `UpSetR`

```{r}
# install.packages("UpSetR")  # if needed
library(UpSetR)
sets <- c("LVC_CL","MVC_CL","HVC_CL","HVC_EG","LRT")

df_up <- upset_data
df_up[sets] <- lapply(df_up[sets], function(x) as.integer(as.logical(x)))

UpSetR::upset(df_up, sets = sets, keep.order = TRUE)

```

# Perform clustering with `DEGreport`

see Roberts Lab Issue [#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering.

```{r}
clustering_sig_genes <- LRT_sig_genes %>%
  arrange(padj) #genes in rows, stats in columns

matrix <- vsd_sig %>% 
  column_to_rownames(var = "gene_id") #genes in rows, sample ID in columns

meta <- metadata_or %>%
  dplyr::select(sample_id, stage, leachate) %>% #sample id and stage, and leachate columns
  column_to_rownames(var = "sample_id") %>%
  dplyr::select(stage, leachate) #sample ID in rows, one column of lifestage 

# Obtain transformed values for those significant genes
cluster_vst <- matrix[clustering_sig_genes$gene_id, ] #rows are DEGs and columns are sample ID
```

Run with minimum 4 genes per cluster to detect because we have few DEGs.\
- DEGreport vignette, [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)-,Detect%20patterns%20of%20expression,-In%20this%20section)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_clusters <- degPatterns(cluster_vst, metadata = meta, time = "stage", col=NULL, plot=TRUE, minc = 4)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
clusters <- degPatterns(cluster_vst, metadata = meta, time = "leachate", col=NULL, plot=TRUE, minc = 3)
```
```{r}
27+8+4+5
```

```{r}
ggsave(filename="../../output/06_deg/interactions/degPatternscluster_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for leachate (stage grouped all together in one big group, so we can focus in on leachate with the cluster analysis here).

```{r}
# Rename and extract information 
groups <- clusters$df %>%
          rename(gene_id=genes) %>% 
          rename(leachate_cluster=cluster)
```

Save the gene cluster annotation

```{r}
write.csv(groups, "../../output/06_deg/interactions/clusters.csv", row.names = FALSE)
```

Plot is showing Z score of VST normalized gene counts

```{r, warning=FALSE}
cluster_plot<-ggplot(clusters[["normalized"]],
        aes(x=leachate, y=value)) +
  facet_wrap(~ cluster)+
   geom_point(colour="darkgray", size=1) +
  ylab("Z-score expression")+
  xlab("")+
   # change the method to make it smoother
   geom_smooth(aes(group=genes), method = "loess", colour="red", linewidth=0.1, se=FALSE)+
   geom_smooth(aes(group=1), method = "loess", colour="black")+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_vline(xintercept=3, linetype="dotted")+
  geom_vline(xintercept=6, linetype="dotted")+
  theme_classic()+
  theme(
    axis.text.x = element_text(angle=90, hjust=1, vjust=1)
  );cluster_plot

ggsave(cluster_plot, filename="../../output/06_deg/interactions/cluster_plot.png", width=12, height=10)
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!
