---
title: "Clustering gene lists"
subtitle: "Grouping genes by metadata and expression patterns"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background
## Inputs

So we have 11 different 'significant gene' lists. Each list indicates that leachate had a "significant" effect somewhere:

LRT 
- `sig_cat_LRT`, 44 DEGs, this list are genes that had an overall impact by leachate interacting with stage (full vs reduced, testing interaction term as a whole)

Interaction effects
- `sig_prawnchip_high.int`, 1 DEG
- `sig_prawnchip_mid.int`, 5 DEGs
- `sig_prawnchip_low.int`, 12 DEGs
- `sig_earlygastrula_high.int`, 3 DEGs
- `sig_earlygastrula_mid.int`, 11 DEGs
- `sig_earlygastrula_low.int`, 11 DEGs

Stage x leachate effects
- `sig_cleavage_low.vs.control`, 39 DEGs
- `sig_cleavage_mid.vs.control`, 77 DEGs
- `sig_cleavage_high.vs.control`, 18 DEGs
- `sig_earlygastrula_high.vs.control`, 22 DEGs


This totals 151 unique DE genes. 

## Outputs
```{r}
output_path <- "../../output/07_cluster/"
```

# Setup

## Load packages
```{r}
library(ComplexUpset)
library(patchwork)
library(tidyverse)
library(DEGreport)
```

## Load inputs

```{r}
input_path <- "../../output/06_deg/results/"
files <- list.files(input_path, pattern = "\\.csv$", full.names = TRUE)
nm    <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}
```

```{r}
# metadata
metadata_or <- read_csv("../../metadata/metadata_or.csv")

# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")
```
# Perform clustering with `DEGreport`

see Roberts Lab Issue [#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering.
You'll need: 
- a list of your significant genes of interest
- a gene count matrix that has been variance stabilized and filtered to your significant genes of interest 
```{r}
# a list of significant genes
clustering_sig_genes <- sig_all

# significant gene variance stabilized transformed counts matrix
matrix <- vst_mat  #genes in rows, sample ID in columns

meta <- metadata_or %>%
  dplyr::select(sample_id, stage, leachate) %>% #sample id and stage, and leachate columns
  column_to_rownames(var = "sample_id") %>%
  dplyr::select(stage, leachate) #sample ID in rows, one column of lifestage 

# Obtain transformed values for those significant genes
cluster_vst <- matrix[clustering_sig_genes$gene_id, ] #rows are DEGs and columns are sample ID
```

Run with minimum 4 genes per cluster to detect because we have few DEGs.\
- DEGreport vignette, [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)-,Detect%20patterns%20of%20expression,-In%20this%20section)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_clusters <- degPatterns(cluster_vst, metadata = meta, time = "stage", col= "leachate", plot=TRUE, minc = 4)
```
```{r}
ggsave(filename="../../output/07_cluster/degStagePatternscluster_plot.png", width=12, height=10)
```
Broadly most genes displayed an increased transcript abundance as embryos developed. 


```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
leachate_clusters <- degPatterns(cluster_vst, metadata = meta, time = "leachate", col="stage", plot=TRUE, minc = 3)
```

```{r}
ggsave(filename="../../output/07_cluster/degLeachatePatternscluster_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for leachate (stage grouped all together in one big group, so we can focus in on leachate with the cluster analysis here).

```{r}
# Rename and extract information 
gene_clusters <- leachate_clusters$df %>%
          rename(gene_id=genes)
```

Save the gene cluster annotation

```{r}
write_csv(gene_clusters, file.path(output, "gene_clusters.csv"))
```

Plot is showing Z score of VST normalized gene counts

```{r}
leachate_cluster_data<-leachate_clusters[["normalized"]]
leachate_cluster_data<-as.data.frame(leachate_cluster_data)
```

```{r, warning=FALSE}
library(tidyverse)

# label text per facet: "Group: k - genes: N"
facet_labs <- leachate_cluster_data %>%
  distinct(cluster, genes) %>%
  count(cluster, name = "n_genes") %>%
  transmute(cluster = as.character(cluster),
            lab = paste0("Cluster: ", cluster, " - genes: ", n_genes)) %>%
  deframe()

ggplot(leachate_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high"))),
       aes(x = leachate, y = value, color = stage)) +
  geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) +
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = stage), se = FALSE) +
  facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = facet_labs)) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "stage") +
  theme_minimal(base_size = 12) +
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_text(face = "bold"))

```


```{r, warning=FALSE}

# Strip labels like "Group: 1 - genes: 97"
facet_labs <- leachate_cluster_data %>%
  distinct(cluster, genes) %>%
  count(cluster, name = "n_genes") %>%
  transmute(cluster, lab = paste0("Group: ", cluster, " - genes: ", n_genes)) %>%
  deframe()
  
ggplot(leachate_cluster_data, aes(x = leachate, y = value, color = stage)) +
    # per-gene trajectories (thin & transparent)
    geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) +
    # raw points (jittered)
    geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
    geom_boxplot(alpha = 0.1)+
    geom_smooth(aes(group=stage))+
    facet_wrap(~ cluster, nrow = 1, labeller = labeller(group_id = facet_labs)) +
    labs(
      x = NULL,
      y = "Z-score of gene abundance",
      color = "stage"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )


# Example:
# p <- plot_profiles(my_df)
# p

```


```{r, warning=FALSE}
cluster_plot<-ggplot(leachate_clusters[["normalized"]],
        aes(x=leachate, y=value, color=stage)) +
  facet_wrap(~ cluster)+
  geom_point(size=1, alpha = 0.1)+
  geom_boxplot()+
  geom_line(alpha = 0.1)+
  ylab("Z-score expression")+
  xlab("")+
   # change the method to make it smoother
   geom_smooth(aes(group=stage), method = "loess", linewidth=0.1, se=FALSE)+
   #geom_smooth(aes(group=1), method = "loess", colour="black")+
  geom_hline(yintercept=0, linetype="dashed")+
  theme_classic()+
  theme(
    axis.text.x = element_text(angle=45, hjust=1, vjust=1)
  )

cluster_plot

ggsave(cluster_plot, filename="../../output/07_cluster/cluster_plot.png", width=12, height=10)
```



```{r}


leachate_cluster_data<-leachate_cluster_data%>%
  dplyr::select(genes, merge, value, stage, leachate, cluster)

cluster_plot2<-ggplot(leachate_cluster_data,
        aes(x=leachate, y=value, color = stage)) +
  facet_grid(~cluster, scales = "free", space = "free")+
   geom_point(colour="lightgray", size=1) +
  ylab("Z-Score Expression")+
  theme_minimal()

cluster_plot2
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!
