---
title: "Group DEGs by Expression Pattern"
subtitle: "Using DEGreport and WGCNA"
author: "Sarah Tanja"
date: 2026-01-05
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE) # Hide messages
```

# Load Libraries
```{r}
library(tidyverse)
library(DEGreport)
library(WGCNA)
```


# Inputs

```{r}
output_path <- "../../output/06_deg/cluster/"
metadata_path <- "../../metadata/metadata_or.csv"
```

```{r}
# Load metadata (outlier samples removed, 61 samples remain)
metadata_or <- read_csv(metadata_path)

# set factors
metadata_or <- metadata_or %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula")),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high")),
    spawn_night = factor(spawn_night, levels  = c("July_6th", "July_7th", "July_8th"))
    )

# View metadata structure
str(metadata_or)
```

```{r}
# Load VST normalized gene count matrices for leachate and stage DEGs
vst_mat_130_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_130_df.csv")
vst_mat_stage_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_stage_df.csv")
```

# DEGreport Clustering

Detect Patterns of Expression see Roberts Lab Issue
[#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering. You'll need: 

- a metadata matrix with sample_id as rows 
```{r}
meta_mat <- metadata_or %>%
  column_to_rownames(var = "sample_id")

head(meta_mat)
```

- a gene count matrix that has been variance stabilized and
filtered to your significant genes of interest with genes in
rows and sample ID in columns

```{r}
vst_mat_130 <- vst_mat_130_df %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
dim(vst_mat_130)
vst_mat_130[1:6,1:6]
```

```{r}
vst_mat_stage <- vst_mat_stage_df %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
dim(vst_mat_stage)
vst_mat_stage[1:6,1:6]
```

## Leachate DEGs 
Run with minimum 10 genes per cluster to detect because we have few
DEGs.

- DEGreport vignette,
[degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across developmental stage
leachate130_GEPs <- degPatterns(vst_mat_130, metadata = meta_mat, time = "stage", col= "leachate", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bystage_plot.png", width=12, height=10)
```


```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across leachate dose
degPatterns(vst_mat_130, metadata = meta_mat, time = "leachate", col="stage", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bydose_plot.png", width=12, height=10)
```

### Save leachate DEG clusters
```{r}
# Rename and extract information 
leachate130_GEP_df <- leachate130_GEPs$df %>%
          rename(gene_id=genes)
```

```{r, eval=FALSE}
write_csv(leachate130_GEP_df, file.path("../../output/06_deg/cluster/leachate_DEG_GEP.csv"))
```
*GEP: gene expression pattern

### Extract z scores

Dataframe values are Z score of VST normalized gene counts
```{r}
leachate_cluster_z <- leachate130_GEPs[["normalized"]]
leachate_cluster_z <- as.data.frame(leachate_cluster_z) %>% 
  select(1:12) %>% 
  rename(zscore = value)
```

Save dataframe of zscores and cluster for each DEG across each stageXdose
```{r, eval=FALSE}
write_csv(leachate_cluster_z, "../../output/06_deg/cluster/leachate_cluster_zscores.csv")
```


Extract the list of genes in each cluster.

The genes that have high relative transcript abundance in the earlygastrula stage group in group
1, high relative transcript abundance in prawnchip are group 2, and high
relative abundance in cleavage are group 3. There is a general trend of
more transcripts present as development progresses... which makes sense
because the embryo as it grows is doing more. The subtle leachate
effects we do see (non monotonic patterns in cleavage in group 1 and 2,
slight increase under high pvc leachate exposure in early gastrula in
group 3) appear to impact the transcript abundances of the stage with
the lowest abundances of those genes... so the lower the gene abundance
the more effect leachate has on those genes? Is that just an artifact of
increased variability at lower transcript abundance? 

### Plot leachate DEG clusters

```{r}
leachate_cluster_z <- read_csv("../../output/06_deg/cluster/leachate_cluster_zscores.csv")
```

Set color scheme
```{r}
leachate.colors <- c(control = "#AEF1FF", 
                low = "#BBC7FF",
                mid = "#7D8BFF", 
                high = "#592F7D")

stage.colors <- c(cleavage = "#EBA600", 
                  prawnchip = "#D9685B", 
                  earlygastrula = "#A2223C")
```

Wrangle and label 
```{r}
# reverse gene cluster order to plot from cleavage to eg
leachate_cluster_zordered <- leachate_cluster_z %>%
  mutate(cluster = factor(cluster, levels = rev(sort(unique(cluster)))))

facet_labs <- leachate_cluster_zordered %>%
  summarise(n_genes = n_distinct(genes), .by = cluster) %>%        # counts unique genes per cluster
  mutate(lab = sprintf("Group: %s - genes: %d", cluster, n_genes)) %>%
  dplyr::select(cluster, lab) %>%
  deframe()                                                        # named vector: names = cluster, values = lab
```

Plot and patch
```{r, warning=FALSE, fig.height=8, fig.width = 12}

# leachate as x plot
pl <- ggplot(leachate_cluster_zordered %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high"))),
       aes(x = leachate, y = zscore, color = stage)) +
  geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = stage), se = FALSE) + # trend
  facet_wrap(~ cluster, nrow = 1) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "stage") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = stage.colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_blank(),
  strip.background = element_blank())

# stage as x plot
ps <- ggplot(leachate_cluster_zordered %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high")),
                stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula"))),
       aes(x = stage, y = zscore, color = leachate)) +
  geom_line(aes(group = interaction(genes, leachate)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = leachate), se = FALSE) + #trend pattern 
  facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = facet_labs)) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "leachate") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = leachate.colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_text(face = "bold"))


# combine with patchwork
pc <- ps / pl

pc

```

Save plots as pngs
```{r}
ggsave("../../output/06_deg/cluster/leachate_bystage_clusters.png", plot = pl, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/leachate_bydose_clusters.png", plot = ps, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/leachate_combo_clusters.png", plot = pc, width=18, height=12 , dpi=600)
```


## Stage DEGs clustering
```{r}
# Use the `degPatterns` function from the 'DEGreport' package
stage_GEPbystage <- degPatterns(
  vst_mat_stage, 
  metadata = meta_mat, 
  time = "stage", 
  plot = FALSE, 
  minc = 10)
```

::: callout-note
STUCK! 
WGCNA blockwiseModules (network clustering, handles 10k+ well)

If you want modules that behave like coexpression “communities” (often great for GO enrichment), WGCNA’s blockwise workflow is designed to scale and can use multiple threads.
:::

```{r}
ggsave(filename="../../output/06_deg/cluster/DEG_StageClusterPatterns_bystage_plot.png", width=12, height=10)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_GEPbydose <- degPatterns(
  vst_mat_130, 
  metadata = meta_mat, 
  time = "leachate", 
  col= "stage", 
  plot=TRUE, 
  minc = 10)
```

```{r}
ggsave(filename="../../output/06_deg/cluster/DEG_StageClusterPatterns_bydose_plot.png", width=12, height=10)
```

# WGCNA

- [tutorial by Jennifer Chang](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0)

Install WGCNA & package dependencies
```{r}
#install.packages(c("fields", "impute", "dynamicTreeCut", "qvalue", "flashClust", "Hmisc") )
#BiocManager::install("impute", ask = FALSE, update = FALSE)
#install.packages("WGCNA")
```

```{r}
library(WGCNA)
enableWGCNAThreads(nThreads = 8)
```

Transpose VST count matrix such that rows are samples and columns are genes
```{r}
vst_t_stage <- t(vst_mat_stage)  # WGCNA expects samples x genes
```

Pick a threshold value 


Create the network using `blockwiseModules`

```{r}
net <- blockwiseModules(
  vst_t_stage,
  power = 6,                 # choose via pickSoftThreshold
  TOMType = "signed",
  minModuleSize = 30,
  mergeCutHeight = 0.25,
  numericLabels = TRUE,
  saveTOMs = FALSE,
  verbose = 2
)
```

```{r}
stage_deg_modules <- data.frame(
  gene_id = rownames(vst_mat),            # your DEG gene IDs
  module  = labels2colors(net$colors)     # convert numeric → color
) %>%
  mutate(
    module = factor(module)
  )
```


```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  net$dendrograms[[1]],
  mergedColors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

## Relate module (color cluster) assignments to treatment groups 

```{r}
module <- data.frame(
  gene_id = names(net$colors),  # per-gene module labels
  colors = labels2colors(net$colors)
)

module %>%
  dplyr::group_by(colors) %>%
  dplyr::summarise(n_genes = dplyr::n(), .groups = "drop")
```

> WGCNA will calcuate an Eigengene (hypothetical central gene) for each module

```{r}
# Get Module Eigengenes per cluster
eig <- moduleEigengenes(vst_t_stage, mergedColors)$eigengenes
```

```{r}
# Reorder modules so similar modules are next to each other
eig <- orderMEs(eig)
module_order = names(eig) %>% gsub("ME","", .)

# Add treatment annotations
metaeig <- cbind(eig, meta_mat[rownames(eig), , drop = FALSE]) # reorders meta_mat rownames to match eig rownames

# make dataframe
meig_df <- metaeig %>% 
  rownames_to_column(var = "sample_id")
```

```{r}
# tidy & plot data
meig_df_long <-  meig_df %>%
  pivot_longer(-c(sample_id, 11:18)) %>% 
  mutate(
    name = gsub("ME", "", name)
  )

meig_df_long %>% ggplot(., aes(x=sample_id, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```

## Examine Expression Profiles

```{r}

```


# Summary & Next Steps

We can perform functional enrichment on these clusters now!
