---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE) # Hide messages
```

# Load Libraries
```{r}
library(tidyverse)
library(DEGreport)
library(WGCNA)
```


# Inputs

```{r}
output_path <- "../../output/06_deg/cluster/"
metadata_path <- "../../metadata/metadata_or.csv"
```

```{r}
# Load metadata (outlier samples removed, 61 samples remain)
metadata_or <- read_csv(metadata_path)

# set factors
metadata_or <- metadata_or %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula")),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high")),
    spawn_night = factor(spawn_night, levels  = c("July_6th", "July_7th", "July_8th"))
    )

# View metadata structure
str(metadata_or)
```

```{r}
# Load VST normalized gene count matrices for leachate and stage DEGs
vst_mat_130_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_130_df.csv")
vst_mat_stage_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_stage_df.csv")
```

```{r}
vst_mat_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_df.csv")
```

break stage DEGs into three matrices for each stage



# Clustering

Detect Patterns of Expression see Roberts Lab Issue
[#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering. You'll need: 

- a metadata matrix with sample_id as rows 
```{r}
meta_mat <- metadata_or %>%
  column_to_rownames(var = "sample_id")

head(meta_mat)
```

- a gene count matrix that has been variance stabilized and
filtered to your significant genes of interest with genes in
rows and sample ID in columns

```{r}
vst_mat_130 <- vst_mat_130_df %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
dim(vst_mat_130)
vst_mat_130[1:6,1:6]
```

```{r}
vst_mat_stage <- vst_mat_stage_df %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
dim(vst_mat_stage)
vst_mat_stage[1:6,1:6]
```

## Leachate DEGs clustering
Run with minimum 10 genes per cluster to detect because we have few
DEGs.

- DEGreport vignette,
[degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
leachate130_GEPbystage <- degPatterns(vst_mat_130, metadata = meta_mat, time = "stage", col= "leachate", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bystage_plot.png", width=12, height=10)
```

Broadly most genes displayed an increased transcript abundance as
embryos developed.

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
leachate130_GEPbydose <- degPatterns(vst_mat_130, metadata = meta_mat, time = "leachate", col="stage", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bydose_plot.png", width=12, height=10)
```

## Stage DEGs clustering


```{r}
# Use the `degPatterns` function from the 'DEGreport' package
stage_GEPbystage <- degPatterns(
  vst_mat_stage, 
  metadata = meta_mat, 
  time = "stage", 
  plot = FALSE, 
  minc = 10)
```

::: callout-note
STUCK! 
WGCNA blockwiseModules (network clustering, handles 10k+ well)

If you want modules that behave like coexpression “communities” (often great for GO enrichment), WGCNA’s blockwise workflow is designed to scale and can use multiple threads.
:::

# WGCNA

- [tutorial by Jennifer Chang](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0)

Install WGCNA & package dependencies
```{r}
#install.packages(c("fields", "impute", "dynamicTreeCut", "qvalue", "flashClust", "Hmisc") )
#BiocManager::install("impute", ask = FALSE, update = FALSE)
#install.packages("WGCNA")
```

```{r}
library(WGCNA)
enableWGCNAThreads(nThreads = 8)
```

Transpose VST count matrix such that rows are samples and columns are genes
```{r}
vst_t_stage <- t(vst_mat_stage)  # WGCNA expects samples x genes
```

Pick a threshold value 


Create the network using `blockwiseModules`

```{r}
net <- blockwiseModules(
  vst_t_stage,
  power = 6,                 # choose via pickSoftThreshold
  TOMType = "signed",
  minModuleSize = 30,
  mergeCutHeight = 0.25,
  numericLabels = TRUE,
  saveTOMs = FALSE,
  verbose = 2
)
```

```{r}
stage_deg_modules <- data.frame(
  gene_id = rownames(vst_mat),            # your DEG gene IDs
  module  = labels2colors(net$colors)     # convert numeric → color
) %>%
  mutate(
    module = factor(module)
  )
```


```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  net$dendrograms[[1]],
  mergedColors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

## Relate module (color cluster) assignments to treatment groups 

```{r}
module <- data.frame(
  gene_id = names(net$colors),  # per-gene module labels
  colors = labels2colors(net$colors)
)

module %>%
  dplyr::group_by(colors) %>%
  dplyr::summarise(n_genes = dplyr::n(), .groups = "drop")
```

> WGCNA will calcuate an Eigengene (hypothetical central gene) for each module

```{r}
# Get Module Eigengenes per cluster
eig <- moduleEigengenes(vst_t_stage, mergedColors)$eigengenes
```

```{r}
# Reorder modules so similar modules are next to each other
eig <- orderMEs(eig)
module_order = names(eig) %>% gsub("ME","", .)

# Add treatment annotations
metaeig <- cbind(eig, meta_mat[rownames(eig), , drop = FALSE]) # reorders meta_mat rownames to match eig rownames

# make dataframe
meig_df <- metaeig %>% 
  rownames_to_column(var = "sample_id")
```

```{r}
# tidy & plot data
meig_df_long <-  meig_df %>%
  pivot_longer(-c(sample_id, 11:18)) %>% 
  mutate(
    name = gsub("ME", "", name)
  )

meig_df_long %>% ggplot(., aes(x=sample_id, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```

## Examine Expression Profiles


```{r}
ggsave(filename="../../output/06_deg/cluster/DEG_StageClusterPatterns_bystage_plot.png", width=12, height=10)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_GEPbydose <- degPatterns(
  vst_mat_130, 
  metadata = meta_mat, 
  time = "leachate", 
  col= "stage", 
  plot=TRUE, 
  minc = 10)
```

```{r}
ggsave(filename="../../output/06_deg/cluster/DEG_StageClusterPatterns_bydose_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for stage and for leachate.
Notice that the gene counts for the patterns (1, 2, 3) are the same
across stage and leachate... very interesting. The genes that have high
relative transcript abundance in the earlygastrula stage group in group
1, high relative transcript abundance in prawnchip are group 2, and high
relative abundance in cleavage are group 3. There is a general trend of
more transcripts present as development progresses... which makes sense
because the embryo as it grows is doing more. The subtle leachate
effects we do see (non monotonic patterns in cleavage in group 1 and 2,
slight increase under high pvc leachate exposure in early gastrula in
group 3) appear to impact the transcript abundances of the stage with
the lowest abundances of those genes... so the lower the gene abundance
the more effect leachate has on those genes? Is that just an artifact of
increased variability at lower transcript abundance? Are they the same
genes? The following code should return TRUE if that is the case.

```{r}
# Rename and extract information 
leachate130_GEP_df <- leachate130_GEPbystage$df %>%
          rename(gene_id=genes)

stage_GEP_df <- stage_GEPbydose$df %>% 
  rename(gene_id=genes)
```

Save the gene expression patterns

```{r}
write_csv(leachate_GEP_clusters, file.path("../../output/06_deg/cluster/gene_clusters.csv"))

write_csv(stage_GEP_clusters, file.path("../../output/06_deg/cluster/gene_clusters.csv"))
```

Plot is showing Z score of VST normalized gene counts

```{r}
gene_cluster_data <- leachate_clusters[["normalized"]]
gene_cluster_data <- as.data.frame(gene_cluster_data) %>% 
  select(1:12) %>% 
  rename(zscore = value)

write_csv(gene_cluster_data, "../../output/06_deg/cluster/gene_cluster_zscores.csv")
```

```{r}
gene_cluster_data <- read_csv("../../output/06_deg/cluster/gene_cluster_zscores.csv")
```

```{r}
leachate_colors <- c(control = "#BFE8D9",
                     low = "#DFB431",
                     mid = "#8E6B01",
                     high = "#503D02"
                     )

stage_colors <- c(cleavage = "#B7B5AC", 
                  prawnchip = "#6F6E69", 
                  earlygastrula = "#343331"
                  )
```

## Plot gene clusters

```{r}
# reverse gene cluster order to plot from cleavage to eg
gene_cluster_data <- gene_cluster_data %>%
  mutate(cluster = factor(cluster, levels = rev(sort(unique(cluster)))))

facet_labs <- gene_cluster_data %>%
  summarise(n_genes = n_distinct(genes), .by = cluster) %>%        # counts unique genes per cluster
  mutate(lab = sprintf("Group: %s - genes: %d", cluster, n_genes)) %>%
  dplyr::select(cluster, lab) %>%
  deframe()                                                        # named vector: names = cluster, values = lab
```

```{r, warning=FALSE, fig.height=8, fig.width = 12}

# leachate as x plot
pl <- ggplot(gene_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high"))),
       aes(x = leachate, y = zscore, color = stage)) +
  geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = stage), se = FALSE) + # trend
  facet_wrap(~ cluster, nrow = 1) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "stage") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = stage_colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_blank(),
  strip.background = element_blank())

# stage as x plot
ps <- ggplot(gene_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high")),
                stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula"))),
       aes(x = stage, y = zscore, color = leachate)) +
  geom_line(aes(group = interaction(genes, leachate)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = leachate), se = FALSE) + #trend pattern 
  facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = facet_labs)) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "leachate") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = leachate_colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_text(face = "bold"))


# combine with patchwork
pc <- ps / pl

pc

```

```{r}
ggsave("../../output/06_deg/cluster/leachate_cluster.png", plot = pl, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/stage_cluster.png", plot = ps, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/combo_cluster.png", plot = pc, width=18, height=12 , dpi=600)
```

hmmm gotta arrange so its group 3 , 2, 1 ... and more space between
groups so that its obvious to compare the top and bottom graphs. Pick
distinct and pretty leachate and stage colors...

## Add gene cluster annotation to gene list summaries

```{r}
# leachate
sig_summary_leachate <- sig_summary_leachate %>%
  left_join(gene_clusters, by = "gene_id")

write_csv(sig_summary_leachate, file.path(output_path, "sig_summary_leachate.csv"))

# stage 
sig_summary_stage <- sig_summary_stage %>% 
  left_join(gene_clusters, by = "gene_id")

write_csv(sig_summary_stage, file.path(output_path, "sig_summary_stage.csv"))
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!
