---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the differential transcript abundance between embryonic **stage** in *Montipora capitata* using a zero intercept DESeq2 design"
author: "Sarah Tanja"
date: 2025-21-08
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

Here we are looking at a multifactorial (2 factors, multiple levels) differential gene expression analysis for *Montipora capitata* coral embryos at 3 developmental stages (4 hours post fertilization = cleavage , 9 hours post fertilization = prawnchip, and 14 hours post fertilization = early gastrula). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of ubiquitous non-point source PVC leachate pollution on coral reefs.

Testing zero-intercept design as discussed by Dr. Atakan Ekiz [here](https://www.atakanekiz.com/technical/a-guide-to-designs-and-contrasts-in-DESeq2/) 

## Inputs

-   `gcm_filtor.csv` is the filtered gene count matrix for 61 samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers removed)

## Outputs

-   `output/06_deg/interactions/int_wald_heatmap.png`
-   `volcano_facet_interaction.png`

# Setup

## Install packages
```{r, eval=FALSE}         
# CRAN packages
install.packages(c(
  "tidyverse",
  "viridis",
  "plotly",
  "cowplot",
  "pheatmap",
  "vsn",
  "ashr",
  "remotes"  # required for GitHub installs
))

# GitHub package
remotes::install_github("mdscheuerell/flexoki")

# Bioconductor
if (!"BiocManager" %in% rownames(installed.packages())) {
  install.packages("BiocManager")
}
BiocManager::install(c("DESeq2", "DEGreport"))
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(viridis)
library(tidyverse)
library(DEGreport)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
library(vsn)
library(pheatmap)
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r, eval=TRUE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE)
head(gcm_filtor)
```

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filtor` is the gene count matrix that has been filtered down to only contain rows where counts are at least 15 in 75% of the samples and 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA are already removed
:::

## Import metadata

```{r, eval=TRUE}
metadata_or <- read_csv("../../metadata/metadata_or.csv")
str(metadata_or)
```

```{r, eval=TRUE}
# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")

str(metadata_or)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples with explanatory variables: - `stage` is a factor with 3 levels (cleavage, prawnchip, earlygastrula) - `hpf` is a discrete numerical variable representing the hours post fertilization 4, 9, and 14 (corresponding to `stage`)

-   `leachate` is a factor with 4 levels (control, low, mid, high)
-   `leachate_mgL` is a continuous numerical variable representing leachate concentrations in mg/L with values 0 , 0.01 , 0.1 , and 1 (corresponding to `leachate`)
:::

# Make DESeq Data Set (dds)
:::: column-margin
::: callout-note
Continuous vs categorical

Embryonic stage should be treated as a categorical variable because DESeq2 will estimate expression differences between stages, rather than assuming a linear progression from 4 → 14 hpf. It will fit different intercepts for each stage (i.e., different baseline expression levels). If you encode stage as numeric, the model would assume a linear change in expression per hour of development, which is biologically risky unless you expect linearity (note: we do not).
:::
::::

```{r, eval=TRUE}
# design without intercept (no reference)
dds <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~0 + stage))
```

::: column-margin
> "Contrasts from design without intercept:
Sometimes it is easier to work with no-reference designs. You can tell the algorithm not to have an intercept (ie. control group) by using `~ 0 + condition` formula. Conceptually, the zero intercept model in DESeq2 does not compare conditions against each other, instead, each condition’s coefficient represents its own log2-fold expression estimate relative to zero." - *sourced from technical blog post by Dr. Atakin Ekiz*

::: 

## Run DESeq LRT Test

> Does modeling leachate as a factor vs. a continuous dose better explain gene expression changes across stages?

```{r, eval=TRUE}
dds <- DESeq(dds, test = "Wald")
```


## Results

Plot dispersion

```{r}
plotDispEsts(dds)
```


# Make contrasts

```{r}
resultsNames(dds)
```

```{r}
# cleavage to prawnchip
res_1 <- results(dds, contrast = list("stagecleavage", "stageprawnchip")) 
summary(res_1, alpha = 0.05)
```

```{r}
# prawnchip to early gastrula
res_2 <- results(dds, contrast = list("stageprawnchip", "stageearlygastrula"))
summary(res_2, alpha = 0.05)
```

```{r}
# cleavage to early gastrula
res_3 <- results(dds, contrast = list("stagecleavage", "stageearlygastrula"))
summary(res_3, alpha = 0.05)
```
Save significant genes
```{r}
# From res 1, 2 & 3
sig_1 <- res_1 %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") %>% 
  mutate(contrast = 1)

sig_2 <- res_2 %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") %>% 
  mutate(contrast = 2)

sig_3 <- res_3 %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") %>% 
  mutate(contrast = 3)

# merge all together
sig_genes <- merge(sig_1, sig_2, by = "gene_id")
```

# Plot
Variance stabilize
```{r}
vsd <- vst(dds, blind=FALSE)
meanSdPlot(assay(vsd))
```
```{r}
# Extract the transformed count matrix
vsd_mat <- assay(vsd)
# Make a quick heatmap of all genes across samples
pheatmap(vsd_mat, 
         cluster_rows = TRUE, 
         cluster_cols = TRUE, 
         show_rownames = FALSE, 
         annotation_col = as.data.frame(colData(dds)[, "stage", drop=FALSE]))
```

### Extract results

```{r}
# Combine into a summary
sig_genes <- data.frame(
  gene_id = rownames(res_high_pc),
  _log2FC = sig_1$log2FoldChange,
  prawnchip_high_padj = sig_1$padj,
  prawnchip_mid_log2FC = res_mid_pc$log2FoldChange)

```

### Plot heatmap

```{r}
library(pheatmap)

# Step 1: Order and extract log2FC values
log2fc_cols <- c(
  "prawnchip_low_log2FC", "prawnchip_mid_log2FC", "prawnchip_high_log2FC",
  "earlygastrula_low_log2FC", "earlygastrula_mid_log2FC", "earlygastrula_high_log2FC"
)

mat <- as.matrix(int_wald_df_sig[, log2fc_cols])
rownames(mat) <- int_wald_df_sig$gene_id

# Step 2: Rename columns for clarity
colnames(mat) <- c(
  "Prawnchip: Low", "Prawnchip: Mid", "Prawnchip: High",
  "Early Gastrula: Low", "Early Gastrula: Mid", "Early Gastrula: High"
)

# Step 3: Create annotation for stage/leachate
annotation_col <- data.frame(
  Stage = rep(c("Prawnchip", "Early Gastrula"), each = 3),
  Leachate = rep(c("Low", "Mid", "High"), times = 2)
)
rownames(annotation_col) <- colnames(mat)

# Step 4: Prepare significance (padj < 0.05) annotation matrix
padj_cols <- c(
  "prawnchip_low_padj", "prawnchip_mid_padj", "prawnchip_high_padj",
  "earlygastrula_low_padj", "earlygastrula_mid_padj", "earlygastrula_high_padj"
)
padj_mat <- as.matrix(int_wald_df_sig[, padj_cols])
rownames(padj_mat) <- int_wald_df_sig$gene_id
colnames(padj_mat) <- colnames(mat)

# Significance stars with NA handling
sig_labels <- ifelse(is.na(padj_mat), "",
                      ifelse(padj_mat < 0.01, "**",
                        ifelse(padj_mat < 0.05, "*",
                          ifelse(padj_mat < 0.1, ".", ""))))

# Step 5: Save heatmap with significance stars
png("../../output/06_deg/interactions/int_wald_heatmap.png", width = 10, height = 12, units = "in", res = 600)

pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         display_numbers = sig_labels,
         number_color = "black",
         fontsize = 10,
         fontsize_number = 8,
         angle_col = 45,
         border_color = NA)

dev.off()
```

Interpretation notes: \> 🧬 What does the heatmap show? Each cell in the heatmap shows the scaled log2 fold change for a gene in a specific interaction condition. \> \> ✅ Scaling Because you set scale = "row", each gene's values are z-scored across the six interaction conditions \> \> Blue = low relative expression change in that condition \> \> Red = high relative expression change in that condition \> \> The actual color scale is relative within each row, not across the whole matrix. \> \> This helps visualize patterns of expression changes across conditions for each gene, regardless of absolute size. \> \> 📊 What can you learn from this? Are there groups of genes responding similarly? Row clustering may reveal gene modules with similar interaction response profiles. \> \> Which stage + leachate condition drives the interaction? Column patterns (e.g., lots of red in prawnchip_high_log2FC) suggest strong interaction effects in that condition. \> \> Which genes have the strongest differential interactions? Genes with high contrast (strong red/blue) have more dynamic interaction effects. \> \> 🧠 Interpretation Reminder These log2 fold changes are interaction effects --- they reflect how gene expression at a specific `stage` & `leachate` level deviates from the expected additive effect of `stage` and `leachate` alone. \> \> So, a large positive log2FC means: \> \> "At this `stage` and `leachate` level, this gene is more highly expressed than we would expect based on `stage` and `leachate` effects individually." \> \> Likewise, a large negative value = stronger-than-expected suppression.

```{r}
int_wald_df_sig$any_sig <- apply(int_wald_df_sig[, grepl("padj", names(int_wald_df_sig))], 1, function(p) any(p < 0.05, na.rm = TRUE))
```

### PCA of significant genes only

```{r}
# Subset DESeqDataSet to just significant genes
dds_wald_sig <- dds_wald[rownames(dds_wald) %in% sig_cat, ]
vsd_wald_sig <- varianceStabilizingTransformation(dds_wald_sig)

# Get PCA data (only 44 sig DEGs)
pca_sig <- plotPCA(vsd_wald_sig, intgroup = "group", returnData = TRUE)
percentVar_sig <- round(100*attr(pca_sig, "percentVar"))

# Merge with metadata
pca_sig <- merge(pca_sig, metadata_or, by.x = "name", by.y = "sample_id", all.x = TRUE)

# Assign specific colors to each hpf
hpf_colors <- c(
  "cleavage" = flex("cyan", sat=500),
  "prawnchip" = flex("green", sat =500),
  "earlygastrula" = flex("yellow", sat=500)
)

pca_sig$label <- paste0("sample: ", pca_sig$sample_id,
                        "\nStage: ", pca_sig$stage,
                        "\nLeachate: ", pca_sig$leachate)

# Indicate leachate level by saturation (alpha)
leachate_alpha_map <- c("control" = 0.5, "low" = 0.6, "mid" = 0.7, "high" = 0.9)
pca_sig$alpha_val <- leachate_alpha_map[pca_sig$leachate]
```

```{r}
# Original ggplot with a tooltip aesthetic
PCA_sig <- ggplot(pca_sig, aes(PC1, PC2, color = stage, label = name, alpha = alpha_val)) +
  geom_point(size = 1,  stroke = 1) +
  ggtitle("M.cap embryos exposed to pvc leachate, 44 significant genes VST expression") +
  xlab(paste0("PC1: ", percentVar_sig[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_sig[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_sig, tooltip = "label")

```
```{r}
# Original ggplot with a tooltip aesthetic
PCA_sig <- ggplot(pca_sig, aes(PC1, PC2, color = stage, shape = leachate, label = name, alpha = alpha_val)) +
  geom_point(size = 1,  stroke = 1) +
  ggtitle("M.cap embryos exposed to pvc leachate, 44 significant genes VST expression") +
  xlab(paste0("PC1: ", percentVar_sig[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_sig[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_sig, tooltip = "label")
```

# Save gene lists

- `sig_cat` : List of the 44 genes that show significant interactions between leachate and stage (saved above) 

- `LRT_sig_genes` : List of LRT significant genes, LFC, & padj values (saved above) 

- `Wald_sig_genes` : List of Wald test results, LFC & padj values for each interaction contrast 

- `vsd_sig` : Variance stabilized count matrix of genes from significant list 

- `all_genes` : List of all genes in the dataset (the filtered input to the DESeq object) which will serve as a reference for functional enrichment

```{r}
Wald_sig_genes <- int_wald_df_sig

write_csv(Wald_sig_genes, "../../output/06_deg/interactions/Wald_sig_genes.csv")
```

```{r}
vsd_sig <-  vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  filter(gene_id %in% sig_cat)

write_csv(vsd_sig, "../../output/06_deg/interactions/vsd_sig.csv")
```

```{r}
all_genes <- vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  select(gene_id)

write_csv(all_genes, "../../output/06_deg/interactions/all_genes.csv")
```

::: callout-note
12,565 genes listed in the `all_genes.csv`
:::

# Perform clustering
see Roberts Lab Issue [#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering. 

```{r}
clustering_sig_genes <- LRT_sig_genes %>%
  arrange(padj) #genes in rows, stats in columns

matrix <- vsd #genes in rows, sample ID in columns

meta <- metadata_or %>%
  dplyr::select(sample_id, stage, leachate) %>% #sample id and stage, and leachate columns
  column_to_rownames(var = "sample_id")

meta<-meta%>%
  dplyr::select(stage, leachate) #sample ID in rows, one column of lifestage 

# Obtain transformed values for those significant genes
cluster_vst <- matrix[clustering_sig_genes$gene_id, ] #rows are DEGs and columns are sample ID
```

Run with minimum 3 genes per cluster to detect because we have few DEGs.  
- DEGreport vignette, [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)-,Detect%20patterns%20of%20expression,-In%20this%20section)
```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_clusters <- degPatterns(cluster_vst, metadata = meta, time = "stage", col=NULL, plot=TRUE, minc = 3)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
clusters <- degPatterns(cluster_vst, metadata = meta, time = "leachate", col=NULL, plot=TRUE, minc = 3)
```
Extract the list of genes in each cluster for leachate (stage grouped all together in one big group, so we can focus in on leachate with the cluster analysis here). 
```{r}
# Rename and extract information 
groups <- clusters$df %>%
          rename(gene_id=genes) %>% 
          rename(leachate_cluster=cluster)
``` 

Save the gene cluster annotation
```{r}
write.csv(groups, "../../output/06_deg/interactions/clusters.csv")
```

Plot is showing Z score of VST normalized gene counts  
```{r, warning=FALSE}
cluster_plot<-ggplot(clusters[["normalized"]],
        aes(x=leachate, y=value)) +
  facet_wrap(~ cluster)+
   geom_point(colour="darkgray", size=1) +
  ylab("Z-score expression")+
  xlab("")+
   # change the method to make it smoother
   geom_smooth(aes(group=genes), method = "loess", colour="red", linewidth=0.1, se=FALSE)+
   geom_smooth(aes(group=1), method = "loess", colour="black")+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_vline(xintercept=3, linetype="dotted")+
  geom_vline(xintercept=6, linetype="dotted")+
  theme_classic()+
  theme(
    axis.text.x = element_text(angle=90, hjust=1, vjust=1)
  );cluster_plot

ggsave(cluster_plot, filename="../../output/06_deg/interactions/cluster_plot.png", width=12, height=10)
```
# Summary & Next Steps

We can perform functional enrichment on these clusters now!
