---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output)
  dev = "jpeg",
  dpi = 600,
  fig.retina = 2)        
```

# Background

Here we are looking at a multifactorial differential gene expression analysis for *Montipora capitata* coral embryos at 3 developmental stages (4 hours post fertilization = cleavage , 9 hours post fertilization = prawnchip, and 14 hours post fertilization = early gastrula). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of ubiquitous non-point source PVC leachate pollution on coral reefs.

## Inputs

-   `gcm_filtor.csv` is the filtered gene count matrix for 61 samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers removed)

## Outputs

-   `output/06_deg/interactions/int_wald_heatmap.png`
-   `volcano_facet_interaction.png`

# Setup

## Install packages

```{r, eval=FALSE}
# CRAN packages
#install.packages(c(
#  "tidyverse",
#  "viridis",
#  "plotly",
#  "cowplot",
#  "pheatmap",
#  "vsn",
#  "ashr",
#  "remotes"  # required for GitHub installs
#))

#install.packages("ComplexUpset")    # ggplot2-based, prettier
# GitHub package
#remotes::install_github("mdscheuerell/flexoki")

# Bioconductor
#if (!"BiocManager" %in% rownames(installed.packages())) {
#  install.packages("BiocManager")
#}
#BiocManager::install(c("DESeq2", "DEGreport"))
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(viridis)
library(tidyverse)
library(DEGreport)
library(ashr)
library(flexoki)
library(plotly)
library(vsn)
library(pheatmap)
library(ComplexUpset)
library(ggplot2)
library(ggplotify)
library(patchwork)   # for wrap_plots(...)
library(cowplot)     # for plot_grid & get_legend
library(dplyr)
library(grid) 
library(purrr)
library(forcats)
library(ggforce)   # for facet_wrap_paginate
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a sequenced RNA library (i.e. a sample), and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r, eval=TRUE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE) 
head(gcm_filtor)
```

-   The count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filtor` is the gene count matrix that has been filtered down to only contain rows where counts are at least 15 in 75% of the samples and 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA are already removed. This step was completed in `code/05_explore`
:::

## Import metadata

```{r, eval=TRUE}
metadata_or <- read_csv("../../metadata/metadata_or.csv")
str(metadata_or)
```

### Reorder metadata factors

```{r, eval=TRUE}
# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")

str(metadata_or)
```

::: callout-note
`metadata_or` contains metadata for 61 samples (*o*utliers *r*emoved) with explanatory variables: - `stage` is a factor with 3 levels (cleavage, prawnchip, earlygastrula) - `hpf` is a discrete numerical variable representing the hours post fertilization 4, 9, and 14 (corresponding to `stage`) - `leachate` is a factor with 4 levels (control, low, mid, high) - `leachate_mgL` is a continuous numerical variable representing leachate concentrations in mg/L with values 0 , 0.01 , 0.1 , and 1 (corresponding to `leachate`)
:::

# LRT

> [Likelihood ratio test](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#likelihood-ratio-test:~:text=seq%20workflow.-,Likelihood%20ratio%20test,-DESeq2%20offers%20two): DESeq2 offers two kinds of hypothesis tests: the Wald test, where we use the estimated standard error of a log2 fold change to test if it is equal to zero, and the likelihood ratio test (LRT). The LRT examines two models for the counts, a full model with a certain number of terms and a reduced model, in which some of the terms of the full model are removed. The test determines if the increased likelihood of the data using the extra terms in the full model is more than expected if those extra terms are truly zero. The LRT is therefore useful for testing multiple terms at once, for example testing 3 or more levels of a factor at once, or all interactions between two variables. The LRT for count data is conceptually similar to an analysis of variance (ANOVA) calculation in linear regression, except that in the case of the Negative Binomial GLM, we use an analysis of deviance (ANODEV), where the deviance captures the difference in likelihood between a full and a reduced model. The likelihood ratio test can be performed by specifying test="LRT" when using the DESeq function, and providing a reduced design formula, e.g. one in which a number of terms from design(dds) are removed. The degrees of freedom for the test is obtained from the difference between the number of parameters in the two models.

> For a control and treatment time series, one can use a design formula containing the condition factor, the time factor, and the interaction of the two. *In this case, using the likelihood ratio test with a reduced model which does not contain the interaction terms will test whether the condition induces a change in gene expression at any time point after the reference level time point (time 0). An example of the later analysis is provided in our [RNA-seq workflow](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html)*.

This compares:

-   full model: `~ stage + leachate + stage:leachate`

-   reduced model: `~ stage + leachate`

This tells you if there are any interaction effects between stage and leachate, but it won't tell you which interaction is significant, or in which direction --- you need Wald contrasts for that.

::: callout-caution
Order matters! The first term (`stage`) is what the model controls for, and the second term (`leachate`) is the main effect. Meaning the model adjusts for stage first and then assesses the leachate effect.
:::

# Categorical vs. continuous

Stage & Leachate will be treated as categorical variables

When embryonic `stage` is treated as a categorical variable DESeq2 will estimate expression differences between stages, rather than assuming a linear progression from 4 → 14 hpf. DESeq2 will fit different intercepts for each stage (i.e., different baseline expression levels). If you encode stage as a continuous numeric, the model would assume a linear change in expression per hour of development, which is biologically risky unless you expect linearity (note: we do not).

::: callout-note
`stage` will be treated as a categorical variable
:::

Whether `leachate` is treated as a factor or a numeric (continuous) variable radically changes the interpretation of both the main effect and the interaction. If a variable is numeric, DESeq2 treats it as a continuous covariate. This fits a regression slope, estimating the change in gene expression per unit increase in the variable. Coding `leachate` as a categorical factor allows each stage to have its own independent leachate effect. ✅ Use this if you expect nonlinear, stage-specific effects of exposure (note: we do).

```{r, eval=TRUE}

dds_cat <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~0 + stage + leachate + stage:leachate))
```

Coding `leachate_mgL` as a continuous numeric tests for a change in expression per unit increase in leachate, and the interaction term tests whether the slope of the dose-response differs depending on stage. ✅ Use this if you expect a linear or dose-dependent trend, and want to ask if the strength or direction of the trend varies by stage (exploring if this is the case).

```{r, eval=TRUE}
dds_con <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~0 + stage + leachate_mgL + stage:leachate_mgL))
```

## Run DESeq LRT Tests

> Does modeling leachate as a factor vs. a continuous dose better explain gene expression changes across stages?

::: callout-note
\~0 zero-intercept model
:::

```{r, eval=TRUE}
dds_cat_LRT <- DESeq(dds_cat, test = "LRT", reduced = ~0 + stage + leachate)
dds_con_LRT <- DESeq(dds_con, test = "LRT", reduced = ~0 + stage + leachate_mgL)
```

| Model         | Full model                                       | Reduced model               | LRT tests...                                               |
|-----------------|-----------------|-----------------|--------------------|
| `dds_cat_LRT` | `~0 + stage + leachate + stage:leachate`         | `~0 + stage + leachate`     | Are there **stage-specific effects of leachate (factor)**? |
| `dds_con_LRT` | `~0 + stage + leachate_mgL + stage:leachate_mgL` | `~0 + stage + leachate_mgL` | Are there **stage-specific trends** in dose response?      |

Plot dispersion estimates

```{r}
plotDispEsts(dds_cat_LRT)
plotDispEsts(dds_con_LRT)
```

::: callout-note
What does the categorical models dispersion (.01 - 1) vs the continuous model (.005 - 0.5) show?! Pay attention to the scale of the Y axis when comparing them.
:::

## Results

> When one performs a likelihood ratio test, the *p* values and the test statistic (the `stat` column) are values for the test that removes all of the variables which are present in the full design and not in the reduced design. This tests the null hypothesis that all the coefficients from these variables and levels of these factors are equal to zero.
>
> The likelihood ratio test *p* values therefore represent a test of *all the variables and all the levels of factors* which are among these variables. However, the results table only has space for one column of log fold change, so a single variable and a single comparison is shown (among the potentially multiple log fold changes which were tested in the likelihood ratio test). This is indicated at the top of the results table with the text, e.g., log2 fold change (MLE): condition C vs A, followed by, LRT p-value: '\~ batch + condition' vs '\~ batch'. This indicates that the *p* value is for the likelihood ratio test of *all the variables and all the levels*, while the log fold change is a single comparison from among those variables and levels. See the help page for *results* for more details.

```{r}
?results
```

> [Building the results table](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#multiple-testing:~:text=5.2-,Building%20the%20results%20table,-Calling%20results%20without): Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level.
>
> Genes with small *p* values from the LRT test are those which at one or more time points after time 0 showed a strain-specific effect. Note therefore that this will not give small *p* values to genes that moved up or down over time in the same way in both strains.

### categorical

```{r}
resultsNames(dds_cat_LRT)
```

```{r, eval=TRUE}
res_cat_LRT <- results(dds_cat_LRT, alpha = 0.05)
res_cat_LRT
```

```{r}
mcols(res_cat_LRT, use.names = TRUE)
```

```{r}
summary(res_cat_LRT)
```

### continuous

```{r}
resultsNames(dds_con_LRT)
```

```{r, eval=TRUE}
res_con_LRT <- results(dds_con_LRT, alpha = 0.05)
summary(res_con_LRT)
```

### Save significant LRT genes

```{r}
sig_cat_LRT <- res_cat_LRT %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

sig_con_LRT <- res_con_LRT %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

overlap_ids <- intersect(sig_cat_LRT$gene_id, sig_con_LRT$gene_id)
length(overlap_ids)        # how many
overlap_ids                # which ones
```

::: callout-note
The 2 significant genes ("Montipora_capitata_HIv3___RNAseq.g11129.t1" & "Montipora_capitata_HIv3___RNAseq.g21373.t1") found using leachate as a continuous factor are also included in the 44 significant genes found when leachate was coded as a categorical factor.

This means that 2 out of 44 significant genes may have a linear relationship (compared to a non-linear one!) as we look at the interactions between leachate and stage. **All 44 of these genes respond to leachate differently depending on stage. Response to leachate pollution differs by embryonic stage.**

44 genes show significant stage-by-leachate interaction when leachate is treated categorically → meaning the response to leachate varies by stage, but not necessarily in a linear/dose-dependent way.

2 genes show significant stage-specific dose-response effects when leachate is treated as continuous → they show a linear or monotonic trend across leachate concentrations that varies by stage.

The 2 overlapping genes are well-modeled by a linear trend in leachate (i.e., a slope, not just group differences). The 42 other genes show non-linear or non-monotonic behavior --- meaning their response does not follow a consistent increase/decrease with dose. Instead, they respond differently to each specific leachate level depending on the stage (e.g., they might increase at 0.1 mg/L but decrease at 1 mg/L, or vice versa).
:::

Because the 2 genes that are captured in the continuous model are also caputured in the categorical model we will just save the significant categorical list
```{r}
write.csv(sig_cat_LRT, "../../output/06_deg/interactions/sig_cat_LRT.csv")
```

::: callout-note
`sig_cat_LRT` is the DESeq LRT result that shows 44 genes with significant interactions between stage and leachate!
:::

> Of the 44 genes showing significant stage-specific responses to leachate (based on the categorical model), only 2 also show a significant linear dose-response interaction with stage (based on the continuous model). This suggests that while stage modulates leachate response broadly, most genes (42/44) exhibit nonlinear or non-monotonic patterns, rather than a simple linear trend across leachate levels.

## LRT result plots

 Pivot longer for gene count plots
 
 > `counts(dds, normalized = TRUE)` gives size-factor–normalized raw counts (recommended for per-sample plots). vst() / rlog() give transformed values suitable for across-gene visualizations.
 
```{r}
# 1. extract normalized counts (matrix)
norm_mat_LRT <- counts(dds_cat_LRT, normalized = TRUE)
# column ids are sample_id from metadata_or
# rownames are gene_id

# extract significant gene_id character string from DESeq2 result
sig_LRT <- as.character(sig_cat_LRT$gene_id)


# 2. convert -> long tidy dataframe
norm_long_sigLRT <- as.data.frame(norm_mat_LRT) %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "sample_id",
    values_to = "expression"
  ) %>% 
  filter(gene_id %in% sig_LRT)

# 3. join sample metadata
norm_long_sigLRT <- norm_long_sigLRT %>% left_join(metadata_or, by = "sample_id")
```

Check to see that there are 44 unique genes in the `norm_long_sigLRT` dataframe
```{r}
unique(norm_long_sigLRT$gene_id)
```

Add additional columns to the long form dataframe 
```{r}
# add padj
norm_long_sigLRT <- norm_long_sigLRT %>% left_join(sig_cat_LRT, by = "gene_id")

# Assign numbers to genes by min padj
gene_order <- norm_long_sigLRT %>%
  group_by(gene_id) %>%
  summarise(min_padj = min(padj, na.rm = TRUE), .groups = "drop") %>%
  arrange(min_padj) %>%
  mutate(padj_rank = row_number())

# Join back to norm_long_sigLRT
norm_long_sigLRT <- norm_long_sigLRT %>%
  left_join(gene_order %>% dplyr::select(gene_id, padj_rank), by = "gene_id")

# View the structure of the long form dataset 
str(norm_long_sigLRT %>% distinct() %>% arrange(padj))
```

How many samples are in each treatment for each significant gene?

This bit of code lets us see how many samples are contributing to the mean expression of a gene broken down by group (aka condition ... leachate x stage)

```{r}
sample_counts <- norm_long_sigLRT %>% 
  dplyr::count(gene_id, group) %>%   # count rows per gene_id + group
  pivot_wider(names_from = group,              # make each group a column
              values_from = n, 
              values_fill = 0)                 # fill missing group-gene combos with 0

summary(sample_counts)
```

::: callout-note
At least 4 samples represent each group per significant gene. Most groups are represented by 5 samples... some have 4, 6, and 1 group has 7 samples. This is important to check because we want to be able to compare the groups, but if they have vastly different sample sizes that could be problematic. This confirms that the sample sizes are reasonably similar for comparisons.
:::

### Gene plot and print values
```{r}
# Parameters: change if your column names differ
df <- norm_long_sigLRT         # your long dataframe
expr_var <- "expression"       # or "vsd_expression"
stage_col <- "stage"
leachate_col <- "leachate"
sample_col <- "sample_id"

# paginate print values
ncol <- 3
nrow <- 3
per_page <- ncol * nrow
```

### Plot function
```{r, plot-function}


# Plot function: returns a ggplot object
make_gene_plot <- function(gene, 
                           data = df, 
                           expr = expr_var,
                           stage = stage_col, 
                           leach = leachate_col, 
                           sample = sample_col,
                           loess = TRUE,
                           title = TRUE, 
                           show_legend = TRUE) {
  gene_df <- data %>% filter(.data$gene_id == gene)
  if (nrow(gene_df) == 0) stop("No rows for gene: ", gene)
  
  p <- ggplot(gene_df, aes_string(x = leach,
                                  y = expr, 
                                  color = stage, 
                                  label = sample)) +
    geom_point(size = 2, 
               position = position_jitter(width = 0.05), 
               alpha = 0.9) +
    labs(x = NULL, y = "Normalized Expression", color = NULL) +
    scale_x_discrete(expand = expansion(add = c(0.08, 0.08))) +  # shrink horizontal padding (left/right)
    theme_minimal() +
    theme(
      plot.title = element_text(size = 10),
      plot.subtitle = element_text(size = 8),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
      )
  
  if (loess) {
    p <- p + geom_smooth(aes(group = stage, 
                             fill = stage, 
                             color = stage),
                         method = "loess", 
                         se = TRUE, 
                         alpha = 0.35, 
                         linewidth = 0.4, 
                         show.legend = FALSE)
  }
  
  if (title) {
    # include padj if available in df or in a separate table; here we try to grab padj if present
    padj_val <- unique(gene_df$padj)
    subtitle <- if (!all(is.na(padj_val))) paste0("padj = ", signif(min(padj_val, na.rm = TRUE), 3)) else NULL
    p <- p + ggtitle(paste0("Gene: ", gene), subtitle = subtitle)
  }
  
  if (!show_legend) p <- p + theme(legend.position = "none")
  p
}
```

### Plot preview
```{r, preview-plot, fig.width=6}
# --- Examples: preview one plot inline ---

# 1) Preview the first gene in your dataset
genes <- df %>% distinct(gene_id) %>% arrange(gene_id) %>% pull(gene_id)
p1 <- make_gene_plot(genes[1])
print(p1)        # prints inline in R/Quarto/RStudio

# 2) Preview a specific gene by id (replace with a real id)
# p2 <- make_gene_plot("Montipora_capitata_HIv3___RNAseq.g11129.t1")
# print(p2)

# 3) If you want an interactive preview (zoom/hover), use plotly:
# library(plotly)
# ggplotly(p1)

# 4) Save the single preview to file
# ggsave("preview_gene1.png", plot = p1, width = 8, height = 4, dpi = 300)
```
### Paginate facet wrap
```{r, paginate-wrap}
for(page in seq_len(n_pages)) {
  p_page <- p_base +
    ggforce::facet_wrap_paginate(~ gene_id, ncol = ncol, nrow = nrow, page = page, scales = "free_y")
  out_file <- file.path(out_dir, sprintf("all_genes_page_%02d.png", page))
  ggsave(out_file, plot = p_page, width = 16, height = 11 * 1.6, units = "in", dpi = 300)
  message("Saved page ", page, " / ", n_pages, " -> ", out_file)
}
```

```{r}
# Display paginated PNGs in GFM-friendly format (collapsible <details> blocks)
# Run this chunk after the PNGs have been generated by the previous chunk.
files <- list.files("../../output/06_deg/interactions/gene_panels", pattern = "^all_genes_page_\\d+\\.png$", full.names = FALSE)
if (length(files) == 0) {
  cat("No paginated images found in the output folder. Run the generation chunk first.")
} else {
  # print a short intro
  cat("\n**Paginated gene panels** — click a page to expand the image.\n\n")
  # Print each image inside a collapsible details block (GFM-friendly)
  for (f in sort(files)) {
    # path used in Markdown should be the relative path from the rendered .md file location to the image file
    img_path <- file.path("output/06_deg/interactions/gene_panels", f)  # adjust if your site layout differs
    # Emit collapsible block
    cat(sprintf("<details>\n<summary>%s</summary>\n\n![](%s)\n\n</details>\n\n", f, img_path))
  }
}
```


::: callout-note
We have 44 genes with significant interaction effects overall.

Next step: 👉 Use Wald tests to dig into which interactions (e.g., stage × leachate level) are responsible --- gene by gene.
:::

# Wald

> Wald tests on interaction terms assess whether the interaction coefficient **differs significantly from 0** --- i.e., whether the combined effect of that `hpf` and `leachate` level deviates from what would be expected by the additive effects alone.

| What You Want to Know                                               | Use            | Explanation                                                                   |
|-------------------------|------------------|-----------------------------|
| Specific interaction term significant?                              | Wald           | `results(dds, name = "stageX.leachateY")`                                     |
| Total effect of a leachate treatment at a specific stage vs control | Wald contrasts | `results(dds, contrast = list(c("leachateY_vs_control", "stageX.leachateY"))` |

Run the Wald test on the categorical dds object.

```{r}
dds_wald <- DESeq(dds_cat, test = "Wald")
```


```{r}
resultsNames(dds_wald)
```

## Interaction terms
-   "stageprawnchip.leachatelow"

-   "stageearlygastrula.leachatelow"

-   "stageprawnchip.leachatemid"

-   "stageearlygastrula.leachatemid"

-   "stageprawnchip.leachatehigh"

-   "stageearlygastrula.leachatehigh"

Each of the above interaction terms tells us the extra effect of the leachate level at a stage (vs the additive sum of the stage and leachate effects) in comparison to a control baseline count. For the interaction terms, we are trying to get the "difference of the differences". For this analysis, we are interested in genes that are differentially expressed in treatment conditions, but not in the control condition (ie. genes uniquely changing in a stage due to treatment).

### Prawnchip

```{r}
pch <- results(dds_wald, name = "stageprawnchip.leachatehigh", alpha = 0.05)
summary(pch)
```

```{r}
sig_pch <- pch %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r}
# make long expression dataframe of significant interaction genes
sig_long <- vsd[sig_cat, ] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "sample_id", values_to = "vsd_expression")

# add metadata
sig_long <- sig_long %>% left_join(metadata_or, by = "sample_id")
# add padj
sig_long <- sig_long %>% left_join(LRT_sig_genes, by = "gene_id")

# Assign numbers to genes by min padj
gene_order <- sig_long %>%
  group_by(gene_id) %>%
  summarise(min_padj = min(padj, na.rm = TRUE), .groups = "drop") %>%
  arrange(min_padj) %>%
  mutate(padj_rank = row_number())

# Join back to sig_long
sig_long <- sig_long %>%
  left_join(gene_order %>% dplyr::select(gene_id, padj_rank), by = "gene_id")
```

```{r}
res_m_int_pc  <- results(dds_wald, name = "stageprawnchip.leachatemid", alpha = 0.05)
summary(res_m_int_pc)
```

```{r}
res_l_int_pc  <- results(dds_wald, name = "stageprawnchip.leachatelow", alpha = 0.05)
summary(res_l_int_pc)
```

### Early gastrula

```{r}
res_h_int_eg <- results(dds_wald, name = "stageearlygastrula.leachatehigh", alpha = 0.05)
summary(res_h_int_eg)
```

```{r}
res_m_int_eg  <- results(dds_wald, name = "stageearlygastrula.leachatemid", alpha = 0.05)
summary(res_m_int_eg)
```

```{r}
res_l_int_eg  <- results(dds_wald, name = "stageearlygastrula.leachatelow", alpha = 0.05)
summary(res_l_int_eg)
```

## vc. Control contrasts

See [A guide to designs and contrasts in DESeq2](https://www.atakanekiz.com/technical/a-guide-to-designs-and-contrasts-in-DESeq2/) by Dr. Atakan Ekiz

```{r}
resultsNames(dds_wald)
```

## low leachate vs control (lvc)

for cleavage (baseline stage)

```{r}
res_lvc_cl <- results(dds_wald, name = "leachate_low_vs_control", alpha = 0.05)
summary(res_lvc_cl)
```

::: callout-note
39 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the low leachate treatment compared to the control. 14 LFC \> 0 & 25 LFC \< 0.
:::

```{r}
sig_lvc_cl <- as.data.frame(subset(res_lvc_cl, padj<0.05))
sig_lvc_cl <- sig_lvc_cl %>% rownames_to_column(., var = "gene_id") 
write_csv(sig_lvc_cl, "../../output/06_deg/contrasts/sig_lvc_cl.csv")
```

for prawnchip

```{r}
res_lvc_pc <- results(dds_wald, contrast = list(c("leachate_low_vs_control",
                                                  "stageprawnchip.leachatelow")), 
                      alpha = 0.05)
summary(res_lvc_pc)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the low leachate treatment when compared to the control.
:::

for early gastrula

```{r}
res_lvc_eg <- results(dds_wald, contrast = list(c("leachate_low_vs_control",
                                                 "stageearlygastrula.leachatelow")),
                      alpha = 0.05)
summary(res_lvc_eg)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early gatrula stage in the low leachate treatment when compared to the control.
:::

## mid leachate vs. control (mvc)

for cleavage

```{r}
res_mvc_cl <- results(dds_wald, name = "leachate_mid_vs_control", alpha = 0.05)
summary(res_mvc_cl)
```

::: callout-note
77 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the mid leachate treatment compared to the control. 77 LFC \> 0. These 77 genes were all found in more abundance compared to the controls.
:::

```{r}
sig_mvc_cl <- as.data.frame(subset(res_mvc_cl, padj<0.05))
sig_mvc_cl <- sig_mvc_cl %>% rownames_to_column(., var = "gene_id")
write_csv(sig_mvc_cl, "../../output/06_deg/contrasts/sig_mvc_cl.csv")
```

for prawnchip

```{r}
res_mvc_pc <- results(dds_wald, contrast = list(c("leachate_mid_vs_control",
                                                  "stageprawnchip.leachatemid")))
summary(res_mvc_pc)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the mid leachate treatment when compared to the control.
:::

for early gastrula

```{r}
res_mvc_eg <- results(dds_wald, contrast = list(c("leachate_mid_vs_control",
                                                  "stageearlygastrula.leachatemid")))
summary(res_mvc_eg)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the mid leachate treatment when compared to the control.
:::

## high leachate vs. control (hvc)

for cleavage

```{r}
res_hvc_cl <- results(dds_wald, name = "leachate_high_vs_control")
summary(res_hvc_cl)
```

::: callout-note
24 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the high leachate treatment compared to the control. 8 LFC \> 0 , 16 LFC \<0.
:::

```{r}
sig_hvc_cl <- as.data.frame(subset(res_hvc_cl, padj<0.05))
sig_hvc_cl <- sig_hvc_cl %>% rownames_to_column(., var = "gene_id")
write_csv(sig_hvc_cl, "../../output/06_deg/contrasts/sig_hvc_cl.csv")
```

for prawnchip

```{r}
res_hvc_pc <- results(dds_wald, contrast = list(c("leachate_high_vs_control",
                                                  "stageprawnchip.leachatehigh")))
summary(res_hvc_pc)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the high leachate treatment when compared to the control.
:::

for early gastrula

```{r}
res_hvc_eg <- results(dds_wald, contrast = list(c("leachate_high_vs_control",
                                                  "stageearlygastrula.leachatehigh")))
summary(res_hvc_eg)
```

::: callout-note
32 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the high leachate treatment compared to the control. 32 LFC \> 0 .
:::

```{r}
sig_hvc_eg <- as.data.frame(subset(res_hvc_eg, padj<0.05))
sig_hvc_eg <- sig_hvc_eg %>% rownames_to_column(., var = "gene_id")
write_csv(sig_hvc_eg, "../../output/06_deg/contrasts/sig_hvc_eg.csv")
```

# Combine results

```{r}
# Combine into a summary
int_wald_df <- data.frame(
  gene_id = rownames(res_high_pc),
  pc_h_log2FC = res_high_pc$log2FoldChange,
  pc_high_padj = res_high_pc$padj,
  prawnchip_mid_log2FC = res_mid_pc$log2FoldChange,
  prawnchip_mid_padj = res_mid_pc$padj,
  prawnchip_low_log2FC = res_low_pc$log2FoldChange,
  prawnchip_low_padj = res_low_pc$padj,
  earlygastrula_high_log2FC = res_high_eg$log2FoldChange,
  earlygastrula_high_padj = res_high_eg$padj,
  earlygastrula_mid_log2FC = res_mid_eg$log2FoldChange,
  earlygastrula_mid_padj = res_mid_eg$padj,
  earlygastrula_low_log2FC = res_low_eg$log2FoldChange,
  earlygastrula_low_padj = res_low_eg$padj
)

# Focus on the 44 genes
int_wald_df_sig <- int_wald_df[int_wald_df$gene_id %in% LRT_sig_genes$gene_id, ]
```

### Plot heatmap

```{r}
library(pheatmap)

# Step 1: Order and extract log2FC values
log2fc_cols <- c(
  "prawnchip_low_log2FC", "prawnchip_mid_log2FC", "prawnchip_high_log2FC",
  "earlygastrula_low_log2FC", "earlygastrula_mid_log2FC", "earlygastrula_high_log2FC"
)

mat <- as.matrix(int_wald_df_sig[, log2fc_cols])
rownames(mat) <- int_wald_df_sig$gene_id

# Step 2: Rename columns for clarity
colnames(mat) <- c(
  "Prawnchip: Low", "Prawnchip: Mid", "Prawnchip: High",
  "Early Gastrula: Low", "Early Gastrula: Mid", "Early Gastrula: High"
)

# Step 3: Create annotation for stage/leachate
annotation_col <- data.frame(
  Stage = rep(c("Prawnchip", "Early Gastrula"), each = 3),
  Leachate = rep(c("Low", "Mid", "High"), times = 2)
)
rownames(annotation_col) <- colnames(mat)

# Step 4: Prepare significance (padj < 0.05) annotation matrix
padj_cols <- c(
  "prawnchip_low_padj", "prawnchip_mid_padj", "prawnchip_high_padj",
  "earlygastrula_low_padj", "earlygastrula_mid_padj", "earlygastrula_high_padj"
)
padj_mat <- as.matrix(int_wald_df_sig[, padj_cols])
rownames(padj_mat) <- int_wald_df_sig$gene_id
colnames(padj_mat) <- colnames(mat)

# Significance stars with NA handling
sig_labels <- ifelse(is.na(padj_mat), "",
                      ifelse(padj_mat < 0.01, "**",
                        ifelse(padj_mat < 0.05, "*",
                          ifelse(padj_mat < 0.1, ".", ""))))

# Step 5: Save heatmap with significance stars
png("../../output/06_deg/interactions/int_wald_heatmap.png", width = 10, height = 12, units = "in", res = 600)

pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         display_numbers = sig_labels,
         number_color = "black",
         fontsize = 10,
         fontsize_number = 8,
         angle_col = 45,
         border_color = NA)
```

Interpretation notes: \> 🧬 What does the heatmap show? Each cell in the heatmap shows the scaled log2 fold change for a gene in a specific interaction condition. \> \> ✅ Scaling Because you set scale = "row", each gene's values are z-scored across the six interaction conditions \> \> Blue = low relative expression change in that condition \> \> Red = high relative expression change in that condition \> \> The actual color scale is relative within each row, not across the whole matrix. \> \> This helps visualize patterns of expression changes across conditions for each gene, regardless of absolute size. \> \> 📊 What can you learn from this? Are there groups of genes responding similarly? Row clustering may reveal gene modules with similar interaction response profiles. \> \> Which stage + leachate condition drives the interaction? Column patterns (e.g., lots of red in prawnchip_high_log2FC) suggest strong interaction effects in that condition. \> \> Which genes have the strongest differential interactions? Genes with high contrast (strong red/blue) have more dynamic interaction effects. \> \> 🧠 Interpretation Reminder These log2 fold changes are interaction effects --- they reflect how gene expression at a specific `stage` & `leachate` level deviates from the expected additive effect of `stage` and `leachate` alone. \> \> So, a large positive log2FC means: \> \> "At this `stage` and `leachate` level, this gene is more highly expressed than we would expect based on `stage` and `leachate` effects individually." \> \> Likewise, a large negative value = stronger-than-expected suppression.

```{r}
int_wald_df_sig$any_sig <- apply(int_wald_df_sig[, grepl("padj", names(int_wald_df_sig))], 1, function(p) any(p < 0.05, na.rm = TRUE))
```

### PCA of significant genes only

```{r}
# Subset DESeqDataSet to just significant genes
dds_wald_sig <- dds_wald[rownames(dds_wald) %in% sig_cat, ]
vsd_wald_sig <- varianceStabilizingTransformation(dds_wald_sig)

# Get PCA data (only 44 sig DEGs)
pca_sig <- plotPCA(vsd_wald_sig, intgroup = "group", returnData = TRUE)
percentVar_sig <- round(100*attr(pca_sig, "percentVar"))

# Merge with metadata
pca_sig <- merge(pca_sig, metadata_or, by.x = "name", by.y = "sample_id", all.x = TRUE)

# Assign specific colors to each hpf
hpf_colors <- c(
  "cleavage" = flex("cyan", sat=500),
  "prawnchip" = flex("green", sat =500),
  "earlygastrula" = flex("yellow", sat=500)
)

pca_sig$label <- paste0("sample: ", pca_sig$sample_id,
                        "\nStage: ", pca_sig$stage,
                        "\nLeachate: ", pca_sig$leachate)

# Indicate leachate level by saturation (alpha)
leachate_alpha_map <- c("control" = 0.5, "low" = 0.6, "mid" = 0.7, "high" = 0.9)
pca_sig$alpha_val <- leachate_alpha_map[pca_sig$leachate]
```

```{r}
# Original ggplot with a tooltip aesthetic
PCA_sig <- ggplot(pca_sig, aes(PC1, PC2, color = stage, label = name, alpha = alpha_val)) +
  geom_point(size = 1,  stroke = 1) +
  ggtitle("M.cap embryos exposed to pvc leachate, 44 significant genes VST expression") +
  xlab(paste0("PC1: ", percentVar_sig[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_sig[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_sig, tooltip = "label")

```

```{r}
# Original ggplot with a tooltip aesthetic
PCA_sig <- ggplot(pca_sig, aes(PC1, PC2, color = stage, shape = leachate, label = name, alpha = alpha_val)) +
  geom_point(size = 1,  stroke = 1) +
  ggtitle("M.cap embryos exposed to pvc leachate, 44 significant genes VST expression") +
  xlab(paste0("PC1: ", percentVar_sig[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_sig[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_sig, tooltip = "label")
```

# Save gene lists

-   `sig_cat` : List of the 44 genes that show significant interactions between leachate and stage (saved above)

-   `LRT_sig_genes` : List of LRT significant genes, LFC, & padj values (saved above)

-   `Wald_sig_genes` : List of Wald test results, LFC & padj values for each interaction contrast

-   `vsd_sig` : Variance stabilized count matrix of genes from significant list

-   `all_genes` : List of all genes in the dataset (the filtered input to the DESeq object) which will serve as a reference for functional enrichment

```{r}
# The 39 genes that are sig diff from control at low cleavage 
sig_lvc_cl <- sig_lvc_cl %>% 
  mutate(low_cleavage = 1)

# The 77 genes that are sig diff from control at mid cleavage
sig_mvc_cl <- sig_mvc_cl %>% 
  mutate(mid_cleavage = 1)

# The 18 genes that are sig diff from control at high cleavage
sig_hvc_cl <- sig_hvc_cl %>% 
  mutate(high_cleavage = 1)

# The 13 genes that are sig diff from control at high early gastrula
sig_hvc_eg <- sig_hvc_eg %>% 
  mutate(high_earlygastrula = 1)
```

```{r}
common_genes <- Reduce(intersect, list(
  sig_lvc_cl$gene_id,
  sig_mvc_cl$gene_id,
  sig_hvc_cl$gene_id
))

length(common_genes)  # Number of genes common to all cleavage, just 1!
print(common_genes) # And it is...
```

::: callout-note
Montipora_capitata_HIv3\_\_\_TS.g50653.t2" is the only gene common across cleavage with differential expression due to low, mid, and high leachate treatments
:::

# Gene set overlap: Venn Diagram

```{r}
genes1 <- sig_lvc_cl$gene_id
genes2 <- sig_mvc_cl$gene_id
genes3 <- sig_hvc_cl$gene_id
genes4 <- sig_hvc_eg$gene_id
genes5 <- LRT_sig_genes$gene_id
```

Upset requires a binary membership table (presence/absence dataframe)

```{r}
all_genes <- unique(c(genes1, genes2, genes3, genes4, genes5))

upset_data <- data.frame(
  gene_id = all_genes,
  LVC_CL  = all_genes %in% genes1,
  MVC_CL  = all_genes %in% genes2,
  HVC_CL  = all_genes %in% genes3,
  HVC_EG  = all_genes %in% genes4,
  LRT     = all_genes %in% genes5
)
```

```{r}
upset_plot <- upset_data %>% 
  upset(
    intersect = c("LVC_CL", "MVC_CL", "HVC_CL", "HVC_EG", "LRT"),
    base_annotations = list(
      'Intersection size' = intersection_size(
        counts = TRUE
      )
    )
  ) +
  labs(title = "UpSet Plot of Significant Genes (P < 0.05)")

upset_plot
```

```{r}
Wald_sig_genes <- int_wald_df_sig

write_csv(Wald_sig_genes, "../../output/06_deg/interactions/Wald_sig_genes.csv")
```

```{r}
vsd_sig <-  vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  filter(gene_id %in% sig_cat)

write_csv(vsd_sig, "../../output/06_deg/interactions/vsd_sig.csv")
```

```{r}
all_genes <- vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  select(gene_id)

write_csv(all_genes, "../../output/06_deg/interactions/all_genes.csv")
```

::: callout-note
12,565 genes listed in the `all_genes.csv`
:::

# Perform clustering

see Roberts Lab Issue [#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering.

```{r}
clustering_sig_genes <- LRT_sig_genes %>%
  arrange(padj) #genes in rows, stats in columns

matrix <- vsd #genes in rows, sample ID in columns

meta <- metadata_or %>%
  dplyr::select(sample_id, stage, leachate) %>% #sample id and stage, and leachate columns
  column_to_rownames(var = "sample_id")

meta<-meta%>%
  dplyr::select(stage, leachate) #sample ID in rows, one column of lifestage 

# Obtain transformed values for those significant genes
cluster_vst <- matrix[clustering_sig_genes$gene_id, ] #rows are DEGs and columns are sample ID
```

Run with minimum 3 genes per cluster to detect because we have few DEGs.\
- DEGreport vignette, [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)-,Detect%20patterns%20of%20expression,-In%20this%20section)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_clusters <- degPatterns(cluster_vst, metadata = meta, time = "stage", col=NULL, plot=TRUE, minc = 4)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
clusters <- degPatterns(cluster_vst, metadata = meta, time = "leachate", col=NULL, plot=TRUE, minc = 3)
```

```{r}
ggsave(filename="../../output/06_deg/interactions/degPatternscluster_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for leachate (stage grouped all together in one big group, so we can focus in on leachate with the cluster analysis here).

```{r}
# Rename and extract information 
groups <- clusters$df %>%
          rename(gene_id=genes) %>% 
          rename(leachate_cluster=cluster)
```

Save the gene cluster annotation

```{r}
write.csv(groups, "../../output/06_deg/interactions/clusters.csv", row.names = FALSE)
```

Plot is showing Z score of VST normalized gene counts

```{r, warning=FALSE}
cluster_plot<-ggplot(clusters[["normalized"]],
        aes(x=leachate, y=value)) +
  facet_wrap(~ cluster)+
   geom_point(colour="darkgray", size=1) +
  ylab("Z-score expression")+
  xlab("")+
   # change the method to make it smoother
   geom_smooth(aes(group=genes), method = "loess", colour="red", linewidth=0.1, se=FALSE)+
   geom_smooth(aes(group=1), method = "loess", colour="black")+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_vline(xintercept=3, linetype="dotted")+
  geom_vline(xintercept=6, linetype="dotted")+
  theme_classic()+
  theme(
    axis.text.x = element_text(angle=90, hjust=1, vjust=1)
  );cluster_plot

ggsave(cluster_plot, filename="../../output/06_deg/interactions/cluster_plot.png", width=12, height=10)
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!
