---
title: "Group DEGs by Expression Pattern"
subtitle: "Using DEGreport and WGCNA"
author: "Sarah Tanja"
date: 2026-01-05
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE) # Hide messages
```

# Load Libraries

```{r}
library(tidyverse)
library(DEGreport)
library(patchwork)
```

Install WGCNA & package dependencies

::: callout-note
If you have not installed WGCNA before, you may need to install some
dependencies first. Uncomment and run the code chunk below to install
them, and then install WGCNA.
:::

```{r}
#install.packages(c("fields", "dynamicTreeCut", "qvalue", "flashClust", "Hmisc") )
#BiocManager::install("impute", ask = FALSE, update = FALSE)
#install.packages("WGCNA")
```

```{r}
library(WGCNA)
enableWGCNAThreads(nThreads = 8)
```

# Inputs

```{r}
output_path <- "../../output/06_deg/cluster/"
metadata_path <- "../../metadata/metadata_or.csv"
```

```{r}
# Load metadata (outlier samples removed, 61 samples remain)
metadata_or <- read_csv(metadata_path)

# set factors
metadata_or <- metadata_or %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula")),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high")),
    spawn_night = factor(spawn_night, levels  = c("July_6th", "July_7th", "July_8th"))
    )

# View metadata structure
str(metadata_or)
```

```{r}
# Load VST normalized gene count matrices for leachate and stage DEGs
vst_mat_130_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_130_df.csv")
vst_mat_stage_df <- read_csv("../../output/06_deg/spawnnight/vst_mat_stage_df.csv")
```

# DEGreport Clustering

Detect Patterns of Expression see Roberts Lab Issue
[#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering. You'll need:

-   a metadata matrix with sample_id as rows

```{r}
meta_mat <- metadata_or %>%
  column_to_rownames(var = "sample_id")

head(meta_mat)
```

-   a gene count matrix that has been variance stabilized and filtered
    to your significant genes of interest with genes in rows and sample
    ID in columns

```{r}
vst_mat_130 <- vst_mat_130_df %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
dim(vst_mat_130)
vst_mat_130[1:6,1:6]
```

```{r}
vst_mat_stage <- vst_mat_stage_df %>%
  column_to_rownames(var = "gene_id") %>%
  as.matrix()
dim(vst_mat_stage)
vst_mat_stage[1:6,1:6]
```

## Leachate DEGs

Run with minimum 10 genes per cluster to detect because we have few
DEGs.

-   DEGreport vignette,
    [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across developmental stage
leachate130_GEPs <- degPatterns(vst_mat_130, metadata = meta_mat, time = "stage", col= "leachate", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bystage_plot.png", width=12, height=10)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across leachate dose
degPatterns(vst_mat_130, metadata = meta_mat, time = "leachate", col="stage", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bydose_plot.png", width=12, height=10)
```

### Save leachate DEG clusters

```{r}
# Rename and extract information 
leachate130_GEP_df <- leachate130_GEPs$df %>%
          rename(gene_id=genes)
```

```{r, eval=FALSE}
write_csv(leachate130_GEP_df, file.path("../../output/06_deg/cluster/leachate_DEG_GEP.csv"))
```

\*GEP: gene expression pattern

### Extract z scores

Dataframe values are Z score of VST normalized gene counts

```{r}
leachate_cluster_z <- leachate130_GEPs[["normalized"]]
leachate_cluster_z <- as.data.frame(leachate_cluster_z) %>% 
  select(1:12) %>% 
  rename(zscore = value)
```

Save dataframe of zscores and cluster for each DEG across each
stageXdose

```{r, eval=FALSE}
write_csv(leachate_cluster_z, "../../output/06_deg/cluster/leachate_cluster_zscores.csv")
```

Extract the list of genes in each cluster.

The genes that have high relative transcript abundance in the
earlygastrula stage group in group 1, high relative transcript abundance
in prawnchip are group 2, and high relative abundance in cleavage are
group 3. There is a general trend of more transcripts present as
development progresses... which makes sense because the embryo as it
grows is doing more. The subtle leachate effects we do see (non
monotonic patterns in cleavage in group 1 and 2, slight increase under
high pvc leachate exposure in early gastrula in group 3) appear to
impact the transcript abundances of the stage with the lowest abundances
of those genes... so the lower the gene abundance the more effect
leachate has on those genes? Is that just an artifact of increased
variability at lower transcript abundance?

### Plot leachate DEG clusters

Set color scheme

```{r}
leachate.colors <- c(control = "#AEF1FF", 
                low = "#BBC7FF",
                mid = "#7D8BFF", 
                high = "#592F7D")

stage.colors <- c(cleavage = "#EBA600", 
                  prawnchip = "#D9685B", 
                  earlygastrula = "#A2223C")
```

Wrangle and label

```{r}
# reverse gene cluster order to plot from cleavage to eg
leachate_cluster_zordered <- leachate_cluster_z %>%
  mutate(cluster = factor(cluster, levels = rev(sort(unique(cluster)))))

facet_labs <- leachate_cluster_zordered %>%
  summarise(n_genes = n_distinct(genes), .by = cluster) %>%        # counts unique genes per cluster
  mutate(lab = sprintf("Group: %s - genes: %d", cluster, n_genes)) %>%
  dplyr::select(cluster, lab) %>%
  deframe()                                                        # named vector: names = cluster, values = lab
```

Plot and patch

```{r, warning=FALSE, fig.height=8, fig.width = 12}

# leachate as x plot
pl <- ggplot(leachate_cluster_zordered %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high"))),
       aes(x = leachate, y = zscore, color = stage)) +
  geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = stage), se = FALSE) + # trend
  facet_wrap(~ cluster, nrow = 1) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "stage") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = stage.colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_blank(),
  strip.background = element_blank())

# stage as x plot
ps <- ggplot(leachate_cluster_zordered %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high")),
                stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula"))),
       aes(x = stage, y = zscore, color = leachate)) +
  geom_line(aes(group = interaction(genes, leachate)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = leachate), se = FALSE) + #trend pattern 
  facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = facet_labs)) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "leachate") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = leachate.colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_text(face = "bold"))


# combine with patchwork
pc <- ps / pl

pc

```

Save plots as pngs

```{r, eval=FALSE}
ggsave("../../output/06_deg/cluster/leachate_bystage_clusters.png", plot = pl, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/leachate_bydose_clusters.png", plot = ps, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/leachate_combo_clusters.png", plot = pc, width=18, height=12 , dpi=600)
```

## Stage DEGs clustering

```{r, eval=FALSE}
# Use the `degPatterns` function from the 'DEGreport' package
#stage_GEPbystage <- degPatterns(
#  vst_mat_stage, 
#  metadata = meta_mat, 
#  time = "stage", 
#  plot = FALSE, 
#  minc = 10)
```

::: callout-note
STUCK! WGCNA blockwiseModules (network clustering, handles 10k+ well)

If you want modules that behave like coexpression “communities” (often
great for GO enrichment), WGCNA’s blockwise workflow is designed to
scale and can use multiple threads.
:::

# WGCNA

-   [tutorial by Jennifer
    Chang](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0)

::: callout-note
Note that here we are starting with a subset of genes that have already
been normalized using a variance stabilizing transformation and are
differentially expressed.
:::

Transpose VST count matrix such that rows are samples and columns are
genes

```{r}
vst_t_stage <- t(vst_mat_stage)  # WGCNA expects samples x genes

vst_t_leachate <- t(vst_mat_130)
```

Examine the dataframe structure to confirm that rows are samples and
columns are genes and values are VST normalized counts

```{r}
vst_t_stage %>%
  as.data.frame() %>%
  dplyr::select(1:3) %>%
  head() %>%
  knitr::kable()
```

```{r}
vst_t_leachate %>% 
  as.data.frame() %>% 
  dplyr::select(1:3) %>% 
  head() %>% 
  knitr::kable()
```

## Pick soft threshhold 

> We can see now that the rows = treatments and columns = gene probes.
> We’re ready to start WGCNA. A correlation network will be a complete
> network (all genes are connected to all other genes). Ergo we will
> need to pick a threshhold value (if correlation is below threshold,
> remove the edge). We assume the true biological network follows a
> scale-free structure (see papers by Albert Barabasi). To do that,
> WGCNA will try a range of soft thresholds and create a diagnostic
> plot. - [Jennifer
> Chang](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0:~:text=We%20can%20see,a%20diagnostic%20plot.)

::: callout-note
This step will take several minutes to run
:::

```{r}
# Choose a set of soft-thresholding powers
powers = c(c(1:10), seq(from = 12, to = 20, by = 2))

# Call the network topology analysis function
sft_stage = pickSoftThreshold(
  vst_t_stage,             # <= Input data
  powerVector = powers,
  networkType = "signed",
  #corFnc = cor,
  verbose = 5)

# Call the network topology analysis function
sft_leachate = pickSoftThreshold(
  vst_t_leachate,             # <= Input data
  powerVector = powers,
  networkType = "signed",
  #corFnc = cor,
  verbose = 5)
```

```{r}
par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft_stage$fitIndices[, 1],
     -sign(sft_stage$fitIndices[, 3]) * sft_stage$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence stage")
)
text(sft_stage$fitIndices[, 1],
     -sign(sft_stage$fitIndices[, 3]) * sft_stage$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.6, col = "red")
plot(sft_stage$fitIndices[, 1],
     sft_stage$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity stage")
)
text(sft_stage$fitIndices[, 1],
     sft_stage$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```

```{r}
par(mfrow = c(1,2));
cex1 = 0.9;

plot(sft_leachate$fitIndices[, 1],
     -sign(sft_leachate$fitIndices[, 3]) * sft_leachate$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit, signed R^2",
     main = paste("Scale independence leachate")
)
text(sft_leachate$fitIndices[, 1],
     -sign(sft_leachate$fitIndices[, 3]) * sft_leachate$fitIndices[, 2],
     labels = powers, cex = cex1, col = "red"
)
abline(h = 0.6, col = "red")
plot(sft_leachate$fitIndices[, 1],
     sft_leachate$fitIndices[, 5],
     xlab = "Soft Threshold (power)",
     ylab = "Mean Connectivity",
     type = "n",
     main = paste("Mean connectivity leachate")
)
text(sft_leachate$fitIndices[, 1],
     sft_leachate$fitIndices[, 5],
     labels = powers,
     cex = cex1, col = "red")
```
::: callout-note
For stage DEGs (~10k genes)

WGCNA is appropriate

You should run enrichment on those modules

You should treat those as “real” developmental programs

✅ For leachate DEGs (~130 genes)

degPatterns() / clustering on expression profiles is the right tool

These clusters are pattern-based, not network-based

You should not claim scale-free structure or co-expression modules

Your instinct to keep these conceptually separate was correct.

How I would describe this in methods (important)

You don’t need to show this plot in the paper, but if asked:

“Because the leachate-responsive gene set was small and pre-filtered for differential expression, scale-free topology criteria were not met across soft-threshold powers. Therefore, leachate-associated expression patterns were summarized using trajectory-based clustering (degReport) rather than co-expression network inference.”
:::

For **signed networks**, the authors explicitly recommend:

-    Accepting **lower** R^2 (often 0.6–0.7)

-   Prioritizing **reasonable connectivity**

-   Choosing a power near the **knee**, not the ceiling

<!-- -->

-   Each red number is a candidate power (1–20)
-   The red horizontal line is the cut-off of 0.6 for the scale-free
    topology fit index
-   As power increases, the network becomes more scale-free[^1]
-   We want to find the lowest power where the curve crosses the red
    line or plateaus near it

[^1]:
    ## what “scale-free” means

    In a **scale-free network**:

    -   **Most genes have very few connections**

    -   **A small number of genes (hubs) have many connections**

    -    There is **no “typical” number of connections** (no
        characteristic scale)

    -    This contrasts with a random or regular network, where most
        nodes have roughly the same number of connections.

::: callout-note
“A soft-thresholding power of 8 was selected, as it was the lowest value
achieving a high scale-free topology fit (R2≈0.9) while maintaining
sufficient mean connectivity.”
:::

## Create the network using `blockwiseModules`

Network type + TOM type (most important conceptual choice)

 **`signed`**`network & corType`

Signed networks are recommended for gene expression because they
preserve direction: positively co-expressed genes cluster together;
negative correlations don’t create “same module” structure. If you care
about “genes moving together” (typical): **signed.** If you want modules
that allow anti-correlated genes together: **unsigned**

Use:

-    `networkType = "signed"`

-   `TOMType = "signed"`

Correlation choice for VST

For VST expression, the standard is:

-   `corType = "pearson"` **if you trust the data are clean**

### `minModuleSize`

Because you’re using **DEGs only**, you often have fewer genes and
stronger signal. Common choices:

-   20–30 for DEG lists (\~100–3000 genes)

-   50 if you want only big, robust modules

### `deepSplit`

Controls how aggressively the dendrogram is cut:

-    1 = conservative (fewer, larger modules)

-   2 = default sweet spot

-   3–4 = more modules (often too fragmented for DEG-only sets)

-   **Start with 2.** If you get one giant module, go to 3. If you get
    tons of tiny modules, go to 1.

### `mergeCutHeight`

-   Controls module merging by eigengene similarity. Smaller = stricter
    (less merging).

-   0.25 = common default (more merging)

-    **0.15** = stricter (keeps modules separate)

-   0.10 = very strict

-   For DEG-only, **0.15** is often good (keeps distinct patterns).

```{r}
net <- blockwiseModules(
  vst_t_stage,
  power = 10, # choose via pickSoftThreshold
  networkType = "signed", # cluster positively co-expressed genes together
  TOMType = "signed",
  corType = "pearson",
  deepSplit = 1,
  minModuleSize = 50,
  mergeCutHeight = 0.3,
  numericLabels = TRUE,
  saveTOMs = FALSE,
  verbose = 3
)
```

```{r}
stage_deg_modules <- data.frame(
  gene_id = colnames(vst_t_stage),            # your DEG gene IDs
  WGCNAmodule  = labels2colors(net$colors)     # convert numeric → color
) %>%
  mutate(
    WGCNAmodule = factor(WGCNAmodule)
  )
```

```{r}
# Convert labels to colors for plotting
mergedColors = labels2colors(net$colors)

# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  net$dendrograms[[1]],
  mergedColors[net$blockGenes[[1]]],
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05 )
```

## Relate module (color cluster) assignments to treatment groups

```{r}
stage_deg_modules %>%
  dplyr::group_by(WGCNAmodule) %>% 
  dplyr::summarise(n_genes = dplyr::n(), .groups = "drop") %>% 
  arrange(desc(n_genes)) %>% 
  knitr::kable()
```



## Model Eigengenes


> WGCNA will calculate an Eigengene (hypothetical central gene) for each
> module

```{r}
# Get Module Eigengenes per cluster
eig <- moduleEigengenes(vst_t_stage, mergedColors)$eigengenes
```

```{r}
# Reorder modules so similar modules are next to each other
eig <- orderMEs(eig)
module_order = names(eig) %>% gsub("ME","", .)

# Add treatment annotations
metaeig <- cbind(eig, meta_mat[rownames(eig), , drop = FALSE]) # reorders meta_mat rownames to match eig rownames

# make dataframe
ME_df <- metaeig %>% 
  rownames_to_column(var = "sample_id")
```

```{r}
# tidy & plot data
ME_df_long <-  ME_df %>%
  pivot_longer(c(2:7)) %>% 
  mutate(
    name = gsub("ME", "", name),
    name = factor(name, levels = module_order)
  )

ME_df_long %>% ggplot(., aes(x=sample_id, y=name, fill=value)) +
  geom_tile() +
  theme_bw() +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1,1)) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title = "Module-trait Relationships", y = "Modules", fill="corr")
```




## Examine Expression Profiles

```{r}
expr_pat <- stage_deg_modules %>% 
left_join(vst_mat_stage_df)

expr_pat_long <- expr_pat %>% 
  pivot_longer(
    cols = -c(gene_id, WGCNAmodule),
    names_to = "sample_id",
    values_to = "vst"
  ) %>% 
  left_join(metadata_or, by = "sample_id") %>% 
  group_by(gene_id) %>% 
  mutate(
    z = as.numeric(scale(vst))
  ) %>% 
  ungroup()
```

```{r}
ggplot(expr_pat_long, aes(x=stage, y=z))+
  geom_line(aes(group = gene_id), alpha = 0.05, linewidth = 0.01)+
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  facet_wrap(~WGCNAmodule)
```

Cleavage-high → decreases over time

turquoise: strongest “cleavage program” (high at cleavage, steadily down)

Prawnchip-peak (hump-shaped)

yellow: classic “mid-stage / prawnchip program” (up at prawnchip, down again by early gastrula)

Prawnchip-dip (valley-shaped)
green and red: cleavage-high → dip at prawnchip → rebound a bit at early gastrula
These often represent “processes suppressed mid-stage then reactivated” or a mixed module that could merge/split depending on parameters.

Early gastrula-high (increasing)

blue and brown: “late program” (low at cleavage, increasing toward early gastrula)
grey

grey is the WGCNA “unassigned” bucket (genes that didn’t fit cleanly). Don’t enrich/interpret grey as a biological module.

We do have the three big biological programs we expected (early / mid / late), but WGCNA is splitting them into submodules (early = turquoise; mid = yellow/green+red; late = blue+brown)... with the mid stage having a up/dip distinction.

Keep modules but label them as belonging to one of 4 programs

Assign each module a “stage label” based on expression pattern (cleavage peak / prawn peak / prawn dip / early gastrula peak)

Then do enrichment by the grouped “meta-programs”

### Pick peaks & dips via Model Eigengene
Summarize ME by stage and assign objective module labels
- Compute stage means for each ME
```{r}
ME_stage_means <- ME_df %>%
  pivot_longer(cols = starts_with("ME"),
               names_to = "ME",
               values_to = "eigengene") %>%
  group_by(ME, stage) %>%
  summarize(mean_ME = mean(eigengene, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = stage, values_from = mean_ME)

ME_stage_means %>% knitr::kable()
```

Convert ME name → module_color and classify pattern

This uses a simple, transparent rule set:

peak stage = stage with highest mean ME

also classify shape: monotonic up/down vs mid peak vs mid dip
```{r}
stage_levels <- c("cleavage", "prawnchip", "earlygastrula")  # adjust to your exact names

ME_classified <- ME_stage_means %>%
  mutate(
    module_color = sub("^ME", "", ME),

    # pull into numeric vectors for easy comparisons
    cleav = .data[[stage_levels[1]]],
    prawn = .data[[stage_levels[2]]],
    gastr = .data[[stage_levels[3]]],

    peak_stage = stage_levels[max.col(cbind(cleav, prawn, gastr), ties.method = "first")],

    pattern_type = case_when(
      cleav > prawn & prawn > gastr ~ "monotonic_down",
      cleav < prawn & prawn < gastr ~ "monotonic_up",
      prawn > cleav & prawn > gastr ~ "hump",
      prawn < cleav & prawn < gastr ~ "valley",
      TRUE                          ~ "other"
    ),

    # This is your final grouping variable (edit labels as you like)
    module_stage = case_when(
      pattern_type == "monotonic_down" ~ "cleavage_up",
      pattern_type == "monotonic_up"   ~ "earlygastrula_up",
      pattern_type == "hump"           ~ "prawnchip_up",
      pattern_type == "valley"         ~ "prawnchip_dip",
      TRUE                             ~ "unassigned"
    )
  ) %>%
  select(module_color, peak_stage, pattern_type, module_stage)

ME_classified %>% knitr::kable()
```
```{r}
stage_deg_modules_ME <- stage_deg_modules %>%
  left_join(ME_classified, by = c("WGCNAmodule" = "module_color"))

ME_df %>%
  pivot_longer(cols = starts_with("ME"), names_to = "ME", values_to = "eigengene") %>%
  mutate(module_color = sub("^ME", "", ME)) %>%
  left_join(ME_classified, by = "module_color") %>%
  ggplot(aes(x = stage, y = eigengene, group = sample_id)) +
  geom_line(alpha = 0.1) +
  stat_summary(aes(group = 1), fun = median, geom = "line", linewidth = 1) +
  facet_wrap(~ module_color, scales = "free_y") +
  theme_bw()

```



## Extract & save 

### Leachate clusters from degReport
Per-gene meta-dataframe with cluster assignments
Make a dataframe with one row per gene and its cluster assignment
```{r}
leachate_deg_clusters <- leachate130_GEPs$df

head(leachate_deg_clusters)
```
tidy save to csv without rownames
```{r, eval=FALSE}
write_csv(leachate_deg_clusters, file.path(output_path,
                                            "leachate_DEG_clusters.csv"))
```

Save zscores
```{r. eval=FALSE}
write_csv(leachate_cluster_z, file.path(output_path,
                                    "leachate_DEG_cluster_zscores.csv"))
```


### Stage modules from WGCNA
          
Per-gene meta-dataframe with module assignments
Make a dataframe with one row per gene and its module assignment and "stage label" 

```{r}
head(stage_deg_modules_ME) %>% knitr::kable()
```

```{r, eval=FALSE}
write_csv(stage_deg_modules_ME,
          file.path(output_path,"stage_DEG_WGCNA_modules.csv"))
```

# Summary & Next Steps
```{r}
module_summary <- stage_deg_modules_ME %>%
  count(module_stage, WGCNAmodule, name = "n_genes") %>%
  arrange(module_stage, desc(n_genes))

module_summary %>% knitr::kable()
```
We generated and saved:

- stage_DEG_WGCNA_modules.csv
- leachate_DEG_clusters.csv

These define membership into groups of genes with similar expression patterns. 
We can perform functional enrichment on these modules and clusters now!

```{r}
sessionInfo()
```

