---
title: "Look for GOIs in DEGs"
subtitle: "in comparison to Morgan et al. 2022 "
author: "Sarah Tanja"
date: 2026-05-17
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE) # Hide messages
```

# Background

> The GOIs are:
>
> 1.  17β-hydroxysteroid dehydrogenases type 14 (17β HSD14)
>
> 2.  17β-hydroxysteroid dehydrogenases type 12 (17β HSD12)
>
> 3.   Niemann-Pick C type 2 (NPC2)
>
> 4.  Equistatin (EI)
>
> 5.  Complement component C3 (C3)
>
> 6.  Cathepsin L (CTSL)
>
> 7.  Patched domain-containing protein 3 (PTCH3)
>
> 8.  Smoothened (SMO)
>
> 9.  Desert Hedgehog (DHH)
>
> 10. Zinc finger protein GLI2 (GLI2)
>
> 11. Vitellogenin (VTG)
>
> These GOIs were selected because of functional associations with
> steroid hormone biosynthesis; cholesterol binding/transport; immunity;
> phagocytosis; or Hedgehog signaling.

# Load Libraries

```{r libraries}
library(tidyverse)
library(GenomicRanges)
library(Biostrings)
library(kableExtra)
library(ggsidekick)
library(ggrepel)
```

# Load DEG lists

```{r}
sig_summary <- read_csv("../../output/06_deg/spawnnight/sig_summary.csv")

length(unique(sig_summary$gene_id))
```

130 DEGs including Wald and LRT contrasts

```{r}
sig_summary_Wald <- sig_summary %>% 
  dplyr::filter(!list %in% c("joint","int"))

length(unique(sig_summary_Wald$gene_id))
```

122 DEGs only Wald contrasts (specific stagexlevel vs control)

##### Read in gene lists

```{r}
# significant genes list
sig_genes_leachate <- read.csv("../../output/06_deg/spawnnight/sig_genes_leachate.csv")
```

##### Import annotated EggNog GO annotation dataframe

```{r}
EggNog <- read_csv("../../output/07_enrichment/goseq/universe_genes_anno_semi.csv") 

EggNog %>% names()
```

Filter for our 130 DEGs

```{r}
EggNog_leachate_degs <- EggNog %>% 
  dplyr::filter(gene_id %in% sig_genes_leachate$gene_id) %>% 
  dplyr::select(gene_id, Description, Preferred_name, PFAMs, peak_stage)

length(unique(EggNog_leachate_degs$gene_id))
```

Save it!

```{r}
write_csv(EggNog_leachate_degs, 
          "../../output/06_deg/spawnnight/EggNog_leachate_degs.csv")
```

How do we look for those 11 GOIs??

Let's look across EggNog Descriptions & PFAMs... 

```{r}
goi_patterns <- tribble(
  ~GOI, ~pattern,
  "HSD17B14 / 17β-HSD14", "HSD17B14|HSD14|17.?beta.*hydroxysteroid|hydroxysteroid.*14",
  "HSD17B12 / 17β-HSD12", "HSD17B12|HSD12|17.?beta.*hydroxysteroid|hydroxysteroid.*12",
  "NPC2", "NPC2|Niemann.?Pick|cholesterol.*binding|sterol.*transport",
  "Equistatin / EI", "equistatin|cystatin|thyropin|protease inhibitor|cathepsin inhibitor",
  "C3", "\\bC3\\b|complement component|complement C3",
  "CTSL", "CTSL|cathepsin.?L|cathepsin",
  "PTCH3", "PTCH3|patched|hedgehog receptor",
  "SMO", "\\bSMO\\b|smoothened",
  "DHH", "\\bDHH\\b|desert hedgehog|hedgehog",
  "GLI2", "GLI2|zinc finger.*GLI|GLI zinc finger|glioma-associated",
  "VTG", "\\bVTG\\b|vitellogenin"
)

search_cols <- c(
  "Description", "Preferred_name", "PFAMs"
)

deg_search <- EggNog_leachate_degs %>%
  mutate(search_text = str_c(across(any_of(search_cols)), collapse = " | "))

goi_hits <- goi_patterns %>%
  mutate(matches = map(pattern, ~ deg_search %>%
                         filter(str_detect(search_text, regex(.x, ignore_case = TRUE))))) %>%
  unnest(matches)

goi_hits %>%
  select(GOI, gene_id, Preferred_name, Description, PFAMs)
```
Hmmmm.... non model organism frustration


Let's describe the top log2fold change genes instead?

```{r top-log2FC}
sig_df <- sig_summary %>% 
left_join(., EggNog_leachate_degs, by = "gene_id") %>% 
  filter(gene_id %in% sig_genes_leachate$gene_id)
  
# Group by cluster/peak_stage and print the top 3 log2Fold changes....
sig_df %>% 
  group_by(peak_stage) %>% 
  slice_max(order_by = abs(log2FoldChange), n=7) %>% 
  summarize(top_hits = paste0(
      Preferred_name,
      ": ",
      Description,
      collapse = "; "
    ),
    .groups = "drop") %>% 
  kable()
```


Prepare to join

```{r}
gene2go <- read_csv("../../output/07_enrichment/goseq/gene2go.csv")

gene2go_join <- gene2go %>% 
  as.tibble() %>% 
  dplyr::rename(go = category) %>% 
  dplyr::filter(gene %in% sig_summary$gene_id)

length(unique(gene2go_join$gene))
length(unique(gene2go_join$go))

gene2go_join %>% names()
```
Ok we have 110 DE leachate-responsive genes, that map to 3295 go terms

Now we want the descriptions for those go terms!


```{r}
# read in the terms
termsGO <- read_csv("../../output/07_enrichment/goseq/universe_reducedTerms_GO.csv") 

# column names
termsGO %>% names()
length(unique(termsGO$go))

# Do these terms match what we have in our gene2go_join df?
setdiff(gene2go_join$go, termsGO$go)
```


```{r}
EggNog_join <- EggNog_leachate_degs %>% 
  dplyr::rename(gene = gene_id)
EggNog_join %>% names()
```
```{r}
sig_sum_join <- sig_summary %>% 
  dplyr::rename(gene = gene_id)

sig_sum_join %>% names()
```

```{r big-join}

gene2go2bp <- gene2go %>% 
  left_join(., reducedGO, by = "go") %>% 
  left_join(., EggNog_leachate_degs, by = c("gene" = "gene_id")) %>% 
  left_join(., sig_summary, by = c("gene" = "gene_id")) %>% 
  dplyr::select(gene, pattern, list, log2FoldChange, lfcSE, padj, Description, Preferred_name, PFAMs, term, parentTerm)

length(unique(gene2go$gene))
length(unique(gene2go2bp$gene))
```

# Look for GPCR-associated go terms in DEG list
```{r}
# read in the terms
termsGO <- read_csv("../../output/07_enrichment/goseq/GO_leachate.csv") 

termsGO %>% names()

length(unique(termsGO$category))
```
12,599 GO terms... really? 130 DEGs expand to *that many* GO terms??

No.. that is all GO terms in universe!
We have to filter for NumDEInCat > 0!
```{r}
de_terms <- termsGO %>% 
  dplyr::filter(numDEInCat > 0)

length(unique(de_terms$category))
```

```{r gpcr-terms}
gpcr_terms <- c(
  "G protein",
  "GPCR",
  "signal transduction",
  "receptor",
  "transmembrane signaling",
  "second messenger",
  "hedgehog",
  "adenylate cyclase",
  "Wnt",
  "cell signaling"
)

gpcrterms <- de_terms %>% 
  filter(
    str_detect(
      term,
      regex(paste(gpcr_terms, collapse = "|"), ignore_case = TRUE)
    )
  )

head(gpcrterms)

length(gpcrterms$term)
```
74 GO terms ... Which DE genes do those terms map back to???

Some special ones to check:

GO:0008277 - regulation of G protein-coupled receptor signaling pathway
GO:0043401 - steroid hormone receptor signaling pathway
GO:0030518 - nuclear receptor-mediated steroid hormone signaling pathway
GO:0007186 - G protein-coupled receptor signaling pathway
GO:0030111 - regulation of Wnt signaling pathway
GO:0060828 - regulation of canonical Wnt signaling pathway
.
.
.


```{r}
gene2go %>% names()
```
```{r summarize-ngos}
gene2go %>% 
  filter(gene %in% sig_summary$gene_id) %>% 
  group_by(gene) %>% 
  summarize(
    n_go_terms = n_distinct(category),
    .groups = "drop"
  )
```
```{r}
gpcr_flagged <- sig_summary %>% 
  mutate(
    gpcr_adjacent = gene_id %in%
      gene2go$gene[gene2go$category %in% gpcrterms$category]
  )


length(unique(gpcr_flagged$gene_id))
length(unique(gpcr_flagged$gene_id))
```

How many of the 130 are GPCR adjacent?
```{r}
gpcr_flagged %>% 
  group_by(cluster) %>% 
  summarize(
    n_gpcr_genes = n_distinct(gene_id[gpcr_adjacent == TRUE]),  
    .groups = "drop"
  )
```
```{r}
16+8+6+2
```


```{r}
gpcr_genes <- gpcr_flagged %>% 
  dplyr::filter(gpcr_adjacent == TRUE) %>% 
  left_join(., EggNog_leachate_degs, by = "gene_id")
```

Save list of GPCR genes and descriptions:
```{r}
write_csv(gpcr_genes, "../../output/07_enrichment/gpcr/gpcr_genes.csv")
```


# Volcano plots
```{r custom-theme}
theme_sleek_axe <- function() {
  theme_sleek() +
    theme(
      panel.border = element_rect(
        color = "grey70",
        fill = NA,
        linewidth = 0.5
      ),
      axis.line.x  = element_line(color = "grey70"),
      axis.line.y  = element_line(color = "grey70"),
      
      plot.title = element_text(
        size  = 10,
        color = "grey40",
        hjust = 0.5,
        margin = margin(b = 20)
      ),
      
      strip.text = element_text(size = 8, color = "grey40"),
      
      plot.subtitle = element_text(
        size   = 8,
        color  = "grey40",
        hjust  = 0.5,
        margin = margin(b = 40)
      ),
      
      axis.title = element_text(size = 8, color = "grey40"),
      axis.text = element_text(size = 7, color = "grey40"),
      legend.title = element_text(size = 8, color = "grey40"),
      legend.text = element_text(size = 7, color = "grey40")
    )
}

```


```{r volcano}
volcano_df <- gpcr_flagged %>% 
  mutate(
    gpcr_adjacent = if_else(gpcr_adjacent == TRUE, "GPCR-adjacent", "Other"),
    neg_log10_padj = -log10(padj)
  )

ggplot(volcano_df, aes(x = log2FoldChange, y = neg_log10_padj)) +
  geom_point(
    aes(color = gpcr_adjacent),
    alpha = 0.7,
    size = 2
  ) +
  scale_color_manual(
    values = c("GPCR-adjacent" = "red", "Other" = "grey70")
  ) +
  facet_wrap(~ cluster) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x = expression(log[2]~fold~change),
    y = expression(-log[10]~adjusted~italic(p)),
    color = NULL
  ) +
  theme_sleek_axe()
```


