---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "", # Prevents appending '##' to beginning of lines in code output)
  dev = "jpeg",
  dpi = 600,
  fig.retina = 2)        
```

# Background

So we have 11 different 'significant gene' lists. Each list indicates that leachate had a significant effect somewhere:

LRT 
- `sig_cat_LRT`, 44 DEGs, this list are genes that had an overall impact by leachate interacting with stage (full vs reduced, testing interaction term as a whole)

Interaction effects
- `sig_prawnchip_high.int`, 1 DEG
- `sig_prawnchip_mid.int`, 5 DEGs
- `sig_prawnchip_low.int`, 12 DEGs
- `sig_earlygastrula_high.int`, 3 DEGs
- `sig_earlygastrula_mid.int`, 11 DEGs
- `sig_earlygastrula_low.int`, 11 DEGs

Stage x leachate effects
- `sig_cleavage_low.vs.control`, 39 DEGs
- `sig_cleavage_mid.vs.control`, 77 DEGs
- `sig_cleavage_high.vs.control`, 18 DEGs
- `sig_earlygastrula_high.vs.control`, 22 DEGs

We also have a normalized count matrix generated via DESeq2
- `norm_count_mat.csv`


## Inputs

Everything from 06_deg/results
```{r}
library(readr)

path <- "../../output/06_deg/results/"
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
nm    <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}

```

Metadata
```{r}
metadata_or <- read.csv("../../metadata/metadata_or.csv")
```

## Outputs
- Gene count plots
- PCA of significant genes
- Heatmap(s) showing significant genes across DEG lists

# Setup

## Load packages

```{r}
# Load packages
library(viridis)
library(tidyverse)
library(DEGreport)
library(ashr)
library(flexoki)
library(plotly)
library(vsn)
library(pheatmap)
library(ComplexUpset)
library(ggplot2)
library(ggplotify)
library(patchwork)   # for wrap_plots(...)
library(cowplot)     # for plot_grid & get_legend
library(dplyr)
library(grid) 
library(purrr)
library(forcats)
library(ggforce)   # for facet_wrap_paginate
```

## Long data for plotting
```{r}
# convert -> long tidy dataframe
norm_count_long <- as.data.frame(norm_count_mat) %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "sample_id",
    values_to = "expression"
  )

# 3. join sample metadata
norm_count_long <- norm_count_long %>% left_join(metadata_or, by = "sample_id")
```

How many samples are in each treatment for each significant gene?

This bit of code lets us see how many samples are contributing to the mean expression of a gene broken down by group (aka condition ... leachate x stage)

```{r}
sample_counts <- norm_count_long %>% 
  dplyr::count(gene_id, group) %>%   # count rows per gene_id + group
  pivot_wider(names_from = group,              # make each group a column
              values_from = n, 
              values_fill = 0)                 # fill missing group-gene combos with 0

summary(sample_counts)
```
::: callout-note
At least 4 samples represent each group per significant gene. Most groups are represented by 5 samples... some have 4, 6, and 1 group has 7 samples. This is important to check because we want to be able to compare the groups, but if they have vastly different sample sizes that could be problematic. This confirms that the sample sizes are reasonably similar for comparisons.
:::

#  Gene count plot function
```{r}
plot_sig_genes <- function(norm_count_long, 
                           sig,                      # vector of gene_ids OR df with gene_id (+ optional padj)
                           gene_col = "gene_id",
                           padj_col = NULL,         # set to the padj column name if available (e.g., "padj")
                           title = NULL,
                           subtitle = NULL,
                           ci_level = 0.95) {

  # --- Normalize the significant list to a tibble with gene_id and optional padj ---
  if (is.vector(sig)) {
    sig_df <- tibble(gene_id = as.character(sig), padj = NA_real_)
  } else if (is.data.frame(sig)) {
    stopifnot(gene_col %in% names(sig))
    sig_df <- sig %>%
      rename(gene_id = !!gene_col) %>%
      mutate(gene_id = as.character(gene_id))
    if (!is.null(padj_col) && padj_col %in% names(sig_df)) {
      sig_df <- sig_df %>% 
        rename(padj = !!padj_col) %>%
        group_by(gene_id) %>%
        summarise(padj = suppressWarnings(min(padj, na.rm = TRUE)), .groups = "drop")
      sig_df$padj[!is.finite(sig_df$padj)] <- NA_real_  # handle all-NA
    } else {
      sig_df <- sig_df %>% mutate(padj = NA_real_)
    }
    sig_df <- sig_df %>% distinct(gene_id, .keep_all = TRUE)
  } else {
    stop("`sig` must be a character vector of gene_ids or a data.frame.")
  }

  # --- Keep only rows present in norm_count_long and prep labels/factors ---
  plot_df <- norm_count_long %>%
    semi_join(sig_df, by = "gene_id") %>%
    left_join(dplyr::select(sig_df, gene_id, padj), by = "gene_id") %>%
    mutate(
      gene_label = ifelse(
        is.na(padj), gene_id,
        paste0(gene_id, "\nFDR: ", formatC(padj, format = "g", digits = 3))
      ),
      leachate = forcats::fct_relevel(as.factor(leachate), c("control","low","mid","high"))
    )

  # --- Summary stats per facet/group/level (mean ± 95% CI) ---
  alpha <- 1 - ci_level
  sum_df <- plot_df %>%
    dplyr::group_by(gene_id, gene_label, stage, leachate) %>%
    dplyr::summarise(
      n   = dplyr::n(),
      mean = mean(expression, na.rm = TRUE),
      sd   = sd(expression,  na.rm = TRUE),
      se   = sd / sqrt(n),
      tcrit = qt(1 - alpha/2, pmax(n - 1, 1)),
      ci   = tcrit * se,
      .groups = "drop"
    ) %>%
    mutate(
      ymin = mean - ci,
      ymax = mean + ci
    )

  # --- Fallback title/subtitle ---
  if (is.null(title)) {
    title <- paste0("Significant genes (n = ", dplyr::n_distinct(plot_df$gene_id), ")")
  }
  if (is.null(subtitle)) {
    subtitle <- paste0("Group means ± ", round(ci_level*100), "% CI per stage; points are raw observations")
  }

  # --- Plot: raw points + summary line/points + CI error bars ---
  ggplot(plot_df, aes(x = leachate, y = expression, color = stage)) +
    labs(
      title = title,
      subtitle = subtitle,
      y = "Normalized gene counts",
      x = "PVC leachate concentration (milligrams of microplastic/ Liter of seawater)"
    ) +
    # raw points
    geom_jitter(width = 0.12, alpha = 0.6, size = 1, show.legend = FALSE) +
    # summary error bars + mean points + connecting line
    geom_errorbar(
      data = sum_df,
      aes(y = mean, ymin = ymin, ymax = ymax, group = stage),
      width = 0.08,
      linewidth = 0.5,
      position = position_dodge2(width = 0.0, preserve = "single"),
      show.legend = FALSE
    ) +
    geom_point(
      data = sum_df,
      aes(y = mean, group = stage),
      size = 1.8,
      stroke = 0.4,
      show.legend = TRUE
    ) +
    geom_line(
      data = sum_df,
      aes(y = mean, group = stage),
      linewidth = 0.6,
      show.legend = FALSE
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 12),
      plot.subtitle = element_text(size = 8),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      strip.text = element_text(size = 7, lineheight = 0.9, margin = margin(2,2,2,2))
    ) +
    scale_x_discrete(
      expand = expansion(mult = c(0.1, 0.1)),
      labels = c(control = "Control", low = "Low (0.01)", mid = "Mid (0.1)", high = "High (1)")
    ) +
    facet_wrap(~ gene_label, scales = "free_y")
}

```

```{r}
plot_sig_genes(norm_count_long, sig_prawnchip_mid.int, gene_col = "gene_id", padj_col = "padj")
```


```{r}
resultsNames(dds_wald)
```

## Interaction terms
-   "stageprawnchip.leachatelow"

-   "stageearlygastrula.leachatelow"

-   "stageprawnchip.leachatemid"

-   "stageearlygastrula.leachatemid"

-   "stageprawnchip.leachatehigh"

-   "stageearlygastrula.leachatehigh"

Each of the above interaction terms tells us the extra effect of the leachate level at a stage (vs the additive sum of the stage and leachate effects) in comparison to a control baseline count. For the interaction terms, we are trying to get the "difference of the differences". For this analysis, we are interested in genes that are differentially expressed in treatment conditions, but not in the control condition (ie. genes uniquely changing in a stage due to treatment).

```{r}
# prawnchip interaction terms
res_prawnchip_high.int <- results(dds_wald, name = "stageprawnchip.leachatehigh", alpha = 0.05)
res_prawnchip_mid.int <- results(dds_wald, name = "stageprawnchip.leachatemid", alpha = 0.05)
res_prawnchip_low.int  <- results(dds_wald, name = "stageprawnchip.leachatelow", alpha = 0.05)

# early gastrula interaction terms
res_earlygastrula_high.int <- results(dds_wald, name = "stageearlygastrula.leachatehigh", alpha = 0.05)
res_earlygastrula_mid.int <- results(dds_wald, name = "stageearlygastrula.leachatemid", alpha = 0.05)
res_earlygastrula_low.int <- results(dds_wald, name = "stageearlygastrula.leachatelow", alpha = 0.05)

```

### Prawnchip
#### high
```{r}
summary(res_prawnchip_high.int)
```
::: callout-note
1 gene was found to have significant stage x leachate interaction (adjusted p value \< 0.05)
:::

```{r}
sig_prawnchip_high.int <- res_prawnchip_high.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

sig_prawnchip_high.int
```

::: callout-note
Only one gene showed a significant stage × leachate interaction in the prawnchip stage under the high treatment (Wald test, BH-adjusted p < 0.05). Specifically, **Montipora_capitata_HIv3___RNAseq.g11129.t1** was up-regulated during the prawnchip stage in the high treatment (log2FC = +5.76, padj = 0.026) relative to the expected count levels...
:::

```{r}
plot_sig_genes(norm_count_long, sig_prawnchip_high.int, gene_col = "gene_id", padj_col = "padj")
```

```{r}
ggplot(sig_prawnchip_high.int, aes(x=leachate, y=expression, color = stage, group = stage)) +
  stat_summary(fun = mean, geom = "line") +
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.1) +
  stat_summary(fun = mean, geom = "point", size = 2) +
  facet_wrap(~ gene_id, scales = "free_y")

```


::: callout-note
Visually... I can't see how this indicates an interaction effect in the prawnchip x high samples
:::

::: column-margin

> DESeq2 uses size-factor normalization (the “median-ratio method”). It corrects for library size and composition so you can compare counts across samples. How it works (median-ratio method) Per-gene geometric mean across samples (skip genes with all zeros). Per-sample ratios: for each gene, divide a sample’s count by that gene’s geometric mean. Size factor: the median of those ratios within each sample. Normalized counts: normalized = raw_count / size_factor. These size factors are used as an offset in the negative binomial GLM for the Wald/LRT tests. Normalization is independent of the design formula.

:::

#### mid
```{r}
summary(res_prawnchip_mid.int)
```
::: callout-note
5 genes were found to have significant prawnchip stage x mid leachate interaction (adjusted p value \< 0.05)
:::

```{r}
sig_prawnchip_mid.int <- res_prawnchip_mid.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r}
plot_sig_genes(norm_count_long, sig_prawnchip_mid.int, gene_col = "gene_id", padj_col = "padj")
```

#### low
```{r}
summary(pcl)
```

```{r}
sig_pcl <- pcl %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

### Early gastrula
#### high
```{r}
summary(res_earlygastrula_high.int)
```

```{r}
sig_earlygastrula_high.int_egh <- res_earlygastrula_high.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

#### mid
```{r}
summary(res_earlygastrula_mid.int)
```

```{r}
sig_earlygastrula_mid.int <- res_earlygastrula_mid.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_earlygastrula_mid.int, "../../output/06_deg/contrasts/sig_earlygastrula_mid.int.csv")
```
#### low
```{r}
summary(res_earlygastrula_low.int)
```

```{r}
sig_earlygastrula_low.int <- res_earlygastrula_low.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_earlygastrula_low.int, "../../output/06_deg/contrasts/sig_earlygastrula_low.int.csv")
```

## vs. Control contrasts

See [A guide to designs and contrasts in DESeq2](https://www.atakanekiz.com/technical/a-guide-to-designs-and-contrasts-in-DESeq2/) by Dr. Atakan Ekiz

```{r}
resultsNames(dds_wald)
```
```{r}
# cleavage 
res_cleavage_low.vs.control <- results(dds_wald, name = "leachatelow", alpha = 0.05)
res_cleavage_mid.vs.control  <- results(dds_wald, name = "leachatemid", alpha = 0.05)
res_cleavage_high.vs.control <- results(dds_wald, name = "leachatehigh", alpha = 0.05)

# prawnchip
res_prawnchip_low.vs.control <- results(dds_wald, contrast = list(c("leachatelow", "stageprawnchip.leachatelow")), alpha = 0.05)
res_prawnchip_mid.vs.control  <- results(dds_wald, contrast = list(c("leachatemid",  "stageprawnchip.leachatemid")), alpha = 0.05)
res_prawnchip_high.vs.control <- results(dds_wald, contrast = list(c("leachatehigh", "stageprawnchip.leachatehigh")), alpha = 0.05)

# earlygastrula
res_earlygastrula_low.vs.control <- results(dds_wald, contrast = list(c("leachatelow", "stageearlygastrula.leachatelow")), alpha = 0.05)
res_earlygastrula_mid.vs.control  <- results(dds_wald, contrast = list(c("leachatemid",  "stageearlygastrula.leachatemid")), alpha = 0.05)
res_earlygastrula_high.vs.control <- results(dds_wald, contrast = list(c("leachatehigh", "stageearlygastrula.leachatehigh")), alpha = 0.05)

```

## low leachate vs control

for cleavage (baseline stage)
> Because you fit ~ 0 + stage + leachate + stage:leachate, **the coefficient leachatelow is the low vs control effect at the reference stage**. From your resultsNames() (no stagecleavage.leachatelow term is present) the reference stage is cleavage, so Low vs control at cleavage is simply the leachatelow coefficient.

```{r}
summary(res_cleavage_low.vs.control)
```

::: callout-note
39 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the low leachate treatment compared to the control. 14 LFC \> 0 & 25 LFC \< 0.
:::

```{r}
sig_cleavage_low.vs.control <- res_cleavage_low.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_cleavage_low.vs.control, "../../output/06_deg/contrasts/sig_cleavage_low.vs.control.csv")
```

for prawnchip

```{r}
res_lvc_pc <- results(dds_wald, contrast = list(c("leachatelow",
                                                  "stageprawnchip.leachatelow")), 
                      alpha = 0.05)
summary(res_lvc_pc)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the low leachate treatment when compared to the control.
:::

for early gastrula

```{r}
res_lvc_eg <- results(dds_wald, contrast = list(c("leachatelow",
                                                 "stageearlygastrula.leachatelow")),
                      alpha = 0.05)
summary(res_lvc_eg)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early gatrula stage in the low leachate treatment when compared to the control.
:::

## mid leachate vs. control (mvc)

for cleavage

```{r}
res_mvc_cl <- results(dds_wald, name = "leachatemid", alpha = 0.05)
summary(res_mvc_cl)
```

::: callout-note
77 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the mid leachate treatment compared to the control. 77 LFC \> 0. These 77 genes were all found in more abundance compared to the controls.
:::

```{r}
sig_mvc_cl <- as.data.frame(subset(res_mvc_cl, padj<0.05))
sig_mvc_cl <- sig_mvc_cl %>% rownames_to_column(., var = "gene_id")
write_csv(sig_mvc_cl, "../../output/06_deg/contrasts/sig_mvc_cl.csv")
```

for prawnchip

```{r}
res_mvc_pc <- results(dds_wald, contrast = list(c("leachatemid",
                                                  "stageprawnchip.leachatemid")))
summary(res_mvc_pc)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the mid leachate treatment when compared to the control.
:::

for early gastrula

```{r}
res_mvc_eg <- results(dds_wald, contrast = list(c("leachatemid",
                                                  "stageearlygastrula.leachatemid")))
summary(res_mvc_eg)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the mid leachate treatment when compared to the control.
:::

## high leachate vs. control (hvc)

for cleavage

```{r}
res_hvc_cl <- results(dds_wald, name = "leachatehigh", alpha=0.05)
summary(res_hvc_cl)
```

::: callout-note
18 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the high leachate treatment compared to the control. 8 LFC \> 0 , 16 LFC \<0.
:::

```{r}
sig_hvc_cl <- as.data.frame(subset(res_hvc_cl, padj<0.05))
sig_hvc_cl <- sig_hvc_cl %>% rownames_to_column(., var = "gene_id")
write_csv(sig_hvc_cl, "../../output/06_deg/contrasts/sig_hvc_cl.csv")
```

for prawnchip

```{r}
res_hvc_pc <- results(dds_wald, contrast = list(c("leachate_high_vs_control",
                                                  "stageprawnchip.leachatehigh")))
summary(res_hvc_pc)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the high leachate treatment when compared to the control.
:::

for early gastrula

```{r}
res_hvc_eg <- results(dds_wald, contrast = list(c("leachate_high_vs_control",
                                                  "stageearlygastrula.leachatehigh")))
summary(res_hvc_eg)
```

::: callout-note
32 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the high leachate treatment compared to the control. 32 LFC \> 0 .
:::

```{r}
sig_hvc_eg <- as.data.frame(subset(res_hvc_eg, padj<0.05))
sig_hvc_eg <- sig_hvc_eg %>% rownames_to_column(., var = "gene_id")
write_csv(sig_hvc_eg, "../../output/06_deg/contrasts/sig_hvc_eg.csv")
```

# Summarize results

# Presence/absence matrix
```{r}
presence_mat <- sig_summary %>%
  distinct(list, gene_id) %>%
  mutate(value = 1L) %>%
  tidyr::pivot_wider(names_from = list, values_from = value, values_fill = 0)
```


```{r}
plot_df <- tidyr::pivot_longer(presence_mat, -gene_id,
                               names_to = "list", values_to = "present")

ggplot(plot_df, aes(x = list, y = gene_id, fill = factor(present))) +
  geom_tile() +
  scale_fill_manual(values = c("0" = "white", "1" = "black"), guide = "none") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# helper to standardize any DESeqResults into a tidy df
as_tbl <- function(res, name){
  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    transmute(
      gene_id,
      result_name = name,
      log2FC      = log2FoldChange,
      pval        = pvalue,
      padj        = padj
    )
}
```

```{r}
# EXAMPLE: replace the ... with your actual DESeqResults objects
res_list <- list(
  # Wald main effects at reference stage (cleavage)
  leachate_low  = results(dds_wald, name = "leachatelow"),
  leachate_mid  = results(dds_wald, name = "leachatemid"),
  leachate_high = results(dds_wald, name = "leachatehigh"),

  # Stage-specific low/mid/high (additive + interaction)
  prawn_low  = results(dds_wald, contrast = list(c("leachatelow",  "stageprawnchip.leachatelow"))),
  prawn_mid  = results(dds_wald, contrast = list(c("leachatemid",  "stageprawnchip.leachatemid"))),
  prawn_high = results(dds_wald, contrast = list(c("leachatehigh", "stageprawnchip.leachatehigh"))),

  eg_low  = results(dds_wald, contrast = list(c("leachatelow",  "stageearlygastrula.leachatelow"))),
  eg_mid  = results(dds_wald, contrast = list(c("leachatemid",  "stageearlygastrula.leachatemid"))),
  eg_high = results(dds_wald, contrast = list(c("leachatehigh", "stageearlygastrula.leachatehigh"))),

  # Pure interaction terms (differences of effects vs cleavage)
  int_prwn_low  = results(dds_wald, name = "stageprawnchip.leachatelow"),
  sig_prwn_mid  = results(dds_wald, name = "stageprawnchip.leachatemid"),
  int_prwn_high = results(dds_wald, name = "stageprawnchip.leachatehigh"),
  int_eg_low    = results(dds_wald, name = "stageearlygastrula.leachatelow"),
  int_eg_mid   
  int_eg_high  

  # LRT (e.g., test of interaction block or full vs reduced)
  sig_cat_LRT
)
```


```{r}
# Combine into a summary
int_wald_df <- data.frame(
  gene_id = rownames(res_high_pc),
  pc_h_log2FC = res_high_pc$log2FoldChange,
  pc_high_padj = res_high_pc$padj,
  prawnchip_mid_log2FC = res_mid_pc$log2FoldChange,
  prawnchip_mid_padj = res_mid_pc$padj,
  prawnchip_low_log2FC = res_low_pc$log2FoldChange,
  prawnchip_low_padj = res_low_pc$padj,
  earlygastrula_high_log2FC = res_high_eg$log2FoldChange,
  earlygastrula_high_padj = res_high_eg$padj,
  earlygastrula_mid_log2FC = res_mid_eg$log2FoldChange,
  earlygastrula_mid_padj = res_mid_eg$padj,
  earlygastrula_low_log2FC = res_low_eg$log2FoldChange,
  earlygastrula_low_padj = res_low_eg$padj
)

# Focus on the 44 genes
int_wald_df_sig <- int_wald_df[int_wald_df$gene_id %in% LRT_sig_genes$gene_id, ]
```

### Plot heatmap

```{r}
library(pheatmap)

# Step 1: Order and extract log2FC values
log2fc_cols <- c(
  "prawnchip_low_log2FC", "prawnchip_mid_log2FC", "prawnchip_high_log2FC",
  "earlygastrula_low_log2FC", "earlygastrula_mid_log2FC", "earlygastrula_high_log2FC"
)

mat <- as.matrix(int_wald_df_sig[, log2fc_cols])
rownames(mat) <- int_wald_df_sig$gene_id

# Step 2: Rename columns for clarity
colnames(mat) <- c(
  "Prawnchip: Low", "Prawnchip: Mid", "Prawnchip: High",
  "Early Gastrula: Low", "Early Gastrula: Mid", "Early Gastrula: High"
)

# Step 3: Create annotation for stage/leachate
annotation_col <- data.frame(
  Stage = rep(c("Prawnchip", "Early Gastrula"), each = 3),
  Leachate = rep(c("Low", "Mid", "High"), times = 2)
)
rownames(annotation_col) <- colnames(mat)

# Step 4: Prepare significance (padj < 0.05) annotation matrix
padj_cols <- c(
  "prawnchip_low_padj", "prawnchip_mid_padj", "prawnchip_high_padj",
  "earlygastrula_low_padj", "earlygastrula_mid_padj", "earlygastrula_high_padj"
)
padj_mat <- as.matrix(int_wald_df_sig[, padj_cols])
rownames(padj_mat) <- int_wald_df_sig$gene_id
colnames(padj_mat) <- colnames(mat)

# Significance stars with NA handling
sig_labels <- ifelse(is.na(padj_mat), "",
                      ifelse(padj_mat < 0.01, "**",
                        ifelse(padj_mat < 0.05, "*",
                          ifelse(padj_mat < 0.1, ".", ""))))

# Step 5: Save heatmap with significance stars
png("../../output/06_deg/interactions/int_wald_heatmap.png", width = 10, height = 12, units = "in", res = 600)

pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         display_numbers = sig_labels,
         number_color = "black",
         fontsize = 10,
         fontsize_number = 8,
         angle_col = 45,
         border_color = NA)
```

Interpretation notes: \> 🧬 What does the heatmap show? Each cell in the heatmap shows the scaled log2 fold change for a gene in a specific interaction condition. \> \> ✅ Scaling Because you set scale = "row", each gene's values are z-scored across the six interaction conditions \> \> Blue = low relative expression change in that condition \> \> Red = high relative expression change in that condition \> \> The actual color scale is relative within each row, not across the whole matrix. \> \> This helps visualize patterns of expression changes across conditions for each gene, regardless of absolute size. \> \> 📊 What can you learn from this? Are there groups of genes responding similarly? Row clustering may reveal gene modules with similar interaction response profiles. \> \> Which stage + leachate condition drives the interaction? Column patterns (e.g., lots of red in prawnchip_high_log2FC) suggest strong interaction effects in that condition. \> \> Which genes have the strongest differential interactions? Genes with high contrast (strong red/blue) have more dynamic interaction effects. \> \> 🧠 Interpretation Reminder These log2 fold changes are interaction effects --- they reflect how gene expression at a specific `stage` & `leachate` level deviates from the expected additive effect of `stage` and `leachate` alone. \> \> So, a large positive log2FC means: \> \> "At this `stage` and `leachate` level, this gene is more highly expressed than we would expect based on `stage` and `leachate` effects individually." \> \> Likewise, a large negative value = stronger-than-expected suppression.

```{r}
int_wald_df_sig$any_sig <- apply(int_wald_df_sig[, grepl("padj", names(int_wald_df_sig))], 1, function(p) any(p < 0.05, na.rm = TRUE))
```

### PCA of significant genes only

```{r}
# Subset DESeqDataSet to just significant genes
dds_wald_sig <- dds_wald[rownames(dds_wald) %in% sig_cat, ]
vsd_wald_sig <- varianceStabilizingTransformation(dds_wald_sig)

# Get PCA data (only 44 sig DEGs)
pca_sig <- plotPCA(vsd_wald_sig, intgroup = "group", returnData = TRUE)
percentVar_sig <- round(100*attr(pca_sig, "percentVar"))

# Merge with metadata
pca_sig <- merge(pca_sig, metadata_or, by.x = "name", by.y = "sample_id", all.x = TRUE)

# Assign specific colors to each hpf
hpf_colors <- c(
  "cleavage" = flex("cyan", sat=500),
  "prawnchip" = flex("green", sat =500),
  "earlygastrula" = flex("yellow", sat=500)
)

pca_sig$label <- paste0("sample: ", pca_sig$sample_id,
                        "\nStage: ", pca_sig$stage,
                        "\nLeachate: ", pca_sig$leachate)

# Indicate leachate level by saturation (alpha)
leachate_alpha_map <- c("control" = 0.5, "low" = 0.6, "mid" = 0.7, "high" = 0.9)
pca_sig$alpha_val <- leachate_alpha_map[pca_sig$leachate]
```

```{r}
# Original ggplot with a tooltip aesthetic
PCA_sig <- ggplot(pca_sig, aes(PC1, PC2, color = stage, label = name, alpha = alpha_val)) +
  geom_point(size = 1,  stroke = 1) +
  ggtitle("M.cap embryos exposed to pvc leachate, 44 significant genes VST expression") +
  xlab(paste0("PC1: ", percentVar_sig[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_sig[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_sig, tooltip = "label")

```

```{r}
# Original ggplot with a tooltip aesthetic
PCA_sig <- ggplot(pca_sig, aes(PC1, PC2, color = stage, shape = leachate, label = name, alpha = alpha_val)) +
  geom_point(size = 1,  stroke = 1) +
  ggtitle("M.cap embryos exposed to pvc leachate, 44 significant genes VST expression") +
  xlab(paste0("PC1: ", percentVar_sig[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar_sig[2], "% variance")) + 
  coord_fixed() +
  scale_color_manual(values = hpf_colors) +
  stat_ellipse() + 
  guides(alpha = "none") +  # hide alpha legend (optional)
  theme_minimal()

# Convert to interactive plot
ggplotly(PCA_sig, tooltip = "label")
```

# Save gene lists

-   `sig_cat` : List of the 44 genes that show significant interactions between leachate and stage (saved above)

-   `LRT_sig_genes` : List of LRT significant genes, LFC, & padj values (saved above)

-   `Wald_sig_genes` : List of Wald test results, LFC & padj values for each interaction contrast

-   `vsd_sig` : Variance stabilized count matrix of genes from significant list

-   `all_genes` : List of all genes in the dataset (the filtered input to the DESeq object) which will serve as a reference for functional enrichment

```{r}
# The 39 genes that are sig diff from control at low cleavage 
sig_lvc_cl <- sig_lvc_cl %>% 
  mutate(low_cleavage = 1)

# The 77 genes that are sig diff from control at mid cleavage
sig_mvc_cl <- sig_mvc_cl %>% 
  mutate(mid_cleavage = 1)

# The 18 genes that are sig diff from control at high cleavage
sig_hvc_cl <- sig_hvc_cl %>% 
  mutate(high_cleavage = 1)

# The 13 genes that are sig diff from control at high early gastrula
sig_hvc_eg <- sig_hvc_eg %>% 
  mutate(high_earlygastrula = 1)
```

```{r}
common_genes <- Reduce(intersect, list(
  sig_lvc_cl$gene_id,
  sig_mvc_cl$gene_id,
  sig_hvc_cl$gene_id
))

length(common_genes)  # Number of genes common to all cleavage, just 1!
print(common_genes) # And it is...
```

::: callout-note
Montipora_capitata_HIv3\_\_\_TS.g50653.t2" is the only gene common across cleavage with differential expression due to low, mid, and high leachate treatments
:::

# Gene set overlap: Venn Diagram

```{r}
genes1 <- sig_lvc_cl$gene_id
genes2 <- sig_mvc_cl$gene_id
genes3 <- sig_hvc_cl$gene_id
genes4 <- sig_hvc_eg$gene_id
genes5 <- LRT_sig_genes$gene_id
```

Upset requires a binary membership table (presence/absence dataframe)

```{r}
all_genes <- unique(c(genes1, genes2, genes3, genes4, genes5))

upset_data <- data.frame(
  gene_id = all_genes,
  LVC_CL  = all_genes %in% genes1,
  MVC_CL  = all_genes %in% genes2,
  HVC_CL  = all_genes %in% genes3,
  HVC_EG  = all_genes %in% genes4,
  LRT     = all_genes %in% genes5
)
```

```{r}
upset_plot <- upset_data %>% 
  upset(
    intersect = c("LVC_CL", "MVC_CL", "HVC_CL", "HVC_EG", "LRT"),
    base_annotations = list(
      'Intersection size' = intersection_size(
        counts = TRUE
      )
    )
  ) +
  labs(title = "UpSet Plot of Significant Genes (P < 0.05)")

upset_plot
```

```{r}
Wald_sig_genes <- int_wald_df_sig

write_csv(Wald_sig_genes, "../../output/06_deg/interactions/Wald_sig_genes.csv")
```

```{r}
vsd_sig <-  vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  filter(gene_id %in% sig_cat)

write_csv(vsd_sig, "../../output/06_deg/interactions/vsd_sig.csv")
```

```{r}
all_genes <- vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  select(gene_id)

write_csv(all_genes, "../../output/06_deg/interactions/all_genes.csv")
```

::: callout-note
12,565 genes listed in the `all_genes.csv`
:::

# Perform clustering

see Roberts Lab Issue [#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering.

```{r}
clustering_sig_genes <- LRT_sig_genes %>%
  arrange(padj) #genes in rows, stats in columns

matrix <- vsd #genes in rows, sample ID in columns

meta <- metadata_or %>%
  dplyr::select(sample_id, stage, leachate) %>% #sample id and stage, and leachate columns
  column_to_rownames(var = "sample_id")

meta<-meta%>%
  dplyr::select(stage, leachate) #sample ID in rows, one column of lifestage 

# Obtain transformed values for those significant genes
cluster_vst <- matrix[clustering_sig_genes$gene_id, ] #rows are DEGs and columns are sample ID
```

Run with minimum 3 genes per cluster to detect because we have few DEGs.\
- DEGreport vignette, [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)-,Detect%20patterns%20of%20expression,-In%20this%20section)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_clusters <- degPatterns(cluster_vst, metadata = meta, time = "stage", col=NULL, plot=TRUE, minc = 4)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
clusters <- degPatterns(cluster_vst, metadata = meta, time = "leachate", col=NULL, plot=TRUE, minc = 3)
```

```{r}
ggsave(filename="../../output/06_deg/interactions/degPatternscluster_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for leachate (stage grouped all together in one big group, so we can focus in on leachate with the cluster analysis here).

```{r}
# Rename and extract information 
groups <- clusters$df %>%
          rename(gene_id=genes) %>% 
          rename(leachate_cluster=cluster)
```

Save the gene cluster annotation

```{r}
write.csv(groups, "../../output/06_deg/interactions/clusters.csv", row.names = FALSE)
```

Plot is showing Z score of VST normalized gene counts

```{r, warning=FALSE}
cluster_plot<-ggplot(clusters[["normalized"]],
        aes(x=leachate, y=value)) +
  facet_wrap(~ cluster)+
   geom_point(colour="darkgray", size=1) +
  ylab("Z-score expression")+
  xlab("")+
   # change the method to make it smoother
   geom_smooth(aes(group=genes), method = "loess", colour="red", linewidth=0.1, se=FALSE)+
   geom_smooth(aes(group=1), method = "loess", colour="black")+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_vline(xintercept=3, linetype="dotted")+
  geom_vline(xintercept=6, linetype="dotted")+
  theme_classic()+
  theme(
    axis.text.x = element_text(angle=90, hjust=1, vjust=1)
  );cluster_plot

ggsave(cluster_plot, filename="../../output/06_deg/interactions/cluster_plot.png", width=12, height=10)
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!
