---
title: "Multifactorial Differential Gene Expression"
subtitle: "using DESeq2 contrasts"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: full
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

Here we are looking at a multifactorial differential gene expression analysis for developing coral embryos at 3 developmental stages (4 hours post fertilization , 9 hours post fertilization, and 14 hours post fertiliation). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of microplastic pollution on coral reefs.

## Inputs

-   `gcm_filtor.csv` is the unfiltered gene count matrix for 61 samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers removed)

## Outputs

-   `volcano_facet_interaction.png`

# Setup

## Install packages

```{r, eval=FALSE}
if ("tidyverse" %in% rownames(installed.packages()) == 'FALSE') 
  install.packages('tidyverse')
if ("DESeq2" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('DESeq2')
if ("ashr" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('ashr')
if ("plotly" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('plotly')
if ("cowplot" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('cowplot')
if ("pak" %in% rownames(installed.packages()) == 'FALSE')
  install.packages('pak')
```

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("EnhancedVolcano")
#pak::pak("mdscheuerell/flexoki")
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(tidyverse)
library(EnhancedVolcano)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names = FALSE)
head(gcm_filtor)
```

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filt` is the gene count matrix that has been filtered to remove 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA
:::

## Import metadata

```{r}
metadata_filt <- read_csv("../metadata/metadata_filt.csv")
str(metadata_filt)
```

```{r}
# Reorder factors
metadata_filt$embryonic_stage <- as.factor(metadata_filt$embryonic_stage)
metadata_filt$embryonic_stage <- fct_relevel(metadata_filt$embryonic_stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_filt$leachate <- as.factor(metadata_filt$leachate)
metadata_filt$leachate <- fct_relevel(metadata_filt$leachate, "control", "low", "mid", "high")

metadata_filt$hpf <- as.factor(metadata_filt$hpf)
metadata_filt$hpf <- fct_relevel(metadata_filt$hpf, "4", "9", "14")
str(metadata_filt)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples where explanatory variables `hpf` is a factor with 3 levels (4, 9, 14) and `leachate` is a continuous numerical variable representing leachate concentrations with values 0 mg/L, 0.01 mg/L, 0.1 mg/L, and 1 mg/L
:::


# Make DESeq2 object

```{r}

dds <- DESeqDataSetFromMatrix(countData = gcm_filt,
                              colData = metadata_filt,
                              design = formula( ~ 1 + hpf + leachate + hpf:leachate))
```

# Run DESeq Test

This calculates the log2fold change between factors

```{r}
dds <- DESeq(dds, test = "LRT", reduced = ~ hpf + leachate)
```

# Results

> **Likelihood ratio test**
>
> DESeq2 offers two kinds of hypothesis tests: the Wald test, where we use the estimated standard error of a log2 fold change to test if it is equal to zero, and the likelihood ratio test (LRT). The LRT examines two models for the counts, a *full* model with a certain number of terms and a *reduced* model, in which some of the terms of the *full* model are removed. The test determines if the increased likelihood of the data using the extra terms in the *full* model is more than expected if those extra terms are truly zero.
>
> The LRT is therefore useful for testing multiple terms at once, for example testing 3 or more levels of a factor at once, or **all interactions between two variables**. The LRT for count data is conceptually similar to an analysis of variance (ANOVA) calculation in linear regression, except that in the case of the Negative Binomial GLM, we use an analysis of deviance (ANODEV), where the *deviance* captures the difference in likelihood between a full and a reduced model.
>
> The likelihood ratio test can be performed by specifying `test="LRT"` when using the *DESeq* function, and providing a reduced design formula, e.g. one in which a number of terms from `design(dds)` are removed. The degrees of freedom for the test is obtained from the difference between the number of parameters in the two models.

```{r}
dds_or <- DESeq(dds_or, test="LRT", reduced= ~ hpf + leachate)
```

```{r}
data.frame(resultsNames <- resultsNames(dds_or))
```
This compares:

Full model: \~ hpf + leachate + hpf:leachate

Reduced model: \~ hpf + leachate

So it asks: Does including the interaction explain significantly more variation? If yes, then some genes respond to leachate differently at different hpf stages.

## Plot dispersion

```{r}
plotDispEsts(dds)
```

::: callout-note
Dispersion trend line is hovering around 0.1, and bulk spread of tag-wise gene estimates is from 0.02 to 0.5
:::

# Build contrasts

```{r}
resultsNames(dds_or)  # see what contrast names are available
```

| Coef | Name                         | Meaning                                                               |
|------|------------------------------|-----------------------------------------------------------------------|
| 1    | `"Intercept"`                | Baseline expression: `hpf = 4` and `leachate = control`               |
| 2    | `"hpf_9_vs_4"`               | Change from 4 hpf → 9 hpf, at **control** exposure                    |
| 3    | `"hpf_14_vs_4"`              | Change from 4 hpf → 14 hpf, at **control** exposure                   |
| 4    | `"leachate_low_vs_control"`  | Change from control → low exposure, at **4 hpf**                      |
| 5    | `"leachate_mid_vs_control"`  | Change from control → mid exposure, at **4 hpf**                      |
| 6    | `"leachate_high_vs_control"` | Change from control → high exposure, at **4 hpf**                     |
| 7    | `"hpf9.leachatelow"`         | **Interaction term**: extra effect of low exposure at 9 hpf vs 4 hpf  |
| 8    | `"hpf14.leachatelow"`        | **Interaction term**: extra effect of low exposure at 14 hpf vs 4 hpf |
| 9    | `"hpf9.leachatemid"`         | **Interaction term**: extra effect of mid exposure at 9 hpf vs 4 hpf  |
| 10   | `"hpf14.leachatemid"`        | **Interaction term**: extra effect of mid exposure at 14 hpf vs 4 hpf |
| 11   | `"hpf9.leachatehigh"`        | **Interaction term**:... of high exposure at 9 hpf vs 4 hp            |
| 12   | `"hpf14.leachatehigh"`       | **Interaction term**:... of high exposure at 14 hpf vs 4 hpf          |

Contrast summary table \| What you want \| Contrast \| \| ------------------------- \| ----------------------------------------------------- \| \| High vs Control at 4 hpf \| `"leachate_high_vs_control"` \| \| Mid vs Control at 4 hpf \| `"leachate_mid_vs_control"` \| \| Low vs Control at 4 hpf \| `"leachate_low_vs_control"` \| \| High vs Control at 9 hpf \| `c("leachate_high_vs_control", "hpf9.leachatehigh")` \| \| Mid vs Control at 9 hpf \| `c("leachate_mid_vs_control", "hpf9.leachatemid")` \| \| Low vs Control at 9 hpf \| `c("leachate_low_vs_control", "hpf9.leachatelow")` \| \| High vs Control at 14 hpf \| `c("leachate_high_vs_control", "hpf14.leachatehigh")` \| \| Mid vs Control at 14 hpf \| `c("leachate_mid_vs_control", "hpf14.leachatemid")` \| \| Low vs Control at 14 hpf \| `c("leachate_low_vs_control", "hpf14.leachatelow")` \|

# Effect of high exposure at 14 hpf

When you test for main effects of leachate (e.g., "leachate_high_vs_control"), it's at the reference timepoint (hpf = 4). The interaction terms model how the leachate effect changes relative to that baseline as development progresses.

What's the effect of high exposure at 14 hpf? You must combine: - "leachate_high_vs_control": main leachate effect at 4 hpf - "hpf14.leachatehigh": additional effect at 14 hpf

```{r}
# output results of contrast
res_14hpf_high <- (results(dds, contrast = list(
  c("leachate_high_vs_control", "hpf14.leachatehigh")
)))

summary(res_14hpf_high)
```

### Extract significant genes

```{r}
padj.cutoff <- 0.05
lfc.cutoff <- 0.5
```

```{r}
# Convert results table into tibble
res_14hpf_high_tb <- res_14hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_high_noLFCcutoff <- res_14hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_high <- sig_14hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at high exposure level:", nrow(sig_14hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at high exposure level:", nrow(sig_14hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_high, "../output/09_deg_multifactorial/sig_14hpf_high.csv")
write_csv(sig_14hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_14hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf14.leachatehigh")), type = "ashr")

plotMA(res_14hpf_high)
plotMA(res_14hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_high,
    lab = rownames(res_14hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14high <- 
  ggplot(res_14hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_high[sig_14hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_high[sig_14hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14high
```

## Effect of mid exposure at 14 hpf

```{r}
# output results of contrast
res_14hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf14.leachatemid")
)))

summary(res_14hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_14hpf_mid_tb <- res_14hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_mid_noLFCcutoff <- res_14hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_mid <- sig_14hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at mid exposure level:", nrow(sig_14hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at mid exposure level:", nrow(sig_14hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_mid, "../output/09_deg_multifactorial/sig_14hpf_mid.csv")
write_csv(sig_14hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}

# shrink lfc for visualizations
res_14hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf14.leachatemid")), type = "ashr")

plotMA(res_14hpf_mid)
plotMA(res_14hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_mid,
    lab = rownames(res_14hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14mid <- 
  ggplot(res_14hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_mid[sig_14hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_mid[sig_14hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14mid
```

## Effect of low exposure at 14 hpf

```{r}
# output results of contrast
res_14hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf14.leachatelow")
)))

summary(res_14hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_14hpf_low_tb <- res_14hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_low_noLFCcutoff <- res_14hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_low <- sig_14hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at low exposure level:", nrow(sig_14hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at low exposure level:", nrow(sig_14hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_low, "../output/09_deg_multifactorial/sig_14hpf_low.csv")
write_csv(sig_14hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_14hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf14.leachatelow")), type = "ashr")

plotMA(res_14hpf_low)
plotMA(res_14hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_low,
    lab = rownames(res_14hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14low <- 
  ggplot(res_14hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_low[sig_14hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_low[sig_14hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14low
```

## Effect of high exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_high <- (results(dds, contrast = list(
  c("leachate_high_vs_control", "hpf9.leachatehigh")
)))

summary(res_9hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_high_tb <- res_9hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_high_noLFCcutoff <- res_9hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_high <- sig_9hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_high, "../output/09_deg_multifactorial/sig_9hpf_high.csv")
write_csv(sig_9hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf9.leachatehigh")), type = "ashr")

plotMA(res_9hpf_high)
plotMA(res_9hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_high,
    lab = rownames(res_9hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9high <- 
  ggplot(res_9hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9high
```

## Effect of mid exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf9.leachatemid")
)))

summary(res_9hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_mid_tb <- res_9hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_mid_noLFCcutoff <- res_9hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_mid <- sig_9hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_mid, "../output/09_deg_multifactorial/sig_9hpf_mid.csv")
write_csv(sig_9hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf9.leachatemid")), type = "ashr")

plotMA(res_9hpf_mid)
plotMA(res_9hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_mid,
    lab = rownames(res_9hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9mid <- 
  ggplot(res_9hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9mid
```

## Effect of low exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf9.leachatelow")
)))

summary(res_9hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_low_tb <- res_9hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_low_noLFCcutoff <- res_9hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_low <- sig_9hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_low, "../output/09_deg_multifactorial/sig_9hpf_low.csv")
write_csv(sig_9hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf9.leachatelow")), type = "ashr")

plotMA(res_9hpf_low)
plotMA(res_9hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_low,
    lab = rownames(res_9hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9low <- 
  ggplot(res_9hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9low
```

## Effect of high exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_high <- (results(dds, name = "leachate_high_vs_control")
)

summary(res_4hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_high_tb <- res_4hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_high_noLFCcutoff <- res_4hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_high <- sig_4hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_high, "../output/09_deg_multifactorial/sig_4hpf_high.csv")
write_csv(sig_4hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_high_ashr <- lfcShrink(dds, coef = 6 , type = "ashr")

plotMA(res_4hpf_high)
plotMA(res_4hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_high,
    lab = rownames(res_4hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4high <- 
  ggplot(res_4hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4high
```

## Effect of mid exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_mid <- (results(dds, name = "leachate_mid_vs_control")
)

summary(res_4hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_mid_tb <- res_4hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_mid_noLFCcutoff <- res_4hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_mid <- sig_4hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_mid, "../output/09_deg_multifactorial/sig_4hpf_mid.csv")
write_csv(sig_4hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_mid_ashr <- lfcShrink(dds, coef = 5 , type = "ashr")

plotMA(res_4hpf_mid)
plotMA(res_4hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_mid,
    lab = rownames(res_4hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4mid <- 
  ggplot(res_4hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4mid
```

## Effect of low exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_low <- (results(dds, name = "leachate_low_vs_control", alpha=0.05)
)

summary(res_4hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_low_tb <- res_4hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_low_noLFCcutoff <- res_4hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_low <- sig_4hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_low, "../output/09_deg_multifactorial/sig_4hpf_low.csv")
write_csv(sig_4hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_low_ashr <- lfcShrink(dds, coef = 4 , type = "ashr")
plotMA(res_4hpf_low)
plotMA(res_4hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_low,
    lab = rownames(res_4hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4low <- 
  ggplot(res_4hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4low
```

# Summary

## Facet volcano plot

```{r}
# First, strip labels from all individual ggplot objects
# (You can do this in a loop if the plots are in a list)

remove_labels <- function(p) {
  p +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 8)  # adjust title size here
    )
}

# Example: if your plots are named v4control, v4low, etc.
plots <- list( 
   v4low, v9low, v14low,
   v4mid, v9mid, v14mid,
   v4high, v9high, v14high
)

plots_clean <- lapply(plots, remove_labels)

# Now plot in 3 rows × 3 columns
volcano_facet <- plot_grid(
  plotlist = plots_clean,
  nrow = 3,
  ncol = 3,
  labels = NULL,
  label_size = 0,
  align = "hv"
)

volcano_facet
```

```{r}
ggsave("../output/09_deg_multifactorial/volcano_facet.png", volcano_facet, width = 16, height = 12, dpi = 300)
```

::: callout-warning
Why do volcano plots not look like volcanos!?

Batch Effects or Low Biological Signal If your experimental design has confounding factors (e.g., batch effects), or the leachate treatment has low impact, then few genes will show strong changes.

Extreme p-values Your plot shows points with -log10(padj) values above 20--30. That means padj is very close to 0, which could indicate: - Very large sample size - Very consistent effect - Or something off with the dispersion estimate or p-value adjustment (e.g., if some genes had zero variance across groups)
:::

# Next steps

Re-run the model without the interaction term. The likelihood ratio test (LRT) shows that the interaction (hpf:leachate) is not significant! We can now test main effects of leachate across stages (i.e., an overall leachate effect) using `~ hpf + leachate`

## Exlpore cooks distance

> DESeq2 relies on the negative binomial distribution to make estimates and perform statistical inference on differences. While the negative binomial is versatile in having a mean and dispersion parameter, extreme counts in individual samples might not fit well to the negative binomial. For this reason, we perform automatic detection of count outliers. We use Cook's distance, which is a measure of how much the fitted coefficients would change if an individual sample were removed (Cook 1977). For more on the implementation of Cook's distance see the manual page for the *results* function. Below we plot the maximum value of Cook's distance for each row over the rank of the test statistic to justify its use as a filtering criterion.

## Explore RUVSeq
