---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the **interactions** between embryonic stage and PVC leachate in *Montipora capitata* using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

Here we are looking at a multifactorial differential gene expression analysis for *Montipora capitata* coral embryos at 3 developmental stages (4 hours post fertilization = cleavage , 9 hours post fertilization = prawnchip, and 14 hours post fertilization = early gastrula). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of ubiquitous non-point source PVC leachate pollution on coral reefs.

## Inputs

-   `gcm_filtor.csv` is the filtered gene count matrix for 61 samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers removed)

## Outputs

-   `output/06_deg/interactions/int_wald_heatmap.png`
-   `volcano_facet_interaction.png`

# Setup

## Install packages

Copy and run this once in the console

```         
# CRAN packages
install.packages(c(
  "tidyverse",
  "viridis",
  "plotly",
  "cowplot",
  "pheatmap",
  "vsn",
  "ashr",
  "remotes"  # required for GitHub installs
))

# GitHub package
remotes::install_github("mdscheuerell/flexoki")

# Bioconductor
if (!"BiocManager" %in% rownames(installed.packages())) {
  install.packages("BiocManager")
}
BiocManager::install(c("DESeq2", "EnhancedVolcano"))
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(viridis)
library(tidyverse)
library(EnhancedVolcano)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
library(vsn)
library(pheatmap)
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r, eval=TRUE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE)
head(gcm_filtor)
```

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filtor` is the gene count matrix that has been filtered down to only contain rows where counts are at least 15 in 75% of the samples and 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA are already removed
:::

## Import metadata

```{r, eval=TRUE}
metadata_or <- read_csv("../../metadata/metadata_or.csv")
str(metadata_or)
```

```{r, eval=TRUE}
# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")

str(metadata_or)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples with explanatory variables: - `stage` is a factor with 3 levels (cleavage, prawnchip, earlygastrula) - `hpf` is a discrete numerical variable representing the hours post fertilization 4, 9, and 14 (corresponding to `stage`)

-   `leachate` is a factor with 4 levels (control, low, mid, high)
-   `leachate_mgL` is a continuous numerical variable representing leachate concentrations in mg/L with values 0 , 0.01 , 0.1 , and 1 (corresponding to `leachate`)
:::

# LRT

This compares:

Full model: \~ stage + leachate + stage:leachate

Reduced model: \~ stage + leachate

This tells you: Is there any interaction effect between stage and leachate?

But it won't tell you which interaction is significant, or in which direction --- you need Wald contrasts for that.

## Continuous vs categorical

Embryonic stage should be treated as a categorical variable because DESeq2 will estimate expression differences between stages, rather than assuming a linear progression from 4 → 14 hpf. It will fit different intercepts for each stage (i.e., different baseline expression levels). If you encode stage as numeric, the model would assume a linear change in expression per hour of development, which is biologically risky unless you expect linearity (note: we do not).

Whether leachate is treated as a factor or a numeric (continuous) variable radically changes the interpretation of both the main effect and the interaction. If a variable is numeric, DESeq2 treats it as a continuous covariate. This fits a regression slope, estimating the change in gene expression per unit increase in the variable.

Coding `leachate` as a categorical factor allows each stage to have its own independent leachate effect. ✅ Use this if you expect nonlinear, stage-specific effects of exposure.

```{r, eval=TRUE}

dds_cat <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ stage + leachate + stage:leachate))
```

Coding `leachate_mgL` as a continuous numeric tests for a change in expression per unit increase in leachate, and the interaction term tests whether the slope of the dose-response differs depending on stage. ✅ Use this if you expect a linear or dose-dependent trend, and want to ask if the strength or direction of the trend varies by stage.

```{r, eval=TRUE}
dds_con <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ stage + leachate_mgL + stage:leachate_mgL))
```

## Run DESeq LRT Tests

> Does modeling leachate as a factor vs. a continuous dose better explain gene expression changes across stages?

```{r, eval=TRUE}
dds_cat <- DESeq(dds_cat, test = "LRT", reduced = ~ stage + leachate)
dds_con <- DESeq(dds_con, test = "LRT", reduced = ~ stage + leachate_mgL)
```

| Model         | Full model                                  | Reduced model          | LRT tests...                                               |
|------------------|------------------|------------------|-------------------|
| `dds_cat_LRT` | `stage + leachate + stage:leachate`         | `stage + leachate`     | Are there **stage-specific effects of leachate (factor)**? |
| `dds_con_LRT` | `stage + leachate_mgL + stage:leachate_mgL` | `stage + leachate_mgL` | Are there **stage-specific trends** in dose response?      |

## Results

Plot dispersion

```{r}
plotDispEsts(dds_cat)
plotDispEsts(dds_con)
```

::: callout-note
Categorical model has less dispersion than the continuous model! Pay attention to the scale of the Y axis when comparing them.
:::

View summary

```{r, eval=TRUE}
res_cat <- results(dds_cat, alpha = 0.05)
summary(res_cat)
```

```{r, eval=TRUE}
res_con <- results(dds_con, alpha = 0.05)
summary(res_con)
```

```{r, eval=TRUE}
sig_cat <- rownames(results(dds_cat, alpha = 0.05)[which(results(dds_cat)$padj < 0.05), ])
sig_con <- rownames(results(dds_con, alpha = 0.05)[which(results(dds_con)$padj < 0.05), ])
length(intersect(sig_cat, sig_con))
```

::: callout-note
2 out of 44 genes have a linear relationship (compared to a non-linear one!) as we look at the interactions between leachate and stage. All 44 of these genes respond to leachate differently depending on stage.

44 genes show significant stage-by-leachate interaction when leachate is treated categorically → meaning the response to leachate varies by stage, but not necessarily in a linear/dose-dependent way.

2 genes show significant stage-specific dose-response effects when leachate is treated as continuous → they show a linear or monotonic trend across leachate concentrations that varies by stage.

The 2 overlapping genes are well-modeled by a linear trend in leachate (i.e., a slope, not just group differences). The 42 other genes show non-linear or non-monotonic behavior --- meaning their response does not follow a consistent increase/decrease with dose. Instead, they respond differently to each specific leachate level depending on the stage (e.g., they might increase at 0.1 mg/L but decrease at 1 mg/L, or vice versa).
:::

## Save significant genes

```{r}
sig_genes <- res_cat %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write.csv(sig_genes, "../../output/06_deg/interactions/sig_genes.csv")
```

> Of the 44 genes showing significant stage-specific responses to leachate (based on the categorical model), only 2 also show a significant linear dose-response interaction with stage (based on the continuous model). This suggests that while stage modulates leachate response broadly, most genes (42/44) exhibit nonlinear or non-monotonic patterns, rather than a simple linear trend across leachate levels. keep.order = TRUE)

```{r, eval=TRUE}
vsd <- assay(vst(dds_cat))

# make long expression dataframe of significant interaction genes
sig_long <- vsd[sig_cat, ] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "sample_id", values_to = "expression")

# add metadata
sig_long <- sig_long %>% left_join(metadata_or, by = "sample_id")
# add padj
sig_long <- sig_long %>% left_join(sig_genes, by = "gene_id")
```

### Plot top 3 most significant interaction genes


### Loop plots of significant interaction genes

```{r}
# create output directory
dir.create("../../output/06_deg/interaction/gene_plots", recursive = TRUE, showWarnings = FALSE)

for (gene in unique(sig_long$gene_id)) {

  gene_df <- sig_long %>% filter(gene_id == gene)

  p <- ggplot(gene_df, aes(x = leachate, y = expression, color = stage, group = stage)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_smooth(method = "loess", se = FALSE) +
    labs(
      title = paste("Gene:", gene),
      x = "Leachate (mg/L)",
      y = "VST expression"
    ) +
    
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    )
  
  # Render as interactive plotly
  print(plotly::ggplotly(p))

  # Save to PNG
  ggsave(
    filename = paste0("../../output/06_deg/gene_plots/", gene, ".png"),
    plot = p, width = 6, height = 4, dpi = 300
  )
}
```

::: panel-tabset
```{r, eval=TRUE}
for (gene in unique(sig_long$gene_id)) {

  gene_df <- sig_long %>% filter(gene_id == gene)

  cat("### ", gene, "\n\n")  # Quarto tab header

  p <- ggplot(gene_df, aes(x = leachate, y = expression, color = stage, group = stage)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_line(stat = "summary", fun = mean, linewidth = 0.8) +
    labs(
      x = "Leachate (mg/L)",
      y = "Rlog expression"
    ) +
    theme_minimal(base_size = 12)

  print(plotly::ggplotly(p))
}
```
:::

::: callout-note
We have 44 genes with significant interaction effects overall.

Next step: 👉 Use Wald tests to dig into which interactions (e.g., stage × leachate level) are responsible --- gene by gene.
:::

# Wald

> Wald tests on interaction terms assess whether the interaction coefficient **differs significantly from 0** --- i.e., whether the combined effect of that `hpf` and `leachate` level deviates from what would be expected by the additive effects alone.

| What You Want to Know                                 | Use                           | Explanation                                     |
|----------------------------|------------------|--------------------------|
| Any interaction overall?                              | LRT                           | `reduced = ~ stage + leachate`                  |
| Specific interaction term significant?                | Wald                          | `results(dds, name = "stageX.leachateY")`       |
| Total effect of leachate at specific stage vs control | Wald with combined contrast   | `c("leachateY_vs_control", "stageX.leachateY")` |
| Difference in leachate effect between two stages      | Wald contrast of interactions | `list("stageX.leachateY", "stageZ.leachateY")`  |

```{r}
dds <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ stage + leachate + stage:leachate))
```

```{r}
dds_wald <- DESeq(dds, test = "Wald")
```

## Interaction terms

```{r}
resultsNames(dds_wald)
```

-   "stageprawnchip.leachatelow"

-   "stageearlygastrula.leachatelow"

-   "stageprawnchip.leachatemid"

-   "stageearlygastrula.leachatemid"

-   "stageprawnchip.leachatehigh"

-   "stageearlygastrula.leachatehigh"

> Each of the above interaction terms tells us the extra effect of leachate at this stage (vs the additive sum of the stage and leachate effects).

### Extract Wald results for all interaction terms

```{r}
# Get the interaction Wald results
res_high_pc <- results(dds_wald, name = "stageprawnchip.leachatehigh")
res_mid_pc  <- results(dds_wald, name = "stageprawnchip.leachatemid")
res_low_pc  <- results(dds_wald, name = "stageprawnchip.leachatelow")
res_high_eg <- results(dds_wald, name = "stageearlygastrula.leachatehigh")
res_mid_eg  <- results(dds_wald, name = "stageearlygastrula.leachatemid")
res_low_eg  <- results(dds_wald, name = "stageearlygastrula.leachatelow")

# Combine into a summary
int_wald_df <- data.frame(
  gene_id = rownames(res_high_pc),
  prawnchip_high_log2FC = res_high_pc$log2FoldChange,
  prawnchip_high_padj = res_high_pc$padj,
  prawnchip_mid_log2FC = res_mid_pc$log2FoldChange,
  prawnchip_mid_padj = res_mid_pc$padj,
  prawnchip_low_log2FC = res_low_pc$log2FoldChange,
  prawnchip_low_padj = res_low_pc$padj,
  earlygastrula_high_log2FC = res_high_eg$log2FoldChange,
  earlygastrula_high_padj = res_high_eg$padj,
  earlygastrula_mid_log2FC = res_mid_eg$log2FoldChange,
  earlygastrula_mid_padj = res_mid_eg$padj,
  earlygastrula_low_log2FC = res_low_eg$log2FoldChange,
  earlygastrula_low_padj = res_low_eg$padj
)

# Focus on the 44 genes
int_wald_df_sig <- int_wald_df[int_wald_df$gene_id %in% sig_genes$gene_id, ]
```

```{r}
library(pheatmap)

# Step 1: Order and extract log2FC values
log2fc_cols <- c(
  "prawnchip_low_log2FC", "prawnchip_mid_log2FC", "prawnchip_high_log2FC",
  "earlygastrula_low_log2FC", "earlygastrula_mid_log2FC", "earlygastrula_high_log2FC"
)

mat <- as.matrix(int_wald_df_sig[, log2fc_cols])
rownames(mat) <- int_wald_df_sig$gene_id

# Step 2: Rename columns for clarity
colnames(mat) <- c(
  "Prawnchip: Low", "Prawnchip: Mid", "Prawnchip: High",
  "Early Gastrula: Low", "Early Gastrula: Mid", "Early Gastrula: High"
)

# Step 3: Create annotation for stage/leachate
annotation_col <- data.frame(
  Stage = rep(c("Prawnchip", "Early Gastrula"), each = 3),
  Leachate = rep(c("Low", "Mid", "High"), times = 2)
)
rownames(annotation_col) <- colnames(mat)

# Step 4: Prepare significance (padj < 0.05) annotation matrix
padj_cols <- c(
  "prawnchip_low_padj", "prawnchip_mid_padj", "prawnchip_high_padj",
  "earlygastrula_low_padj", "earlygastrula_mid_padj", "earlygastrula_high_padj"
)
padj_mat <- as.matrix(int_wald_df_sig[, padj_cols])
rownames(padj_mat) <- int_wald_df_sig$gene_id
colnames(padj_mat) <- colnames(mat)

# Significance stars with NA handling
sig_labels <- ifelse(is.na(padj_mat), "",
                      ifelse(padj_mat < 0.01, "**",
                        ifelse(padj_mat < 0.05, "*",
                          ifelse(padj_mat < 0.1, ".", ""))))

# Step 5: Save heatmap with significance stars
png("../../output/06_deg/interactions/int_wald_heatmap.png", width = 10, height = 12, units = "in", res = 600)

pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         display_numbers = sig_labels,
         number_color = "black",
         fontsize = 10,
         fontsize_number = 8,
         angle_col = 45,
         border_color = NA)

dev.off()
```

```{r}
# Print the heatmap inline
pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         display_numbers = sig_labels,
         number_color = "black",
         fontsize = 10,
         fontsize_number = 8,
         angle_col = 45,
         border_color = NA)
```

> 🧬 What does the heatmap show? Each cell in the heatmap shows the scaled log2 fold change for a gene in a specific interaction condition.
>
> ✅ Scaling Because you set scale = "row", each gene's values are z-scored across the six interaction conditions
>
> Blue = low relative expression change in that condition
>
> Red = high relative expression change in that condition
>
> The actual color scale is relative within each row, not across the whole matrix.
>
> This helps visualize patterns of expression changes across conditions for each gene, regardless of absolute size.
>
> 📊 What can you learn from this? Are there groups of genes responding similarly? Row clustering may reveal gene modules with similar interaction response profiles.
>
> Which stage + leachate condition drives the interaction? Column patterns (e.g., lots of red in prawnchip_high_log2FC) suggest strong interaction effects in that condition.
>
> Which genes have the strongest differential interactions? Genes with high contrast (strong red/blue) have more dynamic interaction effects.
>
> 🧠 Interpretation Reminder These log2 fold changes are interaction effects --- they reflect how gene expression at a specific `stage` & `leachate` level deviates from the expected additive effect of `stage` and `leachate` alone.
>
> So, a large positive log2FC means:
>
> "At this `stage` and `leachate` level, this gene is more highly expressed than we would expect based on `stage` and `leachate` effects individually."
>
> Likewise, a large negative value = stronger-than-expected suppression.

```{r}
int_wald_df_sig$any_sig <- apply(int_wald_df_sig[, grepl("padj", names(int_wald_df_sig))], 1, function(p) any(p < 0.05, na.rm = TRUE))

head(int_wald_df_sig)
```

