---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

Here we are looking at a multifactorial differential gene expression analysis for *Montipora capitata* coral embryos at 3 developmental stages (4 hours post fertilization = cleavage , 9 hours post fertilization = prawnchip, and 14 hours post fertilization = early gastrula). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of ubiquitous non-point source PVC leachate pollution on coral reefs.

https://chatgpt.com/share/68ed79f1-e768-8008-9caa-0df1ca68b4ef

## Inputs

-   `gcm_filtor.csv` is the filtered gene count matrix (genes were kept in which 85% of the samples had a minimum count of 15) for 61 samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers removed)

## Outputs

-   `output/06_deg/resul/int_wald_heatmap.png`
-   `volcano_facet_interaction.png`

# Setup

```{r}
output_path <- "../../output/06_deg/spawnnight/"
metadata_path <- "../../metadata/metadata_or.csv"
```

## Install packages

```{r, eval=FALSE}
# CRAN packages
#install.packages(c(
#  "tidyverse",
#  "viridis",
#  "plotly",
#  "cowplot",
#  "pheatmap",
#  "vsn",
#  "ashr",
#  "remotes"  # required for GitHub installs
#))

#install.packages("ComplexUpset")    # ggplot2-based, prettier
# GitHub package
#remotes::install_github("mdscheuerell/flexoki")

# Bioconductor
#if (!"BiocManager" %in% rownames(installed.packages())) {
#  install.packages("BiocManager")
#}
#BiocManager::install(c("DESeq2", "DEGreport"))
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(viridis)
library(tidyverse)
library(DEGreport)
library(ashr)
library(flexoki)
library(plotly)
library(vsn)
library(pheatmap)
library(ComplexUpset)
library(ggplot2)
library(ggplotify)
library(patchwork)   # for wrap_plots(...)
library(cowplot)     # for plot_grid & get_legend
library(dplyr)
library(grid) 
library(purrr)
library(forcats)
library(ggforce)   # for facet_wrap_paginate
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a sequenced RNA library (i.e. a sample), and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r, eval=TRUE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE) 
head(gcm_filtor)
```

-   The count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filtor` is the gene count matrix that has been filtered down to only contain rows where counts are at least 15 in 85% of the samples and 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA are already removed. This step was completed in `code/05_explore`
:::

## Import metadata

```{r, eval=TRUE}
# Load metadata (outlier samples removed, 61 samples remain)
metadata_or <- read_csv(metadata_path)

# set factors
metadata_or <- metadata_or %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula")),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high")),
    spawn_night = factor(spawn_night, levels  = c("July_6th", "July_7th", "July_8th"))
    )

# View metadata structure
str(metadata_or)
```

::: callout-note
`metadata_or` contains metadata for 61 samples (*o*utliers *r*emoved) with explanatory variables: - `stage` is a factor with 3 levels (cleavage, prawnchip, earlygastrula) - `hpf` is a numerical variable representing the hours post fertilization 4, 9, and 14 (corresponding to `stage`)

-   `leachate` is a factor with 4 levels (control, low, mid, high)

-   `leachate_mgL` is a continuous numerical variable representing leachate concentrations in mg/L with values 0 , 0.01 , 0.1 , and 1 (corresponding to `leachate`)

-   `collection_date` is a date-formatted **"%d-%b-%Y"** column with the date of the night embryos were spawned

-   `spawn_night` is a factor character string of the spawn night date (either July 6th, July 7th, or July 8th)
:::

# LRT

> [Likelihood ratio test](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#likelihood-ratio-test:~:text=seq%20workflow.-,Likelihood%20ratio%20test,-DESeq2%20offers%20two): DESeq2 offers two kinds of hypothesis tests: the Wald test, where we use the estimated standard error of a log2 fold change to test if it is equal to zero, and the likelihood ratio test (LRT). The LRT examines two models for the counts, a full model with a certain number of terms and a reduced model, in which some of the terms of the full model are removed. The test determines if the increased likelihood of the data using the extra terms in the full model is more than expected if those extra terms are truly zero. The LRT is therefore useful for testing multiple terms at once, for example testing 3 or more levels of a factor at once, or all interactions between two variables. The LRT for count data is conceptually similar to an analysis of variance (ANOVA) calculation in linear regression, except that in the case of the Negative Binomial GLM, we use an analysis of deviance (ANODEV), where the deviance captures the difference in likelihood between a full and a reduced model. The likelihood ratio test can be performed by specifying test="LRT" when using the DESeq function, and providing a reduced design formula, e.g. one in which a number of terms from design(dds) are removed. The degrees of freedom for the test is obtained from the difference between the number of parameters in the two models.

> For a control and treatment time series, one can use a design formula containing the condition factor, the time factor, and the interaction of the two. *In this case, using the likelihood ratio test with a reduced model which does not contain the interaction terms will test whether the condition induces a change in gene expression at any time point after the reference level time point (time 0). An example of the later analysis is provided in our [RNA-seq workflow](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html)*.

We run a few LRT tests :

-   full model: `~ spawn_night + stage + leachate + stage:leachate`

-   reduced model: `~ spawn_night + stage + leachate` (effect of interaction) **INT**
-   reduced model: `~ spawn_night + stage ` (effects of leachate and stage specific leachate together) **JOINT**

& 

-   full model: `~ spawn_night + stage + leachate`

-   reduced model: `~ spawn_night + stage ` (tests for main effect of leachate) **MAIN**

This tells you if there are any interaction effects between stage and leachate, but it won't tell you which interaction is significant, or in which direction --- It helps you decide which model to use. If there aren't any differences in significant gene results between models, use the simpler model.

::: callout-caution
Order matters! The first term (`stage`) is what the model controls for, and the second term (`leachate`) is the main effect. Meaning the model adjusts for stage first and then assesses the leachate effect.
:::


Stage & Leachate will be treated as categorical variables

When embryonic `stage` is treated as a categorical variable DESeq2 will estimate expression differences between stages, rather than assuming a linear progression from 4 → 14 hpf. DESeq2 will fit different intercepts for each stage (i.e., different baseline expression levels). If you encode stage as a continuous numeric, the model would assume a linear change in expression per hour of development, which is biologically risky unless you expect linearity (note: we do not).

::: callout-note
`stage` will be treated as a categorical variable
:::

Whether `leachate` is treated as a factor or a numeric (continuous) variable radically changes the interpretation of both the main effect and the interaction. If a variable is numeric, DESeq2 treats it as a continuous covariate. This fits a regression slope, estimating the change in gene expression per unit increase in the variable. Coding `leachate` as a categorical factor allows each stage to have its own independent leachate effect. ✅ Use this if you expect nonlinear, stage-specific effects of exposure (note: we do).

# Make DESeqDataSet

```{r, eval=TRUE}

dds <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ spawn_night + stage + leachate + stage:leachate))
```

dds \<- DESeqDataSetFromMatrix(countData = gcm_filtor,

colData = metadata_or,

design = formula( \~ spawn_night + stage + leachate + stage:leachate))

## Run DESeq LRT Tests

### Spawn night 

Here we treat spawn night as a covariate as a way to control for it as a nuisance factor.

```{r}
dds_sp_LRT <- DESeq(dds, test = "LRT", reduced = ~ stage + leachate + stage:leachate)
```

| Model name     | Reduced model formula                 | What it tests for                         | Interpretation of significant genes                                                                                             |
|----------------|---------------------------------------|-------------------------------------------|---------------------------------------------------------------------------------------------------------------------------------|
| **dds_sp_LRT** | `~ stage + leachate + stage:leachate` | Tests the contribution of **spawn_night** | Genes whose expression significantly varies **among spawn nights**, after accounting for stage, leachate, and their interaction |

```{r}
res_spawn <- results(dds_sp_LRT, alpha = 0.05)
summary(res_spawn)
```

So this test asks:

"Does including spawn_night explain additional variation in gene expression, after accounting for stage and leachate (and their interaction)?"

The fact that you found \~3,551 genes (≈29%) significant means that spawn_night contributes meaningfully to expression variability. That's strong evidence that it is a real batch or parental effect --- not random noise. You're not interested in biological differences among spawn nights; you just want to adjust for them so they don't confound your leachate comparisons. You include spawn_night in the design formula to control for it statistically. You do not test it (no LRT against it; no results table for it). You interpret the coefficients (or LRT results) for leachate, stage, and stage:leachate --- these now represent effects after adjusting for spawn_night.

### Interaction term 

```{r}
dds_int_LRT <- DESeq(dds, test = "LRT", reduced = ~ spawn_night + stage + leachate)
```

| Model name      | Reduced model formula              | What it tests for                                              | Interpretation of significant genes                                                               |
|-----------------|------------------------------------|----------------------------------------------------------------|---------------------------------------------------------------------------------------------------|
| **dds_int_LRT** | `~ spawn_night + stage + leachate` | Tests the contribution of the **stage × leachate interaction** | Genes whose expression patterns differ across **leachate treatments in a stage-dependent manner** |

```{r}
res_int <- results(dds_int_LRT, alpha = 0.05)
summary(res_int)
```

Out of 12,221 genes tested:

-    Only 13 genes (≈ 0.1 %) were significant (FDR \< 0.05).
-   6 up-regulated, 7 down-regulated.
-   Tests **interaction only** (`stage:leachate`).
-   You got **13 DEGs** ⇒ This subset show stage-specific responses.

For nearly all genes, the relationship between leachate level and expression does not change significantly across stages --- the leachate effect is fairly consistent through development. So the interaction term adds very little explanatory power after controlling for spawn night and main effects.

This implies `leachate` affects gene expression, but the direction/magnitude of that effect doesn't vary much among embryonic stages (4, 9, 14 hpf).

If you were expecting strong stage-specific responses, this result suggests that such effects are **minimal** or **limited to a small subset of genes**.

### Main leachate effect

```{r}
dds_main <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ spawn_night + stage + leachate))
```

```{r, eval=TRUE}
dds_main_LRT <- DESeq(dds_main, test = "LRT", reduced = ~ spawn_night + stage)
```

```{r}
res_main <- results(dds_main_LRT, alpha = 0.05)
summary(res_main)
```

### Joint 

```{r}
dds_joint_LRT <- DESeq(dds, test = "LRT", reduced = ~ spawn_night + stage)
```

```{r}
res_joint <- results(dds_joint_LRT, alpha = 0.05)
summary(res_joint)
```

→ Tests the joint contribution of all leachate terms (main effect and stage:leachate).

→ We got 43 DEGs ⇒ there is leachate-associated signal somewhere (main and/or interaction taken together).

This indicates some genes (30) have **weak main + weak interaction** effects that don\'t cross FDR alone but **do** when evaluated together.

::: callout-important
In summary: results are consistent with stage-specific leachate responses (interaction), not a uniform, stage-averaged leachate shift (main effect). Hence: 0 (main-only), 13 (interaction-only), 43 (any leachate signal jointly) is coherent.
:::


> When one performs a likelihood ratio test, the *p* values and the test statistic (the `stat` column) are values for the test that removes all of the variables which are present in the full design and not in the reduced design. This tests the null hypothesis that all the coefficients from these variables and levels of these factors are equal to zero.
>
> The likelihood ratio test *p* values therefore represent a test of *all the variables and all the levels of factors* which are among these variables. However, the results table only has space for one column of log fold change, so a single variable and a single comparison is shown (among the potentially multiple log fold changes which were tested in the likelihood ratio test). This is indicated at the top of the results table with the text, e.g., log2 fold change (MLE): condition C vs A, followed by, LRT p-value: '\~ batch + condition' vs '\~ batch'. This indicates that the *p* value is for the likelihood ratio test of *all the variables and all the levels*, while the log fold change is a single comparison from among those variables and levels.

```{r}
?results
```

> [Building the results table](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#multiple-testing:~:text=5.2-,Building%20the%20results%20table,-Calling%20results%20without): Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level.
>
> Genes with small *p* values from the LRT test are those which at one or more time points after time 0 showed a strain-specific effect. Note therefore that this will not give small *p* values to genes that moved up or down over time in the same way in both strains.


```{r}
sig_int <- res_int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

sig_joint <- res_joint%>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

overlap_ids <- intersect(sig_int$gene_id, sig_joint$gene_id)
length(overlap_ids)        # how many
overlap_ids                # which ones
```

::: callout-note
The 13 genes found to have stage specific effects (interaction term) are also present in the joint list that catches subtle effects of both interaction and main leachate. We have 43 genes with significant leachate effects overall. This is subtle (43 out of ~12K genes may be p-hacked...) but significant based on adjusted pvalues all less than 0.05.
:::


> The 43 genes showing significant stage-specific responses to leachate exhibit nonlinear or non-monotonic patterns, rather than a simple linear trend across leachate levels.

# Wald

> Wald tests assess whether the coefficient **differs significantly from 0**


Run the Wald test on the dds object.

```{r}
dds <- DESeq(dds, test = "Wald")
```

```{r}
resultsNames(dds)
```

## Results

**We're focusing only on leachate effects in this analysis** 

### Interaction effects

-   "stageprawnchip.leachatelow"

-   "stageearlygastrula.leachatelow"

-   "stageprawnchip.leachatemid"

-   "stageearlygastrula.leachatemid"

-   "stageprawnchip.leachatehigh"

-   "stageearlygastrula.leachatehigh"

Each of the above interaction terms tells us the extra effect of the leachate level at a stage (vs the additive sum of the stage and leachate effects) in comparison to a control baseline count. For the interaction terms, we are trying to get the "difference of the differences". 

```{r}
# prawnchip interaction terms
res_prawnchip_high.int <- results(dds, name = "stageprawnchip.leachatehigh", alpha = 0.05)
res_prawnchip_mid.int <- results(dds, name = "stageprawnchip.leachatemid", alpha = 0.05)
res_prawnchip_low.int  <- results(dds, name = "stageprawnchip.leachatelow", alpha = 0.05)

# early gastrula interaction terms
res_earlygastrula_high.int <- results(dds, name = "stageearlygastrula.leachatehigh", alpha = 0.05)
res_earlygastrula_mid.int <- results(dds, name = "stageearlygastrula.leachatemid", alpha = 0.05)
res_earlygastrula_low.int <- results(dds, name = "stageearlygastrula.leachatelow", alpha = 0.05)
```

#### Prawnchip

##### high

```{r}
summary(res_prawnchip_high.int)
```

::: callout-note
0 genes showed a significant stage × leachate interaction in the prawnchip stage under the high treatment (Wald test, BH-adjusted p \< 0.05). 
:::


::: column-margin
> DESeq2 uses size-factor normalization (the "median-ratio method"). It corrects for library size and composition so you can compare counts across samples. How it works (median-ratio method) Per-gene geometric mean across samples (skip genes with all zeros). Per-sample ratios: for each gene, divide a sample's count by that gene's geometric mean. Size factor: the median of those ratios within each sample. Normalized counts: normalized = raw_count / size_factor. These size factors are used as an offset in the negative binomial GLM for the Wald/LRT tests. Normalization is independent of the design formula.
:::

##### mid

```{r}
summary(res_prawnchip_mid.int)
```

::: callout-note
0 genes were found to have significant prawnchip stage x mid leachate interaction (adjusted p value \< 0.05)
:::

##### low

```{r}
summary(res_prawnchip_low.int)
```

::: callout-note
7 genes were found to have significant prawnchip stage x low leachate interaction (adjusted p value \< 0.05)
:::

```{r}
sig_prawnchip_low.int <- res_prawnchip_low.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_prawnchip_low.int, file.path(output_path, "sig_prawnchip_low.int.csv"))
```

#### Early gastrula

##### high

```{r}
summary(res_earlygastrula_high.int)
```

::: callout-note
1 gene was found to have a significant interaction effect between early gastrula and high leachate
:::

```{r}
sig_earlygastrula_high.int <- res_earlygastrula_high.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_earlygastrula_high.int, file.path(output_path, "sig_earlygastrula_high.int.csv"))
```

##### mid

```{r}
summary(res_earlygastrula_mid.int)
```

::: callout-note
0 genes were found to have a significant interaction between early gastrula and mid leachate
:::


##### low

```{r}
summary(res_earlygastrula_low.int)
```

::: callout-note
8 genes were found to have significant interaction between early gastrula and low leachate
:::

```{r}
sig_earlygastrula_low.int <- res_earlygastrula_low.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_earlygastrula_low.int, file.path(output_path, "sig_earlygastrula_low.int.csv"))
```

## Contrasts

See [A guide to designs and contrasts in DESeq2](https://www.atakanekiz.com/technical/a-guide-to-designs-and-contrasts-in-DESeq2/) by Dr. Atakan Ekiz

```{r}
resultsNames(dds)
```

```{r}
# cleavage 
res_cleavage_low.vs.control <- results(dds, name = "leachate_low_vs_control", alpha = 0.05)
res_cleavage_mid.vs.control  <- results(dds, name = "leachate_mid_vs_control", alpha = 0.05)
res_cleavage_high.vs.control <- results(dds, name = "leachate_high_vs_control", alpha = 0.05)

# prawnchip
res_prawnchip_low.vs.control <- results(dds, contrast = list(c("leachate_low_vs_control", "stageprawnchip.leachatelow")), alpha = 0.05)
res_prawnchip_mid.vs.control  <- results(dds, contrast = list(c("leachate_mid_vs_control",  "stageprawnchip.leachatemid")), alpha = 0.05)
res_prawnchip_high.vs.control <- results(dds, contrast = list(c("leachate_high_vs_control", "stageprawnchip.leachatehigh")), alpha = 0.05)

# earlygastrula
res_earlygastrula_low.vs.control <- results(dds, contrast = list(c("leachate_low_vs_control", "stageearlygastrula.leachatelow")), alpha = 0.05)
res_earlygastrula_mid.vs.control  <- results(dds, contrast = list(c("leachate_mid_vs_control",  "stageearlygastrula.leachatemid")), alpha = 0.05)
res_earlygastrula_high.vs.control <- results(dds, contrast = list(c("leachate_high_vs_control", "stageearlygastrula.leachatehigh")), alpha = 0.05)

```

### Cleavage

#### high

```{r}
summary(res_cleavage_high.vs.control)
```

::: callout-note
7 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the high leachate treatment compared to the control. 
:::

```{r}
sig_cleavage_high.vs.control <- res_cleavage_high.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_cleavage_high.vs.control, file.path(output_path, "sig_cleavage_high.vs.control.csv"))
```

#### mid

mid leachate vs. control (mvc) for cleavage

```{r}
summary(res_cleavage_mid.vs.control)
```

::: callout-note
58 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the mid leachate treatment compared to the control. 57 LFC \> 0. 57/58 total genes were found in more abundance compared to the controls.
:::

```{r}
sig_cleavage_mid.vs.control <- res_cleavage_mid.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_cleavage_mid.vs.control, file.path(output_path, "sig_cleavage_mid.vs.control.csv"))
```

#### low

for cleavage (baseline stage) \> Because you fit \~ 0 + stage + leachate + stage:leachate, **the coefficient leachatelow is the low vs control effect at the reference stage**. From your resultsNames() (no stagecleavage.leachatelow term is present) the reference stage is cleavage, so Low vs control at cleavage is simply the leachatelow coefficient.

```{r}
summary(res_cleavage_low.vs.control)
```

::: callout-note
34 genes were found to be significantly differentially expressed (adjusted p value \< 0.05) at the cleavage stage in the low leachate treatment compared to the control. 13 LFC \> 0 & 21 LFC \< 0.
:::

```{r}
sig_cleavage_low.vs.control <- res_cleavage_low.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_cleavage_low.vs.control, file.path(output_path, "sig_cleavage_low.vs.control.csv"))
```

### Prawnchip

#### high

for prawnchip

```{r}
summary(res_prawnchip_high.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the high leachate treatment when compared to the control.
:::

#### mid

```{r}
summary(res_prawnchip_mid.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the high leachate treatment when compared to the control.
:::

#### low

```{r}
summary(res_prawnchip_low.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn chip stage in the high leachate treatment when compared to the control.
:::

### Early gastrula

#### high

```{r}
summary(res_earlygastrula_high.vs.control)
```

::: callout-note
27 genes differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the high leachate treatment when compared to the control. All 27 of these genes had more abundant transcripts in treated samples compared to controls.
:::

```{r}
sig_earlygastrula_high.vs.control <- res_earlygastrula_high.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

write_csv(sig_earlygastrula_high.vs.control, file.path(output_path, "sig_earlygastrula_high.vs.control.csv"))
```

#### mid

```{r}
summary(res_earlygastrula_mid.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the mid leachate treatment when compared to the control.
:::

#### low

```{r}
summary(res_earlygastrula_low.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early gastrula stage in the low leachate treatment when compared to the control.
:::

# Summary

So we have 9 different 'significant gene' lists. 
Each list indicates genes in which leachate had a significant effect.

LRT:
- `sig_int`, 13 DEGs
- `sig_joint`. 44 DEGs

Interaction effects 
- `sig_prawnchip_low.int`, 7 DEGs 
- `sig_earlygastrula_high.int`, 1 DEGs 
- `sig_earlygastrula_low.int`, 8 DEGs

Contrasts
- `sig_cleavage_low.vs.control`, 34 DEGs 
- `sig_cleavage_mid.vs.control`, 58 DEGs 
- `sig_cleavage_high.vs.control`, 7 DEGs 
- `sig_earlygastrula_high.vs.control`, 27 DEGs

```{r}
# 1) Put all of your sig data frames into a *named* list
sig_lists <- list(
  int                           = sig_int,
  joint                         = sig_joint,
  prawnchip_low.int             = sig_prawnchip_low.int,
  earlygastrula_high.int        = sig_earlygastrula_high.int,
  earlygastrula_low.int         = sig_earlygastrula_low.int,
  cleavage_low.vs.control       = sig_cleavage_low.vs.control,
  cleavage_mid.vs.control       = sig_cleavage_mid.vs.control,
  cleavage_high.vs.control      = sig_cleavage_high.vs.control,
  earlygastrula_high.vs.control = sig_earlygastrula_high.vs.control
)

# 2) Smoosh: add a 'list' column indicating the source and stack them
sig_summary <- bind_rows(sig_lists, .id = "list") %>%
  relocate(list, gene_id)

write_csv(sig_summary, file.path(output_path, "sig_summary.csv"))
```

```{r}
sig_all <- dplyr::distinct(sig_summary, gene_id)
nrow(sig_all)
write_csv(sig_all, file.path(output_path, "sig_all.csv"))
```
There are 130 total DEGs across LRT and Wald tests where leachate popped up as a significant covariate. 
# Save count matrices

Make variance stabilized count matrix of these 130 genes of interest
```{r}
# VST transform
vsd <- vst(dds, blind = FALSE) # DESeqTransform Object
vst_mat <- assay(vsd) # VST count matrix with genes as rows and samples as columns

# Keep the 130 genes of interest
vst_mat_130 <- vst_mat[sig_all$gene_id, , drop = FALSE]
dim(vst_mat_130)
```

```{r}
# Convert to dataframes and save 
vst_mat_df <- vst_mat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

vst_mat_130_df <- vst_mat_130 %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

write_csv(vst_mat_df, file.path(output_path,"vst_mat_df.csv"))
write_csv(vst_mat_130_df, file.path(output_path,"vst_mat_130_df.csv"))
```

Make library-size adjusted count matrix of these 130 genes of interest
```{r}
norm <- counts(dds, normalized = TRUE) # size factor normalized count matrix with genes as rows and samples as columns

# Keep the 130 genes of interest
norm_130 <- norm[sig_all$gene_id, , drop = FALSE]
dim(norm_130)
```

```{r}
# Convert to dataframes and save 
norm_df <- norm %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

norm_130_df <- norm_130 %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

write_csv(norm_df, file.path(output_path,"norm_df.csv"))
write_csv(norm_130_df, file.path(output_path,"norm_130_df.csv"))
```
Make raw count matrix of these 130 genes of interest
```{r}
raw <- counts(dds) # raw count matrix with genes as rows and samples as columns

# Keep the 130 genes of interest
raw_130 <- raw[sig_all$gene_id, , drop = FALSE]
dim(raw_130)
```

```{r}
# Convert to dataframes and save 
raw_df <- raw %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

raw_130_df <- raw_130 %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

write_csv(raw_df, file.path(output_path,"raw_df.csv"))
write_csv(raw_130_df, file.path(output_path,"raw_130_df.csv"))
```


# Next steps
-   Make a PCA and PERMANOVA of the DEGs ... 
-   I'd simply like an easy absence/presence heatmap that shows all the genes, and which lists they are significant in.
-   I'd then like to show on the heatmap the log2FC, and add astericks for significance levels p\<0.05, p\<0.01, p\<0.001.
-   I'd like gene count plots visually summarizing each list type (LRT, Interaction, Stagexleachate effects) in tiny tidy multiples
-   Do these genes cluster into expression patterns? ... Clustering
-   What do these genes do ? ... GO enrichment
-   Are there correlations between these differentially expressed genes and differentially abundant taxa?
