---
title: "Multifactorial Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  cache = TRUE) # chunk re-runs if inputs change
```

# Background

Here we are looking at a multifactorial differential gene expression
analysis for *Montipora capitata* coral embryos at 3 developmental
stages (4 hours post fertilization = cleavage , 9 hours post
fertilization = prawnchip, and 14 hours post fertilization = early
gastrula). Embryos at each stage were exposed to either a control, low,
mid, or high concentration of PVC leachate. This work aims to understand
sublethal effects of ubiquitous non-point source PVC leachate pollution
on coral reefs.

What we really want is effect of leachate within each stage.

https://chatgpt.com/share/68ed79f1-e768-8008-9caa-0df1ca68b4ef

## Inputs

-   `gcm_filtor.csv` is the filtered gene count matrix (genes were kept
    in which 85% of the samples had a minimum count of 15) for 61
    samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers
    removed)

## Outputs

-   `output/06_deg/resul/int_wald_heatmap.png`
-   `volcano_facet_interaction.png`

# Setup

```{r}
output_path <- "../../output/06_deg/spawnnight/"
metadata_path <- "../../metadata/metadata_or.csv"
```

## Install packages

```{r, eval=FALSE}
# CRAN packages
#install.packages(c(
#  "tidyverse",
#  "viridis",
#  "plotly",
#  "cowplot",
#  "pheatmap",
#  "ashr",
#  "remotes"  # required for GitHub installs
#))

#install.packages("ComplexUpset")    # ggplot2-based, prettier
# GitHub package
#remotes::install_github("mdscheuerell/flexoki")

# Bioconductor
#if (!"BiocManager" %in% rownames(installed.packages())) {
#  install.packages("BiocManager")
#}
#BiocManager::install(c("DESeq2", "DEGreport","vsn"))
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(viridis)
library(tidyverse)
library(DEGreport)
library(ashr)
library(plotly)
library(vsn)
library(pheatmap)
library(ComplexUpset)
library(ggplotify)
library(patchwork)   # for wrap_plots(...)
library(cowplot)     # for plot_grid & get_legend
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a
sequenced RNA library (i.e. a sample), and the values give the estimated
counts of fragments that were assigned to the respective gene in each
library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's
    in the column headers

```{r, eval=TRUE, cache=FALSE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE) 
head(gcm_filtor)
```

-   The count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of
    numbers), the pollution exposure (C, L, M, H), and the embryonic
    stage (4, 9, 14)

::: callout-note
`gcm_filtor` is the gene count matrix that has been filtered down to
only contain rows where counts are at least 15 in 85% of the samples and
2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT
alignment and did not group in the exploratory PCA are already removed.
This step was completed in `code/05_explore`
:::

## Import metadata

```{r, eval=TRUE}
# Load metadata (outlier samples removed, 61 samples remain)
metadata_or <- read_csv(metadata_path)

# set factors
metadata_or <- metadata_or %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula")),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high")),
    spawn_night = factor(spawn_night, levels  = c("July_6th", "July_7th", "July_8th"))
    )

# View metadata structure
str(metadata_or)
```

::: callout-note
`metadata_or` contains metadata for 61 samples (*o*utliers *r*emoved)
with explanatory variables: - `stage` is a factor with 3 levels
(cleavage, prawnchip, earlygastrula) - `hpf` is a numerical variable
representing the hours post fertilization 4, 9, and 14 (corresponding to
`stage`)

-   `leachate` is a factor with 4 levels (control, low, mid, high)

-   `leachate_mgL` is a continuous numerical variable representing
    leachate concentrations in mg/L with values 0 , 0.01 , 0.1 , and 1
    (corresponding to `leachate`)

-   `collection_date` is a date-formatted **"%d-%b-%Y"** column with the
    date of the night embryos were spawned

-   `spawn_night` is a factor character string of the spawn night date
    (either July 6th, July 7th, or July 8th)
:::

# LRT

> [Likelihood ratio
> test](https://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#likelihood-ratio-test:~:text=seq%20workflow.-,Likelihood%20ratio%20test,-DESeq2%20offers%20two):
> DESeq2 offers two kinds of hypothesis tests: the Wald test, where we
> use the estimated standard error of a log2 fold change to test if it
> is equal to zero, and the likelihood ratio test (LRT). The LRT
> examines two models for the counts, a full model with a certain number
> of terms and a reduced model, in which some of the terms of the full
> model are removed. The test determines if the increased likelihood of
> the data using the extra terms in the full model is more than expected
> if those extra terms are truly zero. The LRT is therefore useful for
> testing multiple terms at once, for example testing 3 or more levels
> of a factor at once, or all interactions between two variables. The
> LRT for count data is conceptually similar to an analysis of variance
> (ANOVA) calculation in linear regression, except that in the case of
> the Negative Binomial GLM, we use an analysis of deviance (ANODEV),
> where the deviance captures the difference in likelihood between a
> full and a reduced model. The likelihood ratio test can be performed
> by specifying test="LRT" when using the DESeq function, and providing
> a reduced design formula, e.g. one in which a number of terms from
> design(dds) are removed. The degrees of freedom for the test is
> obtained from the difference between the number of parameters in the
> two models.

> For a control and treatment time series, one can use a design formula
> containing the condition factor, the time factor, and the interaction
> of the two. *In this case, using the likelihood ratio test with a
> reduced model which does not contain the interaction terms will test
> whether the condition induces a change in gene expression at any time
> point after the reference level time point (time 0). An example of the
> later analysis is provided in our [RNA-seq
> workflow](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html)*.

We run a few LRT tests :

-   full model: `~ spawn_night + stage + leachate + stage:leachate`

-   reduced model: `~ spawn_night + stage + leachate` (effect of
    interaction) **INT**

-   reduced model: `~ spawn_night + stage` (effects of leachate and
    stage specific leachate together) **JOINT**

&

-   full model: `~ spawn_night + stage + leachate`

-   reduced model: `~ spawn_night + stage` (tests for main effect of
    leachate) **MAIN**

This tells you if there are any interaction effects between stage and
leachate, but it won't tell you which interaction is significant, or in
which direction --- It helps you decide which model to use. If there
aren't any differences in significant gene results between models, use
the simpler model.

::: callout-caution
Order matters! The first term (`stage`) is what the model controls for,
and the second term (`leachate`) is the main effect. Meaning the model
adjusts for stage first and then assesses the leachate effect.
:::

Stage & Leachate will be treated as categorical variables

When embryonic `stage` is treated as a categorical variable DESeq2 will
estimate expression differences between stages, rather than assuming a
linear progression from 4 → 14 hpf. DESeq2 will fit different intercepts
for each stage (i.e., different baseline expression levels). If you
encode stage as a continuous numeric, the model would assume a linear
change in expression per hour of development, which is biologically
risky unless you expect linearity (note: we do not).

::: callout-note
`stage` will be treated as a categorical variable
:::

Whether `leachate` is treated as a factor or a numeric (continuous)
variable radically changes the interpretation of both the main effect
and the interaction. If a variable is numeric, DESeq2 treats it as a
continuous covariate. This fits a regression slope, estimating the
change in gene expression per unit increase in the variable. Coding
`leachate` as a categorical factor allows each stage to have its own
independent leachate effect. ✅ Use this if you expect nonlinear,
stage-specific effects of exposure (note: we do).

# Make DESeqDataSet

```{r, eval=TRUE}

dds <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ spawn_night + stage + leachate + stage:leachate))
```

dds \<- DESeqDataSetFromMatrix(countData = gcm_filtor,

colData = metadata_or,

design = formula( \~ spawn_night + stage + leachate + stage:leachate))

## Run DESeq LRT Tests

### Stage

```{r}
dds_stage_LRT <- DESeq(dds, test = "LRT", reduced = ~ spawn_night + leachate)
```

| Model name | Reduced model formula | What it tests for | Interpretation of significant genes |
|-----------------|-----------------|-----------------|---------------------|
| **dds_stage_LRT** | `~ spawn_night + leachate + stage:leachate` | Tests the contribution of **stage** | Genes whose expression significantly varies **across stage**, after accounting for spawn_night and leachate |

```{r}
res_stage <- results(dds_stage_LRT, alpha = 0.05)
summary(res_stage)
```

There are 10687 DEGs across stages.

```{r}
sig_stage <- res_stage %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

### Spawn night

Here we treat spawn night as a covariate as a way to control for it as a
nuisance factor.

```{r}
dds_sp_LRT <- DESeq(dds, test = "LRT", reduced = ~ stage + leachate + stage:leachate)
```

| Model name | Reduced model formula | What it tests for | Interpretation of significant genes |
|-----------------|-----------------|-----------------|---------------------|
| **dds_sp_LRT** | `~ stage + leachate + stage:leachate` | Tests the contribution of **spawn_night** | Genes whose expression significantly varies **across spawn nights**, after accounting for stage, leachate, and their interaction |

```{r}
res_spawn <- results(dds_sp_LRT, alpha = 0.05)
summary(res_spawn)
```

So this test asks:

"Does including spawn_night explain additional variation in gene
expression, after accounting for stage and leachate (and their
interaction)?"

The fact that we found \~3,551 genes (≈29%) significant means that
spawn_night contributes meaningfully to expression variability. That's
strong evidence that it is a real batch or parental effect --- not
random noise. BUT we're not interested in biological differences among
spawn nights in *this* study (although it's an interesting question!);
for now we stay focused on our hypothesis, looking for evidence of
low-dose **leachate** effects. In this case, we just want to adjust for
spawn nights so they don't confound our leachate comparisons. To do this
we include spawn_night in the final design formula to control for it,
but we do not extract any test results for it (no LRT against it; no
results table for it). When we interpret the coefficients (or LRT
results) for leachate, stage, and stage:leachate --- these now represent
effects after adjusting for spawn_night.

### Interaction term

```{r}
dds_int_LRT <- DESeq(dds, test = "LRT", reduced = ~ spawn_night + stage + leachate)
```

| Model name | Reduced model formula | What it tests for | Interpretation of significant genes |
|-----------------|-----------------|-----------------|--------------------|
| **dds_int_LRT** | `~ spawn_night + stage + leachate` | Tests the contribution of the **stage × leachate interaction** | Genes whose expression patterns differ across **leachate treatments in a stage-dependent manner** |

```{r}
res_int <- results(dds_int_LRT, alpha = 0.05)
summary(res_int)
```

Out of 12,221 genes tested:

-   Only 13 genes (≈ 0.1 %) were significant (FDR \< 0.05).
-   6 up-regulated, 7 down-regulated.
-   Tests **interaction only** (`stage:leachate`).
-   We got **13 DEGs** ⇒ This subset only show stage-specific responses.

For nearly all genes, the relationship between leachate level and
expression does not change significantly across stages --- . So the
interaction term adds very little explanatory power after controlling
for spawn night and main effects.

This result suggests that stage:leachate effects are largely non
significant, **minimal** or **limited to a small subset of genes**.

### Main leachate effect

```{r}
dds_main <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ spawn_night + stage + leachate))
```

```{r}
dds_main_LRT <- DESeq(dds_main, test = "LRT", reduced = ~ spawn_night + stage)
```

```{r}
res_main <- results(dds_main_LRT, alpha = 0.05)
summary(res_main)
```

Here we find zero (0) differentially expressed genes when we only model
the main effects and then reduce the model by leachate(effectively
asking, does leachate contribute significantly to the variation among
samples?) And we see that no, the stage and spawn_night variables
explain the variation well enough...

### Joint

```{r}
dds_joint_LRT <- DESeq(dds, test = "LRT", reduced = ~ spawn_night + stage)
```

```{r}
res_joint <- results(dds_joint_LRT, alpha = 0.05)
summary(res_joint)
```

→ Tests the joint contribution of all leachate terms (main effect and
stage:leachate).

→ We got 43 DEGs ⇒ there is leachate-associated signal somewhere (main
and/or interaction taken together).

This indicates some genes (30) have **weak main + weak interaction**
effects that don't cross FDR alone but **do** when evaluated together.

::: callout-important
In summary: results are consistent with stage-specific leachate
responses (interaction), not a uniform, stage-averaged leachate shift
(main effect). Hence: 0 (main-only), 13 (interaction-only), 43 (any
leachate signal jointly) is coherent.
:::

> When one performs a likelihood ratio test, the *p* values and the test
> statistic (the `stat` column) are values for the test that removes all
> of the variables which are present in the full design and not in the
> reduced design. This tests the null hypothesis that all the
> coefficients from these variables and levels of these factors are
> equal to zero.
>
> The likelihood ratio test *p* values therefore represent a test of
> *all the variables and all the levels of factors* which are among
> these variables. However, the results table only has space for one
> column of log fold change, so a single variable and a single
> comparison is shown (among the potentially multiple log fold changes
> which were tested in the likelihood ratio test). This is indicated at
> the top of the results table with the text, e.g., log2 fold change
> (MLE): condition C vs A, followed by, LRT p-value: '\~ batch +
> condition' vs '\~ batch'. This indicates that the *p* value is for the
> likelihood ratio test of *all the variables and all the levels*, while
> the log fold change is a single comparison from among those variables
> and levels.

```         
?results
```

> [Building the results
> table](https://master.bioconductor.org/packages/release/workflows/vignettes/rnaseqGene/inst/doc/rnaseqGene.html#multiple-testing:~:text=5.2-,Building%20the%20results%20table,-Calling%20results%20without):
> Calling results without any arguments will extract the estimated log2
> fold changes and p values for the last variable in the design formula.
> If there are more than 2 levels for this variable, results will
> extract the results table for a comparison of the last level over the
> first level.
>
> Genes with small *p* values from the LRT test are those which at one
> or more time points after time 0 showed a strain-specific effect. Note
> therefore that this will not give small *p* values to genes that moved
> up or down over time in the same way in both strains.

```{r}
sig_int <- res_int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

sig_joint <- res_joint%>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")

overlap_ids <- intersect(sig_int$gene_id, sig_joint$gene_id)
length(overlap_ids)        # how many
overlap_ids                # which ones
```

::: callout-note
The 13 genes found to have stage specific effects (interaction term) are
also present in the joint list that catches subtle effects of both
interaction and main leachate. We have 43 genes with significant
leachate effects overall. This is subtle (43 out of \~12K genes may be
p-hacked...) but significant based on adjusted pvalues all less than
0.05.
:::

```{r}
dim(sig_joint)
```

> The 43 genes showing significant responses to leachate exhibit
> nonlinear or non-monotonic patterns, rather than a simple linear trend
> across leachate levels.

# Wald

> Wald tests assess whether the coefficient **differs significantly from
> 0**

Run the Wald test on the dds object.

```{r}
dds <- DESeq(dds, test = "Wald")
```

```{r}
resultsNames(dds)
```

## Leachate Results

**We're focusing only on leachate effects in this analysis**

### Extract Results for interaction effects

-   "stageprawnchip.leachatelow"

-   "stageearlygastrula.leachatelow"

-   "stageprawnchip.leachatemid"

-   "stageearlygastrula.leachatemid"

-   "stageprawnchip.leachatehigh"

-   "stageearlygastrula.leachatehigh"

Each of the above interaction terms tells us the extra effect of the
leachate level at a stage (vs the additive sum of the stage and leachate
effects) in comparison to a control baseline count. For the interaction
terms, we are trying to get the "difference of the differences".

```{r}
# prawnchip interaction terms
res_prawnchip_high.int <- results(dds, name = "stageprawnchip.leachatehigh", alpha = 0.05)
res_prawnchip_mid.int <- results(dds, name = "stageprawnchip.leachatemid", alpha = 0.05)
res_prawnchip_low.int  <- results(dds, name = "stageprawnchip.leachatelow", alpha = 0.05)

# early gastrula interaction terms
res_earlygastrula_high.int <- results(dds, name = "stageearlygastrula.leachatehigh", alpha = 0.05)
res_earlygastrula_mid.int <- results(dds, name = "stageearlygastrula.leachatemid", alpha = 0.05)
res_earlygastrula_low.int <- results(dds, name = "stageearlygastrula.leachatelow", alpha = 0.05)
```

#### Prawnchip

##### high

```{r}
summary(res_prawnchip_high.int)
```

::: callout-note
0 genes showed a significant stage × leachate interaction in the
prawnchip stage under the high treatment (Wald test, BH-adjusted p \<
0.05).
:::

::: column-margin
> DESeq2 uses size-factor normalization (the "median-ratio method"). It
> corrects for library size and composition so you can compare counts
> across samples. How it works (median-ratio method) Per-gene geometric
> mean across samples (skip genes with all zeros). Per-sample ratios:
> for each gene, divide a sample's count by that gene's geometric mean.
> Size factor: the median of those ratios within each sample. Normalized
> counts: normalized = raw_count / size_factor. These size factors are
> used as an offset in the negative binomial GLM for the Wald/LRT tests.
> Normalization is independent of the design formula.
:::

##### mid

```{r}
summary(res_prawnchip_mid.int)
```

::: callout-note
0 genes were found to have significant prawnchip stage x mid leachate
interaction (adjusted p value \< 0.05)
:::

##### low

```{r}
summary(res_prawnchip_low.int)
```

::: callout-note
7 genes were found to have significant prawnchip stage x low leachate
interaction (adjusted p value \< 0.05)
:::

```{r}
sig_prawnchip_low.int <- res_prawnchip_low.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_prawnchip_low.int, file.path(output_path, "sig_prawnchip_low.int.csv"))
```

#### Early gastrula

##### high

```{r}
summary(res_earlygastrula_high.int)
```

::: callout-note
1 gene was found to have a significant interaction effect between early
gastrula and high leachate
:::

```{r}
sig_earlygastrula_high.int <- res_earlygastrula_high.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_earlygastrula_high.int, file.path(output_path, "sig_earlygastrula_high.int.csv"))
```

##### mid

```{r}
summary(res_earlygastrula_mid.int)
```

::: callout-note
0 genes were found to have a significant interaction between early
gastrula and mid leachate
:::

##### low

```{r}
summary(res_earlygastrula_low.int)
```

::: callout-note
8 genes were found to have significant interaction between early
gastrula and low leachate
:::

```{r}
sig_earlygastrula_low.int <- res_earlygastrula_low.int %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_earlygastrula_low.int, file.path(output_path, "sig_earlygastrula_low.int.csv"))
```

## Extract results for stage by leachate contrasts

See [A guide to designs and contrasts in
DESeq2](https://www.atakanekiz.com/technical/a-guide-to-designs-and-contrasts-in-DESeq2/)
by Dr. Atakan Ekiz

```{r}
resultsNames(dds)
```

```{r}
# cleavage 
res_cleavage_low.vs.control <- results(dds, name = "leachate_low_vs_control", alpha = 0.05)
res_cleavage_mid.vs.control  <- results(dds, name = "leachate_mid_vs_control", alpha = 0.05)
res_cleavage_high.vs.control <- results(dds, name = "leachate_high_vs_control", alpha = 0.05)

# prawnchip
res_prawnchip_low.vs.control <- results(dds, contrast = list(c("leachate_low_vs_control", "stageprawnchip.leachatelow")), alpha = 0.05)
res_prawnchip_mid.vs.control  <- results(dds, contrast = list(c("leachate_mid_vs_control",  "stageprawnchip.leachatemid")), alpha = 0.05)
res_prawnchip_high.vs.control <- results(dds, contrast = list(c("leachate_high_vs_control", "stageprawnchip.leachatehigh")), alpha = 0.05)

# earlygastrula
res_earlygastrula_low.vs.control <- results(dds, contrast = list(c("leachate_low_vs_control", "stageearlygastrula.leachatelow")), alpha = 0.05)
res_earlygastrula_mid.vs.control  <- results(dds, contrast = list(c("leachate_mid_vs_control",  "stageearlygastrula.leachatemid")), alpha = 0.05)
res_earlygastrula_high.vs.control <- results(dds, contrast = list(c("leachate_high_vs_control", "stageearlygastrula.leachatehigh")), alpha = 0.05)

```

### Cleavage

#### high

```{r}
summary(res_cleavage_high.vs.control)
```

::: callout-note
7 genes were found to be significantly differentially expressed
(adjusted p value \< 0.05) at the cleavage stage in the high leachate
treatment compared to the control.
:::

```{r}
sig_cleavage_high.vs.control <- res_cleavage_high.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_cleavage_high.vs.control, file.path(output_path, "sig_cleavage_high.vs.control.csv"))
```

#### mid

mid leachate vs. control (mvc) for cleavage

```{r}
summary(res_cleavage_mid.vs.control)
```

::: callout-note
58 genes were found to be significantly differentially expressed
(adjusted p value \< 0.05) at the cleavage stage in the mid leachate
treatment compared to the control. 57 LFC \> 0. 57/58 total genes were
found in more abundance compared to the controls.
:::

```{r}
sig_cleavage_mid.vs.control <- res_cleavage_mid.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_cleavage_mid.vs.control, file.path(output_path, "sig_cleavage_mid.vs.control.csv"))
```

#### low

for cleavage (baseline stage) \> Because you fit \~ 0 + stage +
leachate + stage:leachate, **the coefficient leachatelow is the low vs
control effect at the reference stage**. From your resultsNames() (no
stagecleavage.leachatelow term is present) the reference stage is
cleavage, so Low vs control at cleavage is simply the leachatelow
coefficient.

```{r}
summary(res_cleavage_low.vs.control)
```

::: callout-note
34 genes were found to be significantly differentially expressed
(adjusted p value \< 0.05) at the cleavage stage in the low leachate
treatment compared to the control. 13 LFC \> 0 & 21 LFC \< 0.
:::

```{r}
sig_cleavage_low.vs.control <- res_cleavage_low.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_cleavage_low.vs.control, file.path(output_path, "sig_cleavage_low.vs.control.csv"))
```

### Prawnchip

#### high

```{r}
summary(res_prawnchip_high.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn
chip stage in the high leachate treatment when compared to the control.
:::

#### mid

```{r}
summary(res_prawnchip_mid.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn
chip stage in the high leachate treatment when compared to the control.
:::

#### low

```{r}
summary(res_prawnchip_low.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the prawn
chip stage in the high leachate treatment when compared to the control.
:::

### Early gastrula

#### high

```{r}
summary(res_earlygastrula_high.vs.control)
```

::: callout-note
27 genes (0.22%) differentially expressed (adjusted p value \< 0.05) at
the early gastrula stage in the high leachate treatment when compared to
the control. All 27 of these genes had more abundant transcripts in
treated samples compared to controls.
:::

```{r}
sig_earlygastrula_high.vs.control <- res_earlygastrula_high.vs.control %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
write_csv(sig_earlygastrula_high.vs.control, file.path(output_path, "sig_earlygastrula_high.vs.control.csv"))
```

#### mid

```{r}
summary(res_earlygastrula_mid.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early
gastrula stage in the mid leachate treatment when compared to the
control.
:::

#### low

```{r}
summary(res_earlygastrula_low.vs.control)
```

::: callout-note
0 genes differentially expressed (adjusted p value \< 0.05) at the early
gastrula stage in the low leachate treatment when compared to the
control.
:::

# Stage Results

```{r}
resultsNames(dds)
```

```{r}
# 1. Get results for each stage vs cleavage
res_prawnchip   <- results(dds, name = "stage_prawnchip_vs_cleavage", alpha = 0.05)
res_earlygastrula   <- results(dds, name = "stage_earlygastrula_vs_cleavage", alpha = 0.05)

# Compare prawnchip vs earlygastrula
res_pc.vs.eg <- results(dds, contrast = list(c("stage_prawnchip_vs_cleavage", "stage_earlygastrula_vs_cleavage")), alpha = 0.05)

```

```{r}
summary(res_prawnchip)
```

There are 6880 DEGs between the cleavage and prawnchip stages. Of those,
3665 genes (30%) are up-regulated and 3215 genes (26%) are
down-regulated.

```{r}
sig_prawnchip <- res_prawnchip %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r}
summary(res_earlygastrula)
```

There are 8007 DEGs between the cleavage and earlygastrula stages. Of
those, 4492 genes (37%) are up-regulated and 3515 genes (29%) are
down-regulated.

```{r}
sig_earlygastrula <- res_earlygastrula %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id")
```

```{r}
summary(res_pc.vs.eg)
```

There are 7780 DEGs between the prawnchip and earlygastrula stages. Of
those, 4331 genes (35%) are up-regulated and 3449 genes (28%) are
down-regulated.

```{r}
sig_pc.vs.eg <- res_pc.vs.eg %>%
  as.data.frame() %>%
  filter(padj < 0.05) %>% 
  rownames_to_column(var = "gene_id") 

nrow(sig_pc.vs.eg)
```

```{r, eval=FALSE}
write_csv(sig_stage, file.path(output_path, "sig_stage.csv"))
write_csv(sig_prawnchip, file.path(output_path, "sig_prawnchip.csv"))
write_csv(sig_earlygastrula, file.path(output_path, "sig_earlygastrula.csv"))
write_csv(sig_pc.vs.eg, file.path(output_path, "sig_pc.vs.eg.csv"))
```

# Results gene lists

## Leachate

We have 9 different 'significant gene' lists for leachate. Each list
indicates genes in which leachate treatment contributed to significantly
differential expression.

LRT:

-   `sig_int`, 13 DEGs
-   `sig_joint`, 44 DEGs

::: callout-note
No main effects due to leachate detected - leachate effects were only
detected within interactions and within stage contrasts
:::

Interaction effects:

-   `sig_prawnchip_low.int`, 7 DEGs
-   `sig_earlygastrula_high.int`, 1 DEG
-   `sig_earlygastrula_low.int`, 8 DEGs

Contrasts:

-   `sig_cleavage_low.vs.control`, 34 DEGs
-   `sig_cleavage_mid.vs.control`, 58 DEGs
-   `sig_cleavage_high.vs.control`, 7 DEGs
-   `sig_earlygastrula_high.vs.control`, 27 DEGs

## Stage

LRT:

-   `sig_stage`, 10687 DEGs

Main effects:

-   `sig_prawnchip`, 6880 DEGs
-   `sig_earlygastrula`, 8007 DEGs

Contrasts:

-   `sig_pc.vs.eg`, 7780 DEGs

# Save gene lists

## Leachate

Make a list of all genes that significantly change with leachate
exposure

```{r}
# 1) Put all of your sig data frames into a *named* list
sig_lists_leachate <- list(
  int                           = sig_int,
  joint                         = sig_joint,
  prawnchip_low.int             = sig_prawnchip_low.int,
  earlygastrula_high.int        = sig_earlygastrula_high.int,
  earlygastrula_low.int         = sig_earlygastrula_low.int,
  cleavage_low.vs.control       = sig_cleavage_low.vs.control,
  cleavage_mid.vs.control       = sig_cleavage_mid.vs.control,
  cleavage_high.vs.control      = sig_cleavage_high.vs.control,
  earlygastrula_high.vs.control = sig_earlygastrula_high.vs.control
)

# 2) Smoosh: add a 'list' column indicating the source and stack them
sig_summary_leachate <- bind_rows(sig_lists_leachate, .id = "list") %>%
  relocate(list, gene_id)
```

```{r}
sig_genes_leachate <- dplyr::distinct(sig_summary_leachate, gene_id)
nrow(sig_genes_leachate)
```

```{r, eval=FALSE}
write_csv(sig_genes_leachate, file.path(output_path, "sig_genes_leachate.csv"))
```

::: callout-note
There are 130 total DEGs across LRT and Wald tests where leachate popped
up as a significant covariate.
:::

## Stage

Make a list of all genes that significantly change across stages

```{r}
sig_lists_stage <- list(
  stage_LRT      = sig_stage,
  pc_v_cl        = sig_prawnchip,
  eg_v_cl        = sig_earlygastrula,
  pc_v_eg        = sig_pc.vs.eg
)

# 2) Smoosh: add a 'list' column indicating the source and stack them
sig_summary_stage <- bind_rows(sig_lists_stage, .id = "list") %>%
  relocate(list, gene_id)
```

```{r}
sig_genes_stage <- dplyr::distinct(sig_summary_stage, gene_id)
nrow(sig_genes_stage)
```

```{r, eval=FALSE}
write_csv(sig_genes_stage, file.path(output_path, "sig_genes_stage.csv"))
```

::: callout-note
There are 10793 total DEGs across LRT and Wald tests where stage was a
significant contributor to differential expression.
:::

## Gene universe

Make a list of all genes tested in DESeq2 analysis, to be used as 'gene
universe' background for functional enrichment.

```{r}
all_genes <- rownames(dds) # a character vector of all genes used in DESeq2 analysis
# save this for functional enrichment 
all_genes %>% tibble(gene_id = all_genes) %>% nrow()
```

```{r, eval=FALSE}
write_csv(tibble(gene_id = all_genes),  file.path(output_path, "all_genes.csv"))
```

::: callout-note
There are 12,221 total genes tested in DESeq2 after filtering low count
genes.
:::

# Save count matrices

Make variance stabilized count matrix of these 130 significant leachate
genes of interest, the significant stage genes of interest, and save the
entire gene universe VST count matrix.

```{r}
# VST transform
vsd <- vst(dds, blind = FALSE) # DESeqTransform Object
vst_mat <- assay(vsd) # VST count matrix with genes as rows and samples as columns

# Keep the 130 genes of interest
vst_mat_130 <- vst_mat[sig_genes_leachate$gene_id, , drop = FALSE]
dim(vst_mat_130)

# Keep the stage genes of interest
vst_mat_stage <- vst_mat[sig_genes_stage$gene_id, , drop=FALSE]
dim(vst_mat_stage)
```

```{r}
# Convert to dataframes 
vst_mat_df <- vst_mat %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

vst_mat_130_df <- vst_mat_130 %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

vst_mat_stage_df <- vst_mat_stage %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
# and save 
write_csv(vst_mat_df, file.path(output_path,"vst_mat_df.csv"))
write_csv(vst_mat_130_df, file.path(output_path,"vst_mat_130_df.csv"))
write_csv(vst_mat_stage_df, file.path(output_path,"vst_mat_stage_df.csv"))
```

Make library-size adjusted count matrices

```{r}
norm <- counts(dds, normalized = TRUE) # size factor normalized count matrix with genes as rows and samples as columns

# Keep the 130 leachate genes of interest
norm_130 <- norm[sig_genes_leachate$gene_id, , drop = FALSE]
dim(norm_130)

# Keep the stage genes of interest
norm_stage <- norm[sig_genes_stage$gene_id, , drop = FALSE]
dim(norm_stage)
```

```{r}
# Convert to dataframes 
norm_df <- norm %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

norm_130_df <- norm_130 %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

norm_stage_df <- norm_stage %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
# and save 
write_csv(norm_df, file.path(output_path,"norm_df.csv"))
write_csv(norm_130_df, file.path(output_path,"norm_130_df.csv"))
write_csv(norm_stage_df, file.path(output_path,"norm_stage_df.csv"))
```

Make raw count matrices

```{r}
raw <- counts(dds) # raw count matrix with genes as rows and samples as columns
dim(raw)

# Keep the 130 genes of interest
raw_130 <- raw[sig_genes_leachate$gene_id, , drop = FALSE]
dim(raw_130)

# Keep the stage genes of interest
raw_stage <- raw[sig_genes_stage$gene_id, , drop = FALSE]
dim(raw_stage)
```

```{r}
# Convert to dataframes  
raw_df <- raw %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

raw_130_df <- raw_130 %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")

raw_stage_df <- raw_stage %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "gene_id")
```

```{r, eval=FALSE}
# and save
write_csv(raw_df, file.path(output_path,"raw_df.csv"))
write_csv(raw_130_df, file.path(output_path,"raw_130_df.csv"))
write_csv(raw_stage_df, file.path(output_path,"raw_stage_df.csv"))
```

# QC

Covariates effect on count data

```{r, fig.height=6, fig.width=12}
resCov <- degCovariates(log2(counts(dds)+0.5),
                        colData(dds))
```

Covariates correlation with metrics

```{r}
cor <- degCorCov(colData(dds))
```

# Clustering

Detect Patterns of Expression see Roberts Lab Issue
[#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering. You'll need: - a metadata matrix with sample_id as
rows - a gene count matrix that has been variance stabilized and
filtered to your significant genes of interest `vst_mat_130` genes in
rows, sample ID in columns

```{r}
meta_mat <- metadata_or %>%
  column_to_rownames(var = "sample_id")
```

Run with minimum 10 genes per cluster to detect because we have few
DEGs.\
- DEGreport vignette,
[degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
leachate130_GEPbystage <- degPatterns(vst_mat_130, metadata = meta_mat, time = "stage", col= "leachate", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bystage_plot.png", width=12, height=10)
```

Broadly most genes displayed an increased transcript abundance as
embryos developed.

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
leachate130_GEPbydose <- degPatterns(vst_mat_130, metadata = meta_mat, time = "leachate", col="stage", plot=TRUE, minc = 10)
```

```{r, eval=FALSE}
ggsave(filename="../../output/06_deg/cluster/DEG_LeachateClusterPatterns_bydose_plot.png", width=12, height=10)
```

We have many DEGs for stage so here we are going to increase our minimum
grouping of genes from 10 to 100

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_GEPbystage <- degPatterns(vst_mat_stage, metadata = meta_mat, time = "stage", col= "leachate", plot=TRUE, minc = 100)
```

```{r}
ggsave(filename="../../output/06_deg/cluster/DEG_StageClusterPatterns_bystage_plot.png", width=12, height=10)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_GEPbydose <- degPatterns(vst_mat_130, metadata = meta_mat, time = "leachate", col= "stage", plot=TRUE, minc = 100)
```

```{r}
ggsave(filename="../../output/06_deg/cluster/DEG_StageClusterPatterns_bydose_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for stage and for leachate.
Notice that the gene counts for the patterns (1, 2, 3) are the same
across stage and leachate... very interesting. The genes that have high
relative transcript abundance in the earlygastrula stage group in group
1, high relative transcript abundance in prawnchip are group 2, and high
relative abundance in cleavage are group 3. There is a general trend of
more transcripts present as development progresses... which makes sense
because the embryo as it grows is doing more. The subtle leachate
effects we do see (non monotonic patterns in cleavage in group 1 and 2,
slight increase under high pvc leachate exposure in early gastrula in
group 3) appear to impact the transcript abundances of the stage with
the lowest abundances of those genes... so the lower the gene abundance
the more effect leachate has on those genes? Is that just an artifact of
increased variability at lower transcript abundance? Are they the same
genes? The following code should return TRUE if that is the case.

```{r}
# Rename and extract information 
leachate130_GEP_df <- leachate130_GEPbystage$df %>%
          rename(gene_id=genes)

stage_GEP_df <- stage_GEPbydose$df %>% 
  rename(gene_id=genes)
```

Save the gene expression patterns

```{r, eval=FALSE}
write_csv(leachate_GEP_clusters, file.path("../../output/06_deg/cluster/gene_clusters.csv"))

write_csv(stage_GEP_clusters, file.path("../../output/06_deg/cluster/gene_clusters.csv"))
```

Plot is showing Z score of VST normalized gene counts

```{r}
gene_cluster_data <- leachate_clusters[["normalized"]]
gene_cluster_data <- as.data.frame(gene_cluster_data) %>% 
  select(1:12) %>% 
  rename(zscore = value)
```

```{r, eval=FALSE}
write_csv(gene_cluster_data, "../../output/06_deg/cluster/gene_cluster_zscores.csv")
```

```{r}
gene_cluster_data <- read_csv("../../output/06_deg/cluster/gene_cluster_zscores.csv")
```

```{r}
leachate_colors <- c(control = "#BFE8D9",
                     low = "#DFB431",
                     mid = "#8E6B01",
                     high = "#503D02"
                     )

stage_colors <- c(cleavage = "#B7B5AC", 
                  prawnchip = "#6F6E69", 
                  earlygastrula = "#343331"
                  )
```

## Plot gene clusters

```{r}
# reverse gene cluster order to plot from cleavage to eg
gene_cluster_data <- gene_cluster_data %>%
  mutate(cluster = factor(cluster, levels = rev(sort(unique(cluster)))))

facet_labs <- gene_cluster_data %>%
  summarise(n_genes = n_distinct(genes), .by = cluster) %>%        # counts unique genes per cluster
  mutate(lab = sprintf("Group: %s - genes: %d", cluster, n_genes)) %>%
  dplyr::select(cluster, lab) %>%
  deframe()                                                        # named vector: names = cluster, values = lab
```

```{r, warning=FALSE, fig.height=8, fig.width = 12}

# leachate as x plot
pl <- ggplot(gene_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high"))),
       aes(x = leachate, y = zscore, color = stage)) +
  geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = stage), se = FALSE) + # trend
  facet_wrap(~ cluster, nrow = 1) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "stage") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = stage_colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_blank(),
  strip.background = element_blank())

# stage as x plot
ps <- ggplot(gene_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high")),
                stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula"))),
       aes(x = stage, y = zscore, color = leachate)) +
  geom_line(aes(group = interaction(genes, leachate)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = leachate), se = FALSE) + #trend pattern 
  facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = facet_labs)) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "leachate") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = leachate_colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_text(face = "bold"))


# combine with patchwork
pc <- ps / pl

pc

```

```{r, eval=FALSE}
ggsave("../../output/06_deg/cluster/leachate_cluster.png", plot = pl, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/stage_cluster.png", plot = ps, width=18, height=6 , dpi=600)
ggsave("../../output/06_deg/cluster/combo_cluster.png", plot = pc, width=18, height=12 , dpi=600)
```

hmmm gotta arrange so its group 3 , 2, 1 ... and more space between
groups so that its obvious to compare the top and bottom graphs. Pick
distinct and pretty leachate and stage colors...

## Add gene cluster annotation to gene list summaries

```{r}
# leachate
sig_summary_leachate <- sig_summary_leachate %>%
  left_join(gene_clusters, by = "gene_id")

write_csv(sig_summary_leachate, file.path(output_path, "sig_summary_leachate.csv"))

# stage 
sig_summary_stage <- sig_summary_stage %>% 
  left_join(gene_clusters, by = "gene_id")

write_csv(sig_summary_stage, file.path(output_path, "sig_summary_stage.csv"))
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!

# Next steps

-   Make a PCA and PERMANOVA of the DEGs ...
-   I'd simply like an easy absence/presence heatmap that shows all the
    genes, and which lists they are significant in.
-   I'd then like to show on the heatmap the log2FC, and add astericks
    for significance levels p\<0.05, p\<0.01, p\<0.001.
-   I'd like gene count plots visually summarizing each list type (LRT,
    Interaction, Stagexleachate effects) in tiny tidy multiples
-   Do these genes cluster into expression patterns? ... Clustering
-   What do these genes do ? ... GO enrichment
-   Are there correlations between these differentially expressed genes
    and differentially abundant taxa?
