---
title: "Multifactorial Differential Gene Expression"
subtitle: "using DESeq2 contrasts"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = FALSE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = ""         # Prevents appending '##' to beginning of lines in code output
)
```

# Background

Here we are looking at a multifactorial differential gene expression analysis for developing coral embryos at 3 developmental stages (4 hours post fertilization , 9 hours post fertilization, and 14 hours post fertiliation). Embryos at each stage were exposed to either a control, low, mid, or high concentration of PVC leachate. This work aims to understand sublethal effects of microplastic pollution on coral reefs.

## Inputs

-   `gcm_filtor.csv` is the filtered gene count matrix for 61 samples (two outliers removed).
-   `metadata_or.csv` is the metadata for 61 samples (two outliers removed)

## Outputs

-   `volcano_facet_interaction.png`

# Setup

## Install packages

Copy and run this once in the console

```         
# CRAN packages
install.packages(c(
  "tidyverse",
  "viridis",
  "plotly",
  "cowplot",
  "pheatmap",
  "vsn",
  "ashr",
  "remotes"  # required for GitHub installs
))

# GitHub package
remotes::install_github("mdscheuerell/flexoki")

# Bioconductor
if (!"BiocManager" %in% rownames(installed.packages())) {
  install.packages("BiocManager")
}
BiocManager::install(c("DESeq2", "EnhancedVolcano"))
```

## Load packages

```{r}
# Load packages
library(DESeq2)
library(viridis)
library(tidyverse)
library(EnhancedVolcano)
library(ashr)
library(flexoki)
library(plotly)
library(cowplot)
library(vsn)
library(pheatmap)
```

## Import filtered gene count matrix

In this count matrix, each row represents a gene, each column a sequenced RNA library, and the values give the estimated counts of fragments that were assigned to the respective gene in each library by `HISAT2`

-   `check.names = FALSE` removed the 'X' from in front of sample id's in the column headers

```{r, eval=TRUE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE)
head(gcm_filtor)
```

-   The raw unfiltered, untransformed, count matrix shows counts of whole numbers (integers)
-   Each column is a sample
-   Each row is a gene
-   A sample id describes parental crosses (the first string of numbers), the pollution exposure (C, L, M, H), and the embryonic stage (4, 9, 14)

::: callout-note
`gcm_filtor` is the gene count matrix that has been filtered down to only contain rows where counts are at least 15 in 75% of the samples and 2 outlier samples ( `131415L4` and `789C4` ) that had poor HISAT alignment and did not group in the exploratory PCA are already removed
:::

## Import metadata

```{r, eval=TRUE}
metadata_or <- read_csv("../../metadata/metadata_or.csv")
str(metadata_or)
```

```{r, eval=TRUE}
# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")

str(metadata_or)
```

::: callout-note
`metadata_filt` contains metadata for 61 samples with explanatory variables: - `stage` is a factor with 3 levels (cleavage, prawnchip, earlygastrula) - `hpf` is a discrete numerical variable representing the hours post fertilization 4, 9, and 14 (corresponding to `stage`)

-   `leachate` is a factor with 4 levels (control, low, mid, high)
-   `leachate_mgL` is a continuous numerical variable representing leachate concentrations in mg/L with values 0 , 0.01 , 0.1 , and 1 (corresponding to `leachate`)
:::

# Continuous vs categorical

Embryonic stage should be treated as a categorical variable because DESeq2 will estimate expression differences between stages, rather than assuming a linear progression from 4 → 14 hpf. It will fit different intercepts for each stage (i.e., different baseline expression levels). If you encode stage as numeric, the model would assume a linear change in expression per hour of development, which is biologically risky unless you expect linearity (note: we do not).

Whether leachate is treated as a factor or a numeric (continuous) variable radically changes the interpretation of both the main effect and the interaction. If a variable is numeric, DESeq2 treats it as a continuous covariate. This fits a regression slope, estimating the change in gene expression per unit increase in the variable.

Coding `leachate` as a categorical factor allows each stage to have its own independent leachate effect. ✅ Use this if you expect nonlinear, stage-specific effects of exposure.

```{r, eval=TRUE}

dds_cat <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ stage + leachate + stage:leachate))
```

Coding `leachate_mgL` as a continuous numeric tests for a change in expression per unit increase in leachate, and the interaction term tests whether the slope of the dose-response differs depending on stage. ✅ Use this if you expect a linear or dose-dependent trend, and want to ask if the strength or direction of the trend varies by stage.

```{r, eval=TRUE}
dds_con <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ stage + leachate_mgL + stage:leachate_mgL))
```

## Run DESeq LRT Tests

> Does modeling leachate as a factor vs. a continuous dose better explain gene expression changes across stages?

```{r, eval=TRUE}
dds_cat <- DESeq(dds_cat, test = "LRT", reduced = ~ stage + leachate)
dds_con <- DESeq(dds_con, test = "LRT", reduced = ~ stage + leachate_mgL)
```

| Model         | Full model                                  | Reduced model          | LRT tests...                                               |
|----------------|-----------------|----------------|-----------------------|
| `dds_cat_LRT` | `stage + leachate + stage:leachate`         | `stage + leachate`     | Are there **stage-specific effects of leachate (factor)**? |
| `dds_con_LRT` | `stage + leachate_mgL + stage:leachate_mgL` | `stage + leachate_mgL` | Are there **stage-specific trends** in dose response?      |

## Results

```{r, eval=TRUE}
res_cat <- results(dds_cat, alpha = 0.05)
summary(res_cat)
```

```{r, eval=TRUE}
res_con <- results(dds_con, alpha = 0.05)
summary(res_con)
```

```{r, eval=TRUE}
sig_cat <- rownames(results(dds_cat, alpha = 0.05)[which(results(dds_cat)$padj < 0.05), ])
sig_con <- rownames(results(dds_con, alpha = 0.05)[which(results(dds_con)$padj < 0.05), ])
length(intersect(sig_cat, sig_con))
```

::: callout-note
2 out of 44 genes have a linear relationship (compared to a non-linear one!) as we look at the interactions between leachate and stage. All 44 of these genes respond to leachate differently depending on stage.

44 genes show significant stage-by-leachate interaction when leachate is treated categorically → meaning the response to leachate varies by stage, but not necessarily in a linear/dose-dependent way.

2 genes show significant stage-specific dose-response effects when leachate is treated as continuous → they show a linear or monotonic trend across leachate concentrations that varies by stage.

The 2 overlapping genes are well-modeled by a linear trend in leachate (i.e., a slope, not just group differences). The 42 other genes show non-linear or non-monotonic behavior --- meaning their response does not follow a consistent increase/decrease with dose. Instead, they respond differently to each specific leachate level depending on the stage (e.g., they might increase at 0.1 mg/L but decrease at 1 mg/L, or vice versa).
:::

> Of the 44 genes showing significant stage-specific responses to leachate (based on the categorical model), only 2 also show a significant linear dose-response interaction with stage (based on the continuous model). This suggests that while stage modulates leachate response broadly, most genes (42/44) exhibit nonlinear or non-monotonic patterns, rather than a simple linear trend across leachate levels. keep.order = TRUE)

```{r, eval=TRUE}
vsd <- assay(vst(dds_cat))

# make long expression dataframe of significant interaction genes
sig_long <- vsd[sig_cat, ] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "sample_id", values_to = "expression")

# add metadata
sig_long <- sig_long %>% left_join(metadata_or, by = "sample_id")
```

### Loop plots of significant interaction genes

```{r}
# create output directory
dir.create("../../output/06_deg/gene_plots", recursive = TRUE, showWarnings = FALSE)

for (gene in unique(sig_long$gene_id)) {

  gene_df <- sig_long %>% filter(gene_id == gene)

  p <- ggplot(gene_df, aes(x = leachate, y = expression, color = stage, group = stage)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_smooth(method = "loess", se = FALSE) +
    labs(
      title = paste("Gene:", gene),
      x = "Leachate (mg/L)",
      y = "VST expression"
    ) +
    
    theme_minimal(base_size = 12) +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5)
    )
  
  # Render as interactive plotly
  print(plotly::ggplotly(p))

  # Save to PNG
  ggsave(
    filename = paste0("../../output/06_deg/gene_plots/", gene, ".png"),
    plot = p, width = 6, height = 4, dpi = 300
  )
}
```

::: panel-tabset
```{r, eval=TRUE}
for (gene in unique(sig_long$gene_id)) {

  gene_df <- sig_long %>% filter(gene_id == gene)

  cat("### ", gene, "\n\n")  # Quarto tab header

  p <- ggplot(gene_df, aes(x = leachate, y = expression, color = stage, group = stage)) +
    geom_point(size = 2, alpha = 0.8) +
    geom_line(stat = "summary", fun = mean, linewidth = 0.8) +
    labs(
      x = "Leachate (mg/L)",
      y = "Rlog expression"
    ) +
    theme_minimal(base_size = 12)

  print(plotly::ggplotly(p))
}
```
:::

```{r}
resultsNames(dds_cat)
```

# LRT vs Wald

> Wald tests on interaction terms assess whether the interaction coefficient **differs significantly from 0** --- i.e., whether the combined effect of that `hpf` and `leachate` level deviates from what would be expected by the additive effects alone.

```{r}
dds <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~ stage + leachate + stage:leachate))
```

```{r}
dds_wald <- DESeq(dds, test = "Wald")
```

```{r}
resultsNames(dds_wald)
```

```{r}
resultsNames(dds_cat)
```

## Plot dispersion

```{r}
plotDispEsts(dds_cat)
plotDispEsts(dds_wald)
```

::: callout-note
Dispersion trend line is hovering around 0.1, and bulk spread of tag-wise gene estimates is from 0.01 to 1.0
:::

# Build contrasts

```{r}
resultsNames(dds_or)  # see what contrast names are available
```

| Coef | Name                         | Meaning                                                               |
|------------------|------------------|-------------------------------------|
| 1    | `"Intercept"`                | Baseline expression: `hpf = 4` and `leachate = control`               |
| 2    | `"hpf_9_vs_4"`               | Change from 4 hpf → 9 hpf, at **control** exposure                    |
| 3    | `"hpf_14_vs_4"`              | Change from 4 hpf → 14 hpf, at **control** exposure                   |
| 4    | `"leachate_low_vs_control"`  | Change from control → low exposure, at **4 hpf**                      |
| 5    | `"leachate_mid_vs_control"`  | Change from control → mid exposure, at **4 hpf**                      |
| 6    | `"leachate_high_vs_control"` | Change from control → high exposure, at **4 hpf**                     |
| 7    | `"hpf9.leachatelow"`         | **Interaction term**: extra effect of low exposure at 9 hpf vs 4 hpf  |
| 8    | `"hpf14.leachatelow"`        | **Interaction term**: extra effect of low exposure at 14 hpf vs 4 hpf |
| 9    | `"hpf9.leachatemid"`         | **Interaction term**: extra effect of mid exposure at 9 hpf vs 4 hpf  |
| 10   | `"hpf14.leachatemid"`        | **Interaction term**: extra effect of mid exposure at 14 hpf vs 4 hpf |
| 11   | `"hpf9.leachatehigh"`        | **Interaction term**:... of high exposure at 9 hpf vs 4 hp            |
| 12   | `"hpf14.leachatehigh"`       | **Interaction term**:... of high exposure at 14 hpf vs 4 hpf          |

Contrast summary table \| What you want \| Contrast \| \| ------------------------- \| ----------------------------------------------------- \| \| High vs Control at 4 hpf \| `"leachate_high_vs_control"` \| \| Mid vs Control at 4 hpf \| `"leachate_mid_vs_control"` \| \| Low vs Control at 4 hpf \| `"leachate_low_vs_control"` \| \| High vs Control at 9 hpf \| `c("leachate_high_vs_control", "hpf9.leachatehigh")` \| \| Mid vs Control at 9 hpf \| `c("leachate_mid_vs_control", "hpf9.leachatemid")` \| \| Low vs Control at 9 hpf \| `c("leachate_low_vs_control", "hpf9.leachatelow")` \| \| High vs Control at 14 hpf \| `c("leachate_high_vs_control", "hpf14.leachatehigh")` \| \| Mid vs Control at 14 hpf \| `c("leachate_mid_vs_control", "hpf14.leachatemid")` \| \| Low vs Control at 14 hpf \| `c("leachate_low_vs_control", "hpf14.leachatelow")` \|

# Effect of high exposure at 14 hpf

When you test for main effects of leachate (e.g., "leachate_high_vs_control"), it's at the reference timepoint (hpf = 4). The interaction terms model how the leachate effect changes relative to that baseline as development progresses.

What's the effect of high exposure at 14 hpf? You must combine: - "leachate_high_vs_control": main leachate effect at 4 hpf - "hpf14.leachatehigh": additional effect at 14 hpf

```{r}
# output results of contrast
res_14hpf_high <- (results(dds, contrast = list(
  c("leachate_high_vs_control", "hpf14.leachatehigh")
)))

summary(res_14hpf_high)
```

### Extract significant genes

```{r}
padj.cutoff <- 0.05
lfc.cutoff <- 0.5
```

```{r}
# Convert results table into tibble
res_14hpf_high_tb <- res_14hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_high_noLFCcutoff <- res_14hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_high <- sig_14hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at high exposure level:", nrow(sig_14hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at high exposure level:", nrow(sig_14hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_high, "../output/09_deg_multifactorial/sig_14hpf_high.csv")
write_csv(sig_14hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_14hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf14.leachatehigh")), type = "ashr")

plotMA(res_14hpf_high)
plotMA(res_14hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_high,
    lab = rownames(res_14hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14high <- 
  ggplot(res_14hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_high[sig_14hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_high[sig_14hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14high
```

## Effect of mid exposure at 14 hpf

```{r}
# output results of contrast
res_14hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf14.leachatemid")
)))

summary(res_14hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_14hpf_mid_tb <- res_14hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_mid_noLFCcutoff <- res_14hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_mid <- sig_14hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at mid exposure level:", nrow(sig_14hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at mid exposure level:", nrow(sig_14hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_mid, "../output/09_deg_multifactorial/sig_14hpf_mid.csv")
write_csv(sig_14hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}

# shrink lfc for visualizations
res_14hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf14.leachatemid")), type = "ashr")

plotMA(res_14hpf_mid)
plotMA(res_14hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_mid,
    lab = rownames(res_14hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14mid <- 
  ggplot(res_14hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_mid[sig_14hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_mid[sig_14hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14mid
```

## Effect of low exposure at 14 hpf

```{r}
# output results of contrast
res_14hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf14.leachatelow")
)))

summary(res_14hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_14hpf_low_tb <- res_14hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_14hpf_low_noLFCcutoff <- res_14hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_14hpf_low <- sig_14hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 14 hpf at low exposure level:", nrow(sig_14hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 14 hpf at low exposure level:", nrow(sig_14hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_14hpf_low, "../output/09_deg_multifactorial/sig_14hpf_low.csv")
write_csv(sig_14hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_14hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_14hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf14.leachatelow")), type = "ashr")

plotMA(res_14hpf_low)
plotMA(res_14hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_14hpf_low,
    lab = rownames(res_14hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 14 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v14low <- 
  ggplot(res_14hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_14hpf_low[sig_14hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_14hpf_low[sig_14hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 14 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v14low
```

## Effect of high exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_high <- (results(dds, contrast = list(
  c("leachate_high_vs_control", "hpf9.leachatehigh")
)))

summary(res_9hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_high_tb <- res_9hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_high_noLFCcutoff <- res_9hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_high <- sig_9hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at high exposure level:", nrow(sig_9hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_high, "../output/09_deg_multifactorial/sig_9hpf_high.csv")
write_csv(sig_9hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_high_ashr <- lfcShrink(dds, contrast = list(c("leachate_high_vs_control", "hpf9.leachatehigh")), type = "ashr")

plotMA(res_9hpf_high)
plotMA(res_9hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_high,
    lab = rownames(res_9hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9high <- 
  ggplot(res_9hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_high[sig_9hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9high
```

## Effect of mid exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_mid <- (results(dds, contrast = list(
  c("leachate_mid_vs_control", "hpf9.leachatemid")
)))

summary(res_9hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_mid_tb <- res_9hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_mid_noLFCcutoff <- res_9hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_mid <- sig_9hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at mid exposure level:", nrow(sig_9hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_mid, "../output/09_deg_multifactorial/sig_9hpf_mid.csv")
write_csv(sig_9hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_mid_ashr <- lfcShrink(dds, contrast = list(c("leachate_mid_vs_control", "hpf9.leachatemid")), type = "ashr")

plotMA(res_9hpf_mid)
plotMA(res_9hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_mid,
    lab = rownames(res_9hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9mid <- 
  ggplot(res_9hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_mid[sig_9hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9mid
```

## Effect of low exposure at 9 hpf

```{r}
# output results of contrast
res_9hpf_low <- (results(dds, contrast = list(
  c("leachate_low_vs_control", "hpf9.leachatelow")
)))

summary(res_9hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_9hpf_low_tb <- res_9hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_9hpf_low_noLFCcutoff <- res_9hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_9hpf_low <- sig_9hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 9 hpf at low exposure level:", nrow(sig_9hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_9hpf_low, "../output/09_deg_multifactorial/sig_9hpf_low.csv")
write_csv(sig_9hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_9hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_9hpf_low_ashr <- lfcShrink(dds, contrast = list(c("leachate_low_vs_control", "hpf9.leachatelow")), type = "ashr")

plotMA(res_9hpf_low)
plotMA(res_9hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_9hpf_low,
    lab = rownames(res_9hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 9 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v9low <- 
  ggplot(res_9hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_9hpf_low[sig_9hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 9 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v9low
```

## Effect of high exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_high <- (results(dds, name = "leachate_high_vs_control")
)

summary(res_4hpf_high)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_high_tb <- res_4hpf_high %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_high_noLFCcutoff <- res_4hpf_high_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_high <- sig_4hpf_high_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at high exposure level:", nrow(sig_4hpf_high), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_high, "../output/09_deg_multifactorial/sig_4hpf_high.csv")
write_csv(sig_4hpf_high_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_high_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_high_ashr <- lfcShrink(dds, coef = 6 , type = "ashr")

plotMA(res_4hpf_high)
plotMA(res_4hpf_high_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_high,
    lab = rownames(res_4hpf_high),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of high exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4high <- 
  ggplot(res_4hpf_high) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_high[sig_4hpf_high$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of high leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4high
```

## Effect of mid exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_mid <- (results(dds, name = "leachate_mid_vs_control")
)

summary(res_4hpf_mid)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_mid_tb <- res_4hpf_mid %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_mid_noLFCcutoff <- res_4hpf_mid_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_mid <- sig_4hpf_mid_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at mid exposure level:", nrow(sig_4hpf_mid), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_mid, "../output/09_deg_multifactorial/sig_4hpf_mid.csv")
write_csv(sig_4hpf_mid_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_mid_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_mid_ashr <- lfcShrink(dds, coef = 5 , type = "ashr")

plotMA(res_4hpf_mid)
plotMA(res_4hpf_mid_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_mid,
    lab = rownames(res_4hpf_mid),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of mid exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4mid <- 
  ggplot(res_4hpf_mid) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_mid[sig_4hpf_mid$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of mid leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4mid
```

## Effect of low exposure at 4 hpf

```{r}
# output results of contrast
res_4hpf_low <- (results(dds, name = "leachate_low_vs_control", alpha=0.05)
)

summary(res_4hpf_low)
```

### Extract significant genes

```{r}
# Convert results table into tibble
res_4hpf_low_tb <- res_4hpf_low %>%
  data.frame() %>%
  rownames_to_column(var="gene") %>%
  as_tibble()

# subset that table to only keep the significant genes using our pre-defined thresholds:
sig_4hpf_low_noLFCcutoff <- res_4hpf_low_tb %>%
  filter(padj < padj.cutoff)

sig_4hpf_low <- sig_4hpf_low_noLFCcutoff %>% 
  filter(abs(log2FoldChange) > lfc.cutoff)
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low_noLFCcutoff), "(padj<", padj.cutoff, ")")
```

```{r}
paste("Number of significant DEGs 4 hpf at low exposure level:", nrow(sig_4hpf_low), "(padj<", padj.cutoff, ", log-fold change >", lfc.cutoff, ")")
```

```{r}
write_csv(sig_4hpf_low, "../output/09_deg_multifactorial/sig_4hpf_low.csv")
write_csv(sig_4hpf_low_noLFCcutoff, "../output/09_deg_multifactorial/sig_4hpf_low_noLFCcutoff.csv")
```

### Volcano plots

```{r}
# shrink lfc for visualizations
res_4hpf_low_ashr <- lfcShrink(dds, coef = 4 , type = "ashr")
plotMA(res_4hpf_low)
plotMA(res_4hpf_low_ashr)
```

```{r}
EnhancedVolcano(res_4hpf_low,
    lab = rownames(res_4hpf_low),
    x = 'log2FoldChange',
    y = 'padj',  # or 'padj' for adjusted p-values
    pCutoff = 0.05,
    FCcutoff = 0.5,
    title = 'Volcano plot',
    subtitle = 'Effect of low exposure at 4 hpf',
    pointSize = 1.0,
    labSize = 3.0,
    col = c('grey30', 'forestgreen', 'royalblue', 'red2')
)

```

```{r}
v4low <- 
  ggplot(res_4hpf_low) +
  # Plot all
  geom_point(aes(x=log2FoldChange, y=-log10(padj),color="unchanged"),
             size=.5) +
  # Overlay all significantly upregulated in red
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange > 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="upregulated"), 
             size=.5) +
  # Overlay all significantly downregulated in blue
  geom_point(data = sig_4hpf_low[sig_4hpf_low$log2FoldChange < 0, ], 
             aes(x=log2FoldChange, y=-log10(padj), color="downregulated"), 
             size=.5) +
  ggtitle("Effect of low leachate exposure at 4 hpf") +
  xlab("log2 fold change") + 
  ylab("-log10 adjusted p-value") +
  scale_x_continuous(limits = c(-2,2)) +
  scale_y_continuous(limits = c(0,100)) +
  scale_color_manual(values = c("unchanged" = "darkgrey", "upregulated" = "red", "downregulated" = "blue"),
                     labels = c("unchanged" = "Unchanged", "upregulated" = "Upregulated", "downregulated" = "Downregulated"),
                     name = NULL) +
  theme(legend.position = "top",
        plot.title = element_text(size = rel(1.5), hjust = 0.5),
        axis.title = element_text(size = rel(1.25)))

v4low
```

# Summary

## Facet volcano plot

```{r}
# First, strip labels from all individual ggplot objects
# (You can do this in a loop if the plots are in a list)

remove_labels <- function(p) {
  p +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      legend.position = "none",
      plot.title = element_text(size = 8)  # adjust title size here
    )
}

# Example: if your plots are named v4control, v4low, etc.
plots <- list( 
   v4low, v9low, v14low,
   v4mid, v9mid, v14mid,
   v4high, v9high, v14high
)

plots_clean <- lapply(plots, remove_labels)

# Now plot in 3 rows × 3 columns
volcano_facet <- plot_grid(
  plotlist = plots_clean,
  nrow = 3,
  ncol = 3,
  labels = NULL,
  label_size = 0,
  align = "hv"
)

volcano_facet
```

```{r}
ggsave("../output/09_deg_multifactorial/volcano_facet.png", volcano_facet, width = 16, height = 12, dpi = 600)
```

::: callout-warning
Why do volcano plots not look like volcanos!?

Batch Effects or Low Biological Signal If your experimental design has confounding factors (e.g., batch effects), or the leachate treatment has low impact, then few genes will show strong changes.

Extreme p-values Your plot shows points with -log10(padj) values above 20--30. That means padj is very close to 0, which could indicate: - Very large sample size - Very consistent effect - Or something off with the dispersion estimate or p-value adjustment (e.g., if some genes had zero variance across groups)
:::

# Next steps

Re-run the model without the interaction term. The likelihood ratio test (LRT) shows that the interaction (hpf:leachate) is not significant! We can now test main effects of leachate across stages (i.e., an overall leachate effect) using `~ hpf + leachate`

## Exlpore cooks distance

> DESeq2 relies on the negative binomial distribution to make estimates and perform statistical inference on differences. While the negative binomial is versatile in having a mean and dispersion parameter, extreme counts in individual samples might not fit well to the negative binomial. For this reason, we perform automatic detection of count outliers. We use Cook's distance, which is a measure of how much the fitted coefficients would change if an individual sample were removed (Cook 1977). For more on the implementation of Cook's distance see the manual page for the *results* function. Below we plot the maximum value of Cook's distance for each row over the rank of the test statistic to justify its use as a filtering criterion.

## Explore RUVSeq
