---
title: "Plotting Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

So we have 9 different 'significant gene' lists. 
Each list indicates genes in which leachate had a significant effect.

## Inputs
LRT:
- `sig_int`, 13 DEGs
- `sig_joint`. 44 DEGs

Interaction effects 
- cleavage_low.int , 0 DEGs
- cleavage_mid.int , 0 DEGs
- cleavage_high.int, 0 DEGs

- `sig_prawnchip_low.int`, 7 DEGs 
- prawnchip_mid.int , 0 DEGs
- prawnchip_high.int, 0 DEGs

- `sig_earlygastrula_low.int`, 8 DEGs
- earlygastrula_mid.int, 0 DEGs
- `sig_earlygastrula_high.int`, 1 DEGs 

Contrasts
- `sig_cleavage_low.vs.control`, 34 DEGs 
- `sig_cleavage_mid.vs.control`, 58 DEGs 
- `sig_cleavage_high.vs.control`, 7 DEGs 

- prawnchip_low.vs.control, 0 DEGs
- prawnchip_mid.vs.control, 0 DEGs
- prawnchip_high.vs.control, 0 DEGs

- earlygastrula_low.vs.control, 0 DEGs
- earlygastrula_mid.vs.control, 0 DEGs
- `sig_earlygastrula_high.vs.control`, 27 DEGs

- `sig_summary.csv` 

We also have VST count matrices generated via DESeq2 of the full gene set and the combined DEGs
- `vst_mat_df.csv`
- `vst_mat_130_df.csv`

## Outputs
- Gene count plots
- PCA of significant genes
- Heatmap(s) showing significant genes across DEG lists
- Heatmap of genes for each cluster?
- Plots of log2FC for each contrast
- Plots of log2FC for each cluster
- VennDiagram of gene set overlaps

# Setup

## Load packages

```{r}
# Load packages
library(viridis)
library(tidyverse)
library(readr)
library(DEGreport)
library(ashr)
library(plotly)
library(vsn)
library(pheatmap)
library(ComplexUpset)
library(ggplot2)
library(ggplotify)
library(patchwork)   # for wrap_plots(...)
library(cowplot)     # for plot_grid & get_legend
library(dplyr)
library(grid) 
library(purrr)
library(forcats)
library(ggforce)   # for facet_wrap_paginate
library(ggnewscale) # to display 2 padj scales on same plot
library(colorspace)
```

## Load inputs
Everything from output/06_deg/spawnnight
```{r}
path <- "../../output/06_deg/spawnnight/"
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
nm    <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}
```

```{r}
gene_cluster_data <- read_csv("../../output/06_deg/cluster/gene_cluster_zscores.csv")
```

Metadata
```{r}
metadata_or <- read.csv("../../metadata/metadata_or.csv")
str(metadata_or)
```
```{r}
duplicated(norm_130_df$gene_id)
```

## Long data for plotting
```{r}
# convert -> long tidy dataframe
norm_130_long <- as.data.frame(norm_130_df) %>%
  pivot_longer(
    cols = -gene_id,
    names_to = "sample_id",
    values_to = "norm.expression"
  )

# 3. join sample metadata
norm_130_long <- norm_130_long %>% left_join(metadata_or, by = "sample_id")
```

How many samples are in each treatment for each significant gene?

This bit of code lets us see how many samples are contributing to the mean expression of a gene broken down by group (aka condition ... leachate x stage)

```{r}
sample_counts <- norm_130_long %>% 
  dplyr::count(gene_id, group) %>%   # count rows per gene_id + group
  pivot_wider(names_from = group,              # make each group a column
              values_from = n, 
              values_fill = 0)                 # fill missing group-gene combos with 0

summary(sample_counts)
```

::: callout-note
At least 4 samples represent each group per significant gene. Most groups are represented by 5 samples... some have 4, 6, and 1 group has 7 samples. This is important to check because we want to be able to compare the groups, but if they have vastly different sample sizes that could be problematic. This confirms that the sample sizes are reasonably similar for comparisons.
:::

# Fold Change 

Here I want to visualize how many DEGs there are in each contrast, and their magnitude and direction.  

The interaction (`int`) and joint (`joint`) lists have valid p values but not valid log2FoldChange values... those are taken from a contrast... therefore visualizing an LRT test log2FoldChange isn't useful. We will try again and exlude them... 

The main takeaway for this plot is that the cleavage stage and low dose groups have more DEGs, plus the early gastrula high dose group... not necessarily comparing the exact genes from one group to another ... though that is interesting too.... it just seems that the data is too sparse... need more data ink! This plot would be more interesting if it included what the genes do... but the problem with 1 gene to many GO terms make that difficult.

Here we add rows that will be populated with 'NA' for the categories that had 0 DEGs. It's important to show these to contrast the groups that did pop DEGs!
```{r}
sig_summary <-  sig_summary %>% 
  add_row(`list` = c("cleavage_low.int" ,
                    "cleavage_mid.int" , 
                    "cleavage_high.int",
                    "prawnchip_mid.int", 
                    "prawnchip_high.int",
                    "earlygastrula_mid.int",
                    "prawnchip_low.vs.control", 
                    'prawnchip_mid.vs.control', 
                    'prawnchip_high.vs.control', 
                    'earlygastrula_low.vs.control', 
                    'earlygastrula_mid.vs.control'
                    ))
```

Ok here we're going to only look at the 9 contrasts (sig effect of leachate on gene expression of an embryo at a specific stage and dose)
```{r}
sig_FC <- sig_summary %>% 
  dplyr::filter(!list %in% c("int", "joint", 
                             "cleavage_low.int",
                             "cleavage_mid.int" , 
                             "cleavage_high.int",
                             "prawnchip_mid.int",
                             "prawnchip_low.int",
                             "prawnchip_high.int",
                             "earlygastrula_low.int",
                             "earlygastrula_mid.int",
                             "earlygastrula_high.int")) %>%
  mutate(list = factor(list, levels = c("cleavage_high.vs.control",
                                        "prawnchip_high.vs.control",
                                        "earlygastrula_high.vs.control",
                                        "cleavage_mid.vs.control",
                                        "prawnchip_mid.vs.control",
                                        "earlygastrula_mid.vs.control",
                                        "cleavage_low.vs.control",
                                        "prawnchip_low.vs.control",
                                        "earlygastrula_low.vs.control"))) %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_wrap(~list, nrow = 3)+
  geom_point(size=1)+
  scale_colour_gradientn(name="P(adj) main leachate effect", colours = colorspace::heat_hcl(7))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id), linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") + # Add vertical line
  theme_classic() +
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(color = "grey60"),
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size=12, color = "grey60"),
        strip.text = element_text(size=6),
        strip.background = element_blank(),
        axis.title.x = element_text(size = 14, color = "grey60"),
        plot.title = element_text(size = 16, color = "grey60"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC
```

And here we're going to look at the interaction effects
```{r}
sig_FC_int <- sig_summary %>% 
  dplyr::filter(!list %in% c("int", "joint", 
                             "cleavage_high.vs.control",
                             "prawnchip_high.vs.control",
                             "earlygastrula_high.vs.control",
                             "cleavage_mid.vs.control",
                             "prawnchip_mid.vs.control",
                             "earlygastrula_mid.vs.control",
                             "cleavage_low.vs.control",
                             "prawnchip_low.vs.control",
                             "earlygastrula_low.vs.control")) %>%
  mutate(list = factor(list, levels = c("cleavage_high.int",
                                        "prawnchip_high.int",
                                        "earlygastrula_high.int",
                                        "cleavage_mid.int", 
                                        "prawnchip_mid.int",
                                        "earlygastrula_mid.int",
                                        "cleavage_low.int",
                                        "prawnchip_low.int",
                                        "earlygastrula_low.int"))) %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_wrap(~list, nrow = 3)+
  geom_point(size=1)+
  scale_colour_gradientn(name="P(adj) stage x leachate interaction", colours = colorspace::sequential_hcl(7, palette = "Viridis"))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id), linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") + # Add vertical line
  theme_classic() +
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(color = "grey60"),
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size=12, color = "grey60"),
        strip.text = element_text(size=6),
        strip.background = element_blank(),
        axis.title.x = element_text(size = 14, color = "grey60"),
        plot.title = element_text(size = 16, color = "grey60"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC_int

```
Cool! Next let's try to combine the info about main stage specific contrasts and interactions into one plot (thereby concentrating more information in less space!)

We can use `ggnewscale` to map two scales to the same facet plot. 
To do this we'll need a distinct color palette for each different adjusted pvalue scale.
Checkout colorspace palette options: https://colorspace.r-forge.r-project.org/articles/hcl_palettes.html
```{r}
colorspace::hcl_palettes(type = "sequential")
```
Make another column by which we want to facet called plot_groups
```{r}
groups <- c(
  "cleavage_low", "cleavage_mid", "cleavage_high",
  "prawnchip_low", "prawnchip_mid", "prawnchip_high",
  "earlygastrula_low", "earlygastrula_mid", "earlygastrula_high"
)

sig_summary <- sig_summary %>%
  mutate(
    plot_group = str_extract(
      list,
      paste0("^(", paste(groups, collapse = "|"), ")")
    )
  )
```

Make a dataframe for plotting that breaks down the row into main or interaction contrast categories
```{r}
plot_levels <- c(
  "cleavage_high", "prawnchip_high", "earlygastrula_high",
  "cleavage_mid",  "prawnchip_mid",  "earlygastrula_mid",
  "cleavage_low",  "prawnchip_low",  "earlygastrula_low"
)

fc_plot <- sig_summary %>%
  filter(!list %in% c("int", "joint")) %>%  # keep your current filter
  mutate(
    effect_type = case_when(
      str_detect(list, "\\.int$")         ~ "interaction",
      str_detect(list, "\\.vs\\.control") ~ "main",
      TRUE                                ~ "other"
    )
  ) %>%
  filter(effect_type %in% c("main", "interaction")) %>%
  arrange(abs(log2FoldChange)) %>%
  mutate(
    gene_id_fac = factor(gene_id, levels = unique(gene_id)),
    plot_group  = factor(plot_group, levels = plot_levels)
  ) %>% 
  left_join(gene_clusters, by = "gene_id") %>% 
  mutate(cluster = factor(cluster, levels = c(3, 2, 1), labels = c("cleavage", "prawnchip", "earlygastrula")))
```

Make the plot that combines main and interaction effects 
```{r, fig.height=8, fig.width=11}
p_FC <- ggplot() +
  # ===== MAIN EFFECT: segments + points, 1st colour scale =====
  geom_segment(
    data = filter(fc_plot, effect_type == "main"),
    aes(
      x    = 0,
      xend = log2FoldChange,
      y    = gene_id_fac,
      yend = gene_id_fac,
      colour = padj
    ),
    linewidth = 1
  ) +
  geom_point(
    data = filter(sig_plot, effect_type == "main"),
    aes(
      x = log2FoldChange,
      y = gene_id_fac,
      colour = padj
    ),
    size = 1
  ) +
  scale_colour_gradientn(
    name    = "P(adj) stage specific leachate effect",
    colours = colorspace::sequential_hcl(7, palette = "TealGrn"),
    guide   = guide_colourbar(order = 1,
                            direction = "horizontal",
                            title.position = "top",
                            title.hjust = 0.5)
  ) +

  ggnewscale::new_scale_colour() +   # << reset colour scale here

  # ===== INTERACTION: segments + points, 2nd colour scale =====
  geom_segment(
    data = filter(sig_plot, effect_type == "interaction"),
    aes(
      x    = 0,
      xend = log2FoldChange,
      y    = gene_id_fac,
      yend = gene_id_fac,
      colour = padj
    ),
    linewidth = 1
  ) +
  geom_point(
    data = filter(sig_plot, effect_type == "interaction"),
    aes(
      x = log2FoldChange,
      y = gene_id_fac,
      colour = padj
    ),
    size = 1
  ) +
  scale_colour_gradientn(
    name    = "P(adj) stage × leachate interaction effect",
    colours = colorspace::sequential_hcl(7, palette = "Peach"),
    guide   = guide_colourbar(order = 2,
                            direction = "horizontal",
                            title.position = "top",
                            title.hjust = 0.5)
  ) +

  # ===== facets + theme =====
  facet_wrap(~ plot_group, nrow = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  theme_classic() +
  theme(
    legend.text      = element_text(size = 8, color = "grey60"),
    legend.key.width = grid::unit(1.25, "cm"),  # increase until labels stop colliding
    legend.title     = element_text(size = 10, color = "grey30"),
    legend.position  = "top",
    axis.text.y      = element_blank(),
    axis.title.y     = element_blank(),
    axis.line.y      = element_blank(),
    axis.line.x      = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.text.x      = element_text(size = 12, color = "grey60"),
    strip.background = element_blank(),
    axis.title.x     = element_text(size = 14, color = "grey30"),
    plot.title       = element_text(size = 16, color = "grey20", hjust = 0.5)
  ) +
  ggtitle("Log2 Fold Change of significant DEGs")+
  labs(x = "Log2 Fold Change")

p_FC

ggsave("../../output/06_deg/figs/log2FC.png", width = 11, height = 8)
```
I like this plot! This will be what I tweak for the final manuscript. I'd like to add symbols for the embryonic stages and leachate dose in place of the facet titles. If possible, I'd like to link the genes to enriched function on the y axis... but gene semantic similarity doesn't necessarily map to log2Fold change or significance... 

Should I include the number of up and down regulated DEGs in each group?
How many genes are shared among these groups with log2FC values?

```{r}
dplyr::n_distinct(fc_plot$gene_id, na.rm = TRUE)
```
Ok so ... 8 genes are present in the joint/int list that are not described by the contrasts... why is that? 

To re-iterate, the main point of this plot is that cleavage experienced the most differential expression due to leachate, and low dose also shows a signal across stages... then there is this senescense of response in the prawnchip stage (a narrowing?), and another signal in the longest exposure and highest dose (early gatrula, high).


# Gene Clusters

```{r}
gene_cluster_data <- read_csv("../../output/06_deg/cluster/gene_cluster_zscores.csv")
str(gene_cluster_data)
```


```{r}
leachate_colors <- c(control = "#BFE8D9", #cyan100
                     low = "#A699D8", #purple300
                     mid = "#5E409D", #purple600
                     high = "#39172B" #magenta900
                     )

stage_colors <- c(cleavage = "#AF3029", #red600
                  prawnchip = "#AD8301", #yellow600
                  earlygastrula = "#71320D" #orange600
                  )
```

```{r}
# reverse gene cluster order to plot from cleavage to eg
gene_cluster_data <- gene_cluster_data %>%
  mutate(cluster = factor(cluster, levels = rev(sort(unique(cluster)))))

facet_labs <- gene_cluster_data %>%
  summarise(n_genes = n_distinct(genes), .by = cluster) %>%        # counts unique genes per cluster
  mutate(lab = sprintf("Group: %s - genes: %d", cluster, n_genes)) %>%
  dplyr::select(cluster, lab) %>%
  deframe()                                                        # named vector: names = cluster, values = lab
```

```{r, warning=FALSE, fig.height=4, fig.width = 8}

# leachate as x plot
pl <- ggplot(gene_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high")),
                stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula"))),
       aes(x = leachate, y = zscore, color = stage)) +
  geom_line(aes(group = interaction(genes, stage)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = stage), se = FALSE) + # trend
  facet_wrap(~ cluster, nrow = 1) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "stage") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = stage_colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_blank(),
  strip.background = element_blank())

# stage as x plot
ps <- ggplot(gene_cluster_data %>%
         mutate(leachate = factor(leachate, levels = c("control","low","mid","high")),
                stage = factor(stage, levels = c("cleavage", "prawnchip", "earlygastrula"))),
       aes(x = stage, y = zscore, color = leachate)) +
  geom_line(aes(group = interaction(genes, leachate)), alpha = 0.1, linewidth = 0.4) + # per-gene(thin & transparent)
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.2, size = 1.2) +
  geom_boxplot(alpha = 0.10, outlier.shape = NA) +
  geom_smooth(aes(group = leachate), se = FALSE) + #trend pattern 
  facet_wrap(~ cluster, nrow = 1, labeller = labeller(cluster = facet_labs)) +
  labs(x = NULL, y = "Z-score of gene abundance", color = "leachate") +
  theme_minimal(base_size = 12) +
  scale_color_manual(values = leachate_colors)+
  scale_y_continuous(breaks = function(x) pretty(x, n = 3)) +
  theme(panel.grid = element_blank(),
  panel.grid.major.y = element_line(color = "grey80", linewidth = 0.3),   # keep major horizontal
  panel.grid.minor.y = element_blank(),  # drop minor horizontal
  panel.grid.major.x = element_blank(),  # drop vertical
  panel.grid.minor.x = element_blank(),
  strip.text = element_text(face = "bold"))


# combine with patchwork
pc <- ps / pl

pc

```

Can I annotate the DEG FC plot with cluster info? 

```{r, fig.width=6, fig.height=7}
pfc <- ggplot() +
  # ===== MAIN EFFECT: segments + points, 1st colour scale =====
  geom_segment(
    data = filter(fc_plot),
    aes(
      x    = 0,
      xend = log2FoldChange,
      y    = gene_id_fac,
      yend = gene_id_fac,
      colour = cluster
    ),
    linewidth = 0.9
  ) +
  scale_colour_discrete(
    name  = "Gene cluster pattern",
    type  = stage_colors,
    guide = guide_legend(order = 2)   # << put this legend last
  ) +
  ggnewscale::new_scale_colour() +

  # ===== MAIN EFFECT PADJ =====
  geom_point(
    data = filter(fc_plot, effect_type == "main"),
    aes(
      x = log2FoldChange,
      y = gene_id_fac,
      colour = padj
    ),
    size = 1.5
  ) +
  scale_colour_gradientn(
    name    = "P(adj) stage specific leachate effect",
    colours = colorspace::sequential_hcl(7, palette = "TealGrn"),
    guide   = guide_colourbar(
      order         = 1,                 # first row
      direction     = "horizontal",
      title.position = "top",
      title.hjust   = 0.5
    )
  ) +

  ggnewscale::new_scale_colour() +

  # ===== INTERACTION PADJ =====
  geom_point(
    data = filter(fc_plot, effect_type == "interaction"),
    aes(
      x = log2FoldChange,
      y = gene_id_fac,
      colour = padj
    ), size = 1.5) +
  scale_colour_gradientn(
    name    = "P(adj) stage × leachate interaction effect",
    colours = colorspace::sequential_hcl(7, palette = "OrYel"),
    guide   = guide_colourbar(
      order         = 1,                 # second in top row
      direction     = "horizontal",
      title.position = "top",
      title.hjust   = 0.5
    )
  ) +

  # ===== facets + theme =====
  facet_wrap(~ plot_group, nrow = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +
  theme_classic() +
  theme(
    #legend.box       = "vertical",  # << stack legend rows
    legend.position  = "top",
    legend.text      = element_text(size = 8, color = "grey60"),
    legend.key.width = grid::unit(.75, "cm"),
    legend.title     = element_text(size = 10, color = "grey30"),
    axis.text.y      = element_blank(),
    axis.title.y     = element_blank(),
    axis.line.y      = element_blank(),
    axis.line.x      = element_blank(),
    axis.ticks.y     = element_blank(),
    axis.ticks.x     = element_blank(),
    axis.text.x      = element_text(size = 12, color = "grey60"),
    strip.background = element_blank(),
    axis.title.x     = element_text(size = 14, color = "grey30"),
    plot.title       = element_text(size = 16, color = "grey20", hjust = 0.5)
  ) +
  ggtitle("Log2 Fold Change of significant DEGs") +
  labs(x = "Log2 Fold Change")

pfc
```
combine with cowplot
```{r, fig.width = 8, fig.height = 12}
pc / pfc
```


# Summary & Next Steps

We can perform functional enrichment on these clusters now!
