---
title: "Plotting Differential Gene Expression"
subtitle: "Exploring the interactions between embryonic stage and PVC leachate in Montipora capitata using DESeq2"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

So we have 9 different 'significant gene' lists. 
Each list indicates genes in which leachate had a significant effect.

## Inputs
LRT:
- `sig_int`, 13 DEGs
- `sig_joint`. 44 DEGs

Interaction effects 
- cleavage_low.int , 0 DEGs
- cleavage_mid.int , 0 DEGs
- cleavage_high.int, 0 DEGs

- `sig_prawnchip_low.int`, 7 DEGs 
- prawnchip_mid.int , 0 DEGs
- prawnchip_high.int, 0 DEGs

- `sig_earlygastrula_low.int`, 8 DEGs
- earlygastrula_mid.int, 0 DEGs
- `sig_earlygastrula_high.int`, 1 DEGs 

Contrasts
- `sig_cleavage_low.vs.control`, 34 DEGs 
- `sig_cleavage_mid.vs.control`, 58 DEGs 
- `sig_cleavage_high.vs.control`, 7 DEGs 

- prawnchip_low.vs.control, 0 DEGs
- prawnchip_mid.vs.control, 0 DEGs
- prawnchip_high.vs.control, 0 DEGs

- earlygastrula_low.vs.control, 0 DEGs
- earlygastrula_mid.vs.control, 0 DEGs
- `sig_earlygastrula_high.vs.control`, 27 DEGs

- `sig_summary.csv` 

We also have VST count matrices generated via DESeq2 of the full gene set and the combined DEGs
- `vst_mat_df.csv`
- `vst_mat_130_df.csv`

## Outputs
- Gene count plots
- PCA of significant genes
- Heatmap(s) showing significant genes across DEG lists
- Heatmap of genes for each cluster?
- Plots of log2FC for each contrast
- Plots of log2FC for each cluster
- VennDiagram of gene set overlaps

# Setup

## Load packages

```{r}
# Load packages
library(viridis)
library(tidyverse)
library(readr)
library(DEGreport)
library(ashr)
library(plotly)
library(vsn)
library(pheatmap)
library(ComplexUpset)
library(ggplot2)
library(ggplotify)
library(patchwork)   # for wrap_plots(...)
library(cowplot)     # for plot_grid & get_legend
library(dplyr)
library(grid) 
library(purrr)
library(forcats)
library(ggforce)   # for facet_wrap_paginate
```

## Load inputs
Everything from output/06_deg/spawnnight
```{r}
path <- "../../output/06_deg/spawnnight/"
files <- list.files(path, pattern = "\\.csv$", full.names = TRUE)
nm    <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}
```

```{r}
gene_cluster_data <- read_csv("../../output/06_deg/cluster/gene_cluster_zscores.csv")
```

Metadata
```{r}
metadata_or <- read.csv("../../metadata/metadata_or.csv")
str(metadata_or)
```

## Long data for plotting
```{r}
# convert -> long tidy dataframe
norm_130_long <- as.data.frame(norm_130_df) %>%
  rownames_to_column("gene_id") 



%>%
  pivot_longer(
    cols = -gene_id,
    names_to = "sample_id",
    values_to = "expression"
  )

# 3. join sample metadata
norm_130_long <- norm_count_long %>% left_join(metadata_or, by = "sample_id")
```

How many samples are in each treatment for each significant gene?

This bit of code lets us see how many samples are contributing to the mean expression of a gene broken down by group (aka condition ... leachate x stage)

```{r}
sample_counts <- norm_count_long %>% 
  dplyr::count(gene_id, group) %>%   # count rows per gene_id + group
  pivot_wider(names_from = group,              # make each group a column
              values_from = n, 
              values_fill = 0)                 # fill missing group-gene combos with 0

summary(sample_counts)
```

::: callout-note
At least 4 samples represent each group per significant gene. Most groups are represented by 5 samples... some have 4, 6, and 1 group has 7 samples. This is important to check because we want to be able to compare the groups, but if they have vastly different sample sizes that could be problematic. This confirms that the sample sizes are reasonably similar for comparisons.
:::

# Fold Change 

Here I want to visualize how many DEGs there are in each contrast, and their magnitude and direction. 
```{r}
sig_FC <- sig_summary %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_grid(~list, scales="free_y")+
  geom_point(size=1)+
  scale_colour_gradientn(name="P(adj)", colours = colorspace::heat_hcl(7))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id),linewidth=1) +
  geom_vline(xintercept = 0, linetype = "dashed") + # Add vertical line
  theme_classic() +
  ylab("Gene")+
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.text.x = element_text(size=12),
        strip.text=element_text(size=14),
        axis.title.y = element_text(size=14, face="bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC


```
The interaction (`int`) and joint (`joint`) lists have valid p values but not valid log2FoldChange values... those are taken from a contrast... therefore visualizing an LRT test log2FoldChange isn't useful. We will try again and exlude them... 
```{r}
sig_FC <- sig_summary %>% 
  dplyr::filter(!list %in% c("int", "joint")) %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_grid(~list, scales="free_y")+
  geom_point(size=2)+
  scale_colour_gradientn(name="P(adj)", colours = colorspace::heat_hcl(7))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id), linewidth=1.5) +
  geom_vline(xintercept = 0, linetype = "dashed") + # Add vertical line
  theme_classic() +
  ylab("Gene")+
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.text.x = element_text(size=12),
        strip.text=element_text(size=6),
        axis.title.y = element_text(size=14, face="bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC
```
I think i want this plot to show how many and in which direction... not necessarily comparing the exact genes from one group to another ... though that is interesting too.... it just seems that the data is too sparse... need more data ink! This plot would be more interesting if it included what the genes do... but the problem with 1 gene to many GO terms make that difficult... Should I show all the contrasts that have no DEGs? Should I show the interaction effects separately from the contrast effects?

```{r}
sig_summary <-  sig_summary %>% 
  add_row(`list` = c("cleavage_low.int" ,
                    "cleavage_mid.int" , 
                    "cleavage_high.int",
                    "prawnchip_mid.int", 
                    "prawnchip_high.int",
                    "earlygastrula_mid.int",
                    "prawnchip_low.vs.control", 
                    'prawnchip_mid.vs.control', 
                    'prawnchip_high.vs.control', 
                    'earlygastrula_low.vs.control', 
                    'earlygastrula_mid.vs.control'
                    ))
```

Ok here we're going to only look at the 9 contrasts (sig effect of leachate on gene expression of an embryo at a specific stage and dose)

```{r}
sig_FC <- sig_summary %>% 
  dplyr::filter(!list %in% c("int", "joint", 
                             "cleavage_low.int",
                             "cleavage_mid.int" , 
                             "cleavage_high.int",
                             "prawnchip_mid.int",
                             "prawnchip_low.int",
                             "prawnchip_high.int",
                             "earlygastrula_low.int",
                             "earlygastrula_mid.int",
                             "earlygastrula_high.int")) %>%
  mutate(list = factor(list, levels = c("earlygastrula_low.vs.control",
                                        "earlygastrula_mid.vs.control",
                                        "earlygastrula_high.vs.control", 
                                        "prawnchip_low.vs.control", 
                                        "prawnchip_mid.vs.control", 
                                        "prawnchip_high.vs.control", 
                                        "cleavage_low.vs.control", 
                                        "cleavage_mid.vs.control",
                                        "cleavage_high.vs.control"))) %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_wrap(~list, nrow = 3)+
  geom_point(size=1)+
  scale_colour_gradientn(name="P(adj)", colours = colorspace::heat_hcl(7))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id),linewidth=1) +
  geom_vline(xintercept = 0, linetype = "dashed") + # Add vertical line
  theme_classic() +
  ylab("Gene")+
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.text.x = element_text(size=12),
        strip.text=element_text(size=6),
        axis.title.y = element_text(size=14, face="bold"),
        axis.title.x = element_text(size = 14, face = "bold"),
        plot.title = element_text(size = 16, face = "bold"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC
```
```{r}
sig_FC <- sig_summary %>% 
  dplyr::filter(!list %in% c("int", "joint", 
                             "cleavage_low.int",
                             "cleavage_mid.int" , 
                             "cleavage_high.int",
                             "prawnchip_mid.int",
                             "prawnchip_low.int",
                             "prawnchip_high.int",
                             "earlygastrula_low.int",
                             "earlygastrula_mid.int",
                             "earlygastrula_high.int")) %>%
  mutate(list = factor(list, levels = c("cleavage_high.vs.control",
                                        "prawnchip_high.vs.control",
                                        "earlygastrula_high.vs.control",
                                        "cleavage_mid.vs.control",
                                        "prawnchip_mid.vs.control",
                                        "earlygastrula_mid.vs.control",
                                        "cleavage_low.vs.control",
                                        "prawnchip_low.vs.control",
                                        "earlygastrula_low.vs.control"))) %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_wrap(~list, nrow = 3)+
  geom_point(size=1)+
  scale_colour_gradientn(name="P(adj) main leachate effect", colours = colorspace::heat_hcl(7))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id), linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") + # Add vertical line
  theme_classic() +
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(color = "grey60"),
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size=12, color = "grey60"),
        strip.text = element_text(size=6),
        strip.background = element_blank(),
        axis.title.x = element_text(size = 14, color = "grey60"),
        plot.title = element_text(size = 16, color = "grey60"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC
```


```{r}
sif_FC_int <- sig_summary %>% 
  dplyr::filter(!list %in% c("int", "joint", 
                             "cleavage_high.vs.control",
                             "prawnchip_high.vs.control",
                             "earlygastrula_high.vs.control",
                             "cleavage_mid.vs.control",
                             "prawnchip_mid.vs.control",
                             "earlygastrula_mid.vs.control",
                             "cleavage_low.vs.control",
                             "prawnchip_low.vs.control",
                             "earlygastrula_low.vs.control")) %>%
  mutate(list = factor(list, levels = c("cleavage_high.int",
                                        "prawnchip_high.int",
                                        "earlygastrula_high.int",
                                        "cleavage_mid.int", 
                                        "prawnchip_mid.int",
                                        "earlygastrula_mid.int",
                                        "cleavage_low.int",
                                        "prawnchip_low.int",
                                        "earlygastrula_low.int"))) %>% 
  dplyr::arrange(log2FoldChange) %>%
  ggplot(aes(x=log2FoldChange, y=reorder(gene_id, -log2FoldChange), colour=padj)) +
  facet_wrap(~list, nrow = 3)+
  geom_point(size=1)+
  scale_colour_gradientn(name="P(adj) stage x leachate interaction", colours = colorspace::sequential_hcl(7, palette = "Viridis"))+
  geom_segment(aes(x=0, xend=log2FoldChange, y=gene_id, yend=gene_id), linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") + # Add vertical line
  theme_classic() +
  theme(legend.position = "right",
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.line.y = element_blank(),
        axis.line.x = element_line(color = "grey60"),
        axis.ticks.y = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size=12, color = "grey60"),
        strip.text = element_text(size=6),
        strip.background = element_blank(),
        axis.title.x = element_text(size = 14, color = "grey60"),
        plot.title = element_text(size = 16, color = "grey60"))+
  ggtitle('Log2 Fold Change of significant DEGs')

sig_FC_int

```
```{r}
colorspace::hcl_palettes(type = "sequential")
```

#  Gene count plot function
```{r}
plot_sig_genes <- function(norm_count_long, 
                           sig,                      # vector of gene_ids OR df with gene_id (+ optional padj)
                           gene_col = "gene_id",
                           padj_col = NULL,         # set to the padj column name if available (e.g., "padj")
                           title = NULL,
                           subtitle = NULL,
                           ci_level = 0.95) {

  # --- Normalize the significant list to a tibble with gene_id and optional padj ---
  if (is.vector(sig)) {
    sig_df <- tibble(gene_id = as.character(sig), padj = NA_real_)
  } else if (is.data.frame(sig)) {
    stopifnot(gene_col %in% names(sig))
    sig_df <- sig %>%
      rename(gene_id = !!gene_col) %>%
      mutate(gene_id = as.character(gene_id))
    if (!is.null(padj_col) && padj_col %in% names(sig_df)) {
      sig_df <- sig_df %>% 
        rename(padj = !!padj_col) %>%
        group_by(gene_id) %>%
        summarise(padj = suppressWarnings(min(padj, na.rm = TRUE)), .groups = "drop")
      sig_df$padj[!is.finite(sig_df$padj)] <- NA_real_  # handle all-NA
    } else {
      sig_df <- sig_df %>% mutate(padj = NA_real_)
    }
    sig_df <- sig_df %>% distinct(gene_id, .keep_all = TRUE)
  } else {
    stop("`sig` must be a character vector of gene_ids or a data.frame.")
  }

  # --- Keep only rows present in norm_count_long and prep labels/factors ---
  plot_df <- norm_count_long %>%
    semi_join(sig_df, by = "gene_id") %>%
    left_join(dplyr::select(sig_df, gene_id, padj), by = "gene_id") %>%
    mutate(
      gene_label = ifelse(
        is.na(padj), gene_id,
        paste0(gene_id, "\nFDR: ", formatC(padj, format = "g", digits = 3))
      ),
      leachate = forcats::fct_relevel(as.factor(leachate), c("control","low","mid","high"))
    )

  # --- Summary stats per facet/group/level (mean ± 95% CI) ---
  alpha <- 1 - ci_level
  sum_df <- plot_df %>%
    dplyr::group_by(gene_id, gene_label, stage, leachate) %>%
    dplyr::summarise(
      n   = dplyr::n(),
      mean = mean(expression, na.rm = TRUE),
      sd   = sd(expression,  na.rm = TRUE),
      se   = sd / sqrt(n),
      tcrit = qt(1 - alpha/2, pmax(n - 1, 1)),
      ci   = tcrit * se,
      .groups = "drop"
    ) %>%
    mutate(
      ymin = mean - ci,
      ymax = mean + ci
    )

  # --- Fallback title/subtitle ---
  if (is.null(title)) {
    title <- paste0("Significant genes (n = ", dplyr::n_distinct(plot_df$gene_id), ")")
  }
  if (is.null(subtitle)) {
    subtitle <- paste0("Group means ± ", round(ci_level*100), "% CI per stage; points are raw observations")
  }

  # --- Plot: raw points + summary line/points + CI error bars ---
  ggplot(plot_df, aes(x = leachate, y = expression, color = stage)) +
    labs(
      title = title,
      subtitle = subtitle,
      y = "Normalized gene counts",
      x = "PVC leachate concentration (milligrams of microplastic/ Liter of seawater)"
    ) +
    # raw points
    geom_jitter(width = 0.12, alpha = 0.6, size = 1, show.legend = FALSE) +
    # summary error bars + mean points + connecting line
    geom_errorbar(
      data = sum_df,
      aes(y = mean, ymin = ymin, ymax = ymax, group = stage),
      width = 0.08,
      linewidth = 0.5,
      position = position_dodge2(width = 0.0, preserve = "single"),
      show.legend = FALSE
    ) +
    geom_point(
      data = sum_df,
      aes(y = mean, group = stage),
      size = 1.8,
      stroke = 0.4,
      show.legend = TRUE
    ) +
    geom_line(
      data = sum_df,
      aes(y = mean, group = stage),
      linewidth = 0.6,
      show.legend = FALSE
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(size = 12),
      plot.subtitle = element_text(size = 8),
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank(),
      strip.text = element_text(size = 7, lineheight = 0.9, margin = margin(2,2,2,2))
    ) +
    scale_x_discrete(
      expand = expansion(mult = c(0.1, 0.1)),
      labels = c(control = "Control", low = "Low (0.01)", mid = "Mid (0.1)", high = "High (1)")
    ) +
    facet_wrap(~ gene_label, scales = "free_y")
}

```


# Presence/absence matrix
```{r}
presence_mat <- sig_summary %>%
  distinct(list, gene_id) %>%
  mutate(value = 1L) %>%
  tidyr::pivot_wider(names_from = list, values_from = value, values_fill = 0)
```


```{r}
plot_df <- tidyr::pivot_longer(presence_mat, -gene_id,
                               names_to = "list", values_to = "present")

ggplot(plot_df, aes(x = list, y = gene_id, fill = factor(present))) +
  geom_tile() +
  scale_fill_manual(values = c("0" = "white", "1" = "black"), guide = "none") +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# helper to standardize any DESeqResults into a tidy df
as_tbl <- function(res, name){
  as.data.frame(res) %>%
    rownames_to_column("gene_id") %>%
    transmute(
      gene_id,
      result_name = name,
      log2FC      = log2FoldChange,
      pval        = pvalue,
      padj        = padj
    )
}
```

```{r}
# EXAMPLE: replace the ... with your actual DESeqResults objects
res_list <- list(
  # Wald main effects at reference stage (cleavage)
  leachate_low  = results(dds_wald, name = "leachatelow"),
  leachate_mid  = results(dds_wald, name = "leachatemid"),
  leachate_high = results(dds_wald, name = "leachatehigh"),

  # Stage-specific low/mid/high (additive + interaction)
  prawn_low  = results(dds_wald, contrast = list(c("leachatelow",  "stageprawnchip.leachatelow"))),
  prawn_mid  = results(dds_wald, contrast = list(c("leachatemid",  "stageprawnchip.leachatemid"))),
  prawn_high = results(dds_wald, contrast = list(c("leachatehigh", "stageprawnchip.leachatehigh"))),

  eg_low  = results(dds_wald, contrast = list(c("leachatelow",  "stageearlygastrula.leachatelow"))),
  eg_mid  = results(dds_wald, contrast = list(c("leachatemid",  "stageearlygastrula.leachatemid"))),
  eg_high = results(dds_wald, contrast = list(c("leachatehigh", "stageearlygastrula.leachatehigh"))),

  # Pure interaction terms (differences of effects vs cleavage)
  int_prwn_low  = results(dds_wald, name = "stageprawnchip.leachatelow"),
  sig_prwn_mid  = results(dds_wald, name = "stageprawnchip.leachatemid"),
  int_prwn_high = results(dds_wald, name = "stageprawnchip.leachatehigh"),
  int_eg_low    = results(dds_wald, name = "stageearlygastrula.leachatelow"),
  int_eg_mid   
  int_eg_high  

  # LRT (e.g., test of interaction block or full vs reduced)
  sig_cat_LRT
)
```


```{r}
# Combine into a summary
int_wald_df <- data.frame(
  gene_id = rownames(res_high_pc),
  pc_h_log2FC = res_high_pc$log2FoldChange,
  pc_high_padj = res_high_pc$padj,
  prawnchip_mid_log2FC = res_mid_pc$log2FoldChange,
  prawnchip_mid_padj = res_mid_pc$padj,
  prawnchip_low_log2FC = res_low_pc$log2FoldChange,
  prawnchip_low_padj = res_low_pc$padj,
  earlygastrula_high_log2FC = res_high_eg$log2FoldChange,
  earlygastrula_high_padj = res_high_eg$padj,
  earlygastrula_mid_log2FC = res_mid_eg$log2FoldChange,
  earlygastrula_mid_padj = res_mid_eg$padj,
  earlygastrula_low_log2FC = res_low_eg$log2FoldChange,
  earlygastrula_low_padj = res_low_eg$padj
)

# Focus on the 44 genes
int_wald_df_sig <- int_wald_df[int_wald_df$gene_id %in% LRT_sig_genes$gene_id, ]
```

### Plot heatmap

```{r}
# Step 1: Order and extract log2FC values
log2fc_cols <- c(
  "prawnchip_low_log2FC", "prawnchip_mid_log2FC", "prawnchip_high_log2FC",
  "earlygastrula_low_log2FC", "earlygastrula_mid_log2FC", "earlygastrula_high_log2FC"
)

mat <- as.matrix(int_wald_df_sig[, log2fc_cols])
rownames(mat) <- int_wald_df_sig$gene_id

# Step 2: Rename columns for clarity
colnames(mat) <- c(
  "Prawnchip: Low", "Prawnchip: Mid", "Prawnchip: High",
  "Early Gastrula: Low", "Early Gastrula: Mid", "Early Gastrula: High"
)

# Step 3: Create annotation for stage/leachate
annotation_col <- data.frame(
  Stage = rep(c("Prawnchip", "Early Gastrula"), each = 3),
  Leachate = rep(c("Low", "Mid", "High"), times = 2)
)
rownames(annotation_col) <- colnames(mat)

# Step 4: Prepare significance (padj < 0.05) annotation matrix
padj_cols <- c(
  "prawnchip_low_padj", "prawnchip_mid_padj", "prawnchip_high_padj",
  "earlygastrula_low_padj", "earlygastrula_mid_padj", "earlygastrula_high_padj"
)
padj_mat <- as.matrix(int_wald_df_sig[, padj_cols])
rownames(padj_mat) <- int_wald_df_sig$gene_id
colnames(padj_mat) <- colnames(mat)

# Significance stars with NA handling
sig_labels <- ifelse(is.na(padj_mat), "",
                      ifelse(padj_mat < 0.01, "**",
                        ifelse(padj_mat < 0.05, "*",
                          ifelse(padj_mat < 0.1, ".", ""))))

# Step 5: Save heatmap with significance stars
png("../../output/06_deg/interactions/int_wald_heatmap.png", width = 10, height = 12, units = "in", res = 600)

pheatmap(mat,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         display_numbers = sig_labels,
         number_color = "black",
         fontsize = 10,
         fontsize_number = 8,
         angle_col = 45,
         border_color = NA)
```

Interpretation notes: \> 🧬 What does the heatmap show? Each cell in the heatmap shows the scaled log2 fold change for a gene in a specific interaction condition. \> \> ✅ Scaling Because you set scale = "row", each gene's values are z-scored across the six interaction conditions \> \> Blue = low relative expression change in that condition \> \> Red = high relative expression change in that condition \> \> The actual color scale is relative within each row, not across the whole matrix. \> \> This helps visualize patterns of expression changes across conditions for each gene, regardless of absolute size. \> \> 📊 What can you learn from this? Are there groups of genes responding similarly? Row clustering may reveal gene modules with similar interaction response profiles. \> \> Which stage + leachate condition drives the interaction? Column patterns (e.g., lots of red in prawnchip_high_log2FC) suggest strong interaction effects in that condition. \> \> Which genes have the strongest differential interactions? Genes with high contrast (strong red/blue) have more dynamic interaction effects. \> \> 🧠 Interpretation Reminder These log2 fold changes are interaction effects --- they reflect how gene expression at a specific `stage` & `leachate` level deviates from the expected additive effect of `stage` and `leachate` alone. \> \> So, a large positive log2FC means: \> \> "At this `stage` and `leachate` level, this gene is more highly expressed than we would expect based on `stage` and `leachate` effects individually." \> \> Likewise, a large negative value = stronger-than-expected suppression.

```{r}
int_wald_df_sig$any_sig <- apply(int_wald_df_sig[, grepl("padj", names(int_wald_df_sig))], 1, function(p) any(p < 0.05, na.rm = TRUE))
```


```{r}
common_genes <- Reduce(intersect, list(
  sig_lvc_cl$gene_id,
  sig_mvc_cl$gene_id,
  sig_hvc_cl$gene_id
))

length(common_genes)  # Number of genes common to all cleavage, just 1!
print(common_genes) # And it is...
```

::: callout-note
Montipora_capitata_HIv3\_\_\_TS.g50653.t2" is the only gene common across cleavage with differential expression due to low, mid, and high leachate treatments
:::

# Gene set overlap: Venn Diagram
## Venn Diagram with ComplexUpset

```{r}
genes1 <- sig_lvc_cl$gene_id
genes2 <- sig_mvc_cl$gene_id
genes3 <- sig_hvc_cl$gene_id
genes4 <- sig_hvc_eg$gene_id
genes5 <- LRT_sig_genes$gene_id
```

Upset requires a binary membership table (presence/absence dataframe)

```{r}
all_genes <- unique(c(genes1, genes2, genes3, genes4, genes5))

upset_data <- data.frame(
  gene_id = all_genes,
  LVC_CL  = all_genes %in% genes1,
  MVC_CL  = all_genes %in% genes2,
  HVC_CL  = all_genes %in% genes3,
  HVC_EG  = all_genes %in% genes4,
  LRT     = all_genes %in% genes5
)
```

```{r}
install.packages("ComplexUpset")
```

```{r}
ComplexUpset::upset(
  upset_data,
    intersect = c("LVC_CL", "MVC_CL", "HVC_CL", "HVC_EG", "LRT"),
    base_annotations = list('Intersection size'= 
                              ComplexUpset::intersection_size(counts = TRUE)))

```

## Gene set overlaps with `UpSetR`

```{r}
# install.packages("UpSetR")  # if needed
library(UpSetR)
sets <- c("LVC_CL","MVC_CL","HVC_CL","HVC_EG","LRT")

df_up <- upset_data
df_up[sets] <- lapply(df_up[sets], function(x) as.integer(as.logical(x)))

UpSetR::upset(df_up, sets = sets, keep.order = TRUE)

```

```{r}
genes1 <- sig_lvc_cl$gene_id
genes2 <- sig_mvc_cl$gene_id
genes3 <- sig_hvc_cl$gene_id
genes4 <- sig_hvc_eg$gene_id
genes5 <- LRT_sig_genes$gene_id
```

Upset requires a binary membership table (presence/absence dataframe)

```{r}
all_genes <- unique(c(genes1, genes2, genes3, genes4, genes5))

upset_data <- data.frame(
  gene_id = all_genes,
  LVC_CL  = all_genes %in% genes1,
  MVC_CL  = all_genes %in% genes2,
  HVC_CL  = all_genes %in% genes3,
  HVC_EG  = all_genes %in% genes4,
  LRT     = all_genes %in% genes5
)
```

```{r}
upset_plot <- upset_data %>% 
  upset(
    intersect = c("LVC_CL", "MVC_CL", "HVC_CL", "HVC_EG", "LRT"),
    base_annotations = list(
      'Intersection size' = intersection_size(
        counts = TRUE
      )
    )
  ) +
  labs(title = "UpSet Plot of Significant Genes (P < 0.05)")

upset_plot
```

```{r}
Wald_sig_genes <- int_wald_df_sig

write_csv(Wald_sig_genes, "../../output/06_deg/interactions/Wald_sig_genes.csv")
```

```{r}
vsd_sig <-  vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  filter(gene_id %in% sig_cat)

write_csv(vsd_sig, "../../output/06_deg/interactions/vsd_sig.csv")
```

```{r}
all_genes <- vsd %>%
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>%
  select(gene_id)

write_csv(all_genes, "../../output/06_deg/interactions/all_genes.csv")
```

::: callout-note
12,565 genes listed in the `all_genes.csv`
:::

# Perform clustering

see Roberts Lab Issue [#1916](https://github.com/RobertsLab/resources/issues/1916)

Perform clustering.

```{r}
clustering_sig_genes <- LRT_sig_genes %>%
  arrange(padj) #genes in rows, stats in columns

matrix <- vsd #genes in rows, sample ID in columns

meta <- metadata_or %>%
  dplyr::select(sample_id, stage, leachate) %>% #sample id and stage, and leachate columns
  column_to_rownames(var = "sample_id")

meta<-meta%>%
  dplyr::select(stage, leachate) #sample ID in rows, one column of lifestage 

# Obtain transformed values for those significant genes
cluster_vst <- matrix[clustering_sig_genes$gene_id, ] #rows are DEGs and columns are sample ID
```

Run with minimum 3 genes per cluster to detect because we have few DEGs.\
- DEGreport vignette, [degPatterns](https://www.bioconductor.org/packages/release/bioc/vignettes/DEGreport/inst/doc/DEGreport.html#:~:text=%22expression%22)-,Detect%20patterns%20of%20expression,-In%20this%20section)

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
stage_clusters <- degPatterns(cluster_vst, metadata = meta, time = "stage", col=NULL, plot=TRUE, minc = 4)
```

```{r}
# Use the `degPatterns` function from the 'DEGreport' package to show gene clusters across sample groups
clusters <- degPatterns(cluster_vst, metadata = meta, time = "leachate", col=NULL, plot=TRUE, minc = 3)
```

```{r}
ggsave(filename="../../output/06_deg/interactions/degPatternscluster_plot.png", width=12, height=10)
```

Extract the list of genes in each cluster for leachate (stage grouped all together in one big group, so we can focus in on leachate with the cluster analysis here).

```{r}
# Rename and extract information 
groups <- clusters$df %>%
          rename(gene_id=genes) %>% 
          rename(leachate_cluster=cluster)
```

Save the gene cluster annotation

```{r}
write.csv(groups, "../../output/06_deg/interactions/clusters.csv", row.names = FALSE)
```

Plot is showing Z score of VST normalized gene counts

```{r, warning=FALSE}
cluster_plot<-ggplot(clusters[["normalized"]],
        aes(x=leachate, y=value)) +
  facet_wrap(~ cluster)+
   geom_point(colour="darkgray", size=1) +
  ylab("Z-score expression")+
  xlab("")+
   # change the method to make it smoother
   geom_smooth(aes(group=genes), method = "loess", colour="red", linewidth=0.1, se=FALSE)+
   geom_smooth(aes(group=1), method = "loess", colour="black")+
  geom_hline(yintercept=0, linetype="dashed")+
  geom_vline(xintercept=3, linetype="dotted")+
  geom_vline(xintercept=6, linetype="dotted")+
  theme_classic()+
  theme(
    axis.text.x = element_text(angle=90, hjust=1, vjust=1)
  );cluster_plot

ggsave(cluster_plot, filename="../../output/06_deg/interactions/cluster_plot.png", width=12, height=10)
```

# Summary & Next Steps

We can perform functional enrichment on these clusters now!
