# Background

## Inputs

## Outputs

## Load packages

```{r}
library(DESeq2)
library(vegan)
library(tidyverse)
```

## Load inputs

```{r, eval=TRUE}
gcm_filtor <- read.csv("../../output/05_explore/gcm_filtor.csv", row.names="gene_id", check.names=FALSE) 
head(gcm_filtor)
```

```{r, eval=TRUE}
# Load metadata (outliers removed)
metadata_or <- read_csv("../../metadata/metadata_or.csv")

# Reorder factors
metadata_or$stage <- as.factor(metadata_or$stage)
metadata_or$stage <- fct_relevel(metadata_or$stage, "cleavage", "prawnchip", "earlygastrula")
 
metadata_or$leachate <- as.factor(metadata_or$leachate)
metadata_or$leachate <- fct_relevel(metadata_or$leachate, "control", "low", "mid", "high")

str(metadata_or)
```

# Make DESeqData Object

```{r}
dds_cat <- DESeqDataSetFromMatrix(countData = gcm_filtor,
                              colData = metadata_or,
                              design = formula( ~0 + stage + leachate + stage:leachate))
```

# Estimate size factors

check that they are all less than 4 (should return TRUE)

```{r}
dds_cat <- estimateSizeFactors(dds_cat)
hist(sizeFactors(dds_cat))
all(sizeFactors(dds_cat))<4
```

# Apply a variance stabilizing transformation (VST)

We apply a variance stabilizing transformation to minimize effects of small counts and normalize for library size

Note: t() transposes. With DESeq2's VST:

t(assay(vst(dds_cat, blind = TRUE))) flips it to samples × genes.

Why it matters:

Functions like vegan::vegdist() and adonis2() expect rows = samples, columns = features.

Use the transposed version for PERMANOVA: rows (samples) will match your meta data frame.

```{r}
# extract variance stabilized count matrix
vst_mat <- assay(vst(dds_cat, blind = TRUE))

# transpose
vst_t_mat <- t(vst_mat)
str(vst_t_mat)
```

# Compute distance between samples

```{r}
eu_dist <- vegdist((vst_t_mat), method = "eu")
```

# Run `adonis2` to build PERMANOVA model

::: callout-note
Because adonis2() uses random permutations to get p-values. Without fixing the seed, each run draws a different set/order of permutations, which results in tiny differences in F-stats/p-values. By adding `set.seed(1)` the permutation stream is reproducible so you (and future you) get the same numbers.
:::

```{r}
adonis2(eu_dist ~ leachate * stage, 
        data = metadata_or,     
        permutations = 999,
        by = "terms")
```

```{r}
adonis2(eu_dist ~ leachate * stage, 
        data = metadata_or,
        permutations = 999,
        strata = metadata_or$parents, # <-- control for parental cross
        by = "terms") 
```

::: callout-note
A PERMANOVA with 999 permutations (permutations restricted within parental cross) revealed that developmental stage explained the majority of variation in multivariate gene-expression profiles ($F_{2,49} = 45.0$, $R^2 = 0.613$, $p = 0.001$). Neither leachate concentration ($F_{3,49} = 1.09$, $R^2 = 0.022$, $p = 0.375$) nor the leachate,×,stage interaction ($F_{6,49} = 0.78$, $R^2 = 0.032$, $p = 0.697$) had a significant effect. Overall, the model explained 66.7% of the variance in gene-expression dissimilarity, with stage accounting for nearly all of the explained variation.

These results indicate that transcriptomic profiles differed strongly among developmental stages but not across leachate concentrations or their interaction, suggesting that developmental processes dominate gene-expression variation in this dataset.
:::

Why does the permanova say `leachate` and `leachate:stage` is insignificant but `DESeq2` results return significantly differentially expressed genes for those terms?

**PERMANOVA** is a **single** multivariate test on a distance among **samples**

It asks whether group **centroids** differ (and a bit of spread, depending on dispersion). If `leachate` shifts are modest, bidirectional, or limited to a subset of genes, the overall centroid shift can be tiny → non-sig.

**Dominant developmental stage signal**

Your term table shows **`stage`** explains \~61% of variance; that can swamp smaller `leachate` signals. After blocking by parents, the residual degrees of freedom and variability left to detect `leachate` are limited.

Your DESeq2 results might be **stage-specific** contrasts (e.g., leachate within stage), while your PERMANOVA tested a **single global** leachate and interaction term. The global test can be ns even if **some stages** have effects.

## Global heatmap

```{r}
pak::pak("mdscheuerell/flexoki") # installs flexoki
library(flexoki) # load the version from mdscheuerell

# Specify colors
# Assign specific colors to each hpf
hpf_colors <- c(
  "4" = flex("cyan", sat=500),
  "9" = flex("green", sat =500),
  "14" = flex("yellow", sat=500)
)

# Assign specific colors to each leachate level
leachate_colors <- c(
  "control" = flex("blue", sat=400),
  "low" = flex("magenta", sat =500),
  "mid" = flex("magenta", sat=700),
  "high" = flex("magenta", sat=950)
)

# Annotation colors for pheatmap
ann_colors = list(
  leachate = leachate_colors,
  hpf = hpf_colors
)

# Indicate leachate level by saturation (alpha)
leachate_alpha_map <- c("control" = 0.5, "low" = 0.6, "mid" = 0.7, "high" = 0.9)
pca_500$alpha_val <- leachate_alpha_map[pca_500$leachate]
```

```{r}
# Extract the annotation dataframe
df <- colData(vst_mat)[ , c("leachate", "hpf")]

legend_names_col = colnames(vst_mat)
names(df)<- c("leachate", "hpf")
rownames(df) <- legend_names_col

# Specify colors
#leachate
# Assign specific colors to each hpf
leachate_colors <- c(
  "control" = flex("blue", sat=400),
  "low" = flex("magenta", sat =500),
  "mid" = flex("magenta", sat=700),
  "high" = flex("magenta", sat=950)
)

ann_colors = list(
  leachate = leachate_colors,
  hpf = hpf_colors
)

levels(df$leachate) 
levels(df$hpf)

```

```{r}
pheatmap(
  vst_mat, 
  cluster_rows=TRUE, 
  show_rownames=FALSE, 
  color=inferno(10), 
  show_colnames=FALSE, 
  fontsize_row=3, 
  scale="row", 
  cluster_cols=TRUE, 
  annotation_col=df, 
  annotation_colors = ann_colors, 
  labels_col=legend_names_col, 
  clustering_distance_rows = "euclidean", 
  clustering_distance_cols = "euclidean", 
  cutree_rows = 3)
```
