---
title: "PERMANOVA of RNA-seq gene counts"
subtitle: "A statistical measure to answer 'are the differences between groups bigger than distances within groups?'"
author: "Sarah Tanja"
date: 2024-12-02
date-modified: today
format:
  gfm: 
    toc: true
    number-sections: true
  html:
    theme: journal
    highlight-style: github
    page-layout: article
    code-background: true
    code-tools: 
      source: true
      toggle: true
    toc: true
    toc-depth: 2
    toc-location: left
    number-sections: true
    df-print: kable
    smooth-scroll: true
    link-external-icon: true
    link-external-newwindow: true
    reference-location: margin
    citation-location: margin
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, # Display code chunks
  eval = TRUE, # Evaluate code chunks
  warning = FALSE, # Hide warnings
  message = FALSE, # Hide messages
  comment = "") # Prevents appending '##' to beginning of lines in code output))        
```

# Background

# Setup

## Inputs

-   metadata `metadata_or`
-   filtered gene count matrix `gcm_filtor`
-   list of all significantly expressed genes `sig_all`

## Outputs

## Set paths
```{r}
metadata_path <- "../../metadata/metadata_or.csv"
output_path <- "../../output/06_deg/permanova/"
input_path <- "../../output/06_deg/spawnnight/"
```

## Load packages

```{r}
library(DESeq2)
library(vegan)
library(tidyverse)
```

## Load inputs

```{r}

files <- file.path(input_path, c("vst_mat_df.csv", "vst_mat_130_df.csv"))

nm <- make.names(tools::file_path_sans_ext(basename(files)), unique = TRUE)

for (i in seq_along(files)) {
  assign(nm[i], readr::read_csv(files[i]), envir = .GlobalEnv)
}
```

### Metadata
```{r, eval=TRUE}
# Load metadata (outlier samples removed, 61 samples remain)
metadata_or <- read_csv(metadata_path)

# set factors
metadata_or <- metadata_or %>% 
  mutate(
    collection_date = as.Date(collection_date, format = "%d-%b-%Y"),
    stage    = factor(stage,    levels = c("cleavage", "prawnchip", "earlygastrula")),
    leachate = factor(leachate, levels = c("control", "low", "mid", "high")),
    spawn_night = factor(spawn_night, levels  = c("July_6th", "July_7th", "July_8th"))
    )

# View metadata structure
str(metadata_or)
```

# Data prep

### Turn the data.frames back to matrices

```{r}
vst_mat <- vst_mat_df %>% 
  column_to_rownames(var = "gene_id")

vst_mat_130 <- vst_mat_130_df %>% 
  column_to_rownames(var = "gene_id")
```


### Transpose count matrices

The variance stabilizing transformation (vst) minimized effects of small counts and normalized for library size.
Note: `t()` transposes. The code `t()` flips the matrix to samples as rows × genes as columns. Functions like `vegan::vegdist()` and `adonis2()` expect rows = samples, columns = features. Use the transposed matrix for PERMANOVA such that rows are samples ordered to match the sample names in our metadata. 

```{r}
# transpose
## Full set of genes
t_vst_mat <- t(vst_mat) # rows are samples, columns are genes
dim(t_vst_mat)

## DEGs only
t_vst_mat_130 <- t(vst_mat_130) # rows are samples, columns are genes
dim(t_vst_mat_130)
```

### Compute Euclidean distance between samples

```{r}
eu_dist_full <- vegdist((t_vst_mat), method = "eu")

eu_dist_deg <- vegdist((t_vst_mat_130), method = "eu")
```

# Run PERMANOVA

Run `adonis2` to build PERMANOVA model

::: callout-note
Because adonis2() uses random permutations to get p-values. Without fixing the seed, each run draws a different set/order of permutations, which results in tiny differences in F-stats/p-values. By adding `set.seed(###)` the permutation stream is reproducible so you (and future you) get the same numbers.
:::

## Full gene set
```{r}
set.seed(11052025)
adonis2(eu_dist_full ~ leachate * stage, 
        data = metadata_or,
        permutations = 999,
        strata = metadata_or$spawn_night, # <-- control for spawn night
        by = "terms") 
```

::: callout-note
A PERMANOVA with 999 permutations (permutations restricted within spawn night) revealed that developmental stage explained the majority of variation in multivariate gene-expression profiles ($F_{2,49} = 45.0$, $R^2 = 0.613$, $p = 0.001$). Neither leachate concentration ($F_{3,49} = 1.09$, $R^2 = 0.022$, $p = 0.375$) nor the leachate,×,stage interaction ($F_{6,49} = 0.78$, $R^2 = 0.032$, $p = 0.697$) had a significant effect. Overall, the model explained 66.7% of the variance in gene-expression dissimilarity, with stage accounting for nearly all of the explained variation.

These results indicate that transcriptomic profiles differed strongly among developmental stages but not across leachate concentrations or their interaction, suggesting that developmental processes dominate gene-expression variation in this dataset.
:::

Why does the permanova say `leachate` and `leachate:stage` is insignificant but `DESeq2` results return significantly differentially expressed genes for those terms?

**PERMANOVA** is a **single** multivariate test on a distance among **samples**

It asks whether group **centroids** differ (and a bit of spread, depending on dispersion). If `leachate` shifts are modest, bidirectional, or limited to a subset of genes, the overall centroid shift can be tiny → non-sig.

**Dominant developmental stage signal**

Your term table shows **`stage`** explains \~61% of variance; that can swamp smaller `leachate` signals. After blocking by parents, the residual degrees of freedom and variability left to detect `leachate` are limited.

Your DESeq2 results might be **stage-specific** contrasts (e.g., leachate within stage), while your PERMANOVA tested a **single global** leachate and interaction term. The global test can be ns even if **some stages** have effects.

## DEGs

```{r}
adonis2(eu_dist_deg ~ leachate * stage, 
        data = metadata_or,
        permutations = 999,
        strata = metadata_or$spawn_night, # <-- control for spawn_night
        by = "terms") 
```

::: callout-note
Using PERMANOVA on Euclidean distances calculated from the 130 differentially expressed genes (DEGs) (n = 61 samples; 999 permutations; stratified by parental cross), we detected a strong effect of developmental stage and smaller but significant effects of leachate and their interaction. Stage explained R² = 0.680 of the variance and was highly significant (Df = 2, F = 76.83, p = 0.001). Leachate explained R² = 0.039 (Df = 3, F = 2.96, p = 0.034), and the leachate × stage interaction explained R² = 0.063 (Df = 6, F = 2.38, p = 0.023). Residual variation accounted for R² = 0.217.
:::

# PCAs
## Full gene set
## DEGs